### 2025-02-25 paper/calculus.tex
cleanup main.tex

\section{The Calculus}
\label{sec:calculus}

\begin{table}
    \caption{Syntax of the calculus of pure sessions.}
    \label{tab:syntax}
    \begin{math}
        \displaystyle
        \begin{array}[t]{rcll}
            P, Q & ::= & \Link\x\y \\
                & | & \Close\x \\
                & | & \Fail\x \\
                & | & \Wait\x.P \\
                & | & \Select\x{\InTag_i}\z.P \\
                & | & \Case\x\z{P}{Q} \\
                & | & \Fork\x\y\z{P}{Q} \\
                & | & \Join\x\y\z.P \\
                & | & \Cut\x{P}{Q} \\
        \end{array}
    \end{math}
\end{table}

The calculus of pure sessions that is formalized in our development is closely
related to the linear fragment of GV~\citep{Wadler14}, without shared channels.
Its syntax makes use of an infinite set of \emph{channel names}, ranged over by
$x$, $y$ and $z$, and is shown in \cref{tab:syntax}.

\[
    \begin{array}{rrcll}
      \SLink &
      \Link\x\y & \pcong & \Link\y\x \\
      \SComm &
      \Cut\x{P}{Q} & \pcong & \Cut\x{Q}{P} \\
      \SWait &
      \Cut\x{\Wait\y.P}{Q} & \pcong & \Wait\y.\Cut\x{P}{Q} & x \ne y \\
      \SFail &
      \Cut\x{\Fail\y}{P} & \pcong & \Fail\y & x \ne y \\
      \SSelect &
      \Cut\x{\Select\y{\InTag_i}\z.P}{Q} & \pcong & \Select\y{\InTag_i}\z.\Cut\x{P}{Q} & x \ne y,z \\
      \SCase &
      \Cut\x{\Case\y\z{P}{Q}}{R} & \pcong & \Case\y\z{\Cut\x{P}{R}}{\Cut\x{Q}{R}} & x \ne y,z \\
      \SForkL &
      \Cut\x{\Fork\y\u\v{P}{Q}}{R} & \pcong & \Fork\y\u\v{\Cut\x{P}{R}}{Q} & x \in \fn{P}\setminus\set{y,u,v} \\
      \SForkR &
      \Cut\x{\Fork\y\u\v{P}{Q}}{R} & \pcong & \Fork\y\u\v{P}{\Cut\x{Q}{R}} & x \in \fn{Q}\setminus\set{y,u,v} \\
      \SJoin &
      \Cut\x{\Join\y\u\v.P}{Q} & \pcong & \Join\y\u\v.\Cut\x{P}{Q} & x \ne y,u,v \\
      \\
      \RLink &
      \Cut\x{\Link\x\y}{P} & \red & P\subst\y\x \\
      \RFail &
      \Cut\x{\Fail\y}{P} & \red & \Fail\y & x \ne y \\
      \RClose &
      \Cut\x{\Close\x}{\Wait\x.P} & \red & P \\
      \RSelect &
      \Cut\x{\Select\x{\Tag_i}\y.P}{\Case\x\y{Q_1}{Q_2}} & \red & \Cut\y{P}{Q_i} & i\in\set{1,2} \\
      \RFork &
      \Cut\x{\Fork\x\y\z{P}{Q}}{\Join\x\y\z.R} & \red & \Cut\y{P}{\Cut\z{Q}{R}} \\
      \RCut &
      \Cut\x{P}{R} & \red & \Cut\x{Q}{R} & P \red Q \\
      \RCong &
      P & \red & Q & P \pcong R \red Q \\
    \end{array}
\]

### 2025-02-25 paper/main.tex
cleanup main.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%=======================================================%%
%% to print line numbers in the margin use lineno option %%
%%=======================================================%%

\documentclass[pdflatex,sn-basic]{sn-jnl}

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{amsthm}
\usepackage{cleveref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{agda}
\usepackage{xspace}

\input{macros}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}

\theoremstyle{remark}
\newtheorem{remark}{Remark}

\raggedbottom
%%\unnumbered% uncomment this for unnumbered level heads

\begin{document}

\title{A Logical Solution of the Linearity Challenge}

%%=============================================================%%
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% \author*[1,2]{\fnm{Joergen W.} \spfx{van der} \sur{Ploeg} 
%%  \sfx{IV}}\email{iauthor@gmail.com}
%%=============================================================%%

\author*[1]{\fnm{Luca} \sur{Padovani}}\email{luca.padovani2@unibo.it}

\author[2]{\fnm{Claudia} \sur{Raffaelli}}\email{xxx@yyy.zzz}
\equalcont{These authors contributed equally to this work.}

\affil*[1]{\orgdiv{Department of Computer Science and Engineering}, \orgname{Università di Bologna}, \orgaddress{\street{Mura Anteo Zamboni 7}, \city{Bologna}, \postcode{40138}, \state{BO}, \country{Italy}}}

\affil[2]{\orgdiv{Computer Science Division}, \orgname{Università di Camerino}, \orgaddress{\street{via Madonna delle Carceri 7}, \city{Camerino}, \postcode{62032}, \state{MC}, \country{Italy}}}

\abstract{blah}

\keywords{keyword1, Keyword2, Keyword3, Keyword4}

\maketitle

\section{Introduction}
\label{sec:introduction}

\input{calculus}

\bibliography{main}

\end{document}

Introduction section, of referenced text \cite{bib1} expands on the background
of the work (some overlap with the Abstract is acceptable). The introduction
should not include subheadings.

Springer Nature does not impose a strict layout as standard however authors are
advised to check the individual requirements for the journal they are planning
to submit to as there may be journal-level preferences. When preparing your text
please also be aware that some stylistic choices are not supported in full text
XML (publication version), including coloured font. These will not be replicated
in the typeset article if it is accepted.

\section{Results}\label{sec2}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.

\section{This is an example for first level head---section head}\label{sec3}

\subsection{This is an example for second level head---subsection head}\label{subsec2}

\subsubsection{This is an example for third level head---subsubsection head}\label{subsubsec2}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.

\section{Equations}\label{sec4}

Equations in \LaTeX\ can either be inline or on-a-line by itself (``display equations''). For
inline equations use the \verb+$...$+ commands. E.g.: The equation
$H\psi = E \psi$ is written via the command \verb+$H \psi = E \psi$+.

For display equations (with auto generated equation numbers)
one can use the equation or align environments:
\begin{equation}
\|\tilde{X}(k)\|^2 \leq\frac{\sum\limits_{i=1}^{p}\left\|\tilde{Y}_i(k)\right\|^2+\sum\limits_{j=1}^{q}\left\|\tilde{Z}_j(k)\right\|^2 }{p+q}.\label{eq1}
\end{equation}
where,
\begin{align}
D_\mu &=  \partial_\mu - ig \frac{\lambda^a}{2} A^a_\mu \nonumber \\
F^a_{\mu\nu} &= \partial_\mu A^a_\nu - \partial_\nu A^a_\mu + g f^{abc} A^b_\mu A^a_\nu \label{eq2}
\end{align}
Notice the use of \verb+\nonumber+ in the align environment at the end
of each line, except the last, so as not to produce equation numbers on
lines where no equation numbers are required. The \verb+\label{}+ command
should only be used at the last line of an align environment where
\verb+\nonumber+ is not used.
\begin{equation}
Y_\infty = \left( \frac{m}{\textrm{GeV}} \right)^{-3}
    \left[ 1 + \frac{3 \ln(m/\textrm{GeV})}{15}
    + \frac{\ln(c_2/5)}{15} \right]
\end{equation}
The class file also supports the use of \verb+\mathbb{}+, \verb+\mathscr{}+ and
\verb+\mathcal{}+ commands. As such \verb+\mathbb{R}+, \verb+\mathscr{R}+
and \verb+\mathcal{R}+ produces $\mathbb{R}$, $\mathscr{R}$ and $\mathcal{R}$
respectively (refer Subsubsection~\ref{subsubsec2}).

\section{Tables}\label{sec5}

Tables can be inserted via the normal table and tabular environment. To put
footnotes inside tables you should use \verb+\footnotetext[]{...}+ tag.
The footnote appears just below the table itself (refer Tables~\ref{tab1} and \ref{tab2}). 
For the corresponding footnotemark use \verb+\footnotemark[...]+

\begin{table}[h]
\caption{Caption text}\label{tab1}%
\begin{tabular}{@{}llll@{}}
\toprule
Column 1 & Column 2  & Column 3 & Column 4\\
\midrule
row 1    & data 1   & data 2  & data 3  \\
row 2    & data 4   & data 5\footnotemark[1]  & data 6  \\
row 3    & data 7   & data 8  & data 9\footnotemark[2]  \\
\botrule
\end{tabular}
\footnotetext{Source: This is an example of table footnote. This is an example of table footnote.}
\footnotetext[1]{Example for a first table footnote. This is an example of table footnote.}
\footnotetext[2]{Example for a second table footnote. This is an example of table footnote.}
\end{table}

\noindent
The input format for the above table is as follows:

%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. Please ignore this.       %%
%%=============================================%%
\bigskip
\begin{verbatim}
\begin{table}[<placement-specifier>]
\caption{<table-caption>}\label{<table-label>}%
\begin{tabular}{@{}llll@{}}
\toprule
Column 1 & Column 2 & Column 3 & Column 4\\
\midrule
row 1 & data 1 & data 2	 & data 3 \\
row 2 & data 4 & data 5\footnotemark[1] & data 6 \\
row 3 & data 7 & data 8	 & data 9\footnotemark[2]\\
\botrule
\end{tabular}
\footnotetext{Source: This is an example of table footnote. 
This is an example of table footnote.}
\footnotetext[1]{Example for a first table footnote.
This is an example of table footnote.}
\footnotetext[2]{Example for a second table footnote. 
This is an example of table footnote.}
\end{table}
\end{verbatim}
\bigskip
%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. Please ignore this.       %%
%%=============================================%%

In case of double column layout, tables which do not fit in single column width should be set to full text width. For this, you need to use \verb+\begin{table*}+ \verb+...+ \verb+\end{table*}+ instead of \verb+\begin{table}+ \verb+...+ \verb+\end{table}+ environment. Lengthy tables which do not fit in textwidth should be set as rotated table. For this, you need to use \verb+\begin{sidewaystable}+ \verb+...+ \verb+\end{sidewaystable}+ instead of \verb+\begin{table*}+ \verb+...+ \verb+\end{table*}+ environment. This environment puts tables rotated to single column width. For tables rotated to double column width, use \verb+\begin{sidewaystable*}+ \verb+...+ \verb+\end{sidewaystable*}+.

\section{Figures}\label{sec6}

As per the \LaTeX\ standards you need to use eps images for \LaTeX\ compilation and \verb+pdf/jpg/png+ images for \verb+PDFLaTeX+ compilation. This is one of the major difference between \LaTeX\ and \verb+PDFLaTeX+. Each image should be from a single input .eps/vector image file. Avoid using subfigures. The command for inserting images for \LaTeX\ and \verb+PDFLaTeX+ can be generalized. The package used to insert images in \verb+LaTeX/PDFLaTeX+ is the graphicx package. Figures can be inserted via the normal figure environment as shown in the below example:

In case of double column layout, the above format puts figure captions/images to single column width. To get spanned images, we need to provide \verb+\begin{figure*}+ \verb+...+ \verb+\end{figure*}+.

For sample purpose, we have included the width of images in the optional argument of \verb+\includegraphics+ tag. Please ignore this.

\section{Algorithms, Program codes and Listings}\label{sec7}

Packages \verb+algorithm+, \verb+algorithmicx+ and \verb+algpseudocode+ are used for setting algorithms in \LaTeX\ using the format:

%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. Please ignore this.       %%
%%=============================================%%

You may refer above listed package documentations for more details before setting \verb+algorithm+ environment. For program codes, the ``verbatim'' package is required and the command to be used is \verb+\begin{verbatim}+ \verb+...+ \verb+\end{verbatim}+.

Similarly, for \verb+listings+, use the \verb+listings+ package. \verb+\begin{lstlisting}+ \verb+...+ \verb+\end{lstlisting}+ is used to set environments similar to \verb+verbatim+ environment. Refer to the \verb+lstlisting+ package documentation for more details.

A fast exponentiation procedure:

\lstset{texcl=true,basicstyle=\small\sf,commentstyle=\small\rm,mathescape=true,escapeinside={(*}{*)}}
\begin{lstlisting}
begin
  for $i:=1$ to $10$ step $1$ do
      expt($2,i$);  
      newline() od                (*\textrm{Comments will be set flush to the right margin}*)
where
proc expt($x,n$) $\equiv$
  $z:=1$;
  do if $n=0$ then exit fi;
     do if odd($n$) then exit fi;                 
        comment: (*\textrm{This is a comment statement;}*)
        $n:=n/2$; $x:=x*x$ od;
     { $n>0$ };
     $n:=n-1$; $z:=z*x$ od;
  print($z$). 
end
\end{lstlisting}

%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. Please ignore this.       %%
%%=============================================%%
\bigskip
\begin{minipage}{\hsize}%
\lstset{frame=single,framexleftmargin=-1pt,framexrightmargin=-17pt,framesep=12pt,linewidth=0.98\textwidth,language=pascal}% Set your language (you can change the language for each code-block optionally)
%%% Start your code-block
\begin{lstlisting}
for i:=maxint to 0 do
begin
{ do nothing }
end;
Write('Case insensitive ');
Write('Pascal keywords.');
\end{lstlisting}
\end{minipage}

\section{Cross referencing}\label{sec8}

Environments such as figure, table, equation and align can have a label
declared via the \verb+\label{#label}+ command. For figures and table
environments use the \verb+\label{}+ command inside or just
below the \verb+\caption{}+ command. You can then use the
\verb+\ref{#label}+ command to cross-reference them. As an example, consider
the label declared for Figure~\ref{fig1} which is
\verb+\label{fig1}+. To cross-reference it, use the command 
\verb+Figure \ref{fig1}+, for which it comes up as
``Figure~\ref{fig1}''.

To reference line numbers in an algorithm, consider the label declared for the line number 2 of Algorithm~\ref{algo1} is \verb+\label{algln2}+. To cross-reference it, use the command \verb+\ref{algln2}+ for which it comes up as line~\ref{algln2} of Algorithm~\ref{algo1}.

\subsection{Details on reference citations}\label{subsec7}

Standard \LaTeX\ permits only numerical citations. To support both numerical and author-year citations this template uses \verb+natbib+ \LaTeX\ package. For style guidance please refer to the template user manual.

Here is an example for \verb+\cite{...}+: \cite{bib1}. Another example for \verb+\citep{...}+: \citep{bib2}. For author-year citation mode, \verb+\cite{...}+ prints Jones et al. (1990) and \verb+\citep{...}+ prints (Jones et al., 1990).

All cited bib entries are printed at the end of this article: \cite{bib3}, \cite{bib4}, \cite{bib5}, \cite{bib6}, \cite{bib7}, \cite{bib8}, \cite{bib9}, \cite{bib10}, \cite{bib11}, \cite{bib12} and \cite{bib13}.

\section{Examples for theorem like environments}\label{sec10}

For theorem like environments, we require \verb+amsthm+ package. There are three types of predefined theorem styles exists---\verb+thmstyleone+, \verb+thmstyletwo+ and \verb+thmstylethree+

%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. Please ignore this.       %%
%%=============================================%%
\bigskip
\begin{tabular}{|l|p{19pc}|}
\hline
\verb+thmstyleone+ & Numbered, theorem head in bold font and theorem text in italic style \\\hline
\verb+thmstyletwo+ & Numbered, theorem head in roman font and theorem text in italic style \\\hline
\verb+thmstylethree+ & Numbered, theorem head in bold font and theorem text in roman style \\\hline
\end{tabular}
\bigskip
%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. Please ignore this.       %%
%%=============================================%%

For mathematics journals, theorem styles can be included as shown in the following examples:

\begin{theorem}[Theorem subhead]\label{thm1}
Example theorem text. Example theorem text. Example theorem text. Example theorem text. Example theorem text. 
Example theorem text. Example theorem text. Example theorem text. Example theorem text. Example theorem text. 
Example theorem text. 
\end{theorem}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.

\begin{proposition}
Example proposition text. Example proposition text. Example proposition text. Example proposition text. Example proposition text. 
Example proposition text. Example proposition text. Example proposition text. Example proposition text. Example proposition text. 
\end{proposition}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.

\begin{example}
Phasellus adipiscing semper elit. Proin fermentum massa
ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem. 
\end{example}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.

\begin{remark}
Phasellus adipiscing semper elit. Proin fermentum massa
ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem. 
\end{remark}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.

\begin{definition}[Definition sub head]
Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. 
\end{definition}

Additionally a predefined ``proof'' environment is available: \verb+\begin{proof}+ \verb+...+ \verb+\end{proof}+. This prints a ``Proof'' head in italic font style and the ``body text'' in roman font style with an open square at the end of each proof environment.

\begin{proof}
Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. 
\end{proof}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.

\begin{proof}[Proof of Theorem~{\upshape\ref{thm1}}]
Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. 
\end{proof}

\noindent
For a quote environment, use \verb+\begin{quote}...\end{quote}+
\begin{quote}
Quoted text example. Aliquam porttitor quam a lacus. Praesent vel arcu ut tortor cursus volutpat. In vitae pede quis diam bibendum placerat. Fusce elementum
convallis neque. Sed dolor orci, scelerisque ac, dapibus nec, ultricies ut, mi. Duis nec dui quis leo sagittis commodo.
\end{quote}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text (refer Figure~\ref{fig1}). Sample body text. Sample body text. Sample body text (refer Table~\ref{tab3}).

\section{Methods}\label{sec11}

Topical subheadings are allowed. Authors must ensure that their Methods section includes adequate experimental and characterization data necessary for others in the field to reproduce their work. Authors are encouraged to include RIIDs where appropriate.

\textbf{Ethical approval declarations} (only required where applicable) Any article reporting experiment/s carried out on (i)~live vertebrate (or higher invertebrates), (ii)~humans or (iii)~human samples must include an unambiguous statement within the methods section that meets the following requirements:

\begin{enumerate}[1.]
\item Approval: a statement which confirms that all experimental protocols were approved by a named institutional and/or licensing committee. Please identify the approving body in the methods section

\item Accordance: a statement explicitly saying that the methods were carried out in accordance with the relevant guidelines and regulations

\item Informed consent (for experiments involving humans or human tissue samples): include a statement confirming that informed consent was obtained from all participants and/or their legal guardian/s
\end{enumerate}

If your manuscript includes potentially identifying patient/participant information, or if it describes human transplantation research, or if it reports results of a clinical trial then  additional information will be required. Please visit (\url{https://www.nature.com/nature-research/editorial-policies}) for Nature Portfolio journals, (\url{https://www.springer.com/gp/authors-editors/journal-author/journal-author-helpdesk/publishing-ethics/14214}) for Springer Nature journals, or (\url{https://www.biomedcentral.com/getpublished/editorial-policies\#ethics+and+consent}) for BMC.

\section{Discussion}\label{sec12}

Discussions should be brief and focused. In some disciplines use of Discussion or `Conclusion' is interchangeable. It is not mandatory to use both. Some journals prefer a section `Results and Discussion' followed by a section `Conclusion'. Please refer to Journal-level guidance for any specific requirements.

\section{Conclusion}\label{sec13}

Conclusions may be used to restate your hypothesis or research question, restate your major findings, explain the relevance and the added value of your work, highlight any limitations of your study, describe future directions for research and recommendations.

In some disciplines use of Discussion or 'Conclusion' is interchangeable. It is not mandatory to use both. Please refer to Journal-level guidance for any specific requirements.

\backmatter

\bmhead{Supplementary information}

If your article has accompanying supplementary file/s please state so here.

Authors reporting data from electrophoretic gels and blots should supply the full unprocessed scans for key as part of their Supplementary information. This may be requested by the editorial team/s if it is missing.

Please refer to Journal-level guidance for any specific requirements.

\bmhead{Acknowledgements}

Acknowledgements are not compulsory. Where included they should be brief. Grant or contribution numbers may be acknowledged.

Please refer to Journal-level guidance for any specific requirements.

\section*{Declarations}

Some journals require declarations to be submitted in a standardised format. Please check the Instructions for Authors of the journal to which you are submitting to see if you need to complete this section. If yes, your manuscript must contain the following sections under the heading `Declarations':

\begin{itemize}
\item Funding
\item Conflict of interest/Competing interests (check journal-specific guidelines for which heading to use)
\item Ethics approval and consent to participate
\item Consent for publication
\item Data availability 
\item Materials availability
\item Code availability 
\item Author contribution
\end{itemize}

\noindent
If any of the sections are not relevant to your manuscript, please include the heading and write `Not applicable' for that section.

%%===================================================%%
%% For presentation purpose, we have included        %%
%% \bigskip command. Please ignore this.             %%
%%===================================================%%
\bigskip
\begin{flushleft}%
Editorial Policies for:

\bigskip\noindent
Springer journals and proceedings: \url{https://www.springer.com/gp/editorial-policies}

\bigskip\noindent
Nature Portfolio journals: \url{https://www.nature.com/nature-research/editorial-policies}

\bigskip\noindent
\textit{Scientific Reports}: \url{https://www.nature.com/srep/journal-policies/editorial-policies}

\bigskip\noindent
BMC journals: \url{https://www.biomedcentral.com/getpublished/editorial-policies}
\end{flushleft}

\begin{appendices}

\section{Section title of first appendix}\label{secA1}

An appendix contains supplementary information that is not an essential part of the text itself but which may be helpful in providing a more comprehensive understanding of the research problem or it is information that is too cumbersome to be included in the body of the paper.

%%=============================================%%
%% For submissions to Nature Portfolio Journals %%
%% please use the heading ``Extended Data''.   %%
%%=============================================%%

%%=============================================================%%
%% Sample for another appendix section			       %%
%%=============================================================%%

%% \section{Example of another appendix section}\label{secA2}%
%% Appendices may be used for helpful, supporting or essential material that would otherwise 
%% clutter, break up or be distracting to the text. Appendices can consist of sections, figures, 
%% tables and equations etc.

\end{appendices}

%%===========================================================================================%%
%% If you are submitting to one of the Nature Portfolio journals, using the eJP submission   %%
%% system, please include the references within the manuscript file itself. You may do this  %%
%% by copying the reference list from your .bbl file, paste it into the main manuscript .tex %%
%% file, and delete the associated \verb+\bibliography+ commands.                            %%
%%===========================================================================================%%

\bibliography{sn-bibliography}% common bib file
%% if required, the content of .bbl file can be included here once bbl is generated
%%\input sn-article.bbl

\end{document}

### 2025-06-12 paper/calculus.tex
snapshot

@@
 \begin{table}
-    \caption{Syntax of the calculus of pure sessions.}
+    \caption{Syntax of \Calculus.}
     \label{tab:syntax}
@@
 The calculus of pure sessions that is formalized in our development is closely
-related to the linear fragment of GV~\citep{Wadler14}, without shared channels.
-Its syntax makes use of an infinite set of \emph{channel names}, ranged over by
-$x$, $y$ and $z$, and is shown in \cref{tab:syntax}.
+related to the multiplicative, additive fragment of CP~\citep{Wadler14}, without
+shared channels. Its syntax makes use of an infinite set of \emph{channel
+names}, ranged over by $x$, $y$ and $z$, and is shown in \cref{tab:syntax}.
+%
+A link $\Link\x\y$ denotes the merging of the $x$ and $y$ channels, so that
+every message sent on one of the channels is forwarded to the other. This form
+is useful for modeling \emph{free outputs}, as we will see later.\Luca{Mettere
+esempio!}
+%
+The $\Close\x$ and $\Wait\x.P$ processes respectively model the output and input of a termination signal on the channel $x$.
+The former process has no continuation, while the latter process continues as $P$ once the session has been terminated.
+%
+The $\Fail\x$ process denotes a runtime error concerning the session $x$.
+%
+The $\Select\x{\InTag_i}\z.P$ and $\Case\x\z{Q_1}{Q_2}$ processes respectively
+denote the output and input of a label $\InTag_i$ along with a fresh
+continuation $z$ on the channel $x$. Depending on the label $\InTag_i$, the
+receiver continues as $Q_i$. In practice it is useful to consider arbitrary
+(domain-specific) sets of tags instead of the canonical two $\InTag_1$ and
+$\InTag_2$. This generalization has no substantial impact on the formalization,
+so we stick to the classical case of ``binary choices'' in line with the usual
+presentations of linear logic.
+%
+The $\Fork\x\y\z{P}{Q}$ and $\Join\x\y\z.R$ processes respectively denote the
+output and input of a pair of fresh channels $y$ and $z$ on the channel $x$.
+These forms are often used in conjunction with links to model \emph{free
+outputs}, but can also be used to model bifurcating sessions whereby a
+sequential protocol continues as two parallel sub-protocols.
 
+\begin{table}
+    \caption{Operational semantics of \Calculus.}
     \[
@@
     \]
+\end{table}

### 2025-06-16 paper/calculus.tex
semantics

@@
 %
-The $\Close\x$ and $\Wait\x.P$ processes respectively model the output and input of a termination signal on the channel $x$.
-The former process has no continuation, while the latter process continues as $P$ once the session has been terminated.
+The $\Close\x$ and $\Wait\x.P$ processes respectively model the output and input
+of a termination signal on the channel $x$. The former process has no
+continuation, while the latter process continues as $P$ once the session has
+been terminated.
 %
@@
 
+free and bound names
+
+we consideer processes equal up to renaming of bound names
+
 \begin{table}
     \caption{Operational semantics of \Calculus.}
+    \label{tab:semantics}
     \[
@@
         \Cut\x{P}{Q} & \pcong & \Cut\x{Q}{P} \\
+        \SFail &
+        \Cut\x{\Fail\y}{P} & \pcong & \Fail\y & x \ne y \\
         \SWait &
         \Cut\x{\Wait\y.P}{Q} & \pcong & \Wait\y.\Cut\x{P}{Q} & x \ne y \\
-        \SFail &
-        \Cut\x{\Fail\y}{P} & \pcong & \Fail\y & x \ne y \\
         \SSelect &
@@
         \Cut\x{\Link\x\y}{P} & \red & P\subst\y\x \\
-        \RFail &
-        \Cut\x{\Fail\y}{P} & \red & \Fail\y & x \ne y \\
+        % \RFail &
+        % \Cut\x{\Fail\y}{P} & \red & \Fail\y & x \ne y \\
         \RClose &
@@
         \RSelect &
-        \Cut\x{\Select\x{\Tag_i}\y.P}{\Case\x\y{Q_1}{Q_2}} & \red & \Cut\y{P}{Q_i} & i\in\set{1,2} \\
+        \Cut\x{\Select\x{\Tag_i}\z.P}{\Case\x\z{Q_1}{Q_2}} & \red & \Cut\z{P}{Q_i} & i\in\set{1,2} \\
         \RFork &
@@
 \end{table}
+
+The operational semantics of \Calculus is shown in \cref{tab:semantics} and is
+given by two relations: a \emph{structural pre-congruence} relation $\pcong$ and
+a \emph{reduction} relation $\red$. Structural pre-congruence relates processes
+that we consider to be essentially indistinguishable. Reduction relates
+processes in which a communication occurs. In some presentations of linear
+logic~\cite{Doumane}, these relations are respectively dubbed \emph{external}
+and \emph{internal} reductions. Let us describe the two relations more in
+detail.
+
+Structural pre-congruence is the least pre-congruence defined by rule
+\rulename{s-*} rules.
+%
+\SLink and \SComm assert that links and parallel compositions are symmetric. The
+remaining rules, when read from left to right, push a cut on $x$ underneath the
+topmost prefix on $y$ of one of its sub-processes when $x \ne y$. These rules
+are key to float input/output actions to the top-level of a process, so that
+they can interact with corresponding complementary actions in the context
+surrounding the process.
+%
+Note that some rules have additional side-conditions preventing bound names to
+become free.\Luca{Non proprio, sono tutti nomi bound.} Also, there are two
+versions of \SForkL and \SForkR depending on which of the two continuations $P$
+and $Q$ contains occurrences of the restricted channel $x$.
+%
+Notably and somewhat surprisingly, there is no pre-congruence rule asserting
+that parallel composition is associative. While this is a sound and expected
+property of parallel composition, it turns out to be inessential for proving all
+of the properties of \Calculus that we consider. We leave it out in light of the
+forthcoming formalization of the calculus, in which every simplication is
+welcome.
+
+The reduction rules \RLink, \RClose, \RSelect and \RFork are in one-to-one
+correspondence with the well-known reductions of session calculis and the cut
+reductions of linear logic. \RLink eliminates a link $\Link\x\y$ by unifying $x$
+and $y$. \RClose closes a session $x$ when a termination signal is exchanged on
+$x$. \RSelect and \RFork respectively model the communication of a tag $\Tag_i$
+and of a fresh channel $y$. Since channels are linear in \Calculus, each
+non-terminating communication carries along a \emph{fresh continuation} $z$ on
+which the rest of the conversation may take place, in line with the semantics of
+the linear $\pi$-calculus~\cite{KobayashiPierceTurner99}.
+%
+The rule~\RCut propagates reductions through cuts and the rules \RCong closes
+reduction by structural pre-congruence.

### 2025-06-19 paper/calculus.tex
semantics and typing

@@
-\section{The Calculus}
+\section{\Calculus}
 \label{sec:calculus}
+
+\subsection{Syntax}
+\label{sec:syntax}
 
@@
             P, Q & ::= & \Link\x\y \\
+                & | & \Fail\x \\
                 & | & \Close\x \\
-                & | & \Fail\x \\
                 & | & \Wait\x.P \\
@@
 
-free and bound names
+The notions of free and bound names are fairly standard. Just bear in mind that
+all action prefixes bind continuation channels in (some) continuation processes.
+In particular, $\Select\x{\Tag_i}\z.P$ binds $z$ in $P$, $\Case\x\z{P}{Q}$ binds
+$z$ in both $P$ and $Q$, $\Fork\x\y\z{P}{Q}$ binds $y$ in $P$ but not in $Q$ and
+$z$ in $Q$ but not in $P$, and $\Join\x\y\z.P$ binds both $y$ and $z$ in $P$.
+%
+We write $\fn{P}$ for the set of channel names occurring free in $P$ and we
+identify processes up to renaming of bound names.
 
-we consideer processes equal up to renaming of bound names
+\subsection{Operational Semantics}
+\label{sec:semantics}
 
@@
 of the properties of \Calculus that we consider. We leave it out in light of the
-forthcoming formalization of the calculus, in which every simplication is
-welcome.
+forthcoming formalization of the calculus, in which every simplification is
+welcome, although standard presentations of classical linear logic admit such
+reduction~\cite{Doumane17}.
 
 The reduction rules \RLink, \RClose, \RSelect and \RFork are in one-to-one
-correspondence with the well-known reductions of session calculis and the cut
+correspondence with the well-known reductions of session calculi and the cut
 reductions of linear logic. \RLink eliminates a link $\Link\x\y$ by unifying $x$
@@
 reduction by structural pre-congruence.
+
+\subsection{Type System}
+
+Types, ranged over by $A$, $B$, \dots, are propositions of multiplicative,
+additive linear logic:
+\[
+    A, B ::= \Zero \mid \Top \mid \One \mid \Bot \mid A \Plus B \mid A \With B \mid A \Ten B \mid A \Par B
+\]
+
+context syntax and notation
+
+\begin{table}
+    \caption{Operational semantics of \Calculus.}
+    \label{tab:semantics}
+    \begin{mathpar}
+        \inferrule[\LinkRule]{~}{
+            \wtp{
+                \Link\x\y
+            }{
+                x : A, y : \dual{A}
+            }
+        }
+        \and
+        \inferrule[\FailRule]{~}{
+            \wtp{
+                \Fail\x
+            }{
+                \Context, x : \Top
+            }
+        }
+        \and
+        \inferrule[\CloseRule]{~}{
+            \wtp{
+                \Close\x
+            }{
+                x : \One
+            }
+        }
+        \and
+        \inferrule[\WaitRule]{
+            \wtp{
+                P
+            }{
+                \Context
+            }
+        }{
+            \wtp{
+                \Wait\x.P
+            }{
+                \Context, x : \Bot
+            }
+        }
+        \and
+        \inferrule[\SelectRule]{
+            \wtp{
+                P
+            }{
+                \Context, y : A_i
+            }
+        }{
+            \wtp{
+                \Select\x{\Tag_i}\y.P
+            }{
+                \Context, x : A_1 \Plus A_2
+            }
+        }
+        \and
+        \inferrule[\CaseRule]{
+            \wtp{
+                P
+            }{
+                \Context, y : A
+            }
+            \\
+            \wtp{
+                Q
+            }{
+                \Context, y : B
+            }
+        }{
+            \wtp{
+                \Case\x\y{P}{Q}
+            }{
+                \Context, x : A \With B
+            }
+        }
+        \and
+        \inferrule[\ForkRule]{
+            \wtp{
+                P
+            }{
+                \ContextC, y : A
+            }
+            \\
+            \wtp{
+                Q
+            }{
+                \ContextD, z : B
+            }
+        }{
+            \wtp{
+                \Join\x\y\z{P}{Q}
+            }{
+                \ContextC, \ContextD, x : A \Ten B
+            }
+        }
+        \and
+        \inferrule[\JoinRule]{
+            \wtp{
+                P
+            }{
+                \Context, y : A, z : B
+            }
+        }{
+            \wtp{
+                \Join\x\y\z.P
+            }{
+                \Context, x : A \Par B
+            }
+        }
+        \and
+        \inferrule[\CutRule]{
+            \wtp{
+                P
+            }{
+                \ContextC, x : A
+            }
+            \\
+            \wtp{
+                Q
+            }{
+                \ContextD, x : \dual{A}
+            }
+        }{
+            \wtp{
+                \Cut\x{P}{Q}
+            }{
+                \ContextC, \ContextD
+            }
+        }
+    \end{mathpar}
+\end{table}

### 2025-06-19 paper/calculus.tex
++

@@
 
-The calculus of pure sessions that is formalized in our development is closely
-related to the multiplicative, additive fragment of CP~\citep{Wadler14}, without
-shared channels. Its syntax makes use of an infinite set of \emph{channel
-names}, ranged over by $x$, $y$ and $z$, and is shown in \cref{tab:syntax}.
+The calculus of pure sessions that is formalized in our development is dubbed
+\Calculus and is closely related to the multiplicative, additive fragment of
+CP~\citep{Wadler14}, without shared channels. Its syntax makes use of an
+infinite set of \emph{channel names}, ranged over by $x$, $y$ and $z$, and is
+shown in \cref{tab:syntax}.
 %

### 2025-07-05 paper/calculus.tex
++

@@
 
-Types, ranged over by $A$, $B$, \dots, are propositions of multiplicative,
-additive linear logic:
+The types for \Calculus, ranged over by $A$, $B$, \dots, are propositions of
+\MALL generated by the following grammar:
 \[
@@
 \]
+
+The interpretation of propositions as session types is the usual
+one~\cite{Wadler14}: the constants $\Bot$ and $\One$ describe the behavior of a
+process sending/receiving a termination signal; the additive connectives $A
+\Plus B$ and $\A \With B$ describe the behavior of a process sending/receiving a
+boolean value and then adhering to either $A$ and $B$ accordingly; the
+multiplicative connectives $A \Ten B$ and $A \Par B$ describe the behavior of a
+process sending/receiving a channel (to be used according to $A$) and then
+continuing as described by $B$. The additive constants $\Zero$ and $\Top$ do not
+describe useful protocols, but they can serve the role of smallest/largest type
+in type systems that support a notion of subtyping~\citep{HornePadovani24}.
 

### 2025-07-20 paper/calculus.tex
client server syntax and semantics

@@
 \begin{table}
-    \caption{Syntax of \Calculus.}
-    \label{tab:syntax}
+    \caption{Syntax of \Calculus.}\label{tab:syntax}
+    \begin{tabular}{c}
         \begin{math}
@@
             \begin{array}[t]{rcll}
-            P, Q & ::= & \Link\x\y \\
-                & | & \Fail\x \\
-                & | & \Close\x \\
-                & | & \Wait\x.P \\
-                & | & \Select\x{\InTag_i}\z.P \\
-                & | & \Case\x\z{P}{Q} \\
-                & | & \Fork\x\y\z{P}{Q} \\
-                & | & \Join\x\y\z.P \\
-                & | & \Cut\x{P}{Q} \\
+                P, Q & ::= & \Link\x\y            & \text{link} \\
+                    & | & \Fail\x                 & \text{fail} \\
+                    & | & \Close\x                & \text{close} \\
+                    & | & \Wait\x.P               & \text{wait} \\
+                    & | & \Select\x{\InTag_i}\z.P & \text{select} \\
+                    & | & \Case\x\z{P}{Q}         & \text{branch} \\
+                    & | & \Fork\x\y\z{P}{Q}       & \text{fork} \\
+                    & | & \Join\x\y\z.P           & \text{join} \\
+                    & | & \Server\x\y.P           & \text{accept} \\
+                    & | & \Client\x\y.P           & \text{request} \\
+                    & | & \Weaken\x.P             & \text{weaken} \\
+                    & | & \Contract\x\y\z.P       & \text{contract} \\
+                    & | & \Cut\x{P}{Q}            & \text{cut} \\
             \end{array}
         \end{math}
+    \end{tabular}
 \end{table}
@@
 The calculus of pure sessions that is formalized in our development is dubbed
-\Calculus and is closely related to the multiplicative, additive fragment of
-CP~\citep{Wadler14}, without shared channels. Its syntax makes use of an
-infinite set of \emph{channel names}, ranged over by $x$, $y$ and $z$, and is
-shown in \cref{tab:syntax}.
+\Calculus and is a linear version of \CP~\citep{Wadler14,GayVasconcelos25}. Its
+syntax makes use of an infinite set of \emph{channel names}, ranged over by $x$,
+$y$ and $z$, and is shown in Table~\ref{tab:syntax}.
 %
@@
 of a termination signal on the channel $x$. The former process has no
-continuation, while the latter process continues as $P$ once the session has
-been terminated.
+continuation, while the latter process continues as $P$ once the signa has been
+received.
 %
-The $\Fail\x$ process denotes a runtime error concerning the session $x$.
+The $\Fail\x$ process denotes a runtime error concerning the channel $x$.
 %
@@
 The $\Fork\x\y\z{P}{Q}$ and $\Join\x\y\z.R$ processes respectively denote the
-output and input of a pair of fresh channels $y$ and $z$ on the channel $x$.
-These forms are often used in conjunction with links to model \emph{free
-outputs}, but can also be used to model bifurcating sessions whereby a
-sequential protocol continues as two parallel sub-protocols.
+output and input of a pair of fresh channels $y$ and $z$ on the channel $x$. The
+former process forks into $P$ and $Q$, each using $y$ and $z$ respectively. The
+latter process continues as $R$. These forms are often used in conjunction with
+links to model \emph{free outputs}, but can also be used to model bifurcating
+sessions whereby a sequential protocol continues as two parallel sub-protocols.
+
+Next we have process forms dealing with shared (non-linear) channels. The
+$\Server\x\y.P$ and $\Client\x\y.P$ processes respectively denote \emph{servers}
+and \emph{clients} acting on the $x$. Each request (from a client) spawns a copy
+of the server's body using the continuation channel $y$. The process
+$\Weaken\x.P$ denotes an explicit \emph{weakening}, that is a client that does
+\emph{not} use the server waiting on $x$. The process $\Contract\x\y\z.P$
+denotes an explicit \emph{contraction}, that is a client that uses the server
+$x$ multiple times on the channels $y$ and $z$.
+
+\emph{Cuts} of the form $\Cut\x{P}{Q}$ represent the parallel composition of $P$
+and $Q$ connected by a channel $x$.
 
@@
 $z$ in $Q$ but not in $P$, and $\Join\x\y\z.P$ binds both $y$ and $z$ in $P$.
+%
+Also, $\Server\x\y.P$ and $\Client\x\y.P$ bind $y$ in $P$ and
+$\Contract\x\y\z.P$ binds $y$ and $z$ in $P$.
 %
@@
     \label{tab:semantics}
-    \[
+    \begin{tabular}{@{}c@{}}
+        \begin{math}
+            \displaystyle
             \begin{array}{rrcll}
@@
             \Cut\x{\Join\y\u\v.P}{Q} & \pcong & \Join\y\u\v.\Cut\x{P}{Q} & x \ne y,u,v \\
+            \SClient &
+            \Cut\x{\Client\y\z.P}{Q} & \pcong & \Client\y\z.\Cut\x{P}{Q} & x \ne y,z \\
+            \SWeaken &
+            \Cut\x{\Weaken\y.P}{Q} & \pcong & \Weaken\y.\Cut\x{P}{Q} & x \ne y \\
+            \SContract &
+            \Cut\x{\Contract\y\u\v.P}{Q} & \pcong & \Contract\y\u\v.\Cut\x{P}{Q} & x \ne y,u,v \\
+            \SServer &
+            \Cut\x{\Server\y\u.P}{\Server\x\v.Q} & \pcong & \Server\y\u.\Cut\x{P}{\Server\x\v.Q} & x \ne y,u \\
             \\
@@
             \Cut\x{\Link\x\y}{P} & \red & P\subst\y\x \\
-        % \RFail &
-        % \Cut\x{\Fail\y}{P} & \red & \Fail\y & x \ne y \\
             \RClose &
@@
             \Cut\x{\Fork\x\y\z{P}{Q}}{\Join\x\y\z.R} & \red & \Cut\y{P}{\Cut\z{Q}{R}} \\
+            \RConnect &
+            \Cut\x{\Server\x\y.P}{\Client\x\y.Q} & \red & \Cut\y{P}{Q} \\
+            \RWeaken &
+            \Cut\x{\Server\x\y.P}{\Weaken\x.Q} & \red & \Weaken{\seqof\z}.Q
+            & \fn{P} = \set{y,\seqof\z} \\
+            \RContract &
+            \Cut\x{\Server\x\y.P}{\Contract\x{x'}{x''}.Q} & \red &
+            \multicolumn{2}{l}{
+                \Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.\Cut{x'}{
+                    \Server{x'}\y.P'
+                }{
+                    \Cut{x''}{
+                        \Server{x''}\y.P''
+                    }{
+                        Q
+                    }
+                }
+                ~^*
+            }
+            \\
             \RCut &
@@
             P & \red & Q & P \pcong R \red Q \\
+            \\
+            \multicolumn{5}{c}{
+                ~^*
+                \fn{P} = \set{y,\seqof\z},
+                P' = P\subst{\seqof\z'}{\seqof\z},
+                P'' = P\subst{\seqof\z''}{\seqof\z}
+            }
             \end{array}
-    \]
+        \end{math}
+    \end{tabular}
 \end{table}
 
-The operational semantics of \Calculus is shown in \cref{tab:semantics} and is
-given by two relations: a \emph{structural pre-congruence} relation $\pcong$ and
-a \emph{reduction} relation $\red$. Structural pre-congruence relates processes
-that we consider to be essentially indistinguishable. Reduction relates
-processes in which a communication occurs. In some presentations of linear
-logic~\cite{Doumane}, these relations are respectively dubbed \emph{external}
-and \emph{internal} reductions. Let us describe the two relations more in
-detail.
+The operational semantics of \Calculus is shown in Table~\ref{tab:semantics} and
+is given by two relations: a \emph{structural pre-congruence} relation $\pcong$
+and a \emph{reduction} relation $\red$. Structural pre-congruence relates
+processes that we consider to be essentially indistinguishable. Reduction
+relates processes in which a communication occurs. In some presentations of
+linear logic~\cite{Doumane}, these relations are respectively dubbed
+\emph{external} and \emph{internal} reductions. Let us describe the two
+relations more in detail.
 
@@
 
+The notion of \emph{duality} is standard and we write $\dual{A}$ for the dual of
+$A$.
+
 context syntax and notation
-
-\begin{table}
-    \caption{Operational semantics of \Calculus.}
-    \label{tab:semantics}
-    \begin{mathpar}
-        \inferrule[\LinkRule]{~}{
-            \wtp{
-                \Link\x\y
-            }{
-                x : A, y : \dual{A}
-            }
-        }
-        \and
-        \inferrule[\FailRule]{~}{
-            \wtp{
-                \Fail\x
-            }{
-                \Context, x : \Top
-            }
-        }
-        \and
-        \inferrule[\CloseRule]{~}{
-            \wtp{
-                \Close\x
-            }{
-                x : \One
-            }
-        }
-        \and
-        \inferrule[\WaitRule]{
-            \wtp{
-                P
-            }{
-                \Context
-            }
-        }{
-            \wtp{
-                \Wait\x.P
-            }{
-                \Context, x : \Bot
-            }
-        }
-        \and
-        \inferrule[\SelectRule]{
-            \wtp{
-                P
-            }{
-                \Context, y : A_i
-            }
-        }{
-            \wtp{
-                \Select\x{\Tag_i}\y.P
-            }{
-                \Context, x : A_1 \Plus A_2
-            }
-        }
-        \and
-        \inferrule[\CaseRule]{
-            \wtp{
-                P
-            }{
-                \Context, y : A
-            }
-            \\
-            \wtp{
-                Q
-            }{
-                \Context, y : B
-            }
-        }{
-            \wtp{
-                \Case\x\y{P}{Q}
-            }{
-                \Context, x : A \With B
-            }
-        }
-        \and
-        \inferrule[\ForkRule]{
-            \wtp{
-                P
-            }{
-                \ContextC, y : A
-            }
-            \\
-            \wtp{
-                Q
-            }{
-                \ContextD, z : B
-            }
-        }{
-            \wtp{
-                \Join\x\y\z{P}{Q}
-            }{
-                \ContextC, \ContextD, x : A \Ten B
-            }
-        }
-        \and
-        \inferrule[\JoinRule]{
-            \wtp{
-                P
-            }{
-                \Context, y : A, z : B
-            }
-        }{
-            \wtp{
-                \Join\x\y\z.P
-            }{
-                \Context, x : A \Par B
-            }
-        }
-        \and
-        \inferrule[\CutRule]{
-            \wtp{
-                P
-            }{
-                \ContextC, x : A
-            }
-            \\
-            \wtp{
-                Q
-            }{
-                \ContextD, x : \dual{A}
-            }
-        }{
-            \wtp{
-                \Cut\x{P}{Q}
-            }{
-                \ContextC, \ContextD
-            }
-        }
-    \end{mathpar}
-\end{table}

### 2025-07-20 paper/main.tex
client server syntax and semantics

@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                                                 %%
-%% Please do not use \input{...} to include other tex files.       %%
-%% Submit your LaTeX manuscript as one .tex document.              %%
-%%                                                                 %%
-%% All additional figures and files should be attached             %%
-%% separately and not embedded in the \TeX\ document itself.       %%
-%%                                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%=======================================================%%
-%% to print line numbers in the margin use lineno option %%
-%%=======================================================%%
-
-\documentclass[pdflatex,sn-basic]{sn-jnl}
+\documentclass[pdflatex,sn-mathphys-num]{sn-jnl}
 
@@
 \usepackage{amsthm}
-\usepackage{cleveref}
 \usepackage{xcolor}
+\usepackage{textcomp}
 \usepackage{listings}
+\usepackage{mathtools}
+\usepackage{stmaryrd}
+\usepackage{mathpartir}
 \usepackage{agda}
 \usepackage{xspace}
-
+\usepackage{upgreek}
+%%%%
 \input{macros}
 
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}
-\newtheorem{proposition}[theorem]{Proposition}
+%%%%%=============================================================================%%%%
+%%%%  Remarks: This template is provided to aid authors with the preparation
+%%%%  of original research articles intended for submission to journals published 
+%%%%  by Springer Nature. The guidance has been prepared in partnership with 
+%%%%  production teams to conform to Springer Nature technical requirements. 
+%%%%  Editorial and presentation requirements differ among journal portfolios and 
+%%%%  research disciplines. You may find sections in this template are irrelevant 
+%%%%  to your work and are empowered to omit any such section if allowed by the 
+%%%%  journal you intend to submit to. The submission guidelines and policies 
+%%%%  of the journal take precedence. A detailed User Manual is available in the 
+%%%%  template package for technical guidance.
+%%%%%=============================================================================%%%%
 
-\theoremstyle{definition}
-\newtheorem{example}{Example}
-\newtheorem{definition}{Definition}
+%% as per the requirement new theorem styles can be included as shown below
+%\theoremstyle{theorem}%
+\newtheorem{theorem}{Theorem}%  meant for continuous numbers
+%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
+%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
+\newtheorem{proposition}[theorem]{Proposition}% 
+%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
 
-\theoremstyle{remark}
-\newtheorem{remark}{Remark}
+% \theoremstyle{thmstyletwo}%
+% \newtheorem{example}{Example}%
+% \newtheorem{remark}{Remark}%
+
+% \theoremstyle{thmstylethree}%
+% \newtheorem{definition}{Definition}%
 
@@
 
+%%==================================%%
+%% Sample for unstructured abstract %%
+%%==================================%%
+
 \abstract{blah}
 
+%%================================%%
+%% Sample for structured abstract %%
+%%================================%%
+
+% \abstract{\textbf{Purpose:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
+The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations.
+As a guide the abstract should not exceed 200 words.
+Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
+% 
+% \textbf{Methods:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
+The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations.
+As a guide the abstract should not exceed 200 words.
+Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
+% 
+% \textbf{Results:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
+The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations.
+As a guide the abstract should not exceed 200 words.
+Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
+% 
+% \textbf{Conclusion:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
+The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations.
+As a guide the abstract should not exceed 200 words.
+Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.}
+
 \keywords{keyword1, Keyword2, Keyword3, Keyword4}
+
+%%\pacs[JEL Classification]{D8, H51}
+
+%%\pacs[MSC Classification]{35A01, 65L10, 65L12, 65L20, 65L70}
 
@@
 \end{document}
-
-Introduction section, of referenced text \cite{bib1} expands on the background
-of the work (some overlap with the Abstract is acceptable). The introduction
-should not include subheadings.
-
-Springer Nature does not impose a strict layout as standard however authors are
-advised to check the individual requirements for the journal they are planning
-to submit to as there may be journal-level preferences. When preparing your text
-please also be aware that some stylistic choices are not supported in full text
-XML (publication version), including coloured font. These will not be replicated
-in the typeset article if it is accepted.
-
-\section{Results}\label{sec2}
-
-Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.
-
-\section{This is an example for first level head---section head}\label{sec3}
-
-\subsection{This is an example for second level head---subsection head}\label{subsec2}
-
-\subsubsection{This is an example for third level head---subsubsection head}\label{subsubsec2}
-
-Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.
-
-\section{Equations}\label{sec4}
-
-Equations in \LaTeX\ can either be inline or on-a-line by itself (``display equations''). For
-inline equations use the \verb+$...$+ commands. E.g.: The equation
-$H\psi = E \psi$ is written via the command \verb+$H \psi = E \psi$+.
-
-For display equations (with auto generated equation numbers)
-one can use the equation or align environments:
-\begin{equation}
-\|\tilde{X}(k)\|^2 \leq\frac{\sum\limits_{i=1}^{p}\left\|\tilde{Y}_i(k)\right\|^2+\sum\limits_{j=1}^{q}\left\|\tilde{Z}_j(k)\right\|^2 }{p+q}.\label{eq1}
-\end{equation}
-where,
-\begin{align}
-D_\mu &=  \partial_\mu - ig \frac{\lambda^a}{2} A^a_\mu \nonumber \\
-F^a_{\mu\nu} &= \partial_\mu A^a_\nu - \partial_\nu A^a_\mu + g f^{abc} A^b_\mu A^a_\nu \label{eq2}
-\end{align}
-Notice the use of \verb+\nonumber+ in the align environment at the end
-of each line, except the last, so as not to produce equation numbers on
-lines where no equation numbers are required. The \verb+\label{}+ command
-should only be used at the last line of an align environment where
-\verb+\nonumber+ is not used.
-\begin{equation}
-Y_\infty = \left( \frac{m}{\textrm{GeV}} \right)^{-3}
-    \left[ 1 + \frac{3 \ln(m/\textrm{GeV})}{15}
-    + \frac{\ln(c_2/5)}{15} \right]
-\end{equation}
-The class file also supports the use of \verb+\mathbb{}+, \verb+\mathscr{}+ and
-\verb+\mathcal{}+ commands. As such \verb+\mathbb{R}+, \verb+\mathscr{R}+
-and \verb+\mathcal{R}+ produces $\mathbb{R}$, $\mathscr{R}$ and $\mathcal{R}$
-respectively (refer Subsubsection~\ref{subsubsec2}).
-
-\section{Tables}\label{sec5}
-
-Tables can be inserted via the normal table and tabular environment. To put
-footnotes inside tables you should use \verb+\footnotetext[]{...}+ tag.
-The footnote appears just below the table itself (refer Tables~\ref{tab1} and \ref{tab2}). 
-For the corresponding footnotemark use \verb+\footnotemark[...]+
-
-\begin{table}[h]
-\caption{Caption text}\label{tab1}%
-\begin{tabular}{@{}llll@{}}
-\toprule
-Column 1 & Column 2  & Column 3 & Column 4\\
-\midrule
-row 1    & data 1   & data 2  & data 3  \\
-row 2    & data 4   & data 5\footnotemark[1]  & data 6  \\
-row 3    & data 7   & data 8  & data 9\footnotemark[2]  \\
-\botrule
-\end{tabular}
-\footnotetext{Source: This is an example of table footnote. This is an example of table footnote.}
-\footnotetext[1]{Example for a first table footnote. This is an example of table footnote.}
-\footnotetext[2]{Example for a second table footnote. This is an example of table footnote.}
-\end{table}
-
-\noindent
-The input format for the above table is as follows:
-
-%%=============================================%%
-%% For presentation purpose, we have included  %%
-%% \bigskip command. Please ignore this.       %%
-%%=============================================%%
-\bigskip
-\begin{verbatim}
-\begin{table}[<placement-specifier>]
-\caption{<table-caption>}\label{<table-label>}%
-\begin{tabular}{@{}llll@{}}
-\toprule
-Column 1 & Column 2 & Column 3 & Column 4\\
-\midrule
-row 1 & data 1 & data 2	 & data 3 \\
-row 2 & data 4 & data 5\footnotemark[1] & data 6 \\
-row 3 & data 7 & data 8	 & data 9\footnotemark[2]\\
-\botrule
-\end{tabular}
-\footnotetext{Source: This is an example of table footnote. 
-This is an example of table footnote.}
-\footnotetext[1]{Example for a first table footnote.
-This is an example of table footnote.}
-\footnotetext[2]{Example for a second table footnote. 
-This is an example of table footnote.}
-\end{table}
-\end{verbatim}
-\bigskip
-%%=============================================%%
-%% For presentation purpose, we have included  %%
-%% \bigskip command. Please ignore this.       %%
-%%=============================================%%
-
-In case of double column layout, tables which do not fit in single column width should be set to full text width.
-For this, you need to use \verb+\begin{table*}+ \verb+...+ \verb+\end{table*}+ instead of \verb+\begin{table}+ \verb+...+ \verb+\end{table}+ environment.
-Lengthy tables which do not fit in textwidth should be set as rotated table.
-For this, you need to use \verb+\begin{sidewaystable}+ \verb+...+ \verb+\end{sidewaystable}+ instead of \verb+\begin{table*}+ \verb+...+ \verb+\end{table*}+ environment.
-This environment puts tables rotated to single column width.
-For tables rotated to double column width, use \verb+\begin{sidewaystable*}+ \verb+...+ \verb+\end{sidewaystable*}+.
-
-\section{Figures}\label{sec6}
-
-As per the \LaTeX\ standards you need to use eps images for \LaTeX\ compilation and \verb+pdf/jpg/png+ images for \verb+PDFLaTeX+ compilation.
-This is one of the major difference between \LaTeX\ and \verb+PDFLaTeX+.
-Each image should be from a single input .eps/vector image file.
-Avoid using subfigures.
-The command for inserting images for \LaTeX\ and \verb+PDFLaTeX+ can be generalized.
-The package used to insert images in \verb+LaTeX/PDFLaTeX+ is the graphicx package.
-Figures can be inserted via the normal figure environment as shown in the below example:
-
-In case of double column layout, the above format puts figure captions/images to single column width. To get spanned images, we need to provide \verb+\begin{figure*}+ \verb+...+ \verb+\end{figure*}+.
-
-For sample purpose, we have included the width of images in the optional argument of \verb+\includegraphics+ tag. Please ignore this.
-
-\section{Algorithms, Program codes and Listings}\label{sec7}
-
-Packages \verb+algorithm+, \verb+algorithmicx+ and \verb+algpseudocode+ are used for setting algorithms in \LaTeX\ using the format:
-
-%%=============================================%%
-%% For presentation purpose, we have included  %%
-%% \bigskip command. Please ignore this.       %%
-%%=============================================%%
-
-You may refer above listed package documentations for more details before setting \verb+algorithm+ environment.
-For program codes, the ``verbatim'' package is required and the command to be used is \verb+\begin{verbatim}+ \verb+...+ \verb+\end{verbatim}+.
-
-Similarly, for \verb+listings+, use the \verb+listings+ package.
-\verb+\begin{lstlisting}+ \verb+...+ \verb+\end{lstlisting}+ is used to set environments similar to \verb+verbatim+ environment.
-Refer to the \verb+lstlisting+ package documentation for more details.
-
-A fast exponentiation procedure:
-
-\lstset{texcl=true,basicstyle=\small\sf,commentstyle=\small\rm,mathescape=true,escapeinside={(*}{*)}}
-\begin{lstlisting}
-begin
-  for $i:=1$ to $10$ step $1$ do
-      expt($2,i$);  
-      newline() od                (*\textrm{Comments will be set flush to the right margin}*)
-where
-proc expt($x,n$) $\equiv$
-  $z:=1$;
-  do if $n=0$ then exit fi;
-     do if odd($n$) then exit fi;                 
-        comment: (*\textrm{This is a comment statement;}*)
-        $n:=n/2$; $x:=x*x$ od;
-     { $n>0$ };
-     $n:=n-1$; $z:=z*x$ od;
-  print($z$). 
-end
-\end{lstlisting}
-
-%%=============================================%%
-%% For presentation purpose, we have included  %%
-%% \bigskip command. Please ignore this.       %%
-%%=============================================%%
-\bigskip
-\begin{minipage}{\hsize}%
-\lstset{frame=single,framexleftmargin=-1pt,framexrightmargin=-17pt,framesep=12pt,linewidth=0.98\textwidth,language=pascal}% Set your language (you can change the language for each code-block optionally)
-%%% Start your code-block
-\begin{lstlisting}
-for i:=maxint to 0 do
-begin
-{ do nothing }
-end;
-Write('Case insensitive ');
-Write('Pascal keywords.');
-\end{lstlisting}
-\end{minipage}
-
-\section{Cross referencing}\label{sec8}
-
-Environments such as figure, table, equation and align can have a label
-declared via the \verb+\label{#label}+ command. For figures and table
-environments use the \verb+\label{}+ command inside or just
-below the \verb+\caption{}+ command. You can then use the
-\verb+\ref{#label}+ command to cross-reference them. As an example, consider
-the label declared for Figure~\ref{fig1} which is
-\verb+\label{fig1}+. To cross-reference it, use the command 
-\verb+Figure \ref{fig1}+, for which it comes up as
-``Figure~\ref{fig1}''.
-
-To reference line numbers in an algorithm, consider the label declared for the line number 2 of Algorithm~\ref{algo1} is \verb+\label{algln2}+.
-To cross-reference it, use the command \verb+\ref{algln2}+ for which it comes up as line~\ref{algln2} of Algorithm~\ref{algo1}.
-
-\subsection{Details on reference citations}\label{subsec7}
-
-Standard \LaTeX\ permits only numerical citations.
-To support both numerical and author-year citations this template uses \verb+natbib+ \LaTeX\ package.
-For style guidance please refer to the template user manual.
-
-Here is an example for \verb+\cite{...}+: \cite{bib1}.
-Another example for \verb+\citep{...}+: \citep{bib2}.
-For author-year citation mode, \verb+\cite{...}+ prints Jones et al.
-(1990) and \verb+\citep{...}+ prints (Jones et al., 1990).
-
-All cited bib entries are printed at the end of this article: \cite{bib3}, \cite{bib4}, \cite{bib5}, \cite{bib6}, \cite{bib7}, \cite{bib8}, \cite{bib9}, \cite{bib10}, \cite{bib11}, \cite{bib12} and \cite{bib13}.
-
-\section{Examples for theorem like environments}\label{sec10}
-
-For theorem like environments, we require \verb+amsthm+ package. There are three types of predefined theorem styles exists---\verb+thmstyleone+, \verb+thmstyletwo+ and \verb+thmstylethree+
-
-%%=============================================%%
-%% For presentation purpose, we have included  %%
-%% \bigskip command. Please ignore this.       %%
-%%=============================================%%
-\bigskip
-\begin{tabular}{|l|p{19pc}|}
-\hline
-\verb+thmstyleone+ & Numbered, theorem head in bold font and theorem text in italic style \\\hline
-\verb+thmstyletwo+ & Numbered, theorem head in roman font and theorem text in italic style \\\hline
-\verb+thmstylethree+ & Numbered, theorem head in bold font and theorem text in roman style \\\hline
-\end{tabular}
-\bigskip
-%%=============================================%%
-%% For presentation purpose, we have included  %%
-%% \bigskip command. Please ignore this.       %%
-%%=============================================%%
-
-For mathematics journals, theorem styles can be included as shown in the following examples:
-
-\begin{theorem}[Theorem subhead]\label{thm1}
-Example theorem text. Example theorem text. Example theorem text. Example theorem text. Example theorem text. 
-Example theorem text. Example theorem text. Example theorem text. Example theorem text. Example theorem text. 
-Example theorem text. 
-\end{theorem}
-
-Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.
-
-\begin{proposition}
-Example proposition text. Example proposition text. Example proposition text. Example proposition text. Example proposition text. 
-Example proposition text. Example proposition text. Example proposition text. Example proposition text. Example proposition text. 
-\end{proposition}
-
-Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.
-
-\begin{example}
-Phasellus adipiscing semper elit. Proin fermentum massa
-ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
-at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem. 
-\end{example}
-
-Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.
-
-\begin{remark}
-Phasellus adipiscing semper elit. Proin fermentum massa
-ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
-at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem. 
-\end{remark}
-
-Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.
-
-\begin{definition}[Definition sub head]
-Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. 
-\end{definition}
-
-Additionally a predefined ``proof'' environment is available: \verb+\begin{proof}+ \verb+...+ \verb+\end{proof}+.
-This prints a ``Proof'' head in italic font style and the ``body text'' in roman font style with an open square at the end of each proof environment.
-
-\begin{proof}
-Example for proof text.
-Example for proof text.
-Example for proof text.
-Example for proof text.
-Example for proof text.
-Example for proof text.
-Example for proof text.
-Example for proof text.
-Example for proof text.
-Example for proof text.
-\end{proof}
-
-Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.
-
-\begin{proof}[Proof of Theorem~{\upshape\ref{thm1}}]
-Example for proof text.
-Example for proof text.
-Example for proof text.
-Example for proof text.
-Example for proof text.
-Example for proof text.
-Example for proof text.
-Example for proof text.
-Example for proof text.
-Example for proof text.
-\end{proof}
-
-\noindent
-For a quote environment, use \verb+\begin{quote}...\end{quote}+
-\begin{quote}
-Quoted text example. Aliquam porttitor quam a lacus. Praesent vel arcu ut tortor cursus volutpat. In vitae pede quis diam bibendum placerat. Fusce elementum
-convallis neque. Sed dolor orci, scelerisque ac, dapibus nec, ultricies ut, mi. Duis nec dui quis leo sagittis commodo.
-\end{quote}
-
-Sample body text. Sample body text. Sample body text. Sample body text. Sample body text (refer Figure~\ref{fig1}). Sample body text. Sample body text. Sample body text (refer Table~\ref{tab3}).
-
-\section{Methods}\label{sec11}
-
-Topical subheadings are allowed.
-Authors must ensure that their Methods section includes adequate experimental and characterization data necessary for others in the field to reproduce their work.
-Authors are encouraged to include RIIDs where appropriate.
-
-\textbf{Ethical approval declarations} (only required where applicable) Any article reporting experiment/s carried out on (i)~live vertebrate (or higher invertebrates), (ii)~humans or (iii)~human samples must include an unambiguous statement within the methods section that meets the following requirements:
-
-\begin{enumerate}[1.]
-\item Approval: a statement which confirms that all experimental protocols were approved by a named institutional and/or licensing committee. Please identify the approving body in the methods section
-
-\item Accordance: a statement explicitly saying that the methods were carried out in accordance with the relevant guidelines and regulations
-
-\item Informed consent (for experiments involving humans or human tissue samples): include a statement confirming that informed consent was obtained from all participants and/or their legal guardian/s
-\end{enumerate}
-
-If your manuscript includes potentially identifying patient/participant information, or if it describes human transplantation research, or if it reports results of a clinical trial then  additional information will be required.
-Please visit (\url{https://www.nature.com/nature-research/editorial-policies}) for Nature Portfolio journals, (\url{https://www.springer.com/gp/authors-editors/journal-author/journal-author-helpdesk/publishing-ethics/14214}) for Springer Nature journals, or (\url{https://www.biomedcentral.com/getpublished/editorial-policies\#ethics+and+consent}) for BMC.
-
-\section{Discussion}\label{sec12}
-
-Discussions should be brief and focused.
-In some disciplines use of Discussion or `Conclusion' is interchangeable.
-It is not mandatory to use both.
-Some journals prefer a section `Results and Discussion' followed by a section `Conclusion'.
-Please refer to Journal-level guidance for any specific requirements.
-
-\section{Conclusion}\label{sec13}
-
-Conclusions may be used to restate your hypothesis or research question, restate your major findings, explain the relevance and the added value of your work, highlight any limitations of your study, describe future directions for research and recommendations.
-
-In some disciplines use of Discussion or 'Conclusion' is interchangeable. It is not mandatory to use both. Please refer to Journal-level guidance for any specific requirements.
-
-\backmatter
-
-\bmhead{Supplementary information}
-
-If your article has accompanying supplementary file/s please state so here.
-
-Authors reporting data from electrophoretic gels and blots should supply the full unprocessed scans for key as part of their Supplementary information.
-This may be requested by the editorial team/s if it is missing.
-
-Please refer to Journal-level guidance for any specific requirements.
-
-\bmhead{Acknowledgements}
-
-Acknowledgements are not compulsory. Where included they should be brief. Grant or contribution numbers may be acknowledged.
-
-Please refer to Journal-level guidance for any specific requirements.
-
-\section*{Declarations}
-
-Some journals require declarations to be submitted in a standardised format.
-Please check the Instructions for Authors of the journal to which you are submitting to see if you need to complete this section.
-If yes, your manuscript must contain the following sections under the heading `Declarations':
-
-\begin{itemize}
-\item Funding
-\item Conflict of interest/Competing interests (check journal-specific guidelines for which heading to use)
-\item Ethics approval and consent to participate
-\item Consent for publication
-\item Data availability 
-\item Materials availability
-\item Code availability 
-\item Author contribution
-\end{itemize}
-
-\noindent
-If any of the sections are not relevant to your manuscript, please include the heading and write `Not applicable' for that section.
-
-%%===================================================%%
-%% For presentation purpose, we have included        %%
-%% \bigskip command. Please ignore this.             %%
-%%===================================================%%
-\bigskip
-\begin{flushleft}%
-Editorial Policies for:
-
-\bigskip\noindent
-Springer journals and proceedings: \url{https://www.springer.com/gp/editorial-policies}
-
-\bigskip\noindent
-Nature Portfolio journals: \url{https://www.nature.com/nature-research/editorial-policies}
-
-\bigskip\noindent
-\textit{Scientific Reports}: \url{https://www.nature.com/srep/journal-policies/editorial-policies}
-
-\bigskip\noindent
-BMC journals: \url{https://www.biomedcentral.com/getpublished/editorial-policies}
-\end{flushleft}
-
-\begin{appendices}
-
-\section{Section title of first appendix}\label{secA1}
-
-An appendix contains supplementary information that is not an essential part of the text itself but which may be helpful in providing a more comprehensive understanding of the research problem or it is information that is too cumbersome to be included in the body of the paper.
-
-%%=============================================%%
-%% For submissions to Nature Portfolio Journals %%
-%% please use the heading ``Extended Data''.   %%
-%%=============================================%%
-
-%%=============================================================%%
-%% Sample for another appendix section			       %%
-%%=============================================================%%
-
-%% \section{Example of another appendix section}\label{secA2}%
-%% Appendices may be used for helpful, supporting or essential material that would otherwise 
-%% clutter, break up or be distracting to the text. Appendices can consist of sections, figures, 
-%% tables and equations etc.
-
-\end{appendices}
-
-%%===========================================================================================%%
-%% If you are submitting to one of the Nature Portfolio journals, using the eJP submission   %%
-%% system, please include the references within the manuscript file itself. You may do this  %%
-%% by copying the reference list from your .bbl file, paste it into the main manuscript .tex %%
-%% file, and delete the associated \verb+\bibliography+ commands.                            %%
-%%===========================================================================================%%
-
-\bibliography{sn-bibliography}% common bib file
-%% if required, the content of .bbl file can be included here once bbl is generated
-%%\input sn-article.bbl
-
-\end{document}

### 2025-07-21 paper/calculus.tex
progress

@@
 syntax makes use of an infinite set of \emph{channel names}, ranged over by $x$,
-$y$ and $z$, and is shown in Table~\ref{tab:syntax}.
+$y$ and $z$, and is shown in~\Cref{tab:syntax}.
 %
@@
 
-\emph{Cuts} of the form $\Cut\x{P}{Q}$ represent the parallel composition of $P$
-and $Q$ connected by a channel $x$.
+\emph{Cuts} of the form $\Cut\x{P}{Q}$ represent the parallel composition of the
+processes $P$ and $Q$ connected by a channel $x$.
 
-The notions of free and bound names are fairly standard. Just bear in mind that
-all action prefixes bind continuation channels in (some) continuation processes.
-In particular, $\Select\x{\Tag_i}\z.P$ binds $z$ in $P$, $\Case\x\z{P}{Q}$ binds
-$z$ in both $P$ and $Q$, $\Fork\x\y\z{P}{Q}$ binds $y$ in $P$ but not in $Q$ and
-$z$ in $Q$ but not in $P$, and $\Join\x\y\z.P$ binds both $y$ and $z$ in $P$.
-%
-Also, $\Server\x\y.P$ and $\Client\x\y.P$ bind $y$ in $P$ and
-$\Contract\x\y\z.P$ binds $y$ and $z$ in $P$.
+The notions of free and bound names are almost standard. Just bear in mind that
+also output action prefixes bind continuation channels in (some) continuation
+processes. In particular, $\Select\x{\Tag_i}\z.P$ binds $z$ in $P$, while
+$\Fork\x\y\z{P}{Q}$ binds $y$ in $P$ but not in $Q$ and $z$ in $Q$ but not in
+$P$.
 %
@@
 
-The operational semantics of \Calculus is shown in Table~\ref{tab:semantics} and
+The operational semantics of \Calculus is shown in~\Cref{tab:semantics} and
 is given by two relations: a \emph{structural pre-congruence} relation $\pcong$
@@
 relates processes in which a communication occurs. In some presentations of
-linear logic~\cite{Doumane}, these relations are respectively dubbed
-\emph{external} and \emph{internal} reductions. Let us describe the two
-relations more in detail.
+linear logic~\cite{BaeldeDoumaneSaurin16,Doumane17}, these relations are
+respectively dubbed \emph{external} and \emph{internal} reductions. Let us
+describe the two relations more in detail.
 
-Structural pre-congruence is the least pre-congruence defined by rule
+Structural pre-congruence is the least pre-congruence defined by the
 \rulename{s-*} rules.
@@
 %
-Note that some rules have additional side-conditions preventing bound names to
-become free.\Luca{Non proprio, sono tutti nomi bound.} Also, there are two
-versions of \SForkL and \SForkR depending on which of the two continuations $P$
-and $Q$ contains occurrences of the restricted channel $x$.
-%
+Note that some rules have additional side-conditions aimed at preserving the
+meaning of names when binders are moved around. Also, there are two versions of
+\SForkL and \SForkR depending on which of the two continuations $P$ and $Q$
+contains occurrences of the restricted channel $x$.
+
 Notably and somewhat surprisingly, there is no pre-congruence rule asserting
@@
 reduction~\cite{Doumane17}.
+%
+Another rule that deserves attention is \SServer. In this case, a server prefix
+$\Server\y\u$ can be pulled out of a cut only provided that the other process in
+the cut is also a server on the channel restricted by the cut.
 
-The reduction rules \RLink, \RClose, \RSelect and \RFork are in one-to-one
-correspondence with the well-known reductions of session calculi and the cut
-reductions of linear logic. \RLink eliminates a link $\Link\x\y$ by unifying $x$
-and $y$. \RClose closes a session $x$ when a termination signal is exchanged on
-$x$. \RSelect and \RFork respectively model the communication of a tag $\Tag_i$
-and of a fresh channel $y$. Since channels are linear in \Calculus, each
-non-terminating communication carries along a \emph{fresh continuation} $z$ on
-which the rest of the conversation may take place, in line with the semantics of
+The reduction rules \RLink, \RClose, \RSelect, \RFork, \RConnect, \RWeaken and
+\RContract are in one-to-one correspondence with the well-known reductions of
+session calculi and the principal cut reductions of linear logic. \RLink
+eliminates a link $\Link\x\y$ by substituting $y$ for $x$. \RClose closes a
+session $x$ when a termination signal is exchanged on $x$. \RSelect and \RFork
+respectively model the communication of a tag $\Tag_i$ and of a fresh channel
+$y$. Each of these communications carries along a \emph{fresh continuation} $z$
+on which the rest of the conversation takes place, in line with the semantics of
 the linear $\pi$-calculus~\cite{KobayashiPierceTurner99}.
 %
+\RConnect, \RWeaken and \RContract describe interactions on shared channels and
+respectively represent the connection between a server and a client, the
+disposal of an unused server, and the duplication of a server. In \RWeaken and
+\RContract we use some convenient (albeit slightly ambiguous) notation for
+denoting sequences of (pairwise distinct) channels and prefixes. In particular,
+$\seqof\z$ stands for a sequence $z_1, \dots, z_n$ of channels,
+$\Weaken{\seqof\z}.Q$ stands for $\Weaken{z_1}\dots\Weaken{z_n}.Q$ and
+$\Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.R$ stands for
+$\Contract{z_1}{z_1'}{z_1''}\dots\Contract{z_n}{z_n'}{z_n''}.R$.
+
 The rule~\RCut propagates reductions through cuts and the rules \RCong closes
@@
 \subsection{Type System}
+\label{sec:typing-rules}
 
-The types for \Calculus, ranged over by $A$, $B$, \dots, are propositions of
-\MALL generated by the following grammar:
+The types for \Calculus, ranged over by $A$, $B$, \dots, are linear logic
+ propositions generated by the following grammar:
 \[
-    A, B ::= \Zero \mid \Top \mid \One \mid \Bot \mid A \Plus B \mid A \With B \mid A \Ten B \mid A \Par B
+    A, B ::= \Zero \mid \Top \mid \One \mid \Bot \mid A \Plus B \mid A \With B \mid A \Ten B \mid A \Par B \mid \OfCourse A \mid \WhyNot A
 \]
@@
 describe useful protocols, but they can serve the role of smallest/largest type
-in type systems that support a notion of subtyping~\citep{HornePadovani24}.
+in type systems that support a notion of subtyping~\citep{HornePadovani24}. The
+``of course'' modality $\OfCourse A$ and the ``why not'' modality $\WhyNot A$
+describe the behavior of servers and clients accepting and requesting
+connections to be used according to $A$.
 
@@
 
-context syntax and notation
+Typing contexts (\ie sequents) are finite maps from channel names to types
+written as $x_1 : A_1, \dots, x_n : A_n$ and ranged over by $\ContextC$ and
+$\ContextD$. We write $\ContextC, \ContextD$ for the union of $\ContextC$ and
+$\ContextD$ when they have disjoint domain. We write $\WhyNot\Context$ for some
+context $\Context = x_1 : \WhyNot A_1, \dots, x_n : \WhyNot A_n$ where all the
+types in its range are prefixed by the $\WhyNot$ modality.
+
+\begin{table}
+    \caption{Typing rules of \Calculus.}
+    \label{tab:typing-rules}
+    \begin{tabular}{@{}c@{}}
+        \begin{mathpar}
+            \inferrule[\LinkRule]{~}{
+                \wtp{\Link\x\y}{x : A, y : \dual{A}}
+            }
+            \and
+            \inferrule[\FailRule]{~}{
+                \wtp{\Fail\x}{x : \Top, \Context}
+            }
+            \and
+            \inferrule[\CloseRule]{~}{
+                \wtp{\Close\x}{x : \One}
+            }
+            \and
+            \inferrule[\WaitRule]{
+                \wtp{P}{\Context}
+            }{
+                \wtp{\Wait\x.P}{x : \Bot, \Context}
+            }
+            \and
+            \inferrule[\SelectRule]{
+                \wtp{P}{y : A_i, \Context}
+            }{
+                \wtp{\Select\x{\InTag_i}\y.P}{x : A_1 \Plus A_2, \Context}
+            }
+            \and
+            \inferrule[\CaseRule]{
+                \wtp{P}{y : A, \Context}
+                \\
+                \wtp{Q}{y : B, \Context}
+            }{
+                \wtp{\Case\x\y{P}{Q}}{x : A \With B, \Context}
+            }
+            \and
+            \inferrule[\ForkRule]{
+                \wtp{P}{y : A, \ContextC}
+                \\
+                \wtp{Q}{z : B, \ContextD}
+            }{
+                \wtp{\Fork\x\y\z{P}{Q}}{x : A \Ten B, \ContextC, \ContextD}
+            }
+            \and
+            \inferrule[\JoinRule]{
+                \wtp{P}{y : A, z : B, \Context}
+            }{
+                \wtp{\Join\x\y\z.P}{x : A \Par B, \Context}
+            }
+            \and
+            \inferrule[\ServerRule]{
+                \wtp{P}{y : A, \WhyNot\Context}
+            }{
+                \wtp{\Server\x\y.P}{x : \OfCourse A, \WhyNot\Context}
+            }
+            \and
+            \inferrule[\ClientRule]{
+                \wtp{P}{y : A, \Context}
+            }{
+                \wtp{\Client\x\y.P}{x : \WhyNot A, \Context}
+            }
+            \and
+            \inferrule[\WeakenRule]{
+                \wtp{P}\Context
+            }{
+                \wtp{P}{x : \WhyNot A, \Context}
+            }
+            \and
+            \inferrule[\ContractRule]{
+                \wtp{P}{y : \WhyNot A, z : \WhyNot A, \Context}
+            }{
+                \wtp{P}{x : \WhyNot A, \Context}
+            }
+            \and
+            \inferrule[\CutRule]{
+                \wtp{P}{x : A, \ContextC}
+                \\
+                \wtp{Q}{x : \dual{A}, \ContextD}
+            }{
+                \wtp{\Cut\x{P}{Q}}{\ContextC, \ContextD}
+            }
+        \end{mathpar}
+    \end{tabular}
+\end{table}
+
+Typing judgments have the form $\wtp{P}\Context$ meaning that the process $P$ is
+well typed in the context $\Context$. Equivalently, the judgment indicates that
+the sequent $\wtp{}\Context$ is derivable and $P$ is a \emph{proof term}
+corresponding to the derivation for $\wtp{}\Context$.
+%
+The typing rules are shown in \Cref{tab:typing-rules}. Since they are in
+one-to-one correspondence with the proof rules of classical linear logic, the
+reader may refer to the standard literature on linear logic~\cite{} or
+propositions as sessions~\citep{Wadler14} for their interpretation.
+
+\subsection{Properties of Well-Typed Processes}
+\label{sec:properties}
+
+\begin{lemma}
+    \label{lem:pcong}
+    If $P \pcong Q$, then $\wtp{P}\Context$ implies $\wtp{Q}\Context$.
+\end{lemma}
+
+\begin{theorem}[subject reduction]
+    \label{thm:red}
+    If $P \red Q$, then $\wtp{P}\Context$ implies $\wtp{Q}\Context$.
+\end{theorem}
+
+\begin{theorem}[deadlock freedom]
+    \label{thm:df}
+    If $\wtp{P}\Context$, then $P$ is either observable or reducible.
+\end{theorem}
+
+safety

### 2025-07-21 paper/calculus.tex
properties

@@
 The calculus of pure sessions that is formalized in our development is dubbed
-\Calculus and is a linear version of \CP~\citep{Wadler14,GayVasconcelos25}. Its
-syntax makes use of an infinite set of \emph{channel names}, ranged over by $x$,
-$y$ and $z$, and is shown in~\Cref{tab:syntax}.
+\Calculus and is a linear version of \CP~\citep{Wadler14,GayVasconcelos25},
+where every session endpoint is meant to be used \emph{exactly once}. This
+differs from \CP and other session calculi where session endpoints can be used
+repeatedly, albeit sequentially.
+
+The syntax of \Calculus makes use of an infinite set of \emph{channel names},
+ranged over by $x$, $y$ and $z$, and is shown in~\Cref{tab:syntax}.
 %
@@
 
-\begin{lemma}
-    \label{lem:pcong}
-    If $P \pcong Q$, then $\wtp{P}\Context$ implies $\wtp{Q}\Context$.
-\end{lemma}
+The \emph{linearity challenge}~\citep{CarboneEtAl24} aims at formalizing two
+essential properties of well-typed processes: (1) typing is preserved by
+reductions; (2) peer endpoints are always used in complementary ways. This
+latter property is dubbed \emph{well formedness} in the text of the challenge.
+We now formulate these properties using the notation developed so far.
+
+Concerning the preservation of typing, this corresponds to the usual subject
+reduction result which is expressed thus.
 
@@
 
-\begin{theorem}[deadlock freedom]
-    \label{thm:df}
-    If $\wtp{P}\Context$, then $P$ is either observable or reducible.
+In order to formulate well formedness, we start by defining \emph{reduction
+contexts} are partial processes with a single unguarded hole $\Hole$, thus:
+\[
+    \ReductionContext ::= \Hole \mid \Cut\x\ReductionContext{P} \mid \Cut\x{P}\ReductionContext
+\]
+
+As usual, we write $\ReductionContext[P]$ for the process obtained by replacing
+the hole in $\ReductionContext$ with $P$, noting that such replacement may
+capture channel names that are bound in $\ReductionContext$ and occur free in
+$P$.
+
+Next we introduce some terminology concerning the shape of processes. We say
+that $P$ is a \emph{thread} if it is anything but a cut. In other words, a
+thread is either a link or a process that starts with an input/output action of
+some sort. Note that a thread may contain cuts, but these must all guarded
+underneath the topmost action prefix.
+%
+We also say that $P$ is \emph{observable} if $P \pcong Q$ for some thread $Q$.
+An observable process is a process that exposes an action on a free channel name
+and therefore can interact through that channel, if put into some appropriate
+reduction context. We say that $P$ is \emph{reducible} if $P \red Q$ for some
+$Q$. A reducible process may perform a reduction step.
+
+Now we observe that if $P$ and $Q$ both act on the same channel $x$ in
+complementary ways, then their parallel composition $\Cut\x{P}{Q}$ must be able
+to reduce according to one of the principal cut reductions described in
+\Cref{tab:semantics}. Therefore, an alternative (and more convenient) way of
+formulating well formedness is the following:
+
+\begin{definition}[well formedness]
+    \label{def:well-formedness}
+    We say that $P$ is \emph{well formed} if $P \pcong
+    \ReductionContext[\Cut\x{Q_1}{Q_2}]$ where both $Q_1$ and $Q_2$ are threads
+    implies that $\Cut\x{Q_1}{Q_2}$ is either observable or reducible.
+\end{definition}
+
+In general a process may be well-formed even if it is unable to reduce. For
+example, $P \eqdef \Cut\x{\Wait\y.\Close\x}{\Wait\x.\Close\z}$ is irreducible
+but also well formed, because $P \pcong
+\Wait\y.\Cut\x{\Close\x}{\Wait\x.\Close\z}$ which is observable.
+
+\begin{theorem}[type safety]
+    \label{thm:type-safety}
+    If $\wtp{P}\Context$, then $P$ is well formed.
 \end{theorem}
 
-safety
+% \begin{theorem}[deadlock freedom]
+%     \label{thm:df}
+%     If $\wtp{P}\Context$, then $P$ is either observable or reducible.
+% \end{theorem}
+
+% safety

### 2025-07-25 paper/formalization.lagda.tex
agda code

\section{Agda Formalization}
\label{sec:formalization}

\begin{code}
data Type : Set where
  𝟘 𝟙 ⊥ ⊤ : Type
  ¡ ¿ : Type → Type
  _&_ _⊕_ _⊗_ _⅋_ : Type → Type → Type
\end{code}

\begin{code}
data Dual : Type → Type → Set where
  d-𝟘-⊤ : Dual 𝟘 ⊤
  d-⊤-𝟘 : Dual ⊤ 𝟘
  d-𝟙-⊥ : Dual 𝟙 ⊥
  d-⊥-𝟙 : Dual ⊥ 𝟙
  d-!-? : ∀{A B} → Dual A B → Dual (¡ A) (¿ B)
  d-?-! : ∀{A B} → Dual A B → Dual (¿ A) (¡ B)
  d-&-⊕ : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A & B) (A′ ⊕ B′)
  d-⊕-& : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⊕ B) (A′ & B′)
  d-⊗-⅋ : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⊗ B) (A′ ⅋ B′)
  d-⅋-⊗ : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⅋ B) (A′ ⊗ B′)
\end{code}

\begin{code}
dual-symm : ∀{A B} → Dual A B → Dual B A
\end{code}
\begin{code}[hide]
dual-symm d-𝟘-⊤ = d-⊤-𝟘
dual-symm d-⊤-𝟘 = d-𝟘-⊤
dual-symm d-𝟙-⊥ = d-⊥-𝟙
dual-symm d-⊥-𝟙 = d-𝟙-⊥
dual-symm (d-!-? p) = d-?-! (dual-symm p)
dual-symm (d-?-! p) = d-!-? (dual-symm p)
dual-symm (d-&-⊕ p q) = d-⊕-& (dual-symm p) (dual-symm q)
dual-symm (d-⊕-& p q) = d-&-⊕ (dual-symm p) (dual-symm q)
dual-symm (d-⊗-⅋ p q) = d-⅋-⊗ (dual-symm p) (dual-symm q)
dual-symm (d-⅋-⊗ p q) = d-⊗-⅋ (dual-symm p) (dual-symm q)
\end{code}

### 2025-07-25 paper/watch.sh

@@ -1,2 +1,2 @@
 #!/bin/sh
-exec watchexec -d 500ms -i main.tex agda --latex-dir=. --latex main.lagda.tex
+exec watchexec -d 500ms -i formalization.tex agda --latex-dir=. --latex formalization.lagda.tex

### 2025-07-25 paper/formalization.lagda.tex
types and contexts in agda

@@
 \label{sec:formalization}
+
+\begin{code}[hide]
+open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
+open import Relation.Binary.PropositionalEquality using (_≡_; refl; cong; cong₂)
+open import Data.List.Base using (List; []; _∷_; [_]; _++_)
+\end{code}
+
+\subsection{Type Representation}
+\label{sec:type-agda}
+
+The representation of types is standard, with one constructor for each of the
+forms described in \Cref{sec:calculus}.
 
@@
 \end{code}
+
+The notion of duality is formalized as a \emph{relation} \AgdaDatatype{Dual}
+such that $\AgdaDatatype{Dual}~A~B$ holds if and only if $A = \dual{B}$.
 
@@
 
+It is straightforward to prove that duality is a symmetric relation and that it
+behaves as an involution. From this we prove that it acts as the function
+$\dual\cdot$ defined in \Cref{sec:calculus}.
+
 \begin{code}
 dual-symm   : ∀{A B} → Dual A B → Dual B A
+dual-inv    : ∀{A B C} → Dual A B → Dual B C → A ≡ C
+dual-fun-r  : ∀{A B C} → Dual A B → Dual A C → B ≡ C
+dual-fun-l  : ∀{A B C} → Dual B A → Dual C A → B ≡ C
 \end{code}
@@
 \end{code}
+
+\begin{code}[hide]
+dual-inv d-𝟘-⊤ d-⊤-𝟘 = refl
+dual-inv d-⊤-𝟘 d-𝟘-⊤ = refl
+dual-inv d-𝟙-⊥ d-⊥-𝟙 = refl
+dual-inv d-⊥-𝟙 d-𝟙-⊥ = refl
+dual-inv (d-!-? p) (d-?-! q) = cong ¡ (dual-inv p q)
+dual-inv (d-?-! p) (d-!-? q) = cong ¿ (dual-inv p q)
+dual-inv (d-&-⊕ p q) (d-⊕-& r s) = cong₂ _&_ (dual-inv p r) (dual-inv q s)
+dual-inv (d-⊕-& p q) (d-&-⊕ r s) = cong₂ _⊕_ (dual-inv p r) (dual-inv q s)
+dual-inv (d-⊗-⅋ p q) (d-⅋-⊗ r s) = cong₂ _⊗_ (dual-inv p r) (dual-inv q s)
+dual-inv (d-⅋-⊗ p q) (d-⊗-⅋ r s) = cong₂ _⅋_ (dual-inv p r) (dual-inv q s)
+
+dual-fun-r d e = dual-inv (dual-symm d) e
+
+dual-fun-l d e = dual-inv d (dual-symm e)
+\end{code}
+
+\subsection{Context Representation}
+\label{sec:context-agda}
+
+We are going to adopt a nameless representation of channels whereby a channel is
+identified by its position in a typing context. This representation is akin to
+using De Bruijn indices, except that the index, instead of being represented
+explicitly by a natural number, is computable from the \emph{proof} that the
+type of the channel belong to the typing context.
+%
+Typing contexts are represented using \emph{lists} of types, where the
+(polymorphic) \AgdaDatatype{List} type is defined in Agda′s standard library.
+
+\begin{code}
+Context : Set
+Context = List Type
+\end{code}
+
+The most important operation concerning typing contexts is \emph{splitting}. The
+splitting of $\ContextC$ into $\ContextD$ and $\ContextE$, which is denoted by
+$\Splitting\ContextC\ContextD\ContextE$, represents the fact that $\ContextC$
+contains all the types contained in $\ContextD$ and $\ContextE$, preserving both
+their overall \emph{multiplicity} and also their relative \emph{order} within
+$\ContextD$ and $\ContextE$. A \emph{proof} of
+$\Splitting\ContextC\ContextD\ContextE$ shows how the types in $\ContextC$ are
+distributed in $\ContextD$ and $\ContextE$ from left to right.
+
+\begin{code}[hide]
+infix 4 _≃_+_ _≃_,_
+\end{code}
+\begin{code}
+data _≃_+_ : Context → Context → Context → Set where
+  split-e  : [] ≃ [] + []
+  split-l  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ A ∷ Δ + Θ
+  split-r  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ Δ + A ∷ Θ
+\end{code}
+
+As an example, below is the splitting of the context $[A, B, C, D]$ into $[A,
+D]$ and $[B, C]$. Note how the sequence of \AgdaInductiveConstructor{split-l}
+and \AgdaInductiveConstructor{split-r} applications determines where each of the
+types in $[A, B, C, D]$ ends up in $[A, D]$ and $[B, C]$.
+
+\begin{code}[hide]
+module _ where
+  postulate A B C D : Type
+\end{code}
+\begin{code}
+  splitting-example : (A ∷ B ∷ C ∷ D ∷ []) ≃ (A ∷ D ∷ []) + (B ∷ C ∷ [])
+  splitting-example = split-l (split-r (split-r (split-l split-e)))
+\end{code}
+
+It is often the case that the context $\ContextD$ in a splitting
+$\Splitting\ContextC\ContextD\ContextE$ is a singleton list $[A]$. For this
+particular case, we introduce a dedicated notation that allows us to represent
+this case in a more compact and readable way, as
+$\SimpleSplitting\ContextC{A}\ContextE$.
+
+\begin{code}
+_≃_,_ : Context → Type → Context → Set
+Γ ≃ A , Δ = Γ ≃ [ A ] + Δ
+\end{code}
+
+Context splitting enjoys a number of expected properties. In particular, it is
+easy to see that splitting is commutative and that the empty context is both a
+left and right unit of splitting.
+
+\begin{code}
++-comm    : ∀{Γ Δ Θ} → Γ ≃ Δ + Θ → Γ ≃ Θ + Δ
++-unit-l  : ∀{Γ} → Γ ≃ [] + Γ
++-unit-r  : ∀{Γ} → Γ ≃ Γ + []
+\end{code}
+\begin{code}[hide]
++-comm split-e = split-e
++-comm (split-l p) = split-r (+-comm p)
++-comm (split-r p) = split-l (+-comm p)
+
++-unit-l {[]} = split-e
++-unit-l {_ ∷ _} = split-r +-unit-l
+
++-unit-r = +-comm +-unit-l
+\end{code}
+
+Context splitting is also associative in a sense that is made precise below. If
+we write $\ContextD + \ContextE$ for some $\ContextC$ such that $Γ ≃ Δ + Θ$, then
+we can prove that $Γ_1 + (Γ_2 + Γ_3) = (Γ_1 + Γ_2) + Γ_3$.
+
+\begin{code}
++-assoc-r  : ∀{Γ Δ Θ Δ′ Θ′} → Γ ≃ Δ + Θ → Θ ≃ Δ′ + Θ′ →
+             ∃[ Γ′ ] Γ′ ≃ Δ + Δ′ × Γ ≃ Γ′ + Θ′
++-assoc-l  : ∀{Γ Δ Θ Δ′ Θ′} → Γ ≃ Δ + Θ → Δ ≃ Δ′ + Θ′ →
+             ∃[ Γ′ ] Γ′ ≃ Θ′ + Θ × Γ ≃ Δ′ + Γ′
+\end{code}
+\begin{code}[hide]
++-assoc-r split-e split-e = [] , split-e , split-e
++-assoc-r (split-l p) q with +-assoc-r p q
+... | _ , p′ , q′ = _ , split-l p′ , split-l q′
++-assoc-r (split-r p) (split-l q) with +-assoc-r p q
+... | _ , p′ , q′ = _ , split-r p′ , split-l q′
++-assoc-r (split-r p) (split-r q) with +-assoc-r p q
+... | _ , p′ , q′ = _ , p′ , split-r q′
+
++-assoc-l p q with +-assoc-r (+-comm p) (+-comm q)
+... | Δ , r , p′ = Δ , +-comm r , +-comm p′
+
++-empty-l : ∀{Γ Δ} → Γ ≃ [] + Δ → Γ ≡ Δ
++-empty-l split-e = refl
++-empty-l (split-r p) = cong (_ ∷_) (+-empty-l p)
+
++-sing-l : ∀{A B Γ} → [ A ] ≃ B , Γ → A ≡ B × Γ ≡ []
++-sing-l (split-l split-e) = refl , refl
+\end{code}
+
+\subsection{Context Permutations}
+\label{sec:permutation-agda}

### 2025-07-26 paper/formalization.lagda.tex
description of processes

@@
 \begin{code}[hide]
+open import Data.Bool using (Bool; if_then_else_)
 open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
@@
 \label{sec:permutation-agda}
+
+\begin{code}
+data _#_ : Context → Context → Set where
+  #refl  : ∀{Γ} → Γ # Γ
+  #here  : ∀{A B Γ} → (A ∷ B ∷ Γ) # (B ∷ A ∷ Γ)
+  #next  : ∀{A Γ Δ} → Γ # Δ → (A ∷ Γ) # (A ∷ Δ)
+  #tran  : ∀{Γ Δ Θ} → Γ # Δ → Δ # Θ → Γ # Θ
+
+#sym : ∀{Γ Δ} → Γ # Δ → Δ # Γ
+#sym #refl = #refl
+#sym #here = #here
+#sym (#next π) = #next (#sym π)
+#sym (#tran π π′) = #tran (#sym π′) (#sym π)
+
+#empty-inv : ∀{Γ} → [] # Γ → Γ ≡ []
+#empty-inv #refl = refl
+#empty-inv (#tran π π′) rewrite #empty-inv π | #empty-inv π′ = refl
+
+#singleton-inv : ∀{A Γ} → [ A ] # Γ → Γ ≡ [ A ]
+#singleton-inv {Γ = Γ} #refl = refl
+#singleton-inv {Γ = Γ} (#next π) rewrite #empty-inv π = refl
+#singleton-inv {Γ = Γ} (#tran π π′) rewrite #singleton-inv π | #singleton-inv π′ = refl
+
+#rot : ∀{A B C Γ} → (A ∷ B ∷ C ∷ Γ) # (C ∷ A ∷ B ∷ Γ)
+#rot = #tran (#next #here) #here
+
+#cons : ∀{A Γ Δ} → Γ ≃ A , Δ → (A ∷ Δ) # Γ
+#cons (split-l p) with +-empty-l p
+... | refl = #refl
+#cons (split-r p) = #tran #here (#next (#cons p))
+
+#split : ∀{Γ Γ₁ Γ₂ Δ} → Γ # Δ → Γ ≃ Γ₁ + Γ₂ → ∃[ Δ₁ ] ∃[ Δ₂ ] (Δ ≃ Δ₁ + Δ₂ × Γ₁ # Δ₁ × Γ₂ # Δ₂)
+#split #refl p = _ , _ , p , #refl , #refl
+#split (#next π) (split-l p) with #split π p
+... | Δ₁ , Δ₂ , q , π₁ , π₂ = _ ∷ Δ₁ , Δ₂ , split-l q , #next π₁ , π₂
+#split (#next π) (split-r p) with #split π p
+... | Δ₁ , Δ₂ , q , π₁ , π₂ = Δ₁ , _ ∷ Δ₂ , split-r q , π₁ , #next π₂
+#split #here (split-l (split-l p)) = _ , _ , split-l (split-l p) , #here , #refl
+#split #here (split-l (split-r p)) = _ , _ , split-r (split-l p) , #refl , #refl
+#split #here (split-r (split-l p)) = _ , _ , split-l (split-r p) , #refl , #refl
+#split #here (split-r (split-r p)) = _ , _ , split-r (split-r p) , #refl , #here
+#split (#tran π π′) p with #split π p
+... | Θ₁ , Θ₂ , p′ , π₁ , π₂ with #split π′ p′
+... | Δ₁ , Δ₂ , q , π₁′ , π₂′ = Δ₁ , Δ₂ , q , #tran π₁ π₁′ , #tran π₂ π₂′
+
+#one+ : ∀{A Γ Γ′ Δ} → Γ # Δ → Γ ≃ A , Γ′ → ∃[ Δ′ ] (Δ ≃ A , Δ′ × Γ′ # Δ′)
+#one+ π p with #split π p
+... | _ , _ , q , π₁ , π₂ rewrite #singleton-inv π₁ = _ , q , π₂
+
+#shift : ∀{Γ A Δ} → (Γ ++ A ∷ Δ) # (A ∷ Γ ++ Δ)
+#shift {[]} = #refl
+#shift {B ∷ Γ} = #tran (#next #shift) #here
+
++++# : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → (Γ₁ ++ Γ₂) # Γ
++++# split-e = #refl
++++# (split-l p) = #next (+++# p)
++++# (split-r p) = #tran #shift (#next (+++# p))
+
+#left : ∀{Γ Δ Θ} → Γ # Δ → (Θ ++ Γ) # (Θ ++ Δ)
+#left {Θ = []} π = π
+#left {Θ = _ ∷ Θ} π = #next (#left π)
+
+data Un : Context → Set where
+  un-[]  : Un []
+  un-∷   : ∀{A Γ} → Un Γ → Un (¿ A ∷ Γ)
+
+#un : ∀{Γ Δ} → Γ # Δ → Un Γ → Un Δ
+#un #refl un = un
+#un (#next π) (un-∷ un) = un-∷ (#un π un)
+#un #here (un-∷ (un-∷ un)) = un-∷ (un-∷ un)
+#un (#tran π π′) un = #un π′ (#un π un)
+
+#un+ : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → Un Γ₁ → Un Γ₂ → Un Γ
+#un+ split-e un-[] un-[] = un-[]
+#un+ (split-l p) (un-∷ un₁) un₂ = un-∷ (#un+ p un₁ un₂)
+#un+ (split-r p) un₁ (un-∷ un₂) = un-∷ (#un+ p un₁ un₂)
+\end{code}
+
+\subsection{Process Representation}
+\label{sec:process-agda}
+
+CO-DE BRUIJN INDICES
+
+We adopt an \emph{intrinsically typed} representation of processes, so that a
+process is also a proof derivation showing that the process is well typed. This
+choice increases the effort in the definition of the datatypes for processes and
+their operational semantics, but it pays off in the rest of the formalization
+for a number of reasons:
+
+\begin{itemize}
+\item channel names are encoded in the context splitting proofs occurring within
+  processes, hence they do not need to be represented explicitly.
+\item processes and typing rules are conflated in the same datatype, thus
+  reducing the overall number of datatypes in the formalization;
+\item the typing preservation results are embedded in the definition of the
+  operational semantics of processes and do not need to be proved separately.
+\end{itemize}
+
+We now illustrate the definition of the \AgdaDatatype{Process} datatype and
+describe each constructor in detail. The datatype is indexed by a typing
+context:
+
+\begin{AgdaAlign}
+\begin{code}
+data Process : Context → Set where
+   link      : ∀{Γ A B} (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) → Process Γ
+\end{code}
+
+A \AgdaInductiveConstructor{link} process is well typed in a typing context with
+exactly two types $A$ and $B$ which must be related by duality. We use a proof
+$p$ that $\Context \simeq [A] + [B]$ instead of requiring $\Context$ to be
+simply $[A,B]$ so that it becomes straightforward to define \SLink using
+\AgdaFunction{dual-symm} and \AgdaFunction{+-comm}.
+
+\begin{code}
+   fail      : ∀{Γ Δ} (p : Γ ≃ ⊤ , Δ) → Process Γ
+\end{code}
+
+A \AgdaInductiveConstructor{fail} process simply requires the presence of $\Top$
+in the typing context.
+
+\begin{code}
+   close     : Process [ 𝟙 ]
+\end{code}
+
+A \AgdaInductiveConstructor{close} process is well typed in the singleton
+context $[\One]$.
+
+\begin{code}
+   wait      : ∀{Γ Δ} (p : Γ ≃ ⊥ , Δ) → Process Δ → Process Γ
+\end{code}
+
+A \AgdaInductiveConstructor{wait} process is well typed in any context
+$\ContextC$ that contains $\Bot$ and whose residual $\ContextD$ allows the
+continuation to be well typed.
+
+\begin{code}
+   select    : ∀{Γ Δ A B} (x : Bool) (p : Γ ≃ A ⊕ B , Δ) →
+               Process ((if x then A else B) ∷ Δ) → Process Γ
+\end{code}
+
+We represent the tags $\InTag_1$ and $\InTag_2$ with Agda′s boolean values
+\AgdaInductiveConstructor{True} and \AgdaInductiveConstructor{False}. In this
+way we can use a single constructor \AgdaInductiveConstructor{select} for the
+selections $\Select\x{\InTag_i}.P$. According to the semantics of \Calculus, the
+channel on which the selection is performed is consumed and the continuation
+process uses a fresh continuation channel. As a consequence, the type of the
+continuation channel, which is either $A$ or $B$ depending on the value of the
+tag $x$, is \emph{prepended} in front of the typing context used to type the
+continuation process.
+
+\begin{code}
+   case      : ∀{Γ Δ A B} (p : Γ ≃ A & B , Δ) →
+               Process (A ∷ Δ) → Process (B ∷ Δ) → Process Γ
+\end{code}
+
+A \AgdaInductiveConstructor{case} process has two possible continuations. Note
+again that the continuation channel, which is received along with the tag, has
+either type $A$ or type $B$ depending on the tag and its type is prepended to
+the typing context of the continuation process since it is the newest channel.
+
+\begin{code}
+   fork      : ∀{Γ Δ Γ₁ Γ₂ A B} (p : Γ ≃ A ⊗ B , Δ) (q : Δ ≃ Γ₁ + Γ₂) →
+               Process (A ∷ Γ₁) → Process (B ∷ Γ₂) → Process Γ
+\end{code}
+
+A \AgdaInductiveConstructor{fork} process creates two fresh continuation
+channels and spaws two continuation processes. Since context splitting is a
+ternary relation, we need two splitting proofs $p$ and $q$ to isolate the type
+$A \Ten B$ of channel on which the pair of continuation channels is sent and to
+distribute the remaining channel across the two continuation processes.
+
+\begin{code}
+   join      : ∀{Γ Δ A B} (p : Γ ≃ A ⅋ B , Δ) →
+               Process (B ∷ A ∷ Δ) → Process Γ
+\end{code}
+
+The \AgdaInductiveConstructor{join} process has a single continuation which uses
+the pair of continuation channels received from a channel of type $A \Par B$.
+Note that $B$ is prepended in front of $A$, somewhat implying that the second
+continuation is bound \emph{after} the first even though, from a technical
+standpoint, both continuations are bound simultaneously. We could have prepended
+the types $A$ and $B$ also in the opposite order, provided that the reduction
+rule \RFork (formalized in \Cref{sec:semantics-agda}) is suitably adjusted.
+
+\begin{code}
+   server    : ∀{Γ Δ A} (p : Γ ≃ ¡ A , Δ) (un : Un Δ) →
+               Process (A ∷ Δ) → Process Γ
+\end{code}
+
+A \AgdaInductiveConstructor{server} constructor introduces a channel of type
+$\OfCourse A$ given a continuation process that consumes a channel of type $A$
+and a proof $\mathit{un}$ that every other channel in $\ContextD$ is
+unrestricted.
+
+\begin{code}
+   client    : ∀{Γ Δ A} (p : Γ ≃ ¿ A , Δ) → Process (A ∷ Δ) → Process Γ
+   weaken    : ∀{Γ Δ A} (p : Γ ≃ ¿ A , Δ) → Process Δ → Process Γ
+   contract  : ∀{Γ Δ A} (p : Γ ≃ ¿ A , Δ) → Process (¿ A ∷ ¿ A ∷ Δ) → Process Γ
+\end{code}
+
+The constructors \AgdaInductiveConstructor{client},
+\AgdaInductiveConstructor{weaken} and \AgdaInductiveConstructor{contract}
+introduce a channel of type $\WhyNot A$ and follow the same pattern of the
+previous process forms. Note that, in the case of
+\AgdaInductiveConstructor{contract}, the contraction of $\WhyNot A$ applies to
+the first two channels in the typing context of the continuation. This is
+consistent with the fact that \Calculus uses an explicit prefix
+$\Contract\x\y\z$ that models contraction and binds $y$ and $z$ in the
+continuation process. In a pen-and-paper presentation, like the one by
+Wadler~\citep{Wadler14}, it is arguably more convenient to have weaking and
+contraction typing rules without the corresponding process forms.
+
+\begin{code}
+   cut       : ∀{Γ Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) →
+               Process (A ∷ Γ₁) → Process (B ∷ Γ₂) → Process Γ
+\end{code}
+
+A \AgdaInductiveConstructor{cut} process incorporates a duality proof for the
+peer endpoints of the channel being restricted and a splitting proof that
+distributes the free channels among the two parallel processes.
+
+\end{AgdaAlign}
+
+\begin{code}
+#process : ∀{Γ Δ} → Γ # Δ → Process Γ → Process Δ
+\end{code}
+\begin{code}[hide]
+#process π (link d p) with #one+ π p
+... | Δ′ , q , π′ with #singleton-inv π′
+... | refl = link d q
+#process π close with #singleton-inv π
+... | refl = close
+#process π (fail p) with #one+ π p
+... | Δ′ , q , π′ = fail q
+#process π (wait p P) with #one+ π p
+... | Δ′ , q , π′ = wait q (#process π′ P)
+#process π (select x p P) with #one+ π p
+... | Δ′ , q , π′ = select x q (#process (#next π′) P)
+#process π (case p P Q) with #one+ π p
+... | Δ′ , q , π′ = case q (#process (#next π′) P) (#process (#next π′) Q)
+#process π (fork p q P Q) with #one+ π p
+... | Δ′ , p′ , π′ with #split π′ q
+... | Δ₁ , Δ₂ , q′ , π₁ , π₂ = fork p′ q′ (#process (#next π₁) P) (#process (#next π₂) Q)
+#process π (join p P) with #one+ π p
+... | Δ′ , q , π′ = join q (#process (#next (#next π′)) P)
+#process π (cut d p P Q) with #split π p
+... | Δ₁ , Δ₂ , q , π₁ , π₂ = cut d q (#process (#next π₁) P) (#process (#next π₂) Q)
+#process π (server p un P) with #one+ π p
+... | Δ′ , q , π′ = server q (#un π′ un) (#process (#next π′) P)
+#process π (client p P) with #one+ π p
+... | Δ′ , q , π′ = client q (#process (#next π′) P)
+#process π (weaken p P) with #one+ π p
+... | Δ′ , q , π′ = weaken q (#process π′ P)
+#process π (contract p P) with #one+ π p
+... | Δ′ , q , π′ = contract q (#process (#next (#next π′)) P)
+\end{code}
+
+\subsection{Operational Semantics}
+\label{sec:semantics-agda}

### 2025-07-26 paper/formalization.lagda.tex
completed structural congruence

@@
 \begin{code}[hide]
-open import Data.Bool using (Bool; if_then_else_)
+open import Data.Bool using (Bool; true; false; if_then_else_)
 open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
@@
 We represent the tags $\InTag_1$ and $\InTag_2$ with Agda′s boolean values
-\AgdaInductiveConstructor{True} and \AgdaInductiveConstructor{False}. In this
+\AgdaInductiveConstructor{true} and \AgdaInductiveConstructor{false}. In this
 way we can use a single constructor \AgdaInductiveConstructor{select} for the
@@
 and a proof $\mathit{un}$ that every other channel in $\ContextD$ is
-unrestricted.
+unrestricted.\Luca{Introdurre terminologia e dire che unrestricted vuol dire che
+il canale può essere ``duplicato'' e scartato.}
 
@@
 \label{sec:semantics-agda}
+
+We formalize structural precongruence as a binary relation between processes
+that are well typed in the \emph{same} typing context. This entails that
+structural precongruence preserves typing by definition.
+
+\begin{AgdaAlign}
+\begin{code}
+data _⊒_ : ∀{Γ} → Process Γ → Process Γ → Set where
+\end{code}
+
+The datatype for $\pcong$ has one constructor for each of the structural
+precongruence rules in \Cref{tab:semantics}. Since many aspects recur
+repeatedly, we will focus on just a few representative rules starting from
+\SComm.
+
+\begin{code}
+  s-comm :
+    ∀{Γ Γ₁ Γ₂ A B P Q} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) →
+    cut d p P Q ⊒ cut (dual-symm d) (+-comm p) Q P
+\end{code}
+
+The constructor \AgdaInductiveConstructor{s-comm} models the commutativity
+property of parallel composition. We use \AgdaFunction{dual-symm} and
+\AgdaFunction{+-comm} to compute the proofs of the relation
+$\AgdaDatatype{Dual}~B~A$ and of the splitting $\ContextC \simeq \Context_2 +
+\Context_1$ from $d$ and $p$ respectively.
+
+\begin{code}[hide]
+  s-link :
+    ∀{Γ A B}
+    (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) →
+    link d p ⊒ link (dual-symm d) (+-comm p)
+
+  s-fail :
+    ∀{Γ Γ₁ Γ₂ Δ A B P} (d : Dual A B)
+    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ⊤ , Δ) →
+    let _ , _ , q′ = +-assoc-l p q in
+    cut d p (fail (split-r q)) P ⊒ fail q′
+\end{code}
+
+The constructor \AgdaInductiveConstructor{s-wait} models the \SWait rule:
+
+\begin{code}
+  s-wait :
+    ∀{Γ Γ₁ Γ₂ Δ A B} {P : Process (A ∷ Δ)} {Q : Process (B ∷ Γ₂)}
+    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ⊥ , Δ) →
+    let _ , p′ , q′ = +-assoc-l p q in
+    cut d p (wait (split-r q) P) Q ⊒ wait q′ (cut d p′ P Q)
+\end{code}
+
+There are two non-trivial aspects worth commenting. The first one concerns the
+proof $(\AgdaInductiveConstructor{split-r}~q)$ used in the
+\AgdaInductiveConstructor{wait} process before structural precongruence is
+applied. To understand the meaning of this proof, we must recall three key
+elements:
+\begin{enumerate}
+  \item
+    $(\AgdaInductiveConstructor{wait}~(\AgdaInductiveConstructor{split-r}~q)~P)$
+    is a direct sub-process of the \AgdaInductiveConstructor{cut}, and therefore
+    it is meant to be well typed in the context
+    $A~\AgdaInductiveConstructor{∷}~\Context_1$.
+  \item Being a \AgdaInductiveConstructor{wait} process, such context must
+    contain a $\Bot$ type as per the typing rule \WaitRule. That is
+    $A~\AgdaInductiveConstructor{∷}~\Context_1 \simeq [\Bot] +
+    A~\AgdaInductiveConstructor{∷}~\Delta$ for some $\Delta$.
+  \item The \SWait rule is applicable only provided that the channel restricted
+    by the cut (say $x$, of type $A$) is different from the channel (say $y$, of
+    type $\Bot$) consumed by the \AgdaInductiveConstructor{wait} process.
+    %
+    We enforce the side condition $x \ne y$ of \SWait imposing that the type $A$
+    in front of $A~\AgdaInductiveConstructor{∷}~\Context_1$ necessarily comes
+    from the right component of the splitting $[\Bot] +
+    A~\AgdaInductiveConstructor{∷}~\Delta$ through the use of
+    \AgdaInductiveConstructor{split-r}.
+\end{enumerate}
+
+The other aspect that is worth commenting concerns the rearrangement of the
+splittings in the process after the application of structural precongruence.
+Overall, $p$ and $q$ prove the splittings $([\Bot] + \ContextD) + \ContextC_2$,
+but the precongruence rule requires this splitting to be rearranged as $[\Bot] +
+(\ContextD + \ContextC_2)$. That is, we need to apply the left-to-right
+associativity property of context splitting which we called
+\AgdaFunction{+-assoc-l} in \Cref{sec:context-agda}. The nested
+\AgdaKeyword{let}-\AgdaKeyword{in} allows us to pattern match the result of the
+application $\AgdaFunction{+-assoc-l}~p~q$ and to extract the new proofs $p'$
+and $q'$ for the rearranged splittings.
+
+The constructors \AgdaInductiveConstructor{s-select-l} and
+\AgdaInductiveConstructor{s-select-r} model \SSelect when the selected tag is
+respectively \AgdaInductiveConstructor{true} and
+\AgdaInductiveConstructor{false}.
+
+\begin{code}
+  s-select-l :
+    ∀{Γ Γ₁ Γ₂ Δ A B C D} {P : Process (C ∷ A ∷ Δ)} {Q : Process (B ∷ Γ₂)}
+    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ C ⊕ D , Δ) →
+    let _ , p′ , q′ = +-assoc-l p q in
+    cut d p (select true (split-r q) P) Q ⊒
+    select true q′ (cut d (split-l p′) (#process #here P) Q)
+\end{code}
+
+Here the process
+$(\AgdaInductiveConstructor{select}~\AgdaInductiveConstructor{true}~(\AgdaInductiveConstructor{split-r}~q)~P)$,
+which is found under a cut for $x : A$, is using some channel $y : C \Plus D$ to
+send the tag $\AgdaInductiveConstructor{true}$.
+%
+Unlike \AgdaInductiveConstructor{wait}, which consumes the channel it operates
+on, the continuation process $P$ will be using a fresh continuation channel $z :
+C$. Therefore, $P$ is required to be well typed in the context
+$C~\AgdaInductiveConstructor{∷}~A~\AgdaInductiveConstructor{∷}~\ContextD$, where
+the type of the continuation sits on top of the type of the restricted channel.
+%
+After structural precongruence is applied, however, the type $C$ of the
+continuation channel $z$ ends up underneath that of the restricted channel $x$,
+because now the two channels are introduced in the opposite order. Therefore, we
+are in a situation where we need to turn $P$ into a process that is well typed
+in the context
+$A~\AgdaInductiveConstructor{∷}~C~\AgdaInductiveConstructor{∷}~\ContextD$. This
+is achieved applying the function \AgdaFunction{\#process} to the
+\AgdaInductiveConstructor{\#here} permutation, which swaps the two topmost types
+in a typing context, and finally to the process $P$.
+
+\begin{code}[hide]
+  s-select-r :
+    ∀{Γ Γ₁ Γ₂ Δ A B C D}
+    {P : Process (D ∷ A ∷ Δ)}
+    {Q : Process (B ∷ Γ₂)}
+    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ C ⊕ D , Δ) →
+    let _ , p′ , q′ = +-assoc-l p q in
+    cut d p (select false (split-r q) P) Q ⊒
+    select false q′ (cut d (split-l p′) (#process #here P) Q)
+
+  s-case :
+    ∀{Γ A B A₁ A₂ Γ₁ Γ₂ Δ}
+    {P : Process (A₁ ∷ A ∷ Δ)}
+    {Q : Process (A₂ ∷ A ∷ Δ)}
+    {R : Process (B ∷ Γ₂)}
+    (d : Dual A B)
+    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ A₁ & A₂ , Δ) →
+    let _ , p′ , q′ = +-assoc-l p q in
+    cut d p (case (split-r q) P Q) R ⊒
+    case q′ (cut d (split-l p′) (#process #here P) R)
+            (cut d (split-l p′) (#process #here Q) R)
+\end{code}
+
+\emph{En passant}, we also show the modeling of the \SForkL rule, which is
+interesting because of its complex side conditions:
+
+\begin{code}
+  s-fork-l :
+    ∀{Γ Γ₁ Γ₂ Δ Δ₁ Δ₂ A B C D}
+    {P : Process (C ∷ A ∷ Δ₁)} {Q : Process (D ∷ Δ₂)} {R : Process (B ∷ Γ₂)}
+    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ C ⊗ D , Δ) (r : Δ ≃ Δ₁ + Δ₂) →
+    let _ , p′ , q′ = +-assoc-l p q in
+    let _ , p′′ , r′ = +-assoc-l p′ r in
+    let _ , q′′ , r′′ = +-assoc-r r′ (+-comm p′′) in
+    cut d p (fork (split-r q) (split-l r) P Q) R ⊒
+    fork q′ r′′ (cut d (split-l q′′) (#process #here P) R) Q
+\end{code}
+
+Recall from \Cref{tab:semantics} that we want to apply this rule on a process of
+the form $\Cut\x{\Fork\y\u\v{P}{Q}}{R}$ when $x \in \fn{P}$. We capture this
+condition by means of the splitting $(\AgdaInductiveConstructor{split-l}~r)$,
+meaning that the type $A$ of $x$ ends up in the typing context for $P$ and not
+in the one for $Q$.
+%
+The symmetric rule \SForkR is modeled by another constructor
+\AgdaInductiveConstructor{s-fork-r}, which is not shown here but is similar to
+\AgdaInductiveConstructor{s-fork-l} except that
+$(\AgdaInductiveConstructor{split-l}~r)$ is replaced by
+$(\AgdaInductiveConstructor{split-r}~r)$.
+
+\begin{code}[hide]
+  s-fork-r :
+    ∀{Γ Γ₁ Γ₂ Δ Δ₁ Δ₂ A B C D}
+    {P : Process (C ∷ Δ₁)}
+    {Q : Process (D ∷ A ∷ Δ₂)}
+    {R : Process (B ∷ Γ₂)}
+    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ C ⊗ D , Δ)
+    (r : Δ ≃ Δ₁ + Δ₂) →
+    let _ , p′ , q′ = +-assoc-l p q in
+    let _ , p′′ , r′ = +-assoc-l p′ r in
+    cut d p (fork (split-r q) (split-r r) P Q) R ⊒
+    fork q′ r′ P (cut d (split-l p′′) (#process #here Q) R)
+
+  s-join :
+    ∀{Γ Γ₁ Γ₂ Δ A B C D}
+    {P : Process (D ∷ C ∷ A ∷ Δ)}
+    {Q : Process (B ∷ Γ₂)}
+    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ C ⅋ D , Δ) →
+    let _ , p′ , q′ = +-assoc-l p q in
+    cut d p (join (split-r q) P) Q ⊒
+    join q′ (cut d (split-l (split-l p′)) (#process #rot P) Q)
+
+  s-server :
+    ∀{Γ A B C Γ₁ Γ₂ Δ₁}
+    {P : Process (C ∷ ¿ A ∷ Δ₁)}
+    {Q : Process (B ∷ Γ₂)}
+    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ¡ C , Δ₁) (r : Γ₂ ≃ [] + Γ₂)
+    (un₁ : Un Δ₁) (un₂ : Un Γ₂) →
+    let _ , p′ , q′ = +-assoc-l p q in
+    cut (d-?-! d) p (server (split-r q) (un-∷ un₁) P) (server (split-l r) un₂ Q) ⊒
+    server q′ (#un+ p′ un₁ un₂) (cut (d-?-! d) (split-l p′) (#process #here P) (server (split-l r) un₂ Q))
+
+  s-client :
+    ∀{Γ A B C Γ₁ Γ₂ Δ}
+    {P : Process (C ∷ A ∷ Δ)}
+    {Q : Process (B ∷ Γ₂)}
+    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ¿ C , Δ) →
+    let _ , p′ , q′ = +-assoc-l p q in
+    cut d p (client (split-r q) P) Q ⊒
+    client q′ (cut d (split-l p′) (#process #here P) Q)
+
+  s-weaken :
+    ∀{Γ A B C Γ₁ Γ₂ Δ}
+    {P : Process (A ∷ Δ)}
+    {Q : Process (B ∷ Γ₂)}
+    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ¿ C , Δ) →
+    let _ , p′ , q′ = +-assoc-l p q in
+    cut d p (weaken (split-r q) P) Q ⊒
+    weaken q′ (cut d p′ P Q)
+
+  s-contract :
+    ∀{Γ A B C Γ₁ Γ₂ Δ}
+    {P : Process (¿ C ∷ ¿ C ∷ A ∷ Δ)}
+    {Q : Process (B ∷ Γ₂)}
+    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ¿ C , Δ) →
+    let _ , p′ , q′ = +-assoc-l p q in
+    cut d p (contract (split-r q) P) Q ⊒
+    contract q′ (cut d (split-l (split-l p′)) (#process #rot P) Q)
+\end{code}
+
+Finally, in \Cref{sec:semantics} we have colloquially defined $\pcong$ as a
+``\emph{pre-congruence}'', implicitly implying that it is reflexive, transitive,
+and that it is is preserved by some forms of calculus. In the formalization we
+have to be precise and we introduce specific rules:
+
+\begin{code}
+  s-refl  : ∀{Γ} {P : Process Γ} → P ⊒ P
+  s-tran  : ∀{Γ} {P Q R : Process Γ} → P ⊒ Q → Q ⊒ R → P ⊒ R
+  s-cong  : ∀{Γ Γ₁ Γ₂ A A′} {P Q : Process (A ∷ Γ₁)} {P′ Q′ : Process (A′ ∷ Γ₂)}
+            (d : Dual A A′) (p : Γ ≃ Γ₁ + Γ₂) → P ⊒ Q → P′ ⊒ Q′ →
+            cut d p P P′ ⊒ cut d p Q Q′
+\end{code}
+
+Note that we define a single congruence rule \AgdaInductiveConstructor{s-cong}
+that allows us to apply $\pcong$ within cuts, but not underneath prefixes. This
+limited form of pre-congruence turns out to be sufficient for the development
+that follows.
+\end{AgdaAlign}

### 2025-07-26 paper/watch.sh
++

@@ -1,2 +1,2 @@
 #!/bin/sh
-exec watchexec -d 500ms -i formalization.tex agda --latex-dir=. --latex formalization.lagda.tex
+exec watchexec -d 500ms -e .lagda.tex agda --latex-dir=. --latex formalization.lagda.tex

### 2025-07-26 paper/Context.lagda.tex
split formalization into modules

--- 2025-07-25 paper/formalization.lagda.tex
@@
-\section{Agda Formalization}
-\label{sec:formalization}
-
 \begin{code}[hide]
+open import Data.Bool using (Bool; true; false; if_then_else_)
 open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
@@
 open import Data.List.Base using (List; []; _∷_; [_]; _++_)
-\end{code}
 
-\subsection{Type Representation}
-\label{sec:type-agda}
-
-The representation of types is standard, with one constructor for each of the
-forms described in \Cref{sec:calculus}.
-
-\begin{code}
-data Type : Set where
-  𝟘 𝟙 ⊥ ⊤          : Type
-  ¡ ¿              : Type → Type
-  _&_ _⊕_ _⊗_ _⅋_  : Type → Type → Type
-\end{code}
-
-The notion of duality is formalized as a \emph{relation} \AgdaDatatype{Dual}
-such that $\AgdaDatatype{Dual}~A~B$ holds if and only if $A = \dual{B}$.
-
-\begin{code}
-data Dual : Type → Type → Set where
-  d-𝟘-⊤  : Dual 𝟘 ⊤
-  d-⊤-𝟘  : Dual ⊤ 𝟘
-  d-𝟙-⊥  : Dual 𝟙 ⊥
-  d-⊥-𝟙  : Dual ⊥ 𝟙
-  d-!-?  : ∀{A B} → Dual A B → Dual (¡ A) (¿ B)
-  d-?-!  : ∀{A B} → Dual A B → Dual (¿ A) (¡ B)
-  d-&-⊕  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A & B) (A′ ⊕ B′)
-  d-⊕-&  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⊕ B) (A′ & B′)
-  d-⊗-⅋  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⊗ B) (A′ ⅋ B′)
-  d-⅋-⊗  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⅋ B) (A′ ⊗ B′)
-\end{code}
-
-It is straightforward to prove that duality is a symmetric relation and that it
-behaves as an involution. From this we prove that it acts as the function
-$\dual\cdot$ defined in \Cref{sec:calculus}.
-
-\begin{code}
-dual-symm   : ∀{A B} → Dual A B → Dual B A
-dual-inv    : ∀{A B C} → Dual A B → Dual B C → A ≡ C
-dual-fun-r  : ∀{A B C} → Dual A B → Dual A C → B ≡ C
-dual-fun-l  : ∀{A B C} → Dual B A → Dual C A → B ≡ C
-\end{code}
-\begin{code}[hide]
-dual-symm d-𝟘-⊤ = d-⊤-𝟘
-dual-symm d-⊤-𝟘 = d-𝟘-⊤
-dual-symm d-𝟙-⊥ = d-⊥-𝟙
-dual-symm d-⊥-𝟙 = d-𝟙-⊥
-dual-symm (d-!-? p) = d-?-! (dual-symm p)
-dual-symm (d-?-! p) = d-!-? (dual-symm p)
-dual-symm (d-&-⊕ p q) = d-⊕-& (dual-symm p) (dual-symm q)
-dual-symm (d-⊕-& p q) = d-&-⊕ (dual-symm p) (dual-symm q)
-dual-symm (d-⊗-⅋ p q) = d-⅋-⊗ (dual-symm p) (dual-symm q)
-dual-symm (d-⅋-⊗ p q) = d-⊗-⅋ (dual-symm p) (dual-symm q)
-\end{code}
-
-\begin{code}[hide]
-dual-inv d-𝟘-⊤ d-⊤-𝟘 = refl
-dual-inv d-⊤-𝟘 d-𝟘-⊤ = refl
-dual-inv d-𝟙-⊥ d-⊥-𝟙 = refl
-dual-inv d-⊥-𝟙 d-𝟙-⊥ = refl
-dual-inv (d-!-? p) (d-?-! q) = cong ¡ (dual-inv p q)
-dual-inv (d-?-! p) (d-!-? q) = cong ¿ (dual-inv p q)
-dual-inv (d-&-⊕ p q) (d-⊕-& r s) = cong₂ _&_ (dual-inv p r) (dual-inv q s)
-dual-inv (d-⊕-& p q) (d-&-⊕ r s) = cong₂ _⊕_ (dual-inv p r) (dual-inv q s)
-dual-inv (d-⊗-⅋ p q) (d-⅋-⊗ r s) = cong₂ _⊗_ (dual-inv p r) (dual-inv q s)
-dual-inv (d-⅋-⊗ p q) (d-⊗-⅋ r s) = cong₂ _⅋_ (dual-inv p r) (dual-inv q s)
-
-dual-fun-r d e = dual-inv (dual-symm d) e
-
-dual-fun-l d e = dual-inv d (dual-symm e)
+open import Type
 \end{code}
@@
 +-sing-l (split-l split-e) = refl , refl
+
+data Un : Context → Set where
+  un-[]  : Un []
+  un-∷   : ∀{A Γ} → Un Γ → Un (¿ A ∷ Γ)
 \end{code}
-
-\subsection{Context Permutations}
-\label{sec:permutation-agda}

### 2025-07-26 paper/Permutations.lagda.tex
split formalization into modules

--- 2025-07-25 paper/formalization.lagda.tex
@@
 open import Data.List.Base using (List; []; _∷_; [_]; _++_)
-\end{code}
 
-\subsection{Type Representation}
-\label{sec:type-agda}
-
-The representation of types is standard, with one constructor for each of the
-forms described in \Cref{sec:calculus}.
-
-\begin{code}
-data Type : Set where
-  𝟘 𝟙 ⊥ ⊤          : Type
-  ¡ ¿              : Type → Type
-  _&_ _⊕_ _⊗_ _⅋_  : Type → Type → Type
-\end{code}
-
-The notion of duality is formalized as a \emph{relation} \AgdaDatatype{Dual}
-such that $\AgdaDatatype{Dual}~A~B$ holds if and only if $A = \dual{B}$.
-
-\begin{code}
-data Dual : Type → Type → Set where
-  d-𝟘-⊤  : Dual 𝟘 ⊤
-  d-⊤-𝟘  : Dual ⊤ 𝟘
-  d-𝟙-⊥  : Dual 𝟙 ⊥
-  d-⊥-𝟙  : Dual ⊥ 𝟙
-  d-!-?  : ∀{A B} → Dual A B → Dual (¡ A) (¿ B)
-  d-?-!  : ∀{A B} → Dual A B → Dual (¿ A) (¡ B)
-  d-&-⊕  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A & B) (A′ ⊕ B′)
-  d-⊕-&  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⊕ B) (A′ & B′)
-  d-⊗-⅋  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⊗ B) (A′ ⅋ B′)
-  d-⅋-⊗  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⅋ B) (A′ ⊗ B′)
-\end{code}
-
-It is straightforward to prove that duality is a symmetric relation and that it
-behaves as an involution. From this we prove that it acts as the function
-$\dual\cdot$ defined in \Cref{sec:calculus}.
-
-\begin{code}
-dual-symm   : ∀{A B} → Dual A B → Dual B A
-dual-inv    : ∀{A B C} → Dual A B → Dual B C → A ≡ C
-dual-fun-r  : ∀{A B C} → Dual A B → Dual A C → B ≡ C
-dual-fun-l  : ∀{A B C} → Dual B A → Dual C A → B ≡ C
-\end{code}
-\begin{code}[hide]
-dual-symm d-𝟘-⊤ = d-⊤-𝟘
-dual-symm d-⊤-𝟘 = d-𝟘-⊤
-dual-symm d-𝟙-⊥ = d-⊥-𝟙
-dual-symm d-⊥-𝟙 = d-𝟙-⊥
-dual-symm (d-!-? p) = d-?-! (dual-symm p)
-dual-symm (d-?-! p) = d-!-? (dual-symm p)
-dual-symm (d-&-⊕ p q) = d-⊕-& (dual-symm p) (dual-symm q)
-dual-symm (d-⊕-& p q) = d-&-⊕ (dual-symm p) (dual-symm q)
-dual-symm (d-⊗-⅋ p q) = d-⅋-⊗ (dual-symm p) (dual-symm q)
-dual-symm (d-⅋-⊗ p q) = d-⊗-⅋ (dual-symm p) (dual-symm q)
-\end{code}
-
-\begin{code}[hide]
-dual-inv d-𝟘-⊤ d-⊤-𝟘 = refl
-dual-inv d-⊤-𝟘 d-𝟘-⊤ = refl
-dual-inv d-𝟙-⊥ d-⊥-𝟙 = refl
-dual-inv d-⊥-𝟙 d-𝟙-⊥ = refl
-dual-inv (d-!-? p) (d-?-! q) = cong ¡ (dual-inv p q)
-dual-inv (d-?-! p) (d-!-? q) = cong ¿ (dual-inv p q)
-dual-inv (d-&-⊕ p q) (d-⊕-& r s) = cong₂ _&_ (dual-inv p r) (dual-inv q s)
-dual-inv (d-⊕-& p q) (d-&-⊕ r s) = cong₂ _⊕_ (dual-inv p r) (dual-inv q s)
-dual-inv (d-⊗-⅋ p q) (d-⅋-⊗ r s) = cong₂ _⊗_ (dual-inv p r) (dual-inv q s)
-dual-inv (d-⅋-⊗ p q) (d-⊗-⅋ r s) = cong₂ _⅋_ (dual-inv p r) (dual-inv q s)
-
-dual-fun-r d e = dual-inv (dual-symm d) e
-
-dual-fun-l d e = dual-inv d (dual-symm e)
-\end{code}
-
-\subsection{Context Representation}
-\label{sec:context-agda}
-
-We are going to adopt a nameless representation of channels whereby a channel is
-identified by its position in a typing context. This representation is akin to
-using De Bruijn indices, except that the index, instead of being represented
-explicitly by a natural number, is computable from the \emph{proof} that the
-type of the channel belong to the typing context.
-%
-Typing contexts are represented using \emph{lists} of types, where the
-(polymorphic) \AgdaDatatype{List} type is defined in Agda′s standard library.
-
-\begin{code}
-Context : Set
-Context = List Type
-\end{code}
-
-The most important operation concerning typing contexts is \emph{splitting}. The
-splitting of $\ContextC$ into $\ContextD$ and $\ContextE$, which is denoted by
-$\Splitting\ContextC\ContextD\ContextE$, represents the fact that $\ContextC$
-contains all the types contained in $\ContextD$ and $\ContextE$, preserving both
-their overall \emph{multiplicity} and also their relative \emph{order} within
-$\ContextD$ and $\ContextE$. A \emph{proof} of
-$\Splitting\ContextC\ContextD\ContextE$ shows how the types in $\ContextC$ are
-distributed in $\ContextD$ and $\ContextE$ from left to right.
-
-\begin{code}[hide]
-infix 4 _≃_+_ _≃_,_
-\end{code}
-\begin{code}
-data _≃_+_ : Context → Context → Context → Set where
-  split-e  : [] ≃ [] + []
-  split-l  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ A ∷ Δ + Θ
-  split-r  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ Δ + A ∷ Θ
-\end{code}
-
-As an example, below is the splitting of the context $[A, B, C, D]$ into $[A,
-D]$ and $[B, C]$. Note how the sequence of \AgdaInductiveConstructor{split-l}
-and \AgdaInductiveConstructor{split-r} applications determines where each of the
-types in $[A, B, C, D]$ ends up in $[A, D]$ and $[B, C]$.
-
-\begin{code}[hide]
-module _ where
-  postulate A B C D : Type
-\end{code}
-\begin{code}
-  splitting-example : (A ∷ B ∷ C ∷ D ∷ []) ≃ (A ∷ D ∷ []) + (B ∷ C ∷ [])
-  splitting-example = split-l (split-r (split-r (split-l split-e)))
-\end{code}
-
-It is often the case that the context $\ContextD$ in a splitting
-$\Splitting\ContextC\ContextD\ContextE$ is a singleton list $[A]$. For this
-particular case, we introduce a dedicated notation that allows us to represent
-this case in a more compact and readable way, as
-$\SimpleSplitting\ContextC{A}\ContextE$.
-
-\begin{code}
-_≃_,_ : Context → Type → Context → Set
-Γ ≃ A , Δ = Γ ≃ [ A ] + Δ
-\end{code}
-
-Context splitting enjoys a number of expected properties. In particular, it is
-easy to see that splitting is commutative and that the empty context is both a
-left and right unit of splitting.
-
-\begin{code}
-+-comm    : ∀{Γ Δ Θ} → Γ ≃ Δ + Θ → Γ ≃ Θ + Δ
-+-unit-l  : ∀{Γ} → Γ ≃ [] + Γ
-+-unit-r  : ∀{Γ} → Γ ≃ Γ + []
-\end{code}
-\begin{code}[hide]
-+-comm split-e = split-e
-+-comm (split-l p) = split-r (+-comm p)
-+-comm (split-r p) = split-l (+-comm p)
-
-+-unit-l {[]} = split-e
-+-unit-l {_ ∷ _} = split-r +-unit-l
-
-+-unit-r = +-comm +-unit-l
-\end{code}
-
-Context splitting is also associative in a sense that is made precise below. If
-we write $\ContextD + \ContextE$ for some $\ContextC$ such that $Γ ≃ Δ + Θ$, then
-we can prove that $Γ_1 + (Γ_2 + Γ_3) = (Γ_1 + Γ_2) + Γ_3$.
-
-\begin{code}
-+-assoc-r  : ∀{Γ Δ Θ Δ′ Θ′} → Γ ≃ Δ + Θ → Θ ≃ Δ′ + Θ′ →
-             ∃[ Γ′ ] Γ′ ≃ Δ + Δ′ × Γ ≃ Γ′ + Θ′
-+-assoc-l  : ∀{Γ Δ Θ Δ′ Θ′} → Γ ≃ Δ + Θ → Δ ≃ Δ′ + Θ′ →
-             ∃[ Γ′ ] Γ′ ≃ Θ′ + Θ × Γ ≃ Δ′ + Γ′
-\end{code}
-\begin{code}[hide]
-+-assoc-r split-e split-e = [] , split-e , split-e
-+-assoc-r (split-l p) q with +-assoc-r p q
-... | _ , p′ , q′ = _ , split-l p′ , split-l q′
-+-assoc-r (split-r p) (split-l q) with +-assoc-r p q
-... | _ , p′ , q′ = _ , split-r p′ , split-l q′
-+-assoc-r (split-r p) (split-r q) with +-assoc-r p q
-... | _ , p′ , q′ = _ , p′ , split-r q′
-
-+-assoc-l p q with +-assoc-r (+-comm p) (+-comm q)
-... | Δ , r , p′ = Δ , +-comm r , +-comm p′
-
-+-empty-l : ∀{Γ Δ} → Γ ≃ [] + Δ → Γ ≡ Δ
-+-empty-l split-e = refl
-+-empty-l (split-r p) = cong (_ ∷_) (+-empty-l p)
-
-+-sing-l : ∀{A B Γ} → [ A ] ≃ B , Γ → A ≡ B × Γ ≡ []
-+-sing-l (split-l split-e) = refl , refl
+open import Type
+open import Context
 \end{code}
@@
 \label{sec:permutation-agda}
+
+\begin{code}
+data _#_ : Context → Context → Set where
+  #refl  : ∀{Γ} → Γ # Γ
+  #here  : ∀{A B Γ} → (A ∷ B ∷ Γ) # (B ∷ A ∷ Γ)
+  #next  : ∀{A Γ Δ} → Γ # Δ → (A ∷ Γ) # (A ∷ Δ)
+  #tran  : ∀{Γ Δ Θ} → Γ # Δ → Δ # Θ → Γ # Θ
+
+#sym : ∀{Γ Δ} → Γ # Δ → Δ # Γ
+#sym #refl = #refl
+#sym #here = #here
+#sym (#next π) = #next (#sym π)
+#sym (#tran π π′) = #tran (#sym π′) (#sym π)
+
+#empty-inv : ∀{Γ} → [] # Γ → Γ ≡ []
+#empty-inv #refl = refl
+#empty-inv (#tran π π′) rewrite #empty-inv π | #empty-inv π′ = refl
+
+#singleton-inv : ∀{A Γ} → [ A ] # Γ → Γ ≡ [ A ]
+#singleton-inv {Γ = Γ} #refl = refl
+#singleton-inv {Γ = Γ} (#next π) rewrite #empty-inv π = refl
+#singleton-inv {Γ = Γ} (#tran π π′) rewrite #singleton-inv π | #singleton-inv π′ = refl
+
+#rot : ∀{A B C Γ} → (A ∷ B ∷ C ∷ Γ) # (C ∷ A ∷ B ∷ Γ)
+#rot = #tran (#next #here) #here
+
+#cons : ∀{A Γ Δ} → Γ ≃ A , Δ → (A ∷ Δ) # Γ
+#cons (split-l p) with +-empty-l p
+... | refl = #refl
+#cons (split-r p) = #tran #here (#next (#cons p))
+
+#split : ∀{Γ Γ₁ Γ₂ Δ} → Γ # Δ → Γ ≃ Γ₁ + Γ₂ → ∃[ Δ₁ ] ∃[ Δ₂ ] (Δ ≃ Δ₁ + Δ₂ × Γ₁ # Δ₁ × Γ₂ # Δ₂)
+#split #refl p = _ , _ , p , #refl , #refl
+#split (#next π) (split-l p) with #split π p
+... | Δ₁ , Δ₂ , q , π₁ , π₂ = _ ∷ Δ₁ , Δ₂ , split-l q , #next π₁ , π₂
+#split (#next π) (split-r p) with #split π p
+... | Δ₁ , Δ₂ , q , π₁ , π₂ = Δ₁ , _ ∷ Δ₂ , split-r q , π₁ , #next π₂
+#split #here (split-l (split-l p)) = _ , _ , split-l (split-l p) , #here , #refl
+#split #here (split-l (split-r p)) = _ , _ , split-r (split-l p) , #refl , #refl
+#split #here (split-r (split-l p)) = _ , _ , split-l (split-r p) , #refl , #refl
+#split #here (split-r (split-r p)) = _ , _ , split-r (split-r p) , #refl , #here
+#split (#tran π π′) p with #split π p
+... | Θ₁ , Θ₂ , p′ , π₁ , π₂ with #split π′ p′
+... | Δ₁ , Δ₂ , q , π₁′ , π₂′ = Δ₁ , Δ₂ , q , #tran π₁ π₁′ , #tran π₂ π₂′
+
+#one+ : ∀{A Γ Γ′ Δ} → Γ # Δ → Γ ≃ A , Γ′ → ∃[ Δ′ ] (Δ ≃ A , Δ′ × Γ′ # Δ′)
+#one+ π p with #split π p
+... | _ , _ , q , π₁ , π₂ rewrite #singleton-inv π₁ = _ , q , π₂
+
+#shift : ∀{Γ A Δ} → (Γ ++ A ∷ Δ) # (A ∷ Γ ++ Δ)
+#shift {[]} = #refl
+#shift {B ∷ Γ} = #tran (#next #shift) #here
+
++++# : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → (Γ₁ ++ Γ₂) # Γ
++++# split-e = #refl
++++# (split-l p) = #next (+++# p)
++++# (split-r p) = #tran #shift (#next (+++# p))
+
+#left : ∀{Γ Δ Θ} → Γ # Δ → (Θ ++ Γ) # (Θ ++ Δ)
+#left {Θ = []} π = π
+#left {Θ = _ ∷ Θ} π = #next (#left π)
+
+#un : ∀{Γ Δ} → Γ # Δ → Un Γ → Un Δ
+#un #refl un = un
+#un (#next π) (un-∷ un) = un-∷ (#un π un)
+#un #here (un-∷ (un-∷ un)) = un-∷ (un-∷ un)
+#un (#tran π π′) un = #un π′ (#un π un)
+
+#un+ : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → Un Γ₁ → Un Γ₂ → Un Γ
+#un+ split-e un-[] un-[] = un-[]
+#un+ (split-l p) (un-∷ un₁) un₂ = un-∷ (#un+ p un₁ un₂)
+#un+ (split-r p) un₁ (un-∷ un₂) = un-∷ (#un+ p un₁ un₂)
+\end{code}

### 2025-07-26 paper/Process.lagda.tex
split formalization into modules

--- 2025-07-26 paper/formalization.lagda.tex
@@
-\section{Agda Formalization}
-\label{sec:formalization}
-
 \begin{code}[hide]
-open import Data.Bool using (Bool; if_then_else_)
+open import Data.Bool using (Bool; true; false; if_then_else_)
 open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
@@
 open import Data.List.Base using (List; []; _∷_; [_]; _++_)
-\end{code}
 
-\subsection{Type Representation}
-\label{sec:type-agda}
-
-The representation of types is standard, with one constructor for each of the
-forms described in \Cref{sec:calculus}.
-
-\begin{code}
-data Type : Set where
-  𝟘 𝟙 ⊥ ⊤          : Type
-  ¡ ¿              : Type → Type
-  _&_ _⊕_ _⊗_ _⅋_  : Type → Type → Type
-\end{code}
-
-The notion of duality is formalized as a \emph{relation} \AgdaDatatype{Dual}
-such that $\AgdaDatatype{Dual}~A~B$ holds if and only if $A = \dual{B}$.
-
-\begin{code}
-data Dual : Type → Type → Set where
-  d-𝟘-⊤  : Dual 𝟘 ⊤
-  d-⊤-𝟘  : Dual ⊤ 𝟘
-  d-𝟙-⊥  : Dual 𝟙 ⊥
-  d-⊥-𝟙  : Dual ⊥ 𝟙
-  d-!-?  : ∀{A B} → Dual A B → Dual (¡ A) (¿ B)
-  d-?-!  : ∀{A B} → Dual A B → Dual (¿ A) (¡ B)
-  d-&-⊕  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A & B) (A′ ⊕ B′)
-  d-⊕-&  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⊕ B) (A′ & B′)
-  d-⊗-⅋  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⊗ B) (A′ ⅋ B′)
-  d-⅋-⊗  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⅋ B) (A′ ⊗ B′)
-\end{code}
-
-It is straightforward to prove that duality is a symmetric relation and that it
-behaves as an involution. From this we prove that it acts as the function
-$\dual\cdot$ defined in \Cref{sec:calculus}.
-
-\begin{code}
-dual-symm   : ∀{A B} → Dual A B → Dual B A
-dual-inv    : ∀{A B C} → Dual A B → Dual B C → A ≡ C
-dual-fun-r  : ∀{A B C} → Dual A B → Dual A C → B ≡ C
-dual-fun-l  : ∀{A B C} → Dual B A → Dual C A → B ≡ C
-\end{code}
-\begin{code}[hide]
-dual-symm d-𝟘-⊤ = d-⊤-𝟘
-dual-symm d-⊤-𝟘 = d-𝟘-⊤
-dual-symm d-𝟙-⊥ = d-⊥-𝟙
-dual-symm d-⊥-𝟙 = d-𝟙-⊥
-dual-symm (d-!-? p) = d-?-! (dual-symm p)
-dual-symm (d-?-! p) = d-!-? (dual-symm p)
-dual-symm (d-&-⊕ p q) = d-⊕-& (dual-symm p) (dual-symm q)
-dual-symm (d-⊕-& p q) = d-&-⊕ (dual-symm p) (dual-symm q)
-dual-symm (d-⊗-⅋ p q) = d-⅋-⊗ (dual-symm p) (dual-symm q)
-dual-symm (d-⅋-⊗ p q) = d-⊗-⅋ (dual-symm p) (dual-symm q)
-\end{code}
-
-\begin{code}[hide]
-dual-inv d-𝟘-⊤ d-⊤-𝟘 = refl
-dual-inv d-⊤-𝟘 d-𝟘-⊤ = refl
-dual-inv d-𝟙-⊥ d-⊥-𝟙 = refl
-dual-inv d-⊥-𝟙 d-𝟙-⊥ = refl
-dual-inv (d-!-? p) (d-?-! q) = cong ¡ (dual-inv p q)
-dual-inv (d-?-! p) (d-!-? q) = cong ¿ (dual-inv p q)
-dual-inv (d-&-⊕ p q) (d-⊕-& r s) = cong₂ _&_ (dual-inv p r) (dual-inv q s)
-dual-inv (d-⊕-& p q) (d-&-⊕ r s) = cong₂ _⊕_ (dual-inv p r) (dual-inv q s)
-dual-inv (d-⊗-⅋ p q) (d-⅋-⊗ r s) = cong₂ _⊗_ (dual-inv p r) (dual-inv q s)
-dual-inv (d-⅋-⊗ p q) (d-⊗-⅋ r s) = cong₂ _⅋_ (dual-inv p r) (dual-inv q s)
-
-dual-fun-r d e = dual-inv (dual-symm d) e
-
-dual-fun-l d e = dual-inv d (dual-symm e)
-\end{code}
-
-\subsection{Context Representation}
-\label{sec:context-agda}
-
-We are going to adopt a nameless representation of channels whereby a channel is
-identified by its position in a typing context. This representation is akin to
-using De Bruijn indices, except that the index, instead of being represented
-explicitly by a natural number, is computable from the \emph{proof} that the
-type of the channel belong to the typing context.
-%
-Typing contexts are represented using \emph{lists} of types, where the
-(polymorphic) \AgdaDatatype{List} type is defined in Agda′s standard library.
-
-\begin{code}
-Context : Set
-Context = List Type
-\end{code}
-
-The most important operation concerning typing contexts is \emph{splitting}. The
-splitting of $\ContextC$ into $\ContextD$ and $\ContextE$, which is denoted by
-$\Splitting\ContextC\ContextD\ContextE$, represents the fact that $\ContextC$
-contains all the types contained in $\ContextD$ and $\ContextE$, preserving both
-their overall \emph{multiplicity} and also their relative \emph{order} within
-$\ContextD$ and $\ContextE$. A \emph{proof} of
-$\Splitting\ContextC\ContextD\ContextE$ shows how the types in $\ContextC$ are
-distributed in $\ContextD$ and $\ContextE$ from left to right.
-
-\begin{code}[hide]
-infix 4 _≃_+_ _≃_,_
-\end{code}
-\begin{code}
-data _≃_+_ : Context → Context → Context → Set where
-  split-e  : [] ≃ [] + []
-  split-l  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ A ∷ Δ + Θ
-  split-r  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ Δ + A ∷ Θ
-\end{code}
-
-As an example, below is the splitting of the context $[A, B, C, D]$ into $[A,
-D]$ and $[B, C]$. Note how the sequence of \AgdaInductiveConstructor{split-l}
-and \AgdaInductiveConstructor{split-r} applications determines where each of the
-types in $[A, B, C, D]$ ends up in $[A, D]$ and $[B, C]$.
-
-\begin{code}[hide]
-module _ where
-  postulate A B C D : Type
-\end{code}
-\begin{code}
-  splitting-example : (A ∷ B ∷ C ∷ D ∷ []) ≃ (A ∷ D ∷ []) + (B ∷ C ∷ [])
-  splitting-example = split-l (split-r (split-r (split-l split-e)))
-\end{code}
-
-It is often the case that the context $\ContextD$ in a splitting
-$\Splitting\ContextC\ContextD\ContextE$ is a singleton list $[A]$. For this
-particular case, we introduce a dedicated notation that allows us to represent
-this case in a more compact and readable way, as
-$\SimpleSplitting\ContextC{A}\ContextE$.
-
-\begin{code}
-_≃_,_ : Context → Type → Context → Set
-Γ ≃ A , Δ = Γ ≃ [ A ] + Δ
-\end{code}
-
-Context splitting enjoys a number of expected properties. In particular, it is
-easy to see that splitting is commutative and that the empty context is both a
-left and right unit of splitting.
-
-\begin{code}
-+-comm    : ∀{Γ Δ Θ} → Γ ≃ Δ + Θ → Γ ≃ Θ + Δ
-+-unit-l  : ∀{Γ} → Γ ≃ [] + Γ
-+-unit-r  : ∀{Γ} → Γ ≃ Γ + []
-\end{code}
-\begin{code}[hide]
-+-comm split-e = split-e
-+-comm (split-l p) = split-r (+-comm p)
-+-comm (split-r p) = split-l (+-comm p)
-
-+-unit-l {[]} = split-e
-+-unit-l {_ ∷ _} = split-r +-unit-l
-
-+-unit-r = +-comm +-unit-l
-\end{code}
-
-Context splitting is also associative in a sense that is made precise below. If
-we write $\ContextD + \ContextE$ for some $\ContextC$ such that $Γ ≃ Δ + Θ$, then
-we can prove that $Γ_1 + (Γ_2 + Γ_3) = (Γ_1 + Γ_2) + Γ_3$.
-
-\begin{code}
-+-assoc-r  : ∀{Γ Δ Θ Δ′ Θ′} → Γ ≃ Δ + Θ → Θ ≃ Δ′ + Θ′ →
-             ∃[ Γ′ ] Γ′ ≃ Δ + Δ′ × Γ ≃ Γ′ + Θ′
-+-assoc-l  : ∀{Γ Δ Θ Δ′ Θ′} → Γ ≃ Δ + Θ → Δ ≃ Δ′ + Θ′ →
-             ∃[ Γ′ ] Γ′ ≃ Θ′ + Θ × Γ ≃ Δ′ + Γ′
-\end{code}
-\begin{code}[hide]
-+-assoc-r split-e split-e = [] , split-e , split-e
-+-assoc-r (split-l p) q with +-assoc-r p q
-... | _ , p′ , q′ = _ , split-l p′ , split-l q′
-+-assoc-r (split-r p) (split-l q) with +-assoc-r p q
-... | _ , p′ , q′ = _ , split-r p′ , split-l q′
-+-assoc-r (split-r p) (split-r q) with +-assoc-r p q
-... | _ , p′ , q′ = _ , p′ , split-r q′
-
-+-assoc-l p q with +-assoc-r (+-comm p) (+-comm q)
-... | Δ , r , p′ = Δ , +-comm r , +-comm p′
-
-+-empty-l : ∀{Γ Δ} → Γ ≃ [] + Δ → Γ ≡ Δ
-+-empty-l split-e = refl
-+-empty-l (split-r p) = cong (_ ∷_) (+-empty-l p)
-
-+-sing-l : ∀{A B Γ} → [ A ] ≃ B , Γ → A ≡ B × Γ ≡ []
-+-sing-l (split-l split-e) = refl , refl
-\end{code}
-
-\subsection{Context Permutations}
-\label{sec:permutation-agda}
-
-\begin{code}
-data _#_ : Context → Context → Set where
-  #refl  : ∀{Γ} → Γ # Γ
-  #here  : ∀{A B Γ} → (A ∷ B ∷ Γ) # (B ∷ A ∷ Γ)
-  #next  : ∀{A Γ Δ} → Γ # Δ → (A ∷ Γ) # (A ∷ Δ)
-  #tran  : ∀{Γ Δ Θ} → Γ # Δ → Δ # Θ → Γ # Θ
-
-#sym : ∀{Γ Δ} → Γ # Δ → Δ # Γ
-#sym #refl = #refl
-#sym #here = #here
-#sym (#next π) = #next (#sym π)
-#sym (#tran π π′) = #tran (#sym π′) (#sym π)
-
-#empty-inv : ∀{Γ} → [] # Γ → Γ ≡ []
-#empty-inv #refl = refl
-#empty-inv (#tran π π′) rewrite #empty-inv π | #empty-inv π′ = refl
-
-#singleton-inv : ∀{A Γ} → [ A ] # Γ → Γ ≡ [ A ]
-#singleton-inv {Γ = Γ} #refl = refl
-#singleton-inv {Γ = Γ} (#next π) rewrite #empty-inv π = refl
-#singleton-inv {Γ = Γ} (#tran π π′) rewrite #singleton-inv π | #singleton-inv π′ = refl
-
-#rot : ∀{A B C Γ} → (A ∷ B ∷ C ∷ Γ) # (C ∷ A ∷ B ∷ Γ)
-#rot = #tran (#next #here) #here
-
-#cons : ∀{A Γ Δ} → Γ ≃ A , Δ → (A ∷ Δ) # Γ
-#cons (split-l p) with +-empty-l p
-... | refl = #refl
-#cons (split-r p) = #tran #here (#next (#cons p))
-
-#split : ∀{Γ Γ₁ Γ₂ Δ} → Γ # Δ → Γ ≃ Γ₁ + Γ₂ → ∃[ Δ₁ ] ∃[ Δ₂ ] (Δ ≃ Δ₁ + Δ₂ × Γ₁ # Δ₁ × Γ₂ # Δ₂)
-#split #refl p = _ , _ , p , #refl , #refl
-#split (#next π) (split-l p) with #split π p
-... | Δ₁ , Δ₂ , q , π₁ , π₂ = _ ∷ Δ₁ , Δ₂ , split-l q , #next π₁ , π₂
-#split (#next π) (split-r p) with #split π p
-... | Δ₁ , Δ₂ , q , π₁ , π₂ = Δ₁ , _ ∷ Δ₂ , split-r q , π₁ , #next π₂
-#split #here (split-l (split-l p)) = _ , _ , split-l (split-l p) , #here , #refl
-#split #here (split-l (split-r p)) = _ , _ , split-r (split-l p) , #refl , #refl
-#split #here (split-r (split-l p)) = _ , _ , split-l (split-r p) , #refl , #refl
-#split #here (split-r (split-r p)) = _ , _ , split-r (split-r p) , #refl , #here
-#split (#tran π π′) p with #split π p
-... | Θ₁ , Θ₂ , p′ , π₁ , π₂ with #split π′ p′
-... | Δ₁ , Δ₂ , q , π₁′ , π₂′ = Δ₁ , Δ₂ , q , #tran π₁ π₁′ , #tran π₂ π₂′
-
-#one+ : ∀{A Γ Γ′ Δ} → Γ # Δ → Γ ≃ A , Γ′ → ∃[ Δ′ ] (Δ ≃ A , Δ′ × Γ′ # Δ′)
-#one+ π p with #split π p
-... | _ , _ , q , π₁ , π₂ rewrite #singleton-inv π₁ = _ , q , π₂
-
-#shift : ∀{Γ A Δ} → (Γ ++ A ∷ Δ) # (A ∷ Γ ++ Δ)
-#shift {[]} = #refl
-#shift {B ∷ Γ} = #tran (#next #shift) #here
-
-+++# : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → (Γ₁ ++ Γ₂) # Γ
-+++# split-e = #refl
-+++# (split-l p) = #next (+++# p)
-+++# (split-r p) = #tran #shift (#next (+++# p))
-
-#left : ∀{Γ Δ Θ} → Γ # Δ → (Θ ++ Γ) # (Θ ++ Δ)
-#left {Θ = []} π = π
-#left {Θ = _ ∷ Θ} π = #next (#left π)
-
-data Un : Context → Set where
-  un-[]  : Un []
-  un-∷   : ∀{A Γ} → Un Γ → Un (¿ A ∷ Γ)
-
-#un : ∀{Γ Δ} → Γ # Δ → Un Γ → Un Δ
-#un #refl un = un
-#un (#next π) (un-∷ un) = un-∷ (#un π un)
-#un #here (un-∷ (un-∷ un)) = un-∷ (un-∷ un)
-#un (#tran π π′) un = #un π′ (#un π un)
-
-#un+ : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → Un Γ₁ → Un Γ₂ → Un Γ
-#un+ split-e un-[] un-[] = un-[]
-#un+ (split-l p) (un-∷ un₁) un₂ = un-∷ (#un+ p un₁ un₂)
-#un+ (split-r p) un₁ (un-∷ un₂) = un-∷ (#un+ p un₁ un₂)
+open import Type
+open import Context
+open import Permutations
 \end{code}
@@
 \end{code}
-
-\subsection{Operational Semantics}
-\label{sec:semantics-agda}

### 2025-07-26 paper/Semantics.lagda.tex
split formalization into modules

--- 2025-07-26 paper/formalization.lagda.tex
@@
-\section{Agda Formalization}
-\label{sec:formalization}
-
 \begin{code}[hide]
@@
 open import Data.List.Base using (List; []; _∷_; [_]; _++_)
-\end{code}
 
-\subsection{Type Representation}
-\label{sec:type-agda}
-
-The representation of types is standard, with one constructor for each of the
-forms described in \Cref{sec:calculus}.
-
-\begin{code}
-data Type : Set where
-  𝟘 𝟙 ⊥ ⊤          : Type
-  ¡ ¿              : Type → Type
-  _&_ _⊕_ _⊗_ _⅋_  : Type → Type → Type
-\end{code}
-
-The notion of duality is formalized as a \emph{relation} \AgdaDatatype{Dual}
-such that $\AgdaDatatype{Dual}~A~B$ holds if and only if $A = \dual{B}$.
-
-\begin{code}
-data Dual : Type → Type → Set where
-  d-𝟘-⊤  : Dual 𝟘 ⊤
-  d-⊤-𝟘  : Dual ⊤ 𝟘
-  d-𝟙-⊥  : Dual 𝟙 ⊥
-  d-⊥-𝟙  : Dual ⊥ 𝟙
-  d-!-?  : ∀{A B} → Dual A B → Dual (¡ A) (¿ B)
-  d-?-!  : ∀{A B} → Dual A B → Dual (¿ A) (¡ B)
-  d-&-⊕  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A & B) (A′ ⊕ B′)
-  d-⊕-&  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⊕ B) (A′ & B′)
-  d-⊗-⅋  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⊗ B) (A′ ⅋ B′)
-  d-⅋-⊗  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⅋ B) (A′ ⊗ B′)
-\end{code}
-
-It is straightforward to prove that duality is a symmetric relation and that it
-behaves as an involution. From this we prove that it acts as the function
-$\dual\cdot$ defined in \Cref{sec:calculus}.
-
-\begin{code}
-dual-symm   : ∀{A B} → Dual A B → Dual B A
-dual-inv    : ∀{A B C} → Dual A B → Dual B C → A ≡ C
-dual-fun-r  : ∀{A B C} → Dual A B → Dual A C → B ≡ C
-dual-fun-l  : ∀{A B C} → Dual B A → Dual C A → B ≡ C
-\end{code}
-\begin{code}[hide]
-dual-symm d-𝟘-⊤ = d-⊤-𝟘
-dual-symm d-⊤-𝟘 = d-𝟘-⊤
-dual-symm d-𝟙-⊥ = d-⊥-𝟙
-dual-symm d-⊥-𝟙 = d-𝟙-⊥
-dual-symm (d-!-? p) = d-?-! (dual-symm p)
-dual-symm (d-?-! p) = d-!-? (dual-symm p)
-dual-symm (d-&-⊕ p q) = d-⊕-& (dual-symm p) (dual-symm q)
-dual-symm (d-⊕-& p q) = d-&-⊕ (dual-symm p) (dual-symm q)
-dual-symm (d-⊗-⅋ p q) = d-⅋-⊗ (dual-symm p) (dual-symm q)
-dual-symm (d-⅋-⊗ p q) = d-⊗-⅋ (dual-symm p) (dual-symm q)
-\end{code}
-
-\begin{code}[hide]
-dual-inv d-𝟘-⊤ d-⊤-𝟘 = refl
-dual-inv d-⊤-𝟘 d-𝟘-⊤ = refl
-dual-inv d-𝟙-⊥ d-⊥-𝟙 = refl
-dual-inv d-⊥-𝟙 d-𝟙-⊥ = refl
-dual-inv (d-!-? p) (d-?-! q) = cong ¡ (dual-inv p q)
-dual-inv (d-?-! p) (d-!-? q) = cong ¿ (dual-inv p q)
-dual-inv (d-&-⊕ p q) (d-⊕-& r s) = cong₂ _&_ (dual-inv p r) (dual-inv q s)
-dual-inv (d-⊕-& p q) (d-&-⊕ r s) = cong₂ _⊕_ (dual-inv p r) (dual-inv q s)
-dual-inv (d-⊗-⅋ p q) (d-⅋-⊗ r s) = cong₂ _⊗_ (dual-inv p r) (dual-inv q s)
-dual-inv (d-⅋-⊗ p q) (d-⊗-⅋ r s) = cong₂ _⅋_ (dual-inv p r) (dual-inv q s)
-
-dual-fun-r d e = dual-inv (dual-symm d) e
-
-dual-fun-l d e = dual-inv d (dual-symm e)
-\end{code}
-
-\subsection{Context Representation}
-\label{sec:context-agda}
-
-We are going to adopt a nameless representation of channels whereby a channel is
-identified by its position in a typing context. This representation is akin to
-using De Bruijn indices, except that the index, instead of being represented
-explicitly by a natural number, is computable from the \emph{proof} that the
-type of the channel belong to the typing context.
-%
-Typing contexts are represented using \emph{lists} of types, where the
-(polymorphic) \AgdaDatatype{List} type is defined in Agda′s standard library.
-
-\begin{code}
-Context : Set
-Context = List Type
-\end{code}
-
-The most important operation concerning typing contexts is \emph{splitting}. The
-splitting of $\ContextC$ into $\ContextD$ and $\ContextE$, which is denoted by
-$\Splitting\ContextC\ContextD\ContextE$, represents the fact that $\ContextC$
-contains all the types contained in $\ContextD$ and $\ContextE$, preserving both
-their overall \emph{multiplicity} and also their relative \emph{order} within
-$\ContextD$ and $\ContextE$. A \emph{proof} of
-$\Splitting\ContextC\ContextD\ContextE$ shows how the types in $\ContextC$ are
-distributed in $\ContextD$ and $\ContextE$ from left to right.
-
-\begin{code}[hide]
-infix 4 _≃_+_ _≃_,_
-\end{code}
-\begin{code}
-data _≃_+_ : Context → Context → Context → Set where
-  split-e  : [] ≃ [] + []
-  split-l  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ A ∷ Δ + Θ
-  split-r  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ Δ + A ∷ Θ
-\end{code}
-
-As an example, below is the splitting of the context $[A, B, C, D]$ into $[A,
-D]$ and $[B, C]$. Note how the sequence of \AgdaInductiveConstructor{split-l}
-and \AgdaInductiveConstructor{split-r} applications determines where each of the
-types in $[A, B, C, D]$ ends up in $[A, D]$ and $[B, C]$.
-
-\begin{code}[hide]
-module _ where
-  postulate A B C D : Type
-\end{code}
-\begin{code}
-  splitting-example : (A ∷ B ∷ C ∷ D ∷ []) ≃ (A ∷ D ∷ []) + (B ∷ C ∷ [])
-  splitting-example = split-l (split-r (split-r (split-l split-e)))
-\end{code}
-
-It is often the case that the context $\ContextD$ in a splitting
-$\Splitting\ContextC\ContextD\ContextE$ is a singleton list $[A]$. For this
-particular case, we introduce a dedicated notation that allows us to represent
-this case in a more compact and readable way, as
-$\SimpleSplitting\ContextC{A}\ContextE$.
-
-\begin{code}
-_≃_,_ : Context → Type → Context → Set
-Γ ≃ A , Δ = Γ ≃ [ A ] + Δ
-\end{code}
-
-Context splitting enjoys a number of expected properties. In particular, it is
-easy to see that splitting is commutative and that the empty context is both a
-left and right unit of splitting.
-
-\begin{code}
-+-comm    : ∀{Γ Δ Θ} → Γ ≃ Δ + Θ → Γ ≃ Θ + Δ
-+-unit-l  : ∀{Γ} → Γ ≃ [] + Γ
-+-unit-r  : ∀{Γ} → Γ ≃ Γ + []
-\end{code}
-\begin{code}[hide]
-+-comm split-e = split-e
-+-comm (split-l p) = split-r (+-comm p)
-+-comm (split-r p) = split-l (+-comm p)
-
-+-unit-l {[]} = split-e
-+-unit-l {_ ∷ _} = split-r +-unit-l
-
-+-unit-r = +-comm +-unit-l
-\end{code}
-
-Context splitting is also associative in a sense that is made precise below. If
-we write $\ContextD + \ContextE$ for some $\ContextC$ such that $Γ ≃ Δ + Θ$, then
-we can prove that $Γ_1 + (Γ_2 + Γ_3) = (Γ_1 + Γ_2) + Γ_3$.
-
-\begin{code}
-+-assoc-r  : ∀{Γ Δ Θ Δ′ Θ′} → Γ ≃ Δ + Θ → Θ ≃ Δ′ + Θ′ →
-             ∃[ Γ′ ] Γ′ ≃ Δ + Δ′ × Γ ≃ Γ′ + Θ′
-+-assoc-l  : ∀{Γ Δ Θ Δ′ Θ′} → Γ ≃ Δ + Θ → Δ ≃ Δ′ + Θ′ →
-             ∃[ Γ′ ] Γ′ ≃ Θ′ + Θ × Γ ≃ Δ′ + Γ′
-\end{code}
-\begin{code}[hide]
-+-assoc-r split-e split-e = [] , split-e , split-e
-+-assoc-r (split-l p) q with +-assoc-r p q
-... | _ , p′ , q′ = _ , split-l p′ , split-l q′
-+-assoc-r (split-r p) (split-l q) with +-assoc-r p q
-... | _ , p′ , q′ = _ , split-r p′ , split-l q′
-+-assoc-r (split-r p) (split-r q) with +-assoc-r p q
-... | _ , p′ , q′ = _ , p′ , split-r q′
-
-+-assoc-l p q with +-assoc-r (+-comm p) (+-comm q)
-... | Δ , r , p′ = Δ , +-comm r , +-comm p′
-
-+-empty-l : ∀{Γ Δ} → Γ ≃ [] + Δ → Γ ≡ Δ
-+-empty-l split-e = refl
-+-empty-l (split-r p) = cong (_ ∷_) (+-empty-l p)
-
-+-sing-l : ∀{A B Γ} → [ A ] ≃ B , Γ → A ≡ B × Γ ≡ []
-+-sing-l (split-l split-e) = refl , refl
-\end{code}
-
-\subsection{Context Permutations}
-\label{sec:permutation-agda}
-
-\begin{code}
-data _#_ : Context → Context → Set where
-  #refl  : ∀{Γ} → Γ # Γ
-  #here  : ∀{A B Γ} → (A ∷ B ∷ Γ) # (B ∷ A ∷ Γ)
-  #next  : ∀{A Γ Δ} → Γ # Δ → (A ∷ Γ) # (A ∷ Δ)
-  #tran  : ∀{Γ Δ Θ} → Γ # Δ → Δ # Θ → Γ # Θ
-
-#sym : ∀{Γ Δ} → Γ # Δ → Δ # Γ
-#sym #refl = #refl
-#sym #here = #here
-#sym (#next π) = #next (#sym π)
-#sym (#tran π π′) = #tran (#sym π′) (#sym π)
-
-#empty-inv : ∀{Γ} → [] # Γ → Γ ≡ []
-#empty-inv #refl = refl
-#empty-inv (#tran π π′) rewrite #empty-inv π | #empty-inv π′ = refl
-
-#singleton-inv : ∀{A Γ} → [ A ] # Γ → Γ ≡ [ A ]
-#singleton-inv {Γ = Γ} #refl = refl
-#singleton-inv {Γ = Γ} (#next π) rewrite #empty-inv π = refl
-#singleton-inv {Γ = Γ} (#tran π π′) rewrite #singleton-inv π | #singleton-inv π′ = refl
-
-#rot : ∀{A B C Γ} → (A ∷ B ∷ C ∷ Γ) # (C ∷ A ∷ B ∷ Γ)
-#rot = #tran (#next #here) #here
-
-#cons : ∀{A Γ Δ} → Γ ≃ A , Δ → (A ∷ Δ) # Γ
-#cons (split-l p) with +-empty-l p
-... | refl = #refl
-#cons (split-r p) = #tran #here (#next (#cons p))
-
-#split : ∀{Γ Γ₁ Γ₂ Δ} → Γ # Δ → Γ ≃ Γ₁ + Γ₂ → ∃[ Δ₁ ] ∃[ Δ₂ ] (Δ ≃ Δ₁ + Δ₂ × Γ₁ # Δ₁ × Γ₂ # Δ₂)
-#split #refl p = _ , _ , p , #refl , #refl
-#split (#next π) (split-l p) with #split π p
-... | Δ₁ , Δ₂ , q , π₁ , π₂ = _ ∷ Δ₁ , Δ₂ , split-l q , #next π₁ , π₂
-#split (#next π) (split-r p) with #split π p
-... | Δ₁ , Δ₂ , q , π₁ , π₂ = Δ₁ , _ ∷ Δ₂ , split-r q , π₁ , #next π₂
-#split #here (split-l (split-l p)) = _ , _ , split-l (split-l p) , #here , #refl
-#split #here (split-l (split-r p)) = _ , _ , split-r (split-l p) , #refl , #refl
-#split #here (split-r (split-l p)) = _ , _ , split-l (split-r p) , #refl , #refl
-#split #here (split-r (split-r p)) = _ , _ , split-r (split-r p) , #refl , #here
-#split (#tran π π′) p with #split π p
-... | Θ₁ , Θ₂ , p′ , π₁ , π₂ with #split π′ p′
-... | Δ₁ , Δ₂ , q , π₁′ , π₂′ = Δ₁ , Δ₂ , q , #tran π₁ π₁′ , #tran π₂ π₂′
-
-#one+ : ∀{A Γ Γ′ Δ} → Γ # Δ → Γ ≃ A , Γ′ → ∃[ Δ′ ] (Δ ≃ A , Δ′ × Γ′ # Δ′)
-#one+ π p with #split π p
-... | _ , _ , q , π₁ , π₂ rewrite #singleton-inv π₁ = _ , q , π₂
-
-#shift : ∀{Γ A Δ} → (Γ ++ A ∷ Δ) # (A ∷ Γ ++ Δ)
-#shift {[]} = #refl
-#shift {B ∷ Γ} = #tran (#next #shift) #here
-
-+++# : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → (Γ₁ ++ Γ₂) # Γ
-+++# split-e = #refl
-+++# (split-l p) = #next (+++# p)
-+++# (split-r p) = #tran #shift (#next (+++# p))
-
-#left : ∀{Γ Δ Θ} → Γ # Δ → (Θ ++ Γ) # (Θ ++ Δ)
-#left {Θ = []} π = π
-#left {Θ = _ ∷ Θ} π = #next (#left π)
-
-data Un : Context → Set where
-  un-[]  : Un []
-  un-∷   : ∀{A Γ} → Un Γ → Un (¿ A ∷ Γ)
-
-#un : ∀{Γ Δ} → Γ # Δ → Un Γ → Un Δ
-#un #refl un = un
-#un (#next π) (un-∷ un) = un-∷ (#un π un)
-#un #here (un-∷ (un-∷ un)) = un-∷ (un-∷ un)
-#un (#tran π π′) un = #un π′ (#un π un)
-
-#un+ : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → Un Γ₁ → Un Γ₂ → Un Γ
-#un+ split-e un-[] un-[] = un-[]
-#un+ (split-l p) (un-∷ un₁) un₂ = un-∷ (#un+ p un₁ un₂)
-#un+ (split-r p) un₁ (un-∷ un₂) = un-∷ (#un+ p un₁ un₂)
-\end{code}
-
-\subsection{Process Representation}
-\label{sec:process-agda}
-
-CO-DE BRUIJN INDICES
-
-We adopt an \emph{intrinsically typed} representation of processes, so that a
-process is also a proof derivation showing that the process is well typed. This
-choice increases the effort in the definition of the datatypes for processes and
-their operational semantics, but it pays off in the rest of the formalization
-for a number of reasons:
-
-\begin{itemize}
-\item channel names are encoded in the context splitting proofs occurring within
-  processes, hence they do not need to be represented explicitly.
-\item processes and typing rules are conflated in the same datatype, thus
-  reducing the overall number of datatypes in the formalization;
-\item the typing preservation results are embedded in the definition of the
-  operational semantics of processes and do not need to be proved separately.
-\end{itemize}
-
-We now illustrate the definition of the \AgdaDatatype{Process} datatype and
-describe each constructor in detail. The datatype is indexed by a typing
-context:
-
-\begin{AgdaAlign}
-\begin{code}
-data Process : Context → Set where
-   link      : ∀{Γ A B} (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) → Process Γ
-\end{code}
-
-A \AgdaInductiveConstructor{link} process is well typed in a typing context with
-exactly two types $A$ and $B$ which must be related by duality. We use a proof
-$p$ that $\Context \simeq [A] + [B]$ instead of requiring $\Context$ to be
-simply $[A,B]$ so that it becomes straightforward to define \SLink using
-\AgdaFunction{dual-symm} and \AgdaFunction{+-comm}.
-
-\begin{code}
-   fail      : ∀{Γ Δ} (p : Γ ≃ ⊤ , Δ) → Process Γ
-\end{code}
-
-A \AgdaInductiveConstructor{fail} process simply requires the presence of $\Top$
-in the typing context.
-
-\begin{code}
-   close     : Process [ 𝟙 ]
-\end{code}
-
-A \AgdaInductiveConstructor{close} process is well typed in the singleton
-context $[\One]$.
-
-\begin{code}
-   wait      : ∀{Γ Δ} (p : Γ ≃ ⊥ , Δ) → Process Δ → Process Γ
-\end{code}
-
-A \AgdaInductiveConstructor{wait} process is well typed in any context
-$\ContextC$ that contains $\Bot$ and whose residual $\ContextD$ allows the
-continuation to be well typed.
-
-\begin{code}
-   select    : ∀{Γ Δ A B} (x : Bool) (p : Γ ≃ A ⊕ B , Δ) →
-               Process ((if x then A else B) ∷ Δ) → Process Γ
-\end{code}
-
-We represent the tags $\InTag_1$ and $\InTag_2$ with Agda′s boolean values
-\AgdaInductiveConstructor{true} and \AgdaInductiveConstructor{false}. In this
-way we can use a single constructor \AgdaInductiveConstructor{select} for the
-selections $\Select\x{\InTag_i}.P$. According to the semantics of \Calculus, the
-channel on which the selection is performed is consumed and the continuation
-process uses a fresh continuation channel. As a consequence, the type of the
-continuation channel, which is either $A$ or $B$ depending on the value of the
-tag $x$, is \emph{prepended} in front of the typing context used to type the
-continuation process.
-
-\begin{code}
-   case      : ∀{Γ Δ A B} (p : Γ ≃ A & B , Δ) →
-               Process (A ∷ Δ) → Process (B ∷ Δ) → Process Γ
-\end{code}
-
-A \AgdaInductiveConstructor{case} process has two possible continuations. Note
-again that the continuation channel, which is received along with the tag, has
-either type $A$ or type $B$ depending on the tag and its type is prepended to
-the typing context of the continuation process since it is the newest channel.
-
-\begin{code}
-   fork      : ∀{Γ Δ Γ₁ Γ₂ A B} (p : Γ ≃ A ⊗ B , Δ) (q : Δ ≃ Γ₁ + Γ₂) →
-               Process (A ∷ Γ₁) → Process (B ∷ Γ₂) → Process Γ
-\end{code}
-
-A \AgdaInductiveConstructor{fork} process creates two fresh continuation
-channels and spaws two continuation processes. Since context splitting is a
-ternary relation, we need two splitting proofs $p$ and $q$ to isolate the type
-$A \Ten B$ of channel on which the pair of continuation channels is sent and to
-distribute the remaining channel across the two continuation processes.
-
-\begin{code}
-   join      : ∀{Γ Δ A B} (p : Γ ≃ A ⅋ B , Δ) →
-               Process (B ∷ A ∷ Δ) → Process Γ
-\end{code}
-
-The \AgdaInductiveConstructor{join} process has a single continuation which uses
-the pair of continuation channels received from a channel of type $A \Par B$.
-Note that $B$ is prepended in front of $A$, somewhat implying that the second
-continuation is bound \emph{after} the first even though, from a technical
-standpoint, both continuations are bound simultaneously. We could have prepended
-the types $A$ and $B$ also in the opposite order, provided that the reduction
-rule \RFork (formalized in \Cref{sec:semantics-agda}) is suitably adjusted.
-
-\begin{code}
-   server    : ∀{Γ Δ A} (p : Γ ≃ ¡ A , Δ) (un : Un Δ) →
-               Process (A ∷ Δ) → Process Γ
-\end{code}
-
-A \AgdaInductiveConstructor{server} constructor introduces a channel of type
-$\OfCourse A$ given a continuation process that consumes a channel of type $A$
-and a proof $\mathit{un}$ that every other channel in $\ContextD$ is
-unrestricted.\Luca{Introdurre terminologia e dire che unrestricted vuol dire che
-il canale può essere ``duplicato'' e scartato.}
-
-\begin{code}
-   client    : ∀{Γ Δ A} (p : Γ ≃ ¿ A , Δ) → Process (A ∷ Δ) → Process Γ
-   weaken    : ∀{Γ Δ A} (p : Γ ≃ ¿ A , Δ) → Process Δ → Process Γ
-   contract  : ∀{Γ Δ A} (p : Γ ≃ ¿ A , Δ) → Process (¿ A ∷ ¿ A ∷ Δ) → Process Γ
-\end{code}
-
-The constructors \AgdaInductiveConstructor{client},
-\AgdaInductiveConstructor{weaken} and \AgdaInductiveConstructor{contract}
-introduce a channel of type $\WhyNot A$ and follow the same pattern of the
-previous process forms. Note that, in the case of
-\AgdaInductiveConstructor{contract}, the contraction of $\WhyNot A$ applies to
-the first two channels in the typing context of the continuation. This is
-consistent with the fact that \Calculus uses an explicit prefix
-$\Contract\x\y\z$ that models contraction and binds $y$ and $z$ in the
-continuation process. In a pen-and-paper presentation, like the one by
-Wadler~\citep{Wadler14}, it is arguably more convenient to have weaking and
-contraction typing rules without the corresponding process forms.
-
-\begin{code}
-   cut       : ∀{Γ Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) →
-               Process (A ∷ Γ₁) → Process (B ∷ Γ₂) → Process Γ
-\end{code}
-
-A \AgdaInductiveConstructor{cut} process incorporates a duality proof for the
-peer endpoints of the channel being restricted and a splitting proof that
-distributes the free channels among the two parallel processes.
-
-\end{AgdaAlign}
-
-\begin{code}
-#process : ∀{Γ Δ} → Γ # Δ → Process Γ → Process Δ
-\end{code}
-\begin{code}[hide]
-#process π (link d p) with #one+ π p
-... | Δ′ , q , π′ with #singleton-inv π′
-... | refl = link d q
-#process π close with #singleton-inv π
-... | refl = close
-#process π (fail p) with #one+ π p
-... | Δ′ , q , π′ = fail q
-#process π (wait p P) with #one+ π p
-... | Δ′ , q , π′ = wait q (#process π′ P)
-#process π (select x p P) with #one+ π p
-... | Δ′ , q , π′ = select x q (#process (#next π′) P)
-#process π (case p P Q) with #one+ π p
-... | Δ′ , q , π′ = case q (#process (#next π′) P) (#process (#next π′) Q)
-#process π (fork p q P Q) with #one+ π p
-... | Δ′ , p′ , π′ with #split π′ q
-... | Δ₁ , Δ₂ , q′ , π₁ , π₂ = fork p′ q′ (#process (#next π₁) P) (#process (#next π₂) Q)
-#process π (join p P) with #one+ π p
-... | Δ′ , q , π′ = join q (#process (#next (#next π′)) P)
-#process π (cut d p P Q) with #split π p
-... | Δ₁ , Δ₂ , q , π₁ , π₂ = cut d q (#process (#next π₁) P) (#process (#next π₂) Q)
-#process π (server p un P) with #one+ π p
-... | Δ′ , q , π′ = server q (#un π′ un) (#process (#next π′) P)
-#process π (client p P) with #one+ π p
-... | Δ′ , q , π′ = client q (#process (#next π′) P)
-#process π (weaken p P) with #one+ π p
-... | Δ′ , q , π′ = weaken q (#process π′ P)
-#process π (contract p P) with #one+ π p
-... | Δ′ , q , π′ = contract q (#process (#next (#next π′)) P)
+open import Type
+open import Context
+open import Permutations
+open import Process
 \end{code}

### 2025-07-26 paper/Type.lagda.tex
split formalization into modules

--- 2025-07-25 paper/formalization.lagda.tex
@@
-\section{Agda Formalization}
-\label{sec:formalization}
-
 \begin{code}[hide]
-open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
 open import Relation.Binary.PropositionalEquality using (_≡_; refl; cong; cong₂)
-open import Data.List.Base using (List; []; _∷_; [_]; _++_)
 \end{code}
@@
 \end{code}
-
-\subsection{Context Representation}
-\label{sec:context-agda}
-
-We are going to adopt a nameless representation of channels whereby a channel is
-identified by its position in a typing context. This representation is akin to
-using De Bruijn indices, except that the index, instead of being represented
-explicitly by a natural number, is computable from the \emph{proof} that the
-type of the channel belong to the typing context.
-%
-Typing contexts are represented using \emph{lists} of types, where the
-(polymorphic) \AgdaDatatype{List} type is defined in Agda′s standard library.
-
-\begin{code}
-Context : Set
-Context = List Type
-\end{code}
-
-The most important operation concerning typing contexts is \emph{splitting}. The
-splitting of $\ContextC$ into $\ContextD$ and $\ContextE$, which is denoted by
-$\Splitting\ContextC\ContextD\ContextE$, represents the fact that $\ContextC$
-contains all the types contained in $\ContextD$ and $\ContextE$, preserving both
-their overall \emph{multiplicity} and also their relative \emph{order} within
-$\ContextD$ and $\ContextE$. A \emph{proof} of
-$\Splitting\ContextC\ContextD\ContextE$ shows how the types in $\ContextC$ are
-distributed in $\ContextD$ and $\ContextE$ from left to right.
-
-\begin{code}[hide]
-infix 4 _≃_+_ _≃_,_
-\end{code}
-\begin{code}
-data _≃_+_ : Context → Context → Context → Set where
-  split-e  : [] ≃ [] + []
-  split-l  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ A ∷ Δ + Θ
-  split-r  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ Δ + A ∷ Θ
-\end{code}
-
-As an example, below is the splitting of the context $[A, B, C, D]$ into $[A,
-D]$ and $[B, C]$. Note how the sequence of \AgdaInductiveConstructor{split-l}
-and \AgdaInductiveConstructor{split-r} applications determines where each of the
-types in $[A, B, C, D]$ ends up in $[A, D]$ and $[B, C]$.
-
-\begin{code}[hide]
-module _ where
-  postulate A B C D : Type
-\end{code}
-\begin{code}
-  splitting-example : (A ∷ B ∷ C ∷ D ∷ []) ≃ (A ∷ D ∷ []) + (B ∷ C ∷ [])
-  splitting-example = split-l (split-r (split-r (split-l split-e)))
-\end{code}
-
-It is often the case that the context $\ContextD$ in a splitting
-$\Splitting\ContextC\ContextD\ContextE$ is a singleton list $[A]$. For this
-particular case, we introduce a dedicated notation that allows us to represent
-this case in a more compact and readable way, as
-$\SimpleSplitting\ContextC{A}\ContextE$.
-
-\begin{code}
-_≃_,_ : Context → Type → Context → Set
-Γ ≃ A , Δ = Γ ≃ [ A ] + Δ
-\end{code}
-
-Context splitting enjoys a number of expected properties. In particular, it is
-easy to see that splitting is commutative and that the empty context is both a
-left and right unit of splitting.
-
-\begin{code}
-+-comm    : ∀{Γ Δ Θ} → Γ ≃ Δ + Θ → Γ ≃ Θ + Δ
-+-unit-l  : ∀{Γ} → Γ ≃ [] + Γ
-+-unit-r  : ∀{Γ} → Γ ≃ Γ + []
-\end{code}
-\begin{code}[hide]
-+-comm split-e = split-e
-+-comm (split-l p) = split-r (+-comm p)
-+-comm (split-r p) = split-l (+-comm p)
-
-+-unit-l {[]} = split-e
-+-unit-l {_ ∷ _} = split-r +-unit-l
-
-+-unit-r = +-comm +-unit-l
-\end{code}
-
-Context splitting is also associative in a sense that is made precise below. If
-we write $\ContextD + \ContextE$ for some $\ContextC$ such that $Γ ≃ Δ + Θ$, then
-we can prove that $Γ_1 + (Γ_2 + Γ_3) = (Γ_1 + Γ_2) + Γ_3$.
-
-\begin{code}
-+-assoc-r  : ∀{Γ Δ Θ Δ′ Θ′} → Γ ≃ Δ + Θ → Θ ≃ Δ′ + Θ′ →
-             ∃[ Γ′ ] Γ′ ≃ Δ + Δ′ × Γ ≃ Γ′ + Θ′
-+-assoc-l  : ∀{Γ Δ Θ Δ′ Θ′} → Γ ≃ Δ + Θ → Δ ≃ Δ′ + Θ′ →
-             ∃[ Γ′ ] Γ′ ≃ Θ′ + Θ × Γ ≃ Δ′ + Γ′
-\end{code}
-\begin{code}[hide]
-+-assoc-r split-e split-e = [] , split-e , split-e
-+-assoc-r (split-l p) q with +-assoc-r p q
-... | _ , p′ , q′ = _ , split-l p′ , split-l q′
-+-assoc-r (split-r p) (split-l q) with +-assoc-r p q
-... | _ , p′ , q′ = _ , split-r p′ , split-l q′
-+-assoc-r (split-r p) (split-r q) with +-assoc-r p q
-... | _ , p′ , q′ = _ , p′ , split-r q′
-
-+-assoc-l p q with +-assoc-r (+-comm p) (+-comm q)
-... | Δ , r , p′ = Δ , +-comm r , +-comm p′
-
-+-empty-l : ∀{Γ Δ} → Γ ≃ [] + Δ → Γ ≡ Δ
-+-empty-l split-e = refl
-+-empty-l (split-r p) = cong (_ ∷_) (+-empty-l p)
-
-+-sing-l : ∀{A B Γ} → [ A ] ≃ B , Γ → A ≡ B × Γ ≡ []
-+-sing-l (split-l split-e) = refl , refl
-\end{code}
-
-\subsection{Context Permutations}
-\label{sec:permutation-agda}

### 2025-07-26 paper/agda.tex
split formalization into modules

\section{Agda Formalization}
\label{sec:agda}

\input{Type}
\input{Context}
\input{Permutations}
\input{Process}
\input{Semantics}

### 2025-07-26 paper/watch.sh

@@ -1,2 +1,2 @@
 #!/bin/sh
-exec watchexec -d 500ms -e .lagda.tex agda --latex-dir=. --latex formalization.lagda.tex
+exec watchexec -d 500ms -f "*.lagda.tex" make

### 2025-07-26 paper/Reduction.lagda.tex
included reduction

\begin{code}[hide]
open import Data.Bool using (Bool; true; false; if_then_else_)
open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
open import Relation.Binary.PropositionalEquality using (_≡_; refl; sym; cong; cong₂)
open import Data.List.Base using (List; []; _∷_; [_]; _++_)
open import Data.List.Properties using (++-assoc)

open import Type
open import Context
open import Permutations
open import Process
open import Congruence
\end{code}

\subsection{Reduction}
\label{sec:reduction-agda}

\begin{code}
weakening : ∀{Γ Γ₁ Γ₂} (un : Un Γ₁) → Γ ≃ Γ₁ + Γ₂ → Process Γ₂ → Process Γ
weakening un p P = #process (+++# p) (aux un P)
  where
    aux : ∀{Γ₁ Γ₂} (un : Un Γ₁) → Process Γ₂ → Process (Γ₁ ++ Γ₂)
    aux un-[] P = P
    aux (un-∷ un) P = weaken (split-l +-unit-l) (aux un P)

contraction : ∀{Γ Γ₁ Γ₂} (un : Un Γ₁) → Γ ≃ Γ₁ + Γ₂ → Process (Γ₁ ++ Γ) → Process Γ
contraction un p P = #process (+++# p) (aux un (#process (#left (#sym (+++# p))) P))
  where
    aux : ∀{Γ₁ Γ₂} → Un Γ₁ → Process (Γ₁ ++ Γ₁ ++ Γ₂) → Process (Γ₁ ++ Γ₂)
    aux un-[] P = P
    aux {¿ A ∷ Γ₁} {Γ₂} (un-∷ un) P with contract (split-l +-unit-l) (#process (#shift {¿ A ∷ Γ₁} {¿ A} {Γ₁ ++ Γ₂}) P)
    ... | P₁ rewrite sym (++-assoc (¿ A ∷ Γ₁) Γ₁ Γ₂) with #process (#sym (#shift {Γ₁ ++ Γ₁} {¿ A})) P₁
    ... | P₂ rewrite ++-assoc Γ₁ Γ₁ (¿ A ∷ Γ₂) with aux un P₂
    ... | P₃ = #process #shift P₃

data _↝_ {Γ} : Process Γ → Process Γ → Set where
  r-link :
    ∀{Δ A B} {P : Process (B ∷ Δ)} (d : Dual A B) (e : Dual A B) (p : Γ ≃ B , Δ) →
    cut d p (link e (split-l (split-r split-e))) P ↝ #process (#cons p) P

  r-close :
    ∀{P : Process Γ} (p₀ : Γ ≃ [] + Γ) (q₀ : Γ ≃ [] + Γ) →
    cut d-𝟙-⊥ p₀ close (wait (split-l q₀) P) ↝ P

  r-select-l :
    ∀{Γ₁ Γ₂ A A′ B B′}
    {P : Process (A ∷ Γ₁)} {Q : Process (A′ ∷ Γ₂)} {R : Process (B′ ∷ Γ₂)}
    (d : Dual A A′) (e : Dual B B′) (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
    cut (d-⊕-& d e) p
        (select true (split-l p₀) P)
        (case (split-l q₀) Q R) ↝ cut d p P Q

  r-select-r :
    ∀{Γ₁ Γ₂ A A′ B B′}
    {P : Process (B ∷ Γ₁)} {Q : Process (A′ ∷ Γ₂)} {R : Process (B′ ∷ Γ₂)}
    (d : Dual A A′) (e : Dual B B′) (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
    cut (d-⊕-& d e) p
        (select false (split-l p₀) P)
        (case (split-l q₀) Q R) ↝ cut e p P R

  r-fork :
    ∀{Γ₁ Γ₂ Γ₃ Δ A B A′ B′}
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} {R : Process (B′ ∷ A′ ∷ Γ₃)}
    (d : Dual A A′) (e : Dual B B′) (p : Γ ≃ Δ + Γ₃) (p₀ : Γ₃ ≃ [] + Γ₃)
    (q : Δ ≃ Γ₁ + Γ₂) (q₀ : Δ ≃ [] + Δ) →
    let _ , p′ , q′ = +-assoc-l p q in
    cut (d-⊗-⅋ d e) p
        (fork (split-l q₀) q P Q)
        (join (split-l p₀) R) ↝ cut d q′ P (cut e (split-r p′) Q R)

  r-client :
    ∀{Γ₁ Γ₂ A A′}
    {P : Process (A ∷ Γ₁)} {Q : Process (A′ ∷ Γ₂)} (d : Dual A A′)
    (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
    cut (d-!-? d) p
      (server (split-l p₀) un P)
      (client (split-l q₀) Q) ↝ cut d p P Q

  r-weaken :
    ∀{Γ₁ Γ₂ A A′}
    {P : Process (A ∷ Γ₁)} {Q : Process Γ₂}
    (d : Dual A A′) (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
    cut (d-!-? d) p
        (server (split-l p₀) un P)
        (weaken (split-l q₀) Q) ↝ weakening un p Q

  r-contract :
    ∀{Γ₁ Γ₂ A A′}
    {P : Process (A ∷ Γ₁)} {Q : Process (¿ A′ ∷ ¿ A′ ∷ Γ₂)}
    (d : Dual A A′) (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
    cut (d-!-? d) p
      (server (split-l p₀) un P)
      (contract (split-l q₀) Q) ↝
      contraction un p
        (cut (d-!-? d) ++≃+
             (server (split-l p₀) un P)
             (cut (d-!-? d) (split-r p) (server (split-l p₀) un P) Q))

  r-cut :
    ∀{Γ₁ Γ₂ A B}
    {P Q : Process (A ∷ Γ₁)} {R : Process (B ∷ Γ₂)}
    (d : Dual A B) (q : Γ ≃ Γ₁ + Γ₂) (r : P ↝ Q) →
    cut d q P R ↝ cut d q Q R

  r-cong :
    ∀{P R Q : Process Γ}
    (p : P ⊒ R) (q : R ↝ Q) → P ↝ Q

data _⇒_ {Γ} : Process Γ → Process Γ → Set where
  refl : ∀{P : Process Γ} → P ⇒ P
  tran : ∀{P Q R : Process Γ} → P ↝ Q → Q ⇒ R → P ⇒ R

Reducible : ∀{Γ} → Process Γ → Set
Reducible P = ∃[ Q ] P ↝ Q
\end{code}

### 2025-07-27 paper/Reduction.lagda.tex
reduction

@@
 
-\begin{code}
+\begin{code}[hide]
 weakening : ∀{Γ Γ₁ Γ₂} (un : Un Γ₁) → Γ ≃ Γ₁ + Γ₂ → Process Γ₂ → Process Γ
@@
     ... | P₃ = #process #shift P₃
+\end{code}
 
+Just like structural precongruence, reduction is formalized as a binary relation
+between processes that are well typed in the very same typing context. Thus, the
+definition of reduction embeds the subject reduction result.
+
+\begin{AgdaAlign}
+\begin{code}
 data _↝_ {Γ} : Process Γ → Process Γ → Set where
-  r-link :
-    ∀{Δ A B} {P : Process (B ∷ Δ)} (d : Dual A B) (e : Dual A B) (p : Γ ≃ B , Δ) →
+\end{code}
+
+There is a constructor for each of the reduction rules in \Cref{tab:semantics}.
+Let us comment a representative selection of them.
+
+The constructor \AgdaInductiveConstructor{r-link} models \RLink.
+
+\begin{code}
+  r-link      : ∀{Δ A B} {P : Process (B ∷ Δ)}
+                (d : Dual A B) (e : Dual A B) (p : Γ ≃ B , Δ) →
                 cut d p (link e (split-l (split-r split-e))) P ↝ #process (#cons p) P
+\end{code}
 
-  r-close :
-    ∀{P : Process Γ} (p₀ : Γ ≃ [] + Γ) (q₀ : Γ ≃ [] + Γ) →
+For generality, the rules quantifies over two proofs $d$ and $e$ of the same
+duality relation between $A$ and $B$. Even though such proofs would necessarily
+be the same, distinguishing them saves us a proof that they are equal whenever
+we want to apply this reduction. 
+%
+Also note that the splitting
+$(\AgdaInductiveConstructor{split-l}~(\AgdaInductiveConstructor{split-r}~\AgdaInductiveConstructor{split-e}))$
+to which \AgdaInductiveConstructor{link} is applied is structured consistently
+with \Cref{tab:semantics}, where the link $\Link\x\y$ being reduced is oriented
+so that the restricted channel $x$ occurs on the left. 
+%
+Finally, the process $P\subst\y\x$ resulting from the reduction is subject to a
+channel substitution in \Cref{tab:semantics}. In the Agda representation of
+processes with nameless channels, the same effect is achieved by the permutation
+$\AgdaFunction{\#cons}~p$, which ``renames'' (the position of) the free
+occurrence of $x$ within $P$ with (the position of) $y$ encoded in the splitting
+$p$.
+
+The constructor \AgdaInductiveConstructor{r-close} models \RClose:
+
+\begin{code}
+  r-close     : ∀{P : Process Γ} (p₀ : Γ ≃ [] + Γ) (q₀ : Γ ≃ [] + Γ) →
                 cut d-𝟙-⊥ p₀ close (wait (split-l q₀) P) ↝ P
+\end{code}
 
-  r-select-l :
-    ∀{Γ₁ Γ₂ A A′ B B′}
+While the \AgdaInductiveConstructor{close} constructor implicitly refers to the
+only free channel occurring in a process of the form $\Close\x$, the
+\AgdaInductiveConstructor{wait} constructor uses a splitting proof of the form
+$(\AgdaInductiveConstructor{split-l}~q_0)$ to be sure that the referenced
+channel is also the restricted one, and therefore matches the one of the
+\AgdaInductiveConstructor{close} process.
+%
+Just like \AgdaInductiveConstructor{r-link} quantifies over multiple proofs for
+the same duality relation, also \AgdaInductiveConstructor{r-close} (and several
+other reduction constructors) quantifies over $p_0$ and $q_0$ which both prove
+the splitting $\Context \simeq \AgdaInductiveConstructor{[]} + \Context$. Since
+the left context is empty, these splittings must be equal and made of a sequence
+of \AgdaInductiveConstructor{split-r} applications followed by
+\AgdaInductiveConstructor{split-e}, but it is easier to quantify them separately
+so that we do not have to worry about proving their equality whenever we wish to
+apply this reduction.
+
+The constructor \AgdaInductiveConstructor{r-select-l} models \RSelect when the
+selected tag is $\InTag_1$, that is \AgdaInductiveConstructor{true} in Agda:
+
+\begin{code}
+  r-select-l  : ∀{Γ₁ Γ₂ A A′ B B′}
                 {P : Process (A ∷ Γ₁)} {Q : Process (A′ ∷ Γ₂)} {R : Process (B′ ∷ Γ₂)}
-    (d : Dual A A′) (e : Dual B B′) (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
+                (d : Dual A A′) (e : Dual B B′)
+                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
                 cut (d-⊕-& d e) p
@@
                     (case (split-l q₀) Q R) ↝ cut d p P Q
+\end{code}
 
+There is not much more to note here (and in other omitted reduction rules),
+except again for the multiple quantifications over the trivial splittings
+$\Context_i \simeq \AgdaInductiveConstructor{[]} + \Context_i$ and the use of
+\AgdaInductiveConstructor{split-l} to make sure that the channel referred to by
+\AgdaInductiveConstructor{select} and \AgdaInductiveConstructor{case} is indeed
+the one restricted by the cut.
+
+\begin{code}[hide]
   r-select-r  :
@@
              (cut (d-!-? d) (split-r p) (server (split-l p₀) un P) Q))
+\end{code}
 
-  r-cut :
-    ∀{Γ₁ Γ₂ A B}
-    {P Q : Process (A ∷ Γ₁)} {R : Process (B ∷ Γ₂)}
-    (d : Dual A B) (q : Γ ≃ Γ₁ + Γ₂) (r : P ↝ Q) →
+Reduction is closed under cuts and by structural pre-congruence as per \RCut and
+\RCong. The corresponding constructors that model these features are shown
+below:
+
+\begin{code}
+  r-cut       : ∀{Γ₁ Γ₂ A B} {P Q : Process (A ∷ Γ₁)} {R : Process (B ∷ Γ₂)}
+                (d : Dual A B) (q : Γ ≃ Γ₁ + Γ₂) → P ↝ Q →
                 cut d q P R ↝ cut d q Q R
 
-  r-cong :
-    ∀{P R Q : Process Γ}
-    (p : P ⊒ R) (q : R ↝ Q) → P ↝ Q
+  r-cong      : ∀{P R Q : Process Γ} → P ⊒ R → R ↝ Q → P ↝ Q
+\end{code}
+\end{AgdaAlign}
+In the following it is also useful to work with the reflexive, transitive
+closure of $\red$, which is defined below:
 
+\begin{code}
 data _⇒_ {Γ} : Process Γ → Process Γ → Set where
@@
   tran  : ∀{P Q R : Process Γ} → P ↝ Q → Q ⇒ R → P ⇒ R
+\end{code}
 
+\begin{code}[hide]
 Reducible : ∀{Γ} → Process Γ → Set

### 2025-07-27 paper/DeadlockFreedom.lagda.tex
drafted deadlock freedom

\begin{code}
open import Data.Sum
open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
open import Data.Bool using (Bool; true; false; if_then_else_)
open import Data.List.Base using ([]; _∷_; [_])
open import Relation.Nullary using (¬_; contradiction)
open import Relation.Binary.PropositionalEquality using (refl)

open import Type
open import Context
open import Process
open import Reduction
open import Congruence
open import Permutations

data Cut {Γ} : Process Γ → Set where
  cut :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    Cut (cut d p P Q)

data Link {Γ} : Process Γ → Set where
  link :
    ∀{A B} (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) → Link (link d p)

data Input : ∀{Γ} → Process Γ → Set where
  fail :
    ∀{Γ Δ}
    (p : Γ ≃ [] + Δ) → Input (fail (split-l p))
  wait :
    ∀{Γ Δ} (p : Γ ≃ [] + Δ) {P : Process Δ} → Input (wait (split-l p) P)
  case :
    ∀{Γ Δ A B} (p : Γ ≃ [] + Δ) {P : Process (A ∷ Δ)} {Q : Process (B ∷ Δ)} →
    Input (case (split-l p) P Q)
  join :
    ∀{Γ Δ A B} (p : Γ ≃ [] + Δ) {P : Process (B ∷ A ∷ Δ)} →
    Input (join (split-l p) P)

data Client : ∀{Γ} → Process Γ → Set where
  client :
    ∀{Γ Δ A} (p : Γ ≃ [] + Δ)
    {P : Process (A ∷ Δ)} →
    Client (client (split-l p) P)
  weaken :
    ∀{Γ Δ A} (p : Γ ≃ [] + Δ)
    {P : Process Δ} →
    Client (weaken {A = A} (split-l p) P)
  contract :
    ∀{Γ Δ A} (p : Γ ≃ [] + Δ)
    {P : Process (¿ A ∷ ¿ A ∷ Δ)} →
    Client (contract (split-l p) P)

data Output : ∀{Γ} → Process Γ → Set where
  close : Output close
  select :
    ∀{Γ Δ A B} (x : Bool) (p : Γ ≃ [] + Δ) {P : Process ((if x then A else B) ∷ Δ)} →
    Output (select x (split-l p) P)
  fork :
    ∀{Γ Δ Δ₁ Δ₂ A B} (p : Γ ≃ [] + Δ) (q : Δ ≃ Δ₁ + Δ₂)
    {P : Process (A ∷ Δ₁)} {Q : Process (B ∷ Δ₂)} →
    Output (fork (split-l p) q P Q)

data Delayed : ∀{Γ} → Process Γ → Set where
  fail :
    ∀{A Γ Δ}
    (p : Γ ≃ ⊤ , Δ) → Delayed (fail (split-r {A} p))
  wait :
    ∀{C Γ Δ} (p : Γ ≃ ⊥ , Δ) {P : Process (C ∷ Δ)} → Delayed (wait (split-r p) P)
  case :
    ∀{Γ Δ C A B} (p : Γ ≃ A & B , Δ) {P : Process (A ∷ C ∷ Δ)} {Q : Process (B ∷ C ∷ Δ)} →
    Delayed (case (split-r p) P Q)
  join :
    ∀{Γ Δ C A B} (p : Γ ≃ A ⅋ B , Δ) {P : Process (B ∷ A ∷ C ∷ Δ)} →
    Delayed (join (split-r p) P)
  select :
    ∀{Γ Δ C A B} (x : Bool) (p : Γ ≃ A ⊕ B , Δ) {P : Process ((if x then A else B) ∷ C ∷ Δ)} →
    Delayed (select x (split-r p) P)
  fork-l :
    ∀{Γ Δ Δ₁ Δ₂ C A B} (p : Γ ≃ A ⊗ B , Δ) (q : Δ ≃ Δ₁ + Δ₂)
    {P : Process (A ∷ C ∷ Δ₁)} {Q : Process (B ∷ Δ₂)} →
    Delayed (fork (split-r p) (split-l q) P Q)
  fork-r :
    ∀{Γ Δ Δ₁ Δ₂ C A B} (p : Γ ≃ A ⊗ B , Δ) (q : Δ ≃ Δ₁ + Δ₂)
    {P : Process (A ∷ Δ₁)} {Q : Process (B ∷ C ∷ Δ₂)} →
    Delayed (fork (split-r p) (split-r q) P Q)
  client :
    ∀{Γ Δ A C} (p : Γ ≃ ¿ A , Δ) {P : Process (A ∷ C ∷ Δ)} →
    Delayed (client (split-r p) P)
  weaken :
    ∀{Γ Δ A C} (p : Γ ≃ ¿ A , Δ) {P : Process (C ∷ Δ)} →
    Delayed (weaken (split-r p) P)
  contract :
    ∀{Γ Δ A C} (p : Γ ≃ ¿ A , Δ) {P : Process (¿ A ∷ ¿ A ∷ C ∷ Δ)} →
    Delayed (contract (split-r p) P)

data Server : ∀{Γ} → Process Γ → Set where
  server :
    ∀{Γ Δ A} (p : Γ ≃ [] + Δ) (un : Un Δ) {P : Process (A ∷ Δ)} →
    Server (server (split-l p) un P)

data DelayedServer : ∀{Γ} → Process Γ → Set where
  server :
    ∀{Γ Δ A C} (p : Γ ≃ ¡ A , Δ) (un : Un Δ) {P : Process (A ∷ ¿ C ∷ Δ)} →
    DelayedServer (server (split-r p) (un-∷ un) P)

Thread : ∀{Γ} → Process Γ → Set
Thread P = Link P ⊎ Delayed P ⊎ Output P ⊎ Input P ⊎ Client P ⊎ Server P ⊎ DelayedServer P

link-thread : ∀{Γ A B} (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) → Thread (link d p)
link-thread d p = inj₁ (link d p)

close-thread : Thread close
close-thread = inj₂ (inj₂ (inj₁ close))

wait-thread : ∀{Γ Δ} (p : Γ ≃ [ ⊥ ] + Δ) {P : Process Δ} → Thread (wait p P)
wait-thread (split-l p) = inj₂ (inj₂ (inj₂ (inj₁ (wait p))))
wait-thread (split-r p) = inj₂ (inj₁ (wait p))

fail-thread : ∀{Γ Δ} (p : Γ ≃ [ ⊤ ] + Δ) → Thread (fail p)
fail-thread (split-l p) = inj₂ (inj₂ (inj₂ (inj₁ (fail p))))
fail-thread (split-r p) = inj₂ (inj₁ (fail p))

case-thread :
  ∀{Γ Δ A B} (p : Γ ≃ [ A & B ] + Δ) {P : Process (A ∷ Δ)} {Q : Process (B ∷ Δ)} →
  Thread (case p P Q)
case-thread (split-l p) = inj₂ (inj₂ (inj₂ (inj₁ (case p))))
case-thread (split-r p) = inj₂ (inj₁ (case p))

join-thread :
  ∀{Γ Δ A B} (p : Γ ≃ [ A ⅋ B ] + Δ) {P : Process (B ∷ A ∷ Δ)} →
  Thread (join p P)
join-thread (split-l p) = inj₂ (inj₂ (inj₂ (inj₁ (join p))))
join-thread (split-r p) = inj₂ (inj₁ (join p))

select-thread :
  ∀{Γ Δ A B} (x : Bool) (p : Γ ≃ A ⊕ B , Δ) {P : Process ((if x then A else B) ∷ Δ)} →
  Thread (select x p P)
select-thread x (split-l p) = inj₂ (inj₂ (inj₁ (select x p)))
select-thread x (split-r p) = inj₂ (inj₁ (select x p))

fork-thread :
  ∀{Γ Δ Δ₁ Δ₂ A B} (p : Γ ≃ [ A ⊗ B ] + Δ) (q : Δ ≃ Δ₁ + Δ₂)
  {P : Process (A ∷ Δ₁)} {Q : Process (B ∷ Δ₂)} →
  Thread (fork p q P Q)
fork-thread (split-l p) q = inj₂ (inj₂ (inj₁ (fork p q)))
fork-thread (split-r p) (split-l q) = inj₂ (inj₁ (fork-l p q))
fork-thread (split-r p) (split-r q) = inj₂ (inj₁ (fork-r p q))

client-thread :
  ∀{Γ Δ A} (p : Γ ≃ [ ¿ A ] + Δ)
  {P : Process (A ∷ Δ)} →
  Thread (client p P)
client-thread (split-l p) = inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (client p)))))
client-thread (split-r p) = inj₂ (inj₁ (client p))

weaken-thread :
  ∀{Γ Δ A} (p : Γ ≃ [ ¿ A ] + Δ)
  {P : Process Δ} →
  Thread (weaken {A = A} p P)
weaken-thread (split-l p) = inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (weaken p)))))
weaken-thread (split-r p) = inj₂ (inj₁ (weaken p))

contract-thread :
  ∀{Γ Δ A} (p : Γ ≃ [ ¿ A ] + Δ)
  {P : Process (¿ A ∷ ¿ A ∷ Δ)} →
  Thread (contract p P)
contract-thread (split-l p) = inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (contract p)))))
contract-thread (split-r p) = inj₂ (inj₁ (contract p))

server-thread :
  ∀{Γ Δ A} (p : Γ ≃ [ ¡ A ] + Δ) (un : Un Δ) {P : Process (A ∷ Δ)} →
  Thread (server p un P)
server-thread (split-l p) un = inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (server p un))))))
server-thread (split-r p) (un-∷ un) = inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (server p un))))))

data CanonicalCut {Γ} : Process Γ → Set where
  cc-link :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    Link P → CanonicalCut (cut d p P Q)
  cc-delayed :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    Delayed P → CanonicalCut (cut d p P Q)
  cc-delayed-server :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    DelayedServer P → Server Q → CanonicalCut (cut d p P Q)
  cc-redex :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    Output P → Input Q → CanonicalCut (cut d p P Q)
  cc-connect :
    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
    Server P → Client Q → CanonicalCut (cut d p P Q)

output-output :
  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Output P × Output Q)
output-output d-𝟙-⊥ (close , ())

output-client :
  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Output P × Client Q)
output-client () (close , client p)

output-server :
  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Output P × Server Q)
output-server () (close , server p un)

output-delayed-server :
  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Output P × DelayedServer Q)
output-delayed-server () (close , server p un)

input-input :
  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Input P × Input Q)
input-input d-⊤-𝟘 (fail p , ())

input-client :
  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Input P × Client Q)
input-client () (fail p₁ , client p)

input-server :
  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Input P × Server Q)
input-server () (fail p₁ , server p un)

input-delayed-server :
  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Input P × DelayedServer Q)
input-delayed-server () (fail p₁ , server p un)

client-client :
  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Client P × Client Q)
client-client () (client p , client p₁)

client-delayed-server :
  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Client P × DelayedServer Q)
client-delayed-server () (client p₁ , server p un)

server-server :
  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Server P × Server Q)
server-server () (server p un , server p₁ un₁)

delayed-server-delayed-served :
  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (DelayedServer P × DelayedServer Q)
delayed-server-delayed-served () (server p un , server p₁ un₁)

canonical-cut :
  ∀{Γ Γ₁ Γ₂ A B}
  {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)}
  (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) →
  Thread P → Thread Q →
  ∃[ R ] CanonicalCut R × cut d p P Q ⊒ R
canonical-cut dc pc (inj₁ x) Qt = _ , cc-link dc pc x , s-refl
canonical-cut dc pc (inj₂ x) (inj₁ y) = _ , cc-link (dual-symm dc) (+-comm pc) y , s-comm dc pc
canonical-cut dc pc (inj₂ (inj₁ x)) (inj₂ y) = _ , cc-delayed dc pc x , s-refl
canonical-cut dc pc (inj₂ (inj₂ x)) (inj₂ (inj₁ y)) = _ , cc-delayed (dual-symm dc) (+-comm pc) y , s-comm dc pc
canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₁ y))) = contradiction (x , y) (output-output dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = _ , cc-redex dc pc x y , s-refl
canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) (output-client dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) (output-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) (output-delayed-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₁ y))) = _ , cc-redex (dual-symm dc) (+-comm pc) y x , s-comm dc pc
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) (output-client (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) (output-server (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) (output-delayed-server (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (x , y) (input-input dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) (input-client dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) (input-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) (input-delayed-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) (input-client (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) (input-server (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) (input-delayed-server (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) (client-client dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = _ , cc-connect (dual-symm dc) (+-comm pc) y x , s-comm dc pc
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) (client-delayed-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = _ , cc-connect dc pc x y , s-refl
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (y , x) (client-delayed-server (dual-symm dc))
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) (server-server dc)
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = _ , cc-delayed-server (dual-symm dc) (+-comm pc) y x , s-comm dc pc
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = _ , cc-delayed-server dc pc x y , s-refl
canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) (delayed-server-delayed-served dc)

Observable : ∀{Γ} → Process Γ → Set
Observable P = ∃[ Q ] P ⊒ Q × Thread Q

Live : ∀{Γ} → Process Γ → Set
Live P = Observable P ⊎ Reducible P

DeadlockFree : ∀{Γ} → Process Γ → Set
DeadlockFree {Γ} P = ∀(Q : Process Γ) → P ⇒ Q → Live Q

⊒Live : ∀{Γ} {P Q : Process Γ} → P ⊒ Q → Live Q → Live P
⊒Live pcong (inj₁ (_ , x , th)) = inj₁ (_ , s-tran pcong x , th)
⊒Live pcong (inj₂ (_ , red)) = inj₂ (_ , r-cong pcong red)

live-cut : ∀{Γ} {P : Process Γ} → CanonicalCut P → Live P
live-cut (cc-link d p (link e (split-l (split-r split-e)))) with dual-fun-r e d
... | refl = inj₂ (_ , r-link d e p)
live-cut (cc-link d p (link e (split-r (split-l split-e)))) with dual-fun-l e (dual-symm d)
... | refl = inj₂ (_ , r-cong (s-cong d p (s-link e (split-r (split-l split-e))) s-refl) (r-link d (dual-symm e) p))
live-cut (cc-redex d-𝟙-⊥ p close (wait q)) with +-empty-l p | +-empty-l q
... | refl | refl = inj₂ (_ , r-close p q)
live-cut (cc-redex (d-⊕-& d e) p (select false q) (case r)) with +-empty-l q | +-empty-l r
... | refl | refl = inj₂ (_ , r-select-r d e p q r)
live-cut (cc-redex (d-⊕-& d e) p (select true q) (case r)) with +-empty-l q | +-empty-l r
... | refl | refl = inj₂ (_ , r-select-l d e p q r)
live-cut (cc-redex (d-⊗-⅋ d e) p (fork q r) (join s)) with +-empty-l q | +-empty-l s
... | refl | refl = inj₂ (_ , r-fork d e p s r q)
live-cut (cc-connect (d-!-? d) p (server q un) (client r)) with +-empty-l q | +-empty-l r
... | refl | refl = inj₂ (_ , r-client d p q r un)
live-cut (cc-connect (d-!-? d) p (server q un) (weaken r)) with +-empty-l q | +-empty-l r
... | refl | refl = inj₂ (_ , r-weaken d p q r un)
live-cut (cc-connect (d-!-? d) p (server q un) (contract r)) with +-empty-l q | +-empty-l r
... | refl | refl = inj₂ (_ , r-contract d p q r un)
live-cut (cc-delayed d p (wait q)) =
  let _ , _ , q′ = +-assoc-l p q in
  inj₁ (_ , s-wait d p q , wait-thread q′)
live-cut (cc-delayed d p (case q)) =
  let _ , _ , q′ = +-assoc-l p q in
  inj₁ (_ , s-case d p q , case-thread q′)
live-cut (cc-delayed d p (join q)) =
  let _ , _ , q′ = +-assoc-l p q in
  inj₁ (_ , s-join d p q , join-thread q′)
live-cut (cc-delayed d p (select false q)) =
  let _ , _ , q′ = +-assoc-l p q in
  inj₁ (_ , s-select-r d p q , select-thread false q′)
live-cut (cc-delayed d p (select true q)) =
  let _ , _ , q′ = +-assoc-l p q in
  inj₁ (_ , s-select-l d p q , select-thread true q′)
live-cut (cc-delayed d p (fork-l q r)) =
  let _ , p′ , q′ = +-assoc-l p q in
  let _ , p′′ , r′ = +-assoc-l p′ r in
  let _ , q′′ , r′′ = +-assoc-r r′ (+-comm p′′) in
  inj₁ (_ , s-fork-l d p q r , fork-thread q′ r′′)
live-cut (cc-delayed d p (fork-r q r)) =
  let _ , p′ , q′ = +-assoc-l p q in
  let _ , p′′ , r′ = +-assoc-l p′ r in
  inj₁ (_ , s-fork-r d p q r , fork-thread q′ r′)
live-cut (cc-delayed d p (fail q)) =
  let _ , p′ , q′ = +-assoc-l p q in
  inj₁ (_ , s-fail d p q , fail-thread q′)
live-cut (cc-delayed d p (client q)) =
  let _ , _ , q′ = +-assoc-l p q in
  inj₁ (_ , s-client d p q , client-thread q′)
live-cut (cc-delayed d p (weaken q)) =
  let _ , _ , q′ = +-assoc-l p q in
  inj₁ (_ , s-weaken d p q , weaken-thread q′)
live-cut (cc-delayed d p (contract q)) =
  let _ , _ , q′ = +-assoc-l p q in
  inj₁ (_ , s-contract d p q , contract-thread q′)
live-cut (cc-delayed-server (d-?-! d) p (server q un) (server r un′)) with +-empty-l r
... | refl =
  let _ , p′ , q′ = +-assoc-l p q in
  inj₁ (_ , s-server d p q r un un′ , server-thread q′ (#un+ p′ un un′))

live : ∀{Γ} (P : Process Γ) → Live P
\end{code}
\begin{code}[hide]
live (link d p) = inj₁ (_ , s-refl , link-thread d p)
live (fail p) = inj₁ (_ , s-refl , fail-thread p)
live close = inj₁ (_ , s-refl , close-thread)
live (wait p P) = inj₁ (_ , s-refl , wait-thread p)
live (select x p P) = inj₁ (_ , s-refl , select-thread x p)
live (case p P Q) = inj₁ (_ , s-refl , case-thread p)
live (fork p q P Q) = inj₁ (_ , s-refl , fork-thread p q)
live (join p P) = inj₁ (_ , s-refl , join-thread p)
live (server p un P) = inj₁ (_ , s-refl , server-thread p un)
live (client p P) = inj₁ (_ , s-refl , client-thread p)
live (weaken p P) = inj₁ (_ , s-refl , weaken-thread p)
live (contract p P) = inj₁ (_ , s-refl , contract-thread p)
live (cut d p P Q) with live P
... | inj₂ (_ , red) = inj₂ (_ , r-cut d p red)
... | inj₁ (_ , Pc , Pt) with live Q
... | inj₂ (_ , red) = inj₂ (_ , r-cong (s-comm d p) (r-cut (dual-symm d) (+-comm p) red))
... | inj₁ (_ , Qc , Qt) with canonical-cut d p Pt Qt
... | _ , cc , pcong = ⊒Live (s-tran (s-cong d p Pc Qc) pcong) (live-cut cc)
\end{code}

\begin{code}
deadlock-freedom : ∀{Γ} (P : Process Γ) → DeadlockFree P
deadlock-freedom P Q reds = live Q
\end{code}

### 2025-07-31 paper/calculus.tex
agda code
better definitions

@@
 latter property is dubbed \emph{well formedness} in the text of the challenge.
-We now formulate these properties using the notation developed so far.
+In this work we also consider \emph{deadlock freedom} which follows from the
+fact that \Calculus is based on linear logic. As we will see, well formedness is
+nothing but a particular form of deadlock freedom. We now formulate these
+properties using the notation developed so far.
 
 Concerning the preservation of typing, this corresponds to the usual subject
-reduction result which is expressed thus.
+reduction result, which is expressed thus:
 
@@
 
-In order to formulate well formedness, we start by defining \emph{reduction
-contexts} are partial processes with a single unguarded hole $\Hole$, thus:
+In order to formulate deadlock freedom, we need to introduce some terminology
+for referring to processes that are unable to make any progress. Deadlock is not
+simply the negation of reducibility: there are irreducible processes that are
+not a deadlock because they would be able to reduce if put in a suitable
+context. For example, $\Wait\x.P$ does not reduce, and yet it is not a deadlock
+because it would be able to make progress when composed in parallel with
+$\Close\x$. In general, even a process like $\Cut\y{\Wait\y.P}{\Wait\z.Q}$ where
+$x\ne y,z$ cannot be considered a deadlock, because the prefixes $\Wait\y$ and
+$\Wait\z$ could be exposed using \SWait and possibly \SComm. Let us make all
+this more precise.
+
+We say that $P$ is a \emph{thread} if it is anything but a cut. In other words,
+a thread is either a link or a process that starts with an input/output action
+of some sort. Note that a thread may contain cuts, but these must all guarded
+underneath the topmost action prefix.
+%
+We also say that $P$ is \emph{observable} if $P \pcong Q$ for some thread $Q$.
+An observable process is a process that exposes an action on a free channel name
+and therefore can interact through that channel, if put into some appropriate
+context. We say that $P$ is \emph{reducible} if $P \red Q$ for some $Q$. A
+reducible process may perform a reduction step.
+%
+Now, a process is a \emph{deadlock} if it is neither observable nor reducible. A
+simple example of deadlock is the process $\Cut\x{\Close\x}{\Close\x}$.
+
+\begin{theorem}[deadlock freedom]
+    \label{thm:df}
+    If $\wtp{P}\Context$, then $P$ is not a deadlock.
+\end{theorem}
+
+In order to formulate \emph{well formedness}~\cite{CarboneEtAl}, we define
+\emph{reduction contexts} as partial processes with a single unguarded hole
+$\Hole$, thus:
 \[
@@
 $P$.
-
-Next we introduce some terminology concerning the shape of processes. We say
-that $P$ is a \emph{thread} if it is anything but a cut. In other words, a
-thread is either a link or a process that starts with an input/output action of
-some sort. Note that a thread may contain cuts, but these must all guarded
-underneath the topmost action prefix.
 %
-We also say that $P$ is \emph{observable} if $P \pcong Q$ for some thread $Q$.
-An observable process is a process that exposes an action on a free channel name
-and therefore can interact through that channel, if put into some appropriate
-reduction context. We say that $P$ is \emph{reducible} if $P \red Q$ for some
-$Q$. A reducible process may perform a reduction step.
-
 Now we observe that if $P$ and $Q$ both act on the same channel $x$ in
-complementary ways, then their parallel composition $\Cut\x{P}{Q}$ must be able
-to reduce according to one of the principal cut reductions described in
-\Cref{tab:semantics}. Therefore, an alternative (and more convenient) way of
-formulating well formedness is the following:
-
-\begin{definition}[well formedness]
-    \label{def:well-formedness}
-    We say that $P$ is \emph{well formed} if $P \pcong
-    \ReductionContext[\Cut\x{Q_1}{Q_2}]$ where both $Q_1$ and $Q_2$ are threads
-    implies that $\Cut\x{Q_1}{Q_2}$ is either observable or reducible.
-\end{definition}
-
-In general a process may be well-formed even if it is unable to reduce. For
-example, $P \eqdef \Cut\x{\Wait\y.\Close\x}{\Wait\x.\Close\z}$ is irreducible
-but also well formed, because $P \pcong
-\Wait\y.\Cut\x{\Close\x}{\Wait\x.\Close\z}$ which is observable.
+complementary ways, then their parallel composition $\Cut\x{P}{Q}$ is reducible
+according to one of the principal cut reductions described in
+\Cref{tab:semantics}. Therefore, an alternative (and more general) way of
+formulating well formedness is this:
+%
+we say that $P$ is \emph{well formed} if $P \pcong
+\ReductionContext[\Cut\x{Q_1}{Q_2}]$ implies that $\Cut\x{Q_1}{Q_2}$ is not a
+deadlock.
 
@@
 
-% \begin{theorem}[deadlock freedom]
-%     \label{thm:df}
-%     If $\wtp{P}\Context$, then $P$ is either observable or reducible.
-% \end{theorem}
-
-% safety
+Note that the properties expressed in \Cref{thm:df,thm:type-safety} are
+invariant under reductions thanks to \Cref{thm:red}.

### 2025-08-01 paper/DeadlockFreedom.lagda.tex
drafted deadlock freedom

@@
-\begin{code}
+\subsection{Deadlock Freedom}
+\label{sec:deadlock-freedom-agda}
+
+As we have seen in \Cref{sec:calculus}, the deadlock freedom result
+(\Cref{thm:df}) rests on some notions and predicates about processes which we
+now formalize in Agda. First of all we need to distinguish \emph{threads} from
+\emph{cuts}. As it turns out, in the formalization it is convenient to provide
+an even more fine-grained classification of threads, distinguishing between
+links, input/output actions on linear/shared channels, and sometimes also
+whether such actions operate on free or bound channels. We define predicates for
+each of these process classes:
+
+\begin{code}[hide]
 open import Data.Sum
@@
 open import Context
+open import Permutations
 open import Process
@@
 open import Congruence
-open import Permutations
+\end{code}
 
-data Cut {Γ} : Process Γ → Set where
+\begin{code}
+data Cut            : ∀{Γ} → Process Γ → Set
+data Link           : ∀{Γ} → Process Γ → Set
+data Input          : ∀{Γ} → Process Γ → Set
+data Output         : ∀{Γ} → Process Γ → Set
+data Delayed        : ∀{Γ} → Process Γ → Set
+data Client         : ∀{Γ} → Process Γ → Set
+data Server         : ∀{Γ} → Process Γ → Set
+data DelayedServer  : ∀{Γ} → Process Γ → Set
+\end{code}
+
+These predicates are relatively uninteresting since they are essentially
+isomorphic to the relevant parts of the \AgdaDatatype{Process} datatype. The
+only remarkable aspect is that \AgdaDatatype{Input}, \AgdaDatatype{Output},
+\AgdaDatatype{Client} and \AgdaDatatype{Server} the channel being acted upon by
+the thread is the \emph{first} in $\Context$, hence it is the \emph{most
+recently introduced} channel, whereas in \AgdaDatatype{Delayed} and
+\AgdaDatatype{DelayedServer} the channel is older. This allows us to distinguish
+those threads that, in the context of a cut, operate on the channel bound by the
+cut or on a free channel. As an illustration, consider the
+\AgdaInductiveConstructor{wait} case in \AgdaDatatype{Input} and in
+\AgdaDatatype{Delayed}. In the former predicate we have
+
+\begin{AgdaAlign}
+\begin{code}[hide]
+data Cut where
   cut :
-    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
+    ∀{Γ Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
     {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
@@
 
-data Link {Γ} : Process Γ → Set where
+data Link where
   link :
-    ∀{A B} (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) → Link (link d p)
+    ∀{Γ A B} (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) → Link (link d p)
 
-data Input : ∀{Γ} → Process Γ → Set where
+data Input where
   fail :
@@
     (p : Γ ≃ [] + Δ) → Input (fail (split-l p))
-  wait :
-    ∀{Γ Δ} (p : Γ ≃ [] + Δ) {P : Process Δ} → Input (wait (split-l p) P)
+\end{code}
+
+\begin{code}
+  wait : ∀{Γ Δ} (p : Γ ≃ [] + Δ) {P : Process Δ} → Input (wait (split-l p) P)
+\end{code}
+where the \AgdaInductiveConstructor{split-l} constructor indicates that the
+thread operates on the most recent channel. In the latter predicate we have
+
+\begin{code}[hide]
   case :
@@
 
-data Client : ∀{Γ} → Process Γ → Set where
+data Client where
   client :
@@
 
-data Output : ∀{Γ} → Process Γ → Set where
+data Output where
   close : Output close
@@
 
-data Delayed : ∀{Γ} → Process Γ → Set where
+data Delayed where
   fail :
@@
     (p : Γ ≃ ⊤ , Δ) → Delayed (fail (split-r {A} p))
-  wait :
-    ∀{C Γ Δ} (p : Γ ≃ ⊥ , Δ) {P : Process (C ∷ Δ)} → Delayed (wait (split-r p) P)
+\end{code}
+
+\begin{code}
+  wait : ∀{C Γ Δ} (p : Γ ≃ ⊥ , Δ) {P : Process (C ∷ Δ)} →
+         Delayed (wait (split-r p) P)
+\end{code}
+where the \AgdaInductiveConstructor{split-r} constructor indicates that the
+thread operates on a channel other than the most recent one.
+
+\begin{code}[hide]
   case :
@@
 
-data Server : ∀{Γ} → Process Γ → Set where
+data Server where
   server :
@@
 
-data DelayedServer : ∀{Γ} → Process Γ → Set where
+data DelayedServer where
   server :
@@
     DelayedServer (server (split-r p) (un-∷ un) P)
+\end{code}
+\end{AgdaAlign}
 
+Now the \AgdaDatatype{Thread} predicate is simply the disjunction of all the
+previous ones.
+
+\begin{code}
 Thread : ∀{Γ} → Process Γ → Set
-Thread P = Link P ⊎ Delayed P ⊎ Output P ⊎ Input P ⊎ Client P ⊎ Server P ⊎ DelayedServer P
+Thread P = Link P ⊎ Delayed P ⊎ Output P ⊎ Input P ⊎
+           Client P ⊎ Server P ⊎ DelayedServer P
+\end{code}
 
+We can now define observability, reducibility and aliveness in the expected way:
+
+\begin{code}
+Observable : ∀{Γ} → Process Γ → Set
+Observable P = ∃[ Q ] P ⊒ Q × Thread Q
+
+Reducible : ∀{Γ} → Process Γ → Set
+Reducible P = ∃[ Q ] P ↝ Q
+
+Alive : ∀{Γ} → Process Γ → Set
+Alive P = Observable P ⊎ Reducible P
+\end{code}
+
+\begin{code}[hide]
 link-thread : ∀{Γ A B} (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) → Thread (link d p)
@@
 server-thread (split-r p) (un-∷ un) = inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (server p un))))))
+\end{code}
 
+In order to prove that every (well-typed) process is alive, it is convenient to
+also define a ``canonical'' form for cuts, which is a form that matches at least
+one of the left-hand-sides of the rules for structural pre-congruence or
+reduction (\Cref{tab:semantics}). This is the notion where the fine-grained
+classification of threads introduced earlier comes into play.
+
+\begin{code}
 data CanonicalCut {Γ} : Process Γ → Set where
@@
     Server P → Client Q → CanonicalCut (cut d p P Q)
+\end{code}
 
+A \emph{canonical cut} $\Cut\x{P}{Q}$ has one of these forms:
+\begin{itemize}
+\item $P$ is a link (\AgdaInductiveConstructor{cc-link}), or
+\item $P$ operates on a channel other than $x$ and is not a server
+(\AgdaInductiveConstructor{cc-delayed}), or
+\item $P$ and $Q$ are servers and $P$ operates on a channel other than $x$
+(\AgdaInductiveConstructor{cc-delayed-server}), or
+\item $P$ performs an output on $x$ and $Q$ performs an input on $x$
+(\AgdaInductiveConstructor{cc-redex} and \AgdaInductiveConstructor{cc-connect}).
+\end{itemize}
+
+\begin{code}[hide]
 output-output :
@@
 delayed-server-delayed-served () (server p un , server p₁ un₁)
+\end{code}
 
+Every cut $\Cut\x{P}{Q}$ where both $P$ and $Q$ are threads can be rewritten
+into a canonical cut using structural pre-congruence:
+
+\begin{code}
 canonical-cut :
-  ∀{Γ Γ₁ Γ₂ A B}
-  {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)}
+  ∀{Γ Γ₁ Γ₂ A B} {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)}
   (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) →
-  Thread P → Thread Q →
-  ∃[ R ] CanonicalCut R × cut d p P Q ⊒ R
+  Thread P → Thread Q → ∃[ R ] CanonicalCut R × cut d p P Q ⊒ R
+\end{code}
+
+\begin{code}[hide]
 canonical-cut dc pc (inj₁ x) Qt = _ , cc-link dc pc x , s-refl
@@
 
-Observable : ∀{Γ} → Process Γ → Set
-Observable P = ∃[ Q ] P ⊒ Q × Thread Q
+⊒Alive : ∀{Γ} {P Q : Process Γ} → P ⊒ Q → Alive Q → Alive P
+⊒Alive pcong (inj₁ (_ , x , th)) = inj₁ (_ , s-tran pcong x , th)
+⊒Alive pcong (inj₂ (_ , red)) = inj₂ (_ , r-cong pcong red)
+\end{code}
 
-Live : ∀{Γ} → Process Γ → Set
-Live P = Observable P ⊎ Reducible P
+It is easy to prove that every canonical cut is alive, either reducing it or
+applying structural pre-congruence to rewrite it into a thread.
 
-DeadlockFree : ∀{Γ} → Process Γ → Set
-DeadlockFree {Γ} P = ∀(Q : Process Γ) → P ⇒ Q → Live Q
+\begin{code}
+canonical-cut-alive : ∀{Γ} {P : Process Γ} → CanonicalCut P → Alive P
+\end{code}
 
-⊒Live : ∀{Γ} {P Q : Process Γ} → P ⊒ Q → Live Q → Live P
-⊒Live pcong (inj₁ (_ , x , th)) = inj₁ (_ , s-tran pcong x , th)
-⊒Live pcong (inj₂ (_ , red)) = inj₂ (_ , r-cong pcong red)
-
-live-cut : ∀{Γ} {P : Process Γ} → CanonicalCut P → Live P
-live-cut (cc-link d p (link e (split-l (split-r split-e)))) with dual-fun-r e d
+\begin{code}[hide]
+canonical-cut-alive (cc-link d p (link e (split-l (split-r split-e)))) with dual-fun-r e d
 ... | refl = inj₂ (_ , r-link d e p)
-live-cut (cc-link d p (link e (split-r (split-l split-e)))) with dual-fun-l e (dual-symm d)
+canonical-cut-alive (cc-link d p (link e (split-r (split-l split-e)))) with dual-fun-l e (dual-symm d)
 ... | refl = inj₂ (_ , r-cong (s-cong d p (s-link e (split-r (split-l split-e))) s-refl) (r-link d (dual-symm e) p))
-live-cut (cc-redex d-𝟙-⊥ p close (wait q)) with +-empty-l p | +-empty-l q
+canonical-cut-alive (cc-redex d-𝟙-⊥ p close (wait q)) with +-empty-l p | +-empty-l q
 ... | refl | refl = inj₂ (_ , r-close p q)
-live-cut (cc-redex (d-⊕-& d e) p (select false q) (case r)) with +-empty-l q | +-empty-l r
+canonical-cut-alive (cc-redex (d-⊕-& d e) p (select false q) (case r)) with +-empty-l q | +-empty-l r
 ... | refl | refl = inj₂ (_ , r-select-r d e p q r)
-live-cut (cc-redex (d-⊕-& d e) p (select true q) (case r)) with +-empty-l q | +-empty-l r
+canonical-cut-alive (cc-redex (d-⊕-& d e) p (select true q) (case r)) with +-empty-l q | +-empty-l r
 ... | refl | refl = inj₂ (_ , r-select-l d e p q r)
-live-cut (cc-redex (d-⊗-⅋ d e) p (fork q r) (join s)) with +-empty-l q | +-empty-l s
+canonical-cut-alive (cc-redex (d-⊗-⅋ d e) p (fork q r) (join s)) with +-empty-l q | +-empty-l s
 ... | refl | refl = inj₂ (_ , r-fork d e p s r q)
-live-cut (cc-connect (d-!-? d) p (server q un) (client r)) with +-empty-l q | +-empty-l r
+canonical-cut-alive (cc-connect (d-!-? d) p (server q un) (client r)) with +-empty-l q | +-empty-l r
 ... | refl | refl = inj₂ (_ , r-client d p q r un)
-live-cut (cc-connect (d-!-? d) p (server q un) (weaken r)) with +-empty-l q | +-empty-l r
+canonical-cut-alive (cc-connect (d-!-? d) p (server q un) (weaken r)) with +-empty-l q | +-empty-l r
 ... | refl | refl = inj₂ (_ , r-weaken d p q r un)
-live-cut (cc-connect (d-!-? d) p (server q un) (contract r)) with +-empty-l q | +-empty-l r
+canonical-cut-alive (cc-connect (d-!-? d) p (server q un) (contract r)) with +-empty-l q | +-empty-l r
 ... | refl | refl = inj₂ (_ , r-contract d p q r un)
-live-cut (cc-delayed d p (wait q)) =
+canonical-cut-alive (cc-delayed d p (wait q)) =
   let _ , _ , q′ = +-assoc-l p q in
   inj₁ (_ , s-wait d p q , wait-thread q′)
-live-cut (cc-delayed d p (case q)) =
+canonical-cut-alive (cc-delayed d p (case q)) =
   let _ , _ , q′ = +-assoc-l p q in
   inj₁ (_ , s-case d p q , case-thread q′)
-live-cut (cc-delayed d p (join q)) =
+canonical-cut-alive (cc-delayed d p (join q)) =
   let _ , _ , q′ = +-assoc-l p q in
   inj₁ (_ , s-join d p q , join-thread q′)
-live-cut (cc-delayed d p (select false q)) =
+canonical-cut-alive (cc-delayed d p (select false q)) =
   let _ , _ , q′ = +-assoc-l p q in
   inj₁ (_ , s-select-r d p q , select-thread false q′)
-live-cut (cc-delayed d p (select true q)) =
+canonical-cut-alive (cc-delayed d p (select true q)) =
   let _ , _ , q′ = +-assoc-l p q in
   inj₁ (_ , s-select-l d p q , select-thread true q′)
-live-cut (cc-delayed d p (fork-l q r)) =
+canonical-cut-alive (cc-delayed d p (fork-l q r)) =
   let _ , p′ , q′ = +-assoc-l p q in
@@
   inj₁ (_ , s-fork-l d p q r , fork-thread q′ r′′)
-live-cut (cc-delayed d p (fork-r q r)) =
+canonical-cut-alive (cc-delayed d p (fork-r q r)) =
   let _ , p′ , q′ = +-assoc-l p q in
@@
   inj₁ (_ , s-fork-r d p q r , fork-thread q′ r′)
-live-cut (cc-delayed d p (fail q)) =
+canonical-cut-alive (cc-delayed d p (fail q)) =
   let _ , p′ , q′ = +-assoc-l p q in
   inj₁ (_ , s-fail d p q , fail-thread q′)
-live-cut (cc-delayed d p (client q)) =
+canonical-cut-alive (cc-delayed d p (client q)) =
   let _ , _ , q′ = +-assoc-l p q in
   inj₁ (_ , s-client d p q , client-thread q′)
-live-cut (cc-delayed d p (weaken q)) =
+canonical-cut-alive (cc-delayed d p (weaken q)) =
   let _ , _ , q′ = +-assoc-l p q in
   inj₁ (_ , s-weaken d p q , weaken-thread q′)
-live-cut (cc-delayed d p (contract q)) =
+canonical-cut-alive (cc-delayed d p (contract q)) =
   let _ , _ , q′ = +-assoc-l p q in
   inj₁ (_ , s-contract d p q , contract-thread q′)
-live-cut (cc-delayed-server (d-?-! d) p (server q un) (server r un′)) with +-empty-l r
+canonical-cut-alive (cc-delayed-server (d-?-! d) p (server q un) (server r un′)) with +-empty-l r
 ... | refl =
@@
   inj₁ (_ , s-server d p q r un un′ , server-thread q′ (#un+ p′ un un′))
+\end{code}
 
-live : ∀{Γ} (P : Process Γ) → Live P
+Now deadlock freedom for $P$ can be proved by induction on $P$.
+
+\begin{code}
+deadlock-freedom : ∀{Γ} (P : Process Γ) → Alive P
 \end{code}
+
+When $P$ is a thread, then it is obviously observable and hence alive. When $P$
+is a cut $\Cut\x{P}{Q}$, \AgdaFunction{deadlock-freedom} is applied recursively
+on $P$ and $Q$. If either of these applications yields a reduction, then the
+whole cut is reducible and therefore alive. If both applications yield a thread,
+then we conclude that the cut is alive first rewriting it into a canonical cut
+(\AgdaFunction{canonical-cut}) and then using
+\AgdaFunction{canonical-cut-alive}.
+
 \begin{code}[hide]
-live (link d p) = inj₁ (_ , s-refl , link-thread d p)
-live (fail p) = inj₁ (_ , s-refl , fail-thread p)
-live close = inj₁ (_ , s-refl , close-thread)
-live (wait p P) = inj₁ (_ , s-refl , wait-thread p)
-live (select x p P) = inj₁ (_ , s-refl , select-thread x p)
-live (case p P Q) = inj₁ (_ , s-refl , case-thread p)
-live (fork p q P Q) = inj₁ (_ , s-refl , fork-thread p q)
-live (join p P) = inj₁ (_ , s-refl , join-thread p)
-live (server p un P) = inj₁ (_ , s-refl , server-thread p un)
-live (client p P) = inj₁ (_ , s-refl , client-thread p)
-live (weaken p P) = inj₁ (_ , s-refl , weaken-thread p)
-live (contract p P) = inj₁ (_ , s-refl , contract-thread p)
-live (cut d p P Q) with live P
+deadlock-freedom (link d p) = inj₁ (_ , s-refl , link-thread d p)
+deadlock-freedom (fail p) = inj₁ (_ , s-refl , fail-thread p)
+deadlock-freedom close = inj₁ (_ , s-refl , close-thread)
+deadlock-freedom (wait p P) = inj₁ (_ , s-refl , wait-thread p)
+deadlock-freedom (select x p P) = inj₁ (_ , s-refl , select-thread x p)
+deadlock-freedom (case p P Q) = inj₁ (_ , s-refl , case-thread p)
+deadlock-freedom (fork p q P Q) = inj₁ (_ , s-refl , fork-thread p q)
+deadlock-freedom (join p P) = inj₁ (_ , s-refl , join-thread p)
+deadlock-freedom (server p un P) = inj₁ (_ , s-refl , server-thread p un)
+deadlock-freedom (client p P) = inj₁ (_ , s-refl , client-thread p)
+deadlock-freedom (weaken p P) = inj₁ (_ , s-refl , weaken-thread p)
+deadlock-freedom (contract p P) = inj₁ (_ , s-refl , contract-thread p)
+deadlock-freedom (cut d p P Q) with deadlock-freedom P
 ... | inj₂ (_ , red) = inj₂ (_ , r-cut d p red)
-... | inj₁ (_ , Pc , Pt) with live Q
+... | inj₁ (_ , Pc , Pt) with deadlock-freedom Q
 ... | inj₂ (_ , red) = inj₂ (_ , r-cong (s-comm d p) (r-cut (dual-symm d) (+-comm p) red))
 ... | inj₁ (_ , Qc , Qt) with canonical-cut d p Pt Qt
-... | _ , cc , pcong = ⊒Live (s-tran (s-cong d p Pc Qc) pcong) (live-cut cc)
+... | _ , cc , pcong = ⊒Alive (s-tran (s-cong d p Pc Qc) pcong) (canonical-cut-alive cc)
 \end{code}
-
-\begin{code}
-deadlock-freedom : ∀{Γ} (P : Process Γ) → DeadlockFree P
-deadlock-freedom P Q reds = live Q
-\end{code}

### 2025-08-01 paper/Process.lagda.tex
drafted deadlock freedom

@@
 the types $A$ and $B$ also in the opposite order, provided that the reduction
-rule \RFork (formalized in \Cref{sec:semantics-agda}) is suitably adjusted.
+rule \RFork (formalized in \Cref{sec:reduction-agda}) is suitably adjusted.
 

### 2025-08-01 paper/Reduction.lagda.tex
drafted deadlock freedom

@@
 \end{AgdaAlign}
-In the following it is also useful to work with the reflexive, transitive
-closure of $\red$, which is defined below:
-
-\begin{code}
-data _⇒_ {Γ} : Process Γ → Process Γ → Set where
-  refl  : ∀{P : Process Γ} → P ⇒ P
-  tran  : ∀{P Q R : Process Γ} → P ↝ Q → Q ⇒ R → P ⇒ R
-\end{code}
-
-\begin{code}[hide]
-Reducible : ∀{Γ} → Process Γ → Set
-Reducible P = ∃[ Q ] P ↝ Q
-\end{code}

### 2025-08-01 paper/calculus.tex
drafted deadlock freedom

@@
 reductions; (2) peer endpoints are always used in complementary ways. This
-latter property is dubbed \emph{well formedness} in the text of the challenge.
-In this work we also consider \emph{deadlock freedom} which follows from the
-fact that \Calculus is based on linear logic. As we will see, well formedness is
-nothing but a particular form of deadlock freedom. We now formulate these
-properties using the notation developed so far.
+latter property is called \emph{well formedness} in the challenge. In this work
+we also consider \emph{deadlock freedom}, which is more general than well
+formedness and holds for \Calculus since its type system is based on linear
+logic. We now formulate these properties using the notation developed so far.
 
@@
     \label{thm:red}
-    If $P \red Q$, then $\wtp{P}\Context$ implies $\wtp{Q}\Context$.
+    If $P \red Q$ then $\wtp{P}\Context$ implies $\wtp{Q}\Context$.
 \end{theorem}
 
-In order to formulate deadlock freedom, we need to introduce some terminology
-for referring to processes that are unable to make any progress. Deadlock is not
-simply the negation of reducibility: there are irreducible processes that are
-not a deadlock because they would be able to reduce if put in a suitable
-context. For example, $\Wait\x.P$ does not reduce, and yet it is not a deadlock
-because it would be able to make progress when composed in parallel with
-$\Close\x$. In general, even a process like $\Cut\y{\Wait\y.P}{\Wait\z.Q}$ where
-$x\ne y,z$ cannot be considered a deadlock, because the prefixes $\Wait\y$ and
-$\Wait\z$ could be exposed using \SWait and possibly \SComm. Let us make all
-this more precise.
+In order to formulate deadlock freedom, we first need to introduce some
+terminology for referring to processes that are unable to make any progress. A
+simple example of deadlock is the process $\Cut\x{\Close\x}{\Close\x}$. This
+process is unable to reduce (because a $\Close\x$ process is meant to interact
+with a process of the form $\Wait\x.P$) and, more generally, it is unable to
+interact regardless of the context in which it occurs because the $\Close\x$
+sub-processes it contains are blocked on the channel $x$ bound by the cut.
+%
+In general, the property of being a deadlock is not simply the inability to
+reduce: there are irreducible processes that are not a deadlock because they
+would be able to reduce if put in a suitable context. For example, $\Wait\x.P$
+does not reduce, and yet it is not a deadlock because it would be able to make
+progress when composed in parallel with $\Close\x$. In general, even a process
+like $\Cut\y{\Wait\y.P}{\Wait\z.Q}$ where $x\ne y,z$ cannot be considered a
+deadlock, because the prefixes $\Wait\y$ and $\Wait\z$ could be exposed using
+\SWait and possibly \SComm. Let us make all this more precise.
 
@@
 a thread is either a link or a process that starts with an input/output action
-of some sort. Note that a thread may contain cuts, but these must all guarded
-underneath the topmost action prefix.
+of some sort. Note that a thread may contain cuts, but these cuts must all
+guarded underneath the topmost action prefix.
 %
@@
 %
-Now, a process is a \emph{deadlock} if it is neither observable nor reducible. A
-simple example of deadlock is the process $\Cut\x{\Close\x}{\Close\x}$.
+We say that a process is \emph{alive} if it is either observable or reducible
+and that it is a \emph{deadlock} if it is not alive.
 
@@
     \label{thm:df}
-    If $\wtp{P}\Context$, then $P$ is not a deadlock.
+    If $\wtp{P}\Context$ then $P$ is alive.
 \end{theorem}
 
-In order to formulate \emph{well formedness}~\cite{CarboneEtAl}, we define
+In order to formulate \emph{well formedness}~\cite{CarboneEtAl24}, we define
 \emph{reduction contexts} as partial processes with a single unguarded hole
@@
 we say that $P$ is \emph{well formed} if $P \pcong
-\ReductionContext[\Cut\x{Q_1}{Q_2}]$ implies that $\Cut\x{Q_1}{Q_2}$ is not a
-deadlock.
+\ReductionContext[\Cut\x{Q_1}{Q_2}]$ implies that $\Cut\x{Q_1}{Q_2}$ is alive.
 
@@
     \label{thm:type-safety}
-    If $\wtp{P}\Context$, then $P$ is well formed.
+    If $\wtp{P}\Context$ then $P$ is well formed.
 \end{theorem}

### 2025-08-01 paper/Safety.lagda.tex
drafted type safety

\section{Type Safety}
\label{sec:safety-agda}

\begin{code}[hide]
open import Data.Sum
open import Data.Product using (_,_)
open import Relation.Nullary using (¬_; contradiction)
open import Data.List.Base using ([]; _∷_; [_])

open import Type
open import Context
open import Process
open import Congruence
open import Reduction
open import DeadlockFreedom
\end{code}

We have seen that type safety is a simple instance of deadlock freedom, which is
made even simpler to formalize in our development where processes are
intrinsically typed. We start by defining reduction contexts as processes with a
single hole. In our intrinsically-typed formalization, reduction contexts are
typed and indexed by two typing contexts $\ContextD$ and $\ContextC$. The latter
is the typing context under which the reduction context as a whole is well
typed. The former is the typing context of the hole, which is invariant for a
given reduction context and therefore specified as a parameter of the data type:

\begin{code}
data ReductionContext (Δ : Context) : Context → Set where
  hole   : ReductionContext Δ Δ
  cut-l  : ∀{Γ Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) →
           ReductionContext Δ (A ∷ Γ₁) → Process (B ∷ Γ₂) →
           ReductionContext Δ Γ
  cut-r  : ∀{Γ Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) →
           Process (A ∷ Γ₁) → ReductionContext Δ (B ∷ Γ₂) →
           ReductionContext Δ Γ
\end{code}

Substitution inside a reduction context $\ReductionContext$ is a straightforward
function $\AgdaFunction{\_⟦\_⟧}$ that operates recursively on the structure of
$\ReductionContext$:

\begin{code}
_⟦_⟧ : ∀{Γ Δ} → ReductionContext Δ Γ → Process Δ → Process Γ
hole ⟦ P ⟧           = P
cut-l d p 𝒞 Q ⟦ P ⟧  = cut d p (𝒞 ⟦ P ⟧) Q
cut-r d p Q 𝒞 ⟦ P ⟧  = cut d p Q (𝒞 ⟦ P ⟧)
\end{code}

A process $P$ is well formed if every unguarded sub-process $Q$ in it is alive.

\begin{code}
WellFormed        : ∀{Γ} → Process Γ → Set
WellFormed {Γ} P  = ∀{Δ} {𝒞 : ReductionContext Δ Γ} {Q : Process Δ} →
                    P ⊒ (𝒞 ⟦ Q ⟧) → Alive Q
\end{code}

The proof of type safety ends up being a trivial application of
\AgdaFunction{deadlock-freedom}.

\begin{code}
type-safety : ∀{Γ} (P : Process Γ) → WellFormed P
type-safety P {_} {_} {Q} _ = deadlock-freedom Q
\end{code}

### 2025-08-01 paper/calculus.tex
drafted type safety

@@
 \Cref{tab:semantics}. Therefore, an alternative (and more general) way of
-formulating well formedness is this:
+formulating well formedness is simply this:
 %
-we say that $P$ is \emph{well formed} if $P \pcong
-\ReductionContext[\Cut\x{Q_1}{Q_2}]$ implies that $\Cut\x{Q_1}{Q_2}$ is alive.
+we say that $P$ is \emph{well formed} if $P \pcong \ReductionContext[Q]$ implies
+that $Q$ is alive.
 

### 2025-08-01 paper/conclusion.tex
drafted type safety

\section{Concluding Remarks}
\label{sec:conclusion}

context splitting hard to read but overall doable

structural pre-congruence by far most involved, and it will grow even further in
order to prove more complex properties (eg confluence)

linear pi-calculus is good

### 2025-08-01 paper/Permutations.lagda.tex
drafted deadlock freedom
added prose for permutations

@@
 \subsection{Context Permutations}
-\label{sec:permutation-agda}
+\label{sec:permutations-agda}
+
+According to our nameless representation of channels, it is the \emph{order} of
+channels in a typing context $\Context$ that determines the binding location in
+the structure of a process. When the binding structure of a process changes, \eg
+because a structural pre-congruence rule is applied, or when a channel
+substitution occurs, \cf the right-hand side of the \LinkRule reduction, the
+typing context must be suitably rearranged. Such rearrangement is in fact a
+\emph{permutation} of the elements of $\Context$.
+
+We define typing context permutations inductively, as a binary relation
+\AgdaDatatype{\_\#\_}:
 
@@
   #tran  : ∀{Γ Δ Θ} → Γ # Δ → Δ # Θ → Γ # Θ
+\end{code}
 
+This definition is essentially the same as the notion of list permutation in
+Agda's standard library. Each constructor represents a particular kind of
+permutation: \AgdaInductiveConstructor{\#refl} for the trivial permutation that
+does not change anything; \AgdaInductiveConstructor{\#here} for the permutation
+that swaps the first two elements of a typing context;
+\AgdaInductiveConstructor{\#next} for the permutation that applies a
+sub-permutation to the tail of a typing context;
+\AgdaInductiveConstructor{\#tran} for the sequential composition of
+permutations.
+%
+We have preferred defining our own notion of permutation for simplicity: the
+\AgdaInductiveConstructor{\#here} constructor does not need a sub-permutation
+for the tail of the typing context, which can always be performed, if needed,
+using \AgdaInductiveConstructor{\#next} and \AgdaInductiveConstructor{\#tran}.
+Also, and more importantly, this definition is \emph{monomorphic} (it does not
+need to operate on lists of arbitrary type), resulting in a more streamlined
+usage in the rest of the code.
+
+\begin{code}[hide]
 #sym : ∀{Γ Δ} → Γ # Δ → Δ # Γ
@@
 #rot = #tran (#next #here) #here
+\end{code}
 
+Just like permutations from Agda's standard library, also \AgdaDatatype{\_\#\_}
+enjoys a number of expected (and useful) properties, such as symmetry. Among all
+properties, in the following we will make use of a specific one that connects
+context splitting and permutations: if $\Context \simeq A, \ContextD$, then
+$\Context$ must be a permutation of $A~\AgdaInductiveConstructor{∷}~\ContextD$.
+
+\begin{code}
 #cons : ∀{A Γ Δ} → Γ ≃ A , Δ → (A ∷ Δ) # Γ
@@
 #cons (split-r p) = #tran #here (#next (#cons p))
+\end{code}
 
+\begin{code}[hide]
 #split : ∀{Γ Γ₁ Γ₂ Δ} → Γ # Δ → Γ ≃ Γ₁ + Γ₂ → ∃[ Δ₁ ] ∃[ Δ₂ ] (Δ ≃ Δ₁ + Δ₂ × Γ₁ # Δ₁ × Γ₂ # Δ₂)

### 2025-08-02 paper/calculus.tex
++

@@
 
-The calculus of pure sessions that is formalized in our development is dubbed
-\Calculus and is a linear version of \CP~\citep{Wadler14,GayVasconcelos25},
-where every session endpoint is meant to be used \emph{exactly once}. This
-differs from \CP and other session calculi where session endpoints can be used
-repeatedly, albeit sequentially.
+The calculus of pure sessions we formalized is dubbed \Calculus and is a linear
+version of \CP~\citep{Wadler14,GayVasconcelos25} such that every session
+endpoint is meant to be used \emph{exactly once}. This differs from \CP and
+other session calculi where session endpoints can be used repeatedly, albeit
+sequentially. The key idea is that each message sent on a session endpoint is
+paired with a \emph{fresh continuation channel} on which the rest of the
+conversation takes place.
+%
+This way of encoding sessions using explicit continuations has been first
+described by \citet{Kobayashi} and later studied extensively by
+\citet{DardhaGiachinoSangiorgi17}.
 
@@
 %
-A link $\Link\x\y$ denotes the merging of the $x$ and $y$ channels, so that
-every message sent on one of the channels is forwarded to the other. This form
-is useful for modeling \emph{free outputs}, as we will see later.\Luca{Mettere
-esempio!}
+A link $\Link\x\y$ denotes the merging of the channels $x$ and $y$, so that
+every message sent on one of the channels is forwarded to the other. As
+discussed in the literature, this form is useful for modeling \emph{delegation},
+that is the transmission of an existing channel over another channel.
 %

### 2025-08-02 paper/calculus.tex
revised calculus, biblio

@@
 This way of encoding sessions using explicit continuations has been first
-described by \citet{Kobayashi} and later studied extensively by
+described by \citet[extended version]{Kobayashi02b} and later studied by
 \citet{DardhaGiachinoSangiorgi17}.
@@
 every message sent on one of the channels is forwarded to the other. As
-discussed in the literature, this form is useful for modeling \emph{delegation},
-that is the transmission of an existing channel over another channel.
+discussed in the literature, this form is useful also for modeling
+\emph{delegation}, that is the exchange of an existing session endpoint in
+another session.
 %
@@
 of a termination signal on the channel $x$. The former process has no
-continuation, while the latter process continues as $P$ once the signa has been
+continuation, while the latter process continues as $P$ once the signal has been
 received.
@@
 %
-The $\Select\x{\InTag_i}\z.P$ and $\Case\x\z{Q_1}{Q_2}$ processes respectively
-denote the output and input of a label $\InTag_i$ along with a fresh
-continuation $z$ on the channel $x$. Depending on the label $\InTag_i$, the
-receiver continues as $Q_i$. In practice it is useful to consider arbitrary
-(domain-specific) sets of tags instead of the canonical two $\InTag_1$ and
+The $\Select\x{\InTag_i}\z.P$ for $i=1,2$ and $\Case\x\z{Q_1}{Q_2}$ processes
+respectively denote the output and input of a label $\InTag_i$ along with a
+fresh continuation $z$ on the channel $x$. The receiver continues as $Q_i$
+depending on the received label $\InTag_i$. In practice it is useful to consider
+arbitrary (domain-specific) sets of tags instead of the canonical $\InTag_1$ and
 $\InTag_2$. This generalization has no substantial impact on the formalization,
@@
 latter process continues as $R$. These forms are often used in conjunction with
-links to model \emph{free outputs}, but can also be used to model bifurcating
+links to model \emph{delegation}, but can also be used to model bifurcating
 sessions whereby a sequential protocol continues as two parallel sub-protocols.
@@
 $\Server\x\y.P$ and $\Client\x\y.P$ processes respectively denote \emph{servers}
-and \emph{clients} acting on the $x$. Each request (from a client) spawns a copy
-of the server's body using the continuation channel $y$. The process
-$\Weaken\x.P$ denotes an explicit \emph{weakening}, that is a client that does
-\emph{not} use the server waiting on $x$. The process $\Contract\x\y\z.P$
-denotes an explicit \emph{contraction}, that is a client that uses the server
-$x$ multiple times on the channels $y$ and $z$.
+and \emph{clients} acting on the shared channel $x$. Each request (from a
+client) spawns a copy of the server's body using the continuation channel $y$.
+The process $\Weaken\x.P$ denotes an explicit \emph{weakening}, that is a client
+that \emph{does not} use the server waiting on $x$. The process
+$\Contract\x\y\z.P$ denotes an explicit \emph{contraction}, that is a client
+that uses the server $x$ multiple times on the channels $y$ and $z$.
 
@@
 
-The notions of free and bound names are almost standard. Just bear in mind that
+The notions of free and bound names are fairly standard. Just bear in mind that
 also output action prefixes bind continuation channels in (some) continuation
@@
             \RSelect &
-            \Cut\x{\Select\x{\Tag_i}\z.P}{\Case\x\z{Q_1}{Q_2}} & \red & \Cut\z{P}{Q_i} & i\in\set{1,2} \\
+            \Cut\x{\Select\x{\InTag_i}\z.P}{\Case\x\z{Q_1}{Q_2}} & \red & \Cut\z{P}{Q_i} & i\in\set{1,2} \\
             \RFork &
@@
 
+Reduction is defined by the \rulename{r-*} rules.
+%
 The reduction rules \RLink, \RClose, \RSelect, \RFork, \RConnect, \RWeaken and
@@
 session $x$ when a termination signal is exchanged on $x$. \RSelect and \RFork
-respectively model the communication of a tag $\Tag_i$ and of a fresh channel
+respectively model the communication of a tag $\InTag_i$ and of a fresh channel
 $y$. Each of these communications carries along a \emph{fresh continuation} $z$
-on which the rest of the conversation takes place, in line with the semantics of
-the linear $\pi$-calculus~\cite{KobayashiPierceTurner99}.
+on which the rest of the conversation takes place, in line with the encoding of
+sessions into the linear
+$\pi$-calculus~\cite{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
 %
@@
 
-The rule~\RCut propagates reductions through cuts and the rules \RCong closes
+The rule~\RCut propagates reductions through cuts and the rule \RCong closes
 reduction by structural pre-congruence.
@@
 \Plus B$ and $\A \With B$ describe the behavior of a process sending/receiving a
-boolean value and then adhering to either $A$ and $B$ accordingly; the
+label $\InTag_i$ and then adhering to either $A$ and $B$ accordingly; the
 multiplicative connectives $A \Ten B$ and $A \Par B$ describe the behavior of a
@@
 context $\Context = x_1 : \WhyNot A_1, \dots, x_n : \WhyNot A_n$ where all the
-types in its range are prefixed by the $\WhyNot$ modality.
+types in its range are prefixed by the $\WhyNot$ modality. We call these types
+\emph{unrestricted}, since they are used to denote shared channels that can be
+(explicitly) discarded and duplicated at will.
 
@@
 The typing rules are shown in \Cref{tab:typing-rules}. Since they are in
-one-to-one correspondence with the proof rules of classical linear logic, the
-reader may refer to the standard literature on linear logic~\cite{} or
-propositions as sessions~\citep{Wadler14} for their interpretation.
+one-to-one correspondence with -- and have exactly the same structure of -- the
+proof rules of classical linear logic, the reader may refer to the standard
+literature on linear logic~\cite{} or propositions as sessions~\citep{Wadler14}
+for their interpretation.
 
@@
 interact regardless of the context in which it occurs because the $\Close\x$
-sub-processes it contains are blocked on the channel $x$ bound by the cut.
+sub-processes it contains are blocked on the channel $x$ that is bound by the
+cut.
 %
@@
 reduce: there are irreducible processes that are not a deadlock because they
-would be able to reduce if put in a suitable context. For example, $\Wait\x.P$
+would be able to reduce if put into a suitable context. For example, $\Wait\x.P$
 does not reduce, and yet it is not a deadlock because it would be able to make
@@
 %
-We also say that $P$ is \emph{observable} if $P \pcong Q$ for some thread $Q$.
-An observable process is a process that exposes an action on a free channel name
+We say that $P$ is \emph{observable} if $P \pcong Q$ for some thread $Q$. An
+observable process is a process that exposes an action on a free channel name
 and therefore can interact through that channel, if put into some appropriate
@@
 
+Well-typed \Calculus processes are deadlock free:
+
 \begin{theorem}[deadlock freedom]
@@
 
-In order to formulate \emph{well formedness}~\cite{CarboneEtAl24}, we define
-\emph{reduction contexts} as partial processes with a single unguarded hole
-$\Hole$, thus:
+We now shift the attention to \emph{well formedness}. In the linearity
+challenge~\cite{CarboneEtAl24}, this property ensures that, whenever two
+processes composed in parallel begin with actions concerning the very same
+session, then such actions are \emph{complementary}, in the sense that they
+describe opposite forms of interaction. To define well formedness in our
+setting, we introduce \emph{reduction contexts} as partial processes with a
+single unguarded hole $\Hole$, thus:
 \[
@@
 %
-Now we observe that if $P$ and $Q$ both act on the same channel $x$ in
-complementary ways, then their parallel composition $\Cut\x{P}{Q}$ is reducible
-according to one of the principal cut reductions described in
+Now we observe that if $Q_1$ and $Q_2$ both act on the same channel $x$ in
+complementary ways, then their parallel composition $\Cut\x{Q_1}{Q_2}$ is
+reducible according to one of the principal cut reductions described in
 \Cref{tab:semantics}. Therefore, an alternative (and more general) way of
@@
 we say that $P$ is \emph{well formed} if $P \pcong \ReductionContext[Q]$ implies
-that $Q$ is alive.
+that $Q$ is alive. In the particular case when $Q = \Cut\x{Q_1}{Q_2}$ and both
+$Q_1$ and $Q_2$ start with an action on $x$, then $Q$ is not observable (because
+actions on $x$ cannot be pulled out of the cut that binds $x$) and therefore it
+must be reducible by \Cref{thm:df}.
+
+Well-typed processes are well formed:
 

### 2025-08-02 paper/conclusion.tex
revised calculus, biblio

@@
 
+working with an intrinsically typed calculus drastically reduces the number of
+cases that must be considered in proving properties (eg deadlock freedom)
+
+it also makes it trivial to reason about properties of sub-processes (see type
+safety, where Q is well typed)
+
 structural pre-congruence by far most involved, and it will grow even further in
@@
 linear pi-calculus is good
+
+we think that a simple formalization is a good basis for further investigations.
+For examples, defined cut elimination, strong normalization, soundness of
+subtyping, for the linear fragment of \Calculus

### 2025-08-02 paper/introduction.tex
drafted intro

\section{Introduction}
\label{sec:introduction}

The Concurrent Calculi Formalisation Benchmark~\citep{CarboneEtAl24} is a
collection of challenges concerning the mechanisation of formal models of
concurrent and distributed programming languages.
%
These models often make use of distinctive features that set them apart from the
formal models of \emph{sequential} programming languages, such as the adoption
of substructural (linear, affine) type systems, the dynamic scope of first-class
channels in systems of communicating processes, the extensive use of coinduction
for describing and reasoning on possibly infinite behaviors.
%
The benchmark aims at investigating effective formalization techniques that take
these features into account so as to foster the adoption of machine-checked
proofs in research work concerning concurrent and distributed programming
languages.

This paper reports on a solution to one of the challenges in the benchmark --
hereafter called \emph{linearity challenge} -- which is concerned with the
formalization of a minimal \emph{session calculus}. Sessions and session
types~\citep{Honda93,HondaVasconcelosKubo98,HuttelEtAl16} are established
formalisms for the static analysis of distributed programs based on peer-to-peer
communication channels. Every session type system revolves around three key
ideas: (1) session endpoints are \emph{linear resources} that cannot be
discarded or duplicated without jeopardizing some key safety and liveness
properties of a program; (2) the type of a session endpoint \emph{must be
updated} after each use to reflect the state of the interaction it describes;
(3) peer session endpoints are meant to be used in complementary ways so as to
guarantee the absence of communication errors and, to some extent, progress of
the interaction.

The calculus that we formalize -- called \Calculus in the following -- and the
one in the description of the challenge differ for two main aspects:
\begin{enumerate}
\item The challenge considers a minimal calculus of sessions, which is basically
    a first-order $\pi$-calculus without replication, whereas \Calculus is
    closely related to
    \CP~\citep{CairesPfenning10,Wadler14,CairesPfenningToninho16}, a
    fully-featured calculus of higher-order sessions with both linear and shared
    channels and whose type system is based on linear logic.
    %
    This allows us to target a model of sessions that is well-known in the
    literature, that is based on firm logical grounds, and that includes a
    broader range of features.
\item While the calculus in the challenge is equipped with native sessions,
    \Calculus is a \emph{linear} calculus where each channel is meant to be used
    for a \emph{single} communication. Structured conversations pertaining the
    same session and involving the exchange of several messages can still be
    modeled thanks to the use of \emph{explicit continuation channels}, in the
    same spirit of the encoding of (binary) sessions into the linear
    $\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
    %
    Overall, \Calculus has the same expressive power of \CP but is a lower-level
    calculus in which sessions can be encoded instead of being featured
    natively.
\end{enumerate}

We value these differences because their combined effect allows us to obtain a
remarkably simple formalization of \Calculus.
%
The fact of working with a calculus based on linear logic prevents \emph{by
construction} the same (sequential) process to own both endpoints of a session.
This situation is clearly undesirable since it does not correspond to a useful
pattern of interaction (every meaningful session requires its endpoints to be
used by parallel processes) and is a potential source of deadlocks.
%
From the standpoint of the formalization, where the representation of channel
names is a primary design choice, it spares us the need to \emph{distinguish}
the two endpoints of a session, \eg by means of polarities~\citep{GayHole05} or
by using different names connected by the same
binder~\citep{Vasconcelos12,GayVasconcelos25}.
%
Also, the fact of working with a calculus with \emph{linear} channels and
explicit continuations spares us the need to \emph{update} the type of channels
in typing contexts. When a channel is used it is effectively consumed, therefore
its type is \emph{removed} from the typing context and the type of the
continuation channel is \emph{added} back to the typing context. As is turns
out, removing and adding types is easier than updating them. Even more so
considering that the type of continuation channels can be conveniently added
\emph{at the beginning} of a typing context, since such channels are fresh by
definition.

The rest of the paper is structured as follows. \Cref{sec:calculus} describes
the syntax and the operational semantics of \Calculus and states the properties
of well-typed processes that we formalize and prove. \Cref{sec:agda} outlines
the essential aspects of the formalization with particular emphasis on the
representation of channel names and typing contexts. \Cref{sec:conclusion}
summarizes our contributions, discusses some extensions and future work. The
formalization has been carried out using Agda and its full code is available
from a public
repository.\footnote{\url{https://github.com/boystrange/LinearityChallenge/tree/main/exp}}

### 2025-08-02 paper/introduction.tex
++

@@
 
-This paper reports on a solution to one of the challenges in the benchmark --
-hereafter called \emph{linearity challenge} -- which is concerned with the
-formalization of a minimal \emph{session calculus}. Sessions and session
-types~\citep{Honda93,HondaVasconcelosKubo98,HuttelEtAl16} are established
-formalisms for the static analysis of distributed programs based on peer-to-peer
-communication channels. Every session type system revolves around three key
-ideas: (1) session endpoints are \emph{linear resources} that cannot be
-discarded or duplicated without jeopardizing some key safety and liveness
-properties of a program; (2) the type of a session endpoint \emph{must be
-updated} after each use to reflect the state of the interaction it describes;
-(3) peer session endpoints are meant to be used in complementary ways so as to
-guarantee the absence of communication errors and, to some extent, progress of
-the interaction.
+This paper reports on a solution to one of the challenges in the benchmark
+concerned with the formalization of a minimal \emph{session calculus}. Hereafter
+we will refer to this as the \emph{linearity challenge}.
+%
+Sessions and session types~\citep{Honda93,HondaVasconcelosKubo98,HuttelEtAl16}
+are established formalisms for the static analysis of distributed programs based
+on peer-to-peer communication channels. Every session type system revolves
+around three key ideas: (1) session endpoints are \emph{linear resources} that
+cannot be discarded or duplicated without jeopardizing some safety and liveness
+properties of a program; (2) the type of a session endpoint is \emph{updated}
+after each use to reflect the state of the interaction it describes; (3) peer
+session endpoints are meant to be used in complementary ways so as to guarantee
+the absence of communication errors and, to some extent, progress of the
+interaction.
 
-The calculus that we formalize -- called \Calculus in the following -- and the
-one in the description of the challenge differ for two main aspects:
+The calculus that we formalize, which we call \Calculus, and the one in the
+description of the challenge differ for two main aspects:
 \begin{enumerate}
@@
     %
-    This allows us to target a model of sessions that is well-known in the
-    literature, that is based on firm logical grounds, and that includes a
-    broader range of features.
+    Targeting \Calculus allows us to consider a model of sessions that is
+    well-known in the literature, that is based on firm logical grounds, and
+    that includes a broader range of features including higher-order sessions
+    and shared channels.
 \item While the calculus in the challenge is equipped with native sessions,
@@
     same session and involving the exchange of several messages can still be
-    modeled thanks to the use of \emph{explicit continuation channels}, in the
-    same spirit of the encoding of (binary) sessions into the linear
+    modeled thanks to the use of \emph{explicit continuation channels}, as in
+    the encoding of (binary) sessions into the linear
     $\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
@@
 
-We value these differences because their combined effect allows us to obtain a
-remarkably simple formalization of \Calculus.
+\Luca{Da qualche parte dire quali sono le sfide della sfida: representation of
+channel names, context splitting. Parlare di co de bruijn?}
+
+We value these aspects of \Calculus because their combined effect allows us to
+obtain a remarkably simple formalization.
 %
@@
 in typing contexts. When a channel is used it is effectively consumed, therefore
-its type is \emph{removed} from the typing context and the type of the
+its type can be \emph{removed} from the typing context and the type of the
 continuation channel is \emph{added} back to the typing context. As is turns

### 2025-08-02 paper/introduction.tex
++

--- 2025-08-02 paper/introduction.tex
@@
 formalization has been carried out using Agda and its full code is available
-from a public
+from a public\Luca{Related work? Confronto con altre formalizzazioni?}
 repository.\footnote{\url{https://github.com/boystrange/LinearityChallenge/tree/main/exp}}

### 2025-08-03 paper/introduction.tex
++

@@
 of substructural (linear, affine) type systems, the dynamic scope of first-class
-channels in systems of communicating processes, the extensive use of coinduction
-for describing and reasoning on possibly infinite behaviors.
+channels in systems of communicating processes, the need for coinductive
+definitions and proof methods for describing and reasoning on possibly infinite
+behaviors.
 %
@@
 concerned with the formalization of a minimal \emph{session calculus}. Hereafter
-we will refer to this as the \emph{linearity challenge}.
+we will refer to this challenge as the \emph{linearity challenge}.
 %
@@
 
+BISOGNA PARLARE DI RELATED WORK
+
 The rest of the paper is structured as follows. \Cref{sec:calculus} describes

### 2025-11-20 paper/introduction.tex
standalone source code

@@
 
+\begin{itemize}
+\item lavoro di Simon su formalizzazione pi calcolo con type system (anche
+    lineari?) \cite{Gay01} Il codice non c'è più online, ma si parla di 650 +
+    6500 linee di codice e prove. controllare che risultati vengono dimostrati
+\item \cite{GotoEtAl16} formalizzazione session calculus con polimorfismo, no
+    deadlock freedom, 18000 righe di formalizzazione senza contare gli esempi
+\item \cite{Thiemann19} 3000 loc Agda, linguaggio funzionale con thread ma
+    semantica complicata, c'è una sezione introduttiva con le sfide
+\item \cite{CastroFerreiraYoshida20}
+\item \cite{CicconePadovani20}
+\item \cite{RouvoetPoulsenKrebbersVisser20}
+\item Jacobs Balzer Krebbers 2022, first mechanization of deadlock freedom with
+    connectivity graphs
+\item Sano, Kavanagh, Pientka 2023, "mechanization is an art"
+\end{itemize}
+
 The rest of the paper is structured as follows. \Cref{sec:calculus} describes

### 2025-11-27 paper/Type.lagda.tex
(re)started paper

@@
 \begin{code}[hide]
+{-# OPTIONS --rewriting #-}
+open import Data.Nat
+open import Data.Fin
 open import Relation.Binary.PropositionalEquality using (_≡_; refl; cong; cong₂)
+open import Agda.Builtin.Equality.Rewrite
 \end{code}
@@
 
-The representation of types is standard, with one constructor for each of the
-forms described in \Cref{sec:calculus}.
+The representation of types is fairly standard, with one constructor for each of
+the forms described in \Cref{sec:calculus}. We use de Bruijn indices for
+denoting type variables and we make sure that types are well-scoped by indexing
+them with the number of variables that must be bound.
 
 \begin{code}
-data Type : Set where
-  𝟘 𝟙 ⊥ ⊤          : Type
-  ¡ ¿              : Type → Type
-  _&_ _⊕_ _⊗_ _⅋_  : Type → Type → Type
+data PreType : ℕ → Set where
+  𝟘 𝟙 ⊥ ⊤          : ∀{n} → PreType n
+  var rav          : ∀{n} → Fin n → PreType n
+  ¡ ¿              : ∀{n} → PreType n → PreType n
+  _&_ _⊕_ _⊗_ _⅋_  : ∀{n} → PreType n → PreType n → PreType n
+  $∀ $∃            : ∀{n} → PreType (suc n) → PreType n
 \end{code}
 
-The notion of duality is formalized as a \emph{relation} \AgdaDatatype{Dual}
-such that $\AgdaDatatype{Dual}~A~B$ holds if and only if $A = \dual{B}$.
+Note the constructors \AgdaInductiveConstructor{var} and
+\AgdaInductiveConstructor{rav}, which respectively represent type variables and
+their dual, and the quantifiers \AgdaInductiveConstructor{\$∀} and
+\AgdaInductiveConstructor{\$∃} which increase the number of bound variables in
+the scoped (pre)type.
+
+The dual of a (pre)type is computed by the following \emph{function}
+\AgdaFunction{dual}.
 
 \begin{code}
-data Dual : Type → Type → Set where
-  d-𝟘-⊤  : Dual 𝟘 ⊤
-  d-⊤-𝟘  : Dual ⊤ 𝟘
-  d-𝟙-⊥  : Dual 𝟙 ⊥
-  d-⊥-𝟙  : Dual ⊥ 𝟙
-  d-!-?  : ∀{A B} → Dual A B → Dual (¡ A) (¿ B)
-  d-?-!  : ∀{A B} → Dual A B → Dual (¿ A) (¡ B)
-  d-&-⊕  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A & B) (A′ ⊕ B′)
-  d-⊕-&  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⊕ B) (A′ & B′)
-  d-⊗-⅋  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⊗ B) (A′ ⅋ B′)
-  d-⅋-⊗  : ∀{A B A′ B′} → Dual A A′ → Dual B B′ → Dual (A ⅋ B) (A′ ⊗ B′)
+dual : ∀{n} → PreType n → PreType n
+dual 𝟘        = ⊤
+dual 𝟙        = ⊥
+dual ⊥        = 𝟙
+dual ⊤        = 𝟘
+dual (var x)  = rav x
+dual (rav x)  = var x
+dual (¡ A)    = ¿ (dual A)
+dual (¿ A)    = ¡ (dual A)
+dual (A & B)  = dual A ⊕ dual B
+dual (A ⊕ B)  = dual A & dual B
+dual (A ⊗ B)  = dual A ⅋ dual B
+dual (A ⅋ B)  = dual A ⊗ dual B
+dual ($∀ A)   = $∃ (dual A)
+dual ($∃ A)   = $∀ (dual A)
 \end{code}
 
-It is straightforward to prove that duality is a symmetric relation and that it
-behaves as an involution. From this we prove that it acts as the function
-$\dual\cdot$ defined in \Cref{sec:calculus}.
+It is straightforward to prove that duality is a an involution.
 
 \begin{code}
-dual-symm   : ∀{A B} → Dual A B → Dual B A
-dual-inv    : ∀{A B C} → Dual A B → Dual B C → A ≡ C
-dual-fun-r  : ∀{A B C} → Dual A B → Dual A C → B ≡ C
-dual-fun-l  : ∀{A B C} → Dual B A → Dual C A → B ≡ C
+dual-inv : ∀{n} {A : PreType n} → dual (dual A) ≡ A
 \end{code}
 \begin{code}[hide]
-dual-symm d-𝟘-⊤ = d-⊤-𝟘
-dual-symm d-⊤-𝟘 = d-𝟘-⊤
-dual-symm d-𝟙-⊥ = d-⊥-𝟙
-dual-symm d-⊥-𝟙 = d-𝟙-⊥
-dual-symm (d-!-? p) = d-?-! (dual-symm p)
-dual-symm (d-?-! p) = d-!-? (dual-symm p)
-dual-symm (d-&-⊕ p q) = d-⊕-& (dual-symm p) (dual-symm q)
-dual-symm (d-⊕-& p q) = d-&-⊕ (dual-symm p) (dual-symm q)
-dual-symm (d-⊗-⅋ p q) = d-⅋-⊗ (dual-symm p) (dual-symm q)
-dual-symm (d-⅋-⊗ p q) = d-⊗-⅋ (dual-symm p) (dual-symm q)
+dual-inv {_} {𝟘} = refl
+dual-inv {_} {𝟙} = refl
+dual-inv {_} {⊥} = refl
+dual-inv {_} {⊤} = refl
+dual-inv {_} {var x} = refl
+dual-inv {_} {rav x} = refl
+dual-inv {_} {¡ A} = cong ¡ dual-inv
+dual-inv {_} {¿ A} = cong ¿ dual-inv
+dual-inv {_} {A & B} = cong₂ _&_ dual-inv dual-inv
+dual-inv {_} {A ⊕ B} = cong₂ _⊕_ dual-inv dual-inv
+dual-inv {_} {A ⊗ B} = cong₂ _⊗_ dual-inv dual-inv
+dual-inv {_} {A ⅋ B} = cong₂ _⅋_ dual-inv dual-inv
+dual-inv {_} {$∀ A} = cong $∀ dual-inv
+dual-inv {_} {$∃ A} = cong $∃ dual-inv
 \end{code}
 
-\begin{code}[hide]
-dual-inv d-𝟘-⊤ d-⊤-𝟘 = refl
-dual-inv d-⊤-𝟘 d-𝟘-⊤ = refl
-dual-inv d-𝟙-⊥ d-⊥-𝟙 = refl
-dual-inv d-⊥-𝟙 d-𝟙-⊥ = refl
-dual-inv (d-!-? p) (d-?-! q) = cong ¡ (dual-inv p q)
-dual-inv (d-?-! p) (d-!-? q) = cong ¿ (dual-inv p q)
-dual-inv (d-&-⊕ p q) (d-⊕-& r s) = cong₂ _&_ (dual-inv p r) (dual-inv q s)
-dual-inv (d-⊕-& p q) (d-&-⊕ r s) = cong₂ _⊕_ (dual-inv p r) (dual-inv q s)
-dual-inv (d-⊗-⅋ p q) (d-⅋-⊗ r s) = cong₂ _⊗_ (dual-inv p r) (dual-inv q s)
-dual-inv (d-⅋-⊗ p q) (d-⊗-⅋ r s) = cong₂ _⅋_ (dual-inv p r) (dual-inv q s)
+In fact, this property is so important in the rest of the formalization that it
+is worth defining an implicit rewriting rules so that Agda can autonomously
+apply it whenever possible. This is achieved by means of the following pragma
+directive.
 
-dual-fun-r d e = dual-inv (dual-symm d) e
-
-dual-fun-l d e = dual-inv d (dual-symm e)
+\begin{code}
+{-# REWRITE dual-inv #-}
 \end{code}

### 2025-11-27 paper/Type.lagda.tex
prose for types

@@
 In fact, this property is so important in the rest of the formalization that it
-is worth defining an implicit rewriting rules so that Agda can autonomously
-apply it whenever possible. This is achieved by means of the following pragma
+is worth defining an implicit rewriting rule so that Agda can autonomously apply
+it whenever possible. This is achieved by means of the following pragma
 directive.
@@
 \end{code}
+
+\begin{code}[hide]
+ext : ∀{m n} → (Fin m → Fin n) → Fin (suc m) → Fin (suc n)
+ext ρ zero = zero
+ext ρ (suc k) = suc (ρ k)
+
+rename : ∀{m n} → (Fin m → Fin n) → PreType m → PreType n
+rename ρ 𝟘 = 𝟘
+rename ρ 𝟙 = 𝟙
+rename ρ ⊥ = ⊥
+rename ρ ⊤ = ⊤
+rename ρ (var x) = var (ρ x)
+rename ρ (rav x) = rav (ρ x)
+rename ρ (¡ A) = ¡ (rename ρ A)
+rename ρ (¿ A) = ¿ (rename ρ A)
+rename ρ (A & B) = rename ρ A & rename ρ B
+rename ρ (A ⊕ B) = rename ρ A ⊕ rename ρ B
+rename ρ (A ⊗ B) = rename ρ A ⊗ rename ρ B
+rename ρ (A ⅋ B) = rename ρ A ⅋ rename ρ B
+rename ρ ($∀ A) = $∀ (rename (ext ρ) A)
+rename ρ ($∃ A) = $∃ (rename (ext ρ) A)
+
+exts : ∀{m n} → (Fin m → PreType n) → Fin (suc m) → PreType (suc n)
+exts σ zero = var zero
+exts σ (suc k) = rename suc (σ k)
+\end{code}
+
+Next we define the function \AgdaFunction{subst} that simultaneously substitutes
+the type variables of a (pre)type with other (pre)types. Even though in practice
+we will always substitute one variable at a time, it is technically easier to
+substitutes \emph{all} variables of a (pre)type. The definition of
+\AgdaFunction{subst} relies on some auxiliary functions for \emph{renaming} type
+variables and \emph{lifting} substitutions across quantifiers. All such
+definitions are straightforward adaptations of those describe by
+\cite{KokkeWadler}.
+
+\begin{code}
+subst : ∀{m n} → (Fin m → PreType n) → PreType m → PreType n
+\end{code}
+\begin{code}[hide]
+subst σ 𝟘 = 𝟘
+subst σ 𝟙 = 𝟙
+subst σ ⊥ = ⊥
+subst σ ⊤ = ⊤
+subst σ (var x) = σ x
+subst σ (rav x) = dual (σ x)
+subst σ (¡ A) = ¡ (subst σ A)
+subst σ (¿ A) = ¿ (subst σ A)
+subst σ (A & B) = subst σ A & subst σ B
+subst σ (A ⊕ B) = subst σ A ⊕ subst σ B
+subst σ (A ⊗ B) = subst σ A ⊗ subst σ B
+subst σ (A ⅋ B) = subst σ A ⅋ subst σ B
+subst σ ($∀ A) = $∀ (subst (exts σ) A)
+subst σ ($∃ A) = $∃ (subst (exts σ) A)
+\end{code}
+
+Among all substitutions, we will use the one that substitutes the 0-indexed type
+variable with a (pre)type. It is convenient to introduce this substitution once
+and for all, which we do here.
+
+\begin{code}
+make-subst : ∀{n} → PreType n → Fin (suc n) → PreType n
+make-subst A zero = A
+make-subst A (suc k) = var k
+\end{code}
+
+Duality and substitutions are meant to commute.
+
+\begin{code}
+dual-subst : ∀{m n} {σ : Fin m → PreType n} {A : PreType m} →
+             subst σ (dual A) ≡ dual (subst σ A)
+\end{code}
+\begin{code}[hide]
+dual-subst {_} {_} {σ} {𝟘} = refl
+dual-subst {_} {_} {σ} {𝟙} = refl
+dual-subst {_} {_} {σ} {⊥} = refl
+dual-subst {_} {_} {σ} {⊤} = refl
+dual-subst {_} {_} {σ} {var x} = refl
+dual-subst {_} {_} {σ} {rav x} = refl
+dual-subst {_} {_} {σ} {¡ A} = cong ¿ (dual-subst {σ = σ} {A})
+dual-subst {_} {_} {σ} {¿ A} = cong ¡ (dual-subst {σ = σ} {A})
+dual-subst {_} {_} {σ} {A & B} = cong₂ _⊕_ (dual-subst {σ = σ} {A}) (dual-subst {σ = σ} {B})
+dual-subst {_} {_} {σ} {A ⊕ B} = cong₂ _&_ (dual-subst {σ = σ} {A}) (dual-subst {σ = σ} {B})
+dual-subst {_} {_} {σ} {A ⊗ B} = cong₂ _⅋_ (dual-subst {σ = σ} {A}) (dual-subst {σ = σ} {B})
+dual-subst {_} {_} {σ} {A ⅋ B} = cong₂ _⊗_ (dual-subst {σ = σ} {A}) (dual-subst {σ = σ} {B})
+dual-subst {_} {_} {σ} {$∀ A} = cong $∃ (dual-subst {σ = exts σ} {A})
+dual-subst {_} {_} {σ} {$∃ A} = cong $∀ (dual-subst {σ = exts σ} {A})
+\end{code}
+
+Just like \AgdaFunction{dual-inv}, \AgdaFunction{dual-subst} too is key in the
+formalization that follows. Therefore, we add it to the set of implicit
+rewriting rules used by Agda so that we do not have to think about it again.
+
+\begin{code}
+{-# REWRITE dual-subst #-}
+\end{code}
+
+At this stage we are finally ready to define the type of closed pre-types, those
+having no free type variables. We call \AgdaDatatype{Type} closed pre-types.
+Hereafter, we will seldom use pre-types again.
+
+\begin{code}
+Type : Set
+Type = PreType zero
+\end{code}

### 2025-11-28 paper/calculus.tex
snapshot

@@
 
-\subsection{Syntax}
+The calculus of pure sessions that we formalize is dubbed \Calculus and is a
+linear version of \CP~\citep{Wadler14,GayVasconcelos25} such that every session
+endpoint is meant to be used \emph{exactly once}. This differs from \CP and
+other session calculi where session endpoints can be used repeatedly, albeit
+sequentially. The key idea is that each message sent on a session endpoint is
+paired with a \emph{fresh continuation channel} on which the rest of the
+conversation takes place.
+%
+This way of encoding sessions using explicit continuations has been first
+described by \citet[extended version]{Kobayashi02b} and later studied by
+\citet{DardhaGiachinoSangiorgi17}.
+
+STRUCTURE OF SECTION
+
+\subsection{Types}
+\label{sec:types}
+
+The types for \Calculus, ranged over by $A$, $B$, \dots, are linear logic
+propositions generated by the grammar
+\[
+    A, B ::= X \mid \dual{X} \mid \Zero \mid \Top \mid \One \mid \Bot \mid A \Plus B \mid A \With B \mid A \Ten B \mid A \Par B \mid \OfCourse A \mid \WhyNot A \mid \exists X.A \mid \forall X.A
+\]
+where $X$, $Y$, \dots range over an infinite set of \emph{type variables}.
+
+The interpretation of propositions as session types is the usual
+one~\cite{Wadler14}: the constants $\Bot$ and $\One$ describe the input/output
+of a termination signal; the additive connectives $A \With B$ and $\A \Plus B$
+describe the input/output of a label $\InTag_i$ followed by the corresponding
+protocol $A$ and $B$; the multiplicative connectives $A \Par B$ and $A \Ten B$
+describe the input/output of a channel (to be used according to $A$) followed by
+the protocol $B$. The additive constants $\Zero$ and $\Top$ do not describe
+useful protocols, but they can serve the role of smallest/largest type in type
+systems that support a notion of subtyping~\citep{HornePadovani24}. The ``of
+course'' modality $\OfCourse A$ and the ``why not'' modality $\WhyNot A$
+describe servers and clients accepting and requesting connections to be used
+according to $A$. Finally, the quantifiers $\forall X.A$ and $\exists X.A$
+describe the input/output of a type $X$ followed by the protocol $A$.
+
+The notions of \emph{duality} and of \emph{type substitution} are the usual
+ones~\cite{Wadler14}. Specifically, we write $\dual{A}$ for the dual of $A$ and
+$A\subst{B}X$ for the type obtained by replacing the free occurrences of $X$ in
+$A$ with $B$.
+
+\subsection{Processes}
 \label{sec:syntax}
@@
                     & | & \Fail\x                 & \text{fail} \\
+                    & | & \Wait\x.P               & \text{wait} \\
                     & | & \Close\x                & \text{close} \\
-                    & | & \Wait\x.P               & \text{wait} \\
-                    & | & \Select\x{\InTag_i}\z.P & \text{select} \\
-                    & | & \Case\x\z{P}{Q}         & \text{branch} \\
+                    & | & \Case\x\z{P}{Q}         & \text{case} \\
+                    & | & \Select\x{\InTag_i}\z.P & \text{selection} \\
+                    & | & \Join\x\y\z.P           & \text{join} \\
                     & | & \Fork\x\y\z{P}{Q}       & \text{fork} \\
-                    & | & \Join\x\y\z.P           & \text{join} \\
-                    & | & \Server\x\y.P           & \text{accept} \\
-                    & | & \Client\x\y.P           & \text{request} \\
-                    & | & \Weaken\x.P             & \text{weaken} \\
-                    & | & \Contract\x\y\z.P       & \text{contract} \\
-                    & | & \Cut\x{P}{Q}            & \text{cut} \\
+                    & | & \Forall\x\X\y.P         & \text{type input} \\
+                    & | & \Exists\x{A}\y.P        & \text{type output} \\
+                    & | & \Server\x\y.P           & \text{server accept} \\
+                    & | & \Client\x\y.P           & \text{client request} \\
+                    & | & \Weaken\x.P             & \text{weakening} \\
+                    & | & \Contract\x\y\z.P       & \text{contraction} \\
+                    & | & \Cut[A]\x{P}{Q}         & \text{cut} \\
             \end{array}
@@
 \end{table}
-
-The calculus of pure sessions we formalized is dubbed \Calculus and is a linear
-version of \CP~\citep{Wadler14,GayVasconcelos25} such that every session
-endpoint is meant to be used \emph{exactly once}. This differs from \CP and
-other session calculi where session endpoints can be used repeatedly, albeit
-sequentially. The key idea is that each message sent on a session endpoint is
-paired with a \emph{fresh continuation channel} on which the rest of the
-conversation takes place.
-%
-This way of encoding sessions using explicit continuations has been first
-described by \citet[extended version]{Kobayashi02b} and later studied by
-\citet{DardhaGiachinoSangiorgi17}.
 
@@
 %
-The $\Fail\x$ process denotes a runtime error concerning the channel $x$.
+The process $\Fail\x$ denotes a runtime error concerning the channel $x$.
 %
-The $\Select\x{\InTag_i}\z.P$ for $i=1,2$ and $\Case\x\z{Q_1}{Q_2}$ processes
-respectively denote the output and input of a label $\InTag_i$ along with a
-fresh continuation $z$ on the channel $x$. The receiver continues as $Q_i$
-depending on the received label $\InTag_i$. In practice it is useful to consider
-arbitrary (domain-specific) sets of tags instead of the canonical $\InTag_1$ and
+The processes $\Case\x\z{Q_1}{Q_2}$ and $\Select\x{\InTag_i}\z.P$ for $i=1,2$
+describe the input/output of a label $\InTag_i$ along with a fresh continuation
+$z$ on the channel $x$. The receiver continues as $Q_i$ depending on the
+received label $\InTag_i$. In practice it is useful to consider arbitrary
+(domain-specific) sets of tags instead of the canonical $\InTag_1$ and
 $\InTag_2$. This generalization has no substantial impact on the formalization,
@@
 %
-The $\Fork\x\y\z{P}{Q}$ and $\Join\x\y\z.R$ processes respectively denote the
-output and input of a pair of fresh channels $y$ and $z$ on the channel $x$. The
-former process forks into $P$ and $Q$, each using $y$ and $z$ respectively. The
-latter process continues as $R$. These forms are often used in conjunction with
-links to model \emph{delegation}, but can also be used to model bifurcating
-sessions whereby a sequential protocol continues as two parallel sub-protocols.
+The processes $\Join\x\y\z.R$ and $\Fork\x\y\z{P}{Q}$ describe the input/output
+of a pair of fresh channels $y$ and $z$ on the channel $x$. The former process
+forks into $P$ and $Q$, each using $y$ and $z$ respectively. The latter process
+continues as $R$. These forms are often used in conjunction with links to model
+\emph{delegation}, but can also be used to model bifurcating sessions whereby a
+sequential protocol continues as two parallel sub-protocols.
+%
+The processes $\Exists\x{A}\y.P$ and $\Forall\x\X\y.P$ describe the input/output
+of a type on the channel $x$ along with a fresh continuation $y$. These forms
+are useful to describe polymorphic processes.
 
@@
 
-\emph{Cuts} of the form $\Cut\x{P}{Q}$ represent the parallel composition of the
-processes $P$ and $Q$ connected by a channel $x$.
+\emph{Cuts} of the form $\Cut[A]\x{P}{Q}$ represent the parallel composition of
+the processes $P$ and $Q$ connected by a channel $x$, which has type $A$ in $P$
+and type $\dual{A}$ in $Q$. Henceforth we write $\Cut\x{P}{Q}$ omitting the type
+annotation $A$ when it is irrelevant or clear from the context.
 
@@
             \SLink &
-            \Link\x\y & \pcong & \Link\y\x \\
+            \Link\x\y & \pcong & \Link\y\x
+            \\
             \SComm &
-            \Cut\x{P}{Q} & \pcong & \Cut\x{Q}{P} \\
+            \Cut\x{P}{Q} & \pcong & \Cut\x{Q}{P}
+            \\
             \SFail &
-            \Cut\x{\Fail\y}{P} & \pcong & \Fail\y & x \ne y \\
+            \Cut\x{\Fail\y}{P} & \pcong & \Fail\y % & x \ne y
+            \\
             \SWait &
-            \Cut\x{\Wait\y.P}{Q} & \pcong & \Wait\y.\Cut\x{P}{Q} & x \ne y \\
+            \Cut\x{\Wait\y.P}{Q} & \pcong & \Wait\y.\Cut\x{P}{Q} % & x \ne y
+            \\
+            \SCase &
+            \Cut\x{\Case\y\z{P}{Q}}{R} & \pcong & \Case\y\z{\Cut\x{P}{R}}{\Cut\x{Q}{R}} % & x \ne y,z
+            \\
             \SSelect &
-            \Cut\x{\Select\y{\InTag_i}\z.P}{Q} & \pcong & \Select\y{\InTag_i}\z.\Cut\x{P}{Q} & x \ne y,z \\
-            \SCase &
-            \Cut\x{\Case\y\z{P}{Q}}{R} & \pcong & \Case\y\z{\Cut\x{P}{R}}{\Cut\x{Q}{R}} & x \ne y,z \\
+            \Cut\x{\Select\y{\InTag_i}\z.P}{Q} & \pcong & \Select\y{\InTag_i}\z.\Cut\x{P}{Q} % & x \ne y,z
+            \\
+            \SJoin &
+            \Cut\x{\Join\y\u\v.P}{Q} & \pcong & \Join\y\u\v.\Cut\x{P}{Q} % & x \ne y,u,v
+            \\
             \SForkL &
-            \Cut\x{\Fork\y\u\v{P}{Q}}{R} & \pcong & \Fork\y\u\v{\Cut\x{P}{R}}{Q} & x \in \fn{P}\setminus\set{y,u,v} \\
+            \Cut\x{\Fork\y\u\v{P}{Q}}{R} & \pcong & \Fork\y\u\v{\Cut\x{P}{R}}{Q} & x \in \fn{P}
+            \\
             \SForkR &
-            \Cut\x{\Fork\y\u\v{P}{Q}}{R} & \pcong & \Fork\y\u\v{P}{\Cut\x{Q}{R}} & x \in \fn{Q}\setminus\set{y,u,v} \\
-            \SJoin &
-            \Cut\x{\Join\y\u\v.P}{Q} & \pcong & \Join\y\u\v.\Cut\x{P}{Q} & x \ne y,u,v \\
+            \Cut\x{\Fork\y\u\v{P}{Q}}{R} & \pcong & \Fork\y\u\v{P}{\Cut\x{Q}{R}} & x \in \fn{Q} %\setminus\set{y,u,v}
+            \\
+            \SForall &
+            \Cut\x{\Forall\y\X\z.P}{Q} & \pcong & \Forall\y\X\z.\Cut\x{P}{Q} % & x \ne y,z
+            \\
+            \SExists &
+            \Cut\x{\Exists\y{A}\z.P}{Q} & \pcong & \Exists\y{A}\z.\Cut\x{P}{Q} % & x \ne y,z
+            \\
             \SClient &
-            \Cut\x{\Client\y\z.P}{Q} & \pcong & \Client\y\z.\Cut\x{P}{Q} & x \ne y,z \\
+            \Cut\x{\Client\y\z.P}{Q} & \pcong & \Client\y\z.\Cut\x{P}{Q} % & x \ne y,z
+            \\
             \SWeaken &
-            \Cut\x{\Weaken\y.P}{Q} & \pcong & \Weaken\y.\Cut\x{P}{Q} & x \ne y \\
+            \Cut\x{\Weaken\y.P}{Q} & \pcong & \Weaken\y.\Cut\x{P}{Q} % & x \ne y
+            \\
             \SContract &
-            \Cut\x{\Contract\y\u\v.P}{Q} & \pcong & \Contract\y\u\v.\Cut\x{P}{Q} & x \ne y,u,v \\
+            \Cut\x{\Contract\y\u\v.P}{Q} & \pcong & \Contract\y\u\v.\Cut\x{P}{Q} % & x \ne y,u,v
+            \\
             \SServer &
-            \Cut\x{\Server\y\u.P}{\Server\x\v.Q} & \pcong & \Server\y\u.\Cut\x{P}{\Server\x\v.Q} & x \ne y,u \\
+            \Cut\x{\Server\y\u.P}{\Server\x\v.Q} & \pcong & \Server\y\u.\Cut\x{P}{\Server\x\v.Q} % & x \ne y,u
+            \\
             \\
@@
             \Cut\x{\Fork\x\y\z{P}{Q}}{\Join\x\y\z.R} & \red & \Cut\y{P}{\Cut\z{Q}{R}} \\
+            \RExists &
+            \Cut\x{\Exists\x{A}\z.P}{\Forall\x\X\z.Q} & \red &
+            \Cut\z{P}{Q\subst{A}\X} \\
             \RConnect &
@@
 %
-\SLink and \SComm assert that links and parallel compositions are symmetric. The
-remaining rules, when read from left to right, push a cut on $x$ underneath the
-topmost prefix on $y$ of one of its sub-processes when $x \ne y$. These rules
-are key to float input/output actions to the top-level of a process, so that
-they can interact with corresponding complementary actions in the context
+\SLink and \SComm assert that links and parallel compositions are commutative.
+The remaining rules, when read from left to right, push a cut on $x$ underneath
+the topmost prefix on $y$ of one of its sub-processes when $x \ne y$. These
+rules are key to float input/output actions to the top-level of a process, so
+that they can interact with corresponding complementary actions in the context
 surrounding the process.
 %
-Note that some rules have additional side-conditions aimed at preserving the
-meaning of names when binders are moved around. Also, there are two versions of
-\SForkL and \SForkR depending on which of the two continuations $P$ and $Q$
-contains occurrences of the restricted channel $x$.
+All these rules have implicit side-conditions aimed at preserving the meaning of
+names when binders are moved around: terms entering or exiting the scope of a
+binder for $x$ must not have free occurrences of $x$. This holds also for the
+type variable $X$ in the rule \SForall.
+%
+We content ourselves with such informal description of these critical side
+conditions given that we are going to formalize \Calculus later on.
+%
+Note also that there are two versions of \SForkL and \SForkR depending on which
+of the two continuations $P$ and $Q$ contains free occurrences of the restricted
+channel $x$.
 
@@
 welcome, although standard presentations of classical linear logic admit such
-reduction~\cite{Doumane17}.
+reduction~\cite{Wadler14,Doumane17}.
 %
@@
 %
-The reduction rules \RLink, \RClose, \RSelect, \RFork, \RConnect, \RWeaken and
-\RContract are in one-to-one correspondence with the well-known reductions of
-session calculi and the principal cut reductions of linear logic. \RLink
-eliminates a link $\Link\x\y$ by substituting $y$ for $x$. \RClose closes a
-session $x$ when a termination signal is exchanged on $x$. \RSelect and \RFork
+The reduction rules \RLink, \RClose, \RSelect, \RFork, \RExists, \RConnect,
+\RWeaken and \RContract are in one-to-one correspondence with the well-known
+reductions of session calculi and the principal cut reductions of linear logic.
+\RLink eliminates a link $\Link\x\y$ by substituting $y$ for $x$. \RClose closes
+a session $x$ when a termination signal is exchanged on $x$. \RSelect and \RFork
 respectively model the communication of a tag $\InTag_i$ and of a fresh channel
-$y$. Each of these communications carries along a \emph{fresh continuation} $z$
-on which the rest of the conversation takes place, in line with the encoding of
-sessions into the linear
+$y$. \RExists models the instantiation of a polymorphic variable $X$ as the
+communication of a type $A$. We write $Q\subst{A}\X$ for the process obtained by
+replacing every free occurrence of the type variable $X$ with $A$. Each of these
+communications consumes the channel $x$ and possibly creates and transfers a
+\emph{fresh continuation} $z$ on which the rest of the conversation takes place,
+in line with the encoding of sessions into the linear
 $\pi$-calculus~\cite{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
 %
-\RConnect, \RWeaken and \RContract describe interactions on shared channels and
-respectively represent the connection between a server and a client, the
-disposal of an unused server, and the duplication of a server. In \RWeaken and
-\RContract we use some convenient (albeit slightly ambiguous) notation for
-denoting sequences of (pairwise distinct) channels and prefixes. In particular,
-$\seqof\z$ stands for a sequence $z_1, \dots, z_n$ of channels,
+The rules \RConnect, \RWeaken and \RContract describe interactions on shared
+channels and respectively represent the connection between a server and a
+client, the disposal of an unused server, and the duplication of a server. In
+\RWeaken and \RContract we use some convenient (albeit slightly ambiguous)
+notation for denoting sequences of (pairwise distinct) channels and prefixes. In
+particular, $\seqof\z$ stands for a sequence $z_1, \dots, z_n$ of channels,
 $\Weaken{\seqof\z}.Q$ stands for $\Weaken{z_1}\dots\Weaken{z_n}.Q$ and
@@
 \label{sec:typing-rules}
-
-The types for \Calculus, ranged over by $A$, $B$, \dots, are linear logic
- propositions generated by the following grammar:
-\[
-    A, B ::= \Zero \mid \Top \mid \One \mid \Bot \mid A \Plus B \mid A \With B \mid A \Ten B \mid A \Par B \mid \OfCourse A \mid \WhyNot A
-\]
-
-The interpretation of propositions as session types is the usual
-one~\cite{Wadler14}: the constants $\Bot$ and $\One$ describe the behavior of a
-process sending/receiving a termination signal; the additive connectives $A
-\Plus B$ and $\A \With B$ describe the behavior of a process sending/receiving a
-label $\InTag_i$ and then adhering to either $A$ and $B$ accordingly; the
-multiplicative connectives $A \Ten B$ and $A \Par B$ describe the behavior of a
-process sending/receiving a channel (to be used according to $A$) and then
-continuing as described by $B$. The additive constants $\Zero$ and $\Top$ do not
-describe useful protocols, but they can serve the role of smallest/largest type
-in type systems that support a notion of subtyping~\citep{HornePadovani24}. The
-``of course'' modality $\OfCourse A$ and the ``why not'' modality $\WhyNot A$
-describe the behavior of servers and clients accepting and requesting
-connections to be used according to $A$.
-
-The notion of \emph{duality} is standard and we write $\dual{A}$ for the dual of
-$A$.
 
@@
             \and
-            \inferrule[\CloseRule]{~}{
-                \wtp{\Close\x}{x : \One}
-            }
-            \and
             \inferrule[\WaitRule]{
@@
             \and
-            \inferrule[\SelectRule]{
-                \wtp{P}{y : A_i, \Context}
-            }{
-                \wtp{\Select\x{\InTag_i}\y.P}{x : A_1 \Plus A_2, \Context}
+            \inferrule[\CloseRule]{~}{
+                \wtp{\Close\x}{x : \One}
             }
@@
             \and
+            \inferrule[\SelectRule]{
+                \wtp{P}{y : A_i, \Context}
+            }{
+                \wtp{\Select\x{\InTag_i}\y.P}{x : A_1 \Plus A_2, \Context}
+            }
+            \and
+            \inferrule[\JoinRule]{
+                \wtp{P}{y : A, z : B, \Context}
+            }{
+                \wtp{\Join\x\y\z.P}{x : A \Par B, \Context}
+            }
+            \and
             \inferrule[\ForkRule]{
@@
             \and
-            \inferrule[\JoinRule]{
-                \wtp{P}{y : A, z : B, \Context}
+            \inferrule[\ExistsRule]{
+                \wtp{P}{y : B\subst{A}\X, \Context}
             }{
-                \wtp{\Join\x\y\z.P}{x : A \Par B, \Context}
+                \wtp{\Exists\x{A}\y.P}{x : \exists\X.B, \Context}
             }
+            \and
+            \inferrule[\ForallRule]{
+                \wtp{P}{y : A, \Context}
+            }{
+                \wtp{\Forall\x\X\y.P}{x : \forall\X.A, \Context}
+            }
+            ~X\not\in\fn\Context
             \and
@@
             }{
-                \wtp{\Cut\x{P}{Q}}{\ContextC, \ContextD}
+                \wtp{\Cut[A]\x{P}{Q}}{\ContextC, \ContextD}
             }

### 2025-11-29 paper/calculus.tex
revised description of calculus

@@
 
-The types for \Calculus, ranged over by $A$, $B$, \dots, are linear logic
+The types of \Calculus, ranged over by $A$, $B$, \dots, are the linear logic
 propositions generated by the grammar
 \[
-    A, B ::= X \mid \dual{X} \mid \Zero \mid \Top \mid \One \mid \Bot \mid A \Plus B \mid A \With B \mid A \Ten B \mid A \Par B \mid \OfCourse A \mid \WhyNot A \mid \exists X.A \mid \forall X.A
+    A, B ::= X \mid \dual{X} \mid \Top \mid \Zero \mid \Bot \mid \One \mid A \With B \mid A \Plus B \mid A \Par B \mid A \Ten B \mid \forall X.A \mid \exists X.A \mid \OfCourse A \mid \WhyNot A
 \]
-where $X$, $Y$, \dots range over an infinite set of \emph{type variables}.
+where $X$, $Y$, \dots range over an infinite set of \emph{type (or proposition)
+variables}.
 
 The interpretation of propositions as session types is the usual
-one~\cite{Wadler14}: the constants $\Bot$ and $\One$ describe the input/output
-of a termination signal; the additive connectives $A \With B$ and $\A \Plus B$
-describe the input/output of a label $\InTag_i$ followed by the corresponding
-protocol $A$ and $B$; the multiplicative connectives $A \Par B$ and $A \Ten B$
-describe the input/output of a channel (to be used according to $A$) followed by
-the protocol $B$. The additive constants $\Zero$ and $\Top$ do not describe
-useful protocols, but they can serve the role of smallest/largest type in type
-systems that support a notion of subtyping~\citep{HornePadovani24}. The ``of
-course'' modality $\OfCourse A$ and the ``why not'' modality $\WhyNot A$
-describe servers and clients accepting and requesting connections to be used
-according to $A$. Finally, the quantifiers $\forall X.A$ and $\exists X.A$
-describe the input/output of a type $X$ followed by the protocol $A$.
+one~\cite{Wadler14}: the multiplicative constants $\Bot$ and $\One$ describe
+protocols that terminate with the input/output of a signal; the additive
+connectives $A \With B$ and $\A \Plus B$ describe protocols having two possible
+continuations $A$ and $B$ with are selected/offered; the multiplicative
+connectives $A \Par B$ and $A \Ten B$ describe bifurcating protocols with two
+continuations $A$ and $B$. These forms are typically used to describe the
+input/output of a channel (to be used according to $A$) and then continuing as
+$B$. The additive constants $\Zero$ and $\Top$ do not describe useful protocols,
+but they can play the role of smallest/largest type in type systems that support
+a notion of subtyping~\citep{HornePadovani24}. 
+%
+The quantifiers $\forall X.A$ and $\exists X.A$ describe the input/output of a
+type $X$ followed by the protocol $A$.
+%
+Finally, the ``of course'' modality $\OfCourse A$ and the ``why not'' modality
+$\WhyNot A$ describe servers and clients accepting and requesting connections to
+be used according to $A$.
 
-The notions of \emph{duality} and of \emph{type substitution} are the usual
-ones~\cite{Wadler14}. Specifically, we write $\dual{A}$ for the dual of $A$ and
-$A\subst{B}X$ for the type obtained by replacing the free occurrences of $X$ in
-$A$ with $B$.
+The notions of \emph{free type variables}, \emph{duality} and of \emph{type
+substitution} are standard~\cite{Wadler14}. In particular, we write $\fv{A}$ for
+the set of type variables occurring free in $A$, we write $\dual{A}$ for the
+dual of $A$ and $A\subst{B}X$ for the type obtained by replacing the free
+occurrences of $X$ in $A$ with $B$.
 
@@
                 P, Q & ::= & \Link\x\y            & \text{link} \\
-                    & | & \Fail\x                 & \text{fail} \\
+                    & | & \Fail\x                 & \text{empty choice} \\
                     & | & \Wait\x.P               & \text{wait} \\
@@
 every message sent on one of the channels is forwarded to the other. As
-discussed in the literature, this form is useful also for modeling
-\emph{delegation}, that is the exchange of an existing session endpoint in
-another session.
+discussed in the literature~\cite{LindleyMorris}, this form is also useful for
+modeling \emph{delegation}, that is the exchange of an existing session endpoint
+in another session.
 %
-The $\Close\x$ and $\Wait\x.P$ processes respectively model the output and input
-of a termination signal on the channel $x$. The former process has no
-continuation, while the latter process continues as $P$ once the signal has been
+The $\Wait\x.P$ and $\Close\x$ processes respectively model the input and output
+of a termination signal on the channel $x$. The latter process has no
+continuation, while the former process continues as $P$ once the signal has been
 received.
 %
-The process $\Fail\x$ denotes a runtime error concerning the channel $x$.
+The process $\Fail\x$ can be used to denote a failing protocol concerning the
+channel $x$.
 %
@@
 continues as $R$. These forms are often used in conjunction with links to model
-\emph{delegation}, but can also be used to model bifurcating sessions whereby a
-sequential protocol continues as two parallel sub-protocols.
+\emph{delegation}.
 %
@@
 
-The notions of free and bound names are fairly standard. Just bear in mind that
-also output action prefixes bind continuation channels in (some) continuation
+The notions of free and bound names are fairly standard, bearing in mind that
+output action prefixes bind continuation channels in (some) continuation
 processes. In particular, $\Select\x{\Tag_i}\z.P$ binds $z$ in $P$, while
@@
 welcome, although standard presentations of classical linear logic admit such
-reduction~\cite{Wadler14,Doumane17}.
+law~\cite{Wadler14,Doumane17}.
 %
@@
 
-The rule~\RCut propagates reductions through cuts and the rule \RCong closes
-reduction by structural pre-congruence.
+The rule~\RCut propagates reductions through cuts and the rule \RCong enables
+reduction up to structural pre-congruence.
 
@@
 $\ContextD$. We write $\ContextC, \ContextD$ for the union of $\ContextC$ and
-$\ContextD$ when they have disjoint domain. We write $\WhyNot\Context$ for some
+$\ContextD$ when they have disjoint domains. We write $\WhyNot\Context$ for some
 context $\Context = x_1 : \WhyNot A_1, \dots, x_n : \WhyNot A_n$ where all the
@@
 \[
-    \ReductionContext ::= \Hole \mid \Cut\x\ReductionContext{P} \mid \Cut\x{P}\ReductionContext
+    \ReductionContext ::= \Hole \mid \Cut[A]\x\ReductionContext{P} \mid \Cut[A]\x{P}\ReductionContext
 \]

### 2025-11-29 paper/Type.lagda.tex
revised types

--- 2025-11-27 paper/Type.lagda.tex
@@
 
-The representation of types is fairly standard, with one constructor for each of
-the forms described in \Cref{sec:calculus}. We use de Bruijn indices for
-denoting type variables and we make sure that types are well-scoped by indexing
-them with the number of variables that must be bound.
+The representation of types is standard, with one constructor for each of the
+forms described in \Cref{sec:calculus}. We start by defining \emph{pre-types} as
+an indexed data type $\AgdaDatatype{PreType}~n$ where $n$ is the number of bound
+type variables and we use de Bruijn indices for denoting type variables. In this
+way, we make sure that pre-types are well-scoped.
 
@@
 data PreType : ℕ → Set where
-  𝟘 𝟙 ⊥ ⊤          : ∀{n} → PreType n
+  ⊤ 𝟘 ⊥ 𝟙          : ∀{n} → PreType n
   var rav          : ∀{n} → Fin n → PreType n
-  ¡ ¿              : ∀{n} → PreType n → PreType n
-  _&_ _⊕_ _⊗_ _⅋_  : ∀{n} → PreType n → PreType n → PreType n
-  $∀ $∃            : ∀{n} → PreType (suc n) → PreType n
+  _&_ _⊕_ _⅋_ _⊗_  : ∀{n} → PreType n → PreType n → PreType n
+  `∀ `∃            : ∀{n} → PreType (suc n) → PreType n
+  `! `?            : ∀{n} → PreType n → PreType n
 \end{code}
@@
 \AgdaInductiveConstructor{rav}, which respectively represent type variables and
-their dual, and the quantifiers \AgdaInductiveConstructor{\$∀} and
-\AgdaInductiveConstructor{\$∃} which increase the number of bound variables in
-the scoped (pre)type.
+their dual, and the quantifiers \AgdaInductiveConstructor{`∀} and
+\AgdaInductiveConstructor{`∃} which increase the number of bound type variables
+in the scoped (pre)type.
 
-The dual of a (pre)type is computed by the following \emph{function}
-\AgdaFunction{dual}.
+The dual of a (pre)type is computed by the following function:
 
@@
 dual : ∀{n} → PreType n → PreType n
+dual ⊤        = 𝟘
 dual 𝟘        = ⊤
+dual ⊥        = 𝟙
 dual 𝟙        = ⊥
-dual ⊥        = 𝟙
-dual ⊤        = 𝟘
 dual (var x)  = rav x
 dual (rav x)  = var x
-dual (¡ A)    = ¿ (dual A)
-dual (¿ A)    = ¡ (dual A)
 dual (A & B)  = dual A ⊕ dual B
 dual (A ⊕ B)  = dual A & dual B
+dual (A ⅋ B)  = dual A ⊗ dual B
 dual (A ⊗ B)  = dual A ⅋ dual B
-dual (A ⅋ B)  = dual A ⊗ dual B
-dual ($∀ A)   = $∃ (dual A)
-dual ($∃ A)   = $∀ (dual A)
+dual (`∀ A)   = `∃ (dual A)
+dual (`∃ A)   = `∀ (dual A)
+dual (`! A)    = `? (dual A)
+dual (`? A)    = `! (dual A)
 \end{code}
@@
 \begin{code}[hide]
+dual-inv {_} {⊤} = refl
 dual-inv {_} {𝟘} = refl
+dual-inv {_} {⊥} = refl
 dual-inv {_} {𝟙} = refl
-dual-inv {_} {⊥} = refl
-dual-inv {_} {⊤} = refl
 dual-inv {_} {var x} = refl
 dual-inv {_} {rav x} = refl
-dual-inv {_} {¡ A} = cong ¡ dual-inv
-dual-inv {_} {¿ A} = cong ¿ dual-inv
 dual-inv {_} {A & B} = cong₂ _&_ dual-inv dual-inv
 dual-inv {_} {A ⊕ B} = cong₂ _⊕_ dual-inv dual-inv
+dual-inv {_} {A ⅋ B} = cong₂ _⅋_ dual-inv dual-inv
 dual-inv {_} {A ⊗ B} = cong₂ _⊗_ dual-inv dual-inv
-dual-inv {_} {A ⅋ B} = cong₂ _⅋_ dual-inv dual-inv
-dual-inv {_} {$∀ A} = cong $∀ dual-inv
-dual-inv {_} {$∃ A} = cong $∃ dual-inv
+dual-inv {_} {`∀ A} = cong `∀ dual-inv
+dual-inv {_} {`∃ A} = cong `∃ dual-inv
+dual-inv {_} {`! A} = cong `! dual-inv
+dual-inv {_} {`? A} = cong `? dual-inv
 \end{code}
 
-In fact, this property is so important in the rest of the formalization that it
-is worth defining an implicit rewriting rules so that Agda can autonomously
-apply it whenever possible. This is achieved by means of the following pragma
-directive.
+This property is so important in the rest of the formalization that it is worth
+defining an implicit rewriting rule so that Agda can autonomously apply it
+whenever possible. This is achieved by means of the following pragma
+directive.\footnote{The directive is effective provided that the option
+\texttt{--rewriting} is enabled, either globally when invoking Agda or in the
+module's preamble.}
 
@@
 \end{code}
+
+\begin{code}[hide]
+ext : ∀{m n} → (Fin m → Fin n) → Fin (suc m) → Fin (suc n)
+ext ρ zero = zero
+ext ρ (suc k) = suc (ρ k)
+
+rename : ∀{m n} → (Fin m → Fin n) → PreType m → PreType n
+rename ρ ⊤ = ⊤
+rename ρ 𝟘 = 𝟘
+rename ρ ⊥ = ⊥
+rename ρ 𝟙 = 𝟙
+rename ρ (var x) = var (ρ x)
+rename ρ (rav x) = rav (ρ x)
+rename ρ (A & B) = rename ρ A & rename ρ B
+rename ρ (A ⊕ B) = rename ρ A ⊕ rename ρ B
+rename ρ (A ⅋ B) = rename ρ A ⅋ rename ρ B
+rename ρ (A ⊗ B) = rename ρ A ⊗ rename ρ B
+rename ρ (`∀ A) = `∀ (rename (ext ρ) A)
+rename ρ (`∃ A) = `∃ (rename (ext ρ) A)
+rename ρ (`! A) = `! (rename ρ A)
+rename ρ (`? A) = `? (rename ρ A)
+
+exts : ∀{m n} → (Fin m → PreType n) → Fin (suc m) → PreType (suc n)
+exts σ zero = var zero
+exts σ (suc k) = rename suc (σ k)
+\end{code}
+
+Next we define the function \AgdaFunction{subst} that simultaneously substitutes
+the type variables of a (pre)type with other (pre)types. Even though in practice
+we will always substitute one variable at a time, it is technically easier to
+substitutes \emph{all} variables of a (pre)type. The definition of
+\AgdaFunction{subst} relies on some auxiliary functions for \emph{renaming} type
+variables and \emph{lifting} substitutions across quantifiers. All such
+definitions are straightforward adaptations of those described by
+\cite{KokkeWadler}.
+
+\begin{code}
+subst : ∀{m n} → (Fin m → PreType n) → PreType m → PreType n
+\end{code}
+\begin{code}[hide]
+subst σ ⊤ = ⊤
+subst σ 𝟘 = 𝟘
+subst σ ⊥ = ⊥
+subst σ 𝟙 = 𝟙
+subst σ (var x) = σ x
+subst σ (rav x) = dual (σ x)
+subst σ (A & B) = subst σ A & subst σ B
+subst σ (A ⊕ B) = subst σ A ⊕ subst σ B
+subst σ (A ⅋ B) = subst σ A ⅋ subst σ B
+subst σ (A ⊗ B) = subst σ A ⊗ subst σ B
+subst σ (`∀ A) = `∀ (subst (exts σ) A)
+subst σ (`∃ A) = `∃ (subst (exts σ) A)
+subst σ (`! A) = `! (subst σ A)
+subst σ (`? A) = `? (subst σ A)
+\end{code}
+
+Among all substitutions, we will use the one that substitutes the 0-indexed type
+variable with a (pre)type. It is convenient to introduce this substitution once
+and for all, which we do here.
+
+\begin{code}
+[_/_] : ∀{n} → PreType n → Fin (suc n) → PreType n
+[ A / zero  ]  = A
+[ A / suc k ]  = var k
+\end{code}
+
+Duality and substitutions are meant to commute.
+
+\begin{code}
+dual-subst : ∀{m n} {σ : Fin m → PreType n} {A : PreType m} →
+             subst σ (dual A) ≡ dual (subst σ A)
+\end{code}
+\begin{code}[hide]
+dual-subst {_} {_} {σ} {⊤} = refl
+dual-subst {_} {_} {σ} {𝟘} = refl
+dual-subst {_} {_} {σ} {⊥} = refl
+dual-subst {_} {_} {σ} {𝟙} = refl
+dual-subst {_} {_} {σ} {var x} = refl
+\end{code}
+
+It is worth discussing one case in the proof of \AgdaFunction{dual-subst}:
+\begin{code}
+dual-subst {_} {_} {σ} {rav x} = refl
+\end{code}
+
+Here we are supposed to prove
+$\AgdaFunction{subst}~\sigma~(\AgdaFunction{dual}~(\AgdaInductiveConstructor{rav}~x))
+  \equiv
+  \AgdaFunction{dual}~(\AgdaFunction{subst}~\sigma~(\AgdaInductiveConstructor{rav}~x))$
+  which is definitionally the same of $\sigma~x \equiv
+  \AgdaFunction{dual}~(\AgdaFunction{dual}~(\sigma~x))$.
+%
+We could easily prove this equivalence by invoking \AgdaFunction{dual-inv}, but
+thanks to the added rewriting rule a use of \AgdaInductiveConstructor{refl}
+suffices. In this case the saved effort is negligible, but in later results,
+where it is necessary to use \AgdaFunction{dual-inv} for rewriting part of the
+\emph{index} of some type families, having an implicit rewriting rule allows us
+to avoid writing some quite obscure Agda code.
+
+\begin{code}[hide]
+dual-subst {_} {_} {σ} {A & B} = cong₂ _⊕_ (dual-subst {σ = σ} {A}) (dual-subst {σ = σ} {B})
+dual-subst {_} {_} {σ} {A ⊕ B} = cong₂ _&_ (dual-subst {σ = σ} {A}) (dual-subst {σ = σ} {B})
+dual-subst {_} {_} {σ} {A ⅋ B} = cong₂ _⊗_ (dual-subst {σ = σ} {A}) (dual-subst {σ = σ} {B})
+dual-subst {_} {_} {σ} {A ⊗ B} = cong₂ _⅋_ (dual-subst {σ = σ} {A}) (dual-subst {σ = σ} {B})
+dual-subst {_} {_} {σ} {`∀ A} = cong `∃ (dual-subst {σ = exts σ} {A})
+dual-subst {_} {_} {σ} {`∃ A} = cong `∀ (dual-subst {σ = exts σ} {A})
+dual-subst {_} {_} {σ} {`! A} = cong `? (dual-subst {σ = σ} {A})
+dual-subst {_} {_} {σ} {`? A} = cong `! (dual-subst {σ = σ} {A})
+\end{code}
+
+Just like \AgdaFunction{dual-inv}, \AgdaFunction{dual-subst} too is key in the
+formalization that follows. Therefore, we add it to the set of implicit
+rewriting rules used by Agda so that we do not have to think about it again.
+
+\begin{code}
+{-# REWRITE dual-subst #-}
+\end{code}
+
+At this stage we are finally ready to define the type of closed pre-types, those
+having no free type variables. We call \AgdaDatatype{Type} closed pre-types.
+Hereafter, we will seldom use pre-types again.
+
+\begin{code}
+Type : Set
+Type = PreType zero
+\end{code}

### 2025-11-29 paper/Context.lagda.tex
included reduction
restroed contexts and permutations

@@
 \begin{code}[hide]
-open import Data.Bool using (Bool; true; false; if_then_else_)
+{-# OPTIONS --rewriting #-}
 open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
 open import Relation.Binary.PropositionalEquality using (_≡_; refl; cong; cong₂)
-open import Data.List.Base using (List; []; _∷_; [_]; _++_)
+open import Data.List.Base using (List; []; _∷_; [_]; _++_; map)
 
@@
 Typing contexts are represented using \emph{lists} of types, where the
-(polymorphic) \AgdaDatatype{List} type is defined in Agda′s standard library.
+(polymorphic) \AgdaDatatype{List} type is defined in Agda's standard library.
 
@@
 +-unit-r = +-comm +-unit-l
+
+++≃+ : ∀{Γ Δ} → Γ ++ Δ ≃ Γ + Δ
+++≃+ {[]} = +-unit-l
+++≃+ {_ ∷ _} = split-l ++≃+
 \end{code}
@@
 +-sing-l (split-l split-e) = refl , refl
+\end{code}
 
+To conclude, we define a predicate $\AgdaDatatype{Un}~\Context$ that holds
+whenever $\Context$ is \emph{unrestricted}, namely when it is made of types of
+the form $?A$. We will need this predicate in the definition of a server, which
+must obey the typing rule \ServerRule of \Cref{tab:typing-rules}.
+
+\begin{code}
 data Un : Context → Set where
   un-[]  : Un []
-  un-∷   : ∀{A Γ} → Un Γ → Un (¿ A ∷ Γ)
+  un-∷   : ∀{A Γ} → Un Γ → Un (`? A ∷ Γ)
 \end{code}

### 2025-11-29 paper/Permutations.lagda.tex
restroed contexts and permutations

@@
 \begin{code}[hide]
-open import Data.Bool using (Bool; true; false; if_then_else_)
+{-# OPTIONS --rewriting #-}
 open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
@@
 context splitting and permutations: if $\Context \simeq A, \ContextD$, then
-$\Context$ must be a permutation of $A~\AgdaInductiveConstructor{∷}~\ContextD$.
+$\Context$ must be a permutation of
+$A~\AgdaInductiveConstructor{∷}~\ContextD$.\Luca{Ci sono molte altre proprietà
+non descritte, controllare quelle che servono davvero.}
 
@@
 
-#shift : ∀{Γ A Δ} → (Γ ++ A ∷ Δ) # (A ∷ Γ ++ Δ)
-#shift {[]} = #refl
-#shift {B ∷ Γ} = #tran (#next #shift) #here
+#shift : ∀{A Γ Δ} → (Γ ++ A ∷ Δ) # (A ∷ Γ ++ Δ)
+#shift {_} {[]} = #refl
+#shift {_} {B ∷ Γ} = #tran (#next #shift) #here
 

### 2025-11-30 paper/Context.lagda.tex
new notation for splitting

@@
 data _≃_+_ : Context → Context → Context → Set where
-  split-e  : [] ≃ [] + []
-  split-l  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ A ∷ Δ + Θ
-  split-r  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ Δ + A ∷ Θ
+  •   : [] ≃ [] + []
+  <_  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ A ∷ Δ + Θ
+  >_  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ Δ + A ∷ Θ
 \end{code}
 
-As an example, below is the splitting of the context $[A, B, C, D]$ into $[A,
-D]$ and $[B, C]$. Note how the sequence of \AgdaInductiveConstructor{split-l}
-and \AgdaInductiveConstructor{split-r} applications determines where each of the
-types in $[A, B, C, D]$ ends up in $[A, D]$ and $[B, C]$.
+When splitting a context $\Context$ into $\ContextD + \ContextE$,
+for each type in $\Context$ we use one of the prefix operators \AgdaInductiveConstructor{<\_} and \AgdaInductiveConstructor{>\_} to indicate whether it is meant to be placed in $\ContextD$ or $\ContextE$,
+respectively.
+%
+As an example, below is the splitting of the context $[A, B, C, D]$ into $[B] + [A, C, D]$.
 
@@
 \begin{code}
-  splitting-example : (A ∷ B ∷ C ∷ D ∷ []) ≃ (A ∷ D ∷ []) + (B ∷ C ∷ [])
-  splitting-example = split-l (split-r (split-r (split-l split-e)))
-\end{code}
-
-It is often the case that the context $\ContextD$ in a splitting
-$\Splitting\ContextC\ContextD\ContextE$ is a singleton list $[A]$. For this
-particular case, we introduce a dedicated notation that allows us to represent
-this case in a more compact and readable way, as
-$\SimpleSplitting\ContextC{A}\ContextE$.
-
-\begin{code}
-_≃_,_ : Context → Type → Context → Set
-Γ ≃ A , Δ = Γ ≃ [ A ] + Δ
+  splitting-example₁ : (A ∷ B ∷ C ∷ D ∷ []) ≃ (B ∷ []) + (A ∷ C ∷ D ∷ [])
+  splitting-example₁ = > < > > •
 \end{code}
@@
 +-comm    : ∀{Γ Δ Θ} → Γ ≃ Δ + Θ → Γ ≃ Θ + Δ
-+-unit-l  : ∀{Γ} → Γ ≃ [] + Γ
-+-unit-r  : ∀{Γ} → Γ ≃ Γ + []
+≫         : ∀{Γ} → Γ ≃ [] + Γ
+≪         : ∀{Γ} → Γ ≃ Γ + []
 \end{code}
 \begin{code}[hide]
-+-comm split-e = split-e
-+-comm (split-l p) = split-r (+-comm p)
-+-comm (split-r p) = split-l (+-comm p)
-
-+-unit-l {[]} = split-e
-+-unit-l {_ ∷ _} = split-r +-unit-l
-
-+-unit-r = +-comm +-unit-l
++-comm • = •
++-comm (< p) = > (+-comm p)
++-comm (> p) = < (+-comm p)
 
 ++≃+ : ∀{Γ Δ} → Γ ++ Δ ≃ Γ + Δ
-++≃+ {[]} = +-unit-l
-++≃+ {_ ∷ _} = split-l ++≃+
+++≃+ {[]}    {[]}    = •
+++≃+ {[]}    {_ ∷ _} = > ++≃+
+++≃+ {_ ∷ _} {_}     = < ++≃+
+
+≫ = ++≃+ {[]}
+
+≪ = +-comm ≫
 \end{code}
@@
 \begin{code}[hide]
-+-assoc-r split-e split-e = [] , split-e , split-e
-+-assoc-r (split-l p) q with +-assoc-r p q
-... | _ , p′ , q′ = _ , split-l p′ , split-l q′
-+-assoc-r (split-r p) (split-l q) with +-assoc-r p q
-... | _ , p′ , q′ = _ , split-r p′ , split-l q′
-+-assoc-r (split-r p) (split-r q) with +-assoc-r p q
-... | _ , p′ , q′ = _ , p′ , split-r q′
++-assoc-r • • = [] , • , •
++-assoc-r (< p) q with +-assoc-r p q
+... | _ , p′ , q′ = _ , < p′ , < q′
++-assoc-r (> p) (< q) with +-assoc-r p q
+... | _ , p′ , q′ = _ , > p′ , < q′
++-assoc-r (> p) (> q) with +-assoc-r p q
+... | _ , p′ , q′ = _ , p′ , > q′
 
@@
 ... | Δ , r , p′ = Δ , +-comm r , +-comm p′
+\end{code}
 
+It is often the case that the context $\ContextD$ in a splitting
+$\Splitting\ContextC\ContextD\ContextE$ is a singleton list $[A]$. For this
+particular case, we introduce a dedicated notation that allows us to represent
+this case in a more compact and readable way, as
+$\SimpleSplitting\ContextC{A}\ContextE$.
+
+\begin{code}
+_≃_,_ : Context → Type → Context → Set
+Γ ≃ A , Δ = Γ ≃ [ A ] + Δ
+\end{code}
+\begin{code}[hide]
 +-empty-l : ∀{Γ Δ} → Γ ≃ [] + Δ → Γ ≡ Δ
-+-empty-l split-e = refl
-+-empty-l (split-r p) = cong (_ ∷_) (+-empty-l p)
++-empty-l • = refl
++-empty-l (> p) = cong (_ ∷_) (+-empty-l p)
 
 +-sing-l : ∀{A B Γ} → [ A ] ≃ B , Γ → A ≡ B × Γ ≡ []
-+-sing-l (split-l split-e) = refl , refl
++-sing-l (< •) = refl , refl
+\end{code}
+
+Note that, when specifying a splitting $\SimpleSplitting\ContextC{A}\ContextE$, it may be convenient
+to use \AgdaFunction{≫} as a shortcut for a sequence of applications of \AgdaInductiveConstructor{>\_}
+after the $A$ type has been reached. For instance, the same sample splitting shown earlier can be
+written equivalently and in a more compact way as shown below. More examples will be given in \Cref{sec:examples}.
+
+\begin{code}[hide]
+module _ where
+\end{code}
+\begin{code}
+  splitting-example₂ : (A ∷ B ∷ C ∷ D ∷ []) ≃ (B ∷ []) + (A ∷ C ∷ D ∷ [])
+  splitting-example₂ = > < ≫
 \end{code}

### 2025-11-30 paper/Permutations.lagda.tex
new notation for splitting

@@
 properties, in the following we will make use of a specific one that connects
-context splitting and permutations: if $\Context \simeq A, \ContextD$, then
-$\Context$ must be a permutation of
-$A~\AgdaInductiveConstructor{∷}~\ContextD$.\Luca{Ci sono molte altre proprietà
-non descritte, controllare quelle che servono davvero.}
-
-\begin{code}
-#cons : ∀{A Γ Δ} → Γ ≃ A , Δ → (A ∷ Δ) # Γ
-#cons (split-l p) with +-empty-l p
-... | refl = #refl
-#cons (split-r p) = #tran #here (#next (#cons p))
-\end{code}
+context splitting and permutations: if $\Splitting\ContextC\ContextD\ContextE$,
+then $\ContextC$ is a permutation of the concatenation of $\ContextD$ and
+$\ContextE$.\Luca{Ci sono molte altre proprietà non descritte, controllare
+quelle che servono davvero.}
 
@@
 #split #refl p = _ , _ , p , #refl , #refl
-#split (#next π) (split-l p) with #split π p
-... | Δ₁ , Δ₂ , q , π₁ , π₂ = _ ∷ Δ₁ , Δ₂ , split-l q , #next π₁ , π₂
-#split (#next π) (split-r p) with #split π p
-... | Δ₁ , Δ₂ , q , π₁ , π₂ = Δ₁ , _ ∷ Δ₂ , split-r q , π₁ , #next π₂
-#split #here (split-l (split-l p)) = _ , _ , split-l (split-l p) , #here , #refl
-#split #here (split-l (split-r p)) = _ , _ , split-r (split-l p) , #refl , #refl
-#split #here (split-r (split-l p)) = _ , _ , split-l (split-r p) , #refl , #refl
-#split #here (split-r (split-r p)) = _ , _ , split-r (split-r p) , #refl , #here
+#split (#next π) (< p) with #split π p
+... | Δ₁ , Δ₂ , q , π₁ , π₂ = _ ∷ Δ₁ , Δ₂ , < q , #next π₁ , π₂
+#split (#next π) (> p) with #split π p
+... | Δ₁ , Δ₂ , q , π₁ , π₂ = Δ₁ , _ ∷ Δ₂ , > q , π₁ , #next π₂
+#split #here (< < p) = _ , _ , < < p , #here , #refl
+#split #here (< > p) = _ , _ , > < p , #refl , #refl
+#split #here (> < p) = _ , _ , < > p , #refl , #refl
+#split #here (> > p) = _ , _ , > > p , #refl , #here
 #split (#tran π π′) p with #split π p
@@
 #shift {_} {B ∷ Γ} = #tran (#next #shift) #here
+\end{code}
 
-+++# : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → (Γ₁ ++ Γ₂) # Γ
-+++# split-e = #refl
-+++# (split-l p) = #next (+++# p)
-+++# (split-r p) = #tran #shift (#next (+++# p))
+\begin{code}
+#append : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → (Γ₁ ++ Γ₂) # Γ
+\end{code}
+\begin{code}[hide]
+#append • = #refl
+#append (< p) = #next (#append p)
+#append (> p) = #tran #shift (#next (#append p))
 
@@
 #un+ : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → Un Γ₁ → Un Γ₂ → Un Γ
-#un+ split-e un-[] un-[] = un-[]
-#un+ (split-l p) (un-∷ un₁) un₂ = un-∷ (#un+ p un₁ un₂)
-#un+ (split-r p) un₁ (un-∷ un₂) = un-∷ (#un+ p un₁ un₂)
+#un+ • un-[] un-[] = un-[]
+#un+ (< p) (un-∷ un₁) un₂ = un-∷ (#un+ p un₁ un₂)
+#un+ (> p) un₁ (un-∷ un₂) = un-∷ (#un+ p un₁ un₂)
 \end{code}

### 2025-11-30 paper/Process.lagda.tex
added prose for permutations
new notation for splitting

@@
 \begin{code}[hide]
+{-# OPTIONS --rewriting #-}
 open import Data.Bool using (Bool; true; false; if_then_else_)
@@
 choice increases the effort in the definition of the datatypes for processes and
-their operational semantics, but it pays off in the rest of the formalization
-for a number of reasons:
+their operational semantics, but pays off in the rest of the formalization for a
+number of reasons:
 
@@
 
-\begin{AgdaAlign}
 \begin{code}
 data Process : Context → Set where
-   link      : ∀{Γ A B} (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) → Process Γ
+  link      : ∀{A Γ} → Γ ≃ [ A ] + [ dual A ] → Process Γ
+  fail      : ∀{Γ Δ} → Γ ≃ ⊤ , Δ → Process Γ
+  wait      : ∀{Γ Δ} → Γ ≃ ⊥ , Δ → Process Δ → Process Γ
+  close     : Process [ 𝟙 ]
+  case      : ∀{A B Γ Δ} → Γ ≃ A & B , Δ →
+              Process (A ∷ Δ) → Process (B ∷ Δ) → Process Γ
+  select    : ∀{A B Γ Δ} (x : Bool) → Γ ≃ A ⊕ B , Δ →
+              Process ((if x then A else B) ∷ Δ) → Process Γ
+  join      : ∀{A B Γ Δ} → Γ ≃ A ⅋ B , Δ → Process (B ∷ A ∷ Δ) → Process Γ
+  fork      : ∀{A B Γ Δ Γ₁ Γ₂} → Γ ≃ A ⊗ B , Δ → Δ ≃ Γ₁ + Γ₂ →
+              Process (A ∷ Γ₁) → Process (B ∷ Γ₂) → Process Γ
+  all       : ∀{A Γ Δ} → Γ ≃ `∀ A , Δ →
+              ((B : Type) → Process (subst [ B /_] A ∷ Δ)) → Process Γ
+  ex        : ∀{A B Γ Δ} → Γ ≃ `∃ A , Δ → Process (subst [ B /_] A ∷ Δ) → Process Γ
+  server    : ∀{A Γ Δ} → Γ ≃ `! A , Δ → Un Δ → Process (A ∷ Δ) → Process Γ
+  client    : ∀{A Γ Δ} → Γ ≃ `? A , Δ → Process (A ∷ Δ) → Process Γ
+  weaken    : ∀{A Γ Δ} → Γ ≃ `? A , Δ → Process Δ → Process Γ
+  contract  : ∀{A Γ Δ} → Γ ≃ `? A , Δ → Process (`? A ∷ `? A ∷ Δ) → Process Γ
+  cut       : ∀{A Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ →
+              Process (A ∷ Γ₁) → Process (dual A ∷ Γ₂) → Process Γ
 \end{code}
@@
 
-\begin{code}
-   fail      : ∀{Γ Δ} (p : Γ ≃ ⊤ , Δ) → Process Γ
-\end{code}
+% \begin{code}
+%    fail      : ∀{Γ Δ} (p : Γ ≃ ⊤ , Δ) → Process Γ
+% \end{code}
 
@@
 
-\begin{code}
-   close     : Process [ 𝟙 ]
-\end{code}
+% \begin{code}
+%    close     : Process  [ 𝟙 ]
+% \end{code}
 
@@
 
-\begin{code}
-   wait      : ∀{Γ Δ} (p : Γ ≃ ⊥ , Δ) → Process Δ → Process Γ
-\end{code}
+% \begin{code}
+%    wait      : ∀{Γ Δ} (p : Γ ≃ ⊥ , Δ) → Process Δ → Process Γ
+% \end{code}
 
@@
 
-\begin{code}
-   select    : ∀{Γ Δ A B} (x : Bool) (p : Γ ≃ A ⊕ B , Δ) →
-               Process ((if x then A else B) ∷ Δ) → Process Γ
-\end{code}
+% \begin{code}
+%    select    : ∀{A B Γ Δ} (x : Bool) (p : Γ ≃ A ⊕ B , Δ) →
+%                Process ((if x then A else B) ∷ Δ) → Process Γ
+% \end{code}
 
-We represent the tags $\InTag_1$ and $\InTag_2$ with Agda′s boolean values
+We represent the tags $\InTag_1$ and $\InTag_2$ with Agda's boolean values
 \AgdaInductiveConstructor{true} and \AgdaInductiveConstructor{false}. In this
@@
 
-\begin{code}
-   case      : ∀{Γ Δ A B} (p : Γ ≃ A & B , Δ) →
-               Process (A ∷ Δ) → Process (B ∷ Δ) → Process Γ
-\end{code}
+% \begin{code}
+%    case      : ∀{A B Γ Δ} (p : Γ ≃ A & B , Δ) →
+%                Process (A ∷ Δ) → Process (B ∷ Δ) → Process Γ
+% \end{code}
 
@@
 
-\begin{code}
-   fork      : ∀{Γ Δ Γ₁ Γ₂ A B} (p : Γ ≃ A ⊗ B , Δ) (q : Δ ≃ Γ₁ + Γ₂) →
-               Process (A ∷ Γ₁) → Process (B ∷ Γ₂) → Process Γ
-\end{code}
+% \begin{code}
+%    fork      : ∀{A B Γ Δ Γ₁ Γ₂} (p : Γ ≃ A ⊗ B , Δ) (q : Δ ≃ Γ₁ + Γ₂) →
+%                Process (A ∷ Γ₁) → Process (B ∷ Γ₂) → Process Γ
+% \end{code}
 
@@
 
-\begin{code}
-   join      : ∀{Γ Δ A B} (p : Γ ≃ A ⅋ B , Δ) →
-               Process (B ∷ A ∷ Δ) → Process Γ
-\end{code}
+% \begin{code}
+%    join      : ∀{A B Γ Δ} (p : Γ ≃ A ⅋ B , Δ) → Process (B ∷ A ∷ Δ) →
+%                Process Γ
+% \end{code}
 
@@
 
-\begin{code}
-   server    : ∀{Γ Δ A} (p : Γ ≃ ¡ A , Δ) (un : Un Δ) →
-               Process (A ∷ Δ) → Process Γ
-\end{code}
+% \begin{code}
+%    server    : ∀{A Γ Δ} (p : Γ ≃ `! A , Δ) (un : Un Δ) →
+%                Process (A ∷ Δ) → Process Γ
+% \end{code}
 
@@
 
-\begin{code}
-   client    : ∀{Γ Δ A} (p : Γ ≃ ¿ A , Δ) → Process (A ∷ Δ) → Process Γ
-   weaken    : ∀{Γ Δ A} (p : Γ ≃ ¿ A , Δ) → Process Δ → Process Γ
-   contract  : ∀{Γ Δ A} (p : Γ ≃ ¿ A , Δ) → Process (¿ A ∷ ¿ A ∷ Δ) → Process Γ
-\end{code}
+% \begin{code}
+%    client    : ∀{A Γ Δ} (p : Γ ≃ `? A , Δ) → Process (A ∷ Δ) → Process Γ
+%    weaken    : ∀{A Γ Δ} (p : Γ ≃ `? A , Δ) → Process Δ → Process Γ
+%    contract  : ∀{A Γ Δ} (p : Γ ≃ `? A , Δ) → Process (`? A ∷ `? A ∷ Δ) →
+%                Process Γ
+% \end{code}
 
@@
 
-\begin{code}
-   cut       : ∀{Γ Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) →
-               Process (A ∷ Γ₁) → Process (B ∷ Γ₂) → Process Γ
-\end{code}
+% \begin{code}
+%    cut       : ∀{A Γ Γ₁ Γ₂} (p : Γ ≃ Γ₁ + Γ₂) →
+%                Process (A ∷ Γ₁) → Process (dual A ∷ Γ₂) → Process Γ
+
+%    all       : ∀{A Γ Δ} (p : Γ ≃ `∀ A , Δ) →
+%                ((B : Type) → Process (subst [ B /_] A ∷ Δ)) → Process Γ
+%    ex        : ∀{A B Γ Δ} (p : Γ ≃ `∃ A , Δ) →
+%                Process (subst [ B /_] A ∷ Δ) → Process Γ
+% \end{code}
 
@@
 
-\end{AgdaAlign}
+Permutations preserve process typeability. In a sense we can read this property
+as the fact that we can rename channels at will.
 
@@
 \begin{code}[hide]
-#process π (link d p) with #one+ π p
+#process π (link p) with #one+ π p
 ... | Δ′ , q , π′ with #singleton-inv π′
-... | refl = link d q
-#process π close with #singleton-inv π
-... | refl = close
+... | refl = link q
 #process π (fail p) with #one+ π p
@@
 ... | Δ′ , q , π′ = wait q (#process π′ P)
+#process π close with #singleton-inv π
+... | refl = close
+#process π (case p P Q) with #one+ π p
+... | Δ′ , q , π′ = case q (#process (#next π′) P) (#process (#next π′) Q)
 #process π (select x p P) with #one+ π p
 ... | Δ′ , q , π′ = select x q (#process (#next π′) P)
-#process π (case p P Q) with #one+ π p
-... | Δ′ , q , π′ = case q (#process (#next π′) P) (#process (#next π′) Q)
+#process π (join p P) with #one+ π p
+... | Δ′ , q , π′ = join q (#process (#next (#next π′)) P)
 #process π (fork p q P Q) with #one+ π p
@@
 ... | Δ₁ , Δ₂ , q′ , π₁ , π₂ = fork p′ q′ (#process (#next π₁) P) (#process (#next π₂) Q)
-#process π (join p P) with #one+ π p
-... | Δ′ , q , π′ = join q (#process (#next (#next π′)) P)
-#process π (cut d p P Q) with #split π p
-... | Δ₁ , Δ₂ , q , π₁ , π₂ = cut d q (#process (#next π₁) P) (#process (#next π₂) Q)
+#process π (all p P) with #one+ π p
+... | Δ' , q , π' = all q λ B → #process (#next π') (P B)
+#process π (ex p P) with #one+ π p
+... | Δ' , q , π' = ex q (#process (#next π') P)
 #process π (server p un P) with #one+ π p
@@
 ... | Δ′ , q , π′ = contract q (#process (#next (#next π′)) P)
+#process π (cut p P Q) with #split π p
+... | Δ₁ , Δ₂ , q , π₁ , π₂ = cut q (#process (#next π₁) P) (#process (#next π₂) Q)
 \end{code}

### 2025-11-30 paper/Context.lagda.tex
co de bruijn

@@
 
-To conclude, we define a predicate $\AgdaDatatype{Un}~\Context$ that holds
+To conclude the modeling of typing contexts, we define a predicate $\AgdaDatatype{Un}~\Context$ that holds
 whenever $\Context$ is \emph{unrestricted}, namely when it is made of types of
@@
 \end{code}
+
+It is a simple fact that the merging of two unrestricted contexts
+$\Context_1$ and $\Context_2$ results in an unrestricted context.\Luca{Serve?}
+
+\begin{code}
++-un : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → Un Γ₁ → Un Γ₂ → Un Γ
+\end{code}
+\begin{code}[hide]
++-un • un-[] un-[] = un-[]
++-un (< p) (un-∷ un₁) un₂ = un-∷ (+-un p un₁ un₂)
++-un (> p) un₁ (un-∷ un₂) = un-∷ (+-un p un₁ un₂)
+\end{code}

### 2025-11-30 paper/Permutations.lagda.tex
co de bruijn

@@
 We define typing context permutations inductively, as a binary relation
-\AgdaDatatype{\_\#\_}:
+\AgdaDatatype{\_↭\_}:
 
 \begin{code}
-data _#_ : Context → Context → Set where
-  #refl  : ∀{Γ} → Γ # Γ
-  #here  : ∀{A B Γ} → (A ∷ B ∷ Γ) # (B ∷ A ∷ Γ)
-  #next  : ∀{A Γ Δ} → Γ # Δ → (A ∷ Γ) # (A ∷ Δ)
-  #tran  : ∀{Γ Δ Θ} → Γ # Δ → Δ # Θ → Γ # Θ
+data _↭_ : Context → Context → Set where
+  refl   : ∀{Γ} → Γ ↭ Γ
+  swap   : ∀{A B Γ} → (A ∷ B ∷ Γ) ↭ (B ∷ A ∷ Γ)
+  prep   : ∀{A Γ Δ} → Γ ↭ Δ → (A ∷ Γ) ↭ (A ∷ Δ)
+  trans  : ∀{Γ Δ Θ} → Γ ↭ Δ → Δ ↭ Θ → Γ ↭ Θ
 \end{code}
@@
 Agda's standard library. Each constructor represents a particular kind of
-permutation: \AgdaInductiveConstructor{\#refl} for the trivial permutation that
-does not change anything; \AgdaInductiveConstructor{\#here} for the permutation
+permutation: \AgdaInductiveConstructor{refl} for the trivial permutation that
+does not change anything; \AgdaInductiveConstructor{swap} for the permutation
 that swaps the first two elements of a typing context;
-\AgdaInductiveConstructor{\#next} for the permutation that applies a
+\AgdaInductiveConstructor{prep} for the permutation that applies a
 sub-permutation to the tail of a typing context;
-\AgdaInductiveConstructor{\#tran} for the sequential composition of
-permutations.
+\AgdaInductiveConstructor{trans} for the sequential composition of permutations.
 %
 We have preferred defining our own notion of permutation for simplicity: the
-\AgdaInductiveConstructor{\#here} constructor does not need a sub-permutation
-for the tail of the typing context, which can always be performed, if needed,
-using \AgdaInductiveConstructor{\#next} and \AgdaInductiveConstructor{\#tran}.
-Also, and more importantly, this definition is \emph{monomorphic} (it does not
-need to operate on lists of arbitrary type), resulting in a more streamlined
-usage in the rest of the code.
+\AgdaInductiveConstructor{swap} constructor does not need a sub-permutation for
+the tail of the typing context, which can always be performed, if needed, using
+\AgdaInductiveConstructor{prep} and \AgdaInductiveConstructor{trans}. Also, and
+more importantly, this definition is \emph{monomorphic} (it does not need to
+operate on lists of arbitrary type), resulting in a more streamlined usage in
+the rest of the code.
 
 \begin{code}[hide]
-#sym : ∀{Γ Δ} → Γ # Δ → Δ # Γ
-#sym #refl = #refl
-#sym #here = #here
-#sym (#next π) = #next (#sym π)
-#sym (#tran π π′) = #tran (#sym π′) (#sym π)
+↭sym : ∀{Γ Δ} → Γ ↭ Δ → Δ ↭ Γ
+↭sym refl = refl
+↭sym swap = swap
+↭sym (prep π) = prep (↭sym π)
+↭sym (trans π π′) = trans (↭sym π′) (↭sym π)
 
-#empty-inv : ∀{Γ} → [] # Γ → Γ ≡ []
-#empty-inv #refl = refl
-#empty-inv (#tran π π′) rewrite #empty-inv π | #empty-inv π′ = refl
+↭empty-inv : ∀{Γ} → [] ↭ Γ → Γ ≡ []
+↭empty-inv refl = refl
+↭empty-inv (trans π π′) rewrite ↭empty-inv π | ↭empty-inv π′ = refl
 
-#singleton-inv : ∀{A Γ} → [ A ] # Γ → Γ ≡ [ A ]
-#singleton-inv {Γ = Γ} #refl = refl
-#singleton-inv {Γ = Γ} (#next π) rewrite #empty-inv π = refl
-#singleton-inv {Γ = Γ} (#tran π π′) rewrite #singleton-inv π | #singleton-inv π′ = refl
+↭solo-inv : ∀{A Γ} → [ A ] ↭ Γ → Γ ≡ [ A ]
+↭solo-inv refl = refl
+↭solo-inv (prep π) rewrite ↭empty-inv π = refl
+↭solo-inv (trans π π′) rewrite ↭solo-inv π | ↭solo-inv π′ = refl
 
-#rot : ∀{A B C Γ} → (A ∷ B ∷ C ∷ Γ) # (C ∷ A ∷ B ∷ Γ)
-#rot = #tran (#next #here) #here
+↭rot : ∀{A B C Γ} → (A ∷ B ∷ C ∷ Γ) ↭ (C ∷ A ∷ B ∷ Γ)
+↭rot = trans (prep swap) swap
 \end{code}
 
-Just like permutations from Agda's standard library, also \AgdaDatatype{\_\#\_}
+Just like permutations from Agda's standard library, also \AgdaDatatype{\_↭\_}
 enjoys a number of expected (and useful) properties, such as symmetry. Among all
@@
 \begin{code}[hide]
-#split : ∀{Γ Γ₁ Γ₂ Δ} → Γ # Δ → Γ ≃ Γ₁ + Γ₂ → ∃[ Δ₁ ] ∃[ Δ₂ ] (Δ ≃ Δ₁ + Δ₂ × Γ₁ # Δ₁ × Γ₂ # Δ₂)
-#split #refl p = _ , _ , p , #refl , #refl
-#split (#next π) (< p) with #split π p
-... | Δ₁ , Δ₂ , q , π₁ , π₂ = _ ∷ Δ₁ , Δ₂ , < q , #next π₁ , π₂
-#split (#next π) (> p) with #split π p
-... | Δ₁ , Δ₂ , q , π₁ , π₂ = Δ₁ , _ ∷ Δ₂ , > q , π₁ , #next π₂
-#split #here (< < p) = _ , _ , < < p , #here , #refl
-#split #here (< > p) = _ , _ , > < p , #refl , #refl
-#split #here (> < p) = _ , _ , < > p , #refl , #refl
-#split #here (> > p) = _ , _ , > > p , #refl , #here
-#split (#tran π π′) p with #split π p
-... | Θ₁ , Θ₂ , p′ , π₁ , π₂ with #split π′ p′
-... | Δ₁ , Δ₂ , q , π₁′ , π₂′ = Δ₁ , Δ₂ , q , #tran π₁ π₁′ , #tran π₂ π₂′
+↭split : ∀{Γ Γ₁ Γ₂ Δ} → Γ ↭ Δ → Γ ≃ Γ₁ + Γ₂ → ∃[ Δ₁ ] ∃[ Δ₂ ] (Δ ≃ Δ₁ + Δ₂ × Γ₁ ↭ Δ₁ × Γ₂ ↭ Δ₂)
+↭split refl p = _ , _ , p , refl , refl
+↭split (prep π) (< p) with ↭split π p
+... | Δ₁ , Δ₂ , q , π₁ , π₂ = _ ∷ Δ₁ , Δ₂ , < q , prep π₁ , π₂
+↭split (prep π) (> p) with ↭split π p
+... | Δ₁ , Δ₂ , q , π₁ , π₂ = Δ₁ , _ ∷ Δ₂ , > q , π₁ , prep π₂
+↭split swap (< < p) = _ , _ , < < p , swap , refl
+↭split swap (< > p) = _ , _ , > < p , refl , refl
+↭split swap (> < p) = _ , _ , < > p , refl , refl
+↭split swap (> > p) = _ , _ , > > p , refl , swap
+↭split (trans π π′) p with ↭split π p
+... | Θ₁ , Θ₂ , p′ , π₁ , π₂ with ↭split π′ p′
+... | Δ₁ , Δ₂ , q , π₁′ , π₂′ = Δ₁ , Δ₂ , q , trans π₁ π₁′ , trans π₂ π₂′
 
-#one+ : ∀{A Γ Γ′ Δ} → Γ # Δ → Γ ≃ A , Γ′ → ∃[ Δ′ ] (Δ ≃ A , Δ′ × Γ′ # Δ′)
-#one+ π p with #split π p
-... | _ , _ , q , π₁ , π₂ rewrite #singleton-inv π₁ = _ , q , π₂
+↭solo : ∀{A Γ Γ′ Δ} → Γ ↭ Δ → Γ ≃ A , Γ′ → ∃[ Δ′ ] (Δ ≃ A , Δ′ × Γ′ ↭ Δ′)
+↭solo π p with ↭split π p
+... | _ , _ , q , π₁ , π₂ rewrite ↭solo-inv π₁ = _ , q , π₂
 
-#shift : ∀{A Γ Δ} → (Γ ++ A ∷ Δ) # (A ∷ Γ ++ Δ)
-#shift {_} {[]} = #refl
-#shift {_} {B ∷ Γ} = #tran (#next #shift) #here
+↭shift : ∀{A Γ Δ} → (Γ ++ A ∷ Δ) ↭ (A ∷ Γ ++ Δ)
+↭shift {_} {[]} = refl
+↭shift {_} {_ ∷ _} = trans (prep ↭shift) swap
 \end{code}
@@
 \begin{code}
-#append : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → (Γ₁ ++ Γ₂) # Γ
+↭concat : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → (Γ₁ ++ Γ₂) ↭ Γ
 \end{code}
 \begin{code}[hide]
-#append • = #refl
-#append (< p) = #next (#append p)
-#append (> p) = #tran #shift (#next (#append p))
+↭concat • = refl
+↭concat (< p) = prep (↭concat p)
+↭concat (> p) = trans ↭shift (prep (↭concat p))
 
-#left : ∀{Γ Δ Θ} → Γ # Δ → (Θ ++ Γ) # (Θ ++ Δ)
-#left {Θ = []} π = π
-#left {Θ = _ ∷ Θ} π = #next (#left π)
+↭left : ∀{Γ Δ Θ} → Γ ↭ Δ → (Θ ++ Γ) ↭ (Θ ++ Δ)
+↭left {Θ = []} π = π
+↭left {Θ = _ ∷ _} π = prep (↭left π)
 
-#un : ∀{Γ Δ} → Γ # Δ → Un Γ → Un Δ
-#un #refl un = un
-#un (#next π) (un-∷ un) = un-∷ (#un π un)
-#un #here (un-∷ (un-∷ un)) = un-∷ (un-∷ un)
-#un (#tran π π′) un = #un π′ (#un π un)
-
-#un+ : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → Un Γ₁ → Un Γ₂ → Un Γ
-#un+ • un-[] un-[] = un-[]
-#un+ (< p) (un-∷ un₁) un₂ = un-∷ (#un+ p un₁ un₂)
-#un+ (> p) un₁ (un-∷ un₂) = un-∷ (#un+ p un₁ un₂)
+↭un : ∀{Γ Δ} → Γ ↭ Δ → Un Γ → Un Δ
+↭un refl un = un
+↭un (prep π) (un-∷ un) = un-∷ (↭un π un)
+↭un swap (un-∷ (un-∷ un)) = un-∷ (un-∷ un)
+↭un (trans π π′) un = ↭un π′ (↭un π un)
 \end{code}

### 2025-11-30 paper/Process.lagda.tex
co de bruijn

@@
 
-We adopt an \emph{intrinsically typed} representation of processes, so that a
-process is also a proof derivation showing that the process is well typed. This
-choice increases the effort in the definition of the datatypes for processes and
-their operational semantics, but pays off in the rest of the formalization for a
-number of reasons:
-
+We adopt a \emph{nameless, intrinsically typed} representation of processes,
+that is: a channel is identified simply by the presence of its type in the
+typing context; a process is also a proof derivation showing that the process is
+well typed. This choice increases the effort in the definition of the datatypes
+for processes and their operational semantics, but pays off in the rest of the
+formalization for a number of reasons:
 \begin{itemize}
-\item channel names are encoded in the context splitting proofs occurring within
-  processes, hence they do not need to be represented explicitly.
-\item processes and typing rules are conflated in the same datatype, thus
-  reducing the overall number of datatypes in the formalization;
-\item the typing preservation results are embedded in the definition of the
-  operational semantics of processes and do not need to be proved separately.
+\item we use context splitting not only to separate resources used by parallel
+  processes (\cf \ForkRule and \CutRule), but also to witness that a channel is
+  present in the typing context;
+\item we conflate processes and typing rules in the same datatype, thus reducing
+  the overall number of datatypes defined in the formalization;
+\item we embed the typing preservation results within the very definition of the
+  operational semantics of processes instead of proving them separately
+  (\Cref{sec:agda-semantics}).
 \end{itemize}
 
+Our nameless modeling of channels is known in the literature as the \emph{co-de
+Bruijn} representation~\cite{McBride18}. The proof that the type of a channel is
+found within a typing context essentially encodes the de Bruijn index of the
+channel in the typing context of a given process. However, the typing context
+only contains types for those channels that actually occur in the process.
+Context splitting makes sure that channels are appropriately (and above all
+linearly) distributed among sub-processes, so that each channel occurs (and is
+used) exactly once.\footnote{Strictly speaking, the typing context of a
+\AgdaInductiveConstructor{fail} process may contain channels that do not occur
+(and therefore are not used) by the process, but we consider this case as an
+exception that does not alter in a substantial way the provided intuition
+underlying the use of co-de Bruijn representation.}
+
 We now illustrate the definition of the \AgdaDatatype{Process} datatype and
-describe each constructor in detail. The datatype is indexed by a typing
-context:
+describe its key aspects in detail. The datatype is indexed by a typing context:
 
@@
 data Process : Context → Set where
-  link      : ∀{A Γ} → Γ ≃ [ A ] + [ dual A ] → Process Γ
+  link      : ∀{A Γ} → Γ ≃ A , [ dual A ] → Process Γ
   fail      : ∀{Γ Δ} → Γ ≃ ⊤ , Δ → Process Γ
@@
 \begin{code}
-#process : ∀{Γ Δ} → Γ # Δ → Process Γ → Process Δ
+↭process : ∀{Γ Δ} → Γ ↭ Δ → Process Γ → Process Δ
 \end{code}
 \begin{code}[hide]
-#process π (link p) with #one+ π p
-... | Δ′ , q , π′ with #singleton-inv π′
-... | refl = link q
-#process π (fail p) with #one+ π p
+↭process π (link p) with ↭solo π p
+... | Δ′ , q , π′ rewrite ↭solo-inv π′ = link q
+↭process π (fail p) with ↭solo π p
 ... | Δ′ , q , π′ = fail q
-#process π (wait p P) with #one+ π p
-... | Δ′ , q , π′ = wait q (#process π′ P)
-#process π close with #singleton-inv π
-... | refl = close
-#process π (case p P Q) with #one+ π p
-... | Δ′ , q , π′ = case q (#process (#next π′) P) (#process (#next π′) Q)
-#process π (select x p P) with #one+ π p
-... | Δ′ , q , π′ = select x q (#process (#next π′) P)
-#process π (join p P) with #one+ π p
-... | Δ′ , q , π′ = join q (#process (#next (#next π′)) P)
-#process π (fork p q P Q) with #one+ π p
-... | Δ′ , p′ , π′ with #split π′ q
-... | Δ₁ , Δ₂ , q′ , π₁ , π₂ = fork p′ q′ (#process (#next π₁) P) (#process (#next π₂) Q)
-#process π (all p P) with #one+ π p
-... | Δ' , q , π' = all q λ B → #process (#next π') (P B)
-#process π (ex p P) with #one+ π p
-... | Δ' , q , π' = ex q (#process (#next π') P)
-#process π (server p un P) with #one+ π p
-... | Δ′ , q , π′ = server q (#un π′ un) (#process (#next π′) P)
-#process π (client p P) with #one+ π p
-... | Δ′ , q , π′ = client q (#process (#next π′) P)
-#process π (weaken p P) with #one+ π p
-... | Δ′ , q , π′ = weaken q (#process π′ P)
-#process π (contract p P) with #one+ π p
-... | Δ′ , q , π′ = contract q (#process (#next (#next π′)) P)
-#process π (cut p P Q) with #split π p
-... | Δ₁ , Δ₂ , q , π₁ , π₂ = cut q (#process (#next π₁) P) (#process (#next π₂) Q)
+↭process π (wait p P) with ↭solo π p
+... | Δ′ , q , π′ = wait q (↭process π′ P)
+↭process π close rewrite ↭solo-inv π = close
+↭process π (case p P Q) with ↭solo π p
+... | Δ′ , q , π′ = case q (↭process (prep π′) P) (↭process (prep π′) Q)
+↭process π (select x p P) with ↭solo π p
+... | Δ′ , q , π′ = select x q (↭process (prep π′) P)
+↭process π (join p P) with ↭solo π p
+... | Δ′ , q , π′ = join q (↭process (prep (prep π′)) P)
+↭process π (fork p q P Q) with ↭solo π p
+... | Δ′ , p′ , π′ with ↭split π′ q
+... | Δ₁ , Δ₂ , q′ , π₁ , π₂ = fork p′ q′ (↭process (prep π₁) P) (↭process (prep π₂) Q)
+↭process π (all p P) with ↭solo π p
+... | Δ' , q , π' = all q λ B → ↭process (prep π') (P B)
+↭process π (ex p P) with ↭solo π p
+... | Δ' , q , π' = ex q (↭process (prep π') P)
+↭process π (server p un P) with ↭solo π p
+... | Δ′ , q , π′ = server q (↭un π′ un) (↭process (prep π′) P)
+↭process π (client p P) with ↭solo π p
+... | Δ′ , q , π′ = client q (↭process (prep π′) P)
+↭process π (weaken p P) with ↭solo π p
+... | Δ′ , q , π′ = weaken q (↭process π′ P)
+↭process π (contract p P) with ↭solo π p
+... | Δ′ , q , π′ = contract q (↭process (prep (prep π′)) P)
+↭process π (cut p P Q) with ↭split π p
+... | Δ₁ , Δ₂ , q , π₁ , π₂ = cut q (↭process (prep π₁) P) (↭process (prep π₂) Q)
 \end{code}

### 2025-11-30 paper/Permutations.lagda.tex
++

@@
 that swaps the first two elements of a typing context;
-\AgdaInductiveConstructor{prep} for the permutation that applies a
-sub-permutation to the tail of a typing context;
-\AgdaInductiveConstructor{trans} for the sequential composition of permutations.
+\AgdaInductiveConstructor{prep} for the permutation that applies a permutation
+to the tail of a typing context; \AgdaInductiveConstructor{trans} for the
+sequential composition of permutations.
 %

### 2025-12-02 paper/Permutations.lagda.tex
cleanup

@@
 
-This definition is essentially the same as the notion of list permutation in
-Agda's standard library. Each constructor represents a particular kind of
+Each constructor of \AgdaDatatype{\_↭\_} represents a particular kind of
 permutation: \AgdaInductiveConstructor{refl} for the trivial permutation that
@@
 that swaps the first two elements of a typing context;
-\AgdaInductiveConstructor{prep} for the permutation that applies a permutation
-to the tail of a typing context; \AgdaInductiveConstructor{trans} for the
-sequential composition of permutations.
-%
-We have preferred defining our own notion of permutation for simplicity: the
+\AgdaInductiveConstructor{prep} for the permutation applied to the tail of a
+typing context; \AgdaInductiveConstructor{trans} for the sequential composition
+of permutations.
+
+The definition of \AgdaDatatype{\_↭\_} data type is nearly the same found in
+Agda's standard library. We have preferred defining our own notion of
+permutation for improved simplicity and convenience: the
 \AgdaInductiveConstructor{swap} constructor does not need a sub-permutation for
@@
 \AgdaInductiveConstructor{prep} and \AgdaInductiveConstructor{trans}. Also, and
-more importantly, this definition is \emph{monomorphic} (it does not need to
-operate on lists of arbitrary type), resulting in a more streamlined usage in
-the rest of the code.
+more importantly, the data type is \emph{monomorphic} (it does not need to
+relate lists of arbitrary type) and the quantified arguments $A$ and $B$ of
+\AgdaInductiveConstructor{swap} and \AgdaInductiveConstructor{prep} are
+implicit, which streamlines their usage in the rest of the code.
 
@@
 Just like permutations from Agda's standard library, also \AgdaDatatype{\_↭\_}
-enjoys a number of expected (and useful) properties, such as symmetry. Among all
-properties, in the following we will make use of a specific one that connects
-context splitting and permutations: if $\Splitting\ContextC\ContextD\ContextE$,
-then $\ContextC$ is a permutation of the concatenation of $\ContextD$ and
-$\ContextE$.\Luca{Ci sono molte altre proprietà non descritte, controllare
-quelle che servono davvero.}
+enjoys a number of expected properties, such as symmetry. Among them, in the
+following we will make use of a specific one that connects context splitting and
+permutations: if $\Splitting\ContextC\ContextD\ContextE$, then $\ContextC$ is a
+permutation of the concatenation of $\ContextD$ and $\ContextE$.\Luca{Ci sono
+molte altre proprietà non descritte, controllare quelle che servono davvero.}
 

### 2025-12-02 paper/Context.lagda.tex
restored description of Process

@@
 \begin{code}[hide]
-infix 4 _≃_+_ _≃_,_
+infix 4 _≃_+_ _∋_⊳_
 \end{code}
@@
 It is often the case that the context $\ContextD$ in a splitting
-$\Splitting\ContextC\ContextD\ContextE$ is a singleton list $[A]$. For this
-particular case, we introduce a dedicated notation that allows us to represent
-this case in a more compact and readable way, as
+$\Splitting\ContextC\ContextD\ContextE$ is a singleton list $[A]$. We use this
+particular splitting to describe the presence of a channel of type $A$ in
+$\ContextC$ and giving a name $\ContextE$ to the residual context. For this
+particular case, we introduce a dedicated notation that allows us to write this
+case in a more compact and meaningful way, as
 $\SimpleSplitting\ContextC{A}\ContextE$.
@@
 \begin{code}
-_≃_,_ : Context → Type → Context → Set
-Γ ≃ A , Δ = Γ ≃ [ A ] + Δ
+_∋_⊳_ : Context → Type → Context → Set
+Γ ∋ A ⊳ Δ = Γ ≃ [ A ] + Δ
 \end{code}
@@
 +-empty-l (> p) = cong (_ ∷_) (+-empty-l p)
-
-+-sing-l : ∀{A B Γ} → [ A ] ≃ B , Γ → A ≡ B × Γ ≡ []
-+-sing-l (< •) = refl , refl
 \end{code}
 
-Note that, when specifying a splitting $\SimpleSplitting\ContextC{A}\ContextE$, it may be convenient
-to use \AgdaFunction{≫} as a shortcut for a sequence of applications of \AgdaInductiveConstructor{>\_}
-after the $A$ type has been reached. For instance, the same sample splitting shown earlier can be
-written equivalently and in a more compact way as shown below. More examples will be given in \Cref{sec:examples}.
+Note that, when specifying a splitting $\SimpleSplitting\ContextC{A}\ContextE$,
+it may be convenient to use \AgdaFunction{≫} as a shortcut for a sequence of
+applications of \AgdaInductiveConstructor{>\_} after the $A$ type has been
+reached. For instance, the same sample splitting shown earlier can be written
+equivalently and in a more compact way as shown below. More examples will be
+given in \Cref{sec:examples}.
 
@@
 \begin{code}
-  splitting-example₂ : (A ∷ B ∷ C ∷ D ∷ []) ≃ (B ∷ []) + (A ∷ C ∷ D ∷ [])
+  splitting-example₂ : (A ∷ B ∷ C ∷ D ∷ []) ∋ B ⊳ (A ∷ C ∷ D ∷ [])
   splitting-example₂ = > < ≫

### 2025-12-02 paper/Process.lagda.tex
restored description of Process

@@
 
-CO-DE BRUIJN INDICES
-
-We adopt a \emph{nameless, intrinsically typed} representation of processes,
-that is: a channel is identified simply by the presence of its type in the
-typing context; a process is also a proof derivation showing that the process is
+We adopt a \emph{nameless, intrinsically-typed} representation of processes,
+whereby a channel is identified by the position of its type in the typing
+context and a process is also a proof derivation showing that the process is
 well typed. This choice increases the effort in the definition of the datatypes
-for processes and their operational semantics, but pays off in the rest of the
-formalization for a number of reasons:
+for defining processes and their operational semantics, but pays off in the rest
+of the formalization for a number of reasons:
 \begin{itemize}
@@
 
-Our nameless modeling of channels is known in the literature as the \emph{co-de
-Bruijn} representation~\cite{McBride18}. The proof that the type of a channel is
-found within a typing context essentially encodes the de Bruijn index of the
-channel in the typing context of a given process. However, the typing context
-only contains types for those channels that actually occur in the process.
-Context splitting makes sure that channels are appropriately (and above all
-linearly) distributed among sub-processes, so that each channel occurs (and is
-used) exactly once.\footnote{Strictly speaking, the typing context of a
-\AgdaInductiveConstructor{fail} process may contain channels that do not occur
-(and therefore are not used) by the process, but we consider this case as an
-exception that does not alter in a substantial way the provided intuition
-underlying the use of co-de Bruijn representation.}
+Our nameless representation of channels is known in the literature as the
+\emph{co-de Bruijn} representation of identifiers~\cite{McBride18}. The proof
+that the type of a channel is found within a typing context essentially encodes
+the de Bruijn index of the channel in the typing context of a given process.
+However, the typing context only contains types for those channels that actually
+occur in the process. Context splitting makes sure that channels are
+appropriately (and above all linearly) distributed among sub-processes, so that
+each channel occurs (and is used) exactly once.\footnote{Strictly speaking, the
+typing context of a \AgdaInductiveConstructor{fail} process may contain channels
+that do not occur (and therefore are not used) by the process, but we consider
+this case as an exception that does not alter in a substantial way the provided
+intuition underlying the use of co-de Bruijn representation.}
 
-We now illustrate the definition of the \AgdaDatatype{Process} datatype and
-describe its key aspects in detail. The datatype is indexed by a typing context:
+We now illustrate the definition of the \AgdaDatatype{Process} datatype and then
+describe its key aspects in detail. The datatype is indexed by a typing context
+$\Context$ according to which the process is well typed:
 
@@
 data Process : Context → Set where
-  link      : ∀{A Γ} → Γ ≃ A , [ dual A ] → Process Γ
-  fail      : ∀{Γ Δ} → Γ ≃ ⊤ , Δ → Process Γ
-  wait      : ∀{Γ Δ} → Γ ≃ ⊥ , Δ → Process Δ → Process Γ
+  link      : ∀{A Γ} → Γ ≃ [ A ] + [ dual A ] → Process Γ
+  fail      : ∀{Γ Δ} → Γ ∋ ⊤ ⊳ Δ → Process Γ
+  wait      : ∀{Γ Δ} → Γ ∋ ⊥ ⊳ Δ → Process Δ → Process Γ
   close     : Process [ 𝟙 ]
-  case      : ∀{A B Γ Δ} → Γ ≃ A & B , Δ →
+  case      : ∀{A B Γ Δ} → Γ ∋ A & B ⊳ Δ →
               Process (A ∷ Δ) → Process (B ∷ Δ) → Process Γ
-  select    : ∀{A B Γ Δ} (x : Bool) → Γ ≃ A ⊕ B , Δ →
-              Process ((if x then A else B) ∷ Δ) → Process Γ
-  join      : ∀{A B Γ Δ} → Γ ≃ A ⅋ B , Δ → Process (B ∷ A ∷ Δ) → Process Γ
-  fork      : ∀{A B Γ Δ Γ₁ Γ₂} → Γ ≃ A ⊗ B , Δ → Δ ≃ Γ₁ + Γ₂ →
+  left      : ∀{A B Γ Δ} → Γ ∋ A ⊕ B ⊳ Δ → Process (A ∷ Δ) → Process Γ
+  right     : ∀{A B Γ Δ} → Γ ∋ A ⊕ B ⊳ Δ → Process (B ∷ Δ) → Process Γ
+  join      : ∀{A B Γ Δ} → Γ ∋ A ⅋ B ⊳ Δ → Process (B ∷ A ∷ Δ) → Process Γ
+  fork      : ∀{A B Γ Δ Γ₁ Γ₂} → Γ ∋ A ⊗ B ⊳ Δ → Δ ≃ Γ₁ + Γ₂ →
               Process (A ∷ Γ₁) → Process (B ∷ Γ₂) → Process Γ
-  all       : ∀{A Γ Δ} → Γ ≃ `∀ A , Δ →
-              ((B : Type) → Process (subst [ B /_] A ∷ Δ)) → Process Γ
-  ex        : ∀{A B Γ Δ} → Γ ≃ `∃ A , Δ → Process (subst [ B /_] A ∷ Δ) → Process Γ
-  server    : ∀{A Γ Δ} → Γ ≃ `! A , Δ → Un Δ → Process (A ∷ Δ) → Process Γ
-  client    : ∀{A Γ Δ} → Γ ≃ `? A , Δ → Process (A ∷ Δ) → Process Γ
-  weaken    : ∀{A Γ Δ} → Γ ≃ `? A , Δ → Process Δ → Process Γ
-  contract  : ∀{A Γ Δ} → Γ ≃ `? A , Δ → Process (`? A ∷ `? A ∷ Δ) → Process Γ
+  all       : ∀{A Γ Δ} → Γ ∋ `∀ A ⊳ Δ →
+              ((X : Type) → Process (subst [ X /_] A ∷ Δ)) → Process Γ
+  ex        : ∀{A B Γ Δ} → Γ ∋ `∃ A ⊳ Δ → Process (subst [ B /_] A ∷ Δ) → Process Γ
+  server    : ∀{A Γ Δ} → Γ ∋ `! A ⊳ Δ → Un Δ → Process (A ∷ Δ) → Process Γ
+  client    : ∀{A Γ Δ} → Γ ∋ `? A ⊳ Δ → Process (A ∷ Δ) → Process Γ
+  weaken    : ∀{A Γ Δ} → Γ ∋ `? A ⊳ Δ → Process Δ → Process Γ
+  contract  : ∀{A Γ Δ} → Γ ∋ `? A ⊳ Δ → Process (`? A ∷ `? A ∷ Δ) → Process Γ
   cut       : ∀{A Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ →
@@
 
-A \AgdaInductiveConstructor{link} process is well typed in a typing context with
-exactly two types $A$ and $B$ which must be related by duality. We use a proof
-$p$ that $\Context \simeq [A] + [B]$ instead of requiring $\Context$ to be
-simply $[A,B]$ so that it becomes straightforward to define \SLink using
-\AgdaFunction{dual-symm} and \AgdaFunction{+-comm}.
+Let us start from the constructors representing input/output actions, namely
+those from \AgdaInductiveConstructor{fail} through
+\AgdaInductiveConstructor{contract}. They all basically follow the same pattern:
+their first (explicit) argument is a splitting of the form
+$\SimpleSplitting\ContextC{C}\ContextD$ that identifies the channel on which the
+process is acting within the typing context $\ContextC$ and where the shape of
+$C$ is determined by the typing rule; the subsequent arguments are terms
+corresponding to the continuations of the process being represented, each
+indexed by its own context. Crucially, the type of the continuation channels are
+\emph{prepended} to the context that indexes the continuation processes using
+the \AgdaInductiveConstructor{\_∷\_} list constructor. This is consistent with
+the fact that the continuation channels have been freshly introduced, and the
+channel on which the process operates has been consumed.
+%
+There are just a few deviations from this pattern, which we discuss next.
 
-% \begin{code}
-%    fail      : ∀{Γ Δ} (p : Γ ≃ ⊤ , Δ) → Process Γ
-% \end{code}
+The constructor \AgdaInductiveConstructor{close} lacks a splitting argument:
+since a process $\Close\x$ is well typed in a singleton context $x : \One$, it
+is easier to specify such context directly as the index of
+\AgdaDatatype{Process} rather than through a splitting.
 
-A \AgdaInductiveConstructor{fail} process simply requires the presence of $\Top$
-in the typing context.
+The constructor \AgdaInductiveConstructor{fork}, which represents a process
+$\Fork\x\y\z{P}{Q}$, has an additional splitting argument
+$\Splitting\ContextD{\Context_1}{\Context_2}$ which distributes the resources in
+$\ContextD$ among the two continuation processes $P$ and $Q$.
 
-% \begin{code}
-%    close     : Process  [ 𝟙 ]
-% \end{code}
+The constructor \AgdaInductiveConstructor{all}, which represents a process
+$\Forall\x\X\y.P$, models the continuation $P$ using higher-order abstract
+syntax (HOAS~\cite{}): the continuation $P$ is produced by a function applied to
+a \AgdaInductiveConstructor{Type} $X$. Notice how the constructor enforces the
+side condition of \ForallRule \emph{by definition}:
+%
+$A$ has type
+$\AgdaDatatype{PreType}~(\AgdaInductiveConstructor{suc}~\AgdaInductiveConstructor{zero})$,
+that is a pre-type with \emph{at most one} free type variable (that is $X$),
+whereas $\ContextD$ is a list of $\AgdaDatatype{Type} =
+\AgdaDatatype{PreType}~\AgdaInductiveConstructor{zero}$, that is a list of
+pre-types without free type variables. Therefore, $X$ cannot occur in
+$\ContextD$ as required by \ForallRule.
 
-A \AgdaInductiveConstructor{close} process is well typed in the singleton
-context $[\One]$.
+Finally, the constructor \AgdaInductiveConstructor{server}, which represents a
+process $\Server\x\y.P$, has an additional argument of type
+$\AgdaDatatype{Un}~\ContextD$ enforcing the property that $\ContextD$ is
+unrestricted.
 
-% \begin{code}
-%    wait      : ∀{Γ Δ} (p : Γ ≃ ⊥ , Δ) → Process Δ → Process Γ
-% \end{code}
-
-A \AgdaInductiveConstructor{wait} process is well typed in any context
-$\ContextC$ that contains $\Bot$ and whose residual $\ContextD$ allows the
-continuation to be well typed.
-
-% \begin{code}
-%    select    : ∀{A B Γ Δ} (x : Bool) (p : Γ ≃ A ⊕ B , Δ) →
-%                Process ((if x then A else B) ∷ Δ) → Process Γ
-% \end{code}
-
-We represent the tags $\InTag_1$ and $\InTag_2$ with Agda's boolean values
-\AgdaInductiveConstructor{true} and \AgdaInductiveConstructor{false}. In this
-way we can use a single constructor \AgdaInductiveConstructor{select} for the
-selections $\Select\x{\InTag_i}.P$. According to the semantics of \Calculus, the
-channel on which the selection is performed is consumed and the continuation
-process uses a fresh continuation channel. As a consequence, the type of the
-continuation channel, which is either $A$ or $B$ depending on the value of the
-tag $x$, is \emph{prepended} in front of the typing context used to type the
-continuation process.
-
-% \begin{code}
-%    case      : ∀{A B Γ Δ} (p : Γ ≃ A & B , Δ) →
-%                Process (A ∷ Δ) → Process (B ∷ Δ) → Process Γ
-% \end{code}
-
-A \AgdaInductiveConstructor{case} process has two possible continuations. Note
-again that the continuation channel, which is received along with the tag, has
-either type $A$ or type $B$ depending on the tag and its type is prepended to
-the typing context of the continuation process since it is the newest channel.
-
-% \begin{code}
-%    fork      : ∀{A B Γ Δ Γ₁ Γ₂} (p : Γ ≃ A ⊗ B , Δ) (q : Δ ≃ Γ₁ + Γ₂) →
-%                Process (A ∷ Γ₁) → Process (B ∷ Γ₂) → Process Γ
-% \end{code}
-
-A \AgdaInductiveConstructor{fork} process creates two fresh continuation
-channels and spaws two continuation processes. Since context splitting is a
-ternary relation, we need two splitting proofs $p$ and $q$ to isolate the type
-$A \Ten B$ of channel on which the pair of continuation channels is sent and to
-distribute the remaining channel across the two continuation processes.
-
-% \begin{code}
-%    join      : ∀{A B Γ Δ} (p : Γ ≃ A ⅋ B , Δ) → Process (B ∷ A ∷ Δ) →
-%                Process Γ
-% \end{code}
-
-The \AgdaInductiveConstructor{join} process has a single continuation which uses
-the pair of continuation channels received from a channel of type $A \Par B$.
-Note that $B$ is prepended in front of $A$, somewhat implying that the second
-continuation is bound \emph{after} the first even though, from a technical
-standpoint, both continuations are bound simultaneously. We could have prepended
-the types $A$ and $B$ also in the opposite order, provided that the reduction
-rule \RFork (formalized in \Cref{sec:reduction-agda}) is suitably adjusted.
-
-% \begin{code}
-%    server    : ∀{A Γ Δ} (p : Γ ≃ `! A , Δ) (un : Un Δ) →
-%                Process (A ∷ Δ) → Process Γ
-% \end{code}
-
-A \AgdaInductiveConstructor{server} constructor introduces a channel of type
-$\OfCourse A$ given a continuation process that consumes a channel of type $A$
-and a proof $\mathit{un}$ that every other channel in $\ContextD$ is
-unrestricted.\Luca{Introdurre terminologia e dire che unrestricted vuol dire che
-il canale può essere ``duplicato'' e scartato.}
-
-% \begin{code}
-%    client    : ∀{A Γ Δ} (p : Γ ≃ `? A , Δ) → Process (A ∷ Δ) → Process Γ
-%    weaken    : ∀{A Γ Δ} (p : Γ ≃ `? A , Δ) → Process Δ → Process Γ
-%    contract  : ∀{A Γ Δ} (p : Γ ≃ `? A , Δ) → Process (`? A ∷ `? A ∷ Δ) →
-%                Process Γ
-% \end{code}
-
-The constructors \AgdaInductiveConstructor{client},
-\AgdaInductiveConstructor{weaken} and \AgdaInductiveConstructor{contract}
-introduce a channel of type $\WhyNot A$ and follow the same pattern of the
-previous process forms. Note that, in the case of
-\AgdaInductiveConstructor{contract}, the contraction of $\WhyNot A$ applies to
-the first two channels in the typing context of the continuation. This is
-consistent with the fact that \Calculus uses an explicit prefix
-$\Contract\x\y\z$ that models contraction and binds $y$ and $z$ in the
-continuation process. In a pen-and-paper presentation, like the one by
-Wadler~\citep{Wadler14}, it is arguably more convenient to have weaking and
-contraction typing rules without the corresponding process forms.
-
-% \begin{code}
-%    cut       : ∀{A Γ Γ₁ Γ₂} (p : Γ ≃ Γ₁ + Γ₂) →
-%                Process (A ∷ Γ₁) → Process (dual A ∷ Γ₂) → Process Γ
-
-%    all       : ∀{A Γ Δ} (p : Γ ≃ `∀ A , Δ) →
-%                ((B : Type) → Process (subst [ B /_] A ∷ Δ)) → Process Γ
-%    ex        : ∀{A B Γ Δ} (p : Γ ≃ `∃ A , Δ) →
-%                Process (subst [ B /_] A ∷ Δ) → Process Γ
-% \end{code}
-
-A \AgdaInductiveConstructor{cut} process incorporates a duality proof for the
-peer endpoints of the channel being restricted and a splitting proof that
-distributes the free channels among the two parallel processes.
+The two remaining constructors, \AgdaInductiveConstructor{link} and
+\AgdaInductiveConstructor{cut}, deal with links and cuts in the expected way. A
+link is well typed in a context
+$\Splitting\Context{[A]}{[\AgdaFunction{dual}~A]}$. We use a splitting instead
+of requiring $\Context$ to be $[A,\AgdaFunction{dual}~A]$ so that it becomes
+straightforward to define \SLink using \AgdaFunction{+-comm}.
+%
+The constructor \AgdaInductiveConstructor{cut} models a process $\Cut\x{P}{Q}$
+and uses splitting to distribute the resources in $\Context$ among $P$ and $Q$.
 
@@
 ... | Δ′ , q , π′ = case q (↭process (prep π′) P) (↭process (prep π′) Q)
-↭process π (select x p P) with ↭solo π p
-... | Δ′ , q , π′ = select x q (↭process (prep π′) P)
+↭process π (left p P) with ↭solo π p
+... | Δ′ , q , π′ = left q (↭process (prep π′) P)
+↭process π (right p P) with ↭solo π p
+... | Δ′ , q , π′ = right q (↭process (prep π′) P)
 ↭process π (join p P) with ↭solo π p

### 2025-12-02 paper/calculus.tex
restored description of Process

@@
                 P, Q & ::= & \Link\x\y            & \text{link} \\
-                    & | & \Fail\x                 & \text{empty choice} \\
+                    & | & \Fail\x                 & \text{fail} \\
                     & | & \Wait\x.P               & \text{wait} \\
@@
                     & | & \Case\x\z{P}{Q}         & \text{case} \\
-                    & | & \Select\x{\InTag_i}\z.P & \text{selection} \\
+                    & | & \Select\x{\InTag_i}\z.P & \text{left/right selection} \\
                     & | & \Join\x\y\z.P           & \text{join} \\
                     & | & \Fork\x\y\z{P}{Q}       & \text{fork} \\
-                    & | & \Forall\x\X\y.P         & \text{type input} \\
-                    & | & \Exists\x{A}\y.P        & \text{type output} \\
-                    & | & \Server\x\y.P           & \text{server accept} \\
-                    & | & \Client\x\y.P           & \text{client request} \\
+                    & | & \Forall\x\X\y.P         & \text{for all} \\
+                    & | & \Exists\x{A}\y.P        & \text{exists} \\
+                    & | & \Server\x\y.P           & \text{server} \\
+                    & | & \Client\x\y.P           & \text{client} \\
                     & | & \Weaken\x.P             & \text{weakening} \\

### 2025-12-02 paper/Congruence.lagda.tex
included reduction
restored congruence

--- 2025-07-26 paper/Semantics.lagda.tex
@@
 \begin{code}[hide]
-open import Data.Bool using (Bool; true; false; if_then_else_)
+{-# OPTIONS --rewriting #-}
 open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
-open import Relation.Binary.PropositionalEquality using (_≡_; refl; cong; cong₂)
 open import Data.List.Base using (List; []; _∷_; [_]; _++_)
@@
 
-\subsection{Operational Semantics}
-\label{sec:semantics-agda}
+\subsection{Structural Pre-Congruence}
+\label{sec:congruence-agda}
 
@@
 \begin{code}
-data _⊒_ : ∀{Γ} → Process Γ → Process Γ → Set where
+data _⊒_ {Γ} : Process Γ → Process Γ → Set where
 \end{code}
@@
   s-comm :
-    ∀{Γ Γ₁ Γ₂ A B P Q} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) →
-    cut d p P Q ⊒ cut (dual-symm d) (+-comm p) Q P
+    ∀{A Γ₁ Γ₂ P Q} (p : Γ ≃ Γ₁ + Γ₂) →
+    cut {A} p P Q ⊒ cut (+-comm p) Q P
 \end{code}
@@
 
+implicit rewriting
+
 \begin{code}[hide]
   s-link :
-    ∀{Γ A B}
-    (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) →
-    link d p ⊒ link (dual-symm d) (+-comm p)
-
+    ∀{A} (p : Γ ≃ [ A ] + [ dual A ]) → link p ⊒ link (+-comm p)
   s-fail :
-    ∀{Γ Γ₁ Γ₂ Δ A B P} (d : Dual A B)
-    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ⊤ , Δ) →
+    ∀{A Γ₁ Γ₂ Δ P} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ ⊤ ⊳ Δ) →
     let _ , _ , q′ = +-assoc-l p q in
-    cut d p (fail (split-r q)) P ⊒ fail q′
+    cut {A} p (fail (> q)) P ⊒ fail q′
 \end{code}
@@
   s-wait :
-    ∀{Γ Γ₁ Γ₂ Δ A B} {P : Process (A ∷ Δ)} {Q : Process (B ∷ Γ₂)}
-    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ⊥ , Δ) →
+    ∀{Γ₁ Γ₂ Δ A P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ ⊥ ⊳ Δ) →
     let _ , p′ , q′ = +-assoc-l p q in
-    cut d p (wait (split-r q) P) Q ⊒ wait q′ (cut d p′ P Q)
+    cut {A} p (wait (> q) P) Q ⊒ wait q′ (cut p′ P Q)
 \end{code}
@@
 
+\begin{code}[hide]
+  s-case :
+    ∀{A B C Γ₁ Γ₂ Δ P Q R} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B & C ⊳ Δ) →
+    let _ , p′ , q′ = +-assoc-l p q in
+    cut {A} p (case (> q) P Q) R ⊒ case q′ (cut (< p′) (↭process swap P) R)
+                                           (cut (< p′) (↭process swap Q) R)
+\end{code}
 \begin{code}
-  s-select-l :
-    ∀{Γ Γ₁ Γ₂ Δ A B C D} {P : Process (C ∷ A ∷ Δ)} {Q : Process (B ∷ Γ₂)}
-    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ C ⊕ D , Δ) →
+  s-left :
+    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⊕ C ⊳ Δ) →
     let _ , p′ , q′ = +-assoc-l p q in
-    cut d p (select true (split-r q) P) Q ⊒
-    select true q′ (cut d (split-l p′) (#process #here P) Q)
+    cut {A} p (left (> q) P) Q ⊒ left q′ (cut (< p′) (↭process swap P) Q)
+\end{code}
+\begin{code}[hide]
+  s-right :
+    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⊕ C ⊳ Δ) →
+    let _ , p′ , q′ = +-assoc-l p q in
+    cut {A} p (right (> q) P) Q ⊒ right q′ (cut (< p′) (↭process swap P) Q)
 \end{code}
@@
 
-\begin{code}[hide]
-  s-select-r :
-    ∀{Γ Γ₁ Γ₂ Δ A B C D}
-    {P : Process (D ∷ A ∷ Δ)}
-    {Q : Process (B ∷ Γ₂)}
-    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ C ⊕ D , Δ) →
-    let _ , p′ , q′ = +-assoc-l p q in
-    cut d p (select false (split-r q) P) Q ⊒
-    select false q′ (cut d (split-l p′) (#process #here P) Q)
-
-  s-case :
-    ∀{Γ A B A₁ A₂ Γ₁ Γ₂ Δ}
-    {P : Process (A₁ ∷ A ∷ Δ)}
-    {Q : Process (A₂ ∷ A ∷ Δ)}
-    {R : Process (B ∷ Γ₂)}
-    (d : Dual A B)
-    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ A₁ & A₂ , Δ) →
-    let _ , p′ , q′ = +-assoc-l p q in
-    cut d p (case (split-r q) P Q) R ⊒
-    case q′ (cut d (split-l p′) (#process #here P) R)
-            (cut d (split-l p′) (#process #here Q) R)
-\end{code}
-
 \emph{En passant}, we also show the modeling of the \SForkL rule, which is
@@
 
+\begin{code}[hide]
+  s-join :
+    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⅋ C ⊳ Δ) →
+    let _ , p′ , q′ = +-assoc-l p q in
+    cut {A} p (join (> q) P) Q ⊒
+    join q′ (cut (< < p′) (↭process (↭shift {A} {C ∷ B ∷ []}) P) Q)
+\end{code}
 \begin{code}
   s-fork-l :
-    ∀{Γ Γ₁ Γ₂ Δ Δ₁ Δ₂ A B C D}
-    {P : Process (C ∷ A ∷ Δ₁)} {Q : Process (D ∷ Δ₂)} {R : Process (B ∷ Γ₂)}
-    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ C ⊗ D , Δ) (r : Δ ≃ Δ₁ + Δ₂) →
+    ∀{Γ₁ Γ₂ Δ Δ₁ Δ₂ A B C P Q R}
+    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⊗ C ⊳ Δ) (r : Δ ≃ Δ₁ + Δ₂) →
     let _ , p′ , q′ = +-assoc-l p q in
@@
     let _ , q′′ , r′′ = +-assoc-r r′ (+-comm p′′) in
-    cut d p (fork (split-r q) (split-l r) P Q) R ⊒
-    fork q′ r′′ (cut d (split-l q′′) (#process #here P) R) Q
+    cut {A} p (fork (> q) (< r) P Q) R ⊒
+    fork q′ r′′ (cut (< q′′) (↭process swap P) R) Q
 \end{code}
@@
 Recall from \Cref{tab:semantics} that we want to apply this rule on a process of
-the form $\Cut\x{\Fork\y\u\v{P}{Q}}{R}$ when $x \in \fn{P}$. We capture this
-condition by means of the splitting $(\AgdaInductiveConstructor{split-l}~r)$,
-meaning that the type $A$ of $x$ ends up in the typing context for $P$ and not
-in the one for $Q$.
+the form $\Cut[A]\x{\Fork\y\u\v{P}{Q}}{R}$ when $x \in \fn{P}$. We capture this
+condition by means of the splitting $(\AgdaInductiveConstructor{<}~r)$, meaning
+that the type $A$ of $x$ ends up in the typing context for $P$ and not in the
+one for $Q$.
 %
 The symmetric rule \SForkR is modeled by another constructor
-\AgdaInductiveConstructor{s-fork-r}, which is not shown here but is similar to
+\AgdaInductiveConstructor{s-fork-r}, not shown here but similar to
 \AgdaInductiveConstructor{s-fork-l} except that
-$(\AgdaInductiveConstructor{split-l}~r)$ is replaced by
-$(\AgdaInductiveConstructor{split-r}~r)$.
+$(\AgdaInductiveConstructor{<}~r)$ is replaced by
+$(\AgdaInductiveConstructor{>}~r)$.
 
@@
   s-fork-r :
-    ∀{Γ Γ₁ Γ₂ Δ Δ₁ Δ₂ A B C D}
-    {P : Process (C ∷ Δ₁)}
-    {Q : Process (D ∷ A ∷ Δ₂)}
-    {R : Process (B ∷ Γ₂)}
-    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ C ⊗ D , Δ)
-    (r : Δ ≃ Δ₁ + Δ₂) →
+    ∀{Γ₁ Γ₂ Δ Δ₁ Δ₂ A B C P Q R}
+    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⊗ C ⊳ Δ) (r : Δ ≃ Δ₁ + Δ₂) →
     let _ , p′ , q′ = +-assoc-l p q in
     let _ , p′′ , r′ = +-assoc-l p′ r in
-    cut d p (fork (split-r q) (split-r r) P Q) R ⊒
-    fork q′ r′ P (cut d (split-l p′′) (#process #here Q) R)
-
-  s-join :
-    ∀{Γ Γ₁ Γ₂ Δ A B C D}
-    {P : Process (D ∷ C ∷ A ∷ Δ)}
-    {Q : Process (B ∷ Γ₂)}
-    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ C ⅋ D , Δ) →
-    let _ , p′ , q′ = +-assoc-l p q in
-    cut d p (join (split-r q) P) Q ⊒
-    join q′ (cut d (split-l (split-l p′)) (#process #rot P) Q)
-
+    cut {A} p (fork (> q) (> r) P Q) R ⊒
+    fork q′ r′ P (cut (< p′′) (↭process swap Q) R)
+  s-all :
+    ∀{A B Γ₁ Γ₂ Δ F Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `∀ B ⊳ Δ) ->
+    let _ , p' , q' = +-assoc-l p q in
+    cut {A} p (all (> q) F) Q ⊒ all q' λ σ → cut (< p') (↭process swap (F σ)) Q
+  s-ex :
+    ∀{A B C Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `∃ B ⊳ Δ) ->
+    let _ , p' , q' = +-assoc-l p q in
+    cut {A} p (ex {_} {C} (> q) P) Q ⊒ ex q' (cut (< p') (↭process swap P) Q)
   s-server :
-    ∀{Γ A B C Γ₁ Γ₂ Δ₁}
-    {P : Process (C ∷ ¿ A ∷ Δ₁)}
-    {Q : Process (B ∷ Γ₂)}
-    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ¡ C , Δ₁) (r : Γ₂ ≃ [] + Γ₂)
+    ∀{A B Γ₁ Γ₂ Δ₁ P Q}
+    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `! B ⊳ Δ₁) (r : Γ₂ ≃ [] + Γ₂)
     (un₁ : Un Δ₁) (un₂ : Un Γ₂) →
     let _ , p′ , q′ = +-assoc-l p q in
-    cut (d-?-! d) p (server (split-r q) (un-∷ un₁) P) (server (split-l r) un₂ Q) ⊒
-    server q′ (#un+ p′ un₁ un₂) (cut (d-?-! d) (split-l p′) (#process #here P) (server (split-l r) un₂ Q))
-
+    cut {`? A} p (server (> q) (un-∷ un₁) P) (server (< r) un₂ Q) ⊒
+    server q′ (+-un p′ un₁ un₂) (cut (< p′) (↭process swap P) (server (< r) un₂ Q))
   s-client :
-    ∀{Γ A B C Γ₁ Γ₂ Δ}
-    {P : Process (C ∷ A ∷ Δ)}
-    {Q : Process (B ∷ Γ₂)}
-    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ¿ C , Δ) →
+    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `? B ⊳ Δ) →
     let _ , p′ , q′ = +-assoc-l p q in
-    cut d p (client (split-r q) P) Q ⊒
-    client q′ (cut d (split-l p′) (#process #here P) Q)
-
+    cut {A} p (client (> q) P) Q ⊒ client q′ (cut (< p′) (↭process swap P) Q)
   s-weaken :
-    ∀{Γ A B C Γ₁ Γ₂ Δ}
-    {P : Process (A ∷ Δ)}
-    {Q : Process (B ∷ Γ₂)}
-    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ¿ C , Δ) →
+    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `? B ⊳ Δ) →
     let _ , p′ , q′ = +-assoc-l p q in
-    cut d p (weaken (split-r q) P) Q ⊒
-    weaken q′ (cut d p′ P Q)
-
+    cut {A} p (weaken (> q) P) Q ⊒ weaken q′ (cut p′ P Q)
   s-contract :
-    ∀{Γ A B C Γ₁ Γ₂ Δ}
-    {P : Process (¿ C ∷ ¿ C ∷ A ∷ Δ)}
-    {Q : Process (B ∷ Γ₂)}
-    (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ ¿ C , Δ) →
+    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `? B ⊳ Δ) →
     let _ , p′ , q′ = +-assoc-l p q in
-    cut d p (contract (split-r q) P) Q ⊒
-    contract q′ (cut d (split-l (split-l p′)) (#process #rot P) Q)
+    cut {A} p (contract (> q) P) Q ⊒
+    contract q′ (cut (< < p′) (↭process (↭shift {A} {`? B ∷ `? B ∷ []}) P) Q)
 \end{code}
@@
 \begin{code}
-  s-refl  : ∀{Γ} {P : Process Γ} → P ⊒ P
-  s-tran  : ∀{Γ} {P Q R : Process Γ} → P ⊒ Q → Q ⊒ R → P ⊒ R
-  s-cong  : ∀{Γ Γ₁ Γ₂ A A′} {P Q : Process (A ∷ Γ₁)} {P′ Q′ : Process (A′ ∷ Γ₂)}
-            (d : Dual A A′) (p : Γ ≃ Γ₁ + Γ₂) → P ⊒ Q → P′ ⊒ Q′ →
-            cut d p P P′ ⊒ cut d p Q Q′
+  s-refl  : ∀{P} → P ⊒ P
+  s-tran  : ∀{P Q R} → P ⊒ Q → Q ⊒ R → P ⊒ R
+  s-cong  : ∀{Γ₁ Γ₂ A P Q P′ Q′} (p : Γ ≃ Γ₁ + Γ₂) →
+            P ⊒ Q → P′ ⊒ Q′ → cut {A} p P P′ ⊒ cut p Q Q′
 \end{code}

### 2025-12-02 paper/Congruence.lagda.tex
revised congruence

@@
 The constructor \AgdaInductiveConstructor{s-comm} models the commutativity
-property of parallel composition. We use \AgdaFunction{dual-symm} and
-\AgdaFunction{+-comm} to compute the proofs of the relation
-$\AgdaDatatype{Dual}~B~A$ and of the splitting $\ContextC \simeq \Context_2 +
-\Context_1$ from $d$ and $p$ respectively.
-
-implicit rewriting
+property of parallel composition. We use \AgdaFunction{+-comm} to compute the
+proof of the splitting $\Splitting\ContextC{\Context_2}{\Context_1}$ from $p$
+respectively.
+%
+Notice that this constructor also makes key use of the implicit rewriting rule
+\AgdaFunction{dual-inv} described in \Cref{sec:type-agda}. Indeed $P$ and $Q$
+have type $\AgdaDatatype{Process}~(A~\AgdaInductiveConstructor{∷}~\Context_1)$
+and
+$\AgdaDatatype{Process}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\Context_2)$,
+respectively, but the \AgdaInductiveConstructor{cut} on the r.h.s. of
+\AgdaInductiveConstructor{⊒} expects $P$ to have type
+$\AgdaFunction{dual}~(\AgdaFunction{dual}~A)$. Thanks to
+\AgdaFunction{dual-inv}, Agda considers these types equivalent without requiring
+intricate substitutions in the index of a data type.
 
@@
 There are two non-trivial aspects worth commenting. The first one concerns the
-proof $(\AgdaInductiveConstructor{split-r}~q)$ used in the
+proof $(\AgdaInductiveConstructor{>}~q)$ used in the
 \AgdaInductiveConstructor{wait} process before structural precongruence is
@@
   \item
-    $(\AgdaInductiveConstructor{wait}~(\AgdaInductiveConstructor{split-r}~q)~P)$
-    is a direct sub-process of the \AgdaInductiveConstructor{cut}, and therefore
-    it is meant to be well typed in the context
+    $(\AgdaInductiveConstructor{wait}~(\AgdaInductiveConstructor{>}~q)~P)$ is a
+    direct sub-process of the \AgdaInductiveConstructor{cut}, and therefore it
+    is meant to be well typed in the context
     $A~\AgdaInductiveConstructor{∷}~\Context_1$.
@@
     contain a $\Bot$ type as per the typing rule \WaitRule. That is
-    $A~\AgdaInductiveConstructor{∷}~\Context_1 \simeq [\Bot] +
-    A~\AgdaInductiveConstructor{∷}~\Delta$ for some $\Delta$.
+    $\Splitting{A~\AgdaInductiveConstructor{∷}~\Context_1}{[\Bot]}{A~\AgdaInductiveConstructor{∷}~\ContextD}$
+    for some $\ContextD$.
   \item The \SWait rule is applicable only provided that the channel restricted
@@
     A~\AgdaInductiveConstructor{∷}~\Delta$ through the use of
-    \AgdaInductiveConstructor{split-r}.
+    \AgdaInductiveConstructor{>}.
 \end{enumerate}
@@
 
-The constructors \AgdaInductiveConstructor{s-select-l} and
-\AgdaInductiveConstructor{s-select-r} model \SSelect when the selected tag is
-respectively \AgdaInductiveConstructor{true} and
-\AgdaInductiveConstructor{false}.
+The constructors \AgdaInductiveConstructor{s-left} and
+\AgdaInductiveConstructor{s-right} model \SSelect when the selected tag is
+respectively $\Tag_1$ and $\Tag_2$.
 
@@
 Here the process
-$(\AgdaInductiveConstructor{select}~\AgdaInductiveConstructor{true}~(\AgdaInductiveConstructor{split-r}~q)~P)$,
-which is found under a cut for $x : A$, is using some channel $y : C \Plus D$ to
-send the tag $\AgdaInductiveConstructor{true}$.
+$(\AgdaInductiveConstructor{left}~(\AgdaInductiveConstructor{>}~q)~P)$, which is
+found under a cut for $x : A$, is using some channel $y : B \Plus C$ to send the
+tag $\Tag_1$.
 %
@@
 on, the continuation process $P$ will be using a fresh continuation channel $z :
-C$. Therefore, $P$ is required to be well typed in the context
-$C~\AgdaInductiveConstructor{∷}~A~\AgdaInductiveConstructor{∷}~\ContextD$, where
+B$. Therefore, $P$ is required to be well typed in the context
+$B~\AgdaInductiveConstructor{∷}~A~\AgdaInductiveConstructor{∷}~\ContextD$, where
 the type of the continuation sits on top of the type of the restricted channel.
 %
-After structural precongruence is applied, however, the type $C$ of the
+After structural precongruence is applied, however, the type $B$ of the
 continuation channel $z$ ends up underneath that of the restricted channel $x$,
@@
 in the context
-$A~\AgdaInductiveConstructor{∷}~C~\AgdaInductiveConstructor{∷}~\ContextD$. This
-is achieved applying the function \AgdaFunction{\#process} to the
-\AgdaInductiveConstructor{\#here} permutation, which swaps the two topmost types
+$A~\AgdaInductiveConstructor{∷}~B~\AgdaInductiveConstructor{∷}~\ContextD$. This
+is achieved applying the function \AgdaFunction{↭process} to the
+\AgdaInductiveConstructor{swap} permutation, which swaps the two topmost types
 in a typing context, and finally to the process $P$.
 
-\emph{En passant}, we also show the modeling of the \SForkL rule, which is
-interesting because of its complex side conditions:
+We also show the modeling of the \SForkL rule, which is interesting because of
+its complex side conditions:
 
@@
 Finally, in \Cref{sec:semantics} we have colloquially defined $\pcong$ as a
-``\emph{pre-congruence}'', implicitly implying that it is reflexive, transitive,
+``\emph{pre-congruence}'', implying that it is reflexive, transitive relation
 and that it is is preserved by some forms of calculus. In the formalization we

### 2025-12-02 paper/Reduction.lagda.tex
restored reductions

@@
 \begin{code}[hide]
+{-# OPTIONS --rewriting #-}
 open import Data.Bool using (Bool; true; false; if_then_else_)
-open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
+open import Data.Product using (_×_; _,_; ∃; Σ; Σ-syntax; ∃-syntax)
 open import Relation.Binary.PropositionalEquality using (_≡_; refl; sym; cong; cong₂)
@@
 weakening : ∀{Γ Γ₁ Γ₂} (un : Un Γ₁) → Γ ≃ Γ₁ + Γ₂ → Process Γ₂ → Process Γ
-weakening un p P = #process (+++# p) (aux un P)
+weakening un p P = ↭process (↭concat p) (aux un P)
   where
@@
     aux un-[] P = P
-    aux (un-∷ un) P = weaken (split-l +-unit-l) (aux un P)
+    aux (un-∷ un) P = weaken (< ≫) (aux un P)
 
 contraction : ∀{Γ Γ₁ Γ₂} (un : Un Γ₁) → Γ ≃ Γ₁ + Γ₂ → Process (Γ₁ ++ Γ) → Process Γ
-contraction un p P = #process (+++# p) (aux un (#process (#left (#sym (+++# p))) P))
+contraction un p P = ↭process (↭concat p) (aux un (↭process (↭left (↭sym (↭concat p))) P))
   where
@@
     aux un-[] P = P
-    aux {¿ A ∷ Γ₁} {Γ₂} (un-∷ un) P with contract (split-l +-unit-l) (#process (#shift {¿ A ∷ Γ₁} {¿ A} {Γ₁ ++ Γ₂}) P)
-    ... | P₁ rewrite sym (++-assoc (¿ A ∷ Γ₁) Γ₁ Γ₂) with #process (#sym (#shift {Γ₁ ++ Γ₁} {¿ A})) P₁
-    ... | P₂ rewrite ++-assoc Γ₁ Γ₁ (¿ A ∷ Γ₂) with aux un P₂
-    ... | P₃ = #process #shift P₃
+    aux {`? A ∷ Γ₁} {Γ₂} (un-∷ un) P with contract (< ≫) (↭process (↭shift {`? A} {`? A ∷ Γ₁} {Γ₁ ++ Γ₂}) P)
+    ... | P₁ rewrite sym (++-assoc (`? A ∷ Γ₁) Γ₁ Γ₂) with ↭process (↭sym (↭shift {`? A} {Γ₁ ++ Γ₁})) P₁
+    ... | P₂ rewrite ++-assoc Γ₁ Γ₁ (`? A ∷ Γ₂) with aux un P₂
+    ... | P₃ = ↭process ↭shift P₃
 \end{code}
@@
 \begin{code}
-  r-link      : ∀{Δ A B} {P : Process (B ∷ Δ)}
-                (d : Dual A B) (e : Dual A B) (p : Γ ≃ B , Δ) →
-                cut d p (link e (split-l (split-r split-e))) P ↝ #process (#cons p) P
+  r-link      : ∀{Δ A P} (p : Γ ∋ dual A ⊳ Δ) →
+                cut {A} p (link (< > •)) P ↝ ↭process (↭concat p) P
 \end{code}
@@
 \begin{code}
-  r-close     : ∀{P : Process Γ} (p₀ : Γ ≃ [] + Γ) (q₀ : Γ ≃ [] + Γ) →
-                cut d-𝟙-⊥ p₀ close (wait (split-l q₀) P) ↝ P
+  r-close     : ∀{P} (p₀ : Γ ≃ [] + Γ) (q₀ : Γ ≃ [] + Γ) →
+                cut p₀ close (wait (< q₀) P) ↝ P
 \end{code}
@@
 \begin{code}
-  r-select-l  : ∀{Γ₁ Γ₂ A A′ B B′}
-                {P : Process (A ∷ Γ₁)} {Q : Process (A′ ∷ Γ₂)} {R : Process (B′ ∷ Γ₂)}
-                (d : Dual A A′) (e : Dual B B′)
+  r-left      : ∀{Γ₁ Γ₂ A B P Q R}
                 (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
-                cut (d-⊕-& d e) p
-                    (select true (split-l p₀) P)
-                    (case (split-l q₀) Q R) ↝ cut d p P Q
+                cut {A ⊕ B} p (left (< p₀) P)
+                              (case (< q₀) Q R) ↝ cut p P Q
 \end{code}
@@
 \begin{code}[hide]
-  r-select-r  :
-    ∀{Γ₁ Γ₂ A A′ B B′}
-    {P : Process (B ∷ Γ₁)} {Q : Process (A′ ∷ Γ₂)} {R : Process (B′ ∷ Γ₂)}
-    (d : Dual A A′) (e : Dual B B′) (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
-    cut (d-⊕-& d e) p
-        (select false (split-l p₀) P)
-        (case (split-l q₀) Q R) ↝ cut e p P R
-
-  r-fork      :
-    ∀{Γ₁ Γ₂ Γ₃ Δ A B A′ B′}
-    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} {R : Process (B′ ∷ A′ ∷ Γ₃)}
-    (d : Dual A A′) (e : Dual B B′) (p : Γ ≃ Δ + Γ₃) (p₀ : Γ₃ ≃ [] + Γ₃)
-    (q : Δ ≃ Γ₁ + Γ₂) (q₀ : Δ ≃ [] + Δ) →
+  r-right     : ∀{Γ₁ Γ₂ A B P Q R}
+                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
+                cut {A ⊕ B} p (right (< p₀) P)
+                              (case (< q₀) Q R) ↝ cut p P R
+  r-fork      : ∀{Γ₁ Γ₂ Γ₃ Δ A B P Q R}
+                (p : Γ ≃ Δ + Γ₃) (p₀ : Γ₃ ≃ [] + Γ₃) (q : Δ ≃ Γ₁ + Γ₂) (q₀ : Δ ≃ [] + Δ) →
                 let _ , p′ , q′ = +-assoc-l p q in
-    cut (d-⊗-⅋ d e) p
-        (fork (split-l q₀) q P Q)
-        (join (split-l p₀) R) ↝ cut d q′ P (cut e (split-r p′) Q R)
-
-  r-client    :
-    ∀{Γ₁ Γ₂ A A′}
-    {P : Process (A ∷ Γ₁)} {Q : Process (A′ ∷ Γ₂)} (d : Dual A A′)
+                cut {A ⊗ B} p (fork (< q₀) q P Q)
+                              (join (< p₀) R) ↝ cut q′ P (cut (> p′) Q R)
+  r-client    : ∀{Γ₁ Γ₂ A P Q}
                 (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
-    cut (d-!-? d) p
-      (server (split-l p₀) un P)
-      (client (split-l q₀) Q) ↝ cut d p P Q
-
-  r-weaken    :
-    ∀{Γ₁ Γ₂ A A′}
-    {P : Process (A ∷ Γ₁)} {Q : Process Γ₂}
-    (d : Dual A A′) (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
-    cut (d-!-? d) p
-        (server (split-l p₀) un P)
-        (weaken (split-l q₀) Q) ↝ weakening un p Q
-
-  r-contract  :
-    ∀{Γ₁ Γ₂ A A′}
-    {P : Process (A ∷ Γ₁)} {Q : Process (¿ A′ ∷ ¿ A′ ∷ Γ₂)}
-    (d : Dual A A′) (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
-    cut (d-!-? d) p
-      (server (split-l p₀) un P)
-      (contract (split-l q₀) Q) ↝
+                cut {`! A} p (server (< p₀) un P) (client (< q₀) Q) ↝ cut p P Q
+  r-weaken    : ∀{Γ₁ Γ₂ A P Q}
+                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
+                cut {`! A} p (server (< p₀) un P)
+                             (weaken (< q₀) Q) ↝ weakening un p Q
+  r-contract  : ∀{Γ₁ Γ₂ A P Q}
+                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
+                cut {`! A} p (server (< p₀) un P)
+                             (contract (< q₀) Q) ↝
                     contraction un p
-        (cut (d-!-? d) ++≃+
-             (server (split-l p₀) un P)
-             (cut (d-!-? d) (split-r p) (server (split-l p₀) un P) Q))
+                      (cut ++≃+ (server (< p₀) un P)
+                      (cut (> p) (server (< p₀) un P) Q))
+  r-poly       : ∀{A B Γ₁ Γ₂ P F} (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) ->
+                 cut {`∃ A} p (ex {_} {B} (< p₀) P) (all (< q₀) F) ↝ cut p P (F B)
 \end{code}
@@
 \begin{code}
-  r-cut       : ∀{Γ₁ Γ₂ A B} {P Q : Process (A ∷ Γ₁)} {R : Process (B ∷ Γ₂)}
-                (d : Dual A B) (q : Γ ≃ Γ₁ + Γ₂) → P ↝ Q →
-                cut d q P R ↝ cut d q Q R
-
-  r-cong      : ∀{P R Q : Process Γ} → P ⊒ R → R ↝ Q → P ↝ Q
+  r-cut        : ∀{Γ₁ Γ₂ A P Q R} (q : Γ ≃ Γ₁ + Γ₂) →
+                 P ↝ Q → cut {A} q P R ↝ cut q Q R
+  r-cong       : ∀{P R Q} → P ⊒ R → R ↝ Q → P ↝ Q
 \end{code}

### 2025-12-03 paper/benchmarks/count.sh
snapshot

@@ -1,0 +1,2 @@
+#!/bin/sh
+exec cat `find $1 -name "*.*"` | tr -d [:space:] | wc -c

### 2025-12-03 paper/DeadlockFreedom.lagda.tex
restored DF

@@
 \begin{code}[hide]
+{-# OPTIONS --rewriting #-}
 open import Data.Sum
 open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
-open import Data.Bool using (Bool; true; false; if_then_else_)
 open import Data.List.Base using ([]; _∷_; [_])
@@
 data Cut where
-  cut :
-    ∀{Γ Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
-    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
-    Cut (cut d p P Q)
+  cut : ∀{Γ Γ₁ Γ₂ A P Q} (p : Γ ≃ Γ₁ + Γ₂) -> Cut (cut {A} p P Q)
 
 data Link where
-  link :
-    ∀{Γ A B} (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) → Link (link d p)
+  link : ∀{Γ A} (p : Γ ≃ [ A ] + [ dual A ]) → Link (link p)
 
 data Input where
-  fail :
-    ∀{Γ Δ}
-    (p : Γ ≃ [] + Δ) → Input (fail (split-l p))
+  fail : ∀{Γ Δ} (p : Γ ≃ [] + Δ) → Input (fail (< p))
 \end{code}
@@
 \begin{code}
-  wait : ∀{Γ Δ} (p : Γ ≃ [] + Δ) {P : Process Δ} → Input (wait (split-l p) P)
+  wait : ∀{Γ Δ P} (p : Γ ≃ [] + Δ) → Input (wait (< p) P)
 \end{code}
@@
 \begin{code}[hide]
-  case :
-    ∀{Γ Δ A B} (p : Γ ≃ [] + Δ) {P : Process (A ∷ Δ)} {Q : Process (B ∷ Δ)} →
-    Input (case (split-l p) P Q)
-  join :
-    ∀{Γ Δ A B} (p : Γ ≃ [] + Δ) {P : Process (B ∷ A ∷ Δ)} →
-    Input (join (split-l p) P)
-
-data Client where
-  client :
-    ∀{Γ Δ A} (p : Γ ≃ [] + Δ)
-    {P : Process (A ∷ Δ)} →
-    Client (client (split-l p) P)
-  weaken :
-    ∀{Γ Δ A} (p : Γ ≃ [] + Δ)
-    {P : Process Δ} →
-    Client (weaken {A = A} (split-l p) P)
-  contract :
-    ∀{Γ Δ A} (p : Γ ≃ [] + Δ)
-    {P : Process (¿ A ∷ ¿ A ∷ Δ)} →
-    Client (contract (split-l p) P)
+  case : ∀{Γ Δ A B P Q} (p : Γ ≃ [] + Δ) → Input (case {A} {B} (< p) P Q)
+  join : ∀{Γ Δ A B P} (p : Γ ≃ [] + Δ) → Input (join {A} {B} (< p) P)
+  all  : ∀{A Γ Δ F} (p : Γ ≃ [] + Δ) -> Input (all {A} (< p) F)
 
@@
   close  : Output close
-  select :
-    ∀{Γ Δ A B} (x : Bool) (p : Γ ≃ [] + Δ) {P : Process ((if x then A else B) ∷ Δ)} →
-    Output (select x (split-l p) P)
-  fork :
-    ∀{Γ Δ Δ₁ Δ₂ A B} (p : Γ ≃ [] + Δ) (q : Δ ≃ Δ₁ + Δ₂)
-    {P : Process (A ∷ Δ₁)} {Q : Process (B ∷ Δ₂)} →
-    Output (fork (split-l p) q P Q)
+  left   : ∀{Γ Δ A B P} (p : Γ ≃ [] + Δ) → Output (left {A} {B} (< p) P)
+  right  : ∀{Γ Δ A B P} (p : Γ ≃ [] + Δ) → Output (right {A} {B} (< p) P)
+  fork   : ∀{Γ Δ Δ₁ Δ₂ A B P Q} (p : Γ ≃ [] + Δ) (q : Δ ≃ Δ₁ + Δ₂) -> Output (fork {A} {B} (< p) q P Q)
+  ex     : ∀{A B Γ Δ P} (p : Γ ≃ [] + Δ) -> Output (ex {A} {B} (< p) P)
 
 data Delayed where
-  fail :
-    ∀{A Γ Δ}
-    (p : Γ ≃ ⊤ , Δ) → Delayed (fail (split-r {A} p))
+  fail     : ∀{C Γ Δ} (p : Γ ∋ ⊤ ⊳ Δ) → Delayed (fail (>_ {C} p))
 \end{code}
@@
 \begin{code}
-  wait : ∀{C Γ Δ} (p : Γ ≃ ⊥ , Δ) {P : Process (C ∷ Δ)} →
-         Delayed (wait (split-r p) P)
+  wait     : ∀{C Γ Δ P} (p : Γ ∋ ⊥ ⊳ Δ) → Delayed (wait (>_ {C} p) P)
 \end{code}
@@
 \begin{code}[hide]
-  case :
-    ∀{Γ Δ C A B} (p : Γ ≃ A & B , Δ) {P : Process (A ∷ C ∷ Δ)} {Q : Process (B ∷ C ∷ Δ)} →
-    Delayed (case (split-r p) P Q)
-  join :
-    ∀{Γ Δ C A B} (p : Γ ≃ A ⅋ B , Δ) {P : Process (B ∷ A ∷ C ∷ Δ)} →
-    Delayed (join (split-r p) P)
-  select :
-    ∀{Γ Δ C A B} (x : Bool) (p : Γ ≃ A ⊕ B , Δ) {P : Process ((if x then A else B) ∷ C ∷ Δ)} →
-    Delayed (select x (split-r p) P)
-  fork-l :
-    ∀{Γ Δ Δ₁ Δ₂ C A B} (p : Γ ≃ A ⊗ B , Δ) (q : Δ ≃ Δ₁ + Δ₂)
-    {P : Process (A ∷ C ∷ Δ₁)} {Q : Process (B ∷ Δ₂)} →
-    Delayed (fork (split-r p) (split-l q) P Q)
-  fork-r :
-    ∀{Γ Δ Δ₁ Δ₂ C A B} (p : Γ ≃ A ⊗ B , Δ) (q : Δ ≃ Δ₁ + Δ₂)
-    {P : Process (A ∷ Δ₁)} {Q : Process (B ∷ C ∷ Δ₂)} →
-    Delayed (fork (split-r p) (split-r q) P Q)
-  client :
-    ∀{Γ Δ A C} (p : Γ ≃ ¿ A , Δ) {P : Process (A ∷ C ∷ Δ)} →
-    Delayed (client (split-r p) P)
-  weaken :
-    ∀{Γ Δ A C} (p : Γ ≃ ¿ A , Δ) {P : Process (C ∷ Δ)} →
-    Delayed (weaken (split-r p) P)
-  contract :
-    ∀{Γ Δ A C} (p : Γ ≃ ¿ A , Δ) {P : Process (¿ A ∷ ¿ A ∷ C ∷ Δ)} →
-    Delayed (contract (split-r p) P)
+  case     : ∀{Γ Δ C A B P Q} (p : Γ ∋ A & B ⊳ Δ) → Delayed (case {A} {B} (>_ {C} p) P Q)
+  join     : ∀{Γ Δ C A B P} (p : Γ ∋ A ⅋ B ⊳ Δ) → Delayed (join (>_ {C} p) P)
+  left     : ∀{Γ Δ C A B P} (p : Γ ∋ A ⊕ B ⊳ Δ) → Delayed (left (>_ {C} p) P)
+  right    : ∀{Γ Δ C A B P} (p : Γ ∋ A ⊕ B ⊳ Δ) → Delayed (right (>_ {C} p) P)
+  fork-l   : ∀{Γ Δ Δ₁ Δ₂ C A B P Q} (p : Γ ∋ A ⊗ B ⊳ Δ) (q : Δ ≃ Δ₁ + Δ₂) →
+             Delayed (fork (>_ {C} p) (< q) P Q)
+  fork-r   : ∀{Γ Δ Δ₁ Δ₂ C A B P Q} (p : Γ ∋ A ⊗ B ⊳ Δ) (q : Δ ≃ Δ₁ + Δ₂) →
+             Delayed (fork (>_ {C} p) (> q) P Q)
+  client   : ∀{Γ Δ A C P} (p : Γ ∋ `? A ⊳ Δ) → Delayed (client (>_ {C} p) P)
+  weaken   : ∀{Γ Δ A C P} (p : Γ ∋ `? A ⊳ Δ) → Delayed (weaken (>_ {C} p) P)
+  contract : ∀{Γ Δ A C P} (p : Γ ∋ `? A ⊳ Δ) → Delayed (contract (>_ {C} p) P)
+  ex       : ∀{A B C Γ Δ P} (p : Γ ∋ `∃ A ⊳ Δ) → Delayed (ex {A} {B} (>_ {C} p) P)
+  all      : ∀{A C Γ Δ F} (p : Γ ∋ `∀ A ⊳ Δ) → Delayed (all (>_ {C} p) F)
+
+data Client where
+  client   : ∀{Γ Δ A P} (p : Γ ≃ [] + Δ) → Client (client {A} (< p) P)
+  weaken   : ∀{Γ Δ A P} (p : Γ ≃ [] + Δ) → Client (weaken {A} (< p) P)
+  contract : ∀{Γ Δ A P} (p : Γ ≃ [] + Δ) → Client (contract {A} (< p) P)
 
 data Server where
-  server :
-    ∀{Γ Δ A} (p : Γ ≃ [] + Δ) (un : Un Δ) {P : Process (A ∷ Δ)} →
-    Server (server (split-l p) un P)
+  server : ∀{Γ Δ A P} (p : Γ ≃ [] + Δ) (un : Un Δ) → Server (server {A} (< p) un P)
 
 data DelayedServer where
-  server :
-    ∀{Γ Δ A C} (p : Γ ≃ ¡ A , Δ) (un : Un Δ) {P : Process (A ∷ ¿ C ∷ Δ)} →
-    DelayedServer (server (split-r p) (un-∷ un) P)
+  server : ∀{Γ Δ A C P} (p : Γ ∋ `! A ⊳ Δ) (un : Un Δ) →
+           DelayedServer (server (> p) (un-∷ {C} un) P)
 \end{code}
@@
 \begin{code}[hide]
-link-thread : ∀{Γ A B} (d : Dual A B) (p : Γ ≃ [ A ] + [ B ]) → Thread (link d p)
-link-thread d p = inj₁ (link d p)
+link-thread : ∀{A Γ} (p : Γ ≃ [ A ] + [ dual A ]) → Thread (link p)
+link-thread p = inj₁ (link p)
 
@@
 
-wait-thread : ∀{Γ Δ} (p : Γ ≃ [ ⊥ ] + Δ) {P : Process Δ} → Thread (wait p P)
-wait-thread (split-l p) = inj₂ (inj₂ (inj₂ (inj₁ (wait p))))
-wait-thread (split-r p) = inj₂ (inj₁ (wait p))
+wait-thread : ∀{Γ Δ P} (p : Γ ≃ [ ⊥ ] + Δ) → Thread (wait p P)
+wait-thread (< p) = inj₂ (inj₂ (inj₂ (inj₁ (wait p))))
+wait-thread (> p) = inj₂ (inj₁ (wait p))
 
 fail-thread : ∀{Γ Δ} (p : Γ ≃ [ ⊤ ] + Δ) → Thread (fail p)
-fail-thread (split-l p) = inj₂ (inj₂ (inj₂ (inj₁ (fail p))))
-fail-thread (split-r p) = inj₂ (inj₁ (fail p))
+fail-thread (< p) = inj₂ (inj₂ (inj₂ (inj₁ (fail p))))
+fail-thread (> p) = inj₂ (inj₁ (fail p))
 
-case-thread :
-  ∀{Γ Δ A B} (p : Γ ≃ [ A & B ] + Δ) {P : Process (A ∷ Δ)} {Q : Process (B ∷ Δ)} →
-  Thread (case p P Q)
-case-thread (split-l p) = inj₂ (inj₂ (inj₂ (inj₁ (case p))))
-case-thread (split-r p) = inj₂ (inj₁ (case p))
+case-thread : ∀{A B Γ Δ P Q} (p : Γ ≃ [ A & B ] + Δ) → Thread (case p P Q)
+case-thread (< p) = inj₂ (inj₂ (inj₂ (inj₁ (case p))))
+case-thread (> p) = inj₂ (inj₁ (case p))
 
-join-thread :
-  ∀{Γ Δ A B} (p : Γ ≃ [ A ⅋ B ] + Δ) {P : Process (B ∷ A ∷ Δ)} →
-  Thread (join p P)
-join-thread (split-l p) = inj₂ (inj₂ (inj₂ (inj₁ (join p))))
-join-thread (split-r p) = inj₂ (inj₁ (join p))
+join-thread : ∀{A B Γ Δ P} (p : Γ ≃ [ A ⅋ B ] + Δ) → Thread (join p P)
+join-thread (< p) = inj₂ (inj₂ (inj₂ (inj₁ (join p))))
+join-thread (> p) = inj₂ (inj₁ (join p))
 
-select-thread :
-  ∀{Γ Δ A B} (x : Bool) (p : Γ ≃ A ⊕ B , Δ) {P : Process ((if x then A else B) ∷ Δ)} →
-  Thread (select x p P)
-select-thread x (split-l p) = inj₂ (inj₂ (inj₁ (select x p)))
-select-thread x (split-r p) = inj₂ (inj₁ (select x p))
+left-thread : ∀{A B Γ Δ P} (p : Γ ∋ A ⊕ B ⊳ Δ) → Thread (left p P)
+left-thread (< p) = inj₂ (inj₂ (inj₁ (left p)))
+left-thread (> p) = inj₂ (inj₁ (left p))
 
-fork-thread :
-  ∀{Γ Δ Δ₁ Δ₂ A B} (p : Γ ≃ [ A ⊗ B ] + Δ) (q : Δ ≃ Δ₁ + Δ₂)
-  {P : Process (A ∷ Δ₁)} {Q : Process (B ∷ Δ₂)} →
-  Thread (fork p q P Q)
-fork-thread (split-l p) q = inj₂ (inj₂ (inj₁ (fork p q)))
-fork-thread (split-r p) (split-l q) = inj₂ (inj₁ (fork-l p q))
-fork-thread (split-r p) (split-r q) = inj₂ (inj₁ (fork-r p q))
+right-thread : ∀{A B Γ Δ P} (p : Γ ∋ A ⊕ B ⊳ Δ) → Thread (right p P)
+right-thread (< p) = inj₂ (inj₂ (inj₁ (right p)))
+right-thread (> p) = inj₂ (inj₁ (right p))
 
-client-thread :
-  ∀{Γ Δ A} (p : Γ ≃ [ ¿ A ] + Δ)
-  {P : Process (A ∷ Δ)} →
-  Thread (client p P)
-client-thread (split-l p) = inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (client p)))))
-client-thread (split-r p) = inj₂ (inj₁ (client p))
+fork-thread : ∀{A B Γ Δ Δ₁ Δ₂ P Q} (p : Γ ≃ [ A ⊗ B ] + Δ) (q : Δ ≃ Δ₁ + Δ₂) → Thread (fork p q P Q)
+fork-thread (< p) q = inj₂ (inj₂ (inj₁ (fork p q)))
+fork-thread (> p) (< q) = inj₂ (inj₁ (fork-l p q))
+fork-thread (> p) (> q) = inj₂ (inj₁ (fork-r p q))
 
-weaken-thread :
-  ∀{Γ Δ A} (p : Γ ≃ [ ¿ A ] + Δ)
-  {P : Process Δ} →
-  Thread (weaken {A = A} p P)
-weaken-thread (split-l p) = inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (weaken p)))))
-weaken-thread (split-r p) = inj₂ (inj₁ (weaken p))
+client-thread : ∀{A Γ Δ P} (p : Γ ≃ [ `? A ] + Δ) → Thread (client p P)
+client-thread (< p) = inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (client p)))))
+client-thread (> p) = inj₂ (inj₁ (client p))
 
-contract-thread :
-  ∀{Γ Δ A} (p : Γ ≃ [ ¿ A ] + Δ)
-  {P : Process (¿ A ∷ ¿ A ∷ Δ)} →
-  Thread (contract p P)
-contract-thread (split-l p) = inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (contract p)))))
-contract-thread (split-r p) = inj₂ (inj₁ (contract p))
+weaken-thread : ∀{A Γ Δ P} (p : Γ ≃ [ `? A ] + Δ) → Thread (weaken {A = A} p P)
+weaken-thread (< p) = inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (weaken p)))))
+weaken-thread (> p) = inj₂ (inj₁ (weaken p))
 
-server-thread :
-  ∀{Γ Δ A} (p : Γ ≃ [ ¡ A ] + Δ) (un : Un Δ) {P : Process (A ∷ Δ)} →
-  Thread (server p un P)
-server-thread (split-l p) un = inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (server p un))))))
-server-thread (split-r p) (un-∷ un) = inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (server p un))))))
+contract-thread : ∀{A Γ Δ P} (p : Γ ≃ [ `? A ] + Δ) → Thread (contract p P)
+contract-thread (< p) = inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (contract p)))))
+contract-thread (> p) = inj₂ (inj₁ (contract p))
+
+server-thread : ∀{A Γ Δ P} (p : Γ ≃ [ `! A ] + Δ) (un : Un Δ) → Thread (server p un P)
+server-thread (< p) un = inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (server p un))))))
+server-thread (> p) (un-∷ un) = inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (server p un))))))
+
+ex-thread : ∀{A B Γ Δ P} (p : Γ ∋ `∃ A ⊳ Δ) -> Thread (ex {A} {B} p P)
+ex-thread (< p) = inj₂ (inj₂ (inj₁ (ex p)))
+ex-thread (> p) = inj₂ (inj₁ (ex p))
+
+all-thread : ∀{A Γ Δ F} (p : Γ ∋ `∀ A ⊳ Δ) -> Thread (all p F)
+all-thread (< p) = inj₂ (inj₂ (inj₂ (inj₁ (all p))))
+all-thread (> p) = inj₂ (inj₁ (all p))
 \end{code}
@@
 data CanonicalCut {Γ} : Process Γ → Set where
-  cc-link :
-    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
-    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
-    Link P → CanonicalCut (cut d p P Q)
-  cc-delayed :
-    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
-    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
-    Delayed P → CanonicalCut (cut d p P Q)
-  cc-delayed-server :
-    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
-    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
-    DelayedServer P → Server Q → CanonicalCut (cut d p P Q)
-  cc-redex :
-    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
-    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
-    Output P → Input Q → CanonicalCut (cut d p P Q)
-  cc-connect :
-    ∀{Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂)
-    {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)} →
-    Server P → Client Q → CanonicalCut (cut d p P Q)
+  cc-link           : ∀{Γ₁ Γ₂ A P Q} (p : Γ ≃ Γ₁ + Γ₂) →
+                      Link P → CanonicalCut (cut {A} p P Q)
+  cc-delayed        : ∀{Γ₁ Γ₂ A P Q} (p : Γ ≃ Γ₁ + Γ₂) →
+                      Delayed P → CanonicalCut (cut {A} p P Q)
+  cc-delayed-server : ∀{Γ₁ Γ₂ A P Q} (p : Γ ≃ Γ₁ + Γ₂) →
+                      DelayedServer P → Server Q → CanonicalCut (cut {A} p P Q)
+  cc-redex          : ∀{Γ₁ Γ₂ A P Q} (p : Γ ≃ Γ₁ + Γ₂) →
+                      Output P → Input Q → CanonicalCut (cut {A} p P Q)
+  cc-connect        : ∀{Γ₁ Γ₂ A P Q} (p : Γ ≃ Γ₁ + Γ₂) →
+                      Server P → Client Q → CanonicalCut (cut {A} p P Q)
 \end{code}
@@
 \begin{code}[hide]
-output-output :
-  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Output P × Output Q)
-output-output d-𝟙-⊥ (close , ())
+output-output : ∀{A Γ Δ P Q} → ¬ (Output {A ∷ Γ} P × Output {dual A ∷ Δ} Q)
+output-output (close , ())
 
-output-client :
-  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Output P × Client Q)
-output-client () (close , client p)
+output-client : ∀{A Γ Δ P Q} → ¬ (Output {A ∷ Γ} P × Client {dual A ∷ Δ} Q)
+output-client (close , ())
 
-output-server :
-  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Output P × Server Q)
-output-server () (close , server p un)
+output-server : ∀{A Γ Δ P Q} → ¬ (Output {A ∷ Γ} P × Server {dual A ∷ Δ} Q)
+output-server (close , ())
 
-output-delayed-server :
-  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Output P × DelayedServer Q)
-output-delayed-server () (close , server p un)
+output-delayed-server : ∀{A Γ Δ P Q} → ¬ (Output {A ∷ Γ} P × DelayedServer {dual A ∷ Δ} Q)
+output-delayed-server (close , ())
 
-input-input :
-  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Input P × Input Q)
-input-input d-⊤-𝟘 (fail p , ())
+input-input : ∀{A Γ Δ P Q} → ¬ (Input {A ∷ Γ} P × Input {dual A ∷ Δ} Q)
+input-input (fail _ , ())
 
-input-client :
-  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Input P × Client Q)
-input-client () (fail p₁ , client p)
+input-client : ∀{A Γ Δ P Q} → ¬ (Input {A ∷ Γ} P × Client {dual A ∷ Δ} Q)
+input-client (fail _ , ())
 
-input-server :
-  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Input P × Server Q)
-input-server () (fail p₁ , server p un)
+input-server : ∀{A Γ Δ P Q} → ¬ (Input {A ∷ Γ} P × Server {dual A ∷ Δ} Q)
+input-server (fail _ , ())
 
-input-delayed-server :
-  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Input P × DelayedServer Q)
-input-delayed-server () (fail p₁ , server p un)
+input-delayed-server : ∀{A Γ Δ P Q} → ¬ (Input {A ∷ Γ} P × DelayedServer {dual A ∷ Δ} Q)
+input-delayed-server (fail _ , ())
 
-client-client :
-  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Client P × Client Q)
-client-client () (client p , client p₁)
+client-client : ∀{A Γ Δ P Q} → ¬ (Client {A ∷ Γ} P × Client {dual A ∷ Δ} Q)
+client-client (client _ , ())
 
-client-delayed-server :
-  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Client P × DelayedServer Q)
-client-delayed-server () (client p₁ , server p un)
+client-delayed-server : ∀{A Γ Δ P Q} → ¬ (Client {A ∷ Γ} P × DelayedServer {dual A ∷ Δ} Q)
+client-delayed-server (client _ , ())
 
-server-server :
-  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (Server P × Server Q)
-server-server () (server p un , server p₁ un₁)
+server-server : ∀{A Γ Δ P Q} → ¬ (Server {A ∷ Γ} P × Server {dual A ∷ Δ} Q)
+server-server (server _ _ , ())
 
-delayed-server-delayed-served :
-  ∀{Γ Δ A B} {P : Process (A ∷ Γ)} {Q : Process (B ∷ Δ)} → Dual A B → ¬ (DelayedServer P × DelayedServer Q)
-delayed-server-delayed-served () (server p un , server p₁ un₁)
+delayed-server-delayed-served : ∀{A Γ Δ P Q} → ¬ (DelayedServer {A ∷ Γ} P × DelayedServer {dual A ∷ Δ} Q)
+delayed-server-delayed-served (server _ _ , ())
 \end{code}
@@
 \begin{code}
-canonical-cut :
-  ∀{Γ Γ₁ Γ₂ A B} {P : Process (A ∷ Γ₁)} {Q : Process (B ∷ Γ₂)}
-  (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) →
-  Thread P → Thread Q → ∃[ R ] CanonicalCut R × cut d p P Q ⊒ R
+canonical-cut : ∀{A Γ Γ₁ Γ₂ P Q} (p : Γ ≃ Γ₁ + Γ₂) →
+                Thread P → Thread Q → ∃[ R ] CanonicalCut R × cut {A} p P Q ⊒ R
 \end{code}
@@
 \begin{code}[hide]
-canonical-cut dc pc (inj₁ x) Qt = _ , cc-link dc pc x , s-refl
-canonical-cut dc pc (inj₂ x) (inj₁ y) = _ , cc-link (dual-symm dc) (+-comm pc) y , s-comm dc pc
-canonical-cut dc pc (inj₂ (inj₁ x)) (inj₂ y) = _ , cc-delayed dc pc x , s-refl
-canonical-cut dc pc (inj₂ (inj₂ x)) (inj₂ (inj₁ y)) = _ , cc-delayed (dual-symm dc) (+-comm pc) y , s-comm dc pc
-canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₁ y))) = contradiction (x , y) (output-output dc)
-canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = _ , cc-redex dc pc x y , s-refl
-canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) (output-client dc)
-canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) (output-server dc)
-canonical-cut dc pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) (output-delayed-server dc)
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₁ y))) = _ , cc-redex (dual-symm dc) (+-comm pc) y x , s-comm dc pc
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) (output-client (dual-symm dc))
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) (output-server (dual-symm dc))
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) (output-delayed-server (dual-symm dc))
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (x , y) (input-input dc)
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) (input-client dc)
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) (input-server dc)
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) (input-delayed-server dc)
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) (input-client (dual-symm dc))
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) (input-server (dual-symm dc))
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) (input-delayed-server (dual-symm dc))
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) (client-client dc)
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = _ , cc-connect (dual-symm dc) (+-comm pc) y x , s-comm dc pc
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) (client-delayed-server dc)
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = _ , cc-connect dc pc x y , s-refl
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (y , x) (client-delayed-server (dual-symm dc))
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) (server-server dc)
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = _ , cc-delayed-server (dual-symm dc) (+-comm pc) y x , s-comm dc pc
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = _ , cc-delayed-server dc pc x y , s-refl
-canonical-cut dc pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) (delayed-server-delayed-served dc)
+canonical-cut pc (inj₁ x) Qt = _ , cc-link pc x , s-refl
+canonical-cut pc (inj₂ x) (inj₁ y) = _ , cc-link (+-comm pc) y , s-comm pc
+canonical-cut pc (inj₂ (inj₁ x)) (inj₂ y) = _ , cc-delayed pc x , s-refl
+canonical-cut pc (inj₂ (inj₂ x)) (inj₂ (inj₁ y)) = _ , cc-delayed (+-comm pc) y , s-comm pc
+canonical-cut pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₁ y))) = contradiction (x , y) output-output
+canonical-cut pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = _ , cc-redex pc x y , s-refl
+canonical-cut pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) output-client
+canonical-cut pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) output-server
+canonical-cut pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) output-delayed-server
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₁ y))) = _ , cc-redex (+-comm pc) y x , s-comm pc
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) output-client
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) output-server
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) output-delayed-server
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (x , y) (input-input)
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) input-client
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) input-server
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) input-delayed-server
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) input-client
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) input-server
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) input-delayed-server
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) client-client
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = _ , cc-connect (+-comm pc) y x , s-comm pc
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) client-delayed-server
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = _ , cc-connect pc x y , s-refl
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (y , x) client-delayed-server
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) server-server
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = _ , cc-delayed-server (+-comm pc) y x , s-comm pc
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = _ , cc-delayed-server pc x y , s-refl
+canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) delayed-server-delayed-served
 
@@
 \begin{code}[hide]
-canonical-cut-alive (cc-link d p (link e (split-l (split-r split-e)))) with dual-fun-r e d
-... | refl = inj₂ (_ , r-link d e p)
-canonical-cut-alive (cc-link d p (link e (split-r (split-l split-e)))) with dual-fun-l e (dual-symm d)
-... | refl = inj₂ (_ , r-cong (s-cong d p (s-link e (split-r (split-l split-e))) s-refl) (r-link d (dual-symm e) p))
-canonical-cut-alive (cc-redex d-𝟙-⊥ p close (wait q)) with +-empty-l p | +-empty-l q
+canonical-cut-alive (cc-link p (link (< > •))) = inj₂ (_ , r-link p)
+canonical-cut-alive (cc-link p (link (> < •))) =
+  inj₂ (_ , r-cong (s-cong p (s-link (> < •)) s-refl) (r-link p))
+canonical-cut-alive (cc-redex p close (wait q)) with +-empty-l p | +-empty-l q
 ... | refl | refl = inj₂ (_ , r-close p q)
-canonical-cut-alive (cc-redex (d-⊕-& d e) p (select false q) (case r)) with +-empty-l q | +-empty-l r
-... | refl | refl = inj₂ (_ , r-select-r d e p q r)
-canonical-cut-alive (cc-redex (d-⊕-& d e) p (select true q) (case r)) with +-empty-l q | +-empty-l r
-... | refl | refl = inj₂ (_ , r-select-l d e p q r)
-canonical-cut-alive (cc-redex (d-⊗-⅋ d e) p (fork q r) (join s)) with +-empty-l q | +-empty-l s
-... | refl | refl = inj₂ (_ , r-fork d e p s r q)
-canonical-cut-alive (cc-connect (d-!-? d) p (server q un) (client r)) with +-empty-l q | +-empty-l r
-... | refl | refl = inj₂ (_ , r-client d p q r un)
-canonical-cut-alive (cc-connect (d-!-? d) p (server q un) (weaken r)) with +-empty-l q | +-empty-l r
-... | refl | refl = inj₂ (_ , r-weaken d p q r un)
-canonical-cut-alive (cc-connect (d-!-? d) p (server q un) (contract r)) with +-empty-l q | +-empty-l r
-... | refl | refl = inj₂ (_ , r-contract d p q r un)
-canonical-cut-alive (cc-delayed d p (wait q)) =
+canonical-cut-alive (cc-redex p (right q) (case r)) with +-empty-l q | +-empty-l r
+... | refl | refl = inj₂ (_ , r-right p q r)
+canonical-cut-alive (cc-redex p (left q) (case r)) with +-empty-l q | +-empty-l r
+... | refl | refl = inj₂ (_ , r-left p q r)
+canonical-cut-alive (cc-redex p (fork q r) (join s)) with +-empty-l q | +-empty-l s
+... | refl | refl = inj₂ (_ , r-fork p s r q)
+canonical-cut-alive (cc-redex p (ex q) (all r)) with +-empty-l q | +-empty-l r
+... | refl | refl = inj₂ (_ , r-poly p q r)
+canonical-cut-alive (cc-connect p (server q un) (client r)) with +-empty-l q | +-empty-l r
+... | refl | refl = inj₂ (_ , r-client p q r un)
+canonical-cut-alive (cc-connect p (server q un) (weaken r)) with +-empty-l q | +-empty-l r
+... | refl | refl = inj₂ (_ , r-weaken p q r un)
+canonical-cut-alive (cc-connect p (server q un) (contract r)) with +-empty-l q | +-empty-l r
+... | refl | refl = inj₂ (_ , r-contract p q r un)
+canonical-cut-alive (cc-delayed p (wait q)) =
   let _ , _ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-wait d p q , wait-thread q′)
-canonical-cut-alive (cc-delayed d p (case q)) =
+  inj₁ (_ , s-wait p q , wait-thread q′)
+canonical-cut-alive (cc-delayed p (case q)) =
   let _ , _ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-case d p q , case-thread q′)
-canonical-cut-alive (cc-delayed d p (join q)) =
+  inj₁ (_ , s-case p q , case-thread q′)
+canonical-cut-alive (cc-delayed p (join q)) =
   let _ , _ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-join d p q , join-thread q′)
-canonical-cut-alive (cc-delayed d p (select false q)) =
+  inj₁ (_ , s-join p q , join-thread q′)
+canonical-cut-alive (cc-delayed p (right q)) =
   let _ , _ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-select-r d p q , select-thread false q′)
-canonical-cut-alive (cc-delayed d p (select true q)) =
+  inj₁ (_ , s-right p q , right-thread q′)
+canonical-cut-alive (cc-delayed p (left q)) =
   let _ , _ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-select-l d p q , select-thread true q′)
-canonical-cut-alive (cc-delayed d p (fork-l q r)) =
+  inj₁ (_ , s-left p q , left-thread q′)
+canonical-cut-alive (cc-delayed p (fork-l q r)) =
   let _ , p′ , q′ = +-assoc-l p q in
@@
   let _ , q′′ , r′′ = +-assoc-r r′ (+-comm p′′) in
-  inj₁ (_ , s-fork-l d p q r , fork-thread q′ r′′)
-canonical-cut-alive (cc-delayed d p (fork-r q r)) =
+  inj₁ (_ , s-fork-l p q r , fork-thread q′ r′′)
+canonical-cut-alive (cc-delayed p (fork-r q r)) =
   let _ , p′ , q′ = +-assoc-l p q in
   let _ , p′′ , r′ = +-assoc-l p′ r in
-  inj₁ (_ , s-fork-r d p q r , fork-thread q′ r′)
-canonical-cut-alive (cc-delayed d p (fail q)) =
+  inj₁ (_ , s-fork-r p q r , fork-thread q′ r′)
+canonical-cut-alive (cc-delayed p (fail q)) =
   let _ , p′ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-fail d p q , fail-thread q′)
-canonical-cut-alive (cc-delayed d p (client q)) =
+  inj₁ (_ , s-fail p q , fail-thread q′)
+canonical-cut-alive (cc-delayed p (client q)) =
   let _ , _ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-client d p q , client-thread q′)
-canonical-cut-alive (cc-delayed d p (weaken q)) =
+  inj₁ (_ , s-client p q , client-thread q′)
+canonical-cut-alive (cc-delayed p (weaken q)) =
   let _ , _ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-weaken d p q , weaken-thread q′)
-canonical-cut-alive (cc-delayed d p (contract q)) =
+  inj₁ (_ , s-weaken p q , weaken-thread q′)
+canonical-cut-alive (cc-delayed p (contract q)) =
   let _ , _ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-contract d p q , contract-thread q′)
-canonical-cut-alive (cc-delayed-server (d-?-! d) p (server q un) (server r un′)) with +-empty-l r
+  inj₁ (_ , s-contract p q , contract-thread q′)
+canonical-cut-alive (cc-delayed p (ex q)) =
+  let _ , _ , q' = +-assoc-l p q in
+  inj₁ (_ , s-ex p q , ex-thread q')
+canonical-cut-alive (cc-delayed p (all q)) =
+  let _ , _ , q' = +-assoc-l p q in
+  inj₁ (_ , s-all p q , all-thread q')
+canonical-cut-alive (cc-delayed-server p (server q un) (server r un′)) with +-empty-l r
 ... | refl =
   let _ , p′ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-server d p q r un un′ , server-thread q′ (#un+ p′ un un′))
+  inj₁ (_ , s-server p q r un un′ , server-thread q′ (+-un p′ un un′))
 \end{code}
@@
 \begin{code}[hide]
-deadlock-freedom (link d p) = inj₁ (_ , s-refl , link-thread d p)
+deadlock-freedom (link p) = inj₁ (_ , s-refl , link-thread p)
 deadlock-freedom (fail p) = inj₁ (_ , s-refl , fail-thread p)
@@
 deadlock-freedom (wait p P) = inj₁ (_ , s-refl , wait-thread p)
-deadlock-freedom (select x p P) = inj₁ (_ , s-refl , select-thread x p)
+deadlock-freedom (left p P) = inj₁ (_ , s-refl , left-thread p)
+deadlock-freedom (right p P) = inj₁ (_ , s-refl , right-thread p)
 deadlock-freedom (case p P Q) = inj₁ (_ , s-refl , case-thread p)
@@
 deadlock-freedom (contract p P) = inj₁ (_ , s-refl , contract-thread p)
-deadlock-freedom (cut d p P Q) with deadlock-freedom P
-... | inj₂ (_ , red) = inj₂ (_ , r-cut d p red)
+deadlock-freedom (ex p P) = inj₁ (_ , s-refl , ex-thread p)
+deadlock-freedom (all p F) = inj₁ (_ , s-refl , all-thread p)
+deadlock-freedom (cut p P Q) with deadlock-freedom P
+... | inj₂ (_ , red) = inj₂ (_ , r-cut p red)
 ... | inj₁ (_ , Pc , Pt) with deadlock-freedom Q
-... | inj₂ (_ , red) = inj₂ (_ , r-cong (s-comm d p) (r-cut (dual-symm d) (+-comm p) red))
-... | inj₁ (_ , Qc , Qt) with canonical-cut d p Pt Qt
-... | _ , cc , pcong = ⊒Alive (s-tran (s-cong d p Pc Qc) pcong) (canonical-cut-alive cc)
+... | inj₂ (_ , red) = inj₂ (_ , r-cong (s-comm p) (r-cut (+-comm p) red))
+... | inj₁ (_ , Qc , Qt) with canonical-cut p Pt Qt
+... | _ , cc , pcong = ⊒Alive (s-tran (s-cong p Pc Qc) pcong) (canonical-cut-alive cc)
 \end{code}

### 2025-12-03 paper/DeadlockFreedom.lagda.tex
revised DF

@@
 now formalize in Agda. First of all we need to distinguish \emph{threads} from
-\emph{cuts}. As it turns out, in the formalization it is convenient to provide
-an even more fine-grained classification of threads, distinguishing between
-links, input/output actions on linear/shared channels, and sometimes also
-whether such actions operate on free or bound channels. We define predicates for
-each of these process classes:
+\emph{cuts}. As it turns out, in the formalization it is convenient to provide a
+more fine-grained classification of threads, distinguishing between links,
+input/output actions on linear/shared channels, and sometimes also whether such
+actions operate on free or bound channels. We define predicates for each of
+these process classes:
 
@@
 isomorphic to the relevant parts of the \AgdaDatatype{Process} datatype. The
-only remarkable aspect is that \AgdaDatatype{Input}, \AgdaDatatype{Output},
+only remarkable aspect is that in \AgdaDatatype{Input}, \AgdaDatatype{Output},
 \AgdaDatatype{Client} and \AgdaDatatype{Server} the channel being acted upon by
@@
 \end{code}
-where the \AgdaInductiveConstructor{split-l} constructor indicates that the
-thread operates on the most recent channel. In the latter predicate we have
+where the \AgdaInductiveConstructor{<} constructor indicates that the thread
+operates on the most recent channel. In the latter predicate we have
 
@@
 \end{code}
-where the \AgdaInductiveConstructor{split-r} constructor indicates that the
-thread operates on a channel other than the most recent one.
+where the \AgdaInductiveConstructor{>} constructor indicates that the thread
+operates on a channel other than the most recent one. Note that here we have to
+specify the type $C$ in front of $\Context$ or else Agda is unable to resolve
+some metavariables.
 
@@
 In order to prove that every (well-typed) process is alive, it is convenient to
-also define a ``canonical'' form for cuts, which is a form that matches at least
+also define a ``canonical'' form for cuts, that is a form that matches at least
 one of the left-hand-sides of the rules for structural pre-congruence or
@@
 
-A \emph{canonical cut} $\Cut\x{P}{Q}$ has one of these forms:
+A \emph{canonical cut} $\Cut[A]\x{P}{Q}$ has one of these forms:
 \begin{itemize}
@@
 
-Every cut $\Cut\x{P}{Q}$ where both $P$ and $Q$ are threads can be rewritten
+Every cut $\Cut[A]\x{P}{Q}$ where both $P$ and $Q$ are threads can be rewritten
 into a canonical cut using structural pre-congruence:
@@
 canonical-cut : ∀{A Γ Γ₁ Γ₂ P Q} (p : Γ ≃ Γ₁ + Γ₂) →
-                Thread P → Thread Q → ∃[ R ] CanonicalCut R × cut {A} p P Q ⊒ R
+                Thread P → Thread Q →
+                ∃[ R ] CanonicalCut R × cut {A} p P Q ⊒ R
 \end{code}
@@
 When $P$ is a thread, then it is obviously observable and hence alive. When $P$
-is a cut $\Cut\x{P}{Q}$, \AgdaFunction{deadlock-freedom} is applied recursively
-on $P$ and $Q$. If either of these applications yields a reduction, then the
-whole cut is reducible and therefore alive. If both applications yield a thread,
-then we conclude that the cut is alive first rewriting it into a canonical cut
-(\AgdaFunction{canonical-cut}) and then using
+is a cut $\Cut[A]\x{P}{Q}$, \AgdaFunction{deadlock-freedom} is applied
+recursively on $P$ and $Q$. If either of these applications yields a reduction,
+then the whole cut is reducible and therefore alive. If both applications yield
+a thread, then we conclude that the cut is alive first rewriting it into a
+canonical cut (\AgdaFunction{canonical-cut}) and then using
 \AgdaFunction{canonical-cut-alive}.

### 2025-12-03 paper/Safety.lagda.tex
restored type safety

@@
-\section{Type Safety}
+\subsection{Type Safety}
 \label{sec:safety-agda}
@@
 \begin{code}[hide]
-open import Data.Sum
-open import Data.Product using (_,_)
-open import Relation.Nullary using (¬_; contradiction)
+{-# OPTIONS --rewriting #-}
 open import Data.List.Base using ([]; _∷_; [_])
@@
 open import Congruence
-open import Reduction
 open import DeadlockFreedom
@@
   hole   : ReductionContext Δ Δ
-  cut-l  : ∀{Γ Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) →
-           ReductionContext Δ (A ∷ Γ₁) → Process (B ∷ Γ₂) →
+  cut-l  : ∀{Γ Γ₁ Γ₂ A} (p : Γ ≃ Γ₁ + Γ₂) →
+           ReductionContext Δ (A ∷ Γ₁) → Process (dual A ∷ Γ₂) →
            ReductionContext Δ Γ
-  cut-r  : ∀{Γ Γ₁ Γ₂ A B} (d : Dual A B) (p : Γ ≃ Γ₁ + Γ₂) →
-           Process (A ∷ Γ₁) → ReductionContext Δ (B ∷ Γ₂) →
+  cut-r  : ∀{Γ Γ₁ Γ₂ A} (p : Γ ≃ Γ₁ + Γ₂) →
+           Process (A ∷ Γ₁) → ReductionContext Δ (dual A ∷ Γ₂) →
            ReductionContext Δ Γ
@@
 hole ⟦ P ⟧         = P
-cut-l d p 𝒞 Q ⟦ P ⟧  = cut d p (𝒞 ⟦ P ⟧) Q
-cut-r d p Q 𝒞 ⟦ P ⟧  = cut d p Q (𝒞 ⟦ P ⟧)
+cut-l p 𝒞 Q ⟦ P ⟧  = cut p (𝒞 ⟦ P ⟧) Q
+cut-r p Q 𝒞 ⟦ P ⟧  = cut p Q (𝒞 ⟦ P ⟧)
 \end{code}

### 2025-12-03 paper/Examples.lagda.tex
preparing examples

\section{Examples}
\label{sec:examples}

\begin{code}[hide]
{-# OPTIONS --rewriting #-}
open import Data.Sum hiding (reduce; swap)
open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
open import Data.Nat using (ℕ; zero; suc)
open import Data.Fin using (zero; suc; #_)
open import Data.List.Base using (List; []; _∷_; [_]; _++_)
open import Function using (_$_)
open import Data.Maybe

open import Type
open import Context
open import Permutations
open import Process
open import DeadlockFreedom using (deadlock-freedom)
\end{code}

\begin{code}
𝔹 : Type
𝔹 = 𝟙 ⊕ 𝟙

True : Process [ 𝔹 ]
True = left (< ≫) close

False : Process [ 𝔹 ]
False = right (< ≫) close

Not : Process (dual 𝔹 ∷ 𝔹 ∷ [])
Not = case (< ≫) (wait (< ≫) False) (wait (< ≫) True)

Copy : Process (dual 𝔹 ∷ 𝔹 ∷ [])
Copy = cut (< ≫) (↭process swap Not) Not

Drop : Process (dual 𝔹 ∷ 𝟙 ∷ [])
Drop = case (< ≫) (wait (< ≫) close) (wait (< ≫) close)

And : Process (dual 𝔹 ∷ dual 𝔹 ∷ 𝔹 ∷ [])
And = case (< ≫)
           (wait (< ≫) Copy)
           (wait (< ≫)
                 (cut (< ≫)
                      (↭process swap Drop)
                      (wait (< ≫) False)))

Or : Process (dual 𝔹 ∷ dual 𝔹 ∷ 𝔹 ∷ [])
Or = cut (< < ≫)
         (cut (> < ≫)
              (↭process swap Not)
              (cut (> > < ≫)
                   (↭process swap Not)
                   And))
         Not

ex1 : Process [ 𝔹 ]
ex1 = cut ≫ False (cut ≫ False Or)

reduce : ∀{Γ} → ℕ → Process Γ → Process Γ
reduce zero P = P
reduce (suc n) P with deadlock-freedom P
... | inj₁ (Q , _ , _)  = Q
... | inj₂ (Q , _)      = reduce n Q

_⊸_ : ∀{n} → PreType n → PreType n → PreType n
A ⊸ B = dual A ⅋ B

echo :  let X = var (# 0) in
        Process [ `! (`∀ (X ⊸ X)) ]
echo =  server (< ≫) un-[] $
        all (< ≫) λ X →
        join (< ≫) $
        link (> < ≫)

⊗-comm :  let X = var (# 1) in
          let Y = var (# 0) in
          Process [ `∀ (`∀ ((X ⊗ Y) ⊸ (Y ⊗ X))) ]
⊗-comm =  all (< ≫) λ X →
          all (< ≫) λ Y →
          join (< ≫) $
          join (> < ≫) $
          fork (> > < ≫) (< ≫)
               (link (< ≫))
               (link (< ≫))
\end{code}

### 2025-12-04 paper/Examples.lagda.tex
drafted examples

@@
 
+In this section we illustrate a few simple examples of processes. We start by
+encoding the type of a boolean value being produced as a choice between ``going
+left'' (which we interpret as true) and ``going right'' (which we interpret as
+false).
+
 \begin{code}
@@
 𝔹 = 𝟙 ⊕ 𝟙
+\end{code}
 
-True : Process [ 𝔹 ]
-True = left (< ≫) close
+The boolean constants are processes that perform one of these choices.
 
-False : Process [ 𝔹 ]
-False = right (< ≫) close
+\begin{code}
+true : Process [ 𝔹 ]
+true = left (< ≫) close
 
-Not : Process (dual 𝔹 ∷ 𝔹 ∷ [])
-Not = case (< ≫) (wait (< ≫) False) (wait (< ≫) True)
+false : Process [ 𝔹 ]
+false = right (< ≫) close
+\end{code}
 
-Copy : Process (dual 𝔹 ∷ 𝔹 ∷ [])
-Copy = cut (< ≫) (↭process swap Not) Not
+We can use the host language for programming higher-order processes. For
+example, we can define a conditional process thus:
 
-Drop : Process (dual 𝔹 ∷ 𝟙 ∷ [])
-Drop = case (< ≫) (wait (< ≫) close) (wait (< ≫) close)
+\begin{code}
+if_else : ∀{Γ} → Process Γ → Process Γ → Process (dual 𝔹 ∷ Γ)
+if P else Q = case (< ≫) (wait (< ≫) P) (wait (< ≫) Q)
+\end{code}
 
-And : Process (dual 𝔹 ∷ dual 𝔹 ∷ 𝔹 ∷ [])
-And = case (< ≫)
-           (wait (< ≫) Copy)
-           (wait (< ≫)
-                 (cut (< ≫)
-                      (↭process swap Drop)
-                      (wait (< ≫) False)))
+A term $\AgdaFunction{if}~P~\AgdaFunction{else}~Q$ is a process that waits for a
+boolean value (\cf the $\AgdaFunction{dual}~\mathbb{B}$ type at the front of its
+typing context) and continues as either $P$ or $Q$ depending on whether it
+receives true or false.
+%
+We can then define a process $\AgdaFunction{drop}~P$ that waits for consumes a
+boolean and continues as $P$ regardless of its value.
 
-Or : Process (dual 𝔹 ∷ dual 𝔹 ∷ 𝔹 ∷ [])
-Or = cut (< < ≫)
-         (cut (> < ≫)
-              (↭process swap Not)
-              (cut (> > < ≫)
-                   (↭process swap Not)
-                   And))
-         Not
+\begin{code}
+drop : ∀{Γ} → Process Γ → Process (dual 𝔹 ∷ Γ)
+drop P = if P else P
+\end{code}
 
-ex1 : Process [ 𝔹 ]
-ex1 = cut ≫ False (cut ≫ False Or)
+Using these higher-order forms, it is easy to define the usual boolean
+connectives.
 
-reduce : ∀{Γ} → ℕ → Process Γ → Process Γ
-reduce zero P = P
-reduce (suc n) P with deadlock-freedom P
+\begin{code}
+!!_ : Process [ 𝔹 ] → Process [ 𝔹 ]
+!!_ B = cut ≫ B (if false else true)
+
+_&&_ _||_  : Process [ 𝔹 ] → Process [ 𝔹 ] → Process [ 𝔹 ]
+A && B   = cut ≫ A (cut ≫ B (if (link (< ≫)) else (drop false)))
+A || B   = !! ((!! A) && (!! B))
+\end{code}
+
+Note the use of \AgdaInductiveConstructor{cut}s for combining a process that
+produces a boolean value with a process that consumes a boolean value and the
+use of \AgdaInductiveConstructor{link} for copying a boolean value from one
+channel to another. 
+%
+Here we have chosen to define the disjunction \AgdaFunction{||} from the
+conjunction \AgdaFunction{\&\&} and negation \AgdaFunction{!!} using De Morgan's
+laws.
+
+If we wish to test our definitions, we can implement a simple (lazy) evaluator
+using the deadlock freedom property. We have not proved a termination result,
+but since linear logic enjoys cut elimination we can safely annotate the
+evaluator as terminating.
+
+\begin{code}
+{-# TERMINATING #-}
+eval : ∀{Γ} → Process Γ → Process Γ
+eval P with deadlock-freedom P
 ... | inj₁ (Q , _ , _)  = Q
-... | inj₂ (Q , _)      = reduce n Q
+... | inj₂ (Q , _)      = eval Q
+\end{code}
 
+Now if we ask Agda to normalize the goal
+$\AgdaFunction{eval}~(\AgdaFunction{false}~\AgdaFunction{||}~\AgdaFunction{false})$
+we obtain
+$\AgdaInductiveConstructor{right}~(\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{•})~\AgdaInductiveConstructor{close}$,
+that is the definition of \AgdaFunction{false}, as expected.
+
+For the example that follows, it is convenient to introduce the linear
+implication connective according to the standard definition $A \multimap B
+\eqdef \dual{A} \Par B$.
+
+\begin{code}
 _⊸_ : ∀{n} → PreType n → PreType n → PreType n
 A ⊸ B = dual A ⅋ B
+\end{code}
 
+Below we define a persistent, polymorphic echo \AgdaInductiveConstructor{server}
+that waits for a value and sends it back unchanged. The server is polymorphic in
+the sense that the type of the value is arbitrary and denoted by the type
+variable $X$. The notation $\AgdaFunction{\#}~n$ is defined in Agda's standard
+library and creates an element of \AgdaDatatype{Fin} from the natural number
+$n$. The function \AgdaFunction{\$} is also defined in the standard library and
+denotes function application. We use it here as ``command separator'' to save a
+few parentheses.
+
+\begin{code}
 echo :  let X = var (# 0) in
@@
         join (< ≫) $
-        link (> < ≫)
+        link (< ≫)
+\end{code}
 
-⊗-comm :  let X = var (# 1) in
-          let Y = var (# 0) in
-          Process [ `∀ (`∀ ((X ⊗ Y) ⊸ (Y ⊗ X))) ]
-⊗-comm =  all (< ≫) λ X →
-          all (< ≫) λ Y →
-          join (< ≫) $
-          join (> < ≫) $
-          fork (> > < ≫) (< ≫)
-               (link (< ≫))
-               (link (< ≫))
+We can test the server by connecting it with a \AgdaInductiveConstructor{client}
+that sends true to it.
+
+\begin{code}
+echo-true : Process [ 𝔹 ]
+echo-true = cut ≫ echo (client (< ≫) $
+                       ex (< ≫) $
+                       fork (< ≫) ≫ true (link (< ≫)))
 \end{code}
+
+If we ask Agda to evaluate this process we obtain
+$\AgdaInductiveConstructor{left}~(\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{•})~\AgdaInductiveConstructor{close}$,
+that is the definition of \AgdaFunction{true} as expected.
+
+From these examples it should be clear that writing \Calculus processes in Agda
+is not entirely straightforward, but it is still doable with a little effort.
+The lack of names (for channels) clearly obfuscates the code, but the context
+splittings that must be provided are usually small and fairly easy to
+understand. Most importantly, the use of \AgdaFunction{≫} helps truncating
+context splittings as soon as the seeked channel has been reached.
+%
+Either way, the formalization is meant to study the meta-theory of \Calculus
+rather than implementing a user-friendly DSL for it.
+
+% \begin{code}
+% ⊗-comm : let X = var (# 1) in
+%          let Y = var (# 0) in
+%          Process [ `∀ (`∀ ((X ⊗ Y) ⊸ (Y ⊗ X))) ]
+% ⊗-comm = all (< ≫) λ X →
+%          all (< ≫) λ Y →
+%          join (< ≫) $
+%          join (> < ≫) $
+%          fork (> > < ≫) (< ≫)
+%               (link (< ≫))
+%               (link (< ≫))
+% \end{code}

### 2025-12-04 paper/Reduction.lagda.tex
revised description of reduction rules

@@
 
-\begin{AgdaAlign}
 \begin{code}
@@
 
-The constructor \AgdaInductiveConstructor{r-link} models \RLink.
+The constructor \AgdaInductiveConstructor{r-link} models the reduction
+$\Cut[A]\x{\Link\x\y}{P} \red P\subst\y\x$ called \RLink in
+\Cref{tab:semantics}:
 
 \begin{code}
-  r-link      : ∀{Δ A P} (p : Γ ∋ dual A ⊳ Δ) →
+  r-link      : ∀{Δ A P} (p : Γ ≃ [ dual A ] + Δ) →
                 cut {A} p (link (< > •)) P ↝ ↭process (↭concat p) P
@@
 
-For generality, the rules quantifies over two proofs $d$ and $e$ of the same
-duality relation between $A$ and $B$. Even though such proofs would necessarily
-be the same, distinguishing them saves us a proof that they are equal whenever
-we want to apply this reduction. 
+The splitting $p$ indicates that $y$ (of type $\dual{A}$) occurs on the left
+sub-process of the cut (that is the link $\Link\x\y$) and the splitting
+$(\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{>}~\AgdaInductiveConstructor{•})$
+to which \AgdaInductiveConstructor{link} is applied is structured consistently
+with the syntax of the link $\Link\x\y$ being reduced, which is oriented so that
+the restricted channel $x$ occurs on the left. 
 %
-Also note that the splitting
-$(\AgdaInductiveConstructor{split-l}~(\AgdaInductiveConstructor{split-r}~\AgdaInductiveConstructor{split-e}))$
-to which \AgdaInductiveConstructor{link} is applied is structured consistently
-with \Cref{tab:semantics}, where the link $\Link\x\y$ being reduced is oriented
-so that the restricted channel $x$ occurs on the left. 
-%
-Finally, the process $P\subst\y\x$ resulting from the reduction is subject to a
-channel substitution in \Cref{tab:semantics}. In the Agda representation of
-processes with nameless channels, the same effect is achieved by the permutation
-$\AgdaFunction{\#cons}~p$, which ``renames'' (the position of) the free
-occurrence of $x$ within $P$ with (the position of) $y$ encoded in the splitting
-$p$.
+The process $P$ has type
+$\AgdaDatatype{Process}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\ContextD)$,
+and turns into $P\subst\y\x$ after the reduction. The type
+$\AgdaFunction{dual}~A$ of $x$ in $P$ is the first in the typing context,
+indicating that it is the newest channel that $P$ is aware of. However, after
+the reduction, $x$ is replaced by $y$ which is found \emph{somewhere} within
+$\Context$. The exact location of $y$ in $\ContextC$ is found in the splitting
+$p$. We ``rename'' $x$ into $y$ within $P$ using the permutation
+$\AgdaFunction{↭concat}~p$.
 
@@
 \AgdaInductiveConstructor{wait} constructor uses a splitting proof of the form
-$(\AgdaInductiveConstructor{split-l}~q_0)$ to be sure that the referenced
-channel is also the restricted one, and therefore matches the one of the
+$(\AgdaInductiveConstructor{<}~q_0)$ to be sure that the referenced channel is
+also the restricted one, and therefore matches the one of the
 \AgdaInductiveConstructor{close} process.
-%
-Just like \AgdaInductiveConstructor{r-link} quantifies over multiple proofs for
-the same duality relation, also \AgdaInductiveConstructor{r-close} (and several
-other reduction constructors) quantifies over $p_0$ and $q_0$ which both prove
-the splitting $\Context \simeq \AgdaInductiveConstructor{[]} + \Context$. Since
-the left context is empty, these splittings must be equal and made of a sequence
-of \AgdaInductiveConstructor{split-r} applications followed by
-\AgdaInductiveConstructor{split-e}, but it is easier to quantify them separately
-so that we do not have to worry about proving their equality whenever we wish to
+
+Note that \AgdaInductiveConstructor{r-close} (and several other reduction
+constructors) quantifies over $p_0$ and $q_0$ which both prove the splitting
+$\Context \simeq \AgdaInductiveConstructor{[]} + \Context$. Since the left
+context is empty, these splittings must be equal and made of a sequence of
+\AgdaInductiveConstructor{>} applications followed by
+\AgdaInductiveConstructor{•}. In general, Agda will not be able to ``see'' that
+they are definitionally equal, hence it is easier to quantify them separately so
+that we do not have to worry about proving their equality whenever we wish to
 apply this reduction.
 
-The constructor \AgdaInductiveConstructor{r-select-l} models \RSelect when the
-selected tag is $\InTag_1$, that is \AgdaInductiveConstructor{true} in Agda:
+The constructor \AgdaInductiveConstructor{r-left} models \RSelect when the
+selected tag is $\InTag_1$:
 
@@
 
-There is not much more to note here (and in other omitted reduction rules),
-except again for the multiple quantifications over the trivial splittings
-$\Context_i \simeq \AgdaInductiveConstructor{[]} + \Context_i$ and the use of
-\AgdaInductiveConstructor{split-l} to make sure that the channel referred to by
-\AgdaInductiveConstructor{select} and \AgdaInductiveConstructor{case} is indeed
+There is not much more to note here except again for the multiple
+quantifications over the trivial splittings $\Context_i \simeq
+\AgdaInductiveConstructor{[]} + \Context_i$ and the use of
+\AgdaInductiveConstructor{<} to make sure that the channel referred to by
+\AgdaInductiveConstructor{left} and \AgdaInductiveConstructor{case} is indeed
 the one restricted by the cut.
@@
                       (cut (> p) (server (< p₀) un P) Q))
-  r-poly       : ∀{A B Γ₁ Γ₂ P F} (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) ->
+\end{code}
+
+The remaining constructors that describe the base reductions follow a similar
+pattern. We only glance at \RExists since it involves a non-trivial rewriting of
+types:
+
+\begin{code}
+  r-ex         : ∀{A B Γ₁ Γ₂ P F}
+                 (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
                  cut {`∃ A} p (ex {_} {B} (< p₀) P) (all (< q₀) F) ↝ cut p P (F B)
 \end{code}
+
+Recalling the definitions of \AgdaInductiveConstructor{ex} and
+\AgdaInductiveConstructor{all} from \Cref{sec:process-agda}, we note that $P$
+has type
+$\AgdaDatatype{Process}~(\AgdaFunction{subst}~[~B~/\_]~A~\AgdaInductiveConstructor{∷}~\Context_1)$
+and $(F~B)$ is a process of type
+$\AgdaDatatype{Process}~(\AgdaFunction{subst}~[~B~/\_]~(\AgdaFunction{dual}~A)~\AgdaInductiveConstructor{∷}~\Context_2)$.
+In order for these two process to be composable in a cut, it must be the case
+that $\AgdaFunction{subst}~[~B~/\_]~(\AgdaFunction{dual}~A) \equiv
+\AgdaFunction{dual}~(\AgdaFunction{subst}~[~B~/\_]~A)$, which was proved in
+\Cref{sec:type-agda} under the name \AgdaFunction{dual-subst}. Thanks to the
+implicit rewriting rule, we do not have to rewrite the index in the type of
+$(F~B)$, which is silently accepted as is.
 
@@
 \end{code}
-\end{AgdaAlign}

### 2025-12-04 paper/Context.lagda.tex
general cleanup

@@
 
-When splitting a context $\Context$ into $\ContextD + \ContextE$,
-for each type in $\Context$ we use one of the prefix operators \AgdaInductiveConstructor{<\_} and \AgdaInductiveConstructor{>\_} to indicate whether it is meant to be placed in $\ContextD$ or $\ContextE$,
-respectively.
+When splitting a context $\Context$ into $\ContextD + \ContextE$, for each type
+in $\Context$ we use one of the prefix operators \AgdaInductiveConstructor{<}
+and \AgdaInductiveConstructor{>} to indicate whether the type is meant to be
+placed in $\ContextD$ or $\ContextE$, respectively.
 %
-As an example, below is the splitting of the context $[A, B, C, D]$ into $[B] + [A, C, D]$.
+As an example, below is the splitting of the context $[A, B, C, D]$ into $[B] +
+[A, C, D]$.
 
@@
 $\ContextC$ and giving a name $\ContextE$ to the residual context. For this
-particular case, we introduce a dedicated notation that allows us to write this
-case in a more compact and meaningful way, as
-$\SimpleSplitting\ContextC{A}\ContextE$.
+particular case, we introduce the dedicated notation
+$\SimpleSplitting\ContextC{A}\ContextE$ which we find clearer.
 
@@
 it may be convenient to use \AgdaFunction{≫} as a shortcut for a sequence of
-applications of \AgdaInductiveConstructor{>\_} after the $A$ type has been
-reached. For instance, the same sample splitting shown earlier can be written
-equivalently and in a more compact way as shown below. More examples will be
-given in \Cref{sec:examples}.
+applications of \AgdaInductiveConstructor{>} after the $A$ type has been
+reached. For instance, \AgdaFunction{splitting-example₁} can be written
+equivalently and in a more compact way as shown below. More usages of
+\AgdaFunction{≫} will be given in \Cref{sec:examples}.
 
@@
 
-It is a simple fact that the merging of two unrestricted contexts
-$\Context_1$ and $\Context_2$ results in an unrestricted context.\Luca{Serve?}
+% It is a simple fact that the merging of two unrestricted contexts
+% $\Context_1$ and $\Context_2$ results in an unrestricted context.\Luca{Serve?}
 
-\begin{code}
+\begin{code}[hide]
 +-un : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → Un Γ₁ → Un Γ₂ → Un Γ
-\end{code}
-\begin{code}[hide]
 +-un • un-[] un-[] = un-[]

### 2025-12-04 paper/Permutations.lagda.tex
restored description of Process
general cleanup

@@
 enjoys a number of expected properties, such as symmetry. Among them, in the
-following we will make use of a specific one that connects context splitting and
-permutations: if $\Splitting\ContextC\ContextD\ContextE$, then $\ContextC$ is a
-permutation of the concatenation of $\ContextD$ and $\ContextE$.\Luca{Ci sono
-molte altre proprietà non descritte, controllare quelle che servono davvero.}
+following we will make use of a specific property that relates context
+splitting, list concatenation and permutations: if
+$\Splitting\ContextC\ContextD\ContextE$, then $\ContextC$ is a permutation of
+the concatenation of $\ContextD$ and $\ContextE$.
 
@@
 
-↭solo : ∀{A Γ Γ′ Δ} → Γ ↭ Δ → Γ ≃ A , Γ′ → ∃[ Δ′ ] (Δ ≃ A , Δ′ × Γ′ ↭ Δ′)
+↭solo : ∀{A Γ Γ′ Δ} → Γ ↭ Δ → Γ ∋ A ⊳ Γ′ → ∃[ Δ′ ] (Δ ∋ A ⊳ Δ′ × Γ′ ↭ Δ′)
 ↭solo π p with ↭split π p

### 2025-12-04 paper/Reduction.lagda.tex
general cleanup

@@
 The remaining constructors that describe the base reductions follow a similar
-pattern. We only glance at \RExists since it involves a non-trivial rewriting of
-types:
+pattern, noting that the implementation of \RWeaken and \RContract makes use of
+auxiliary functions that respectively weaken and contract the typing context of
+the resulting process as shown in \Cref{tab:semantics}. It is worth glancing at
+\RExists since it involves a non-trivial rewriting of types:
 
 \begin{code}
-  r-ex         : ∀{A B Γ₁ Γ₂ P F}
+  r-exists     : ∀{A B Γ₁ Γ₂ P F}
                  (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →

### 2025-12-04 paper/related.tex
removed comments

\section{Related Work}
\label{sec:related}

come sono stati misurati

eliminati commenti
eliminati esempi
squeezed spazi (in generale sono significativi)

### 2025-12-08 paper/Process.lagda.tex
restructuring presentation with new code

@@
 {-# OPTIONS --rewriting #-}
-open import Data.Bool using (Bool; true; false; if_then_else_)
+open import Data.Unit using (tt)
+open import Data.Sum
 open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
+open import Data.List.Base using (List; []; _∷_; [_]; _++_)
+open import Relation.Unary
 open import Relation.Binary.PropositionalEquality using (_≡_; refl; cong; cong₂)
-open import Data.List.Base using (List; []; _∷_; [_]; _++_)
 
@@
 
-We adopt a \emph{nameless, intrinsically-typed} representation of processes,
-whereby a channel is identified by the position of its type in the typing
-context and a process is also a proof derivation showing that the process is
-well typed. This choice increases the effort in the definition of the datatypes
-for defining processes and their operational semantics, but pays off in the rest
-of the formalization for a number of reasons:
+We adopt a representation of processes that is \emph{intrinsically typed} and
+based on \emph{co-de Bruijn indexes}. The intrinsically typed representation
+makes sure that only well-typed processes can be constructed. This choice
+increases the effort in the definition of the datatypes for representing
+processes and their operational semantics, but pays off in the rest of the
+formalization for at least three reasons:
 \begin{itemize}
-\item we use context splitting not only to separate resources used by parallel
-  processes (\cf \ForkRule and \CutRule), but also to witness that a channel is
-  present in the typing context;
+\item we can adopt a \emph{nameless} representation of channels, thus avoiding
+  all the technicalities and pitfalls that a named representation entails;
 \item we conflate processes and typing rules in the same datatype, thus reducing
   the overall number of datatypes defined in the formalization;
-\item we embed the typing preservation results within the very definition of the
-  operational semantics of processes instead of proving them separately
+\item the typing preservation results are embedded in the very definition of the
+  operational semantics of processes and need not require separate proofs
   (\Cref{sec:agda-semantics}).
@@
 
-Our nameless representation of channels is known in the literature as the
-\emph{co-de Bruijn} representation of identifiers~\cite{McBride18}. The proof
-that the type of a channel is found within a typing context essentially encodes
-the de Bruijn index of the channel in the typing context of a given process.
-However, the typing context only contains types for those channels that actually
-occur in the process. Context splitting makes sure that channels are
-appropriately (and above all linearly) distributed among sub-processes, so that
-each channel occurs (and is used) exactly once.\footnote{Strictly speaking, the
-typing context of a \AgdaInductiveConstructor{fail} process may contain channels
-that do not occur (and therefore are not used) by the process, but we consider
-this case as an exception that does not alter in a substantial way the provided
-intuition underlying the use of co-de Bruijn representation.}
+We use \emph{co-de Bruijn indexes}~\cite{McBride18} to represent channels. Co-de
+Bruijn indexes are akin to de Bruijn indexes, with the key difference that the
+typing context used to type a certain process $P$ is guaranteed to only contain
+the type of those channels that actually occur in $P$. For this reason, typing
+contexts are split as soon as possible, according to the structure of processes,
+to make sure that channels are appropriately (and above all linearly)
+distributed among sub-processes, so that each channel occurs (and is used)
+exactly once.\footnote{Strictly speaking, the typing context of a
+\AgdaInductiveConstructor{fail} process may contain channels that do not occur
+(and therefore are not used) by the process, but we consider this case as an
+exception that does not alter in a substantial way the provided intuition
+underlying the use of co-de Bruijn representation.}
 
-We now illustrate the definition of the \AgdaDatatype{Process} datatype and then
-describe its key aspects in detail. The datatype is indexed by a typing context
-$\Context$ according to which the process is well typed:
+From this informal introduction to the representation of processes it is clear
+that context splitting is going to be a pervasive notion in the subsequent
+development, to the point that it contributes substantially to the complexity of
+the formalization. For this reason, it is worth investing some effort in a
+defining an abstraction that allows us to express the type of processes in a
+more compact and hopefully more readable way.
+\citet{RouvoetPoulsenKrebbersVisser20} observe that both channels and processes
+can be thought of as \emph{predicates on typing contexts}:
+\begin{itemize}
+\item a channel of type $A$ is a predicate that holds for the singleton context
+  $\AgdaFunction[~A~\AgdaFunction]$;
+\item a process that is well typed in a typing context $\Context$ is a predicate
+  that holds for $\Context$.
+\end{itemize}
+
+Inspired by separation logic~\cite{Hoare},
+\citet{RouvoetPoulsenKrebbersVisser20} define the \emph{separating conjunction}
+$P \sand Q$ of two predicates $P$ and $Q$. The separating conjunction of $P$ and
+$Q$ holds for those contexts $\ContextC$ that can be split into $\ContextD$ and
+$\ContextE$ where $P$ holds for $\ContextD$ and $Q$ holds for $\ContextE$. In
+Agda, \citet{RouvoetPoulsenKrebbersVisser20} define $\sand$ as a record, thus:
+\begin{code}
+record _∗_ (P Q : Pred Context _) (Γ : Context) : Set where
+  constructor _⟨_⟩_
+  field
+    {Δ Θ} : Context
+    px    : P Δ
+    p     : Γ ≃ Δ + Θ
+    qx    : Q Θ
+\end{code}
+\begin{code}[hide]
+infixr 9 _∗_
+\end{code}
+
+The record has five fields: the (implicit) fields $\ContextD$ and $\ContextE$,
+which represent the decomposition of $\ContextC$; the fields \AgdaField{px} and
+\AgdaField{qx}, which are proofs witnessing that $P$ and $Q$ hold respectively
+for $\ContextD$ and $\ContextE$; the field \AgdaField{p}, which witnesses that
+$\ContextD$ and $\ContextE$ are indeed a splitting of $\ContextC$. The
+constructor \AgdaInductiveConstructor{$\_\langle\_\rangle\_$} provides a
+convenient syntax for specifying the (explicit) fields of the record to provide
+(and pattern match) separating conjunctions.
+
+To illustrate a typical usage pattern of separating conjunction we first
+introduce our first predicate \AgdaDatatype{Ch} on typing contexts, which we use
+to describe well-typed channels:
 
 \begin{code}
-data Process : Context → Set where
-  link      : ∀{A Γ} → Γ ≃ [ A ] + [ dual A ] → Process Γ
-  fail      : ∀{Γ Δ} → Γ ∋ ⊤ ⊳ Δ → Process Γ
-  wait      : ∀{Γ Δ} → Γ ∋ ⊥ ⊳ Δ → Process Δ → Process Γ
-  close     : Process [ 𝟙 ]
-  case      : ∀{A B Γ Δ} → Γ ∋ A & B ⊳ Δ →
-              Process (A ∷ Δ) → Process (B ∷ Δ) → Process Γ
-  left      : ∀{A B Γ Δ} → Γ ∋ A ⊕ B ⊳ Δ → Process (A ∷ Δ) → Process Γ
-  right     : ∀{A B Γ Δ} → Γ ∋ A ⊕ B ⊳ Δ → Process (B ∷ Δ) → Process Γ
-  join      : ∀{A B Γ Δ} → Γ ∋ A ⅋ B ⊳ Δ → Process (B ∷ A ∷ Δ) → Process Γ
-  fork      : ∀{A B Γ Δ Γ₁ Γ₂} → Γ ∋ A ⊗ B ⊳ Δ → Δ ≃ Γ₁ + Γ₂ →
-              Process (A ∷ Γ₁) → Process (B ∷ Γ₂) → Process Γ
-  all       : ∀{A Γ Δ} → Γ ∋ `∀ A ⊳ Δ →
-              ((X : Type) → Process (subst [ X /_] A ∷ Δ)) → Process Γ
-  ex        : ∀{A B Γ Δ} → Γ ∋ `∃ A ⊳ Δ → Process (subst [ B /_] A ∷ Δ) → Process Γ
-  server    : ∀{A Γ Δ} → Γ ∋ `! A ⊳ Δ → Un Δ → Process (A ∷ Δ) → Process Γ
-  client    : ∀{A Γ Δ} → Γ ∋ `? A ⊳ Δ → Process (A ∷ Δ) → Process Γ
-  weaken    : ∀{A Γ Δ} → Γ ∋ `? A ⊳ Δ → Process Δ → Process Γ
-  contract  : ∀{A Γ Δ} → Γ ∋ `? A ⊳ Δ → Process (`? A ∷ `? A ∷ Δ) → Process Γ
-  cut       : ∀{A Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ →
-              Process (A ∷ Γ₁) → Process (dual A ∷ Γ₂) → Process Γ
+data Ch (A : Type) : Context → Set where
+  ch : Ch A [ A ]
+\end{code}
+
+The predicate $\AgdaDatatype{Ch}~A$ holds for the singleton typing context
+$[A]$. Hence the predicate $\AgdaDatatype{Ch}~A\sand\AgdaDatatype{Ch}~B$ holds
+for those typing contexts that can be split into $[A]$ and $[B]$, as exemplified
+below:
+\begin{code}
+_ : ∀{A B} → (Ch A ∗ Ch B) (A ∷ B ∷ [])
+_ = ch ⟨ < > • ⟩ ch
+
+_ : ∀{A B} → (Ch A ∗ Ch B) (B ∷ A ∷ [])
+_ = ch ⟨ > < • ⟩ ch
+\end{code}
+
+\begin{table}
+  \begin{tabular}{|lll|}
+    \hline
+    \textbf{Notation} & \textbf{Definition} & \textbf{Meaning} \\
+    \hline
+    $\AgdaFunction{∀[}~P~\AgdaFunction]$ & $\forall\{x\} \to P~x$ & implicit universality \\
+    $P~\AgdaFunction{⇒}~Q$ & $\lambda x \to P~x \to Q~x$ & implication \\
+    $P~\AgdaFunction{∪}~Q$ & $\lambda x \to P~x~\AgdaDatatype{⊎}~Q~x$ & disjunction \\
+    $P~\AgdaFunction{∩}~Q$ & $\lambda x \to P~x~\AgdaDatatype{×}~Q~x$ & conjunction (non-separating) \\
+    $f~\AgdaFunction{⊢}~P$ & $\lambda x \to P~(f~x)$ & update \\
+    $\AgdaFunction{U}$ & $\lambda x \to \AgdaModule{Data.Unit}.\AgdaDatatype{⊤}$ & universal
+    set \\
+    $\AgdaFunction{⋂[}~X~\AgdaFunction{:}~A~\AgdaFunction]~P$ & $\lambda x \to (X : A) \to P~X~x$ & infinitary conjunction \\
+    \hline
+  \end{tabular}
+  \caption{\label{tab:predicates} Used definitions from Agda's \AgdaModule{Relation.Unary} module.}
+\end{table}
+
+We now illustrate the datatype \AgdaDatatype{Proc} for representing processes
+and then describe its key aspects in detail. This datatype uses, in addition to
+$\sand$, a number of definitions from the \AgdaModule{Relation.Unary} module in
+Agda's standard library which are worth recalling in \Cref{tab:predicates}.
+
+\begin{code}
+data Proc : Context → Set where
+  link      : ∀{A} → ∀[ Ch A ∗ Ch (dual A) ⇒ Proc ]
+  fail      : ∀[ Ch ⊤ ∗ U ⇒ Proc ]
+  wait      : ∀[ Ch ⊥ ∗ Proc ⇒ Proc ]
+  close     : ∀[ Ch 𝟙 ⇒ Proc ]
+  case      : ∀{A B} →
+              ∀[ Ch (A & B) ∗ ((A ∷_) ⊢ Proc ∩ (B ∷_) ⊢ Proc) ⇒ Proc ]
+  select    : ∀{A B} →
+              ∀[ Ch (A ⊕ B) ∗ ((A ∷_) ⊢ Proc ∪ (B ∷_) ⊢ Proc) ⇒ Proc ]
+  join      : ∀{A B} →
+              ∀[ Ch (A ⅋ B) ∗ ((A ∷_) ⊢ (B ∷_) ⊢ Proc) ⇒ Proc ]
+  fork      : ∀{A B} →
+              ∀[ Ch (A ⊗ B) ∗ ((A ∷_) ⊢ Proc) ∗ ((B ∷_) ⊢ Proc) ⇒ Proc ]
+  all       : ∀{A} →
+              ∀[ Ch (`∀ A) ∗ ⋂[ X ∶ Type ] ((subst [ X /_] A ∷_) ⊢ Proc) ⇒ Proc ]
+  ex        : ∀{A B} → ∀[ Ch (`∃ A) ∗ ((subst [ B /_] A ∷_) ⊢ Proc) ⇒ Proc ]
+  server    : ∀{A} → ∀[ Ch (`! A) ∗ (Un ∩ ((A ∷_) ⊢ Proc)) ⇒ Proc ]
+  client    : ∀{A} → ∀[ Ch (`? A) ∗ ((A ∷_) ⊢ Proc) ⇒ Proc ]
+  weaken    : ∀{A} → ∀[ Ch (`? A) ∗ Proc ⇒ Proc ]
+  contract  : ∀{A} → ∀[ Ch (`? A) ∗ ((`? A ∷_) ⊢ (`? A ∷_) ⊢ Proc) ⇒ Proc ]
+  cut       : ∀{A} → ∀[ ((A ∷_) ⊢ Proc) ∗ ((dual A ∷_) ⊢ Proc) ⇒ Proc ]
 \end{code}
@@
 \AgdaInductiveConstructor{contract}. They all basically follow the same pattern:
-their first (explicit) argument is a splitting of the form
-$\SimpleSplitting\ContextC{C}\ContextD$ that identifies the channel on which the
-process is acting within the typing context $\ContextC$ and where the shape of
-$C$ is determined by the typing rule; the subsequent arguments are terms
-corresponding to the continuations of the process being represented, each
-indexed by its own context. Crucially, the type of the continuation channels are
-\emph{prepended} to the context that indexes the continuation processes using
-the \AgdaInductiveConstructor{\_∷\_} list constructor. This is consistent with
-the fact that the continuation channels have been freshly introduced, and the
+they possibly quantify over some types $A$ and $B$ and then (implicitly) over a
+typing context $\Context$ through the function \AgdaFunction{∀[\_]} applied to a
+predicate of the form $\cdots~\AgdaFunction{⇒}~\AgdaDatatype{Proc}$. The
+predicate states how to build a process of a particular form, provided that the
+typing context $\Context$ satisfies the predicate on the l.h.s. of
+\AgdaFunction{⇒}.
+
+The constructor \AgdaInductiveConstructor{fail} requires the context $\Context$
+to satisfy the predicate
+$\AgdaDatatype{Ch}~\AgdaInductiveConstructor{⊤}~\AgdaDatatype{∗}~\AgdaFunction{U}$,
+meaning that $\Context$ must contain an entry \AgdaInductiveConstructor{⊤}. That
+form of $\Context$ matches the one in the conclusion of \FailRule.
+
+The constructor \AgdaInductiveConstructor{close} requires the typing context to
+be the singleton list $[\One]$.
+
+Crucially, the type of the continuation channels are \emph{prepended} to the
+context that indexes the continuation processes using the
+\AgdaInductiveConstructor{\_∷\_} list constructor. This is consistent with the
+fact that the continuation channels have been freshly introduced, and the
 channel on which the process operates has been consumed.
@@
 There are just a few deviations from this pattern, which we discuss next.
-
-The constructor \AgdaInductiveConstructor{close} lacks a splitting argument:
-since a process $\Close\x$ is well typed in a singleton context $x : \One$, it
-is easier to specify such context directly as the index of
-\AgdaDatatype{Process} rather than through a splitting.
 
@@
 \begin{code}
-↭process : ∀{Γ Δ} → Γ ↭ Δ → Process Γ → Process Δ
+↭proc : ∀{Γ Δ} → Γ ↭ Δ → Proc Γ → Proc Δ
 \end{code}
 \begin{code}[hide]
-↭process π (link p) with ↭solo π p
-... | Δ′ , q , π′ rewrite ↭solo-inv π′ = link q
-↭process π (fail p) with ↭solo π p
-... | Δ′ , q , π′ = fail q
-↭process π (wait p P) with ↭solo π p
-... | Δ′ , q , π′ = wait q (↭process π′ P)
-↭process π close rewrite ↭solo-inv π = close
-↭process π (case p P Q) with ↭solo π p
-... | Δ′ , q , π′ = case q (↭process (prep π′) P) (↭process (prep π′) Q)
-↭process π (left p P) with ↭solo π p
-... | Δ′ , q , π′ = left q (↭process (prep π′) P)
-↭process π (right p P) with ↭solo π p
-... | Δ′ , q , π′ = right q (↭process (prep π′) P)
-↭process π (join p P) with ↭solo π p
-... | Δ′ , q , π′ = join q (↭process (prep (prep π′)) P)
-↭process π (fork p q P Q) with ↭solo π p
-... | Δ′ , p′ , π′ with ↭split π′ q
-... | Δ₁ , Δ₂ , q′ , π₁ , π₂ = fork p′ q′ (↭process (prep π₁) P) (↭process (prep π₂) Q)
-↭process π (all p P) with ↭solo π p
-... | Δ' , q , π' = all q λ B → ↭process (prep π') (P B)
-↭process π (ex p P) with ↭solo π p
-... | Δ' , q , π' = ex q (↭process (prep π') P)
-↭process π (server p un P) with ↭solo π p
-... | Δ′ , q , π′ = server q (↭un π′ un) (↭process (prep π′) P)
-↭process π (client p P) with ↭solo π p
-... | Δ′ , q , π′ = client q (↭process (prep π′) P)
-↭process π (weaken p P) with ↭solo π p
-... | Δ′ , q , π′ = weaken q (↭process π′ P)
-↭process π (contract p P) with ↭solo π p
-... | Δ′ , q , π′ = contract q (↭process (prep (prep π′)) P)
-↭process π (cut p P Q) with ↭split π p
-... | Δ₁ , Δ₂ , q , π₁ , π₂ = cut q (↭process (prep π₁) P) (↭process (prep π₂) Q)
+↭proc π (link (ch ⟨ p ⟩ ch)) with ↭solo π p
+... | _ , q , π' rewrite ↭solo-inv π' = link (ch ⟨ q ⟩ ch)
+↭proc π (fail (ch ⟨ p ⟩ tt)) with ↭solo π p
+... | _ , q , π' = fail (ch ⟨ q ⟩ tt)
+↭proc π (wait (ch ⟨ p ⟩ P)) with ↭solo π p
+... | _ , q , π' = wait (ch ⟨ q ⟩ ↭proc π' P)
+↭proc π (close ch) rewrite ↭solo-inv π = close ch
+↭proc π (case (ch ⟨ p ⟩ (P , Q))) with ↭solo π p
+... | _ , q , π' = case (ch ⟨ q ⟩ (↭proc (prep π') P , ↭proc (prep π') Q))
+↭proc π (select (ch ⟨ p ⟩ inj₁ P)) with ↭solo π p
+... | _ , q , π' = select (ch ⟨ q ⟩ inj₁ (↭proc (prep π') P))
+↭proc π (select (ch ⟨ p ⟩ inj₂ P)) with ↭solo π p
+... | _ , q , π' = select (ch ⟨ q ⟩ inj₂ (↭proc (prep π') P))
+↭proc π (join (ch ⟨ p ⟩ P)) with ↭solo π p
+... | _ , q , π' = join (ch ⟨ q ⟩ ↭proc (prep (prep π')) P)
+↭proc π (fork (ch ⟨ p ⟩ (P ⟨ q ⟩ Q))) with ↭solo π p
+... | _ , p' , π' with ↭split π' q
+... | Δ₁ , Δ₂ , q' , π₁ , π₂ = fork (ch ⟨ p' ⟩ (↭proc (prep π₁) P ⟨ q' ⟩ ↭proc (prep π₂) Q))
+↭proc π (all (ch ⟨ p ⟩ F)) with ↭solo π p
+... | _ , q , π' = all (ch ⟨ q ⟩ λ X → ↭proc (prep π') (F X))
+↭proc π (ex (ch ⟨ p ⟩ P)) with ↭solo π p
+... | _ , q , π' = ex (ch ⟨ q ⟩ ↭proc (prep π') P)
+↭proc π (server (ch ⟨ p ⟩ (un , P))) with ↭solo π p
+... | _ , q , π' = server (ch ⟨ q ⟩ (↭un π' un , ↭proc (prep π') P))
+↭proc π (client (ch ⟨ p ⟩ P)) with ↭solo π p
+... | _ , q , π' = client (ch ⟨ q ⟩ ↭proc (prep π') P)
+↭proc π (weaken (ch ⟨ p ⟩ P)) with ↭solo π p
+... | _ , q , π' = weaken (ch ⟨ q ⟩ ↭proc π' P)
+↭proc π (contract (ch ⟨ p ⟩ P)) with ↭solo π p
+... | _ , q , π' = contract (ch ⟨ q ⟩ ↭proc (prep (prep π')) P)
+↭proc π (cut (P ⟨ p ⟩ Q)) with ↭split π p
+... | Δ₁ , Δ₂ , q , π₁ , π₂ = cut (↭proc (prep π₁) P ⟨ q ⟩ ↭proc (prep π₂) Q)
 \end{code}

### 2025-12-11 paper/Context.lagda.tex
snapshot

@@
 open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
-open import Relation.Binary.PropositionalEquality using (_≡_; refl; cong; cong₂)
-open import Data.List.Base using (List; []; _∷_; [_]; _++_; map)
+open import Data.List.Base using (List; []; _∷_; [_]; _++_)
+open import Relation.Unary
+open import Relation.Binary.PropositionalEquality using (_≡_; refl; cong)
 
@@
 We are going to adopt a nameless representation of channels whereby a channel is
-identified by its position in a typing context. This representation is akin to
-using De Bruijn indices, except that the index, instead of being represented
-explicitly by a natural number, is computable from the \emph{proof} that the
-type of the channel belong to the typing context.
+identified by the position of its type in a typing context. This representation
+is akin to using De Bruijn indices except that the index, instead of being
+represented explicitly by a natural number, is computable from the proof that
+the type of the channel belongs to the typing context.
 %
-Typing contexts are represented using \emph{lists} of types, where the
-(polymorphic) \AgdaDatatype{List} type is defined in Agda's standard library.
+Typing contexts are represented using lists of types, where the (polymorphic)
+\AgdaDatatype{List} type and its constructors \AgdaInductiveConstructor{[]} and
+\AgdaInductiveConstructor{\_∷\_} are defined in the module
+\AgdaModule{Data.List} of Agda's standard library. We will keep using
+$\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts, even
+though in the Agda formalization they are lists instead of finite maps.
 
@@
 The most important operation concerning typing contexts is \emph{splitting}. The
-splitting of $\ContextC$ into $\ContextD$ and $\ContextE$, which is denoted by
+splitting of $\ContextC$ into $\ContextD$ and $\ContextE$, which we denote by
 $\Splitting\ContextC\ContextD\ContextE$, represents the fact that $\ContextC$
@@
 \begin{code}[hide]
-infix 4 _≃_+_ _∋_⊳_
+infix  4 _≃_+_
+infixr 8 _─∗_
+infixr 9 _∗_
 \end{code}
@@
 and \AgdaInductiveConstructor{>} to indicate whether the type is meant to be
-placed in $\ContextD$ or $\ContextE$, respectively.
+placed in $\ContextD$ or $\ContextE$.
 %
-As an example, below is the splitting of the context $[A, B, C, D]$ into $[B] +
-[A, C, D]$.
+For example, below is a proof of the splitting $\Splitting{[A, B, C,
+D]}{[B]}{[A, C, D]}$.
 
@@
 Context splitting enjoys a number of expected properties. In particular, it is
-easy to see that splitting is commutative and that the empty context is both a
-left and right unit of splitting.
+easy to see that splitting is commutative and that the empty context/list is
+both a left and right unit of splitting.
 
@@
 
-It is often the case that the context $\ContextD$ in a splitting
-$\Splitting\ContextC\ContextD\ContextE$ is a singleton list $[A]$. We use this
-particular splitting to describe the presence of a channel of type $A$ in
-$\ContextC$ and giving a name $\ContextE$ to the residual context. For this
-particular case, we introduce the dedicated notation
-$\SimpleSplitting\ContextC{A}\ContextE$ which we find clearer.
+% It is often the case that the context $\ContextD$ in a splitting
+% $\Splitting\ContextC\ContextD\ContextE$ is a singleton list $[A]$. We use this
+% particular splitting to describe the presence of a channel of type $A$ in
+% $\ContextC$ and giving a name $\ContextE$ to the residual context. For this
+% particular case, we introduce the dedicated notation
+% $\SimpleSplitting\ContextC{A}\ContextE$ which we find clearer.
 
-\begin{code}
-_∋_⊳_ : Context → Type → Context → Set
-Γ ∋ A ⊳ Δ = Γ ≃ [ A ] + Δ
-\end{code}
+% \begin{code}
+% _∋_⊳_ : Context → Type → Context → Set
+% Γ ∋ A ⊳ Δ = Γ ≃ [ A ] + Δ
+% \end{code}
 \begin{code}[hide]
@@
 
-Note that, when specifying a splitting $\SimpleSplitting\ContextC{A}\ContextE$,
-it may be convenient to use \AgdaFunction{≫} as a shortcut for a sequence of
+Note that, when specifying a splitting $\Splitting\ContextC{[A]}\ContextE$, it
+may be convenient to use \AgdaFunction{≫} as a shortcut for a sequence of
 applications of \AgdaInductiveConstructor{>} after the $A$ type has been
@@
 \begin{code}
-  splitting-example₂ : (A ∷ B ∷ C ∷ D ∷ []) ∋ B ⊳ (A ∷ C ∷ D ∷ [])
+  splitting-example₂ : (A ∷ B ∷ C ∷ D ∷ []) ≃ [ B ] + (A ∷ C ∷ D ∷ [])
   splitting-example₂ = > < ≫
@@
 
-To conclude the modeling of typing contexts, we define a predicate $\AgdaDatatype{Un}~\Context$ that holds
-whenever $\Context$ is \emph{unrestricted}, namely when it is made of types of
-the form $?A$. We will need this predicate in the definition of a server, which
-must obey the typing rule \ServerRule of \Cref{tab:typing-rules}.
+\begin{table}
+  \begin{tabular}{|lll|}
+    \hline
+    \textbf{Notation} & \textbf{Definition} & \textbf{Meaning} \\
+    \hline
+    $\AgdaFunction{∀[}~P~\AgdaFunction]$ & $\forall\{x\} \to P~x$ & implicit universality \\
+    $P~\AgdaFunction{⇒}~Q$ & $\lambda x \to P~x \to Q~x$ & implication \\
+    $P~\AgdaFunction{∪}~Q$ & $\lambda x \to P~x~\AgdaDatatype{⊎}~Q~x$ & disjunction \\
+    $P~\AgdaFunction{∩}~Q$ & $\lambda x \to P~x~\AgdaDatatype{×}~Q~x$ & conjunction \\
+    $f~\AgdaFunction{⊢}~P$ & $\lambda x \to P~(f~x)$ & update \\
+    $\AgdaFunction{U}$ & $\lambda x \to \AgdaModule{Data.Unit}.\AgdaDatatype{⊤}$ & universal
+    set \\
+    $\AgdaFunction{⋂[}~X~\AgdaFunction{:}~A~\AgdaFunction]~P$ & $\lambda x \to (X : A) \to P~X~x$ & infinitary conjunction \\
+    \hline
+  \end{tabular}
+  \caption{\label{tab:predicates} Useful definitions from Agda's \AgdaModule{Relation.Unary} module.}
+\end{table}
+
+In the following we will make extensive use of predicates over contexts. For
+this reason, it is worth recalling a number of definitions from the
+\AgdaModule{Relation.Unary} module in Agda's standard library in
+\Cref{tab:predicates}.
+
+Context splitting is going to be pervasive in the subsequent development, to the
+point that it contributes substantially to the complexity of the formalization.
+For this reason, we invest some effort in defining a few abstractions that allow
+us to occasionally hide context splittings and enhance code readability.
+%
+Inspired by separation logic~\cite{Hoare} and
+following~\citet{RouvoetPoulsenKrebbersVisser20}, we define the \emph{separating
+conjunction} $P \sand Q$ of two predicates $P$ and $Q$ over contexts:
+\begin{code}
+data _∗_ (P Q : Pred Context _) (Γ : Context) : Set where
+  _⟨_⟩_ : ∀{Δ Θ} → P Δ → Γ ≃ Δ + Θ → Q Θ → (P ∗ Q) Γ
+\end{code}
+
+The predicate $P \sand Q$ holds for those contexts $\ContextC$ that can be split
+into $\ContextD$ and $\ContextE$ where $P$ holds for $\ContextD$ and $Q$ holds
+for $\ContextE$.
+%
+The constructor \AgdaInductiveConstructor{\_⟨\_⟩\_} has three explicit arguments
+witnessing the splitting $\Splitting\ContextC\ContextD\ContextE$ along with
+$P~\ContextD$ and $Q~\ContextE$.\Luca{Nei related fare presente che queste
+definizioni seguono sostanzialmente \citet{RouvoetPoulsenKrebbersVisser20} ed
+evidenziare le differenze.}
+
+Along with \AgdaDatatype{∗} we define the \emph{separating implication} (also
+known as ``magic wand'')
+\begin{code}
+_─∗_ : Pred Context _ → Pred Context _ → Context → Set
+(P ─∗ Q) Δ = ∀{Θ Γ} → Γ ≃ Δ + Θ → P Θ → Q Γ
+\end{code}
+and prove that \AgdaFunction{─∗} can be used to curry \AgdaDatatype{∗}:
+\begin{code}
+curry∗ : ∀{P Q R} → ∀[ P ∗ Q ⇒ R ] → ∀[ P ⇒ Q ─∗ R ]
+curry∗ F px σ qx = F (px ⟨ σ ⟩ qx)
+\end{code}
+
+To conclude the modeling of typing contexts, we define a predicate
+\AgdaDatatype{Un} that holds for \emph{unrestricted} contexts, namely those
+contexts solely made of types of the form $?A$. We will need this predicate in
+the definition of a server, which must obey the typing rule \ServerRule of
+\Cref{tab:typing-rules}.
 
@@
   un-[]  : Un []
-  un-∷   : ∀{A Γ} → Un Γ → Un (`? A ∷ Γ)
+  un-∷   : ∀{A} → ∀[ Un ⇒ (`? A ∷_) ⊢ Un ]
 \end{code}
 
-% It is a simple fact that the merging of two unrestricted contexts
-% $\Context_1$ and $\Context_2$ results in an unrestricted context.\Luca{Serve?}
+It is easy to prove that $\Splitting\ContextC\ContextD\ContextE$ is unrestricted
+if so are $\ContextD$ and $\ContextE$:
 
+\begin{code}
+∗-un : ∀[ Un ∗ Un ⇒ Un ]
+\end{code}
 \begin{code}[hide]
-+-un : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → Un Γ₁ → Un Γ₂ → Un Γ
-+-un • un-[] un-[] = un-[]
-+-un (< p) (un-∷ un₁) un₂ = un-∷ (+-un p un₁ un₂)
-+-un (> p) un₁ (un-∷ un₂) = un-∷ (+-un p un₁ un₂)
+∗-un (un-[] ⟨ • ⟩ un-[])    = un-[]
+∗-un (un-∷ un ⟨ < σ ⟩ un′)  = un-∷ (∗-un (un ⟨ σ ⟩ un′))
+∗-un (un ⟨ > σ ⟩ un-∷ un′)  = un-∷ (∗-un (un ⟨ σ ⟩ un′))
 \end{code}

### 2025-12-11 paper/Permutations.lagda.tex
snapshot

@@
 
-The definition of \AgdaDatatype{\_↭\_} data type is nearly the same found in
+The definition of \AgdaDatatype{\_↭\_} data type is nearly the same found in the
+module \AgdaModule{Data.List.Relation.Binary.Permutation.Propositional} of
 Agda's standard library. We have preferred defining our own notion of
@@
 ↭solo-inv (trans π π′) rewrite ↭solo-inv π | ↭solo-inv π′ = refl
-
-↭rot : ∀{A B C Γ} → (A ∷ B ∷ C ∷ Γ) ↭ (C ∷ A ∷ B ∷ Γ)
-↭rot = trans (prep swap) swap
 \end{code}
 
-Just like permutations from Agda's standard library, also \AgdaDatatype{\_↭\_}
+Just like permutations in Agda's standard library, also \AgdaDatatype{\_↭\_}
 enjoys a number of expected properties, such as symmetry. Among them, in the
@@
 
-↭solo : ∀{A Γ Γ′ Δ} → Γ ↭ Δ → Γ ∋ A ⊳ Γ′ → ∃[ Δ′ ] (Δ ∋ A ⊳ Δ′ × Γ′ ↭ Δ′)
+↭solo : ∀{A Γ Γ′ Δ} → Γ ↭ Δ → Γ ≃ [ A ] + Γ′ → ∃[ Δ′ ] (Δ ≃ [ A ] + Δ′ × Γ′ ↭ Δ′)
 ↭solo π p with ↭split π p

### 2025-12-11 paper/Process.lagda.tex
snapshot

@@
 open import Data.Sum
-open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
-open import Data.List.Base using (List; []; _∷_; [_]; _++_)
+open import Data.Product using (_,_)
+open import Data.List.Base using ([]; _∷_; [_])
 open import Relation.Unary
-open import Relation.Binary.PropositionalEquality using (_≡_; refl; cong; cong₂)
 
@@
 
-We adopt a representation of processes that is \emph{intrinsically typed} and
-based on \emph{co-de Bruijn indexes}. The intrinsically typed representation
-makes sure that only well-typed processes can be constructed. This choice
-increases the effort in the definition of the datatypes for representing
-processes and their operational semantics, but pays off in the rest of the
-formalization for at least three reasons:
+We adopt an \emph{intrinsically-typed} representation of processes with
+\emph{nameless} channels. The intrinsically-typed representation makes sure that
+only well-typed processes can be constructed. This choice increases the effort
+in the definition of the datatypes for representing processes and their
+operational semantics, but pays off in the rest of the formalization for at
+least three reasons:
 \begin{itemize}
-\item we can adopt a \emph{nameless} representation of channels, thus avoiding
-  all the technicalities and pitfalls that a named representation entails;
+\item we need not give explicit names to channels, thus we avoid all the
+  technicalities and pitfalls that a named representation entails;
 \item we conflate processes and typing rules in the same datatype, thus reducing
@@
 
-We use \emph{co-de Bruijn indexes}~\cite{McBride18} to represent channels. Co-de
-Bruijn indexes are akin to de Bruijn indexes, with the key difference that the
-typing context used to type a certain process $P$ is guaranteed to only contain
-the type of those channels that actually occur in $P$. For this reason, typing
+Channels are not given any name. Instead, they are represented as terms
+witnessing that their type is present in the typing context. This is known as
+\emph{co-de Bruijn syntax}~\cite{McBride18}, according to which the typing
+context associated with a term (a process, a channel) only contains the types of
+the channels that actually occur within the term. For this reason, typing
 contexts are split as soon as possible, according to the structure of processes,
@@
 underlying the use of co-de Bruijn representation.}
-
-From this informal introduction to the representation of processes it is clear
-that context splitting is going to be a pervasive notion in the subsequent
-development, to the point that it contributes substantially to the complexity of
-the formalization. For this reason, it is worth investing some effort in a
-defining an abstraction that allows us to express the type of processes in a
-more compact and hopefully more readable way.
-\citet{RouvoetPoulsenKrebbersVisser20} observe that both channels and processes
-can be thought of as \emph{predicates on typing contexts}:
-\begin{itemize}
-\item a channel of type $A$ is a predicate that holds for the singleton context
-  $\AgdaFunction[~A~\AgdaFunction]$;
-\item a process that is well typed in a typing context $\Context$ is a predicate
-  that holds for $\Context$.
-\end{itemize}
-
-Inspired by separation logic~\cite{Hoare},
-\citet{RouvoetPoulsenKrebbersVisser20} define the \emph{separating conjunction}
-$P \sand Q$ of two predicates $P$ and $Q$. The separating conjunction of $P$ and
-$Q$ holds for those contexts $\ContextC$ that can be split into $\ContextD$ and
-$\ContextE$ where $P$ holds for $\ContextD$ and $Q$ holds for $\ContextE$. In
-Agda, \citet{RouvoetPoulsenKrebbersVisser20} define $\sand$ as a record, thus:
-\begin{code}
-record _∗_ (P Q : Pred Context _) (Γ : Context) : Set where
-  constructor _⟨_⟩_
-  field
-    {Δ Θ} : Context
-    px    : P Δ
-    p     : Γ ≃ Δ + Θ
-    qx    : Q Θ
-\end{code}
-\begin{code}[hide]
-infixr 9 _∗_
-\end{code}
-
-The record has five fields: the (implicit) fields $\ContextD$ and $\ContextE$,
-which represent the decomposition of $\ContextC$; the fields \AgdaField{px} and
-\AgdaField{qx}, which are proofs witnessing that $P$ and $Q$ hold respectively
-for $\ContextD$ and $\ContextE$; the field \AgdaField{p}, which witnesses that
-$\ContextD$ and $\ContextE$ are indeed a splitting of $\ContextC$. The
-constructor \AgdaInductiveConstructor{$\_\langle\_\rangle\_$} provides a
-convenient syntax for specifying the (explicit) fields of the record to provide
-(and pattern match) separating conjunctions.
-
-To illustrate a typical usage pattern of separating conjunction we first
-introduce our first predicate \AgdaDatatype{Ch} on typing contexts, which we use
-to describe well-typed channels:
+%
+Concretely, a channel of type $A$ is a predicate that holds for the singleton
+context $\AgdaFunction[~A~\AgdaFunction]$:
 
@@
 
-The predicate $\AgdaDatatype{Ch}~A$ holds for the singleton typing context
-$[A]$. Hence the predicate $\AgdaDatatype{Ch}~A\sand\AgdaDatatype{Ch}~B$ holds
-for those typing contexts that can be split into $[A]$ and $[B]$, as exemplified
-below:
-\begin{code}
-_ : ∀{A B} → (Ch A ∗ Ch B) (A ∷ B ∷ [])
-_ = ch ⟨ < > • ⟩ ch
-
-_ : ∀{A B} → (Ch A ∗ Ch B) (B ∷ A ∷ [])
-_ = ch ⟨ > < • ⟩ ch
-\end{code}
-
-\begin{table}
-  \begin{tabular}{|lll|}
-    \hline
-    \textbf{Notation} & \textbf{Definition} & \textbf{Meaning} \\
-    \hline
-    $\AgdaFunction{∀[}~P~\AgdaFunction]$ & $\forall\{x\} \to P~x$ & implicit universality \\
-    $P~\AgdaFunction{⇒}~Q$ & $\lambda x \to P~x \to Q~x$ & implication \\
-    $P~\AgdaFunction{∪}~Q$ & $\lambda x \to P~x~\AgdaDatatype{⊎}~Q~x$ & disjunction \\
-    $P~\AgdaFunction{∩}~Q$ & $\lambda x \to P~x~\AgdaDatatype{×}~Q~x$ & conjunction (non-separating) \\
-    $f~\AgdaFunction{⊢}~P$ & $\lambda x \to P~(f~x)$ & update \\
-    $\AgdaFunction{U}$ & $\lambda x \to \AgdaModule{Data.Unit}.\AgdaDatatype{⊤}$ & universal
-    set \\
-    $\AgdaFunction{⋂[}~X~\AgdaFunction{:}~A~\AgdaFunction]~P$ & $\lambda x \to (X : A) \to P~X~x$ & infinitary conjunction \\
-    \hline
-  \end{tabular}
-  \caption{\label{tab:predicates} Used definitions from Agda's \AgdaModule{Relation.Unary} module.}
-\end{table}
-
-We now illustrate the datatype \AgdaDatatype{Proc} for representing processes
-and then describe its key aspects in detail. This datatype uses, in addition to
-$\sand$, a number of definitions from the \AgdaModule{Relation.Unary} module in
-Agda's standard library which are worth recalling in \Cref{tab:predicates}.
+A process that is well typed in a typing context $\Context$ is a predicate that
+holds for $\Context$. Here is the datatype \AgdaDatatype{Proc} for representing
+processes:
 
@@
               ∀[ Ch (A ⊕ B) ∗ ((A ∷_) ⊢ Proc ∪ (B ∷_) ⊢ Proc) ⇒ Proc ]
-  join      : ∀{A B} →
-              ∀[ Ch (A ⅋ B) ∗ ((A ∷_) ⊢ (B ∷_) ⊢ Proc) ⇒ Proc ]
+  join      : ∀{A B} → ∀[ Ch (A ⅋ B) ∗ ((A ∷_) ⊢ (B ∷_) ⊢ Proc) ⇒ Proc ]
   fork      : ∀{A B} →
@@
 
-Let us start from the constructors representing input/output actions, namely
-those from \AgdaInductiveConstructor{fail} through
-\AgdaInductiveConstructor{contract}. They all basically follow the same pattern:
-they possibly quantify over some types $A$ and $B$ and then (implicitly) over a
-typing context $\Context$ through the function \AgdaFunction{∀[\_]} applied to a
-predicate of the form $\cdots~\AgdaFunction{⇒}~\AgdaDatatype{Proc}$. The
-predicate states how to build a process of a particular form, provided that the
-typing context $\Context$ satisfies the predicate on the l.h.s. of
-\AgdaFunction{⇒}.
+All the constructors representing input/output actions, namely those from
+\AgdaInductiveConstructor{fail} through \AgdaInductiveConstructor{contract},
+basically follow the same pattern: they possibly quantify over some types $A$
+and $B$ and then (implicitly) over a typing context $\Context$ through the
+function \AgdaFunction{∀[\_]} applied to a predicate of the form
+$P~\AgdaFunction{⇒}~\AgdaDatatype{Proc}$. The predicate states how to build a
+process that is well typed in $\Context$, provided that $\Context$ satisfies
+$P$. In general, $P$ is a (separating) conjunction of sub-predicates
+corresponding to the channel on which the process is acting and to the premises
+of its typing rule.
 
@@
 
-The constructor \AgdaInductiveConstructor{close} requires the typing context to
-be the singleton list $[\One]$.
+The constructor \AgdaInductiveConstructor{wait} corresponds to a process of the
+form $\Wait\x.P$ and requires $\Context$ to (separately) satisfy
+$\AgdaDatatype{Ch}~\AgdaInductiveConstructor{⊥}$, that is the channel $x$ on
+which the process is operating, as well as \AgdaDatatype{Proc}, that is the
+continuation process $P$, which must be well typed in the remaining typing
+context.
+
+The constructor \AgdaInductiveConstructor{close}, which corresponds to a process
+of the form $\Close\x$, requires the typing context to be the singleton list
+$[\One]$.
 

### 2025-12-11 paper/Type.lagda.tex
snapshot

@@
 
-At this stage we are finally ready to define the type of closed pre-types, those
+At last we are ready to define the type of closed pre-types, those
 having no free type variables. We call \AgdaDatatype{Type} closed pre-types.
-Hereafter, we will seldom use pre-types again.
+From now on, we will seldom use pre-types again.
 

### 2025-12-11 paper/Process.lagda.tex
revised description of processes

@@
 
-\subsection{Process Representation}
+\subsection{Channel and Process Representation}
 \label{sec:process-agda}
@@
 
-The constructor \AgdaInductiveConstructor{fail} requires the context $\Context$
-to satisfy the predicate
+For example, the constructor \AgdaInductiveConstructor{fail}, which corresponds
+to a process of the form $\Fail\x$, requires the context $\Context$ to satisfy
+the predicate
 $\AgdaDatatype{Ch}~\AgdaInductiveConstructor{⊤}~\AgdaDatatype{∗}~\AgdaFunction{U}$,
@@
 
-Crucially, the type of the continuation channels are \emph{prepended} to the
-context that indexes the continuation processes using the
-\AgdaInductiveConstructor{\_∷\_} list constructor. This is consistent with the
-fact that the continuation channels have been freshly introduced, and the
-channel on which the process operates has been consumed.
+Let us move on to the processes that produce a continuation channel. For
+example, the constructor \AgdaInductiveConstructor{case} corresponds to a
+process of the form $\Case\x\y{P}{Q}$, where both $P$ and $Q$ may use the
+continuation channel $y$. In this case $\Context$ must satisfy the predicate
+\[
+  \AgdaDatatype{Ch}~(A~\AgdaInductiveConstructor{\&}~B)~\AgdaDatatype{∗}~((A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}~\AgdaFunction{∩}~(B~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc})
+\]
+which looks intimidating at first but makes perfect sense once we recall
+\CaseRule. Remember that we are trying to establish whether $\Case\x\y{P}{Q}$ is
+well typed in $\Context$.
 %
-There are just a few deviations from this pattern, which we discuss next.
+The predicate $\AgdaDatatype{Ch}~(A~\AgdaInductiveConstructor{\&}~B)$ expresses
+the requirement that the type of $x$ must be of the form $A \With B$ and should
+be found in $\Context$. In other words, it must be the case that
+$\Splitting\Context{[A \With B]}\ContextD$ for some $\ContextD$. Now $P$ and $Q$
+must be well typed in the context $\ContextD$ augmented with the association $y
+: A$ and $y : B$, respectively, whence the use of \AgdaFunction{∩} to verify a
+(non-separating) conjunction of the predicates
+$(A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}$ and
+$(B~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}$. The
+two new contexts are obtained by \emph{adding} either $A$ or $B$ to $\ContextD$,
+which we perform using \AgdaFunction{⊢}.
+%
+Crucially, the types $A$ and $B$ are \emph{prepended} to $\ContextD$, which is
+consistent with the fact that $y$ has been freshly introduced and $x$ has been
+consumed. Notice how easy it is to \emph{prepend} either $A$ or $B$ to
+$\ContextD$ instead of \emph{changing} the type of $x$ in $\Context$ from $A
+\With B$ to either $A$ or $B$ while preserving the position of the type, as we
+would have to do in a ``true'' session type system without explicit continuation
+channels.
 
-The constructor \AgdaInductiveConstructor{fork}, which represents a process
-$\Fork\x\y\z{P}{Q}$, has an additional splitting argument
-$\Splitting\ContextD{\Context_1}{\Context_2}$ which distributes the resources in
-$\ContextD$ among the two continuation processes $P$ and $Q$.
+The interpretation of the remaining constructors is analogous, so we only
+comment \AgdaInductiveConstructor{all}, which represents a process
+$\Forall\x\X\y.P$. This constructor models the continuation $P$ using
+higher-order abstract syntax (HOAS~\cite{}): witnessing that a context
+$\Context$ satisfies the predicate
+\[
+  \AgdaFunction{⋂[}~X~\AgdaFunction{∶}~\AgdaDatatype{Type}~\AgdaFunction]~((\AgdaFunction{subst~[}~X~\AgdaFunction{/\_]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
+\]
+means providing a function that, for every type $X$, produces a witness for the
+predicate
+\[
+  ((\AgdaFunction{subst~[}~X~\AgdaFunction{/\_]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
+\]
+applied to $\Context$. Note that the side condition $X \not\in \fv\Context$ of
+\ForallRule is enforced \emph{by definition}:
+%
+$A$ has type $\AgdaDatatype{PreType}~\AgdaInductiveConstructor{1}$, that is a
+pre-type with \emph{at most one} free type variable ($X$), whereas $\Context$ is
+a typing context, that is a list of $\AgdaDatatype{Type} =
+\AgdaDatatype{PreType}~\AgdaInductiveConstructor{0}$ without free type
+variables. Therefore, $X$ cannot occur in $\Context$.
 
-The constructor \AgdaInductiveConstructor{all}, which represents a process
-$\Forall\x\X\y.P$, models the continuation $P$ using higher-order abstract
-syntax (HOAS~\cite{}): the continuation $P$ is produced by a function applied to
-a \AgdaInductiveConstructor{Type} $X$. Notice how the constructor enforces the
-side condition of \ForallRule \emph{by definition}:
-%
-$A$ has type
-$\AgdaDatatype{PreType}~(\AgdaInductiveConstructor{suc}~\AgdaInductiveConstructor{zero})$,
-that is a pre-type with \emph{at most one} free type variable (that is $X$),
-whereas $\ContextD$ is a list of $\AgdaDatatype{Type} =
-\AgdaDatatype{PreType}~\AgdaInductiveConstructor{zero}$, that is a list of
-pre-types without free type variables. Therefore, $X$ cannot occur in
-$\ContextD$ as required by \ForallRule.
-
-Finally, the constructor \AgdaInductiveConstructor{server}, which represents a
-process $\Server\x\y.P$, has an additional argument of type
-$\AgdaDatatype{Un}~\ContextD$ enforcing the property that $\ContextD$ is
-unrestricted.
-
-The two remaining constructors, \AgdaInductiveConstructor{link} and
-\AgdaInductiveConstructor{cut}, deal with links and cuts in the expected way. A
-link is well typed in a context
-$\Splitting\Context{[A]}{[\AgdaFunction{dual}~A]}$. We use a splitting instead
-of requiring $\Context$ to be $[A,\AgdaFunction{dual}~A]$ so that it becomes
-straightforward to define \SLink using \AgdaFunction{+-comm}.
-%
-The constructor \AgdaInductiveConstructor{cut} models a process $\Cut\x{P}{Q}$
-and uses splitting to distribute the resources in $\Context$ among $P$ and $Q$.
-
-Permutations preserve process typeability. In a sense we can read this property
-as the fact that we can rename channels at will.
+We conclude this module proving that permutations preserve process typeability.
+Since list permutations basically correspond to channel renaming, we can read
+this property as the fact that typeability is preserved by (injective) name
+substitutions.\Luca{Abbiamo detto da qualche parte dove emergono gli indici di
+de Bruijn e di fare attenzione al fatto che lo splitting li altera? Serve questa
+digressione?}\Luca{Da qualche parte mettere un remark per far notare che l'uso
+della separating conjunction introduce rumore, ma avere i costruttori
+currificati ne complica un po' l'interpretazione dei tipi. Forse è meglio fare
+questo rilievo nella prossima sezione quando è evidente il pattern matching che
+serve per esprimere le regole.}
 

### 2025-12-12 paper/Congruence.lagda.tex
revised congruence

@@
 {-# OPTIONS --rewriting #-}
-open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
-open import Data.List.Base using (List; []; _∷_; [_]; _++_)
+open import Data.Unit using (tt)
+open import Data.Sum using (inj₁; inj₂)
+open import Data.Product using (_,_)
+open import Data.List.Base using ([]; _∷_; [_])
 
@@
 \begin{code}
-data _⊒_ {Γ} : Process Γ → Process Γ → Set where
+data _⊒_ {Γ} : Proc Γ → Proc Γ → Set where
 \end{code}
@@
     ∀{A Γ₁ Γ₂ P Q} (p : Γ ≃ Γ₁ + Γ₂) →
-    cut {A} p P Q ⊒ cut (+-comm p) Q P
+    cut {A} (P ⟨ p ⟩ Q) ⊒ cut (Q ⟨ +-comm p ⟩ P)
 \end{code}
@@
 property of parallel composition. We use \AgdaFunction{+-comm} to compute the
-proof of the splitting $\Splitting\ContextC{\Context_2}{\Context_1}$ from $p$
-respectively.
+proof of the splitting $\Splitting\ContextC{\Context_2}{\Context_1}$ from $p$.
 %
@@
   s-link :
-    ∀{A} (p : Γ ≃ [ A ] + [ dual A ]) → link p ⊒ link (+-comm p)
+    ∀{A} (p : Γ ≃ [ A ] + [ dual A ]) →
+    link (ch ⟨ p ⟩ ch) ⊒ link (ch ⟨ +-comm p ⟩ ch)
   s-fail :
-    ∀{A Γ₁ Γ₂ Δ P} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ ⊤ ⊳ Δ) →
+    ∀{A Γ₁ Γ₂ Δ P} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ ⊤ ] + Δ) →
     let _ , _ , q′ = +-assoc-l p q in
-    cut {A} p (fail (> q)) P ⊒ fail q′
+    cut {A} (fail (ch ⟨ > q ⟩ tt) ⟨ p ⟩ P) ⊒ fail (ch ⟨ q′ ⟩ tt)
 \end{code}
@@
   s-wait :
-    ∀{Γ₁ Γ₂ Δ A P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ ⊥ ⊳ Δ) →
+    ∀{Γ₁ Γ₂ Δ A P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ ⊥ ] + Δ) →
     let _ , p′ , q′ = +-assoc-l p q in
-    cut {A} p (wait (> q) P) Q ⊒ wait q′ (cut p′ P Q)
+    cut {A} (wait (ch ⟨ > q ⟩ P) ⟨ p ⟩ Q) ⊒
+    wait (ch ⟨ q′ ⟩ cut (P ⟨ p′ ⟩ Q))
 \end{code}
@@
   \item
-    $(\AgdaInductiveConstructor{wait}~(\AgdaInductiveConstructor{>}~q)~P)$ is a
-    direct sub-process of the \AgdaInductiveConstructor{cut}, and therefore it
-    is meant to be well typed in the context
+    $(\AgdaInductiveConstructor{wait}~(\AgdaInductiveConstructor{ch~⟨~>}~q~\AgdaInductiveConstructor{⟩}~P)$
+    is a direct sub-process of the \AgdaInductiveConstructor{cut}, and therefore
+    it is meant to be well typed in the context
     $A~\AgdaInductiveConstructor{∷}~\Context_1$.
@@
 associativity property of context splitting which we called
-\AgdaFunction{+-assoc-l} in \Cref{sec:context-agda}. The nested
-\AgdaKeyword{let}-\AgdaKeyword{in} allows us to pattern match the result of the
-application $\AgdaFunction{+-assoc-l}~p~q$ and to extract the new proofs $p'$
-and $q'$ for the rearranged splittings.
+\AgdaFunction{+-assoc-l} in \Cref{sec:context-agda}.\Luca{Controllare
+left/right.} The nested \AgdaKeyword{let}-\AgdaKeyword{in} allows us to pattern
+match the result of the application $\AgdaFunction{+-assoc-l}~p~q$ and to
+extract the new proofs $p'$ and $q'$ for the rearranged splittings.
 
-The constructors \AgdaInductiveConstructor{s-left} and
-\AgdaInductiveConstructor{s-right} model \SSelect when the selected tag is
-respectively $\Tag_1$ and $\Tag_2$.
+The constructors \AgdaInductiveConstructor{s-select-l} and
+\AgdaInductiveConstructor{s-select-r} model \SSelect when the selected tag is
+respectively $\InTag_1$ and $\InTag_2$.\Luca{Controllare nomi dei tag con
+sintassi del calcolo. Chiamarli inj come in Agda così c'è consistenza?}
+%
+For example, for \AgdaInductiveConstructor{s-select-l} we have:
 
@@
   s-case :
-    ∀{A B C Γ₁ Γ₂ Δ P Q R} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B & C ⊳ Δ) →
+    ∀{A B C Γ₁ Γ₂ Δ P Q R} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ B & C ] + Δ) →
     let _ , p′ , q′ = +-assoc-l p q in
-    cut {A} p (case (> q) P Q) R ⊒ case q′ (cut (< p′) (↭process swap P) R)
-                                           (cut (< p′) (↭process swap Q) R)
+    cut {A} (case (ch ⟨ > q ⟩ (P , Q)) ⟨ p ⟩ R) ⊒
+    case (ch ⟨ q′ ⟩ (cut (↭proc swap P ⟨ < p′ ⟩ R) ,
+                     cut (↭proc swap Q ⟨ < p′ ⟩ R)))
 \end{code}
 \begin{code}
-  s-left :
-    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⊕ C ⊳ Δ) →
+  s-select-l :
+    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ B ⊕ C ] + Δ) →
     let _ , p′ , q′ = +-assoc-l p q in
-    cut {A} p (left (> q) P) Q ⊒ left q′ (cut (< p′) (↭process swap P) Q)
+    cut {A} (select (ch ⟨ > q ⟩ inj₁ P) ⟨ p ⟩ Q) ⊒
+    select (ch ⟨ q′ ⟩ inj₁ (cut (↭proc swap P ⟨ < p′ ⟩ Q)))
 \end{code}
 \begin{code}[hide]
-  s-right :
-    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⊕ C ⊳ Δ) →
+  s-select-r :
+    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ B ⊕ C ] + Δ) →
     let _ , p′ , q′ = +-assoc-l p q in
-    cut {A} p (right (> q) P) Q ⊒ right q′ (cut (< p′) (↭process swap P) Q)
+    cut {A} (select (ch ⟨ > q ⟩ inj₂ P) ⟨ p ⟩ Q) ⊒
+    select (ch ⟨ q′ ⟩ inj₂ (cut (↭proc swap P ⟨ < p′ ⟩ Q)))
 \end{code}
@@
 Here the process
-$(\AgdaInductiveConstructor{left}~(\AgdaInductiveConstructor{>}~q)~P)$, which is
+$(\AgdaInductiveConstructor{select}~(\AgdaInductiveConstructor{ch ⟨
+>}~q~\AgdaInductiveConstructor{⟩})~\AgdaInductiveConstructor{inj₁}~P)$, which is
 found under a cut for $x : A$, is using some channel $y : B \Plus C$ to send the
-tag $\Tag_1$.
+tag $\InTag_1$.
 %
 Unlike \AgdaInductiveConstructor{wait}, which consumes the channel it operates
-on, the continuation process $P$ will be using a fresh continuation channel $z :
-B$. Therefore, $P$ is required to be well typed in the context
+on, the continuation process $P$ uses a fresh continuation channel $z : B$.
+Therefore, $P$ is required to be well typed in the context
 $B~\AgdaInductiveConstructor{∷}~A~\AgdaInductiveConstructor{∷}~\ContextD$, where
@@
 because now the two channels are introduced in the opposite order. Therefore, we
-are in a situation where we need to turn $P$ into a process that is well typed
-in the context
-$A~\AgdaInductiveConstructor{∷}~B~\AgdaInductiveConstructor{∷}~\ContextD$. This
-is achieved applying the function \AgdaFunction{↭process} to the
+need to rename the channels in $P$ so that it is well typed in the context
+$A~\AgdaInductiveConstructor{∷}~B~\AgdaInductiveConstructor{∷}~\ContextD$. Such
+renaming is achieved applying the function \AgdaFunction{↭proc} to the
 \AgdaInductiveConstructor{swap} permutation, which swaps the two topmost types
@@
 
-We also show the modeling of the \SForkL rule, which is interesting because of
+\begin{code}[hide]
+  s-join :
+    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ B ⅋ C ] + Δ) →
+    let _ , p′ , q′ = +-assoc-l p q in
+    cut {A} (join (ch ⟨ > q ⟩ P) ⟨ p ⟩ Q) ⊒
+    join (ch ⟨ q′ ⟩ cut (↭proc (↭shift {A} {C ∷ B ∷ []}) P ⟨ < < p′ ⟩ Q))
+\end{code}
+
+We also discuss the modeling of the \SForkL rule, which is interesting because of
 its complex side conditions:
 
-\begin{code}[hide]
-  s-join :
-    ∀{Γ₁ Γ₂ Δ A B C P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⅋ C ⊳ Δ) →
-    let _ , p′ , q′ = +-assoc-l p q in
-    cut {A} p (join (> q) P) Q ⊒
-    join q′ (cut (< < p′) (↭process (↭shift {A} {C ∷ B ∷ []}) P) Q)
-\end{code}
 \begin{code}
@@
     ∀{Γ₁ Γ₂ Δ Δ₁ Δ₂ A B C P Q R}
-    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⊗ C ⊳ Δ) (r : Δ ≃ Δ₁ + Δ₂) →
+    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ B ⊗ C ] + Δ) (r : Δ ≃ Δ₁ + Δ₂) →
     let _ , p′ , q′ = +-assoc-l p q in
@@
     let _ , q′′ , r′′ = +-assoc-r r′ (+-comm p′′) in
-    cut {A} p (fork (> q) (< r) P Q) R ⊒
-    fork q′ r′′ (cut (< q′′) (↭process swap P) R) Q
+    cut {A} (fork (ch ⟨ > q ⟩ (P ⟨ < r ⟩ Q)) ⟨ p ⟩ R) ⊒
+    fork (ch ⟨ q′ ⟩ (cut (↭proc swap P ⟨ < q′′ ⟩ R) ⟨ r′′ ⟩ Q))
 \end{code}
@@
     ∀{Γ₁ Γ₂ Δ Δ₁ Δ₂ A B C P Q R}
-    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ B ⊗ C ⊳ Δ) (r : Δ ≃ Δ₁ + Δ₂) →
+    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ B ⊗ C ] + Δ) (r : Δ ≃ Δ₁ + Δ₂) →
     let _ , p′ , q′ = +-assoc-l p q in
     let _ , p′′ , r′ = +-assoc-l p′ r in
-    cut {A} p (fork (> q) (> r) P Q) R ⊒
-    fork q′ r′ P (cut (< p′′) (↭process swap Q) R)
+    cut {A} (fork (ch ⟨ > q ⟩ (P ⟨ > r ⟩ Q)) ⟨ p ⟩ R) ⊒
+    fork (ch ⟨ q′ ⟩ (P ⟨ r′ ⟩ cut (↭proc swap Q ⟨ < p′′ ⟩ R)))
   s-all :
-    ∀{A B Γ₁ Γ₂ Δ F Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `∀ B ⊳ Δ) ->
-    let _ , p' , q' = +-assoc-l p q in
-    cut {A} p (all (> q) F) Q ⊒ all q' λ σ → cut (< p') (↭process swap (F σ)) Q
+    ∀{A B Γ₁ Γ₂ Δ F Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ `∀ B ] + Δ) ->
+    let _ , p′ , q′ = +-assoc-l p q in
+    cut {A} (all (ch ⟨ > q ⟩ F) ⟨ p ⟩ Q) ⊒
+    all (ch ⟨ q′ ⟩ λ X → cut (↭proc swap (F X) ⟨ (< p′) ⟩ Q))
   s-ex :
-    ∀{A B C Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `∃ B ⊳ Δ) ->
-    let _ , p' , q' = +-assoc-l p q in
-    cut {A} p (ex {_} {C} (> q) P) Q ⊒ ex q' (cut (< p') (↭process swap P) Q)
+    ∀{A B C Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ `∃ B ] + Δ) ->
+    let _ , p′ , q′ = +-assoc-l p q in
+    cut {A} (ex {_} {C} (ch ⟨ > q ⟩ P) ⟨ p ⟩ Q) ⊒
+    ex (ch ⟨ q′ ⟩ cut (↭proc swap P ⟨ < p′ ⟩ Q))
   s-server :
     ∀{A B Γ₁ Γ₂ Δ₁ P Q}
-    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `! B ⊳ Δ₁) (r : Γ₂ ≃ [] + Γ₂)
+    (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ `! B ] + Δ₁) (r : Γ₂ ≃ [] + Γ₂)
     (un₁ : Un Δ₁) (un₂ : Un Γ₂) →
     let _ , p′ , q′ = +-assoc-l p q in
-    cut {`? A} p (server (> q) (un-∷ un₁) P) (server (< r) un₂ Q) ⊒
-    server q′ (+-un p′ un₁ un₂) (cut (< p′) (↭process swap P) (server (< r) un₂ Q))
+    cut {`? A} (server (ch ⟨ > q ⟩ (un-∷ un₁ , P)) ⟨ p ⟩ server (ch ⟨ < r ⟩ (un₂ , Q))) ⊒
+    server (ch ⟨ q′ ⟩ (∗-un (un₁ ⟨ p′ ⟩ un₂) , cut (↭proc swap P ⟨ (< p′) ⟩ server (ch ⟨ < r ⟩ (un₂ , Q)))))
   s-client :
-    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `? B ⊳ Δ) →
+    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ `? B ] + Δ) →
     let _ , p′ , q′ = +-assoc-l p q in
-    cut {A} p (client (> q) P) Q ⊒ client q′ (cut (< p′) (↭process swap P) Q)
+    cut {A} (client (ch ⟨ > q ⟩ P) ⟨ p ⟩ Q) ⊒
+    client (ch ⟨ q′ ⟩ cut (↭proc swap P ⟨ < p′ ⟩ Q))
   s-weaken :
-    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `? B ⊳ Δ) →
+    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ `? B ] + Δ) →
     let _ , p′ , q′ = +-assoc-l p q in
-    cut {A} p (weaken (> q) P) Q ⊒ weaken q′ (cut p′ P Q)
+    cut {A} (weaken (ch ⟨ (> q) ⟩ P) ⟨ p ⟩ Q) ⊒ weaken (ch ⟨ q′ ⟩ cut (P ⟨ p′ ⟩ Q))
   s-contract :
-    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ∋ `? B ⊳ Δ) →
+    ∀{A B Γ₁ Γ₂ Δ P Q} (p : Γ ≃ Γ₁ + Γ₂) (q : Γ₁ ≃ [ `? B ] + Δ) →
     let _ , p′ , q′ = +-assoc-l p q in
-    cut {A} p (contract (> q) P) Q ⊒
-    contract q′ (cut (< < p′) (↭process (↭shift {A} {`? B ∷ `? B ∷ []}) P) Q)
+    cut {A} (contract (ch ⟨ > q ⟩ P) ⟨ p ⟩ Q) ⊒
+    contract (ch ⟨ q′ ⟩ cut (↭proc (↭shift {A} {`? B ∷ `? B ∷ []}) P ⟨ < < p′ ⟩ Q))
 \end{code}
@@
   s-cong  : ∀{Γ₁ Γ₂ A P Q P′ Q′} (p : Γ ≃ Γ₁ + Γ₂) →
-            P ⊒ Q → P′ ⊒ Q′ → cut {A} p P P′ ⊒ cut p Q Q′
+            P ⊒ Q → P′ ⊒ Q′ → cut {A} (P ⟨ p ⟩ P′) ⊒ cut (Q ⟨ p ⟩ Q′)
 \end{code}

### 2025-12-12 paper/Process.lagda.tex
revised congruence

@@
   all       : ∀{A} →
-              ∀[ Ch (`∀ A) ∗ ⋂[ X ∶ Type ] ((subst [ X /_] A ∷_) ⊢ Proc) ⇒ Proc ]
-  ex        : ∀{A B} → ∀[ Ch (`∃ A) ∗ ((subst [ B /_] A ∷_) ⊢ Proc) ⇒ Proc ]
+              ∀[ Ch (`∀ A) ∗ ⋂[ X ∶ Type ] ((subst [ X /] A ∷_) ⊢ Proc) ⇒ Proc ]
+  ex        : ∀{A B} → ∀[ Ch (`∃ A) ∗ ((subst [ B /] A ∷_) ⊢ Proc) ⇒ Proc ]
   server    : ∀{A} → ∀[ Ch (`! A) ∗ (Un ∩ ((A ∷_) ⊢ Proc)) ⇒ Proc ]
@@
 The interpretation of the remaining constructors is analogous, so we only
-comment \AgdaInductiveConstructor{all}, which represents a process
+comment \AgdaInductiveConstructor{all}, which represents a polymorphic process
 $\Forall\x\X\y.P$. This constructor models the continuation $P$ using
@@
 \[
-  \AgdaFunction{⋂[}~X~\AgdaFunction{∶}~\AgdaDatatype{Type}~\AgdaFunction]~((\AgdaFunction{subst~[}~X~\AgdaFunction{/\_]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
+  \AgdaFunction{⋂[}~X~\AgdaFunction{∶}~\AgdaDatatype{Type}~\AgdaFunction]~((\AgdaFunction{subst~[}~X~\AgdaFunction{/]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
 \]
@@
 \[
-  ((\AgdaFunction{subst~[}~X~\AgdaFunction{/\_]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
+  ((\AgdaFunction{subst~[}~X~\AgdaFunction{/]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
 \]
@@
 
-We conclude this module proving that permutations preserve process typeability.
-Since list permutations basically correspond to channel renaming, we can read
-this property as the fact that typeability is preserved by (injective) name
+We conclude this module proving that permutations preserve process typing. Since
+list permutations basically correspond to channel renaming, we can read this
+property as the fact that typing is preserved by (injective) name
 substitutions.\Luca{Abbiamo detto da qualche parte dove emergono gli indici di

### 2025-12-13 paper/Reduction.lagda.tex
restored reduction

@@
 {-# OPTIONS --rewriting #-}
-open import Data.Bool using (Bool; true; false; if_then_else_)
-open import Data.Product using (_×_; _,_; ∃; Σ; Σ-syntax; ∃-syntax)
-open import Relation.Binary.PropositionalEquality using (_≡_; refl; sym; cong; cong₂)
-open import Data.List.Base using (List; []; _∷_; [_]; _++_)
+open import Data.Sum using (inj₁; inj₂)
+open import Data.Product using (_,_)
+open import Data.List.Base using ([]; _∷_; [_]; _++_)
 open import Data.List.Properties using (++-assoc)
+open import Relation.Unary
+open import Relation.Binary.PropositionalEquality using (sym)
 
@@
 \begin{code}[hide]
-weakening : ∀{Γ Γ₁ Γ₂} (un : Un Γ₁) → Γ ≃ Γ₁ + Γ₂ → Process Γ₂ → Process Γ
-weakening un p P = ↭process (↭concat p) (aux un P)
+weakening : ∀[ Un ∗ Proc ⇒ Proc ]
+weakening (un ⟨ p ⟩ P) = ↭proc (↭concat p) (aux un P)
   where
-    aux : ∀{Γ₁ Γ₂} (un : Un Γ₁) → Process Γ₂ → Process (Γ₁ ++ Γ₂)
+    aux : ∀{Γ₁ Γ₂} (un : Un Γ₁) → Proc Γ₂ → Proc (Γ₁ ++ Γ₂)
     aux un-[] P = P
-    aux (un-∷ un) P = weaken (< ≫) (aux un P)
+    aux (un-∷ un) P = weaken (ch ⟨ < ≫ ⟩ aux un P)
 
-contraction : ∀{Γ Γ₁ Γ₂} (un : Un Γ₁) → Γ ≃ Γ₁ + Γ₂ → Process (Γ₁ ++ Γ) → Process Γ
-contraction un p P = ↭process (↭concat p) (aux un (↭process (↭left (↭sym (↭concat p))) P))
+contraction : ∀{Γ Γ₁ Γ₂} (un : Un Γ₁) → Γ ≃ Γ₁ + Γ₂ → Proc (Γ₁ ++ Γ) → Proc Γ
+contraction un p P = ↭proc (↭concat p) (aux un (↭proc (↭left (↭sym (↭concat p))) P))
   where
-    aux : ∀{Γ₁ Γ₂} → Un Γ₁ → Process (Γ₁ ++ Γ₁ ++ Γ₂) → Process (Γ₁ ++ Γ₂)
+    aux : ∀{Γ₁ Γ₂} → Un Γ₁ → Proc (Γ₁ ++ Γ₁ ++ Γ₂) → Proc (Γ₁ ++ Γ₂)
     aux un-[] P = P
-    aux {`? A ∷ Γ₁} {Γ₂} (un-∷ un) P with contract (< ≫) (↭process (↭shift {`? A} {`? A ∷ Γ₁} {Γ₁ ++ Γ₂}) P)
-    ... | P₁ rewrite sym (++-assoc (`? A ∷ Γ₁) Γ₁ Γ₂) with ↭process (↭sym (↭shift {`? A} {Γ₁ ++ Γ₁})) P₁
+    aux {`? A ∷ Γ₁} {Γ₂} (un-∷ un) P with contract (ch ⟨ < ≫ ⟩ (↭proc (↭shift {`? A} {`? A ∷ Γ₁} {Γ₁ ++ Γ₂}) P))
+    ... | P₁ rewrite sym (++-assoc (`? A ∷ Γ₁) Γ₁ Γ₂) with ↭proc (↭sym (↭shift {`? A} {Γ₁ ++ Γ₁})) P₁
     ... | P₂ rewrite ++-assoc Γ₁ Γ₁ (`? A ∷ Γ₂) with aux un P₂
-    ... | P₃ = ↭process ↭shift P₃
+    ... | P₃ = ↭proc ↭shift P₃
 \end{code}
@@
 \begin{code}
-data _↝_ {Γ} : Process Γ → Process Γ → Set where
+data _↝_ {Γ} : Proc Γ → Proc Γ → Set where
 \end{code}
@@
   r-link      : ∀{Δ A P} (p : Γ ≃ [ dual A ] + Δ) →
-                cut {A} p (link (< > •)) P ↝ ↭process (↭concat p) P
+                cut {A} (link (ch ⟨ < > • ⟩ ch) ⟨ p ⟩ P) ↝ ↭proc (↭concat p) P
 \end{code}
@@
 The process $P$ has type
-$\AgdaDatatype{Process}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\ContextD)$,
+$\AgdaDatatype{Proc}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\ContextD)$,
 and turns into $P\subst\y\x$ after the reduction. The type
@@
 \begin{code}
-  r-close     : ∀{P} (p₀ : Γ ≃ [] + Γ) (q₀ : Γ ≃ [] + Γ) →
-                cut p₀ close (wait (< q₀) P) ↝ P
+  r-close     : ∀{P} (p₀ q₀ : Γ ≃ [] + Γ) →
+                cut (close ch ⟨ p₀ ⟩ wait (ch ⟨ < q₀ ⟩ P)) ↝ P
 \end{code}
@@
 
-The constructor \AgdaInductiveConstructor{r-left} models \RSelect when the
+The constructor \AgdaInductiveConstructor{r-select-l} models \RSelect when the
 selected tag is $\InTag_1$:
@@
 \begin{code}
-  r-left      : ∀{Γ₁ Γ₂ A B P Q R}
+  r-select-l  : ∀{Γ₁ Γ₂ A B P Q R}
                 (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
-                cut {A ⊕ B} p (left (< p₀) P)
-                              (case (< q₀) Q R) ↝ cut p P Q
+                cut {A ⊕ B} (select (ch ⟨ < p₀ ⟩ inj₁ P) ⟨ p ⟩
+                             case (ch ⟨ < q₀ ⟩ (Q , R))) ↝ cut (P ⟨ p ⟩ Q)
 \end{code}
 
-There is not much more to note here except again for the multiple
-quantifications over the trivial splittings $\Context_i \simeq
-\AgdaInductiveConstructor{[]} + \Context_i$ and the use of
-\AgdaInductiveConstructor{<} to make sure that the channel referred to by
-\AgdaInductiveConstructor{left} and \AgdaInductiveConstructor{case} is indeed
-the one restricted by the cut.
+There is not much to note here except again for the multiple quantifications
+over the trivial splittings $\Context_i \simeq \AgdaInductiveConstructor{[]} +
+\Context_i$ and the use of \AgdaInductiveConstructor{<} to make sure that the
+channel referred to by \AgdaInductiveConstructor{select} and
+\AgdaInductiveConstructor{case} is indeed the one restricted by the cut.
 
 \begin{code}[hide]
-  r-right     : ∀{Γ₁ Γ₂ A B P Q R}
+  r-select-r  : ∀{Γ₁ Γ₂ A B P Q R}
                 (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
-                cut {A ⊕ B} p (right (< p₀) P)
-                              (case (< q₀) Q R) ↝ cut p P R
+                cut {A ⊕ B} (select (ch ⟨ < p₀ ⟩ inj₂ P) ⟨ p ⟩
+                             case (ch ⟨ < q₀ ⟩ (Q , R))) ↝ cut (P ⟨ p ⟩ R)
   r-fork      : ∀{Γ₁ Γ₂ Γ₃ Δ A B P Q R}
@@
                 let _ , p′ , q′ = +-assoc-l p q in
-                cut {A ⊗ B} p (fork (< q₀) q P Q)
-                              (join (< p₀) R) ↝ cut q′ P (cut (> p′) Q R)
+                cut {A ⊗ B} (fork (ch ⟨ < q₀ ⟩ (P ⟨ q ⟩ Q)) ⟨ p ⟩
+                             join (ch ⟨ < p₀ ⟩ R)) ↝ cut (P ⟨ q′ ⟩ cut (Q ⟨ > p′ ⟩ R))
   r-client    : ∀{Γ₁ Γ₂ A P Q}
-                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
-                cut {`! A} p (server (< p₀) un P) (client (< q₀) Q) ↝ cut p P Q
+                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₂) →
+                cut {`? A} (client (ch ⟨ < p₀ ⟩ P) ⟨ p ⟩ server (ch ⟨ < q₀ ⟩ (un , Q))) ↝
+                cut (P ⟨ p ⟩ Q)
   r-weaken    : ∀{Γ₁ Γ₂ A P Q}
-                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
-                cut {`! A} p (server (< p₀) un P)
-                             (weaken (< q₀) Q) ↝ weakening un p Q
+                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₂) →
+                cut {`? A} (weaken (ch ⟨ < p₀ ⟩ P) ⟨ p ⟩ server (ch ⟨ < q₀ ⟩ (un , Q))) ↝
+                weakening (un ⟨ +-comm p ⟩ P)
   r-contract  : ∀{Γ₁ Γ₂ A P Q}
-                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
-                cut {`! A} p (server (< p₀) un P)
-                             (contract (< q₀) Q) ↝
-                    contraction un p
-                      (cut ++≃+ (server (< p₀) un P)
-                      (cut (> p) (server (< p₀) un P) Q))
+                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₂) →
+                cut {`? A} (contract (ch ⟨ < p₀ ⟩ P) ⟨ p ⟩ server (ch ⟨ < q₀ ⟩ (un , Q))) ↝
+                contraction un (+-comm p) (cut (server (ch ⟨ < q₀ ⟩ (un , Q)) ⟨ ++≃+ ⟩
+                                           cut (server (ch ⟨ < q₀ ⟩ (un , Q)) ⟨ > +-comm p ⟩ P)))
 \end{code}
@@
 the resulting process as shown in \Cref{tab:semantics}. It is worth glancing at
-\RExists since it involves a non-trivial rewriting of types:
+the implementation of \RExists since it involves a non-trivial rewriting of
+types:
 
@@
                  (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
-                 cut {`∃ A} p (ex {_} {B} (< p₀) P) (all (< q₀) F) ↝ cut p P (F B)
+                 cut {`∃ A} (ex {A} {B} (ch ⟨ < p₀ ⟩ P) ⟨ p ⟩ all (ch ⟨ < q₀ ⟩ F)) ↝
+                 cut (P ⟨ p ⟩ F B)
 \end{code}
@@
 has type
-$\AgdaDatatype{Process}~(\AgdaFunction{subst}~[~B~/\_]~A~\AgdaInductiveConstructor{∷}~\Context_1)$
+$\AgdaDatatype{Proc}~(\AgdaFunction{subst}~[~B~/]~A~\AgdaInductiveConstructor{∷}~\Context_1)$
 and $(F~B)$ is a process of type
-$\AgdaDatatype{Process}~(\AgdaFunction{subst}~[~B~/\_]~(\AgdaFunction{dual}~A)~\AgdaInductiveConstructor{∷}~\Context_2)$.
+$\AgdaDatatype{Proc}~(\AgdaFunction{subst}~[~B~/]~(\AgdaFunction{dual}~A)~\AgdaInductiveConstructor{∷}~\Context_2)$.
 In order for these two process to be composable in a cut, it must be the case
-that $\AgdaFunction{subst}~[~B~/\_]~(\AgdaFunction{dual}~A) \equiv
-\AgdaFunction{dual}~(\AgdaFunction{subst}~[~B~/\_]~A)$, which was proved in
+that $\AgdaFunction{subst}~[~B~/]~(\AgdaFunction{dual}~A) \equiv
+\AgdaFunction{dual}~(\AgdaFunction{subst}~[~B~/]~A)$, which was proved in
 \Cref{sec:type-agda} under the name \AgdaFunction{dual-subst}. Thanks to the
@@
   r-cut        : ∀{Γ₁ Γ₂ A P Q R} (q : Γ ≃ Γ₁ + Γ₂) →
-                 P ↝ Q → cut {A} q P R ↝ cut q Q R
+                 P ↝ Q → cut {A} (P ⟨ q ⟩ R) ↝ cut (Q ⟨ q ⟩ R)
   r-cong       : ∀{P R Q} → P ⊒ R → R ↝ Q → P ↝ Q

### 2025-12-13 paper/DeadlockFreedom.lagda.tex
revised description of reduction rules
general cleanup
restored deadlock freedom

@@
 (\Cref{thm:df}) rests on some notions and predicates about processes which we
-now formalize in Agda. First of all we need to distinguish \emph{threads} from
-\emph{cuts}. As it turns out, in the formalization it is convenient to provide a
-more fine-grained classification of threads, distinguishing between links,
-input/output actions on linear/shared channels, and sometimes also whether such
-actions operate on free or bound channels. We define predicates for each of
+now formalize in Agda. First of all we need to define the notion of
+\emph{thread}, that is any process other than a cut. As it turns out, it is
+convenient to provide an more fine-grained classification of threads,
+distinguishing between links, input/output actions, and sometimes also whether
+such actions operate on free or bound channels. We define predicates for each of
 these process classes:
@@
 {-# OPTIONS --rewriting #-}
+open import Data.Unit using (tt)
 open import Data.Sum
@@
 open import Relation.Nullary using (¬_; contradiction)
+open import Relation.Unary
 open import Relation.Binary.PropositionalEquality using (refl)
@@
 open import Context
-open import Permutations
 open import Process
@@
 \begin{code}
-data Cut            : ∀{Γ} → Process Γ → Set
-data Link           : ∀{Γ} → Process Γ → Set
-data Input          : ∀{Γ} → Process Γ → Set
-data Output         : ∀{Γ} → Process Γ → Set
-data Delayed        : ∀{Γ} → Process Γ → Set
-data Client         : ∀{Γ} → Process Γ → Set
-data Server         : ∀{Γ} → Process Γ → Set
-data DelayedServer  : ∀{Γ} → Process Γ → Set
+data Link           : ∀{Γ} → Proc Γ → Set
+data Input          : ∀{Γ} → Proc Γ → Set
+data Output         : ∀{Γ} → Proc Γ → Set
+data Delayed        : ∀{Γ} → Proc Γ → Set
+data Server         : ∀{Γ} → Proc Γ → Set
+data DelayedServer  : ∀{Γ} → Proc Γ → Set
 \end{code}
@@
 These predicates are relatively uninteresting since they are essentially
-isomorphic to the relevant parts of the \AgdaDatatype{Process} datatype. The
-only remarkable aspect is that in \AgdaDatatype{Input}, \AgdaDatatype{Output},
-\AgdaDatatype{Client} and \AgdaDatatype{Server} the channel being acted upon by
+isomorphic to the relevant parts of the \AgdaDatatype{Proc} datatype. The only
+relevant aspect that is worth pointing out here is that in \AgdaDatatype{Input},
+\AgdaDatatype{Output} and \AgdaDatatype{Server} the channel being acted upon by
 the thread is the \emph{first} in $\Context$, hence it is the \emph{most
@@
 \begin{code}[hide]
-data Cut where
-  cut : ∀{Γ Γ₁ Γ₂ A P Q} (p : Γ ≃ Γ₁ + Γ₂) -> Cut (cut {A} p P Q)
-
 data Link where
-  link : ∀{Γ A} (p : Γ ≃ [ A ] + [ dual A ]) → Link (link p)
+  link : ∀{Γ A} (p : Γ ≃ [ A ] + [ dual A ]) → Link (link (ch ⟨ p ⟩ ch))
 
 data Input where
-  fail : ∀{Γ Δ} (p : Γ ≃ [] + Δ) → Input (fail (< p))
+  fail : ∀{Γ Δ} (p : Γ ≃ [] + Δ) → Input (fail (ch ⟨ < p ⟩ tt))
 \end{code}
@@
 \begin{code}
-  wait : ∀{Γ Δ P} (p : Γ ≃ [] + Δ) → Input (wait (< p) P)
+  wait : ∀{Γ Δ P} (p : Γ ≃ [] + Δ) → Input (wait (ch ⟨ < p ⟩ P))
 \end{code}
-where the \AgdaInductiveConstructor{<} constructor indicates that the thread
-operates on the most recent channel. In the latter predicate we have
+where the use of the \AgdaInductiveConstructor{<} constructor indicates that the
+thread operates on the most recent channel. In the latter predicate we have
 
 \begin{code}[hide]
-  case : ∀{Γ Δ A B P Q} (p : Γ ≃ [] + Δ) → Input (case {A} {B} (< p) P Q)
-  join : ∀{Γ Δ A B P} (p : Γ ≃ [] + Δ) → Input (join {A} {B} (< p) P)
-  all  : ∀{A Γ Δ F} (p : Γ ≃ [] + Δ) -> Input (all {A} (< p) F)
+  case : ∀{Γ Δ A B P Q} (p : Γ ≃ [] + Δ) → Input (case {A} {B} (ch ⟨ < p ⟩ (P , Q)))
+  join : ∀{Γ Δ A B P} (p : Γ ≃ [] + Δ) → Input (join {A} {B} (ch ⟨ < p ⟩ P))
+  all  : ∀{A Γ Δ F} (p : Γ ≃ [] + Δ) → Input (all {A} (ch ⟨ < p ⟩ F))
 
 data Output where
-  close  : Output close
-  left   : ∀{Γ Δ A B P} (p : Γ ≃ [] + Δ) → Output (left {A} {B} (< p) P)
-  right  : ∀{Γ Δ A B P} (p : Γ ≃ [] + Δ) → Output (right {A} {B} (< p) P)
-  fork   : ∀{Γ Δ Δ₁ Δ₂ A B P Q} (p : Γ ≃ [] + Δ) (q : Δ ≃ Δ₁ + Δ₂) -> Output (fork {A} {B} (< p) q P Q)
-  ex     : ∀{A B Γ Δ P} (p : Γ ≃ [] + Δ) -> Output (ex {A} {B} (< p) P)
+  close    : Output (close ch)
+  select-l : ∀{Γ Δ A B P} (p : Γ ≃ [] + Δ) → Output (select {A} {B} (ch ⟨ < p ⟩ inj₁ P))
+  select-r : ∀{Γ Δ A B P} (p : Γ ≃ [] + Δ) → Output (select {A} {B} (ch ⟨ < p ⟩ inj₂ P))
+  fork     : ∀{Γ Δ Δ₁ Δ₂ A B P Q} (p : Γ ≃ [] + Δ) (q : Δ ≃ Δ₁ + Δ₂) → Output (fork {A} {B} (ch ⟨ < p ⟩ (P ⟨ q ⟩ Q)))
+  ex       : ∀{A B Γ Δ P} (p : Γ ≃ [] + Δ) → Output (ex {A} {B} (ch ⟨ < p ⟩ P))
+  client   : ∀{Γ Δ A P}   (p : Γ ≃ [] + Δ) → Output (client {A} (ch ⟨ < p ⟩ P))
+  weaken   : ∀{Γ Δ A P}   (p : Γ ≃ [] + Δ) → Output (weaken {A} (ch ⟨ < p ⟩ P))
+  contract : ∀{Γ Δ A P}   (p : Γ ≃ [] + Δ) → Output (contract {A} (ch ⟨ < p ⟩ P))
 
 data Delayed where
-  fail     : ∀{C Γ Δ} (p : Γ ∋ ⊤ ⊳ Δ) → Delayed (fail (>_ {C} p))
+  fail     : ∀{C Γ Δ} (p : Γ ≃ [ ⊤ ] + Δ) → Delayed (fail (ch ⟨ >_ {C} p ⟩ tt))
 \end{code}
@@
 \begin{code}
-  wait     : ∀{C Γ Δ P} (p : Γ ∋ ⊥ ⊳ Δ) → Delayed (wait (>_ {C} p) P)
+  wait     : ∀{C Γ Δ P} (p : Γ ≃ [ ⊥ ] + Δ) → Delayed (wait (ch ⟨ >_ {C} p ⟩ P))
 \end{code}
-where the \AgdaInductiveConstructor{>} constructor indicates that the thread
-operates on a channel other than the most recent one. Note that here we have to
-specify the type $C$ in front of $\Context$ or else Agda is unable to resolve
-some metavariables.
+where the use of the \AgdaInductiveConstructor{>} constructor indicates that the
+thread operates on a channel other than the most recent one. Note that here we
+have to specify the type $C$ in front of $\Context$ or else Agda is unable to
+resolve some metavariables.
 
 \begin{code}[hide]
-  case     : ∀{Γ Δ C A B P Q} (p : Γ ∋ A & B ⊳ Δ) → Delayed (case {A} {B} (>_ {C} p) P Q)
-  join     : ∀{Γ Δ C A B P} (p : Γ ∋ A ⅋ B ⊳ Δ) → Delayed (join (>_ {C} p) P)
-  left     : ∀{Γ Δ C A B P} (p : Γ ∋ A ⊕ B ⊳ Δ) → Delayed (left (>_ {C} p) P)
-  right    : ∀{Γ Δ C A B P} (p : Γ ∋ A ⊕ B ⊳ Δ) → Delayed (right (>_ {C} p) P)
-  fork-l   : ∀{Γ Δ Δ₁ Δ₂ C A B P Q} (p : Γ ∋ A ⊗ B ⊳ Δ) (q : Δ ≃ Δ₁ + Δ₂) →
-             Delayed (fork (>_ {C} p) (< q) P Q)
-  fork-r   : ∀{Γ Δ Δ₁ Δ₂ C A B P Q} (p : Γ ∋ A ⊗ B ⊳ Δ) (q : Δ ≃ Δ₁ + Δ₂) →
-             Delayed (fork (>_ {C} p) (> q) P Q)
-  client   : ∀{Γ Δ A C P} (p : Γ ∋ `? A ⊳ Δ) → Delayed (client (>_ {C} p) P)
-  weaken   : ∀{Γ Δ A C P} (p : Γ ∋ `? A ⊳ Δ) → Delayed (weaken (>_ {C} p) P)
-  contract : ∀{Γ Δ A C P} (p : Γ ∋ `? A ⊳ Δ) → Delayed (contract (>_ {C} p) P)
-  ex       : ∀{A B C Γ Δ P} (p : Γ ∋ `∃ A ⊳ Δ) → Delayed (ex {A} {B} (>_ {C} p) P)
-  all      : ∀{A C Γ Δ F} (p : Γ ∋ `∀ A ⊳ Δ) → Delayed (all (>_ {C} p) F)
-
-data Client where
-  client   : ∀{Γ Δ A P} (p : Γ ≃ [] + Δ) → Client (client {A} (< p) P)
-  weaken   : ∀{Γ Δ A P} (p : Γ ≃ [] + Δ) → Client (weaken {A} (< p) P)
-  contract : ∀{Γ Δ A P} (p : Γ ≃ [] + Δ) → Client (contract {A} (< p) P)
+  case     : ∀{Γ Δ C A B P} (p : Γ ≃ [ A & B ] + Δ) → Delayed (case {A} {B} (ch ⟨ >_ {C} p ⟩ P))
+  select-l : ∀{Γ Δ C A B P} (p : Γ ≃ [ A ⊕ B ] + Δ) → Delayed (select (ch ⟨ >_ {C} p ⟩ inj₁ P))
+  select-r : ∀{Γ Δ C A B P} (p : Γ ≃ [ A ⊕ B ] + Δ) → Delayed (select (ch ⟨ >_ {C} p ⟩ inj₂ P))
+  join     : ∀{Γ Δ C A B P} (p : Γ ≃ [ A ⅋ B ] + Δ) → Delayed (join (ch ⟨ >_ {C} p ⟩ P))
+  fork-l   : ∀{Γ Δ Δ₁ Δ₂ C A B P Q} (p : Γ ≃ [ A ⊗ B ] + Δ) (q : Δ ≃ Δ₁ + Δ₂) →
+             Delayed (fork (ch ⟨ >_ {C} p ⟩ (P ⟨ < q ⟩ Q)))
+  fork-r   : ∀{Γ Δ Δ₁ Δ₂ C A B P Q} (p : Γ ≃ [ A ⊗ B ] + Δ) (q : Δ ≃ Δ₁ + Δ₂) →
+             Delayed (fork (ch ⟨ >_ {C} p ⟩ (P ⟨ > q ⟩ Q)))
+  all      : ∀{A C Γ Δ F} (p : Γ ≃ [ `∀ A ] + Δ) → Delayed (all (ch ⟨ >_ {C} p ⟩ F))
+  ex       : ∀{A B C Γ Δ P} (p : Γ ≃ [ `∃ A ] + Δ) → Delayed (ex {A} {B} (ch ⟨ >_ {C} p ⟩ P))
+  client   : ∀{Γ Δ A C P} (p : Γ ≃ [ `? A ] + Δ) → Delayed (client (ch ⟨ >_ {C} p ⟩ P))
+  weaken   : ∀{Γ Δ A C P} (p : Γ ≃ [ `? A ] + Δ) → Delayed (weaken (ch ⟨ >_ {C} p ⟩ P))
+  contract : ∀{Γ Δ A C P} (p : Γ ≃ [ `? A ] + Δ) → Delayed (contract (ch ⟨ >_ {C} p ⟩ P))
 
 data Server where
-  server : ∀{Γ Δ A P} (p : Γ ≃ [] + Δ) (un : Un Δ) → Server (server {A} (< p) un P)
+  server : ∀{Γ Δ A P} (p : Γ ≃ [] + Δ) (un : Un Δ) → Server (server {A} (ch ⟨ < p ⟩ (un , P)))
 
 data DelayedServer where
-  server : ∀{Γ Δ A C P} (p : Γ ∋ `! A ⊳ Δ) (un : Un Δ) →
-           DelayedServer (server (> p) (un-∷ {C} un) P)
+  server : ∀{Γ Δ A C P} (p : Γ ≃ [ `! A ] + Δ) (un : Un Δ) →
+           DelayedServer (server {A} (ch ⟨ >_ p ⟩ (un-∷ {C} un , P)))
 \end{code}
@@
 
-Now the \AgdaDatatype{Thread} predicate is simply the disjunction of all the
+Now the \AgdaDatatype{Thread} predicate is simply the disjoint union of all the
 previous ones.
@@
 \begin{code}
-Thread : ∀{Γ} → Process Γ → Set
-Thread P = Link P ⊎ Delayed P ⊎ Output P ⊎ Input P ⊎
-           Client P ⊎ Server P ⊎ DelayedServer P
+data Thread {Γ} (P : Proc Γ) : Set where
+  link     : Link P → Thread P
+  delayed  : Delayed P → Thread P
+  output   : Output P → Thread P
+  input    : Input P → Thread P
+  server   : Server P → Thread P
+  dserver  : DelayedServer P → Thread P
 \end{code}
@@
 \begin{code}
-Observable : ∀{Γ} → Process Γ → Set
+Observable : ∀{Γ} → Proc Γ → Set
 Observable P = ∃[ Q ] P ⊒ Q × Thread Q
 
-Reducible : ∀{Γ} → Process Γ → Set
+Reducible : ∀{Γ} → Proc Γ → Set
 Reducible P = ∃[ Q ] P ↝ Q
 
-Alive : ∀{Γ} → Process Γ → Set
+Alive : ∀{Γ} → Proc Γ → Set
 Alive P = Observable P ⊎ Reducible P
@@
 \begin{code}[hide]
-link-thread : ∀{A Γ} (p : Γ ≃ [ A ] + [ dual A ]) → Thread (link p)
-link-thread p = inj₁ (link p)
+fail→thread : ∀{Γ Δ} (p : Γ ≃ [ ⊤ ] + Δ) → Thread (fail (ch ⟨ p ⟩ tt))
+fail→thread (< p) = input (fail p)
+fail→thread (> p) = delayed (fail p)
 
-close-thread : Thread close
-close-thread = inj₂ (inj₂ (inj₁ close))
+wait→thread : ∀{Γ Δ P} (p : Γ ≃ [ ⊥ ] + Δ) → Thread (wait (ch ⟨ p ⟩ P))
+wait→thread (< p) = input (wait p)
+wait→thread (> p) = delayed (wait p)
 
-wait-thread : ∀{Γ Δ P} (p : Γ ≃ [ ⊥ ] + Δ) → Thread (wait p P)
-wait-thread (< p) = inj₂ (inj₂ (inj₂ (inj₁ (wait p))))
-wait-thread (> p) = inj₂ (inj₁ (wait p))
+case→thread : ∀{A B Γ Δ P} (p : Γ ≃ [ A & B ] + Δ) → Thread (case (ch ⟨ p ⟩ P))
+case→thread (< p) = input (case p)
+case→thread (> p) = delayed (case p)
 
-fail-thread : ∀{Γ Δ} (p : Γ ≃ [ ⊤ ] + Δ) → Thread (fail p)
-fail-thread (< p) = inj₂ (inj₂ (inj₂ (inj₁ (fail p))))
-fail-thread (> p) = inj₂ (inj₁ (fail p))
+left→thread : ∀{A B Γ Δ P} (p : Γ ≃ [ A ⊕ B ] + Δ) → Thread (select (ch ⟨ p ⟩ inj₁ P))
+left→thread (< p) = output (select-l p)
+left→thread (> p) = delayed (select-l p)
 
-case-thread : ∀{A B Γ Δ P Q} (p : Γ ≃ [ A & B ] + Δ) → Thread (case p P Q)
-case-thread (< p) = inj₂ (inj₂ (inj₂ (inj₁ (case p))))
-case-thread (> p) = inj₂ (inj₁ (case p))
+right→thread : ∀{A B Γ Δ P} (p : Γ ≃ [ A ⊕ B ] + Δ) → Thread (select (ch ⟨ p ⟩ inj₂ P))
+right→thread (< p) = output (select-r p)
+right→thread (> p) = delayed (select-r p)
 
-join-thread : ∀{A B Γ Δ P} (p : Γ ≃ [ A ⅋ B ] + Δ) → Thread (join p P)
-join-thread (< p) = inj₂ (inj₂ (inj₂ (inj₁ (join p))))
-join-thread (> p) = inj₂ (inj₁ (join p))
+join→thread : ∀{A B Γ Δ P} (p : Γ ≃ [ A ⅋ B ] + Δ) → Thread (join (ch ⟨ p ⟩ P))
+join→thread (< p) = input (join p)
+join→thread (> p) = delayed (join p)
 
-left-thread : ∀{A B Γ Δ P} (p : Γ ∋ A ⊕ B ⊳ Δ) → Thread (left p P)
-left-thread (< p) = inj₂ (inj₂ (inj₁ (left p)))
-left-thread (> p) = inj₂ (inj₁ (left p))
+fork→thread : ∀{A B Γ Δ Δ₁ Δ₂ P Q} (p : Γ ≃ [ A ⊗ B ] + Δ) (q : Δ ≃ Δ₁ + Δ₂) → Thread (fork (ch ⟨ p ⟩ (P ⟨ q ⟩ Q)))
+fork→thread (< p) q = output (fork p q)
+fork→thread (> p) (< q) = delayed (fork-l p q)
+fork→thread (> p) (> q) = delayed (fork-r p q)
 
-right-thread : ∀{A B Γ Δ P} (p : Γ ∋ A ⊕ B ⊳ Δ) → Thread (right p P)
-right-thread (< p) = inj₂ (inj₂ (inj₁ (right p)))
-right-thread (> p) = inj₂ (inj₁ (right p))
+all→thread : ∀{A Γ Δ F} (p : Γ ≃ [ `∀ A ] + Δ) → Thread (all (ch ⟨ p ⟩ F))
+all→thread (< p) = input (all p)
+all→thread (> p) = delayed (all p)
 
-fork-thread : ∀{A B Γ Δ Δ₁ Δ₂ P Q} (p : Γ ≃ [ A ⊗ B ] + Δ) (q : Δ ≃ Δ₁ + Δ₂) → Thread (fork p q P Q)
-fork-thread (< p) q = inj₂ (inj₂ (inj₁ (fork p q)))
-fork-thread (> p) (< q) = inj₂ (inj₁ (fork-l p q))
-fork-thread (> p) (> q) = inj₂ (inj₁ (fork-r p q))
+ex→thread : ∀{A B Γ Δ P} (p : Γ ≃ [ `∃ A ] + Δ) → Thread (ex {A} {B} (ch ⟨ p ⟩ P))
+ex→thread (< p) = output (ex p)
+ex→thread (> p) = delayed (ex p)
 
-client-thread : ∀{A Γ Δ P} (p : Γ ≃ [ `? A ] + Δ) → Thread (client p P)
-client-thread (< p) = inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (client p)))))
-client-thread (> p) = inj₂ (inj₁ (client p))
+server→thread : ∀{A Γ Δ P} (p : Γ ≃ [ `! A ] + Δ) (un : Un Δ) → Thread (server (ch ⟨ p ⟩ (un , P)))
+server→thread (< p) un = server (server p un)
+server→thread (> p) (un-∷ un) = dserver (server p un)
 
-weaken-thread : ∀{A Γ Δ P} (p : Γ ≃ [ `? A ] + Δ) → Thread (weaken {A = A} p P)
-weaken-thread (< p) = inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (weaken p)))))
-weaken-thread (> p) = inj₂ (inj₁ (weaken p))
+client→thread : ∀{A Γ Δ P} (p : Γ ≃ [ `? A ] + Δ) → Thread (client (ch ⟨ p ⟩ P))
+client→thread (< p) = output (client p)
+client→thread (> p) = delayed (client p)
 
-contract-thread : ∀{A Γ Δ P} (p : Γ ≃ [ `? A ] + Δ) → Thread (contract p P)
-contract-thread (< p) = inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (contract p)))))
-contract-thread (> p) = inj₂ (inj₁ (contract p))
+weaken→thread : ∀{A Γ Δ P} (p : Γ ≃ [ `? A ] + Δ) → Thread (weaken {A = A} (ch ⟨ p ⟩ P))
+weaken→thread (< p) = output (weaken p)
+weaken→thread (> p) = delayed (weaken p)
 
-server-thread : ∀{A Γ Δ P} (p : Γ ≃ [ `! A ] + Δ) (un : Un Δ) → Thread (server p un P)
-server-thread (< p) un = inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ (server p un))))))
-server-thread (> p) (un-∷ un) = inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (server p un))))))
-
-ex-thread : ∀{A B Γ Δ P} (p : Γ ∋ `∃ A ⊳ Δ) -> Thread (ex {A} {B} p P)
-ex-thread (< p) = inj₂ (inj₂ (inj₁ (ex p)))
-ex-thread (> p) = inj₂ (inj₁ (ex p))
-
-all-thread : ∀{A Γ Δ F} (p : Γ ∋ `∀ A ⊳ Δ) -> Thread (all p F)
-all-thread (< p) = inj₂ (inj₂ (inj₂ (inj₁ (all p))))
-all-thread (> p) = inj₂ (inj₁ (all p))
+contract→thread : ∀{A Γ Δ P} (p : Γ ≃ [ `? A ] + Δ) → Thread (contract (ch ⟨ p ⟩ P))
+contract→thread (< p) = output (contract p)
+contract→thread (> p) = delayed (contract p)
 \end{code}
@@
 \begin{code}
-data CanonicalCut {Γ} : Process Γ → Set where
+data CanonicalCut {Γ} : Proc Γ → Set where
   cc-link     : ∀{Γ₁ Γ₂ A P Q} (p : Γ ≃ Γ₁ + Γ₂) →
-                      Link P → CanonicalCut (cut {A} p P Q)
+                Link P → CanonicalCut (cut {A} (P ⟨ p ⟩ Q))
+  cc-redex    : ∀{Γ₁ Γ₂ A P Q} (p : Γ ≃ Γ₁ + Γ₂) →
+                Output P → (Input ∪ Server) Q →
+                CanonicalCut (cut {A} (P ⟨ p ⟩ Q))
   cc-delayed  : ∀{Γ₁ Γ₂ A P Q} (p : Γ ≃ Γ₁ + Γ₂) →
-                      Delayed P → CanonicalCut (cut {A} p P Q)
-  cc-delayed-server : ∀{Γ₁ Γ₂ A P Q} (p : Γ ≃ Γ₁ + Γ₂) →
-                      DelayedServer P → Server Q → CanonicalCut (cut {A} p P Q)
-  cc-redex          : ∀{Γ₁ Γ₂ A P Q} (p : Γ ≃ Γ₁ + Γ₂) →
-                      Output P → Input Q → CanonicalCut (cut {A} p P Q)
-  cc-connect        : ∀{Γ₁ Γ₂ A P Q} (p : Γ ≃ Γ₁ + Γ₂) →
-                      Server P → Client Q → CanonicalCut (cut {A} p P Q)
+                Delayed P → CanonicalCut (cut {A} (P ⟨ p ⟩ Q))
+  cc-servers  : ∀{Γ₁ Γ₂ A P Q} (p : Γ ≃ Γ₁ + Γ₂) →
+                DelayedServer P → Server Q →
+                CanonicalCut (cut {A} (P ⟨ p ⟩ Q))
 \end{code}
@@
 \item $P$ is a link (\AgdaInductiveConstructor{cc-link}), or
+\item $P$ performs an output on $x$ and $Q$ performs an input on $x$
+  (\AgdaInductiveConstructor{cc-redex}), or
 \item $P$ operates on a channel other than $x$ and is not a server
@@
 \item $P$ and $Q$ are servers and $P$ operates on a channel other than $x$
-(\AgdaInductiveConstructor{cc-delayed-server}), or
-\item $P$ performs an output on $x$ and $Q$ performs an input on $x$
-(\AgdaInductiveConstructor{cc-redex} and \AgdaInductiveConstructor{cc-connect}).
+(\AgdaInductiveConstructor{cc-servers}).
 \end{itemize}
+
+Note that we have to distinguish servers from the other input operations because
+the structural pre-congruence rule \SServer can only be applied when the two
+sub-processes of a cut are both servers.
 
@@
 output-output (close , ())
-
-output-client : ∀{A Γ Δ P Q} → ¬ (Output {A ∷ Γ} P × Client {dual A ∷ Δ} Q)
-output-client (close , ())
-
-output-server : ∀{A Γ Δ P Q} → ¬ (Output {A ∷ Γ} P × Server {dual A ∷ Δ} Q)
-output-server (close , ())
 
@@
 
-input-client : ∀{A Γ Δ P Q} → ¬ (Input {A ∷ Γ} P × Client {dual A ∷ Δ} Q)
-input-client (fail _ , ())
-
 input-server : ∀{A Γ Δ P Q} → ¬ (Input {A ∷ Γ} P × Server {dual A ∷ Δ} Q)
@@
 input-delayed-server (fail _ , ())
-
-client-client : ∀{A Γ Δ P Q} → ¬ (Client {A ∷ Γ} P × Client {dual A ∷ Δ} Q)
-client-client (client _ , ())
-
-client-delayed-server : ∀{A Γ Δ P Q} → ¬ (Client {A ∷ Γ} P × DelayedServer {dual A ∷ Δ} Q)
-client-delayed-server (client _ , ())
 
@@
                 Thread P → Thread Q →
-                ∃[ R ] CanonicalCut R × cut {A} p P Q ⊒ R
+                ∃[ R ] CanonicalCut R × cut {A} (P ⟨ p ⟩ Q) ⊒ R
 \end{code}
@@
 \begin{code}[hide]
-canonical-cut pc (inj₁ x) Qt = _ , cc-link pc x , s-refl
-canonical-cut pc (inj₂ x) (inj₁ y) = _ , cc-link (+-comm pc) y , s-comm pc
-canonical-cut pc (inj₂ (inj₁ x)) (inj₂ y) = _ , cc-delayed pc x , s-refl
-canonical-cut pc (inj₂ (inj₂ x)) (inj₂ (inj₁ y)) = _ , cc-delayed (+-comm pc) y , s-comm pc
-canonical-cut pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₁ y))) = contradiction (x , y) output-output
-canonical-cut pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = _ , cc-redex pc x y , s-refl
-canonical-cut pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) output-client
-canonical-cut pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) output-server
-canonical-cut pc (inj₂ (inj₂ (inj₁ x))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) output-delayed-server
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₁ y))) = _ , cc-redex (+-comm pc) y x , s-comm pc
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) output-client
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) output-server
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₁ y))) = contradiction (y , x) output-delayed-server
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (x , y) (input-input)
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) input-client
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) input-server
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₁ x)))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) input-delayed-server
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) input-client
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) input-server
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₁ y)))) = contradiction (y , x) input-delayed-server
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (x , y) client-client
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = _ , cc-connect (+-comm pc) y x , s-comm pc
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) client-delayed-server
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = _ , cc-connect pc x y , s-refl
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y))))) = contradiction (y , x) client-delayed-server
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = contradiction (x , y) server-server
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = _ , cc-delayed-server (+-comm pc) y x , s-comm pc
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₁ y)))))) = _ , cc-delayed-server pc x y , s-refl
-canonical-cut pc (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ x)))))) (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ (inj₂ y)))))) = contradiction (x , y) delayed-server-delayed-served
+canonical-cut pc (link x) Qt = _ , cc-link pc x , s-refl
+canonical-cut pc Pt (link y) = _ , cc-link (+-comm pc) y , s-comm pc
+canonical-cut pc (delayed x) Qt = _ , cc-delayed pc x , s-refl
+canonical-cut pc Pt (delayed y) = _ , cc-delayed (+-comm pc) y , s-comm pc
+canonical-cut pc (output x) (output y) = contradiction (x , y) output-output
+canonical-cut pc (output x) (input y) = _ , cc-redex pc x (inj₁ y) , s-refl
+canonical-cut pc (output x) (server y) = _ , cc-redex pc x (inj₂ y) , s-refl
+canonical-cut pc (output x) (dserver y) = contradiction (x , y) output-delayed-server
+canonical-cut pc (input x) (output y) = _ , cc-redex (+-comm pc) y (inj₁ x) , s-comm pc
+canonical-cut pc (input x) (input y) = contradiction (x , y) input-input
+canonical-cut pc (input x) (server y) = contradiction (x , y) input-server
+canonical-cut pc (input x) (dserver y) = contradiction (x , y) input-delayed-server
+canonical-cut pc (server x) (output y) = _ , cc-redex (+-comm pc) y (inj₂ x) , s-comm pc
+canonical-cut pc (server x) (input y) = contradiction (y , x) input-server
+canonical-cut pc (server x) (server y) = contradiction (x , y) server-server
+canonical-cut pc (server x) (dserver y) = _ , cc-servers (+-comm pc) y x , s-comm pc
+canonical-cut pc (dserver x) (output y) = contradiction (y , x) output-delayed-server
+canonical-cut pc (dserver x) (input y) = contradiction (y , x) input-delayed-server
+canonical-cut pc (dserver x) (server y) = _ , cc-servers pc x y , s-refl
+canonical-cut pc (dserver x) (dserver y) = contradiction (x , y) delayed-server-delayed-served
 
-⊒Alive : ∀{Γ} {P Q : Process Γ} → P ⊒ Q → Alive Q → Alive P
+⊒Alive : ∀{Γ} {P Q : Proc Γ} → P ⊒ Q → Alive Q → Alive P
 ⊒Alive pcong (inj₁ (_ , x , th)) = inj₁ (_ , s-tran pcong x , th)
@@
 \begin{code}
-canonical-cut-alive : ∀{Γ} {P : Process Γ} → CanonicalCut P → Alive P
+canonical-cut-alive : ∀{Γ} {C : Proc Γ} → CanonicalCut C → Alive C
 \end{code}
@@
 \begin{code}[hide]
-canonical-cut-alive (cc-link p (link (< > •))) = inj₂ (_ , r-link p)
-canonical-cut-alive (cc-link p (link (> < •))) =
-  inj₂ (_ , r-cong (s-cong p (s-link (> < •)) s-refl) (r-link p))
-canonical-cut-alive (cc-redex p close (wait q)) with +-empty-l p | +-empty-l q
-... | refl | refl = inj₂ (_ , r-close p q)
-canonical-cut-alive (cc-redex p (right q) (case r)) with +-empty-l q | +-empty-l r
-... | refl | refl = inj₂ (_ , r-right p q r)
-canonical-cut-alive (cc-redex p (left q) (case r)) with +-empty-l q | +-empty-l r
-... | refl | refl = inj₂ (_ , r-left p q r)
-canonical-cut-alive (cc-redex p (fork q r) (join s)) with +-empty-l q | +-empty-l s
-... | refl | refl = inj₂ (_ , r-fork p s r q)
-canonical-cut-alive (cc-redex p (ex q) (all r)) with +-empty-l q | +-empty-l r
-... | refl | refl = inj₂ (_ , r-poly p q r)
-canonical-cut-alive (cc-connect p (server q un) (client r)) with +-empty-l q | +-empty-l r
-... | refl | refl = inj₂ (_ , r-client p q r un)
-canonical-cut-alive (cc-connect p (server q un) (weaken r)) with +-empty-l q | +-empty-l r
-... | refl | refl = inj₂ (_ , r-weaken p q r un)
-canonical-cut-alive (cc-connect p (server q un) (contract r)) with +-empty-l q | +-empty-l r
-... | refl | refl = inj₂ (_ , r-contract p q r un)
-canonical-cut-alive (cc-delayed p (wait q)) =
-  let _ , _ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-wait p q , wait-thread q′)
-canonical-cut-alive (cc-delayed p (case q)) =
-  let _ , _ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-case p q , case-thread q′)
-canonical-cut-alive (cc-delayed p (join q)) =
-  let _ , _ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-join p q , join-thread q′)
-canonical-cut-alive (cc-delayed p (right q)) =
-  let _ , _ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-right p q , right-thread q′)
-canonical-cut-alive (cc-delayed p (left q)) =
-  let _ , _ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-left p q , left-thread q′)
+canonical-cut-alive (cc-link pc (link (< > •))) = inj₂ (_ , r-link pc)
+canonical-cut-alive (cc-link pc (link (> < •))) =
+  inj₂ (_ , r-cong (s-cong pc (s-link _) s-refl) (r-link pc))
+canonical-cut-alive (cc-redex pc close (inj₁ (wait p))) with +-empty-l p | +-empty-l pc
+... | refl | refl = inj₂ (_ , r-close pc p)
+canonical-cut-alive (cc-redex pc (select-l p) (inj₁ (case q))) with +-empty-l p | +-empty-l q
+... | refl | refl = inj₂ (_ , r-select-l pc p q)
+canonical-cut-alive (cc-redex pc (select-r p) (inj₁ (case q))) with +-empty-l p | +-empty-l q
+... | refl | refl = inj₂ (_ , r-select-r pc p q)
+canonical-cut-alive (cc-redex pc (fork p q) (inj₁ (join r))) with +-empty-l p | +-empty-l r
+... | refl | refl = inj₂ (_ , r-fork pc r q p)
+canonical-cut-alive (cc-redex pc (ex p) (inj₁ (all q))) with +-empty-l p | +-empty-l q
+... | refl | refl = inj₂ (_ , r-exists pc p q)
+canonical-cut-alive (cc-redex pc (client p) (inj₂ (server q un))) with +-empty-l p | +-empty-l q
+... | refl | refl = inj₂ (_ , r-client pc p q un)
+canonical-cut-alive (cc-redex pc (weaken p) (inj₂ (server q un))) with +-empty-l p | +-empty-l q
+... | refl | refl = inj₂ (_ , r-weaken pc p q un)
+canonical-cut-alive (cc-redex pc (contract p) (inj₂ (server q un))) with +-empty-l p | +-empty-l q
+... | refl | refl = inj₂ (_ , r-contract pc p q un)
+canonical-cut-alive (cc-delayed pc (fail p)) =
+  let _ , _ , p' = +-assoc-l pc p in
+  inj₁ (_ , s-fail pc p , fail→thread p')
+canonical-cut-alive (cc-delayed pc (wait p)) =
+  let _ , _ , p' = +-assoc-l pc p in
+  inj₁ (_ , s-wait pc p , wait→thread p')
+canonical-cut-alive (cc-delayed pc (case p)) =
+  let _ , _ , p' = +-assoc-l pc p in
+  inj₁ (_ , s-case pc p , case→thread p')
+canonical-cut-alive (cc-delayed pc (join p)) =
+  let _ , _ , p' = +-assoc-l pc p in
+  inj₁ (_ , s-join pc p , join→thread p')
+canonical-cut-alive (cc-delayed pc (select-l p)) =
+  let _ , _ , p' = +-assoc-l pc p in
+  inj₁ (_ , s-select-l pc p , left→thread p')
+canonical-cut-alive (cc-delayed pc (select-r p)) =
+  let _ , _ , p' = +-assoc-l pc p in
+  inj₁ (_ , s-select-r pc p , right→thread p')
 canonical-cut-alive (cc-delayed p (fork-l q r)) =
-  let _ , p′ , q′ = +-assoc-l p q in
-  let _ , p′′ , r′ = +-assoc-l p′ r in
-  let _ , q′′ , r′′ = +-assoc-r r′ (+-comm p′′) in
-  inj₁ (_ , s-fork-l p q r , fork-thread q′ r′′)
+  let _ , p' , q' = +-assoc-l p q in
+  let _ , p'' , r' = +-assoc-l p' r in
+  let _ , q'' , r'' = +-assoc-r r' (+-comm p'') in
+  inj₁ (_ , s-fork-l p q r , fork→thread q' r'')
 canonical-cut-alive (cc-delayed p (fork-r q r)) =
-  let _ , p′ , q′ = +-assoc-l p q in
-  let _ , p′′ , r′ = +-assoc-l p′ r in
-  inj₁ (_ , s-fork-r p q r , fork-thread q′ r′)
-canonical-cut-alive (cc-delayed p (fail q)) =
-  let _ , p′ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-fail p q , fail-thread q′)
-canonical-cut-alive (cc-delayed p (client q)) =
-  let _ , _ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-client p q , client-thread q′)
-canonical-cut-alive (cc-delayed p (weaken q)) =
-  let _ , _ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-weaken p q , weaken-thread q′)
-canonical-cut-alive (cc-delayed p (contract q)) =
-  let _ , _ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-contract p q , contract-thread q′)
-canonical-cut-alive (cc-delayed p (ex q)) =
-  let _ , _ , q' = +-assoc-l p q in
-  inj₁ (_ , s-ex p q , ex-thread q')
-canonical-cut-alive (cc-delayed p (all q)) =
-  let _ , _ , q' = +-assoc-l p q in
-  inj₁ (_ , s-all p q , all-thread q')
-canonical-cut-alive (cc-delayed-server p (server q un) (server r un′)) with +-empty-l r
+  let _ , p' , q' = +-assoc-l p q in
+  let _ , p'' , r' = +-assoc-l p' r in
+  inj₁ (_ , s-fork-r p q r , fork→thread q' r')
+canonical-cut-alive (cc-delayed pc (all p)) =
+  let _ , _ , p' = +-assoc-l pc p in
+  inj₁ (_ , s-all pc p , all→thread p')
+canonical-cut-alive (cc-delayed pc (ex p)) =
+  let _ , _ , p' = +-assoc-l pc p in
+  inj₁ (_ , s-ex pc p , ex→thread p')
+canonical-cut-alive (cc-delayed pc (client p)) =
+  let _ , _ , p' = +-assoc-l pc p in
+  inj₁ (_ , s-client pc p , client→thread p')
+canonical-cut-alive (cc-delayed pc (weaken p)) =
+  let _ , _ , p' = +-assoc-l pc p in
+  inj₁ (_ , s-weaken pc p , weaken→thread p')
+canonical-cut-alive (cc-delayed pc (contract p)) =
+  let _ , _ , p' = +-assoc-l pc p in
+  inj₁ (_ , s-contract pc p , contract→thread p')
+canonical-cut-alive (cc-servers pc (server p un) (server q un')) with +-empty-l q
 ... | refl =
-  let _ , p′ , q′ = +-assoc-l p q in
-  inj₁ (_ , s-server p q r un un′ , server-thread q′ (+-un p′ un un′))
+  let _ , pc' , p' = +-assoc-l pc p in
+  inj₁ (_ , s-server pc p q un un' , server→thread p' (∗-un (un ⟨ pc' ⟩ un')))
 \end{code}
@@
 \begin{code}
-deadlock-freedom : ∀{Γ} (P : Process Γ) → Alive P
+deadlock-freedom : ∀{Γ} (P : Proc Γ) → Alive P
 \end{code}
@@
 \begin{code}[hide]
-deadlock-freedom (link p) = inj₁ (_ , s-refl , link-thread p)
-deadlock-freedom (fail p) = inj₁ (_ , s-refl , fail-thread p)
-deadlock-freedom close = inj₁ (_ , s-refl , close-thread)
-deadlock-freedom (wait p P) = inj₁ (_ , s-refl , wait-thread p)
-deadlock-freedom (left p P) = inj₁ (_ , s-refl , left-thread p)
-deadlock-freedom (right p P) = inj₁ (_ , s-refl , right-thread p)
-deadlock-freedom (case p P Q) = inj₁ (_ , s-refl , case-thread p)
-deadlock-freedom (fork p q P Q) = inj₁ (_ , s-refl , fork-thread p q)
-deadlock-freedom (join p P) = inj₁ (_ , s-refl , join-thread p)
-deadlock-freedom (server p un P) = inj₁ (_ , s-refl , server-thread p un)
-deadlock-freedom (client p P) = inj₁ (_ , s-refl , client-thread p)
-deadlock-freedom (weaken p P) = inj₁ (_ , s-refl , weaken-thread p)
-deadlock-freedom (contract p P) = inj₁ (_ , s-refl , contract-thread p)
-deadlock-freedom (ex p P) = inj₁ (_ , s-refl , ex-thread p)
-deadlock-freedom (all p F) = inj₁ (_ , s-refl , all-thread p)
-deadlock-freedom (cut p P Q) with deadlock-freedom P
+deadlock-freedom (link (ch ⟨ p ⟩ ch)) = inj₁ (_ , s-refl , link (link p))
+deadlock-freedom (fail (ch ⟨ p ⟩ _)) = inj₁ (_ , s-refl , fail→thread p)
+deadlock-freedom (wait (ch ⟨ p ⟩ _)) = inj₁ (_ , s-refl , wait→thread p)
+deadlock-freedom (close ch) = inj₁ (_ , s-refl , output close)
+deadlock-freedom (case (ch ⟨ p ⟩ _)) = inj₁ (_ , s-refl , case→thread p)
+deadlock-freedom (select (ch ⟨ p ⟩ inj₁ _)) = inj₁ (_ , s-refl , left→thread p)
+deadlock-freedom (select (ch ⟨ p ⟩ inj₂ _)) = inj₁ (_ , s-refl , right→thread p)
+deadlock-freedom (join (ch ⟨ p ⟩ _)) = inj₁ (_ , s-refl , join→thread p)
+deadlock-freedom (fork (ch ⟨ p ⟩ (P ⟨ q ⟩ Q))) = inj₁ (_ , s-refl , fork→thread p q)
+deadlock-freedom (all (ch ⟨ p ⟩ F)) = inj₁ (_ , s-refl , all→thread p)
+deadlock-freedom (ex (ch ⟨ p ⟩ P)) = inj₁ (_ , s-refl , ex→thread p)
+deadlock-freedom (server (ch ⟨ p ⟩ (un , _))) = inj₁ (_ , s-refl , server→thread p un)
+deadlock-freedom (client (ch ⟨ p ⟩ _)) = inj₁ (_ , s-refl , client→thread p)
+deadlock-freedom (weaken (ch ⟨ p ⟩ _)) = inj₁ (_ , s-refl , weaken→thread p)
+deadlock-freedom (contract (ch ⟨ p ⟩ _)) = inj₁ (_ , s-refl , contract→thread p)
+deadlock-freedom (cut (P ⟨ p ⟩ Q)) with deadlock-freedom P
 ... | inj₂ (_ , red) = inj₂ (_ , r-cut p red)

### 2025-12-13 paper/Safety.lagda.tex
restored safety

@@
 open import Data.List.Base using ([]; _∷_; [_])
+open import Relation.Unary
 
@@
   hole   : ReductionContext Δ Δ
-  cut-l  : ∀{Γ Γ₁ Γ₂ A} (p : Γ ≃ Γ₁ + Γ₂) →
-           ReductionContext Δ (A ∷ Γ₁) → Process (dual A ∷ Γ₂) →
-           ReductionContext Δ Γ
-  cut-r  : ∀{Γ Γ₁ Γ₂ A} (p : Γ ≃ Γ₁ + Γ₂) →
-           Process (A ∷ Γ₁) → ReductionContext Δ (dual A ∷ Γ₂) →
-           ReductionContext Δ Γ
+  cut-l  : ∀{A} → ∀[  ((A ∷_) ⊢ ReductionContext Δ) ∗ ((dual A ∷_) ⊢ Proc) ⇒
+                      ReductionContext Δ ]
+  cut-r  : ∀{A} → ∀[  ((A ∷_) ⊢ Proc) ∗ ((dual A ∷_) ⊢ ReductionContext Δ) ⇒
+                      ReductionContext Δ ]
 \end{code}
@@
 \begin{code}
-_⟦_⟧ : ∀{Γ Δ} → ReductionContext Δ Γ → Process Δ → Process Γ
+_⟦_⟧ : ∀{Γ Δ} → ReductionContext Δ Γ → Proc Δ → Proc Γ
 hole               ⟦ P ⟧ = P
-cut-l p 𝒞 Q ⟦ P ⟧  = cut p (𝒞 ⟦ P ⟧) Q
-cut-r p Q 𝒞 ⟦ P ⟧  = cut p Q (𝒞 ⟦ P ⟧)
+cut-l (𝒞 ⟨ p ⟩ Q)  ⟦ P ⟧ = cut ((𝒞 ⟦ P ⟧) ⟨ p ⟩ Q)
+cut-r (Q ⟨ p ⟩ 𝒞)  ⟦ P ⟧ = cut (Q ⟨ p ⟩ (𝒞 ⟦ P ⟧))
 \end{code}
@@
 \begin{code}
-WellFormed        : ∀{Γ} → Process Γ → Set
-WellFormed {Γ} P  = ∀{Δ} {𝒞 : ReductionContext Δ Γ} {Q : Process Δ} →
+WellFormed : ∀{Γ} → Proc Γ → Set
+WellFormed {Γ} P = ∀{Δ} {𝒞 : ReductionContext Δ Γ} {Q : Proc Δ} →
                    P ⊒ (𝒞 ⟦ Q ⟧) → Alive Q
@@
 \begin{code}
-type-safety : ∀{Γ} (P : Process Γ) → WellFormed P
+type-safety : ∀{Γ} (P : Proc Γ) → WellFormed P
 type-safety P {_} {_} {Q} _ = deadlock-freedom Q

### 2025-12-13 paper/Examples.lagda.tex
restored examples

@@
 {-# OPTIONS --rewriting #-}
-open import Data.Sum hiding (reduce; swap)
+open import Function using (_$_)
+open import Data.Sum using (inj₁; inj₂)
 open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
-open import Data.Nat using (ℕ; zero; suc)
 open import Data.Fin using (zero; suc; #_)
-open import Data.List.Base using (List; []; _∷_; [_]; _++_)
-open import Function using (_$_)
-open import Data.Maybe
+open import Data.List.Base using ([]; _∷_; [_])
+open import Relation.Unary
 
@@
 
-In this section we illustrate a few simple examples of processes. We start by
-encoding the type of a boolean value being produced as a choice between ``going
-left'' (which we interpret as true) and ``going right'' (which we interpret as
-false).
+In this section we illustrate a few simple examples of processes encoded in our
+formalization. We start by encoding the type of a boolean value being produced
+as a choice between ``going left'' (which we interpret as true) and ``going
+right'' (which we interpret as false).
 
@@
 \begin{code}
-true : Process [ 𝔹 ]
-true = left (< ≫) close
+true : Proc [ 𝔹 ]
+true = select (ch ⟨ < ≫ ⟩ inj₁ (close ch))
 
-false : Process [ 𝔹 ]
-false = right (< ≫) close
+false : Proc [ 𝔹 ]
+false = select (ch ⟨ < ≫ ⟩ inj₂ (close ch))
 \end{code}
@@
 \begin{code}
-if_else : ∀{Γ} → Process Γ → Process Γ → Process (dual 𝔹 ∷ Γ)
-if P else Q = case (< ≫) (wait (< ≫) P) (wait (< ≫) Q)
+if_else : ∀{Γ} → Proc Γ → Proc Γ → Proc (dual 𝔹 ∷ Γ)
+if P else Q = curry∗ case ch (< ≫) ( wait (ch ⟨ < ≫ ⟩ P)
+                                   , wait (ch ⟨ < ≫ ⟩ Q))
 \end{code}
@@
 typing context) and continues as either $P$ or $Q$ depending on whether it
-receives true or false.
-%
-We can then define a process $\AgdaFunction{drop}~P$ that waits for consumes a
-boolean and continues as $P$ regardless of its value.
+receives true or false. We use \AgdaFunction{curry∗} (defined in
+\Cref{sec:context-agda}) to curry the constructor
+\AgdaInductiveConstructor{case} so that we can supply its arguments one by one
+saving a few parentheses and reducing clutter.
+
+Next we define a process $\AgdaFunction{drop}~P$ that consumes a boolean and
+continues as $P$ regardless of its value.
 
 \begin{code}
-drop : ∀{Γ} → Process Γ → Process (dual 𝔹 ∷ Γ)
+drop : ∀{Γ} → Proc Γ → Proc (dual 𝔹 ∷ Γ)
 drop P = if P else P
@@
 \begin{code}
-!!_ : Process [ 𝔹 ] → Process [ 𝔹 ]
-!!_ B = cut ≫ B (if false else true)
+!!_ : Proc [ 𝔹 ] → Proc [ 𝔹 ]
+!!_ B = curry∗ cut B ≫ (if false else true)
 
-_&&_ _||_  : Process [ 𝔹 ] → Process [ 𝔹 ] → Process [ 𝔹 ]
-A && B   = cut ≫ A (cut ≫ B (if (link (< ≫)) else (drop false)))
+_&&_ _||_  : Proc [ 𝔹 ] → Proc [ 𝔹 ] → Proc [ 𝔹 ]
+A && B   =  curry∗ cut A ≫ $
+            curry∗ cut B ≫ $
+            if (curry∗ link ch (< ≫) ch) else (drop false)
 A || B   = !! ((!! A) && (!! B))
@@
 
+The function \AgdaFunction{\$} is defined in Agda's standard library and denotes
+function application. We use it here as ``separator'' to save a few parentheses.
+%
 Note the use of \AgdaInductiveConstructor{cut}s for combining a process that
@@
 
-If we wish to test our definitions, we can implement a simple (lazy) evaluator
-using the deadlock freedom property. We have not proved a termination result,
-but since linear logic enjoys cut elimination we can safely annotate the
-evaluator as terminating.
+If we wish to test our definitions, we can implement a simple evaluator using
+the deadlock freedom property. We have not proved a termination result, but
+since linear logic enjoys cut elimination we can safely annotate the evaluator
+as terminating.
 
@@
 {-# TERMINATING #-}
-eval : ∀{Γ} → Process Γ → Process Γ
+eval : ∀{Γ} → Proc Γ → Proc Γ
 eval P with deadlock-freedom P
@@
 we obtain
-$\AgdaInductiveConstructor{right}~(\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{•})~\AgdaInductiveConstructor{close}$,
+$\AgdaInductiveConstructor{select}~(\AgdaInductiveConstructor{ch~⟨~<~•~⟩~inj₂}~(\AgdaInductiveConstructor{close~ch}))$,
 that is the definition of \AgdaFunction{false}, as expected.
@@
 library and creates an element of \AgdaDatatype{Fin} from the natural number
-$n$. The function \AgdaFunction{\$} is also defined in the standard library and
-denotes function application. We use it here as ``command separator'' to save a
-few parentheses.
+$n$.
 
@@
 echo : let X = var (# 0) in
-        Process [ `! (`∀ (X ⊸ X)) ]
-echo =  server (< ≫) un-[] $
-        all (< ≫) λ X →
-        join (< ≫) $
-        link (< ≫)
+       Proc [ `! (`∀ (X ⊸ X)) ]
+echo = curry∗ server ch (< ≫)
+             ( un-[]
+             , curry∗ all ch (< ≫) λ X →
+               curry∗ join ch (< ≫) $
+               curry∗ link ch (< ≫) ch)
 \end{code}
@@
 \begin{code}
-echo-true : Process [ 𝔹 ]
-echo-true = cut ≫ echo (client (< ≫) $
-                       ex (< ≫) $
-                       fork (< ≫) ≫ true (link (< ≫)))
+echo-true : Proc [ 𝔹 ]
+echo-true = curry∗ cut echo ≫ $
+            curry∗ client ch (< ≫) $
+            curry∗ ex ch (< ≫) $
+            curry∗ fork ch (< ≫) $ true ⟨ ≫ ⟩ curry∗ link ch (< ≫) ch
 \end{code}
 
-If we ask Agda to evaluate this process we obtain
-$\AgdaInductiveConstructor{left}~(\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{•})~\AgdaInductiveConstructor{close}$,
-that is the definition of \AgdaFunction{true} as expected.
+If we ask Agda to evaluate this process we obtain \AgdaFunction{true}, as
+expected.
 
@@
 rather than implementing a user-friendly DSL for it.
-
-% \begin{code}
-% ⊗-comm : let X = var (# 1) in
-%          let Y = var (# 0) in
-%          Process [ `∀ (`∀ ((X ⊗ Y) ⊸ (Y ⊗ X))) ]
-% ⊗-comm = all (< ≫) λ X →
-%          all (< ≫) λ Y →
-%          join (< ≫) $
-%          join (> < ≫) $
-%          fork (> > < ≫) (< ≫)
-%               (link (< ≫))
-%               (link (< ≫))
-% \end{code}

### 2025-12-13 paper/Context.lagda.tex
cleanup

@@
 
-\begin{table}
+\begin{table}[t]
   \begin{tabular}{|lll|}
@@
 %
-Inspired by separation logic~\cite{Hoare} and
+Inspired by separation logic~\cite{OHearnPym99} and
 following~\citet{RouvoetPoulsenKrebbersVisser20}, we define the \emph{separating

### 2025-12-13 paper/Process.lagda.tex
cleanup

@@
   operational semantics of processes and need not require separate proofs
-  (\Cref{sec:agda-semantics}).
+  (\Cref{sec:congruence-agda,sec:reduction-agda}).
 \end{itemize}

### 2025-12-13 paper/calculus.tex
restored congruence
cleanup

@@
 
-\begin{table}
+\begin{table}[t]
     \caption{Syntax of \Calculus.}\label{tab:syntax}
@@
 every message sent on one of the channels is forwarded to the other. As
-discussed in the literature~\cite{LindleyMorris}, this form is also useful for
+discussed in the literature~\cite{LindleyMorris15}, this form is also useful for
 modeling \emph{delegation}, that is the exchange of an existing session endpoint
@@
 
-\begin{table}
+\begin{table}[t]
     \caption{Operational semantics of \Calculus.}
@@
             \\
+            \SServer &
+            \Cut\x{\Server\y\u.P}{\Server\x\v.Q} & \pcong & \Server\y\u.\Cut\x{P}{\Server\x\v.Q} % & x \ne y,u
+            \\
             \SClient &
@@
             \Cut\x{\Contract\y\u\v.P}{Q} & \pcong & \Contract\y\u\v.\Cut\x{P}{Q} % & x \ne y,u,v
-            \\
-            \SServer &
-            \Cut\x{\Server\y\u.P}{\Server\x\v.Q} & \pcong & \Server\y\u.\Cut\x{P}{\Server\x\v.Q} % & x \ne y,u
             \\
@@
 
-\begin{table}
+\begin{table}[t]
     \caption{Typing rules of \Calculus.}

### 2025-12-16 paper/benchmarks/count.sh
benchmarks

@@ -1,2 +1,0 @@
-#!/bin/sh
-exec cat `find $1 -name "*.*"` | tr -d [:space:] | wc -c

### 2025-12-16 paper/related.tex
related work
more related
++
cleanup
benchmarks

@@
 \label{sec:related}
+
+\Cref{tab:comparison} summarizes a comparison among the formalizations of binary
+session calculus that we are aware of, roughly sorted by publication date. We
+have selected those formalizations that define a calculus of binary sessions
+equipped with a small-step reduction semantics and prove subject reduction and
+possibly deadlock freedom. \fullcite{RouvoetPoulsenKrebbersVisser20} also
+formalize a subset of \CP as case study for their approach based on separation
+logic, which we have partially taken inspiration from for the representation of
+processes (\cref{sec:context-agda,sec:process-agda}). Their formalization adopts
+the same intrinsically-typed representation of processes with nameless channels
+as our own, but their formalization focuses on the implementation of an
+interpreter (no small-step semantics is defined) and subject reduction is
 
@@
 squeezed spazi (in generale sono significativi)
+
+\begin{table}[t]
+    \newcommand{\textlabel}[1]{\rotatebox{90}{#1}}
+    \newcommand{\CHECK}{\checkmark}
+    \newcommand{\SessionPI}{session $\pi$}
+    \newcommand{\SessionLAM}{session $\lambda$}
+    \caption{\label{tab:comparison} Comparison of different formalizations of binary session calculi.}
+    \begin{tabular}{@{}r|c|ccc|rr|rr|r@{}}
+        \textbf{Formalization} &
+        \textlabel{Proof assistant} &
+        \textlabel{Calculus} &
+        \textlabel{Shared channels} &
+        \textlabel{Polymorphism} &
+        \textlabel{Subject reduction} &
+        \textlabel{Library} &
+        \textlabel{Deadlock freedom} &
+        \textlabel{Library} &
+        \textlabel{\bf Total size}
+        \\
+        \hline
+        \fullcite{GotoEtAl16} &
+        Coq &
+        \SessionPI & 
+        \CHECK &
+        \CHECK &
+        543 &
+        -- &
+        -- &
+        -- &
+        \textbf{563}
+        \\
+        \fullcite{Thiemann19} &
+        Agda &
+        \SessionLAM &
+        \CHECK &
+        -- &
+        177 &
+        -- &
+        -- &
+        -- &
+        \textbf{177}
+        \\
+        \fullcite{CastroFerreiraYoshida20} &
+        Coq &
+        \SessionPI &
+        \CHECK &
+        -- &
+        204 &
+        -- &
+        -- &
+        -- &
+        \textbf{204}
+        \\
+        \fullcite{CicconePadovani20} &
+        Agda &
+        linear $\pi$ & 
+        \CHECK &
+        -- &
+        77 &
+        -- &
+        -- &
+        -- &
+        \textbf{77}
+        \\
+        \fullcite{RouvoetPoulsenKrebbersVisser20} &
+        Agda &
+        \CP &
+        -- &
+        -- &
+        27 &
+        55 &
+        -- &
+        -- &
+        \textbf{82}
+        \\
+        \fullcite{ZalakainDardha21} &
+        Agda &
+        graded $\pi$ &
+        \CHECK &
+        -- &
+        82 &
+        8 &
+        -- &
+        -- &
+        \textbf{90}
+        \\
+        \fullcite{JacobsBalzerKrebbers22} &
+        Coq &
+        \SessionLAM &
+        -- &
+        -- &
+        68 &
+        -- &
+        25 &
+        171 &
+        \textbf{264}
+        \\
+        \fullcite{SanoKavanaghPientka23} &
+        Beluga &
+        S\CP &
+        -- &
+        -- &
+        35 &
+        -- &
+        -- &
+        -- &
+        \textbf{35}
+        \\
+        \fullcite{ZackonEtAl25} &
+        Beluga &
+        \CP &
+        -- &
+        -- &
+        56 &
+        73 &
+        -- &
+        -- &
+        \textbf{129}
+        \\
+        this work &
+        Agda &
+        linear \CP &
+        \CHECK &
+        \CHECK &
+        21 &
+        -- &
+        15 &
+        -- &
+        \textbf{36}
+    \end{tabular}
+\end{table}
+
+paper di Marco a ECOOP (multiparty)

### 2025-12-16 paper/related.tex
notes for table

@@
     \caption{\label{tab:comparison} Comparison of different formalizations of binary session calculi.}
-    \begin{tabular}{@{}r|c|ccc|rr|rr|r@{}}
+    \begin{minipage}{.9\textwidth}
+    \begin{tabular}{@{}r|c|ccc|rr|rr|r@{~}l@{}}
         \textbf{Formalization} &
@@
         \textlabel{Library} &
-        \textlabel{\bf Total size}
+        \textlabel{\bf Total size} &
         \\
@@
         -- &
-        \textbf{563}
+        \textbf{563} &
+        \footnote{Excluding safety results.}
         \\
@@
         -- &
-        \textbf{177}
+        \textbf{177} &
+        \footnote{Including support for subtyping and the interpreter.}
         \\
@@
         -- &
-        \textbf{204}
+        \textbf{204} &
+        \footnote{Excluding the original syntax.}
         \\
@@
         -- &
-        \textbf{77}
+        \textbf{77} &
+        \footnote{Including support for dependent session types.}
         \\
@@
         -- &
-        \textbf{82}
+        \textbf{82} &
+        \footnote{Including type-preserving evaluator and library for proof-relevant separation algebra.}
         \\
@@
         -- &
-        \textbf{90}
+        \textbf{90} &
+        \footnote{Including library for algebra of types.}
         \\
@@
         171 &
-        \textbf{264}
+        \textbf{264} &
+        \footnote{Including deadlock freedom and the library for connectivity graphs.}
         \\
@@
         -- &
-        \textbf{129}
+        \textbf{129} &
+        \footnote{Excluding the correspondence between \CP and S\CP.}
         \\
@@
         -- &
-        \textbf{36}
+        \textbf{36} &
+        \footnote{Including deadlock freedom and excluding safety results.}
     \end{tabular}
+    \end{minipage}
 \end{table}

### 2025-12-17 paper/related.tex
more info

--- 2025-12-04 paper/related.tex
@@
 \label{sec:related}
+
+\Cref{tab:comparison} summarizes a comparison among the formalizations of binary
+session calculus that we are aware of, roughly sorted by publication date. We
+have selected those formalizations that define a calculus of binary sessions
+equipped with a reduction/evaluation semantics with proofs of subject reduction
+and possibly of deadlock freedom.
+%
+The formalizations differ substantially in terms of reference calculus, used
+proof assistant, process/expression/channel representation, handling of
+linearity.
+%
+Most importantly, almost all of them predates the publlication of the linearity
+challenge~\cite{CarboneEtAl24} hence were not necessarily targeted to solving
+that particular problem. Nonetheless, they provide a broad spectrum of
+techniques
+
+\fullcite{GotoEtAl16} describe the oldest formalization of a session-based
+calculus for which we were able to access the source code (the formalization of
+\citet{Gay01} predates this work, but the source code is no longer publicly
+available).
+
+fullcite{RouvoetPoulsenKrebbersVisser20} also formalize a subset of \CP as case
+study for their approach based on separation logic, which we have partially
+taken inspiration from for the representation of processes
+(\cref{sec:context-agda,sec:process-agda}). Their formalization adopts the same
+intrinsically-typed representation of processes with nameless channels as our
+own, but their formalization focuses on the implementation of an interpreter (no
+small-step semantics is defined) and subject reduction is
 
@@
 squeezed spazi (in generale sono significativi)
+
+\begin{table}[t]
+    \footnotesize
+    \newcommand{\textlabel}[1]{\rotatebox{90}{#1}}
+    \newcommand{\YES}{$+$}
+    \newcommand{\NO}{$-$}
+    \newcommand{\MAYBE}{$\pm$}
+    \caption{\label{tab:comparison} A summary of different formalizations of binary session calculi. All sizes are in kb.}
+    \begin{minipage}{.9\textwidth}
+    \begin{tabular}{@{}r@{~}|@{~}c@{~}|@{~}c@{~~}c@{~~}c@{~~}c@{~~}c@{~~}c@{~~}c@{~}|@{~}r@{~~}r@{~}|@{~}r@{~~}r@{~}|@{~}r@{~}l@{}}
+        \textbf{Formalization} &
+        \textlabel{Proof assistant} &
+        \textlabel{Calculus} &
+        \textlabel{Intrinsically typed} &
+        \textlabel{Channel linearity} &
+        \textlabel{Channel names} &
+        \textlabel{Continuations} &
+        \textlabel{Shared channels} &
+        \textlabel{Polymorphism} &
+        \textlabel{Subject reduction} &
+        \textlabel{Library} &
+        \textlabel{Deadlock freedom} &
+        \textlabel{Library} &
+        \textlabel{\bf Total size} &
+        \\
+        \hline
+        \fullcite{GotoEtAl16} &
+        Coq &
+        \SessionPI &
+        \NO &
+        splitting &
+        loc. nameless &
+        \NO &
+        \YES &
+        \YES &
+        543 &
+        \NO &
+        \NO &
+        \NO &
+        \textbf{543} &
+        \footnote{Excluding safety results.}
+        \\
+        \fullcite{Thiemann19} &
+        Agda &
+        \SessionLAM &
+        \YES &
+        splitting &
+        co-de Bruijn &
+        \MAYBE &
+        \YES &
+        \NO &
+        177 &
+        \NO &
+        \NO &
+        \NO &
+        \textbf{177} &
+        \footnote{Including support for recursive types, subtyping and the interpreter.}
+        \\
+        \fullcite{CastroFerreiraYoshida20} &
+        Coq &
+        \SessionPI &
+        \NO &
+        splitting &
+        loc. nameless &
+        \NO &
+        \YES &
+        \NO &
+        204 &
+        \NO &
+        \NO &
+        \NO &
+        \textbf{204} &
+        \footnote{Excluding the original syntax.}
+        \\
+        \fullcite{CicconePadovani20} &
+        Agda &
+        \LinearPI & 
+        \YES &
+        splitting &
+        co-de Bruijn &
+        \YES &
+        \YES &
+        \NO &
+        77 &
+        \NO &
+        \NO &
+        \NO &
+        \textbf{77} &
+        \footnote{Including support for recursive and dependent session types.}
+        \\
+        \fullcite{RouvoetPoulsenKrebbersVisser20} &
+        Agda &
+        \CP &
+        \YES &
+        splitting &
+        co-de Bruijn &
+        \YES &
+        \NO &
+        \NO &
+        27 &
+        55 &
+        \NO &
+        \NO &
+        \textbf{82} &
+        \footnote{Including type-preserving evaluator and library for proof-relevant separation algebra.}
+        \\
+        \fullcite{ZalakainDardha21} &
+        Agda &
+        \GradedPI &
+        \NO &
+        leftovers &
+        de Bruijn &
+        \YES &
+        \YES &
+        \NO &
+        82 &
+        8 &
+        \NO &
+        \NO &
+        \textbf{90} &
+        \footnote{Including library for algebra of types.}
+        \\
+        \fullcite{JacobsBalzerKrebbers22} &
+        Coq &
+        \SessionLAM &
+        \NO &
+        splitting &
+        named &
+        \MAYBE &
+        \NO &
+        \NO &
+        68 &
+        \NO &
+        25 &
+        171 &
+        \textbf{264} &
+        \footnote{Including deadlock freedom and the library for connectivity graphs.}
+        \\
+        \fullcite{SanoKavanaghPientka23} &
+        Beluga &
+        \SCP &
+        \NO &
+        predicates &
+        HOAS &
+        \MAYBE &
+        \NO &
+        \NO &
+        35 &
+        \NO &
+        \NO &
+        \NO &
+        \textbf{35}
+        \\
+        \fullcite{ZackonEtAl25} &
+        Beluga &
+        \CP &
+        \NO &
+        tags &
+        HOAS &
+        \NO &
+        \NO &
+        \NO &
+        56 &
+        73 &
+        \NO &
+        \NO &
+        \textbf{129} &
+        \footnote{Excluding the correspondence between \CP and \SCP.}
+        \\
+        this work &
+        Agda &
+        \Calculus &
+        \YES &
+        splitting &
+        co-de Bruijn &
+        \YES &
+        \YES &
+        \YES &
+        21 &
+        \NO &
+        15 &
+        \NO &
+        \textbf{36} &
+        \footnote{Including deadlock freedom and excluding safety results.}
+    \end{tabular}
+    \end{minipage}
+\end{table}
+
+paper di Marco a ECOOP (multiparty)

### 2025-12-18 paper/related.tex
related

@@
 \Cref{tab:comparison} summarizes a comparison among the formalizations of binary
-session calculus that we are aware of, roughly sorted by publication date. We
-have selected those formalizations that define a calculus of binary sessions
-equipped with a reduction/evaluation semantics with proofs of subject reduction
-and possibly of deadlock freedom.
-%
+session calculi and of the linear $\pi$-calculus, sorted by publication date. We
+have selected those formalizations that include the following characteristics:
+\begin{itemize}
+\item syntax and reduction/evaluation semantics for a calculus with support for
+    binary sessions, either natively or through their encoding in terms of
+    linear channels~\cite{KobayashiPierceTurner99,Kobayashi02b};
+\item a type system along with the usual typing preservation results;
+\item a proof of global progress/deadlock freedom, for those type systems that
+    enforce it.
+\end{itemize}
+
 The formalizations differ substantially in terms of reference calculus, used
@@
 %
-Most importantly, almost all of them predates the publlication of the linearity
-challenge~\cite{CarboneEtAl24} hence were not necessarily targeted to solving
-that particular problem. Nonetheless, they provide a broad spectrum of
-techniques
+Most importantly, almost all of them predate the publication of the linearity
+challenge~\citep{CarboneEtAl24}, hence they are not necessarily targeted to
+solving that particular problem.
 
-\fullcite{GotoEtAl16} describe the oldest formalization of a session-based
-calculus for which we were able to access the source code (the formalization of
-\citet{Gay01} predates this work, but the source code is no longer publicly
-available).
+\citet{GotoEtAl16} describe the formalization of a session-based variant of the
+$\pi$-calculus which supports channel polymorphism. This is the oldest
+formalization of a session-based calculus for which we were able to retrieve the
+source code (the work of \citet{Gay01} predates this one, but its source code is
+not publicly available any more).
 
-fullcite{RouvoetPoulsenKrebbersVisser20} also formalize a subset of \CP as case
-study for their approach based on separation logic, which we have partially
-taken inspiration from for the representation of processes
-(\cref{sec:context-agda,sec:process-agda}). Their formalization adopts the same
-intrinsically-typed representation of processes with nameless channels as our
-own, but their formalization focuses on the implementation of an interpreter (no
-small-step semantics is defined) and subject reduction is
+\citet{Thiemann19} formalizes a subset of \GV~\citep{GayVasconcelos10}, a
+functional language extended with session communication primitives, along with
+an interpreter.
+
+\citet{RouvoetPoulsenKrebbersVisser20} present a library of abstractions
+inspired to separation logic aiding the formalization of interpreters for
+computing models supported by linear type systems. One of the presented case
+studies is the formalization of a fragment of \GV~\citep{Wadler14}.
+
+% as case study for their approach based on separation logic, which we have
+% partially taken inspiration from for the representation of processes
+% (\cref{sec:context-agda,sec:process-agda}). Their formalization adopts the same
+% intrinsically-typed representation of processes with nameless channels as our
+% own, but their formalization focuses on the implementation of an interpreter (no
+% small-step semantics is defined) and subject reduction is
+
+\citet{CastroFerreiraYoshida20} describe EMTST, a library for the formalization
+of session type systems which includes as case studies the session calculus of
+\citet{HondaVasconcelosKubo98} (called ``original system'') and a revised
+version of it that is more amenable to be formalized using a locally nameless
+representation of channels.
+
+\citet{CicconePadovani20} formalize a variant of the linear
+$\pi$-calculus~\cite{KobayashiPierceTurner99} that supports dependent types, so
+as to enable the description of communication protocols whose structure depends
+on the content of exchanged messages.
+
+All the formalizations mentioned make use of context splitting. In contrast,
+\citet{ZalakainDardha21} formalize a generalization of the linear $\pi$-calculus
+which is parametric in a \emph{usage algebra} (to account for channel
+sharing/linearity) and that is based on \emph{leftover typing}~\cite{}. The
+basic idea is that a process is typed with respect to an \emph{input context},
+which describes the overall available channels, and an \emph{output context},
+which describes the remaining channels not consumed by the process. In this way,
+it is possible to ``concatenate'' typing judgments instead of splitting them
+beforehand.
+
+Another work based on context splitting is the one by
+\citet{JacobsBalzerKrebbers22}, who formalize a library of \emph{connectivity
+graphs} for reasoning and enforcing deadlock freedom in a variant of
+\GV~\citep{GayVasconcelos10,Wadler14}. This is the first formalization of
+deadlock freedom for a calculus of sessions.
+
+Motivated by the technical difficulties arising from the splitting of typing
+contexts, \citet{SanoKavanaghPientka23} propose an alternative approach based on
+\emph{linearity predicates}. The key idea is to treat typing contexts
+structurally, and to enforce the linear usage of channels by means of separate
+predicates checking the syntactical occurrence of channels.
+
+\citet{ZackonEtAl25} describe yet another approach to the handling of typing
+contexts whereby channels are associated not just with a type but also with a
+\emph{tag} (or \emph{multiplicity}) from a given resource algebra for describing
+their availability. This approach streamlines context splitting in the sense
+that contexts can be treated in an essentially structural way, except for tags
+which are conveninently combined using operations from the resource algebra.
 
@@
 \begin{table}[t]
-    \footnotesize
     \newcommand{\textlabel}[1]{\rotatebox{90}{#1}}
@@
     \begin{minipage}{.9\textwidth}
-    \begin{tabular}{@{}r@{~}|@{~}c@{~}|@{~}c@{~~}c@{~~}c@{~~}c@{~~}c@{~~}c@{~~}c@{~}|@{~}r@{~~}r@{~}|@{~}r@{~~}r@{~}|@{~}r@{~}l@{}}
+    \begin{tabular}{@{}r@{~}|@{~}c@{~}|@{~}c@{~~}c@{~~}c@{~~}c@{~~}c@{~}|@{~}r@{~~}r@{~}|@{~}r@{~~}r@{~}|@{~}r@{~}l@{}}
         \textbf{Formalization} &
@@
         \textlabel{Calculus} &
-        \textlabel{Intrinsically typed} &
         \textlabel{Channel linearity} &
         \textlabel{Channel names} &
+        \textlabel{Intrinsically typed} &
         \textlabel{Continuations} &
-        \textlabel{Shared channels} &
-        \textlabel{Polymorphism} &
+        % \textlabel{Shared channels} &
+        % \textlabel{Polymorphism} &
         \textlabel{Subject reduction} &
@@
         \hline
-        \fullcite{GotoEtAl16} &
+        \citet{GotoEtAl16} &
         Coq &
         \SessionPI &
-        \NO &
         splitting &
@@
         \NO &
-        \YES &
-        \YES &
+        \NO &
+        % \YES &
+        % \YES &
         543 &
@@
         \textbf{543} &
-        \footnote{Excluding safety results.}
+        \footnote{Including shared and polymorphic channels, excluding safety results.}
         \\
-        \fullcite{Thiemann19} &
+        \citet{Thiemann19} &
         Agda &
-        \SessionLAM &
-        \YES &
+        \GV &
         splitting &
         co-de Bruijn &
+        \YES &
         \MAYBE &
-        \YES &
-        \NO &
+        % \YES &
+        % \NO &
         177 &
@@
         \textbf{177} &
-        \footnote{Including support for recursive types, subtyping and the interpreter.}
+        \footnote{Including shared channels, recursive types, subtyping and the interpreter.}
         \\
-        \fullcite{CastroFerreiraYoshida20} &
-        Coq &
-        \SessionPI &
-        \NO &
-        splitting &
-        loc. nameless &
-        \NO &
-        \YES &
-        \NO &
-        204 &
-        \NO &
-        \NO &
-        \NO &
-        \textbf{204} &
-        \footnote{Excluding the original syntax.}
-        \\
-        \fullcite{CicconePadovani20} &
+        \citet{RouvoetPoulsenKrebbersVisser20} &
         Agda &
-        \LinearPI & 
-        \YES &
+        \GV &
         splitting &
@@
         \YES &
-        \YES &
-        \NO &
-        77 &
-        \NO &
-        \NO &
-        \NO &
-        \textbf{77} &
-        \footnote{Including support for recursive and dependent session types.}
-        \\
-        \fullcite{RouvoetPoulsenKrebbersVisser20} &
-        Agda &
-        \CP &
-        \YES &
-        splitting &
-        co-de Bruijn &
-        \YES &
-        \NO &
-        \NO &
+        \MAYBE &
+        % \NO &
+        % \NO &
         27 &
@@
         \\
-        \fullcite{ZalakainDardha21} &
+        \citet{CastroFerreiraYoshida20} &
+        Coq &
+        \SessionPI &
+        splitting &
+        loc. nameless &
+        \NO &
+        \NO &
+        % \YES &
+        % \NO &
+        204 &
+        \NO &
+        \NO &
+        \NO &
+        \textbf{204} &
+        \footnote{Including shared channels, excluding the original syntax.}
+        \\
+        \citet{CicconePadovani20} &
         Agda &
-        \GradedPI &
+        \LinearPI & 
+        splitting &
+        co-de Bruijn &
+        \YES &
+        \YES &
+        % \YES &
+        % \NO &
+        77 &
         \NO &
+        \NO &
+        \NO &
+        \textbf{77} &
+        \footnote{Including shared channels, recursive and dependent session types.}
+        \\
+        \citet{ZalakainDardha21} &
+        Agda &
+        \LinearPI &
         leftovers &
         de Bruijn &
+        \NO &
         \YES &
-        \YES &
-        \NO &
+        % \YES &
+        % \NO &
         82 &
@@
         \textbf{90} &
-        \footnote{Including library for algebra of types.}
+        \footnote{Including shared channels and the library for algebra of types.}
         \\
-        \fullcite{JacobsBalzerKrebbers22} &
+        \citet{JacobsBalzerKrebbers22} &
         Coq &
-        \SessionLAM &
-        \NO &
+        \GV &
         splitting &
         named &
+        \NO &
         \MAYBE &
-        \NO &
-        \NO &
+        % \NO &
+        % \NO &
         68 &
@@
         \\
-        \fullcite{SanoKavanaghPientka23} &
+        \citet{SanoKavanaghPientka23} &
         Beluga &
         \SCP &
-        \NO &
         predicates &
         HOAS &
+        \NO &
         \MAYBE &
-        \NO &
-        \NO &
+        % \NO &
+        % \NO &
         35 &
@@
         \\
-        \fullcite{ZackonEtAl25} &
+        \citet{ZackonEtAl25} &
         Beluga &
         \CP &
-        \NO &
         tags &
@@
         \NO &
-        \NO &
+        % \NO &
+        % \NO &
         56 &
@@
         \Calculus &
-        \YES &
         splitting &
@@
         \YES &
-        \YES &
+        % \YES &
+        % \YES &
         21 &
@@
         \textbf{36} &
-        \footnote{Including deadlock freedom and excluding safety results.}
+        \footnote{Including shared, polymorphic channels and deadlock freedom, excluding safety results.}
     \end{tabular}
@@
 
+instead of counting line numbers, which may be affected by indentation habits,
+readability issues, and other syntactical restrictions of the proof assistant
+being used, we count characters after spaces have been
+squeezed.\footnote{Sequences of two or more consecutive space-like characters
+are collapsed into a single space. The squeezing is obtained by running the
+command \texttt{tr -s [:space:] file} on Unix-like systems.}
+
 paper di Marco a ECOOP (multiparty)

### 2025-12-19 paper/related.tex
progress on related

@@
 
-\Cref{tab:comparison} summarizes a comparison among the formalizations of binary
-session calculi and of the linear $\pi$-calculus, sorted by publication date. We
-have selected those formalizations that include the following characteristics:
+We have compared our formalization with others that have appeared in the
+literature and for which we were able to retrieve the source code. We have
+looked at those formalizations that include the following characteristics:
 \begin{itemize}
@@
 \end{itemize}
-
-The formalizations differ substantially in terms of reference calculus, used
-proof assistant, process/expression/channel representation, handling of
-linearity.
-%
-Most importantly, almost all of them predate the publication of the linearity
-challenge~\citep{CarboneEtAl24}, hence they are not necessarily targeted to
-solving that particular problem.
 
@@
 \emph{linearity predicates}. The key idea is to treat typing contexts
-structurally, and to enforce the linear usage of channels by means of separate
-predicates checking the syntactical occurrence of channels.
+structurally and to enforce the linear usage of channels by checking the
+syntactical occurrence of channels within processes.
 
@@
 
-come sono stati misurati
-
-eliminati commenti
-eliminati esempi
-squeezed spazi (in generale sono significativi)
+\newcommand{\YES}{$+$}
+\newcommand{\NO}{$-$}
+\newcommand{\MAYBE}{$\pm$}
 
@@
     \newcommand{\textlabel}[1]{\rotatebox{90}{#1}}
-    \newcommand{\YES}{$+$}
-    \newcommand{\NO}{$-$}
-    \newcommand{\MAYBE}{$\pm$}
     \caption{\label{tab:comparison} A summary of different formalizations of binary session calculi. All sizes are in kb.}
     \begin{minipage}{.9\textwidth}
-    \begin{tabular}{@{}r@{~}|@{~}c@{~}|@{~}c@{~~}c@{~~}c@{~~}c@{~~}c@{~}|@{~}r@{~~}r@{~}|@{~}r@{~~}r@{~}|@{~}r@{~}l@{}}
+    \begin{tabular}{@{}r@{~}@{~}c@{~~}c@{~}|@{~}c@{~~}c@{~~}c@{~~}c@{~}|@{~}r@{~~}r@{~~}r@{~~}r@{~~}r@{~}l@{}}
         \textbf{Formalization} &
@@
         \textbf{543} &
-        \footnote{Including shared and polymorphic channels, excluding safety results.}
+        \footnote{Includes shared and polymorphic channels. Excluded safety results.}
         \\
@@
         \textbf{177} &
-        \footnote{Including shared channels, recursive types, subtyping and the interpreter.}
+        \footnote{Includes shared channels, recursive types, subtyping and the interpreter.}
         \\
@@
         \textbf{82} &
-        \footnote{Including type-preserving evaluator and library for proof-relevant separation algebra.}
+        \footnote{Includes type-preserving evaluator and library for proof-relevant separation algebra.}
         \\
@@
         \textbf{204} &
-        \footnote{Including shared channels, excluding the original syntax.}
+        \footnote{Includes shared channels. Excluded original syntax.}
         \\
@@
         \textbf{77} &
-        \footnote{Including shared channels, recursive and dependent session types.}
+        \footnote{Includes shared channels, recursive and dependent session types.}
         \\
@@
         \textbf{90} &
-        \footnote{Including shared channels and the library for algebra of types.}
+        \footnote{Includes shared channels and the library for algebra of types.}
         \\
@@
         \textbf{264} &
-        \footnote{Including deadlock freedom and the library for connectivity graphs.}
+        \footnote{Includes deadlock freedom and the library for connectivity graphs.}
         \\
@@
         \textbf{129} &
-        \footnote{Excluding the correspondence between \CP and \SCP.}
+        \footnote{Excluded correspondence between \CP and \SCP.}
         \\
@@
         \textbf{36} &
-        \footnote{Including shared, polymorphic channels and deadlock freedom, excluding safety results.}
+        \footnote{Includes shared, polymorphic channels and deadlock freedom. Excluded safety results.}
     \end{tabular}
@@
 
-instead of counting line numbers, which may be affected by indentation habits,
-readability issues, and other syntactical restrictions of the proof assistant
-being used, we count characters after spaces have been
-squeezed.\footnote{Sequences of two or more consecutive space-like characters
-are collapsed into a single space. The squeezing is obtained by running the
-command \texttt{tr -s [:space:] file} on Unix-like systems.}
+\Cref{tab:comparison} summarizes a comparison of our formalization with the
+aforementioned ones. The three leftmost colums list the reference papers (sorted
+by publication date), the proof assistant and the formalized calculus.
+%
+For simplicity, we have grouped the calculi into classes identified by an
+acronym. Some acronyms, like \CP~\cite{Wadler14} and
+\GV~\cite{GayVasconcelos10,Wadler14} are well known in the literature on session
+types. It should be noted that the actual calculus being formalized usually
+differs from (typically, is a strict subset of) the one identified by the
+acronym, but it retains some of its key features. We use \SessionPI to denote
+(variants of) the session-based $\pi$-calculus presented by
+\citet{HondaVasconcelosKubo98} and \LinearPI to denote (variants of) the linear
+$\pi$-calculus~\cite{KobayashiPierceTurner99}, into which binary sessions can be
+encoded using explicit continuation channels as described by
+\citet{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}. Finally, we use \SCP
+for the structural version of \CP introduced by \citet{SanoKavanaghPientka23}
+and \Calculus for our calculus, a linear version of \CP with explicit
+continuations.
+
+The four middle columns of \Cref{tab:comparison} report the relevant qualitative
+aspects of the formalizations, namely the technique used to enforce the linear
+usage of channels, the technique used to encode channels, whether the
+formalization is intrinsically or extrinsically typed, and whether the calculus
+makes use of continuation channels. Concerning the use of continuations, the
+approaches based directly on the linear $\pi$-calculus (into which sessions can
+be encoded) are marked with {\YES} and those based on a calculus with native
+sessions are marked with {\NO}, as expected. The calculus \SCP is marked with
+{\MAYBE} because, while not being strictly based on the linear $\pi$-calculus,
+it makes use of explicit continuations for defining the linearity predicates
+that check the linear usage of channels. Finally, all the approaches based on
+\GV are also marked with \MAYBE. In principle, \emph{none} of these calculi
+makes use of continuations. However, \GV is designed in such a way that each
+session operation acting on a channel $s$ returns the result of the operation
+(if any) \emph{along with the same channel $s$}. In this way, the type of $s$
+can be ``updated'' to take into account the effect of the operation. As observed
+by \citet{Padovani17}, this way of modelling the communication primitives is
+virtually indistinguishable from one making use of explicit continuation
+channels.
+
+The rightmost columns of \Cref{tab:comparison} report the \emph{size} (in
+kilobytes) of the formalizations as rough (and possibly questionable) estimates
+of their complexity. Instead of counting line numbers, which may depend on code
+indentation styles and syntactical constraints of the proof assistant being
+used, we have counted the number of characters after comments have been removed
+and spaces have been squeezed.\footnote{Sequences of two or more consecutive
+space-like characters are collapsed into a single space. The squeezing is
+obtained by running the command \texttt{tr -s [:space:] file} on Unix-like
+systems.}
+%
+The reported sizes account for the source code of the formalization excluding
+examples and any ``safety'' result, if present. We have noticed that the
+interpretation of ``safety'' varies widely among the formalizations and, except
+for our own, they all differ from the one stated in the linearity
+challenge~\citep{CarboneEtAl24} when available.
+%
+Some formalization define \emph{libraries} which could be reused in different
+contexts. In these cases, the size of the library is reported separately next to
+the size of the part of the development that uses it.
+
+In general, it is difficult to draw firm conclusions on the effectiveness of the
+various approaches because the formalizations differ substantially in terms of
+reference calculus, feature coverage, proof assistant,
+process/expression/channel representation, handling of linearity. Also, most of
+them predate the linearity challenge~\cite{CarboneEtAl24}, hence they are not
+necessarily targeted towards that particular problem. However, some of
+them~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,SanoKavanaghPientka23,ZackonEtAl25}
+are explicitly motivated by the complexity arising from the linear usage of
+resources/channels and from context splitting in particular.
+
+we can make some inference based on the available data.
 

### 2025-12-19 paper/related.tex
related

@@
 The reported sizes account for the source code of the formalization excluding
-examples and any ``safety'' result, if present. We have noticed that the
+examples and any safety result, if present. We have noticed that the
 interpretation of ``safety'' varies widely among the formalizations and, except
@@
 In general, it is difficult to draw firm conclusions on the effectiveness of the
-various approaches because the formalizations differ substantially in terms of
-reference calculus, feature coverage, proof assistant,
-process/expression/channel representation, handling of linearity. Also, most of
-them predate the linearity challenge~\cite{CarboneEtAl24}, hence they are not
-necessarily targeted towards that particular problem. However, some of
+various approaches in addressing the linearity challenge because the
+formalizations differ substantially in terms of reference calculus, feature
+coverage, proof assistant, process/expression/channel representation, handling
+of linearity. Also, most of them predate the publication of the linearity
+challenge~\cite{CarboneEtAl24}, hence they are not necessarily targeted towards
+that particular problem. However, some of
 them~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,SanoKavanaghPientka23,ZackonEtAl25}
@@
 
-we can make some inference based on the available data.
+Based on the available data, we can make the following observations.
+%
+Context splitting, which is very well represented, does not seem to be a good
+indicator of the complexity of the formalization, since it spans the whole range
+of formalization sizes. In particular, our formalization is based on context
+splitting and covers full \CP~\cite{Wadler14}, yet it is by far the smallest
+one.
 
-paper di Marco a ECOOP (multiparty)
+The largest formalizations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are also
+the ones that adopt a locally nameless representation of channels. In this
+formalizations channels are represented in two different ways, depending on
+whether they are free or bound. This necessarily entails some duplication of
+effort as well as some transformation machinery between the two representations
+that contribute to the complexity of the formalization.
+%
+Other channel representations are not strong complexity indicators. Note that
+the adoption of co-de Bruijn syntax implies the use of context splitting, hence
+the two aspects are not completely independent.
+
+There is no strong evidence that the intrinsically typed representation of
+processes reduces the size of the formalization. As observed in
+\Cref{sec:process-agda}, this choice may help reducing the overall number of
+datatypes to be defined, but the definitions are necessarily more involved. In
+the end, it seems plausible to speculate that the effort for representing
+processes, types and typing rules is not substantially different overall.
+
+Finally, it seems that the use of (explicit) continuations helps taming the
+complexity of the formalization more than anything else. Indeed, the six
+smallest formalizations (excluding the deadlock freedom results) -- with an
+average size of around 62kb -- are all based on continuations, whether they are
+explicit (\LinearPI, \SCP, \LCP) or ``virtual'' (\GV), while the remaining ones
+are 263kb on average. This observation is in agreement with the observations we
+made earlier in this paper concerning the difference between \emph{updating}
+bindings versus \emph{removing and adding} bindings in a typing context. The
+adoption of continuations enables a clean linear management of channels and
+avoids the need for strong updates on the typing context.
+
+% It is also interesting to note that
+% whereby \SCP, which the authors~\citet{SanoKavanaghPientka23} do not relate
+% directly to the linear $\pi$-calculus, relies on explicit continuations to tame
+% the complexity of resource management. Our comparison suggests that this aspect,
+% more than the structural treatment of contexts,
+
+% \citet{SanoKavanaghPientka23} do not relate it directly to the linear
+% $\pi$-calculus.
+
+% più che il trattamento strutturale dei contesti (ed evitare lo splitting) è
+% l'uso delle continuazione a fare la differenza
+
+In this work we have focused on models and theories of \emph{binary sessions},
+but there are also formalizations of \emph{multiparty session types}, notably
+those of \citet{JacobsBalzerKrebbers22a} and \citet{TiroreBengstonCarbone25},
+which are notoriously more complex than those of binary session types. Just to
+give an example, the formalization of \citet{TiroreBengstonCarbone25} amounts to
+more than 1Mb of Coq code.

### 2025-12-19 paper/main.tex
progress
agda code
types and contexts in agda
description of processes
completed structural congruence
split formalization into modules
included reduction
reduction
drafted deadlock freedom
drafted type safety
drafted intro
(re)started paper
prose for types
new notation for splitting
co de bruijn
restored description of Process
preparing examples
related work
restructuring presentation with new code
snapshot
revised congruence
restored examples
cleanup

@@
 
-\usepackage{amsmath,amssymb,amsfonts}
-\usepackage{amsthm}
-\usepackage{xcolor}
-\usepackage{textcomp}
-\usepackage{listings}
+\usepackage{amssymb}
 \usepackage{mathtools}
@@
 \usepackage{upgreek}
-%%%%
+\usepackage{cleveref}
+\usepackage{newunicodechar}
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+
 \input{macros}
+
+\newunicodechar{𝟘}{\ensuremath{\Zero}}
+\newunicodechar{𝟙}{\ensuremath{\One}}
+\newunicodechar{⊥}{\ensuremath{\Bot}}
+\newunicodechar{⊤}{\ensuremath{\Top}}
+\newunicodechar{⊕}{\ensuremath{\Plus}}
+\newunicodechar{⊗}{\ensuremath{\Ten}}
+\newunicodechar{⅋}{\ensuremath{\Par}}
+\newunicodechar{∀}{\ensuremath{\forall}}
+\newunicodechar{∩}{\ensuremath\cap}
+\newunicodechar{⋂}{\ensuremath\bigcap}
+\newunicodechar{∶}{\ensuremath{:}}
+\newunicodechar{∪}{\ensuremath\cup}
+\newunicodechar{⊢}{\ensuremath\vdash}
+\newunicodechar{∈}{\ensuremath\in}
+\newunicodechar{─}{\ensuremath{-\mkern-5mu}}
+\newunicodechar{′}{\ensuremath{'}}
+\newunicodechar{≡}{\ensuremath\equiv}
+\newunicodechar{₀}{\ensuremath{_0}}
+\newunicodechar{₁}{\ensuremath{_1}}
+\newunicodechar{₂}{\ensuremath{_2}}
+\newunicodechar{₃}{\ensuremath{_3}}
+\newunicodechar{↝}{\ensuremath\red}
+\newunicodechar{↭}{\ensuremath\leftrightsquigarrow}
+\newunicodechar{⇒}{\ensuremath\Rightarrow}
+\newunicodechar{⊸}{\ensuremath\multimap}
+\newunicodechar{≃}{\ensuremath\simeq}
+\newunicodechar{Γ}{\ensuremath\ContextC}
+\newunicodechar{Δ}{\ensuremath\ContextD}
+\newunicodechar{Θ}{\ensuremath\ContextE}
+\newunicodechar{∗}{\ensuremath\ast}
+\newunicodechar{∷}{\ensuremath{{::}}}
+\newunicodechar{∃}{\ensuremath\exists}
+\newunicodechar{λ}{\ensuremath\lambda}
+\newunicodechar{π}{\ensuremath\pi}
+\newunicodechar{σ}{\ensuremath\sigma}
+\newunicodechar{ρ}{\ensuremath\rho}
+\newunicodechar{⊒}{\ensuremath\pcong}
+\newunicodechar{⊎}{\ensuremath\uplus}
+\newunicodechar{⟦}{\ensuremath\llbracket}
+\newunicodechar{⟧}{\ensuremath\rrbracket}
+\newunicodechar{𝒞}{\ensuremath\ReductionContext}
+\newunicodechar{ℕ}{\ensuremath{\mathbb{N}}}
+\newunicodechar{𝔹}{\ensuremath{\mathbb{B}}}
+\newunicodechar{≫}{\ensuremath\gg}
+\newunicodechar{≪}{\ensuremath\ll}
+\newunicodechar{∋}{\ensuremath\ni}
+\newunicodechar{⊳}{\ensuremath\vartriangleright}
+% \newunicodechar{•}{\bullet}
 
@@
 
-%% as per the requirement new theorem styles can be included as shown below
-%\theoremstyle{theorem}%
-\newtheorem{theorem}{Theorem}%  meant for continuous numbers
-%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
-%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
-\newtheorem{proposition}[theorem]{Proposition}% 
-%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
-
-% \theoremstyle{thmstyletwo}%
-% \newtheorem{example}{Example}%
-% \newtheorem{remark}{Remark}%
-
-% \theoremstyle{thmstylethree}%
-% \newtheorem{definition}{Definition}%
-
 \raggedbottom
@@
 
-\title{A Logical Solution of the Linearity Challenge}
+\title{A Linear Solution of the Linearity Challenge}
 
@@
 
-\author[2]{\fnm{Claudia} \sur{Raffaelli}}\email{xxx@yyy.zzz}
-\equalcont{These authors contributed equally to this work.}
+\author[2]{\fnm{Claudia} \sur{Raffaelli}}\email{claudia.raffaelli@studenti.unicam.it}
 
-\affil*[1]{\orgdiv{Department of Computer Science and Engineering}, \orgname{Università di Bologna}, \orgaddress{\street{Mura Anteo Zamboni 7}, \city{Bologna}, \postcode{40138}, \state{BO}, \country{Italy}}}
+\affil[1]{\orgdiv{Department of Computer Science and Engineering}, \orgname{Università di Bologna}, \orgaddress{\street{Mura Anteo Zamboni 7}, \city{Bologna}, \postcode{40126}, \state{BO}, \country{Italy}}}
 
@@
 
-\keywords{keyword1, Keyword2, Keyword3, Keyword4}
+\keywords{session types, linear logic, deadlock freedom, Agda}
 
@@
 
-\section{Introduction}
-\label{sec:introduction}
-
+\input{introduction}
 \input{calculus}
+\input{agda}
+\input{related}
+\input{conclusion}
 

### 2025-12-21 paper/introduction.tex
cleanup
intro

--- 2025-08-03 paper/introduction.tex
@@
 %
-The benchmark aims at investigating effective formalization techniques that take
+The benchmark aims at identifying effective formalization techniques that take
 these features into account so as to foster the adoption of machine-checked
@@
 
-This paper reports on a solution to one of the challenges in the benchmark
-concerned with the formalization of a minimal \emph{session calculus}. Hereafter
-we will refer to this challenge as the \emph{linearity challenge}.
+One of the challenges in the benchmark, henceforth called \emph{linearity
+challenge}, concerns the formalization of a minimal \emph{session calculus}.
 %
@@
 around three key ideas: (1) session endpoints are \emph{linear resources} that
-cannot be discarded or duplicated without jeopardizing some safety and liveness
+cannot be discarded or duplicated without compromising some safety and liveness
 properties of a program; (2) the type of a session endpoint is \emph{updated}
@@
 interaction.
+%
+The linearity challenge is motivated by the observation that the proper handling
+of linear resources in a formalization often requires a large number of complex
+auxiliary definitions and technical results that divert from the main problem
+under consideration. One of the alleged culprits of such unwanted complexity is
+\emph{context splitting}, namely the operation that partitions a typing context
+in such a way that the linear resources described therein end up in only one of
+the partitions. This observation has led to the development of various
+techniques for dealing with typing contexts without splitting them, including
+leftover typing~\cite{ZalakainDardha21}, the use of linearity
+predicates~\cite{SanoKavanaghPientka23} and contexts with
+tags~\cite{ZackonEtAl25}.
 
+This paper presents a solution to the linearity challenge.
+%
 The calculus that we formalize, which we call \Calculus, and the one in the
-description of the challenge differ for two main aspects:
-\begin{enumerate}
-\item The challenge considers a minimal calculus of sessions, which is basically
-    a first-order $\pi$-calculus without replication, whereas \Calculus is
-    closely related to
-    \CP~\citep{CairesPfenning10,Wadler14,CairesPfenningToninho16}, a
-    fully-featured calculus of higher-order sessions with both linear and shared
-    channels and whose type system is based on linear logic.
+challenge description differ for two key aspects.
 %
-    Targeting \Calculus allows us to consider a model of sessions that is
-    well-known in the literature, that is based on firm logical grounds, and
-    that includes a broader range of features including higher-order sessions
-    and shared channels.
-\item While the calculus in the challenge is equipped with native sessions,
-    \Calculus is a \emph{linear} calculus where each channel is meant to be used
-    for a \emph{single} communication. Structured conversations pertaining the
-    same session and involving the exchange of several messages can still be
-    modeled thanks to the use of \emph{explicit continuation channels}, as in
-    the encoding of (binary) sessions into the linear
-    $\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
+First of all, the challenge only considers a minimal calculus of first-order
+sessions, whereas \LCP is a fully-featured calculus that includes support for
+higher-order, polymorphic sessions and both linear and shared channels. Also, as
+suggested by its name, \LCP is closely related to \CP~\citep{Wadler14} and its
+type system is based on classical linear logic.
 %
-    Overall, \Calculus has the same expressive power of \CP but is a lower-level
-    calculus in which sessions can be encoded instead of being featured
-    natively.
-\end{enumerate}
+The second difference is that the calculus in the challenge is equipped with
+\emph{native} sessions, whereas \LCP is a \emph{linear} calculus where each
+channel is meant to be used for a \emph{single} communication. Structured
+sessions involving the exchange of several messages can be modeled thanks to the
+use of \emph{explicit continuation channels}. This encoding of (binary) sessions
+using linear channels is well known in the
+literature~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
+%
+Overall, \LCP has the same expressive power of \CP, but is a lower-level
+calculus in which sessions can be encoded instead of being featured natively.
 
-\Luca{Da qualche parte dire quali sono le sfide della sfida: representation of
-channel names, context splitting. Parlare di co de bruijn?}
+The formalization we present is the most complete (in terms of features
+supported by the calculus) and, at the same time, the simplest one (in terms of
+code size). It is also one of only two formalizations that prove the deadlock
+freedom property, and it does so with an order of magnitude less code than the
+existing formalization~\cite{JacobsBalzerKrebbers22}.
+
+the outcome of our investigation are somehow surprising in the sense that the
+premises of the challenge and the assumption that the main technical challenge
+is the representation of contexts
+
+congetturiamo che la semplicità della formalizzazione derivi dalle
+caratteristiche del calcolo più che dalle tecniche utilizzate, che alla fine
+sono piuttosto standard.
 
@@
 from a public\Luca{Related work? Confronto con altre formalizzazioni?}
-repository.\footnote{\url{https://github.com/boystrange/LinearityChallenge/tree/main/exp}}
+repository.\footnote{\url{https://github.com/boystrange/LinearityChallenge/tree/main/cp}}

### 2025-12-21 paper/introduction.tex
intro

@@
 One of the challenges in the benchmark, henceforth called \emph{linearity
-challenge}, concerns the formalization of a minimal \emph{session calculus}.
+challenge}, concerns the formalization of a \emph{session calculus}.
 %
@@
 the partitions. This observation has led to the development of various
-techniques for dealing with typing contexts without splitting them, including
-leftover typing~\cite{ZalakainDardha21}, the use of linearity
-predicates~\cite{SanoKavanaghPientka23} and contexts with
-tags~\cite{ZackonEtAl25}.
+alternatives to context splitting including leftover
+typing~\cite{ZalakainDardha21}, the use of linearity
+predicates~\cite{SanoKavanaghPientka23} and tagged contexts~\cite{ZackonEtAl25}.
 
@@
 %
-The calculus that we formalize, which we call \Calculus, and the one in the
-challenge description differ for two key aspects.
+The calculus that we formalize, dubbed \Calculus, and the one in the challenge
+description differ for two key aspects.
 %
@@
 sessions, whereas \LCP is a fully-featured calculus that includes support for
-higher-order, polymorphic sessions and both linear and shared channels. Also, as
-suggested by its name, \LCP is closely related to \CP~\citep{Wadler14} and its
-type system is based on classical linear logic.
+higher-order, polymorphic, linear and shared channels. Also, \LCP is closely
+related to \CP~\citep{Wadler14} and its type system coincides with the proof
+system of classical linear logic.
 %
 The second difference is that the calculus in the challenge is equipped with
-\emph{native} sessions, whereas \LCP is a \emph{linear} calculus where each
-channel is meant to be used for a \emph{single} communication. Structured
-sessions involving the exchange of several messages can be modeled thanks to the
-use of \emph{explicit continuation channels}. This encoding of (binary) sessions
-using linear channels is well known in the
-literature~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
+\emph{native} sessions, whereas \LCP features shared channels, which are
+unrestricted, and linear channels, which are meant to be used for a
+\emph{single} communication. Structured sessions involving the exchange of
+several messages can be modeled using \emph{explicit continuations} according to
+the well-known encoding of (binary) sessions into linear
+channels~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
 %
 Overall, \LCP has the same expressive power of \CP, but is a lower-level
-calculus in which sessions can be encoded instead of being featured natively.
+calculus in which sessions are encodable instead of being featured natively.
 
-The formalization we present is the most complete (in terms of features
-supported by the calculus) and, at the same time, the simplest one (in terms of
-code size). It is also one of only two formalizations that prove the deadlock
-freedom property, and it does so with an order of magnitude less code than the
+Compared to the other analogous formalizations of session/linear
+calculi~\cite{GotoEtAl16,Thiemann19,CicconePadovani20,CastroFerreiraYoshida20,RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,JacobsBalzerKrebbers22,SanoKavanaghPientka23,ZackonEtAl25},
+the one we present here is both the most complete (in terms of features
+supported by the calculus) and the most streamlined (in terms of code size). It
+is also one of only two formalizations that prove the deadlock freedom property
+for a session calculus, and it does so with substantially less code than the
 existing formalization~\cite{JacobsBalzerKrebbers22}.
+
+We acknowledge that \LCP is not
+
+The differences between \LCP and the calculus are functional to the simplicity
+of its formalization.
 

### 2025-12-22 paper/introduction.tex
drafted intro

@@
 
-This paper presents a solution to the linearity challenge.
+This paper proposes a logical and linear solution of the linearity challenge. By
+``logical and linear solution'' we mean that the calculus that we formalize,
+dubbed \Calculus, and the one in the challenge description differ for two key
+aspects.
 %
-The calculus that we formalize, dubbed \Calculus, and the one in the challenge
-description differ for two key aspects.
-%
-First of all, the challenge only considers a minimal calculus of first-order
-sessions, whereas \LCP is a fully-featured calculus that includes support for
+First of all, the challenge considers an \emph{ad hoc}, minimal calculus of
+first-order sessions, whereas \LCP is a fully-featured calculus that supports
 higher-order, polymorphic, linear and shared channels. Also, \LCP is closely
@@
 The second difference is that the calculus in the challenge is equipped with
-\emph{native} sessions, whereas \LCP features shared channels, which are
-unrestricted, and linear channels, which are meant to be used for a
-\emph{single} communication. Structured sessions involving the exchange of
-several messages can be modeled using \emph{explicit continuations} according to
-the well-known encoding of (binary) sessions into linear
+\emph{native} sessions, whereas \LCP features linear channels that are meant to
+be used for a \emph{single} communication. Structured sessions involving the
+exchange of several messages can be modeled using \emph{explicit continuations}
+according to the well-known encoding of (binary) sessions into linear
 channels~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
 %
-Overall, \LCP has the same expressive power of \CP, but is a lower-level
-calculus in which sessions are encodable instead of being featured natively.
+Overall, \LCP is a lower-level calculus in which sessions are encodable instead
+of being featured natively, but it has the same expressive power of \CP.
 
-Compared to the other analogous formalizations of session/linear
-calculi~\cite{GotoEtAl16,Thiemann19,CicconePadovani20,CastroFerreiraYoshida20,RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,JacobsBalzerKrebbers22,SanoKavanaghPientka23,ZackonEtAl25},
-the one we present here is both the most complete (in terms of features
-supported by the calculus) and the most streamlined (in terms of code size). It
-is also one of only two formalizations that prove the deadlock freedom property
-for a session calculus, and it does so with substantially less code than the
-existing formalization~\cite{JacobsBalzerKrebbers22}.
-
-We acknowledge that \LCP is not
-
-The differences between \LCP and the calculus are functional to the simplicity
-of its formalization.
-
-the outcome of our investigation are somehow surprising in the sense that the
-premises of the challenge and the assumption that the main technical challenge
-is the representation of contexts
-
-congetturiamo che la semplicità della formalizzazione derivi dalle
-caratteristiche del calcolo più che dalle tecniche utilizzate, che alla fine
-sono piuttosto standard.
-
-We value these aspects of \Calculus because their combined effect allows us to
-obtain a remarkably simple formalization.
+We value these differences between \LCP and the calculus in the challenge
+description because we think that they contribute to the simplicity of the
+formalization more than the adopted formalization techniques, which end up being
+fairly standard.\Luca{Our work fits the spirit of the challenge identifying not
+only the most appropriate techniques but also the most appropriate models that
+aid the formalization of concurrent/distributed models of computation.}
 %
@@
 
-BISOGNA PARLARE DI RELATED WORK
+% Compared to the other analogous formalizations of session/linear
+% calculi~\cite{GotoEtAl16,Thiemann19,CicconePadovani20,CastroFerreiraYoshida20,RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,JacobsBalzerKrebbers22,SanoKavanaghPientka23,ZackonEtAl25},
+% the one we present here is both the most complete (in terms of features
+% supported by the calculus) and the most streamlined (in terms of code size). It
+% is also one of only two formalizations that prove the deadlock freedom property
+% for a session calculus, and it does so with substantially less code than the
+% existing formalization~\cite{JacobsBalzerKrebbers22}.
 
-The rest of the paper is structured as follows. \Cref{sec:calculus} describes
-the syntax and the operational semantics of \Calculus and states the properties
-of well-typed processes that we formalize and prove. \Cref{sec:agda} outlines
-the essential aspects of the formalization with particular emphasis on the
-representation of channel names and typing contexts. \Cref{sec:conclusion}
-summarizes our contributions, discusses some extensions and future work. The
-formalization has been carried out using Agda and its full code is available
-from a public\Luca{Related work? Confronto con altre formalizzazioni?}
-repository.\footnote{\url{https://github.com/boystrange/LinearityChallenge/tree/main/cp}}
+\subparagraph*{Structure of the paper.}
+%
+\Cref{sec:calculus} describes the syntax and the operational semantics of \LCP
+and states the properties of well-typed processes that we formalize and prove,
+namely typing preservation, deadlock freedom and runtime safety.
+%
+\Cref{sec:agda} presents a roadmap of the Agda formalization and discusses its
+key aspects with particular emphasis on the representation of channel names and
+of typing contexts.
+%
+\Cref{sec:examples} illustrates a few examples of \LCP processes as they are
+represented in Agda and shows how to use Agda as an interpreter for \LCP.
+%
+\Cref{sec:related} discusses related work more in detail by providing a
+qualitative and quantitative comparison between the known formalizations of
+session/linear calculi, including the one for \LCP.
+%
+\Cref{sec:conclusion} summarizes our contributions and discusses some ongoing
+and future work.
+
+The formalization has been carried out using Agda and the source code is
+available on Github at
+\url{https://github.com/boystrange/LinearityChallenge/tree/main/lcp}.

### 2025-12-23 paper/calculus.tex
cleanup
revised intro

@@
-\section{\Calculus}
+\section{Linear \CP}
 \label{sec:calculus}
@@
                     & | & \Fork\x\y\z{P}{Q}       & \text{fork} \\
-                    & | & \Forall\x\X\y.P         & \text{for all} \\
-                    & | & \Exists\x{A}\y.P        & \text{exists} \\
+                    & | & \Forall\x\X\z.P         & \text{for all} \\
+                    & | & \Exists\x{A}\z.P        & \text{exists} \\
                     & | & \Server\x\y.P           & \text{server} \\
@@
 %
-The processes $\Exists\x{A}\y.P$ and $\Forall\x\X\y.P$ describe the input/output
-of a type on the channel $x$ along with a fresh continuation $y$. These forms
+The processes $\Exists\x{A}\z.P$ and $\Forall\x\X\z.P$ describe the input/output
+of a type on the channel $x$ along with a fresh continuation $z$. These forms
 are useful to describe polymorphic processes.
@@
             \SJoin &
-            \Cut\x{\Join\y\u\v.P}{Q} & \pcong & \Join\y\u\v.\Cut\x{P}{Q} % & x \ne y,u,v
+            \Cut\x{\Join\y\u\z.P}{Q} & \pcong & \Join\y\u\z.\Cut\x{P}{Q} % & x \ne y,u,v
             \\
             \SForkL &
-            \Cut\x{\Fork\y\u\v{P}{Q}}{R} & \pcong & \Fork\y\u\v{\Cut\x{P}{R}}{Q} & x \in \fn{P}
+            \Cut\x{\Fork\y\u\z{P}{Q}}{R} & \pcong & \Fork\y\u\z{\Cut\x{P}{R}}{Q} & x \in \fn{P}
             \\
             \SForkR &
-            \Cut\x{\Fork\y\u\v{P}{Q}}{R} & \pcong & \Fork\y\u\v{P}{\Cut\x{Q}{R}} & x \in \fn{Q} %\setminus\set{y,u,v}
+            \Cut\x{\Fork\y\u\z{P}{Q}}{R} & \pcong & \Fork\y\u\z{P}{\Cut\x{Q}{R}} & x \in \fn{Q} %\setminus\set{y,u,v}
             \\

### 2025-12-23 paper/introduction.tex
revised intro

@@
 One of the challenges in the benchmark, henceforth called \emph{linearity
-challenge}, concerns the formalization of a \emph{session calculus}.
+challenge}, concerns the formalization of a \emph{minimal calculus of sessions}.
 %
@@
 The linearity challenge is motivated by the observation that the proper handling
-of linear resources in a formalization often requires a large number of complex
+of linear resources in a formalization often requires a large number of
 auxiliary definitions and technical results that divert from the main problem
-under consideration. One of the alleged culprits of such unwanted complexity is
-\emph{context splitting}, namely the operation that partitions a typing context
-in such a way that the linear resources described therein end up in only one of
-the partitions. This observation has led to the development of various
-alternatives to context splitting including leftover
-typing~\cite{ZalakainDardha21}, the use of linearity
-predicates~\cite{SanoKavanaghPientka23} and tagged contexts~\cite{ZackonEtAl25}.
+under study~\cite{CarboneEtAl24}. One of the alleged culprits of such complexity
+is \emph{context splitting}, namely the operation that partitions a typing
+context in such a way that the linear resources described therein end up in only
+one of the partitions. This observation has led to the exploration of various
+alternatives techniques including leftover typing~\cite{ZalakainDardha21}, the
+use of linearity predicates~\cite{SanoKavanaghPientka23} and tagged
+contexts~\cite{ZackonEtAl25}.
 
-This paper proposes a logical and linear solution of the linearity challenge. By
-``logical and linear solution'' we mean that the calculus that we formalize,
-dubbed \Calculus, and the one in the challenge description differ for two key
-aspects.
+In this paper we attack the linearity challenge from a different perspective:
+\emph{instead of describing new techniques that are effective for the
+formalization of the calculus in the challenge, we describe a new calculus that
+is easy to formalize with known techniques}.\Luca{dobbiamo dire perché lo
+spirito della challenge è rispettato, e dunque giustificare il titolo
+dell'articolo. Dire che il nostro non è un calcolo ad hoc, ma ha fondamenti
+logici importanti, è completo, ed è strettamente in relazione con altri modelli
+analoghi (CP e linear pi)}
 %
-First of all, the challenge considers an \emph{ad hoc}, minimal calculus of
-first-order sessions, whereas \LCP is a fully-featured calculus that supports
-higher-order, polymorphic, linear and shared channels. Also, \LCP is closely
-related to \CP~\citep{Wadler14} and its type system coincides with the proof
-system of classical linear logic.
+More specifically, we present a calculus called \LCP whose type system coincides
+with the proof system of \emph{classical linear logic} and that features
+\emph{linear channels} instead of sessions. Linear channels are meant to be used
+for a \emph{single} communication. Structured sessions involving the exchange of
+several messages can be modeled in \LCP using \emph{explicit continuations}
+according to the well-known encoding of (binary) sessions into the linear
+$\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
 %
-The second difference is that the calculus in the challenge is equipped with
-\emph{native} sessions, whereas \LCP features linear channels that are meant to
-be used for a \emph{single} communication. Structured sessions involving the
-exchange of several messages can be modeled using \emph{explicit continuations}
-according to the well-known encoding of (binary) sessions into linear
-channels~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
+Overall, \LCP is a low-level calculus with a strong connection with linear logic
+and in which sessions are encodable instead of being featured natively.
 %
-Overall, \LCP is a lower-level calculus in which sessions are encodable instead
-of being featured natively, but it has the same expressive power of \CP.
-
-We value these differences between \LCP and the calculus in the challenge
-description because we think that they contribute to the simplicity of the
-formalization more than the adopted formalization techniques, which end up being
-fairly standard.\Luca{Our work fits the spirit of the challenge identifying not
-only the most appropriate techniques but also the most appropriate models that
-aid the formalization of concurrent/distributed models of computation.}
+These features play an important role in taming the complexity of its
+formalization.
 %
@@
 
-% Compared to the other analogous formalizations of session/linear
-% calculi~\cite{GotoEtAl16,Thiemann19,CicconePadovani20,CastroFerreiraYoshida20,RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,JacobsBalzerKrebbers22,SanoKavanaghPientka23,ZackonEtAl25},
-% the one we present here is both the most complete (in terms of features
-% supported by the calculus) and the most streamlined (in terms of code size). It
-% is also one of only two formalizations that prove the deadlock freedom property
-% for a session calculus, and it does so with substantially less code than the
-% existing formalization~\cite{JacobsBalzerKrebbers22}.
+In the end, the formalization of \LCP is both the most complete (in terms of
+features supported by the calculus) and the most streamlined (in terms of code
+size) among the known formalizations of session/linear
+calculi~\cite{GotoEtAl16,Thiemann19,CastroFerreiraYoshida20,CicconePadovani20,RouvoetPoulsenKrebbersVisser20,SanoKavanaghPientka23,JacobsBalzerKrebbers22,ZackonEtAl25,ZalakainDardha21}.
+It is also one of only two formalizations that prove the deadlock freedom
+property for a session calculus, and it does so with substantially less code
+than the existing formalization~\cite{JacobsBalzerKrebbers22}.
 
-\subparagraph*{Structure of the paper.}
+The rest of the paper is organized as follows.
 %

### 2025-12-23 paper/introduction.tex
revised intro

@@
 one of the partitions. This observation has led to the exploration of various
-alternatives techniques including leftover typing~\cite{ZalakainDardha21}, the
+alternative techniques including leftover typing~\cite{ZalakainDardha21}, the
 use of linearity predicates~\cite{SanoKavanaghPientka23} and tagged
@@
 In this paper we attack the linearity challenge from a different perspective:
-\emph{instead of describing new techniques that are effective for the
-formalization of the calculus in the challenge, we describe a new calculus that
-is easy to formalize with known techniques}.\Luca{dobbiamo dire perché lo
-spirito della challenge è rispettato, e dunque giustificare il titolo
-dell'articolo. Dire che il nostro non è un calcolo ad hoc, ma ha fondamenti
-logici importanti, è completo, ed è strettamente in relazione con altri modelli
-analoghi (CP e linear pi)}
+\emph{instead of proposing techniques that make it easy to formalize the
+calculus in the challenge, we propose a (relatively new) calculus that is easy
+to formalize with the known techniques}.\Luca{dobbiamo dire perché lo spirito
+della challenge è rispettato, e dunque giustificare il titolo dell'articolo.
+Dire che il nostro non è un calcolo ad hoc, ma ha fondamenti logici importanti,
+è completo, ed è strettamente in relazione con altri modelli analoghi (CP e
+linear pi)}
 %
-More specifically, we present a calculus called \LCP whose type system coincides
-with the proof system of \emph{classical linear logic} and that features
-\emph{linear channels} instead of sessions. Linear channels are meant to be used
-for a \emph{single} communication. Structured sessions involving the exchange of
-several messages can be modeled in \LCP using \emph{explicit continuations}
-according to the well-known encoding of (binary) sessions into the linear
+More specifically, we describe a calculus called \LCP whose type system
+coincides with the proof system of \emph{classical linear logic} and that
+features \emph{linear channels} instead of sessions. Linear channels are meant
+to be used for a \emph{single} communication. Structured sessions involving the
+exchange of several messages can be modeled in \LCP using \emph{explicit
+continuations} according to the well-known encoding of (binary) sessions into
+the linear
 $\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
 %
-Overall, \LCP is a low-level calculus with a strong connection with linear logic
-and in which sessions are encodable instead of being featured natively.
+Overall, \LCP is a linear version of \CP~\cite{Wadler14} in which sessions are
+encodable instead of being featured natively.
 %
@@
 %
-The fact of working with a calculus based on linear logic prevents \emph{by
-construction} the same (sequential) process to own both endpoints of a session.
-This situation is clearly undesirable since it does not correspond to a useful
-pattern of interaction (every meaningful session requires its endpoints to be
-used by parallel processes) and is a potential source of deadlocks.
+Using a calculus based on linear logic prevents \emph{by construction} the same
+(sequential) process to own both endpoints of a session. This situation is
+clearly undesirable since it does not correspond to a useful pattern of
+interaction (every meaningful session requires its endpoints to be used by
+parallel processes) and is a potential source of deadlocks.
 %
@@
 names is a primary design choice, it spares us the need to \emph{distinguish}
-the two endpoints of a session, \eg by means of polarities~\citep{GayHole05} or
-by using different names connected by the same
-binder~\citep{Vasconcelos12,GayVasconcelos25}.
+the two endpoints of a session, \eg by means of
+polarities~\citep{GayHole05,ZalakainDardha21} or by using different names
+connected by the same binder~\citep{Vasconcelos12,GayVasconcelos25}.
 %
-Also, the fact of working with a calculus with \emph{linear} channels and
-explicit continuations spares us the need to \emph{update} the type of channels
-in typing contexts. When a channel is used it is effectively consumed, therefore
-its type can be \emph{removed} from the typing context and the type of the
-continuation channel is \emph{added} back to the typing context. As is turns
-out, removing and adding types is easier than updating them. Even more so
-considering that the type of continuation channels can be conveniently added
-\emph{at the beginning} of a typing context, since such channels are fresh by
-definition.
+Using a calculus with \emph{linear} channels and explicit continuations spares
+us the need to \emph{update} the type of channels in typing contexts. When a
+channel is used it is effectively consumed, therefore its type can be
+\emph{removed} from the typing context and the type of the continuation channel
+is \emph{added} back to the typing context. As is turns out, removing and adding
+types is easier than updating them. Even more so considering that the type of
+continuation channels can be conveniently added \emph{at the beginning} of a
+typing context, since such channels are fresh by definition.
 
-In the end, the formalization of \LCP is both the most complete (in terms of
+The formalization of \LCP that we obtain is both the most complete (in terms of
 features supported by the calculus) and the most streamlined (in terms of code

### 2025-12-23 paper/calculus.tex
summary of section 2

@@
 
-The calculus of pure sessions that we formalize is dubbed \Calculus and is a
-linear version of \CP~\citep{Wadler14,GayVasconcelos25} such that every session
-endpoint is meant to be used \emph{exactly once}. This differs from \CP and
-other session calculi where session endpoints can be used repeatedly, albeit
-sequentially. The key idea is that each message sent on a session endpoint is
-paired with a \emph{fresh continuation channel} on which the rest of the
-conversation takes place.
+As we have anticipated in \Cref{sec:introduction}, \LCP is a linear version of
+\CP~\citep{Wadler14,GayVasconcelos25} such that every linear channel is meant to
+be used for \emph{exactly one} communication. This differs from \CP and other
+session calculi where session endpoints can be used repeatedly, albeit
+sequentially. A session consisting of several messages exchanged in sequence can
+be encoded using linear channels by pairing each message with a \emph{fresh
+continuation channel} on which the rest of the session takes place.
 %
@@
 
-STRUCTURE OF SECTION
+In this section we present \LCP, starting from its types (\Cref{sec:types}),
+which coincide with classical linear logic propositions, its syntax
+(\Cref{sec:syntax}), the reduction semantics (\Cref{sec:semantics}), the typing
+rules (\Cref{sec:typing-rules}) and the formulation of the properties ensured by
+the type system (\Cref{sec:properties}).
 

### 2025-12-24 paper/calculus.tex
revised calculus

@@
 sequentially. A session consisting of several messages exchanged in sequence can
-be encoded using linear channels by pairing each message with a \emph{fresh
-continuation channel} on which the rest of the session takes place.
+be encoded in \LCP using linear channels by pairing each message with a
+\emph{fresh continuation channel} on which the rest of the session takes place.
 %
@@
 The quantifiers $\forall X.A$ and $\exists X.A$ describe the input/output of a
-type $X$ followed by the protocol $A$.
+type $X$ followed by the protocol $A$ and are useful to describe parametric
+protocol polymorphism.
 %
@@
 
-The syntax of \Calculus makes use of an infinite set of \emph{channel names},
-ranged over by $x$, $y$ and $z$, and is shown in~\Cref{tab:syntax}.
+The syntax of \Calculus makes use of an infinite set of \emph{channels}, ranged
+over by $x$, $y$ and $z$, and is shown in~\Cref{tab:syntax}.
 %
@@
 $\InTag_2$. This generalization has no substantial impact on the formalization,
-so we stick to the classical case of ``binary choices'' in line with the usual
-presentations of linear logic.
+so we stick to binary choices in line with the usual presentations of linear
+logic.
 %
@@
 %
-The processes $\Exists\x{A}\z.P$ and $\Forall\x\X\z.P$ describe the input/output
-of a type on the channel $x$ along with a fresh continuation $z$. These forms
-are useful to describe polymorphic processes.
+The processes $\Forall\x\X\z.P$ and $\Exists\x{A}\z.P$ describe the input/output
+of a type on the channel $x$ along with a fresh continuation $z$.
 
@@
 $\Contract\x\y\z.P$ denotes an explicit \emph{contraction}, that is a client
-that uses the server $x$ multiple times on the channels $y$ and $z$.
+that connects to the server waiting on $x$ multiple times on the channels $y$
+and $z$.
 
@@
 
-The notions of free and bound names are fairly standard, bearing in mind that
-output action prefixes bind continuation channels in (some) continuation
-processes. In particular, $\Select\x{\Tag_i}\z.P$ binds $z$ in $P$, while
-$\Fork\x\y\z{P}{Q}$ binds $y$ in $P$ but not in $Q$ and $z$ in $Q$ but not in
-$P$.
+The notions of free and bound channels are fairly standard, bearing in mind that
+output prefixes bind continuation channels in (some) continuation processes. In
+particular, $\Select\x{\Tag_i}\z.P$ binds $z$ in $P$, while $\Fork\x\y\z{P}{Q}$
+binds $y$ in $P$ but not in $Q$ and $z$ in $Q$ but not in $P$.
 %
-We write $\fn{P}$ for the set of channel names occurring free in $P$ and we
-identify processes up to renaming of bound names.
+We write $\fn{P}$ for the set of channels occurring free in $P$ and we identify
+processes up to renaming of bound channels.
+
+\begin{example}
+    \label{ex:booleans}
+    We can represent the type of a boolean value being produced as the
+    proposition $\Bool \eqdef \One \Plus \One$ and that of a boolean value being
+    consumed as its dual $\dual\Bool = \Bot \With \Bot$. Following these types,
+    the boolean constants can be modeled by the processes
+    \[
+        \TrueP(x) \eqdef \Select\x\LeftTag\z.\Close\z
+        \qquad
+        \FalseP(x) \eqdef \Select\x\RightTag\z.\Close\z
+    \]
+    and the boolean negation function by the process
+    \[
+        \NotP(x,y) \eqdef
+        \BigCase\x\u{
+            \Wait\u.\FalseP(y)
+        }{
+            \Wait\u.\TrueP(y)
+        }
+    \]
+
+    As an example, the composition $\Cut[\Bool]\x{\TrueP(x)}{\NotP(x,y)}$
+    computes the negation of true.
+    %
+    Note the use of explicit continuations in these processes and the fact that
+    each channel is used \emph{exactly once}. The same processes in \CP would be
+    written as
+    \[
+        \TrueP(x) \eqdef \Select\x\LeftTag{}.\Close\x
+        \quad
+        \FalseP(x) \eqdef \Select\x\RightTag{}.\Close\x
+        \quad
+        \NotP(x,y) \eqdef
+        \BigCase\x{}{
+            \Wait\x.\FalseP(y)
+        }{
+            \Wait\x.\TrueP(y)
+        }
+    \]
+    where each channel is used multiple times to indicate the various
+    input/output actions pertaining to the same session.
+    %
+    In general, the \CP version of an \LCP process can be obtained by reusing
+    the same channel $x$ in place of the continuation $z$ in \Cref{tab:syntax}.
+    %
+    \eoe
+\end{example}
 
@@
 %
-We content ourselves with such informal description of these critical side
-conditions given that we are going to formalize \Calculus later on.
+We content ourselves with such informal description of these side conditions
+given that we are going to formalize \Calculus later on.
 %
@@
 
-Notably and somewhat surprisingly, there is no pre-congruence rule asserting
-that parallel composition is associative. While this is a sound and expected
-property of parallel composition, it turns out to be inessential for proving all
-of the properties of \Calculus that we consider. We leave it out in light of the
-forthcoming formalization of the calculus, in which every simplification is
-welcome, although standard presentations of classical linear logic admit such
+There is no pre-congruence rule asserting that parallel composition is
+associative. While this is a sound and expected property of parallel
+composition, it turns out to be inessential for proving the properties of
+\Calculus that we consider. We leave it out in light of the forthcoming
+formalization of the calculus, in which every simplification is welcome,
+although standard presentations of classical linear logic admit such
 law~\cite{Wadler14,Doumane17}.

### 2025-12-24 paper/introduction.tex
revised calculus

@@
 \emph{instead of proposing techniques that make it easy to formalize the
-calculus in the challenge, we propose a (relatively new) calculus that is easy
-to formalize with the known techniques}.\Luca{dobbiamo dire perché lo spirito
-della challenge è rispettato, e dunque giustificare il titolo dell'articolo.
-Dire che il nostro non è un calcolo ad hoc, ma ha fondamenti logici importanti,
-è completo, ed è strettamente in relazione con altri modelli analoghi (CP e
-linear pi)}
+calculus in the challenge, we propose a calculus that is easy to formalize with
+the known techniques}.\Luca{dobbiamo dire perché lo spirito della challenge è
+rispettato, e dunque giustificare il titolo dell'articolo. Dire che il nostro
+non è un calcolo ad hoc, ma ha fondamenti logici importanti, è completo, ed è
+strettamente in relazione con altri modelli analoghi (CP e linear pi)}
 %

### 2025-12-24 paper/calculus.tex
revised calculus, added examples

@@
     \[
-        \TrueP(x) \eqdef \Select\x\LeftTag\z.\Close\z
+        \TrueP(x) \eqdef \Select\x\LeftTag{x'}.\Close{x'}
         \qquad
-        \FalseP(x) \eqdef \Select\x\RightTag\z.\Close\z
+        \FalseP(x) \eqdef \Select\x\RightTag{x'}.\Close{x'}
     \]
@@
         \NotP(x,y) \eqdef
-        \BigCase\x\u{
-            \Wait\u.\FalseP(y)
+        \BigCase\x{x'}{
+            \Wait{x'}.\FalseP(y)
         }{
-            \Wait\u.\TrueP(y)
+            \Wait{x'}.\TrueP(y)
         }
@@
     As an example, the composition $\Cut[\Bool]\x{\TrueP(x)}{\NotP(x,y)}$
-    computes the negation of true.
+    produces false on $y$.
     %
@@
 All these rules have implicit side-conditions aimed at preserving the meaning of
-names when binders are moved around: terms entering or exiting the scope of a
+channels when binders are moved around: terms entering or exiting the scope of a
 binder for $x$ must not have free occurrences of $x$. This holds also for the
@@
 
-Typing contexts (\ie sequents) are finite maps from channel names to types
-written as $x_1 : A_1, \dots, x_n : A_n$ and ranged over by $\ContextC$ and
-$\ContextD$. We write $\ContextC, \ContextD$ for the union of $\ContextC$ and
-$\ContextD$ when they have disjoint domains. We write $\WhyNot\Context$ for some
-context $\Context = x_1 : \WhyNot A_1, \dots, x_n : \WhyNot A_n$ where all the
-types in its range are prefixed by the $\WhyNot$ modality. We call these types
-\emph{unrestricted}, since they are used to denote shared channels that can be
-(explicitly) discarded and duplicated at will.
+We use typing contexts (\ie sequents) to keep track of the type of channels.
+Typing contexts are finite maps from channels to types written as $x_1 : A_1,
+\dots, x_n : A_n$ and ranged over by $\ContextC$ and $\ContextD$. We write
+$\ContextC, \ContextD$ for the union of $\ContextC$ and $\ContextD$ when they
+have disjoint domains. We write $\WhyNot\Context$ for some context $\Context =
+x_1 : \WhyNot A_1, \dots, x_n : \WhyNot A_n$ where all the types in its range
+are prefixed by the $\WhyNot$ modality. We call these types \emph{unrestricted},
+since they are used to denote shared channels that can be (explicitly) discarded
+and duplicated at will.
 
@@
 proof rules of classical linear logic, the reader may refer to the standard
-literature on linear logic~\cite{} or propositions as sessions~\citep{Wadler14}
-for their interpretation.
+literature on linear logic~\cite{Girard87} or propositions as
+sessions~\citep{Wadler14} for their interpretation.
+
+% \begin{example}
+%     \label{ex:typing-booleans}
+%     Below we show the typing derivation for the process
+%     $\Cut[\Bool]{x}{\TrueP(x)}{\NotP(x,y)}$ defined in \Cref{ex:booleans}, where
+%     we have elided some obvious sub-derivations.
+%     \[
+%         \begin{prooftree}
+%             \[
+%                 \[
+%                     \justifies
+%                     \wtp{
+%                         \Close{x'}
+%                     }{
+%                         x' : \One
+%                     }
+%                     \using\CloseRule
+%                 \]
+%                 \justifies
+%                 \wtp{
+%                     \TrueP(x)
+%                 }{
+%                     x : \Bool
+%                 }
+%                 \using\SelectRule
+%             \]
+%             \[
+%                 \[
+%                     \[
+%                         \mathstrut\smash\vdots
+%                         \justifies
+%                         \wtp{
+%                             \FalseP(y)
+%                         }{
+%                             y : \Bool
+%                         }
+%                     \]
+%                     \justifies
+%                     \wtp{
+%                         \Wait{x'}.\FalseP(y)
+%                     }{
+%                         x' : \Bot,
+%                         y : \Bool
+%                     }
+%                     \using\WaitRule
+%                 \]
+%                 \[
+%                     \mathstrut\smash\vdots
+%                     \justifies
+%                     \wtp{
+%                         \Wait{x'}.\TrueP(y)
+%                     }{
+%                         x' : \Bot,
+%                         y : \Bool
+%                     }
+%                 \]
+%                 \justifies
+%                 \wtp{
+%                     \NotP(x,y)
+%                 }{
+%                     x : \dual\Bool,
+%                     y : \Bool
+%                 }
+%                 \using\CaseRule
+%             \]
+%             \justifies
+%             \wtp{
+%                 \Cut[\Bool]{x}{\TrueP(x)}{\NotP(x,y)}
+%             }{
+%                 y : \Bool
+%             }
+%             \using\CutRule
+%         \end{prooftree}
+%     \]
+% \end{example}
+
+\begin{example}
+    \label{ex:echo}
+    In this example we define a polymorphic echo server:
+    \[
+        \EchoP(x) \eqdef
+        \Server\x\y.
+        \Forall\y\X{y'}.
+        \Join{y'}\u{y''}.
+        \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
+    \]
+
+    The server accepts requests on the shared channel $x$. Each request
+    established a new session, initially represented by the channel $y$. The
+    session consists in receiving the type $X$ of the value being echoed
+    followed by the value itself $u$. The value is then sent back unchanged and
+    the session is terminated. Note the succession of different (linear)
+    channels $y, y', y'', y'''$ that chain the session together.
+
+    We establish that $\EchoP(x)$ is well typed with the following deerivation
+    \[
+        \begin{prooftree}
+            \[
+                \[
+                    \[
+                        \[
+                            \justifies
+                            \wtp{
+                                \Link\u\v
+                            }{
+                                u : X,
+                                v : \dual\X
+                            }
+                            \using\LinkRule
+                        \]
+                        \[
+                            \justifies
+                            \wtp{
+                                \Close{y'''}
+                            }{
+                                y''' : \One
+                            }
+                            \using\CloseRule
+                        \]
+                        \justifies
+                        \wtp{
+                            \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
+                        }{
+                            u : X,
+                            y'' : \dual\X \Ten \One
+                        }
+                        \using\ForkRule
+                    \]
+                    \justifies
+                    \wtp{
+                        \Join{y'}\u{y''}.
+                        \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
+                    }{
+                        y' : X \Par (\dual\X \Ten \One)
+                    }
+                    \using\JoinRule
+                \]
+                \justifies
+                \wtp{
+                    \Forall\y\X{y'}.
+                    \Join{y'}\u{y''}.
+                    \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
+                }{
+                    y : \forall\X.X \Par (\dual\X \Ten \One)
+                }
+                \using\ForallRule
+            \]
+            \justifies
+            \wtp{
+                \EchoP(x)
+            }{
+                x : \OfCourse(\forall\X.X \Par (\dual\X \Ten \One))
+            }
+            \using\ServerRule
+        \end{prooftree}
+    \]
+    %
+    where the side condition of the rule \ForallRule is trivially satifised
+    since the typing context does not contain bindings other than the one for
+    $y$.
+    %
+    \eoe
+\end{example}
 
@@
 interact regardless of the context in which it occurs because the $\Close\x$
-sub-processes it contains are blocked on the channel $x$ that is bound by the
-cut.
+sub-processes it contains are blocked on the channel $x$ that is restricted by
+the cut.
 %
@@
 does not reduce, and yet it is not a deadlock because it would be able to make
-progress when composed in parallel with $\Close\x$. In general, even a process
-like $\Cut\y{\Wait\y.P}{\Wait\z.Q}$ where $x\ne y,z$ cannot be considered a
-deadlock, because the prefixes $\Wait\y$ and $\Wait\z$ could be exposed using
-\SWait and possibly \SComm. Let us make all this more precise.
+progress when composed in parallel with $\Close\x$. Even a process like
+$\Cut\y{\Wait\y.P}{\Wait\z.Q}$ where $x\ne y,z$ cannot be considered a deadlock,
+because the prefixes $\Wait\y$ and $\Wait\z$ could be exposed using \SWait and
+possibly \SComm. Let us make all this more precise.
 
@@
 of some sort. Note that a thread may contain cuts, but these cuts must all
-guarded underneath the topmost action prefix.
+guarded underneath the topmost action prefix of the thread.
 %
 We say that $P$ is \emph{observable} if $P \pcong Q$ for some thread $Q$. An
-observable process is a process that exposes an action on a free channel name
-and therefore can interact through that channel, if put into some appropriate
+observable process is a process that exposes an action on a free channel and
+therefore can interact through that channel, if put into some appropriate
 context. We say that $P$ is \emph{reducible} if $P \red Q$ for some $Q$. A
@@
 the hole in $\ReductionContext$ with $P$, noting that such replacement may
-capture channel names that are bound in $\ReductionContext$ and occur free in
-$P$.
+capture channel that are bound in $\ReductionContext$ and occur free in $P$.
 %

### 2025-12-24 paper/introduction.tex
revised calculus, added examples

@@
 
-In this paper we attack the linearity challenge from a different perspective:
-\emph{instead of proposing techniques that make it easy to formalize the
-calculus in the challenge, we propose a calculus that is easy to formalize with
-the known techniques}.\Luca{dobbiamo dire perché lo spirito della challenge è
-rispettato, e dunque giustificare il titolo dell'articolo. Dire che il nostro
-non è un calcolo ad hoc, ma ha fondamenti logici importanti, è completo, ed è
-strettamente in relazione con altri modelli analoghi (CP e linear pi)}
+In this paper we attack the linearity challenge from a different angle:
+\emph{instead of proposing new techniques that make it easy to formalize the
+calculus in the challenge, we propose a (relatively) new calculus that is easy
+to formalize with the existing techniques}.\Luca{dobbiamo dire perché lo spirito
+della challenge è rispettato, e dunque giustificare il titolo dell'articolo.
+Dire che il nostro non è un calcolo ad hoc, ma ha fondamenti logici importanti,
+è completo, ed è strettamente in relazione con altri modelli analoghi (CP e
+linear pi)}
 %
@@
 It is also one of only two formalizations that prove the deadlock freedom
-property for a session calculus, and it does so with substantially less code
-than the existing formalization~\cite{JacobsBalzerKrebbers22}.
+property for a session calculus~\cite{JacobsBalzerKrebbers22} and it does so
+with substantially less code.
 
@@
 The formalization has been carried out using Agda and the source code is
-available on Github at
+available on GitHub at
 \url{https://github.com/boystrange/LinearityChallenge/tree/main/lcp}.

### 2025-12-24 paper/Type.lagda.tex
revised congruence
cleanup
revised types

@@
 The representation of types is standard, with one constructor for each of the
-forms described in \Cref{sec:calculus}. We start by defining \emph{pre-types} as
-an indexed data type $\AgdaDatatype{PreType}~n$ where $n$ is the number of bound
+forms described in \Cref{sec:types}. We start by defining \emph{pre-types} as an
+indexed data type $\AgdaDatatype{PreType}~n$ where $n$ is the number of bound
 type variables and we use de Bruijn indices for denoting type variables. In this
@@
 directive.\footnote{The directive is effective provided that the option
-\texttt{--rewriting} is enabled, either globally when invoking Agda or in the
-module's preamble.}
+\texttt{{-}{-}rewriting} is enabled, either globally when invoking Agda or in
+the module's preamble.}
 
@@
 definitions are straightforward adaptations of those described by
-\cite{KokkeWadler}.
+\citet{KokkeSiekWadler20}.
 
@@
 \begin{code}
-[_/_] : ∀{n} → PreType n → Fin (suc n) → PreType n
-[ A / zero  ]  = A
-[ A / suc k ]  = var k
+[_/] : ∀{n} → PreType n → Fin (suc n) → PreType n
+[ A /] zero     = A
+[ A /] (suc k)  = var k
 \end{code}
@@
 
-It is worth discussing one case in the proof of \AgdaFunction{dual-subst}:
+It is worth looking at one case in the proof of \AgdaFunction{dual-subst},
+namely when the type is a dualized type variable:
+
 \begin{code}
@@
 We could easily prove this equivalence by invoking \AgdaFunction{dual-inv}, but
-thanks to the added rewriting rule a use of \AgdaInductiveConstructor{refl}
-suffices. In this case the saved effort is negligible, but in later results,
-where it is necessary to use \AgdaFunction{dual-inv} for rewriting part of the
-\emph{index} of some type families, having an implicit rewriting rule allows us
-to avoid writing some quite obscure Agda code.
+thanks to the rewriting rule that we have added just above a use of
+\AgdaInductiveConstructor{refl} suffices. In this case the saved effort is
+negligible, but in later results, where it is necessary to use
+\AgdaFunction{dual-inv} for rewriting part of the \emph{index} of some type
+families, having an implicit rewriting rule allows us to avoid writing some
+quite obscure Agda code.
 
@@
 formalization that follows. Therefore, we add it to the set of implicit
-rewriting rules used by Agda so that we do not have to think about it again.
+rewriting rules used by Agda so that we do not have to think about this property
+again.
 

### 2025-12-24 paper/agda.tex
included reduction
drafted deadlock freedom
drafted type safety
(re)started paper
revised types
restroed contexts and permutations
new notation for splitting
restored DF
restored type safety
restructuring presentation with new code
revised congruence
restored reduction
restored examples

@@
 \label{sec:agda}
+
+In this section we describe the formalization of \LCP in Agda. We present in
+detail only the key parts of the code, including the representation of types,
+processes, the definition of the operational semantics and the statement of the
+main results. The interested reader may refer to the public repository for the
+complete source code.
 
@@
 \input{Process}
-\input{Semantics}
+\input{Congruence}
+\input{Reduction}
+\input{DeadlockFreedom}
+\input{Safety}
+\input{Examples}

### 2025-12-24 paper/calculus.tex
revised types

@@
 
-As we have anticipated in \Cref{sec:introduction}, \LCP is a linear version of
-\CP~\citep{Wadler14,GayVasconcelos25} such that every linear channel is meant to
-be used for \emph{exactly one} communication. This differs from \CP and other
-session calculi where session endpoints can be used repeatedly, albeit
-sequentially. A session consisting of several messages exchanged in sequence can
-be encoded in \LCP using linear channels by pairing each message with a
-\emph{fresh continuation channel} on which the rest of the session takes place.
+In this section we give a cursory presentation of \LCP.
+%
+As we have anticipated in \Cref{sec:introduction}, \LCP is very closely related
+to \CP~\citep{Wadler14,GayVasconcelos25} except that \LCP features linear
+channels instead of sessions. A session consisting of several messages exchanged
+in sequence can be encoded in \LCP using linear channels by pairing each message
+with a \emph{fresh continuation channel} on which the rest of the session takes
+place.
 %
@@
 
-In this section we present \LCP, starting from its types (\Cref{sec:types}),
-which coincide with classical linear logic propositions, its syntax
-(\Cref{sec:syntax}), the reduction semantics (\Cref{sec:semantics}), the typing
-rules (\Cref{sec:typing-rules}) and the formulation of the properties ensured by
-the type system (\Cref{sec:properties}).
+We present \LCP starting from its types (\Cref{sec:types}), which coincide with
+classical linear logic propositions, then moving on to the syntax of processes
+(\Cref{sec:syntax}), their reduction semantics (\Cref{sec:semantics}), the
+typing rules (\Cref{sec:typing-rules}) and the formulation of the properties
+ensured by the type system (\Cref{sec:properties}).
 

### 2025-12-24 paper/Context.lagda.tex
revised typing contexts

@@
 
-We are going to adopt a nameless representation of channels whereby a channel is
-identified by the position of its type in a typing context. This representation
-is akin to using De Bruijn indices except that the index, instead of being
-represented explicitly by a natural number, is computable from the proof that
-the type of the channel belongs to the typing context.
-%
-Typing contexts are represented using lists of types, where the (polymorphic)
+We are going to adopt a nameless representation of channels hence typing
+contexts are represented using lists of types, where the (polymorphic)
 \AgdaDatatype{List} type and its constructors \AgdaInductiveConstructor{[]} and
@@
 \AgdaModule{Data.List} of Agda's standard library. We will keep using
-$\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts, even
-though in the Agda formalization they are lists instead of finite maps.
+$\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts as we
+have done in \Cref{sec:typing-rules}, even though in the Agda formalization they
+are lists instead of finite maps.
 
@@
 equivalently and in a more compact way as shown below. More usages of
-\AgdaFunction{≫} will be given in \Cref{sec:examples}.
+\AgdaFunction{≫} will be provided in \Cref{sec:examples}.
 
@@
 
-In the following we will make extensive use of predicates over contexts. For
-this reason, it is worth recalling a number of definitions from the
+Hereafter we will make extensive use of predicates over contexts. For this
+reason, it is worth recalling a number of definitions from the
 \AgdaModule{Relation.Unary} module in Agda's standard library in
@@
 For this reason, we invest some effort in defining a few abstractions that allow
-us to occasionally hide context splittings and enhance code readability.
+us to hide context splittings and enhance code readability.
 %
@@
 witnessing the splitting $\Splitting\ContextC\ContextD\ContextE$ along with
-$P~\ContextD$ and $Q~\ContextE$.\Luca{Nei related fare presente che queste
-definizioni seguono sostanzialmente \citet{RouvoetPoulsenKrebbersVisser20} ed
-evidenziare le differenze.}
+$P~\ContextD$ and $Q~\ContextE$.
+%
+The use of the metavariables $P$ and $Q$ for denoting predicates over contexts
+is not without reason. As we will see shortly, in our formalization processes
+are indeed predicates over typing contexts, albeit not the only ones.
 

### 2025-12-24 paper/Permutations.lagda.tex
revised permutations

@@
 because a structural pre-congruence rule is applied, or when a channel
-substitution occurs, \cf the right-hand side of the \LinkRule reduction, the
-typing context must be suitably rearranged. Such rearrangement is in fact a
-\emph{permutation} of the elements of $\Context$.
+substitution occurs, \cf the right-hand side of the \LinkRule reduction in
+\Cref{tab:semantics}, the typing context must be suitably rearranged. Such
+rearrangement is in fact a \emph{permutation} of the elements of $\Context$.
 
@@
 relate lists of arbitrary type) and the quantified arguments $A$ and $B$ of
-\AgdaInductiveConstructor{swap} and \AgdaInductiveConstructor{prep} are
-implicit, which streamlines their usage in the rest of the code.
+\AgdaInductiveConstructor{swap} and \AgdaInductiveConstructor{prep} can be made
+implicit, which streamlines the usage of these constructors in the rest of the
+code.
 

### 2025-12-24 paper/Process.lagda.tex
revised processes

@@
 
-All the constructors representing input/output actions, namely those from
-\AgdaInductiveConstructor{fail} through \AgdaInductiveConstructor{contract},
-basically follow the same pattern: they possibly quantify over some types $A$
-and $B$ and then (implicitly) over a typing context $\Context$ through the
-function \AgdaFunction{∀[\_]} applied to a predicate of the form
-$P~\AgdaFunction{⇒}~\AgdaDatatype{Proc}$. The predicate states how to build a
-process that is well typed in $\Context$, provided that $\Context$ satisfies
-$P$. In general, $P$ is a (separating) conjunction of sub-predicates
+All the constructors except \AgdaInductiveConstructor{link} and
+\AgdaInductiveConstructor{cut} basically follow the same pattern: they possibly
+quantify over some types $A$ and $B$ and then (implicitly) over a typing context
+$\Context$ through the function \AgdaFunction{∀[\_]} applied to a predicate of
+the form $P~\AgdaFunction{⇒}~\AgdaDatatype{Proc}$. The predicate states how to
+build a process that is well typed in $\Context$, provided that $\Context$
+satisfies $P$. In general, $P$ is a (separating) conjunction of sub-predicates
 corresponding to the channel on which the process is acting and to the premises
@@
 $\AgdaDatatype{Ch}~\AgdaInductiveConstructor{⊤}~\AgdaDatatype{∗}~\AgdaFunction{U}$,
-meaning that $\Context$ must contain an entry \AgdaInductiveConstructor{⊤}. That
-form of $\Context$ matches the one in the conclusion of \FailRule.
+meaning that $\Context$ must contain an entry \AgdaInductiveConstructor{⊤}
+(\AgdaFunction{U} is the universal predicate that holds for every context, \cf
+\Cref{tab:predicates}). That form of $\Context$ matches the one in the
+conclusion of \FailRule.
 
@@
 \]
-which looks intimidating at first but makes perfect sense once we recall
-\CaseRule. Remember that we are trying to establish whether $\Case\x\y{P}{Q}$ is
-well typed in $\Context$.
+which looks intimidating at first but makes perfect sense once we recall the
+typing rule \CaseRule and the definitions of \AgdaFunction{∩} and
+\AgdaFunction{⊢} in \Cref{tab:predicates}. Remember that we are trying to
+establish whether $\Case\x\y{P}{Q}$ is well typed in $\Context$.
 %
@@
 Crucially, the types $A$ and $B$ are \emph{prepended} to $\ContextD$, which is
-consistent with the fact that $y$ has been freshly introduced and $x$ has been
-consumed. Notice how easy it is to \emph{prepend} either $A$ or $B$ to
-$\ContextD$ instead of \emph{changing} the type of $x$ in $\Context$ from $A
-\With B$ to either $A$ or $B$ while preserving the position of the type, as we
-would have to do in a ``true'' session type system without explicit continuation
-channels.
+consistent with the fact that the continuation $y$ has been freshly introduced
+and the channel $x$ has been consumed. Notice how easy it is to \emph{prepend}
+either $A$ or $B$ to $\ContextD$ instead of \emph{changing} the type of $x$ in
+$\Context$ from $A \With B$ to either $A$ or $B$ while preserving the position
+of the type, as we would have to do in a ``true'' session type system without
+explicit continuation channels.
 
@@
 $\Forall\x\X\y.P$. This constructor models the continuation $P$ using
-higher-order abstract syntax (HOAS~\cite{}): witnessing that a context
-$\Context$ satisfies the predicate
+higher-order abstract syntax: witnessing that a context $\Context$ satisfies the
+predicate
 \[
@@
 property as the fact that typing is preserved by (injective) name
-substitutions.\Luca{Abbiamo detto da qualche parte dove emergono gli indici di
-de Bruijn e di fare attenzione al fatto che lo splitting li altera? Serve questa
-digressione?}\Luca{Da qualche parte mettere un remark per far notare che l'uso
+substitutions.\Luca{Da qualche parte mettere un remark per far notare che l'uso
 della separating conjunction introduce rumore, ma avere i costruttori

### 2025-12-24 paper/introduction.tex
revised processes

@@
 key aspects with particular emphasis on the representation of channel names and
-of typing contexts.
+of typing contexts. We assume that the reader is somewhat knowledgeable about
+Agda, but we also recall a few key definitions from Agda's standard library.
 %
-\Cref{sec:examples} illustrates a few examples of \LCP processes as they are
+\Cref{sec:examples} illustrates some examples of \LCP processes as they are
 represented in Agda and shows how to use Agda as an interpreter for \LCP.

### 2025-12-24 paper/Process.lagda.tex
described link and cut

@@
 
-All the constructors except \AgdaInductiveConstructor{link} and
-\AgdaInductiveConstructor{cut} basically follow the same pattern: they possibly
+The constructor \AgdaInductiveConstructor{link} models a process of the form
+$\Link\x\y$. This process is well typed in a context of the form $x : A, y :
+\dual{A}$, namely a context satisfying the predicate
+$\AgdaDatatype{Ch}~A~\AgdaDatatype{∗}~\AgdaDatatype{Ch}~(\AgdaFunction{dual}~A)$
+which we see on the left-hand side of \AgdaFunction{⇒}.
+%
+The constructor \AgdaInductiveConstructor{cut} models a cut $\Cut[A]\x{P}{Q}$.
+For this process to be well typed in a context $\Context$, it must be the case
+that $\Splitting\Context\ContextD\ContextE$ so that $P$ and $Q$ are well typed
+in the contexts $x : A, \ContextD$ and $x : \dual{A}, \ContextE$ respectively.
+Since $x$ is the most recently introduced channel, the types $A$ and $\dual{A}$
+must be added \emph{at the front} of $\ContextD$ and $\ContextE$, which we do by
+means of the functions $(A~\AgdaInductiveConstructor{∷\_})$ and
+$(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷\_})$ (these are partial
+applications of the constructor $\AgdaInductiveConstructor{\_∷\_}$ for lists to
+which we have supplied the left operand).
+
+All the remaining constructors basically follow the same pattern: they possibly
 quantify over some types $A$ and $B$ and then (implicitly) over a typing context

### 2025-12-24 paper/Congruence.lagda.tex
revised congruence

@@
 \AgdaFunction{dual-inv} described in \Cref{sec:type-agda}. Indeed $P$ and $Q$
-have type $\AgdaDatatype{Process}~(A~\AgdaInductiveConstructor{∷}~\Context_1)$
-and
-$\AgdaDatatype{Process}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\Context_2)$,
+have type $\AgdaDatatype{Proc}~(A~\AgdaInductiveConstructor{∷}~\Context_1)$ and
+$\AgdaDatatype{Proc}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\Context_2)$,
 respectively, but the \AgdaInductiveConstructor{cut} on the r.h.s. of
@@
 \AgdaFunction{dual-inv}, Agda considers these types equivalent without requiring
-intricate substitutions in the index of a data type.
+intricate substitutions in the index of \AgdaDatatype{Proc}.
 
@@
 proof $(\AgdaInductiveConstructor{>}~q)$ used in the
-\AgdaInductiveConstructor{wait} process before structural precongruence is
+\AgdaInductiveConstructor{wait} process before structural pre-congruence is
 applied. To understand the meaning of this proof, we must recall three key
@@
     %
-    We enforce the side condition $x \ne y$ of \SWait imposing that the type $A$
-    in front of $A~\AgdaInductiveConstructor{∷}~\Context_1$ necessarily comes
-    from the right component of the splitting $[\Bot] +
-    A~\AgdaInductiveConstructor{∷}~\Delta$ through the use of
-    \AgdaInductiveConstructor{>}.
+    We enforce the side condition $x \ne y$ of \SWait (which we left implicit in
+    \Cref{tab:semantics}) imposing that the type $A$ in front of
+    $A~\AgdaInductiveConstructor{∷}~\Context_1$ necessarily comes from the right
+    component of the splitting $[\Bot] + A~\AgdaInductiveConstructor{∷}~\Delta$
+    through the use of \AgdaInductiveConstructor{>}.
 \end{enumerate}
@@
 associativity property of context splitting which we called
-\AgdaFunction{+-assoc-l} in \Cref{sec:context-agda}.\Luca{Controllare
-left/right.} The nested \AgdaKeyword{let}-\AgdaKeyword{in} allows us to pattern
-match the result of the application $\AgdaFunction{+-assoc-l}~p~q$ and to
-extract the new proofs $p'$ and $q'$ for the rearranged splittings.
+\AgdaFunction{+-assoc-l} in \Cref{sec:context-agda}. The nested
+\AgdaKeyword{let}-\AgdaKeyword{in} allows us to pattern match on the result of
+the application $\AgdaFunction{+-assoc-l}~p~q$ and to extract the new proofs
+$p'$ and $q'$ for the rearranged splittings.
 
@@
 \AgdaInductiveConstructor{s-select-r} model \SSelect when the selected tag is
-respectively $\InTag_1$ and $\InTag_2$.\Luca{Controllare nomi dei tag con
-sintassi del calcolo. Chiamarli inj come in Agda così c'è consistenza?}
+respectively $\InTag_1$ and $\InTag_2$.
 %
@@
 $(\AgdaInductiveConstructor{select}~(\AgdaInductiveConstructor{ch ⟨
->}~q~\AgdaInductiveConstructor{⟩})~\AgdaInductiveConstructor{inj₁}~P)$, which is
-found under a cut for $x : A$, is using some channel $y : B \Plus C$ to send the
-tag $\InTag_1$.
+>}~q~\AgdaInductiveConstructor{⟩})~\AgdaInductiveConstructor{inj₁}~P)$, that is
+$\Select\y\LeftTag\z.P$, is found under a cut for $x : A$ and is using some
+channel $y : B \Plus C$ to send the tag $\InTag_1$.
 %
@@
 $B~\AgdaInductiveConstructor{∷}~A~\AgdaInductiveConstructor{∷}~\ContextD$, where
-the type of the continuation sits on top of the type of the restricted channel.
+the type of the continuation $z$ comes before the type of the restricted channel
+$x$.
 %
 After structural precongruence is applied, however, the type $B$ of the
-continuation channel $z$ ends up underneath that of the restricted channel $x$,
+continuation channel $z$ ends up behind that of the restricted channel $x$,
 because now the two channels are introduced in the opposite order. Therefore, we
@@
 renaming is achieved applying the function \AgdaFunction{↭proc} to the
-\AgdaInductiveConstructor{swap} permutation, which swaps the two topmost types
-in a typing context, and finally to the process $P$.
+\AgdaInductiveConstructor{swap} permutation and to the process $P$.
 

### 2025-12-24 paper/Congruence.lagda.tex
excusatio non petita accusatio manifesta

@@
 \end{AgdaAlign}
+
+We concede that the definition of these pre-congruence rules (including those
+not discussed here) can be difficult to decipher. In part, this is due to the
+fact that splitting proofs are manifest and no longer hidden within separating
+conjunctions as we have done in \Cref{sec:process-agda} because we need them to
+enforce the side conditions of the rules in \Cref{tab:semantics}. Also, using an
+intrinsically-typed representation of processes, pre-congruence and reduction
+rules incorporate the proof that they preserve typing.

### 2025-12-24 paper/Reduction.lagda.tex
revised reduction

@@
 $\Context$. The exact location of $y$ in $\ContextC$ is found in the splitting
-$p$. We ``rename'' $x$ into $y$ within $P$ using the permutation
+$p$, thus we ``rename'' $x$ into $y$ within $P$ using the permutation
 $\AgdaFunction{↭concat}~p$.

### 2025-12-24 paper/DeadlockFreedom.lagda.tex
revised deadlock freedom

@@
 \emph{thread}, that is any process other than a cut. As it turns out, it is
-convenient to provide an more fine-grained classification of threads,
+convenient to provide a more fine-grained classification of threads,
 distinguishing between links, input/output actions, and sometimes also whether
@@
 These predicates are relatively uninteresting since they are essentially
-isomorphic to the relevant parts of the \AgdaDatatype{Proc} datatype. The only
-relevant aspect that is worth pointing out here is that in \AgdaDatatype{Input},
+isomorphic to the relevant fragments of the \AgdaDatatype{Proc} datatype. The
+only aspect that is worth pointing out here is that in \AgdaDatatype{Input},
 \AgdaDatatype{Output} and \AgdaDatatype{Server} the channel being acted upon by
@@
 those threads that, in the context of a cut, operate on the channel bound by the
-cut or on a free channel. As an illustration, consider the
-\AgdaInductiveConstructor{wait} case in \AgdaDatatype{Input} and in
+cut or on a free channel. To clarify this aspect, consider the
+\AgdaInductiveConstructor{wait} constructor in \AgdaDatatype{Input} and in
 \AgdaDatatype{Delayed}. In the former predicate we have
@@
 have to specify the type $C$ in front of $\Context$ or else Agda is unable to
-resolve some metavariables.
+automatically resolve some metavariables.
 
@@
 
-We can now define observability, reducibility and aliveness in the expected way:
+We define observability, reducibility and aliveness in the expected way:
 
@@
 a thread, then we conclude that the cut is alive first rewriting it into a
-canonical cut (\AgdaFunction{canonical-cut}) and then using
+canonical cut with \AgdaFunction{canonical-cut} and then using
 \AgdaFunction{canonical-cut-alive}.

### 2025-12-24 paper/Safety.lagda.tex
revised safety

@@
 single hole. In our intrinsically-typed formalization, reduction contexts are
-typed and indexed by two typing contexts $\ContextD$ and $\ContextC$. The latter
-is the typing context under which the reduction context as a whole is well
-typed. The former is the typing context of the hole, which is invariant for a
-given reduction context and therefore specified as a parameter of the data type:
+parameterized by the typing context $\ContextD$ of the hole, which is invariant,
+and indexed by the typing context $\ContextC$ of the whole reduction context:
 
@@
 \end{code}
+
+Note the presence of two constructors \AgdaInductiveConstructor{cut-l} and
+\AgdaInductiveConstructor{cut-r} which distinguish on whether the hole is found
+in the left (respectively, right) sub-context, as per the grammar of reduction
+contexts in \Cref{sec:properties}.
 
@@
 The proof of type safety ends up being a trivial application of
-\AgdaFunction{deadlock-freedom}.
+\AgdaFunction{deadlock-freedom}, since the process $Q$ in the hole of a
+reduction context is well typed by construction.
 

### 2025-12-24 paper/Examples.lagda.tex
revised examples

@@
 
-In this section we illustrate a few simple examples of processes encoded in our
-formalization. We start by encoding the type of a boolean value being produced
-as a choice between ``going left'' (which we interpret as true) and ``going
-right'' (which we interpret as false).
+In this section we revisit and expand the sample processes discussed in
+\Cref{ex:booleans,ex:echo} and show their encoding in our formalization. The
+encoding of $\Bool$ is straightforward.
 
@@
 
-The boolean constants are processes that perform one of these choices.
+The boolean constants are encoded thus:
+\begin{code}
+True : Proc [ 𝔹 ]
+True = select (ch ⟨ < ≫ ⟩ inj₁ (close ch))
+
+False : Proc [ 𝔹 ]
+False = select (ch ⟨ < ≫ ⟩ inj₂ (close ch))
+\end{code}
+
+We can take advantage of the host language for programming higher-order
+processes. For example, we can define a conditional process thus:
 
 \begin{code}
-true : Proc [ 𝔹 ]
-true = select (ch ⟨ < ≫ ⟩ inj₁ (close ch))
-
-false : Proc [ 𝔹 ]
-false = select (ch ⟨ < ≫ ⟩ inj₂ (close ch))
-\end{code}
-
-We can use the host language for programming higher-order processes. For
-example, we can define a conditional process thus:
-
-\begin{code}
-if_else : ∀{Γ} → Proc Γ → Proc Γ → Proc (dual 𝔹 ∷ Γ)
-if P else Q = curry∗ case ch (< ≫) ( wait (ch ⟨ < ≫ ⟩ P)
+If_Else : ∀[ Proc ⇒ Proc ⇒ (dual 𝔹 ∷_) ⊢ Proc ]
+If P Else Q = curry∗ case ch (< ≫) ( wait (ch ⟨ < ≫ ⟩ P)
                                    , wait (ch ⟨ < ≫ ⟩ Q))
@@
 
-A term $\AgdaFunction{if}~P~\AgdaFunction{else}~Q$ is a process that waits for a
+A term $\AgdaFunction{If}~P~\AgdaFunction{Else}~Q$ is a process that waits for a
 boolean value (\cf the $\AgdaFunction{dual}~\mathbb{B}$ type at the front of its
@@
 
-Next we define a process $\AgdaFunction{drop}~P$ that consumes a boolean and
+Next we define a process $\AgdaFunction{Drop}~P$ that consumes a boolean and
 continues as $P$ regardless of its value.
@@
 \begin{code}
-drop : ∀{Γ} → Proc Γ → Proc (dual 𝔹 ∷ Γ)
-drop P = if P else P
+Drop : ∀[ Proc ⇒ (dual 𝔹 ∷_) ⊢ Proc ]
+Drop P = If P Else P
 \end{code}
@@
 !!_ : Proc [ 𝔹 ] → Proc [ 𝔹 ]
-!!_ B = curry∗ cut B ≫ (if false else true)
+!!_ B = curry∗ cut B ≫ (If False Else True)
 
@@
             curry∗ cut B ≫ $
-            if (curry∗ link ch (< ≫) ch) else (drop false)
+            If (curry∗ link ch (< ≫) ch) Else (Drop False)
 A || B   = !! ((!! A) && (!! B))
@@
 {-# TERMINATING #-}
-eval : ∀{Γ} → Proc Γ → Proc Γ
+eval : ∀[ Proc ⇒ Proc ]
 eval P with deadlock-freedom P
@@
 Now if we ask Agda to normalize the goal
-$\AgdaFunction{eval}~(\AgdaFunction{false}~\AgdaFunction{||}~\AgdaFunction{false})$
+$\AgdaFunction{eval}~(\AgdaFunction{False}~\AgdaFunction{||}~\AgdaFunction{False})$
 we obtain
 $\AgdaInductiveConstructor{select}~(\AgdaInductiveConstructor{ch~⟨~<~•~⟩~inj₂}~(\AgdaInductiveConstructor{close~ch}))$,
-that is the definition of \AgdaFunction{false}, as expected.
+that is the definition of \AgdaFunction{False}, as expected.
 
-For the example that follows, it is convenient to introduce the linear
-implication connective according to the standard definition $A \multimap B
-\eqdef \dual{A} \Par B$.
+For the encoding of the echo server (\Cref{ex:echo}), it is convenient to
+introduce the linear implication connective according to the standard definition
+$A \multimap B \eqdef \dual{A} \Par B$.
 
@@
 
-Below we define a persistent, polymorphic echo \AgdaInductiveConstructor{server}
-that waits for a value and sends it back unchanged. The server is polymorphic in
-the sense that the type of the value is arbitrary and denoted by the type
-variable $X$. The notation $\AgdaFunction{\#}~n$ is defined in Agda's standard
-library and creates an element of \AgdaDatatype{Fin} from the natural number
-$n$.
+Recall that the server is polymorphic in the sense that the type of the value it
+receives and sends back is arbitrary and denoted by a type variable $X$. The
+notation $\AgdaFunction{\#}~n$ is defined in Agda's standard library and creates
+an element of \AgdaDatatype{Fin} from the natural number $n$.
 
 \begin{code}
-echo : let X = var (# 0) in
+Echo : let X = var (# 0) in
        Proc [ `! (`∀ (X ⊸ X)) ]
-echo = curry∗ server ch (< ≫)
+Echo = curry∗ server ch (< ≫)
              ( un-[]
@@
 \begin{code}
-echo-true : Proc [ 𝔹 ]
-echo-true = curry∗ cut echo ≫ $
+Echo-True : Proc [ 𝔹 ]
+Echo-True = curry∗ cut Echo ≫ $
             curry∗ client ch (< ≫) $
             curry∗ ex ch (< ≫) $
-            curry∗ fork ch (< ≫) $ true ⟨ ≫ ⟩ curry∗ link ch (< ≫) ch
+            curry∗ fork ch (< ≫) $ True ⟨ ≫ ⟩ curry∗ link ch (< ≫) ch
 \end{code}
 
-If we ask Agda to evaluate this process we obtain \AgdaFunction{true}, as
+If we ask Agda to evaluate this process we obtain \AgdaFunction{True}, as
 expected.
@@
 From these examples it should be clear that writing \Calculus processes in Agda
-is not entirely straightforward, but it is still doable with a little effort.
-The lack of names (for channels) clearly obfuscates the code, but the context
-splittings that must be provided are usually small and fairly easy to
-understand. Most importantly, the use of \AgdaFunction{≫} helps truncating
-context splittings as soon as the seeked channel has been reached.
+is not entirely straightforward and requires a little effort. The lack of names
+(for channels) clearly obfuscates the code, but the context splittings that must
+be provided are usually small and fairly easy to understand. Most importantly,
+the use of \AgdaFunction{≫} helps truncating context splittings as soon as the
+seeked channel has been reached.
 %

### 2025-12-24 paper/Examples.lagda.tex
sync examples with code

@@
 
-For the encoding of the echo server (\Cref{ex:echo}), it is convenient to
-introduce the linear implication connective according to the standard definition
-$A \multimap B \eqdef \dual{A} \Par B$.
-
-\begin{code}
-_⊸_ : ∀{n} → PreType n → PreType n → PreType n
-A ⊸ B = dual A ⅋ B
-\end{code}
-
-Recall that the server is polymorphic in the sense that the type of the value it
-receives and sends back is arbitrary and denoted by a type variable $X$. The
-notation $\AgdaFunction{\#}~n$ is defined in Agda's standard library and creates
-an element of \AgdaDatatype{Fin} from the natural number $n$.
+For the encoding of the echo server (\Cref{ex:echo}), recall that the server is
+polymorphic in the sense that the type of the value it receives and sends back
+is arbitrary and denoted by a type variable $X$. The notation
+$\AgdaFunction{\#}~n$ is defined in Agda's standard library and creates an
+element of \AgdaDatatype{Fin} from the natural number $n$.
 
@@
 Echo : let X = var (# 0) in
-       Proc [ `! (`∀ (X ⊸ X)) ]
+       Proc [ `! (`∀ (X ⅋ (dual X ⊗ 𝟙))) ]
 Echo = curry∗ server ch (< ≫)
@@
                curry∗ join ch (< ≫) $
-               curry∗ link ch (< ≫) ch)
+               curry∗ (curry∗ fork ch (< ≫)) (curry∗ link ch (< > •) ch) (< ≫) $
+               close ch )
 \end{code}
@@
             curry∗ client ch (< ≫) $
-            curry∗ ex ch (< ≫) $
-            curry∗ fork ch (< ≫) $ True ⟨ ≫ ⟩ curry∗ link ch (< ≫) ch
+            curry∗ (ex {_} {dual 𝔹}) ch (< ≫) $
+            curry∗ (curry∗ fork ch (< ≫)) True ≫ $
+            curry∗ join ch (< ≫) $
+            curry∗ wait ch (< ≫) $
+            curry∗ link ch (< > •) ch
 \end{code}

### 2025-12-24 paper/conclusion.tex
related
revised related work

@@
 \label{sec:conclusion}
+
+In this work we have focused on models and theories of \emph{binary sessions},
+but there are also formalizations of \emph{multiparty session types}, notably
+those of \citet{JacobsBalzerKrebbers22a} and \citet{TiroreBengstonCarbone25},
+which are notoriously more complex than those of binary session types. Just to
+give an example, the formalization of \citet{TiroreBengstonCarbone25} amounts to
+more than 1Mb of Coq code.
+
+FARE CONFRONTO TRA LCP E IL CALCOLO NELLA CHALLENGE PER DIRE CHE NON SI RIESCONO
+A MODELLARE TUTTI I PROCESSI, MAGARI SI PUÒ FARE UN ESEMPIO NELLA SEZIONE SUL
+CALCOLO.
+
+PARLARE DI CP E DI PI CALCOLO LINEARE
+
+BELLIN E SCOTT?
 
@@
 subtyping, for the linear fragment of \Calculus
+
+conclusioni: comparison in \Cref{sec:related} made under the assumption that the
+size of the formalization is a good measure of its complexity. But there are
+several factors that may invalidate this assumption. We should pursue

### 2025-12-24 paper/related.tex
drafted intro
revised calculus, added examples
revised related work

@@
 literature and for which we were able to retrieve the source code. We have
-looked at those formalizations that include the following characteristics:
+looked at those formalizations that include the following features:
 \begin{itemize}
-\item syntax and reduction/evaluation semantics for a calculus with support for
-    binary sessions, either natively or through their encoding in terms of
-    linear channels~\cite{KobayashiPierceTurner99,Kobayashi02b};
+\item reduction/evaluation semantics for a calculus with support for binary
+    sessions, either natively or through their encoding in terms of linear
+    channels~\cite{KobayashiPierceTurner99,Kobayashi02b};
 \item a type system along with the usual typing preservation results;
-\item a proof of global progress/deadlock freedom, for those type systems that
-    enforce it.
+\item optionally, a proof of global progress/deadlock freedom.
 \end{itemize}
@@
 functional language extended with session communication primitives, along with
-an interpreter.
+an interpreter. \citet{CicconePadovani20} have taken inspiration from Thiemann's
+work to formalize a variant of the linear
+$\pi$-calculus~\cite{KobayashiPierceTurner99} that supports dependent types, so
+as to enable the description of communication protocols whose structure depends
+on the content of messages.
+
+\citet{CastroFerreiraYoshida20} describe EMTST, a library for the formalization
+of session type systems which includes as case studies the session calculus of
+\citet{HondaVasconcelosKubo98} (called ``original system'') and a revised
+version of it that is more amenable to be formalized using a locally nameless
+representation of channels.
 
@@
 computing models supported by linear type systems. One of the presented case
-studies is the formalization of a fragment of \GV~\citep{Wadler14}.
+studies is the formalization of a fragment of \GV~\citep{Wadler14}. The
+separating conjunction we have introduced in \Cref{sec:context-agda} and the
+typing of the constructors for the representation of processes in
+\Cref{sec:process-agda} have been heavily inspired by this work.
 
@@
 
-\citet{CastroFerreiraYoshida20} describe EMTST, a library for the formalization
-of session type systems which includes as case studies the session calculus of
-\citet{HondaVasconcelosKubo98} (called ``original system'') and a revised
-version of it that is more amenable to be formalized using a locally nameless
-representation of channels.
-
-\citet{CicconePadovani20} formalize a variant of the linear
-$\pi$-calculus~\cite{KobayashiPierceTurner99} that supports dependent types, so
-as to enable the description of communication protocols whose structure depends
-on the content of exchanged messages.
-
-All the formalizations mentioned make use of context splitting. In contrast,
-\citet{ZalakainDardha21} formalize a generalization of the linear $\pi$-calculus
-which is parametric in a \emph{usage algebra} (to account for channel
-sharing/linearity) and that is based on \emph{leftover typing}~\cite{}. The
-basic idea is that a process is typed with respect to an \emph{input context},
+All the formalizations mentioned so far make use of context splitting. In
+contrast, \citet{ZalakainDardha21} formalize a generalization of the linear
+$\pi$-calculus which is parametric in a \emph{usage algebra} (to account for
+channel sharing/linearity) and that is based on \emph{leftover
+typing}~\cite{Allais17}. 
+%
+With leftover typing a process is typed with respect to an \emph{input context},
 which describes the overall available channels, and an \emph{output context},
-which describes the remaining channels not consumed by the process. In this way,
-it is possible to ``concatenate'' typing judgments instead of splitting them
-beforehand.
+which describes the residual channels not consumed by the process. In this way,
+it is possible to ``concatenate'' typing judgments instead of splitting typing
+contexts beforehand. As \citet{ZalakainDardha21} nicely summarize, ``context
+splits contain usage information that is already present'' in processes, and
+therefore can be omitted. This is true, but only provided that channels are
+given a \emph{name} in processes (\citet{ZalakainDardha21} use de Bruijn indices
+to this aim). In fact, the co-de Bruijn representation of
+processes~\cite{McBride18}, whereby channels are nameless and context splitting
+is performed eagerly, can be seen as the ``dual approach'' of leftover typing:
+channel names provide information that is already present in their (singleton)
+typing context, hence they can be omitted from contexts and processes.
 
@@
 
-Motivated by the technical difficulties arising from the splitting of typing
-contexts, \citet{SanoKavanaghPientka23} propose an alternative approach based on
-\emph{linearity predicates}. The key idea is to treat typing contexts
-structurally and to enforce the linear usage of channels by checking the
-syntactical occurrence of channels within processes.
+Motivated by the technical difficulties arising from context splits,
+\citet{SanoKavanaghPientka23} model a structural version of \CP using an
+alternative approach based on \emph{linearity predicates}. The key idea is to
+treat typing contexts structurally and to enforce the linear usage of sessions
+by checking the syntactical occurrence of channels within processes.
+Interestingly, the approach relies on the \emph{explicit naming of
+continuations} so as to precisely account for the number of times a channel is
+actually used.
 
@@
         \SessionPI &
-        splitting &
+        splits &
         loc. nameless &
@@
         \GV &
-        splitting &
+        splits &
         co-de Bruijn &
@@
         \LinearPI & 
-        splitting &
+        splits &
         co-de Bruijn &
@@
         \GV &
-        splitting &
+        splits &
         named &
@@
         \Calculus &
-        splitting &
+        splits &
         co-de Bruijn &
@@
 
-\Cref{tab:comparison} summarizes a comparison of our formalization with the
+\Cref{tab:comparison} summarizes the comparison of our formalization with the
 aforementioned ones. The three leftmost colums list the reference papers (sorted
@@
 differs from (typically, is a strict subset of) the one identified by the
-acronym, but it retains some of its key features. We use \SessionPI to denote
-(variants of) the session-based $\pi$-calculus presented by
+acronym, but it retains at least some of its key features. We use \SessionPI to
+denote (variants of) the session-based $\pi$-calculus presented by
 \citet{HondaVasconcelosKubo98} and \LinearPI to denote (variants of) the linear
 $\pi$-calculus~\cite{KobayashiPierceTurner99}, into which binary sessions can be
-encoded using explicit continuation channels as described by
-\citet{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}. Finally, we use \SCP
-for the structural version of \CP introduced by \citet{SanoKavanaghPientka23}
-and \Calculus for our calculus, a linear version of \CP with explicit
-continuations.
+encoded using explicit continuation channels as described by \citet[extended
+version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}. Finally, we use
+\SCP for the structural version of \CP introduced by
+\citet{SanoKavanaghPientka23} and \Calculus for our calculus, a linear version
+of \CP with explicit continuations.
 
@@
 be encoded) are marked with {\YES} and those based on a calculus with native
-sessions are marked with {\NO}, as expected. The calculus \SCP is marked with
-{\MAYBE} because, while not being strictly based on the linear $\pi$-calculus,
-it makes use of explicit continuations for defining the linearity predicates
-that check the linear usage of channels. Finally, all the approaches based on
-\GV are also marked with \MAYBE. In principle, \emph{none} of these calculi
-makes use of continuations. However, \GV is designed in such a way that each
-session operation acting on a channel $s$ returns the result of the operation
-(if any) \emph{along with the same channel $s$}. In this way, the type of $s$
-can be ``updated'' to take into account the effect of the operation. As observed
-by \citet{Padovani17}, this way of modelling the communication primitives is
+sessions are marked with {\NO}. The calculus \SCP is marked with {\MAYBE}
+because, while not being strictly based on the linear $\pi$-calculus, it makes
+use of explicit continuations for defining the linearity predicates that check
+the linear usage of channels. Finally, all the approaches based on \GV are also
+marked with \MAYBE. In principle, \emph{none} of these calculi makes use of
+continuations. However, \GV is designed in such a way that each session
+operation acting on a channel $s$ returns the result of the operation (if any)
+\emph{along with the same channel $s$}. In this way, the type of $s$ can be
+``updated'' to take into account the effect of the operation. As observed by
+\citet{Padovani17}, this way of modeling the communication primitives is
 virtually indistinguishable from one making use of explicit continuation
@@
 kilobytes) of the formalizations as rough (and possibly questionable) estimates
-of their complexity. Instead of counting line numbers, which may depend on code
-indentation styles and syntactical constraints of the proof assistant being
+of their complexity. Instead of counting line numbers, which may be affected by
+code indentation styles and syntactical constraints of the proof assistant being
 used, we have counted the number of characters after comments have been removed
@@
 various approaches in addressing the linearity challenge because the
-formalizations differ substantially in terms of reference calculus, feature
-coverage, proof assistant, process/expression/channel representation, handling
-of linearity. Also, most of them predate the publication of the linearity
-challenge~\cite{CarboneEtAl24}, hence they are not necessarily targeted towards
-that particular problem. However, some of
+formalizations differ widely for a variety of entangled factors including the
+reference calculus, supported features, proof assistant,
+process/expression/channel representation, handling of linearity. Also, most
+formalizations predate the publication of the linearity
+challenge~\cite{CarboneEtAl24}, hence they are not necessarily targeted to
+solving that particular problem. However, some of
 them~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,SanoKavanaghPientka23,ZackonEtAl25}
@@
 %
-Context splitting, which is very well represented, does not seem to be a good
-indicator of the complexity of the formalization, since it spans the whole range
-of formalization sizes. In particular, our formalization is based on context
-splitting and covers full \CP~\cite{Wadler14}, yet it is by far the smallest
-one.
+The use of context splitting, which is very well represented, does not seem to
+be a good indicator of the complexity of the formalization, since it spans the
+whole range of sizes. Our formalization is based on context splitting and it is
+the only one that covers \CP in full, yet it is by far the smallest one.
 
@@
 \Cref{sec:process-agda}, this choice may help reducing the overall number of
-datatypes to be defined, but the definitions are necessarily more involved. In
-the end, it seems plausible to speculate that the effort for representing
-processes, types and typing rules is not substantially different overall.
+datatypes to be defined, but the definitions are also more involved because they
+incorporate invariants and the proofs of typing preservation. In the end, it
+seems plausible to speculate that the effort for representing processes, types
+and typing rules is not substantially different overall.
 
@@
 explicit (\LinearPI, \SCP, \LCP) or ``virtual'' (\GV), while the remaining ones
-are 263kb on average. This observation is in agreement with the observations we
-made earlier in this paper concerning the difference between \emph{updating}
+are 263kb on average. This observation is in agreement with the considerations
+we made earlier in this paper concerning the difference between \emph{updating}
 bindings versus \emph{removing and adding} bindings in a typing context. The
@@
 
-% It is also interesting to note that
-% whereby \SCP, which the authors~\citet{SanoKavanaghPientka23} do not relate
-% directly to the linear $\pi$-calculus, relies on explicit continuations to tame
-% the complexity of resource management. Our comparison suggests that this aspect,
-% more than the structural treatment of contexts,
+% It is also interesting to note that whereby \SCP, which the
+% authors~\citet{SanoKavanaghPientka23} do not relate directly to the linear
+% $\pi$-calculus, relies on explicit continuations to tame the complexity of
+% resource management. Our comparison suggests that this aspect, more than the
+% structural treatment of contexts,
 
@@
 % l'uso delle continuazione a fare la differenza
-
-In this work we have focused on models and theories of \emph{binary sessions},
-but there are also formalizations of \emph{multiparty session types}, notably
-those of \citet{JacobsBalzerKrebbers22a} and \citet{TiroreBengstonCarbone25},
-which are notoriously more complex than those of binary session types. Just to
-give an example, the formalization of \citet{TiroreBengstonCarbone25} amounts to
-more than 1Mb of Coq code.

### 2025-12-24 paper/calculus.tex
removed citation to linear logic

@@
 proof rules of classical linear logic, the reader may refer to the standard
-literature on linear logic~\cite{Girard87} or propositions as
-sessions~\citep{Wadler14} for their interpretation.
+literature on linear logic or propositions as sessions~\citep{Wadler14} for
+their interpretation.
 

### 2025-12-24 paper/abstract.tex
revised abstract and conclusion

\abstract{%
    We describe an Agda formalization of a process calculus based on classical
    linear logic that supports the modeling of binary sessions. The
    formalization turns out to be remarkably compact despite the use of
    ``plain'' context splitting for the handling of linear resources. We argue
    that the logical nature of the calculus and the use of explicit
    continuations contribute to the simplicity of its formalization.
    %
}

### 2025-12-24 paper/conclusion.tex
revised abstract and conclusion

--- 2025-08-01 paper/conclusion.tex
@@
 
-context splitting hard to read but overall doable
+We have presented a formalization of \LCP, a variant of the linear
+$\pi$-calculus~\cite{KobayashiPierceTurner99} supported by the same type system
+of \CP~\cite{Wadler14}. Binary sessions can be modeled in \LCP using the
+continuation-passing encoding described by \citet[extended
+version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}.
+%
+Under the assumption that the size of a formalization is a good measure of its
+complexity, the formalization of \LCP can be qualified as ``simple'' compared to
+the existing ones.
+%
+In this respect, our work casts some doubts on the actual contribution of
+context splitting to the complexity of a formalization and hints instead at
+explicit continuations in the underlying model as a simplifying factor.
+%
+It is quite interesting to note that, among the approaches that have been
+proposed to overcome context splitting, the one of \citet{SanoKavanaghPientka23}
+makes use of explicit continuations.
 
-structural pre-congruence by far most involved, and it will grow even further in
-order to prove more complex properties (eg confluence)
+We remark that \LCP and the calculus described in the linearity
+challenge~\cite{CarboneEtAl24} are fundamentally incomparable. On the one hand,
+the challenge only considers a minimal calculus of first-order, monomorphic
+sessions while \LCP is a fully-fledged calculus that supports linear, shared,
+higher-order, polymorphic channels. On the other hand, the logical foundations
+of \LCP prevent the modeling of deadlocks and of sequential processes that own
+both endpoints of a session, both of which can be modeled in the calculus of the
+challenge.
+%
+Because of its logical foundations, we like to think of \LCP as a ``canonical''
+calculus that deserves its own space in the context of the linearity challenge
+alongside with (but not in substitution of) more traditional session calculi.
 
-linear pi-calculus is good
+The simple formalization of \LCP is a good basis for further investigations. We
+have already extended \LCP with support for infinite types and recursive
+processes (the formalization of this extension is publicly available in a
+sibling folder of the one for \LCP). In the future, it would be interesting to
+formalize the strong normalization property of \LCP as a consequence of cut
+elimination of classical linear logic.
+
+In this work we have focused on models and theories of \emph{binary sessions}
+(those connecting exactly two processes), but there are also formalizations of
+\emph{multiparty sessions}, notably those of \citet{JacobsBalzerKrebbers22a} and
+\citet{TiroreBengstonCarbone25}, which are significantly more complex than those
+of binary sessions. Just to give an example, the formalization by
+\citet{TiroreBengstonCarbone25} amounts to more than 1Mb of Coq code.
+Incidentally, multiparty sessions are not supported by a strong connection with
+linear logic and there is no known continuation-passing encoding of multiparty
+sessions. Hence, it is unclear whether the considerations made for this work
+also apply to the multiparty case.

### 2025-12-24 paper/introduction.tex
examples in subsection
removed side note

@@
 calculus in the challenge, we propose a (relatively) new calculus that is easy
-to formalize with the existing techniques}.\Luca{dobbiamo dire perché lo spirito
-della challenge è rispettato, e dunque giustificare il titolo dell'articolo.
-Dire che il nostro non è un calcolo ad hoc, ma ha fondamenti logici importanti,
-è completo, ed è strettamente in relazione con altri modelli analoghi (CP e
-linear pi)}
+to formalize with the existing techniques}.
 %
@@
 %
-\Cref{sec:examples} illustrates some examples of \LCP processes as they are
-represented in Agda and shows how to use Agda as an interpreter for \LCP.
+% \Cref{sec:examples} illustrates some examples of \LCP processes as they are
+% represented in Agda and shows how to use Agda as an interpreter for \LCP.
 %

### 2025-12-24 paper/abstract.tex
revised intro, british spelling

@@
 \abstract{%
-    We describe an Agda formalization of a process calculus based on classical
+    We describe an Agda formalisation of a process calculus based on classical
     linear logic that supports the modeling of binary sessions. The
-    formalization turns out to be remarkably compact despite the use of
+    formalisation turns out to be remarkably compact despite the use of
     ``plain'' context splitting for the handling of linear resources. We argue
     that the logical nature of the calculus and the use of explicit
-    continuations contribute to the simplicity of its formalization.
+    continuations contribute to the simplicity of its formalisation.
     %

### 2025-12-24 paper/introduction.tex
revised intro, british spelling

@@
 %
-The benchmark aims at identifying effective formalization techniques that take
+The benchmark aims at identifying effective formalisation techniques that take
 these features into account so as to foster the adoption of machine-checked
@@
 One of the challenges in the benchmark, henceforth called \emph{linearity
-challenge}, concerns the formalization of a \emph{minimal calculus of sessions}.
+challenge}, concerns the formalisation of a \emph{minimal calculus of sessions}.
 %
 Sessions and session types~\citep{Honda93,HondaVasconcelosKubo98,HuttelEtAl16}
-are established formalisms for the static analysis of distributed programs based
-on peer-to-peer communication channels. Every session type system revolves
+are established abstractions for the static analysis of distributed programs
+based on peer-to-peer communication channels. Every session type system revolves
 around three key ideas: (1) session endpoints are \emph{linear resources} that
@@
 The linearity challenge is motivated by the observation that the proper handling
-of linear resources in a formalization often requires a large number of
+of linear resources in a formalisation often requires a large number of
 auxiliary definitions and technical results that divert from the main problem
@@
 In this paper we attack the linearity challenge from a different angle:
-\emph{instead of proposing new techniques that make it easy to formalize the
+\emph{instead of proposing new techniques that make it easy to formalise the
 calculus in the challenge, we propose a (relatively) new calculus that is easy
-to formalize with the existing techniques}.
+to formalise with the existing techniques}.
 %
 More specifically, we describe a calculus called \LCP whose type system
-coincides with the proof system of \emph{classical linear logic} and that
-features \emph{linear channels} instead of sessions. Linear channels are meant
-to be used for a \emph{single} communication. Structured sessions involving the
-exchange of several messages can be modeled in \LCP using \emph{explicit
-continuations} according to the well-known encoding of (binary) sessions into
-the linear
+coincides with the proof system of classical linear logic and that features
+\emph{linear channels} instead of sessions. Linear channels are meant to be used
+for a \emph{single} communication. Structured communications involving the
+exchange of several messages within the same session can be modeled in \LCP
+using \emph{explicit continuations} according to the well-known encoding of
+(binary) sessions into the linear
 $\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
@@
 encodable instead of being featured natively.
-%
-These features play an important role in taming the complexity of its
-formalization.
+
+The logical foundations of \LCP and its use of explicit continuations play an
+important role in taming the complexity of its formalisation.
 %
@@
 %
-From the standpoint of the formalization, where the representation of channel
-names is a primary design choice, it spares us the need to \emph{distinguish}
-the two endpoints of a session, \eg by means of
+From the standpoint of the formalisation, where the representation of channel
+names is a primary design choice, it spares us the need to distinguish the two
+endpoints of a session, \eg by means of
 polarities~\citep{GayHole05,ZalakainDardha21} or by using different names
@@
 %
-Using a calculus with \emph{linear} channels and explicit continuations spares
-us the need to \emph{update} the type of channels in typing contexts. When a
-channel is used it is effectively consumed, therefore its type can be
+Using a calculus with linear channels and explicit continuations spares us the
+need to \emph{update} the type of channels in typing contexts. Once a channel
+has been used it is effectively consumed, therefore its type can be
 \emph{removed} from the typing context and the type of the continuation channel
@@
 
-The formalization of \LCP that we obtain is both the most complete (in terms of
+The formalisation of \LCP that we obtain is both the most complete (in terms of
 features supported by the calculus) and the most streamlined (in terms of code
-size) among the known formalizations of session/linear
+size) among the known formalisations of session/linear
 calculi~\cite{GotoEtAl16,Thiemann19,CastroFerreiraYoshida20,CicconePadovani20,RouvoetPoulsenKrebbersVisser20,SanoKavanaghPientka23,JacobsBalzerKrebbers22,ZackonEtAl25,ZalakainDardha21}.
-It is also one of only two formalizations that prove the deadlock freedom
+It is also one of only two formalisations that prove the deadlock freedom
 property for a session calculus~\cite{JacobsBalzerKrebbers22} and it does so
@@
 
-The rest of the paper is organized as follows.
+The rest of the paper is organised as follows.
 %
 \Cref{sec:calculus} describes the syntax and the operational semantics of \LCP
-and states the properties of well-typed processes that we formalize and prove,
+and states the properties of well-typed processes that we formalise and prove,
 namely typing preservation, deadlock freedom and runtime safety.
 %
-\Cref{sec:agda} presents a roadmap of the Agda formalization and discusses its
+\Cref{sec:agda} presents a roadmap of the Agda formalisation and discusses its
 key aspects with particular emphasis on the representation of channel names and
@@
 \Cref{sec:related} discusses related work more in detail by providing a
-qualitative and quantitative comparison between the known formalizations of
+qualitative and quantitative comparison between the known formalisations of
 session/linear calculi, including the one for \LCP.
 %
-\Cref{sec:conclusion} summarizes our contributions and discusses some ongoing
+\Cref{sec:conclusion} summarises our contributions and discusses some ongoing
 and future work.
 
-The formalization has been carried out using Agda and the source code is
+The formalisation has been carried out using Agda and the source code is
 available on GitHub at

### 2025-12-24 paper/introduction.tex
revised intro

@@
 %
-\Cref{sec:agda} presents a roadmap of the Agda formalisation and discusses its
-key aspects with particular emphasis on the representation of channel names and
-of typing contexts. We assume that the reader is somewhat knowledgeable about
-Agda, but we also recall a few key definitions from Agda's standard library.
+\Cref{sec:agda} presents the Agda formalisation and illustrates its key aspects
+with particular emphasis on the representation of channels and of typing
+contexts. We assume that the reader is somewhat familiar about Agda, but we also
+recall a few key definitions from Agda's standard library.
 %
@@
 qualitative and quantitative comparison between the known formalisations of
-session/linear calculi, including the one for \LCP.
+session/linear calculi.
 %
-\Cref{sec:conclusion} summarises our contributions and discusses some ongoing
-and future work.
+\Cref{sec:conclusion} summarises our contributions and discusses ongoing and
+future work.
 

### 2025-12-24 paper/abstract.tex
revised abstract

@@
 \abstract{%
-    We describe an Agda formalisation of a process calculus based on classical
+    We describe the Agda formalisation of a process calculus based on classical
     linear logic that supports the modeling of binary sessions. The
     formalisation turns out to be remarkably compact despite the use of
-    ``plain'' context splitting for the handling of linear resources. We argue
-    that the logical nature of the calculus and the use of explicit
-    continuations contribute to the simplicity of its formalisation.
+    ``plain'' context splitting for the proper management of linear channels. We
+    argue that the logical nature of the calculus and the use of explicit
+    continuations are contributing factors to the simplicity of its
+    formalisation.
     %

### 2025-12-24 paper/introduction.tex
removed "names"

@@
 %
-From the standpoint of the formalisation, where the representation of channel
-names is a primary design choice, it spares us the need to distinguish the two
+From the standpoint of the formalisation, where the representation of channels
+is a primary design choice, it spares us the need to distinguish the two
 endpoints of a session, \eg by means of

### 2025-12-24 paper/introduction.tex
github

@@
 with particular emphasis on the representation of channels and of typing
-contexts. We assume that the reader is somewhat familiar about Agda, but we also
-recall a few key definitions from Agda's standard library.
+contexts. We assume that the reader is somewhat familiar about Agda but we
+recall the lesser known definitions from Agda's standard library.
 %
@@
 
-The formalisation has been carried out using Agda and the source code is
-available on GitHub at
-\url{https://github.com/boystrange/LinearityChallenge/tree/main/lcp}.
+The formalisation has been checked with Agda 2.8.0 and the source code is on
+GitHub at \url{https://github.com/boystrange/LinearityChallenge/tree/main/lcp}.

### 2025-12-24 paper/calculus.tex
british spelling

@@
 in sequence can be encoded in \LCP using linear channels by pairing each message
-with a \emph{fresh continuation channel} on which the rest of the session takes
-place.
+with a fresh continuation channel on which the rest of the session takes place.
 %
-This way of encoding sessions using explicit continuations has been first
-described by \citet[extended version]{Kobayashi02b} and later studied by
-\citet{DardhaGiachinoSangiorgi17}.
+This way of encoding sessions has been first described by \citet[extended
+version]{Kobayashi02b} and later studied by \citet{DardhaGiachinoSangiorgi17}.
 
@@
 discussed in the literature~\cite{LindleyMorris15}, this form is also useful for
-modeling \emph{delegation}, that is the exchange of an existing session endpoint
-in another session.
+modeling delegation, that is the exchange of an existing session endpoint in
+another session.
 %
@@
 (domain-specific) sets of tags instead of the canonical $\InTag_1$ and
-$\InTag_2$. This generalization has no substantial impact on the formalization,
+$\InTag_2$. This generalisation has no substantial impact on the formalisation,
 so we stick to binary choices in line with the usual presentations of linear
@@
 \begin{table}[t]
-    \caption{Operational semantics of \Calculus.}
+    \caption{Operational semantics of \Calculus. Some side conditions for the
+    \rulename{s-*} rules are omitted (see text).}
     \label{tab:semantics}
@@
 We content ourselves with such informal description of these side conditions
-given that we are going to formalize \Calculus later on.
+given that we are going to formalise \Calculus later on.
 %
@@
 \Calculus that we consider. We leave it out in light of the forthcoming
-formalization of the calculus, in which every simplification is welcome,
+formalisation of the calculus, in which every simplification is welcome,
 although standard presentations of classical linear logic admit such
@@
 
-The \emph{linearity challenge}~\citep{CarboneEtAl24} aims at formalizing two
+The \emph{linearity challenge}~\citep{CarboneEtAl24} aims at formalising two
 essential properties of well-typed processes: (1) typing is preserved by

### 2025-12-24 paper/Congruence.lagda.tex
british spelling

@@
 
-We formalize structural precongruence as a binary relation between processes
+We formalise structural precongruence as a binary relation between processes
 that are well typed in the \emph{same} typing context. This entails that
@@
 ``\emph{pre-congruence}'', implying that it is reflexive, transitive relation
-and that it is is preserved by some forms of calculus. In the formalization we
+and that it is is preserved by some forms of calculus. In the formalisation we
 have to be precise and we introduce specific rules:

### 2025-12-24 paper/Context.lagda.tex
british spelling

@@
 $\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts as we
-have done in \Cref{sec:typing-rules}, even though in the Agda formalization they
+have done in \Cref{sec:typing-rules}, even though in the Agda formalisation they
 are lists instead of finite maps.
@@
 Context splitting is going to be pervasive in the subsequent development, to the
-point that it contributes substantially to the complexity of the formalization.
+point that it contributes substantially to the complexity of the formalisation.
 For this reason, we invest some effort in defining a few abstractions that allow
@@
 The use of the metavariables $P$ and $Q$ for denoting predicates over contexts
-is not without reason. As we will see shortly, in our formalization processes
+is not without reason. As we will see shortly, in our formalisation processes
 are indeed predicates over typing contexts, albeit not the only ones.

### 2025-12-24 paper/DeadlockFreedom.lagda.tex
british spelling

@@
 (\Cref{thm:df}) rests on some notions and predicates about processes which we
-now formalize in Agda. First of all we need to define the notion of
+now formalise in Agda. First of all we need to define the notion of
 \emph{thread}, that is any process other than a cut. As it turns out, it is

### 2025-12-24 paper/Examples.lagda.tex
examples in subsection
british spelling

@@
-\section{Examples}
+\subsection{Examples}
 \label{sec:examples}
@@
 In this section we revisit and expand the sample processes discussed in
-\Cref{ex:booleans,ex:echo} and show their encoding in our formalization. The
+\Cref{ex:booleans,ex:echo} and show their encoding in our formalisation. The
 encoding of $\Bool$ is straightforward.
@@
 
-Now if we ask Agda to normalize the goal
+Now if we ask Agda to normalise the goal
 $\AgdaFunction{eval}~(\AgdaFunction{False}~\AgdaFunction{||}~\AgdaFunction{False})$
@@
 %
-Either way, the formalization is meant to study the meta-theory of \Calculus
+Either way, the formalisation is meant to study the meta-theory of \Calculus
 rather than implementing a user-friendly DSL for it.

### 2025-12-24 paper/Process.lagda.tex
british spelling

@@
 in the definition of the datatypes for representing processes and their
-operational semantics, but pays off in the rest of the formalization for at
+operational semantics, but pays off in the rest of the formalisation for at
 least three reasons:
@@
 \item we conflate processes and typing rules in the same datatype, thus reducing
-  the overall number of datatypes defined in the formalization;
+  the overall number of datatypes defined in the formalisation;
 \item the typing preservation results are embedded in the very definition of the
@@
 the channels that actually occur within the term. For this reason, typing
-contexts are split as soon as possible, according to the structure of processes,
-to make sure that channels are appropriately (and above all linearly)
-distributed among sub-processes, so that each channel occurs (and is used)
-exactly once.\footnote{Strictly speaking, the typing context of a
+contexts are split eagerly, according to the structure of processes, to make
+sure that channels are appropriately (and above all linearly) distributed among
+sub-processes, so that each channel occurs (and is used) exactly
+once.\footnote{Strictly speaking, the typing context of a
 \AgdaInductiveConstructor{fail} process may contain channels that do not occur

### 2025-12-24 paper/Reduction.lagda.tex
british spelling

@@
 
-Just like structural precongruence, reduction is formalized as a binary relation
+Just like structural precongruence, reduction is formalised as a binary relation
 between processes that are well typed in the very same typing context. Thus, the

### 2025-12-24 paper/Safety.lagda.tex
british spelling

@@
 We have seen that type safety is a simple instance of deadlock freedom, which is
-made even simpler to formalize in our development where processes are
+made even simpler to formalise in our development where processes are
 intrinsically typed. We start by defining reduction contexts as processes with a
-single hole. In our intrinsically-typed formalization, reduction contexts are
-parameterized by the typing context $\ContextD$ of the hole, which is invariant,
+single hole. In our intrinsically-typed formalisation, reduction contexts are
+parameterised by the typing context $\ContextD$ of the hole, which is invariant,
 and indexed by the typing context $\ContextC$ of the whole reduction context:

### 2025-12-24 paper/Type.lagda.tex
british spelling

@@
 
-This property is so important in the rest of the formalization that it is worth
+This property is so important in the rest of the formalisation that it is worth
 defining an implicit rewriting rule so that Agda can autonomously apply it
@@
 It is worth looking at one case in the proof of \AgdaFunction{dual-subst},
-namely when the type is a dualized type variable:
+namely when the type is a dualised type variable:
 
@@
 Just like \AgdaFunction{dual-inv}, \AgdaFunction{dual-subst} too is key in the
-formalization that follows. Therefore, we add it to the set of implicit
+formalisation that follows. Therefore, we add it to the set of implicit
 rewriting rules used by Agda so that we do not have to think about this property

### 2025-12-24 paper/conclusion.tex
british spelling

@@
 
-We have presented a formalization of \LCP, a variant of the linear
+We have presented a formalisation of \LCP, a variant of the linear
 $\pi$-calculus~\cite{KobayashiPierceTurner99} supported by the same type system
@@
 %
-Under the assumption that the size of a formalization is a good measure of its
-complexity, the formalization of \LCP can be qualified as ``simple'' compared to
+Under the assumption that the size of a formalisation is a good measure of its
+complexity, the formalisation of \LCP can be qualified as ``simple'' compared to
 the existing ones.
@@
 In this respect, our work casts some doubts on the actual contribution of
-context splitting to the complexity of a formalization and hints instead at
+context splitting to the complexity of a formalisation and hints instead at
 explicit continuations in the underlying model as a simplifying factor.
@@
 
-The simple formalization of \LCP is a good basis for further investigations. We
+The simple formalisation of \LCP is a good basis for further investigations. We
 have already extended \LCP with support for infinite types and recursive
-processes (the formalization of this extension is publicly available in a
+processes (the formalisation of this extension is publicly available in a
 sibling folder of the one for \LCP). In the future, it would be interesting to
-formalize the strong normalization property of \LCP as a consequence of cut
+formalise the strong normalisation property of \LCP as a consequence of cut
 elimination of classical linear logic.
@@
 In this work we have focused on models and theories of \emph{binary sessions}
-(those connecting exactly two processes), but there are also formalizations of
+(those connecting exactly two processes), but there are also formalisations of
 \emph{multiparty sessions}, notably those of \citet{JacobsBalzerKrebbers22a} and
 \citet{TiroreBengstonCarbone25}, which are significantly more complex than those
-of binary sessions. Just to give an example, the formalization by
+of binary sessions. Just to give an example, the formalisation by
 \citet{TiroreBengstonCarbone25} amounts to more than 1Mb of Coq code.

### 2025-12-24 paper/related.tex
british spelling

@@
 
-We have compared our formalization with others that have appeared in the
+We have compared our formalisation with others that have appeared in the
 literature and for which we were able to retrieve the source code. We have
-looked at those formalizations that include the following features:
+looked at those formalisations that include the following features:
 \begin{itemize}
@@
 
-\citet{GotoEtAl16} describe the formalization of a session-based variant of the
+\citet{GotoEtAl16} describe the formalisation of a session-based variant of the
 $\pi$-calculus which supports channel polymorphism. This is the oldest
-formalization of a session-based calculus for which we were able to retrieve the
+formalisation of a session-based calculus for which we were able to retrieve the
 source code (the work of \citet{Gay01} predates this one, but its source code is
@@
 
-\citet{Thiemann19} formalizes a subset of \GV~\citep{GayVasconcelos10}, a
+\citet{Thiemann19} formalises a subset of \GV~\citep{GayVasconcelos10}, a
 functional language extended with session communication primitives, along with
 an interpreter. \citet{CicconePadovani20} have taken inspiration from Thiemann's
-work to formalize a variant of the linear
+work to formalise a variant of the linear
 $\pi$-calculus~\cite{KobayashiPierceTurner99} that supports dependent types, so
@@
 
-\citet{CastroFerreiraYoshida20} describe EMTST, a library for the formalization
+\citet{CastroFerreiraYoshida20} describe EMTST, a library for the formalisation
 of session type systems which includes as case studies the session calculus of
 \citet{HondaVasconcelosKubo98} (called ``original system'') and a revised
-version of it that is more amenable to be formalized using a locally nameless
+version of it that is more amenable to be formalised using a locally nameless
 representation of channels.
@@
 \citet{RouvoetPoulsenKrebbersVisser20} present a library of abstractions
-inspired to separation logic aiding the formalization of interpreters for
+inspired to separation logic aiding the formalisation of interpreters for
 computing models supported by linear type systems. One of the presented case
-studies is the formalization of a fragment of \GV~\citep{Wadler14}. The
+studies is the formalisation of a fragment of \GV~\citep{Wadler14}. The
 separating conjunction we have introduced in \Cref{sec:context-agda} and the
@@
 % partially taken inspiration from for the representation of processes
-% (\cref{sec:context-agda,sec:process-agda}). Their formalization adopts the same
+% (\cref{sec:context-agda,sec:process-agda}). Their formalisation adopts the same
 % intrinsically-typed representation of processes with nameless channels as our
-% own, but their formalization focuses on the implementation of an interpreter (no
+% own, but their formalisation focuses on the implementation of an interpreter (no
 % small-step semantics is defined) and subject reduction is
 
-All the formalizations mentioned so far make use of context splitting. In
-contrast, \citet{ZalakainDardha21} formalize a generalization of the linear
+All the formalisations mentioned so far make use of context splitting. In
+contrast, \citet{ZalakainDardha21} formalise a generalisation of the linear
 $\pi$-calculus which is parametric in a \emph{usage algebra} (to account for
@@
 it is possible to ``concatenate'' typing judgments instead of splitting typing
-contexts beforehand. As \citet{ZalakainDardha21} nicely summarize, ``context
+contexts beforehand. As \citet{ZalakainDardha21} nicely summarise, ``context
 splits contain usage information that is already present'' in processes, and
@@
 Another work based on context splitting is the one by
-\citet{JacobsBalzerKrebbers22}, who formalize a library of \emph{connectivity
+\citet{JacobsBalzerKrebbers22}, who formalise a library of \emph{connectivity
 graphs} for reasoning and enforcing deadlock freedom in a variant of
-\GV~\citep{GayVasconcelos10,Wadler14}. This is the first formalization of
+\GV~\citep{GayVasconcelos10,Wadler14}. This is the first formalisation of
 deadlock freedom for a calculus of sessions.
@@
     \newcommand{\textlabel}[1]{\rotatebox{90}{#1}}
-    \caption{\label{tab:comparison} A summary of different formalizations of binary session calculi. All sizes are in kb.}
+    \caption{\label{tab:comparison} A summary of different formalisations of binary session calculi. All sizes are in kb.}
     \begin{minipage}{.9\textwidth}
     \begin{tabular}{@{}r@{~}@{~}c@{~~}c@{~}|@{~}c@{~~}c@{~~}c@{~~}c@{~}|@{~}r@{~~}r@{~~}r@{~~}r@{~~}r@{~}l@{}}
-        \textbf{Formalization} &
+        \textbf{Formalisation} &
         \textlabel{Proof assistant} &
@@
 
-\Cref{tab:comparison} summarizes the comparison of our formalization with the
+\Cref{tab:comparison} summarises the comparison of our formalisation with the
 aforementioned ones. The three leftmost colums list the reference papers (sorted
-by publication date), the proof assistant and the formalized calculus.
+by publication date), the proof assistant and the formalised calculus.
 %
@@
 \GV~\cite{GayVasconcelos10,Wadler14} are well known in the literature on session
-types. It should be noted that the actual calculus being formalized usually
+types. It should be noted that the actual calculus being formalised usually
 differs from (typically, is a strict subset of) the one identified by the
@@
 The four middle columns of \Cref{tab:comparison} report the relevant qualitative
-aspects of the formalizations, namely the technique used to enforce the linear
+aspects of the formalisations, namely the technique used to enforce the linear
 usage of channels, the technique used to encode channels, whether the
-formalization is intrinsically or extrinsically typed, and whether the calculus
+formalisation is intrinsically or extrinsically typed, and whether the calculus
 makes use of continuation channels. Concerning the use of continuations, the
@@
 The rightmost columns of \Cref{tab:comparison} report the \emph{size} (in
-kilobytes) of the formalizations as rough (and possibly questionable) estimates
+kilobytes) of the formalisations as rough (and possibly questionable) estimates
 of their complexity. Instead of counting line numbers, which may be affected by
@@
 %
-The reported sizes account for the source code of the formalization excluding
+The reported sizes account for the source code of the formalisation excluding
 examples and any safety result, if present. We have noticed that the
-interpretation of ``safety'' varies widely among the formalizations and, except
+interpretation of ``safety'' varies widely among the formalisations and, except
 for our own, they all differ from the one stated in the linearity
@@
 %
-Some formalization define \emph{libraries} which could be reused in different
+Some formalisation define \emph{libraries} which could be reused in different
 contexts. In these cases, the size of the library is reported separately next to
@@
 various approaches in addressing the linearity challenge because the
-formalizations differ widely for a variety of entangled factors including the
+formalisations differ widely for a variety of entangled factors including the
 reference calculus, supported features, proof assistant,
 process/expression/channel representation, handling of linearity. Also, most
-formalizations predate the publication of the linearity
+formalisations predate the publication of the linearity
 challenge~\cite{CarboneEtAl24}, hence they are not necessarily targeted to
@@
 The use of context splitting, which is very well represented, does not seem to
-be a good indicator of the complexity of the formalization, since it spans the
-whole range of sizes. Our formalization is based on context splitting and it is
+be a good indicator of the complexity of the formalisation, since it spans the
+whole range of sizes. Our formalisation is based on context splitting and it is
 the only one that covers \CP in full, yet it is by far the smallest one.
 
-The largest formalizations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are also
+The largest formalisations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are also
 the ones that adopt a locally nameless representation of channels. In this
-formalizations channels are represented in two different ways, depending on
+formalisations channels are represented in two different ways, depending on
 whether they are free or bound. This necessarily entails some duplication of
 effort as well as some transformation machinery between the two representations
-that contribute to the complexity of the formalization.
+that contribute to the complexity of the formalisation.
 %
@@
 There is no strong evidence that the intrinsically typed representation of
-processes reduces the size of the formalization. As observed in
+processes reduces the size of the formalisation. As observed in
 \Cref{sec:process-agda}, this choice may help reducing the overall number of
@@
 Finally, it seems that the use of (explicit) continuations helps taming the
-complexity of the formalization more than anything else. Indeed, the six
-smallest formalizations (excluding the deadlock freedom results) -- with an
+complexity of the formalisation more than anything else. Indeed, the six
+smallest formalisations (excluding the deadlock freedom results) -- with an
 average size of around 62kb -- are all based on continuations, whether they are

### 2025-12-25 paper/abstract.tex
revised abstract and intro

@@
     We describe the Agda formalisation of a process calculus based on classical
-    linear logic that supports the modeling of binary sessions. The
-    formalisation turns out to be remarkably compact despite the use of
-    ``plain'' context splitting for the proper management of linear channels. We
-    argue that the logical nature of the calculus and the use of explicit
-    continuations are contributing factors to the simplicity of its
-    formalisation.
+    linear logic that supports the modeling of binary sessions through their
+    encoding with explicit continuation channels. The formalisation turns out to
+    be remarkably compact despite the use of context splitting for the proper
+    management of linear channels. We argue that the logical nature of the
+    calculus and the use of explicit continuations are contributing factors to
+    the simplicity of its formalisation.
     %

### 2025-12-25 paper/introduction.tex
tuning
revised abstract and intro

@@
 %
-The linearity challenge is motivated by the observation that the proper handling
-of linear resources in a formalisation often requires a large number of
-auxiliary definitions and technical results that divert from the main problem
+The linearity challenge is motivated by the observation that the proper
+management of linear resources in a formalisation often requires a large number
+of auxiliary definitions and technical results that divert from the main problem
 under study~\cite{CarboneEtAl24}. One of the alleged culprits of such complexity
@@
 
-In this paper we attack the linearity challenge from a different angle:
+In this paper we approach the linearity challenge from a different angle:
 \emph{instead of proposing new techniques that make it easy to formalise the
@@
 for a \emph{single} communication. Structured communications involving the
-exchange of several messages within the same session can be modeled in \LCP
-using \emph{explicit continuations} according to the well-known encoding of
-(binary) sessions into the linear
+exchange of several messages within the same session is modeled in \LCP using
+\emph{explicit continuations} according to the well-known encoding of (binary)
+sessions into the linear
 $\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
@@
 Overall, \LCP is a linear version of \CP~\cite{Wadler14} in which sessions are
-encodable instead of being featured natively.
+encoded instead of being featured natively.
 
-The logical foundations of \LCP and its use of explicit continuations play an
-important role in taming the complexity of its formalisation.
+The logical foundations of \LCP and the use of explicit continuations play an
+important role in taming the complexity of the formalisation.
 %
@@
 types is easier than updating them. Even more so considering that the type of
-continuation channels can be conveniently added \emph{at the beginning} of a
-typing context, since such channels are fresh by definition.
+continuation channels must be added \emph{at the beginning} of a typing context,
+since such channels are fresh by definition.
 
@@
 %
-% \Cref{sec:examples} illustrates some examples of \LCP processes as they are
-% represented in Agda and shows how to use Agda as an interpreter for \LCP.
-%
 \Cref{sec:related} discusses related work more in detail by providing a
@@
 
-The formalisation has been checked with Agda 2.8.0 and the source code is on
+The formalisation has been checked with Agda 2.8.0 and the code is available on
 GitHub at \url{https://github.com/boystrange/LinearityChallenge/tree/main/lcp}.

### 2025-12-25 paper/calculus.tex
revision

@@
 %
-As we have anticipated in \Cref{sec:introduction}, \LCP is very closely related
-to \CP~\citep{Wadler14,GayVasconcelos25} except that \LCP features linear
-channels instead of sessions. A session consisting of several messages exchanged
-in sequence can be encoded in \LCP using linear channels by pairing each message
-with a fresh continuation channel on which the rest of the session takes place.
+As we have anticipated in \Cref{sec:introduction}, \LCP is closely related to
+\CP~\citep{Wadler14,GayVasconcelos25} except that \LCP features linear channels
+instead of sessions. A session consisting of several exchanged messages is
+encoded using linear channels by pairing each message with a fresh continuation
+channel on which the rest of the session takes place.
 %
@@
 
-The interpretation of propositions as session types is the usual
-one~\cite{Wadler14}: the multiplicative constants $\Bot$ and $\One$ describe
+The interpretation of linear logic propositions as protocols is fairly
+usual~\cite{Wadler14}: the multiplicative constants $\Bot$ and $\One$ describe
 protocols that terminate with the input/output of a signal; the additive
 connectives $A \With B$ and $\A \Plus B$ describe protocols having two possible
-continuations $A$ and $B$ with are selected/offered; the multiplicative
+continuations of type $A$ and $B$ which are selected/offered; the multiplicative
 connectives $A \Par B$ and $A \Ten B$ describe bifurcating protocols with two
-continuations $A$ and $B$. These forms are typically used to describe the
-input/output of a channel (to be used according to $A$) and then continuing as
-$B$. The additive constants $\Zero$ and $\Top$ do not describe useful protocols,
-but they can play the role of smallest/largest type in type systems that support
-a notion of subtyping~\citep{HornePadovani24}. 
+continuations of type $A$ and $B$. The additive constants $\Zero$ and $\Top$ do
+not describe useful protocols, but they can play the role of smallest/largest
+type in type systems that support a notion of subtyping~\citep{HornePadovani24}. 
 %
 The quantifiers $\forall X.A$ and $\exists X.A$ describe the input/output of a
-type $X$ followed by the protocol $A$ and are useful to describe parametric
-protocol polymorphism.
+type $X$ with a continuation $A$ and are useful to describe parametric protocol
+polymorphism.
 %
 Finally, the ``of course'' modality $\OfCourse A$ and the ``why not'' modality
-$\WhyNot A$ describe servers and clients accepting and requesting connections to
-be used according to $A$.
+$\WhyNot A$ describe shared channels on which servers and clients accept and
+request connections of type $A$.
 
-The notions of \emph{free type variables}, \emph{duality} and of \emph{type
+The notions of \emph{free type variables}, of \emph{duality} and of \emph{type
 substitution} are standard~\cite{Wadler14}. In particular, we write $\fv{A}$ for
@@
 
-The syntax of \Calculus makes use of an infinite set of \emph{channels}, ranged
+The syntax of processes makes use of an infinite set of \emph{channels}, ranged
 over by $x$, $y$ and $z$, and is shown in~\Cref{tab:syntax}.
@@
 every message sent on one of the channels is forwarded to the other. As
-discussed in the literature~\cite{LindleyMorris15}, this form is also useful for
-modeling delegation, that is the exchange of an existing session endpoint in
-another session.
+discussed in the literature~\cite{LindleyMorris15}, this form is useful for
+modeling higher-order sessions, that is the exchange of an existing channel on
+another channel.
 %
-The $\Wait\x.P$ and $\Close\x$ processes respectively model the input and output
+The processes $\Wait\x.P$ and $\Close\x$ respectively model the input and output
 of a termination signal on the channel $x$. The latter process has no
@@
 %
-The process $\Fail\x$ can be used to denote a failing protocol concerning the
-channel $x$.
+The process $\Fail\x$ can be used to denote a failure concerning the channel
+$x$.
 %
@@
 received label $\InTag_i$. In practice it is useful to consider arbitrary
-(domain-specific) sets of tags instead of the canonical $\InTag_1$ and
+(domain-specific) sets of labels instead of the canonical $\InTag_1$ and
 $\InTag_2$. This generalisation has no substantial impact on the formalisation,
@@
 of a pair of fresh channels $y$ and $z$ on the channel $x$. The former process
-forks into $P$ and $Q$, each using $y$ and $z$ respectively. The latter process
-continues as $R$. These forms are often used in conjunction with links to model
-\emph{delegation}.
+has a single continuation $R$ that can use both $y$ and $z$. The latter process
+forks into $P$ and $Q$, each using $y$ and $z$ respectively. These forms are
+often used in conjunction with links to model higher-order sessions.
 %
@@
 The process $\Weaken\x.P$ denotes an explicit \emph{weakening}, that is a client
-that \emph{does not} use the server waiting on $x$. The process
-$\Contract\x\y\z.P$ denotes an explicit \emph{contraction}, that is a client
-that connects to the server waiting on $x$ multiple times on the channels $y$
-and $z$.
+that \emph{does not} use $x$. The process $\Contract\x\y\z.P$ denotes an
+explicit \emph{contraction} whereby a client uses $x$ multiple times (once with
+name $y$ and once with name $z$).
 
-\emph{Cuts} of the form $\Cut[A]\x{P}{Q}$ represent the parallel composition of
-the processes $P$ and $Q$ connected by a channel $x$, which has type $A$ in $P$
-and type $\dual{A}$ in $Q$. Henceforth we write $\Cut\x{P}{Q}$ omitting the type
-annotation $A$ when it is irrelevant or clear from the context.
+Finally, cuts of the form $\Cut[A]\x{P}{Q}$ represent the parallel composition
+of the processes $P$ and $Q$ connected by a channel $x$, which has type $A$ in
+$P$ and type $\dual{A}$ in $Q$. Henceforth we write $\Cut\x{P}{Q}$ omitting the
+type annotation $A$ when it is irrelevant or clear from the context.
 
@@
 output prefixes bind continuation channels in (some) continuation processes. In
-particular, $\Select\x{\Tag_i}\z.P$ binds $z$ in $P$, while $\Fork\x\y\z{P}{Q}$
-binds $y$ in $P$ but not in $Q$ and $z$ in $Q$ but not in $P$.
+particular, $\Select\x{\InTag_i}\z.P$ binds $z$ in $P$, while
+$\Fork\x\y\z{P}{Q}$ binds $y$ in $P$ but not in $Q$ and $z$ in $Q$ but not in
+$P$.
 %
@@
     Note the use of explicit continuations in these processes and the fact that
-    each channel is used \emph{exactly once}. The same processes in \CP would be
+    each channel is used exactly once. The same processes in \CP would be
     written as
@@
 \begin{table}[t]
-    \caption{Operational semantics of \Calculus. Some side conditions for the
+    \caption{Operational semantics of \Calculus. Many side conditions for the
     \rulename{s-*} rules are omitted (see text).}
@@
 %
-All these rules have implicit side-conditions aimed at preserving the meaning of
-channels when binders are moved around: terms entering or exiting the scope of a
-binder for $x$ must not have free occurrences of $x$. This holds also for the
-type variable $X$ in the rule \SForall.
+All these rules have implicit side-conditions (not shown in
+\Cref{tab:semantics}) aimed at preserving the meaning of channels when binders
+are moved around: terms entering or exiting the scope of a binder for $x$ must
+not have free occurrences of $x$. This holds also for the type variable $X$ in
+the rule \SForall.
 %
@@
 Note also that there are two versions of \SForkL and \SForkR depending on which
-of the two continuations $P$ and $Q$ contains free occurrences of the restricted
-channel $x$.
+of the two continuations (either $P$ or $Q$) contains free occurrences of the
+restricted channel $x$.
 
 There is no pre-congruence rule asserting that parallel composition is
-associative. While this is a sound and expected property of parallel
-composition, it turns out to be inessential for proving the properties of
-\Calculus that we consider. We leave it out in light of the forthcoming
-formalisation of the calculus, in which every simplification is welcome,
-although standard presentations of classical linear logic admit such
-law~\cite{Wadler14,Doumane17}.
+associative. While this is a sound property of parallel composition, it turns
+out to be unnecessary for proving the properties of \LCP that we consider. We
+leave it out in light of the forthcoming formalisation of the calculus, in which
+every simplification is welcome, although standard presentations of classical
+linear logic admit such law~\cite{Wadler14}.
 %
@@
 client, the disposal of an unused server, and the duplication of a server. In
-\RWeaken and \RContract we use some convenient (albeit slightly ambiguous)
+\RWeaken and \RContract we use some convenient (albeit slightly informal)
 notation for denoting sequences of (pairwise distinct) channels and prefixes. In
@@
 
-    We establish that $\EchoP(x)$ is well typed with the following deerivation
+    We establish that $\EchoP(x)$ is well typed with the following derivation
     \[
@@
 essential properties of well-typed processes: (1) typing is preserved by
-reductions; (2) peer endpoints are always used in complementary ways. This
-latter property is called \emph{well formedness} in the challenge. In this work
-we also consider \emph{deadlock freedom}, which is more general than well
-formedness and holds for \Calculus since its type system is based on linear
-logic. We now formulate these properties using the notation developed so far.
+reductions; (2) the peer endpoints of the same session are always used in
+complementary ways. This latter property is called \emph{well formedness} in the
+challenge. In this work we also consider \emph{deadlock freedom}, which is more
+general than well formedness and holds for \Calculus since its type system is
+based on linear logic. We now formulate these properties using the notation
+developed so far.
 
@@
 simple example of deadlock is the process $\Cut\x{\Close\x}{\Close\x}$. This
-process is unable to reduce (because a $\Close\x$ process is meant to interact
+process is unable to reduce (because a process $\Close\x$ is meant to interact
 with a process of the form $\Wait\x.P$) and, more generally, it is unable to
-interact regardless of the context in which it occurs because the $\Close\x$
-sub-processes it contains are blocked on the channel $x$ that is restricted by
-the cut.
+interact regardless of the context in which it occurs because the sub-processes
+it contains are blocked on the channel $x$ that is restricted by the cut.
 %
@@
 We now shift the attention to \emph{well formedness}. In the linearity
-challenge~\cite{CarboneEtAl24}, this property ensures that, whenever two
+challenge~\cite{CarboneEtAl24} this property ensures that, whenever two
 processes composed in parallel begin with actions concerning the very same
-session, then such actions are \emph{complementary}, in the sense that they
+session, then such actions complement each other, in the sense that they
 describe opposite forms of interaction. To define well formedness in our
@@
 the hole in $\ReductionContext$ with $P$, noting that such replacement may
-capture channel that are bound in $\ReductionContext$ and occur free in $P$.
+capture channels that are bound in $\ReductionContext$ and occur free in $P$.
 %

### 2025-12-25 paper/Congruence.lagda.tex
global revision

--- 2025-12-24 paper/Congruence.lagda.tex
@@
 %
-Notice that this constructor also makes key use of the implicit rewriting rule
-\AgdaFunction{dual-inv} described in \Cref{sec:type-agda}. Indeed $P$ and $Q$
-have type $\AgdaDatatype{Proc}~(A~\AgdaInductiveConstructor{∷}~\Context_1)$ and
+Notice that \AgdaInductiveConstructor{s-comm} makes key use of the implicit
+rewriting rule \AgdaFunction{dual-inv} described in \Cref{sec:type-agda}. Indeed
+$P$ and $Q$ have type
+$\AgdaDatatype{Proc}~(A~\AgdaInductiveConstructor{∷}~\Context_1)$ and
 $\AgdaDatatype{Proc}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\Context_2)$,
@@
     $A~\AgdaInductiveConstructor{∷}~\Context_1$ necessarily comes from the right
-    component of the splitting $[\Bot] + A~\AgdaInductiveConstructor{∷}~\Delta$
+    partition of the splitting $[\Bot] + A~\AgdaInductiveConstructor{∷}~\Delta$
     through the use of \AgdaInductiveConstructor{>}.
@@
 Finally, in \Cref{sec:semantics} we have colloquially defined $\pcong$ as a
-``\emph{pre-congruence}'', implying that it is reflexive, transitive relation
-and that it is is preserved by some forms of calculus. In the formalization we
-have to be precise and we introduce specific rules:
+``pre-congruence'', implying that it is a reflexive, transitive relation
+preserved by some forms of the calculus. In the formalisation we have to be
+precise and we introduce specific rules:
 
@@
 \end{AgdaAlign}
+
+We concede that the implementation of the pre-congruence rules (including those
+not discussed here) can be difficult to decipher. In part, this is due to the
+fact that splitting proofs are manifest and no longer hidden by separating
+conjunctions (as in \Cref{sec:process-agda}) because we need them to enforce the
+side conditions of the rules in \Cref{tab:semantics}. We should also bear in
+mind that, using an intrinsically-typed representation of processes,
+pre-congruence and reduction rules (which we describe next) incorporate the
+proof that they preserve typing.

### 2025-12-25 paper/Context.lagda.tex
global revision

@@
 \AgdaModule{Data.List} of Agda's standard library. We will keep using
-$\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts as we
-have done in \Cref{sec:typing-rules}, even though in the Agda formalisation they
-are lists instead of finite maps.
+$\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts as in
+\Cref{sec:typing-rules}, even though in the Agda formalisation they are lists
+instead of finite maps.
 
@@
 contains all the types contained in $\ContextD$ and $\ContextE$, preserving both
-their overall \emph{multiplicity} and also their relative \emph{order} within
-$\ContextD$ and $\ContextE$. A \emph{proof} of
-$\Splitting\ContextC\ContextD\ContextE$ shows how the types in $\ContextC$ are
-distributed in $\ContextD$ and $\ContextE$ from left to right.
+their overall multiplicity and also their relative order within $\ContextD$ and
+$\ContextE$. A \emph{proof} of $\Splitting\ContextC\ContextD\ContextE$ shows how
+the types in $\ContextC$ are distributed in $\ContextD$ and $\ContextE$ from
+left to right.
 
@@
 
-Context splitting enjoys a number of expected properties. In particular, it is
-easy to see that splitting is commutative and that the empty context/list is
-both a left and right unit of splitting.
+It is easy to see that splitting is commutative and that the empty context/list
+is both a left and right unit of splitting.
 
@@
 
-Context splitting is also associative in a sense that is made precise below. If
-we write $\ContextD + \ContextE$ for some $\ContextC$ such that $Γ ≃ Δ + Θ$, then
-we can prove that $Γ_1 + (Γ_2 + Γ_3) = (Γ_1 + Γ_2) + Γ_3$.
+Context splitting is also associative. If we write $\ContextD + \ContextE$ for
+some $\ContextC$ such that $Γ ≃ Δ + Θ$, then we can prove that $Γ_1 + (Γ_2 +
+Γ_3) = (Γ_1 + Γ_2) + Γ_3$.
 
@@
 
-Note that, when specifying a splitting $\Splitting\ContextC{[A]}\ContextE$, it
-may be convenient to use \AgdaFunction{≫} as a shortcut for a sequence of
-applications of \AgdaInductiveConstructor{>} after the $A$ type has been
-reached. For instance, \AgdaFunction{splitting-example₁} can be written
-equivalently and in a more compact way as shown below. More usages of
-\AgdaFunction{≫} will be provided in \Cref{sec:examples}.
+Note that, when specifying a splitting $\Splitting\ContextC{[A]}\ContextE$ where
+the left partition is a singleton $[A]$, it may be convenient to use
+\AgdaFunction{≫} as a shortcut for a sequence of applications of
+\AgdaInductiveConstructor{>} once the $A$ type has been reached in $\ContextC$.
+For instance, \AgdaFunction{splitting-example₁} can be written equivalently and
+in a more compact way as shown below. More usages of \AgdaFunction{≫} will be
+provided in \Cref{sec:examples}.
 
@@
     \hline
+    $\AgdaFunction{Pred}~A~ℓ$ & $A \to \AgdaDatatype{Set}~ℓ$ & predicate over $A$ \\
     $\AgdaFunction{∀[}~P~\AgdaFunction]$ & $\forall\{x\} \to P~x$ & implicit universality \\
@@
   \end{tabular}
-  \caption{\label{tab:predicates} Useful definitions from Agda's \AgdaModule{Relation.Unary} module.}
+  \caption{\label{tab:predicates} Useful definitions in Agda's \AgdaModule{Relation.Unary} module.}
 \end{table}
@@
 reason, it is worth recalling a number of definitions from the
-\AgdaModule{Relation.Unary} module in Agda's standard library in
+\AgdaModule{Relation.Unary} module of Agda's standard library in
 \Cref{tab:predicates}.
 
-Context splitting is going to be pervasive in the subsequent development, to the
-point that it contributes substantially to the complexity of the formalisation.
-For this reason, we invest some effort in defining a few abstractions that allow
-us to hide context splittings and enhance code readability.
+Context splitting is pervasive in the subsequent development. For this reason,
+we define a few abstractions that allow us to hide context splittings and
+enhance code readability, at least in some cases.
 %
@@
 
-The predicate $P \sand Q$ holds for those contexts $\ContextC$ that can be split
-into $\ContextD$ and $\ContextE$ where $P$ holds for $\ContextD$ and $Q$ holds
-for $\ContextE$.
+If $P$ and $Q$ are predicates over contexts, the predicate $P \sand Q$ holds for
+those contexts $\ContextC$ that can be split into $\ContextD$ and $\ContextE$
+such that $P$ holds for $\ContextD$ and $Q$ holds for $\ContextE$.
 %
@@
 witnessing the splitting $\Splitting\ContextC\ContextD\ContextE$ along with
-$P~\ContextD$ and $Q~\ContextE$.
+proofs of $P~\ContextD$ and $Q~\ContextE$.
 %
-The use of the metavariables $P$ and $Q$ for denoting predicates over contexts
-is not without reason. As we will see shortly, in our formalisation processes
-are indeed predicates over typing contexts, albeit not the only ones.
+The use of metavariables $P$ and $Q$ for denoting predicates over contexts is
+appropriate: as we will see shortly, in our formalisation processes are indeed
+an example of predicate over typing contexts.
 
@@
 
-To conclude the modeling of typing contexts, we define a predicate
+To conclude the implementation of typing contexts, we define a predicate
 \AgdaDatatype{Un} that holds for \emph{unrestricted} contexts, namely those

### 2025-12-25 paper/DeadlockFreedom.lagda.tex
global revision

@@
 now formalise in Agda. First of all we need to define the notion of
-\emph{thread}, that is any process other than a cut. As it turns out, it is
-convenient to provide a more fine-grained classification of threads,
-distinguishing between links, input/output actions, and sometimes also whether
-such actions operate on free or bound channels. We define predicates for each of
-these process classes:
+\emph{thread}, that is any process other than a cut. It is convenient to provide
+a more fine-grained classification of threads, distinguishing between links,
+input/output actions, and sometimes also whether such actions operate on free or
+bound channels. We define predicates for each of these process classes:
 
@@
 (\AgdaInductiveConstructor{cc-delayed}), or
-\item $P$ and $Q$ are servers and $P$ operates on a channel other than $x$
+\item both $P$ and $Q$ are servers and $P$ operates on a channel other than $x$
 (\AgdaInductiveConstructor{cc-servers}).
@@
 
-Note that we have to distinguish servers from the other input operations because
-the structural pre-congruence rule \SServer can only be applied when the two
+We have to distinguish servers from the other input operations because the
+structural pre-congruence rule \SServer can only be applied when the two
 sub-processes of a cut are both servers.

### 2025-12-25 paper/Examples.lagda.tex
global revision

@@
 For the encoding of the echo server (\Cref{ex:echo}), recall that the server is
-polymorphic in the sense that the type of the value it receives and sends back
-is arbitrary and denoted by a type variable $X$. The notation
-$\AgdaFunction{\#}~n$ is defined in Agda's standard library and creates an
-element of \AgdaDatatype{Fin} from the natural number $n$.
+polymorphic in the sense that the type $X$ of the value it receives is
+arbitrary. The notation $\AgdaFunction{\#}~n$ is defined in Agda's standard
+library and creates an element of \AgdaDatatype{Fin} from the natural number
+$n$.
 

### 2025-12-25 paper/Permutations.lagda.tex
global revision

@@
 
-The definition of \AgdaDatatype{\_↭\_} data type is nearly the same found in the
-module \AgdaModule{Data.List.Relation.Binary.Permutation.Propositional} of
+The definition of the data type \AgdaDatatype{\_↭\_} is nearly the same found in
+the module \AgdaModule{Data.List.Relation.Binary.Permutation.Propositional} of
 Agda's standard library. We have preferred defining our own notion of
@@
 \AgdaInductiveConstructor{swap} constructor does not need a sub-permutation for
-the tail of the typing context, which can always be performed, if needed, using
-\AgdaInductiveConstructor{prep} and \AgdaInductiveConstructor{trans}. Also, and
-more importantly, the data type is \emph{monomorphic} (it does not need to
-relate lists of arbitrary type) and the quantified arguments $A$ and $B$ of
-\AgdaInductiveConstructor{swap} and \AgdaInductiveConstructor{prep} can be made
-implicit, which streamlines the usage of these constructors in the rest of the
-code.
+the tail of the typing context, which can always be performed, if needed,
+combining \AgdaInductiveConstructor{swap} with \AgdaInductiveConstructor{prep}
+and \AgdaInductiveConstructor{trans}. Also, and more importantly, our data type
+\AgdaDatatype{\_↭\_} is monomorphic (it does not need to relate arbitrary lists)
+and the arguments $A$ and $B$ of \AgdaInductiveConstructor{swap} and
+\AgdaInductiveConstructor{prep} are implicit, which streamlines the usage of
+these constructors in the rest of the code.
 
@@
 Just like permutations in Agda's standard library, also \AgdaDatatype{\_↭\_}
-enjoys a number of expected properties, such as symmetry. Among them, in the
-following we will make use of a specific property that relates context
-splitting, list concatenation and permutations: if
+enjoys a number of expected properties, such as symmetry. In the following we
+will make use of a specific property that relates context splitting, list
+concatenation \AgdaFunction{\_++\_} and permutations: if
 $\Splitting\ContextC\ContextD\ContextE$, then $\ContextC$ is a permutation of

### 2025-12-25 paper/Process.lagda.tex
global revision

@@
 \item the typing preservation results are embedded in the very definition of the
-  operational semantics of processes and need not require separate proofs
+  operational semantics of processes and do not require separate proofs
   (\Cref{sec:congruence-agda,sec:reduction-agda}).
@@
 sure that channels are appropriately (and above all linearly) distributed among
-sub-processes, so that each channel occurs (and is used) exactly
-once.\footnote{Strictly speaking, the typing context of a
-\AgdaInductiveConstructor{fail} process may contain channels that do not occur
-(and therefore are not used) by the process, but we consider this case as an
-exception that does not alter in a substantial way the provided intuition
-underlying the use of co-de Bruijn representation.}
+sub-processes, so that each channel is used exactly once.
 %
@@
 (\AgdaFunction{U} is the universal predicate that holds for every context, \cf
-\Cref{tab:predicates}). That form of $\Context$ matches the one in the
-conclusion of \FailRule.
+\Cref{tab:predicates}). That form of $\Context$ matches the typing context in
+the conclusion of the rule \FailRule.
 
@@
 example, the constructor \AgdaInductiveConstructor{case} corresponds to a
-process of the form $\Case\x\y{P}{Q}$, where both $P$ and $Q$ may use the
+process of the form $\Case\x\y{P}{Q}$, where both $P$ and $Q$ use the
 continuation channel $y$. In this case $\Context$ must satisfy the predicate
@@
 $A$ has type $\AgdaDatatype{PreType}~\AgdaInductiveConstructor{1}$, that is a
-pre-type with \emph{at most one} free type variable ($X$), whereas $\Context$ is
-a typing context, that is a list of $\AgdaDatatype{Type} =
+pre-type with \emph{at most one} free type variable $X$, whereas $\Context$ is a
+typing context, that is a list of $\AgdaDatatype{Type} =
 \AgdaDatatype{PreType}~\AgdaInductiveConstructor{0}$ without free type
@@
 property as the fact that typing is preserved by (injective) name
-substitutions.\Luca{Da qualche parte mettere un remark per far notare che l'uso
-della separating conjunction introduce rumore, ma avere i costruttori
-currificati ne complica un po' l'interpretazione dei tipi. Forse è meglio fare
-questo rilievo nella prossima sezione quando è evidente il pattern matching che
-serve per esprimere le regole.}
+substitutions.
 

### 2025-12-25 paper/Reduction.lagda.tex
global revision

@@
 the reduction, $x$ is replaced by $y$ which is found \emph{somewhere} within
-$\Context$. The exact location of $y$ in $\ContextC$ is found in the splitting
+$\Context$. The exact location of $y$ in $\ContextC$ is encoded in the splitting
 $p$, thus we ``rename'' $x$ into $y$ within $P$ using the permutation
@@
 $\Context \simeq \AgdaInductiveConstructor{[]} + \Context$. Since the left
-context is empty, these splittings must be equal and made of a sequence of
+partition is empty, these splittings must be equal and made of a sequence of
 \AgdaInductiveConstructor{>} applications followed by
@@
 they are definitionally equal, hence it is easier to quantify them separately so
-that we do not have to worry about proving their equality whenever we wish to
-apply this reduction.
+that we do not have to prove their equality whenever we wish to apply this
+reduction.
 
@@
 The remaining constructors that describe the base reductions follow a similar
-pattern, noting that the implementation of \RWeaken and \RContract makes use of
-auxiliary functions that respectively weaken and contract the typing context of
+pattern, except for the implementation of \RWeaken and \RContract which require
+auxiliary functions to respectively weaken and contract the typing context of
 the resulting process as shown in \Cref{tab:semantics}. It is worth glancing at
@@
 $\AgdaDatatype{Proc}~(\AgdaFunction{subst}~[~B~/]~A~\AgdaInductiveConstructor{∷}~\Context_1)$
-and $(F~B)$ is a process of type
+and $F~B$ is a process of type
 $\AgdaDatatype{Proc}~(\AgdaFunction{subst}~[~B~/]~(\AgdaFunction{dual}~A)~\AgdaInductiveConstructor{∷}~\Context_2)$.
-In order for these two process to be composable in a cut, it must be the case
+In order for these two processes to be composable in a cut, it must be the case
 that $\AgdaFunction{subst}~[~B~/]~(\AgdaFunction{dual}~A) \equiv
@@
 implicit rewriting rule, we do not have to rewrite the index in the type of
-$(F~B)$, which is silently accepted as is.
+$F~B$, which is silently accepted as is.
 

### 2025-12-25 paper/Safety.lagda.tex
global revision

@@
 in the left (respectively, right) sub-context, as per the grammar of reduction
-contexts in \Cref{sec:properties}.
+contexts given in \Cref{sec:properties}.
 

### 2025-12-25 paper/Type.lagda.tex
global revision

@@
 type variables and we use de Bruijn indices for denoting type variables. In this
-way, we make sure that pre-types are well-scoped.
+way, we make sure that pre-types are well scoped.
 
@@
 \AgdaInductiveConstructor{`∃} which increase the number of bound type variables
-in the scoped (pre)type.
+in the scoped pre-type.
 
-The dual of a (pre)type is computed by the following function:
+The dual of a pre-type is computed by the following function:
 
@@
 directive.\footnote{The directive is effective provided that the option
-\texttt{{-}{-}rewriting} is enabled, either globally when invoking Agda or in
-the module's preamble.}
+\texttt{{-}{-}rewriting} is enabled, either globally when invoking Agda or
+within a pragma in the module's source code.}
 
@@
   \AgdaFunction{dual}~(\AgdaFunction{subst}~\sigma~(\AgdaInductiveConstructor{rav}~x))$
-  which is definitionally the same of $\sigma~x \equiv
+  which is definitionally the same as $\sigma~x \equiv
   \AgdaFunction{dual}~(\AgdaFunction{dual}~(\sigma~x))$.
@@
 We could easily prove this equivalence by invoking \AgdaFunction{dual-inv}, but
-thanks to the rewriting rule that we have added just above a use of
+thanks to the rewriting rule that we have added earlier a use of
 \AgdaInductiveConstructor{refl} suffices. In this case the saved effort is

### 2025-12-25 paper/agda.tex
british spelling
global revision

--- 2025-07-26 paper/agda.tex
@@
-\section{Agda Formalization}
+\section{Agda Formalisation}
 \label{sec:agda}
+
+In this section we describe the formalisation of \LCP in Agda. We present in
+detail only some key parts of the code including the representation of types and
+processes, the definition of the operational semantics and the statement of the
+main results. The interested reader may find the complete source code in the
+public repository.
 

### 2025-12-25 paper/conclusion.tex
global revision

@@
 $\pi$-calculus~\cite{KobayashiPierceTurner99} supported by the same type system
-of \CP~\cite{Wadler14}. Binary sessions can be modeled in \LCP using the
+of \CP~\cite{Wadler14}. Binary sessions are modeled in \LCP using the
 continuation-passing encoding described by \citet[extended
@@
 We remark that \LCP and the calculus described in the linearity
-challenge~\cite{CarboneEtAl24} are fundamentally incomparable. On the one hand,
-the challenge only considers a minimal calculus of first-order, monomorphic
-sessions while \LCP is a fully-fledged calculus that supports linear, shared,
-higher-order, polymorphic channels. On the other hand, the logical foundations
-of \LCP prevent the modeling of deadlocks and of sequential processes that own
-both endpoints of a session, both of which can be modeled in the calculus of the
-challenge.
+challenge~\cite{CarboneEtAl24} are incomparable in terms of expressiveness. On
+the one hand, the challenge only considers a minimal calculus of first-order,
+monomorphic sessions while \LCP is a fully-fledged calculus that supports
+linear, shared, higher-order, polymorphic channels. On the other hand, the
+logical foundations of \LCP prevent the modeling of deadlocked processes and of
+sequential processes that own both endpoints of a session. These processes can
+be modeled in the calculus of the challenge, where parallel composition and name
+restriction are distinct constructs as in the $\pi$-calculus.
 %
@@
 have already extended \LCP with support for infinite types and recursive
-processes (the formalisation of this extension is publicly available in a
-sibling folder of the one for \LCP). In the future, it would be interesting to
-formalise the strong normalisation property of \LCP as a consequence of cut
-elimination of classical linear logic.
+processes (the source code of this extension is publicly available in a sibling
+folder of the one for \LCP). In the future, it would be interesting to formalise
+the strong normalisation property of \LCP as a consequence of cut elimination of
+classical linear logic.
 

### 2025-12-25 paper/related.tex
global revision

@@
 
-\citet{Thiemann19} formalises a subset of \GV~\citep{GayVasconcelos10}, a
-functional language extended with session communication primitives, along with
-an interpreter. \citet{CicconePadovani20} have taken inspiration from Thiemann's
-work to formalise a variant of the linear
+\citet{Thiemann19} formalises a subset of \GV~\citep{GayVasconcelos10,Wadler14},
+a functional language extended with session communication primitives, along with
+an interpreter. \citet{CicconePadovani20} have taken inspiration from this work
+to formalise a variant of the linear
 $\pi$-calculus~\cite{KobayashiPierceTurner99} that supports dependent types, so
@@
 \citet{RouvoetPoulsenKrebbersVisser20} present a library of abstractions
-inspired to separation logic aiding the formalisation of interpreters for
-computing models supported by linear type systems. One of the presented case
-studies is the formalisation of a fragment of \GV~\citep{Wadler14}. The
-separating conjunction we have introduced in \Cref{sec:context-agda} and the
-typing of the constructors for the representation of processes in
-\Cref{sec:process-agda} have been heavily inspired by this work.
+inspired to separation logic aiding the formalisation of interpreters for models
+supported by linear type systems. One of the presented case studies is the
+formalisation of a fragment of \GV~\citep{GayVasconcelos10,Wadler14}. The
+separating conjunction introduced in \Cref{sec:context-agda} and the typing of
+the constructors for the representation of processes in \Cref{sec:process-agda}
+have been heavily inspired by this work.
 
@@
 % small-step semantics is defined) and subject reduction is
+
+\citet{JacobsBalzerKrebbers22} formalise a library of \emph{connectivity graphs}
+for reasoning and enforcing deadlock freedom in a variant of
+\GV~\citep{GayVasconcelos10,Wadler14}. This is the first formalisation of
+deadlock freedom for a calculus of sessions.
 
@@
 contexts beforehand. As \citet{ZalakainDardha21} nicely summarise, ``context
-splits contain usage information that is already present'' in processes, and
-therefore can be omitted. This is true, but only provided that channels are
-given a \emph{name} in processes (\citet{ZalakainDardha21} use de Bruijn indices
-to this aim). In fact, the co-de Bruijn representation of
-processes~\cite{McBride18}, whereby channels are nameless and context splitting
-is performed eagerly, can be seen as the ``dual approach'' of leftover typing:
-channel names provide information that is already present in their (singleton)
-typing context, hence they can be omitted from contexts and processes.
-
-Another work based on context splitting is the one by
-\citet{JacobsBalzerKrebbers22}, who formalise a library of \emph{connectivity
-graphs} for reasoning and enforcing deadlock freedom in a variant of
-\GV~\citep{GayVasconcelos10,Wadler14}. This is the first formalisation of
-deadlock freedom for a calculus of sessions.
+splits contain usage information that is already present in processes,'' and
+therefore can be omitted. This is true provided that channels are named
+(\citet{ZalakainDardha21} use de Bruijn indices to this aim). In fact, the co-de
+Bruijn representation of processes~\cite{McBride18}, whereby channels are
+nameless and context splitting is performed eagerly, can be seen as the ``dual
+approach'' of leftover typing: channel names provide information that is already
+present in their (singleton) typing context, hence they can be omitted from
+contexts and processes.
 
 Motivated by the technical difficulties arising from context splits,
-\citet{SanoKavanaghPientka23} model a structural version of \CP using an
+\citet{SanoKavanaghPientka23} define a structural version of \CP using an
 alternative approach based on \emph{linearity predicates}. The key idea is to
-treat typing contexts structurally and to enforce the linear usage of sessions
-by checking the syntactical occurrence of channels within processes.
-Interestingly, the approach relies on the \emph{explicit naming of
-continuations} so as to precisely account for the number of times a channel is
-actually used.
+treat typing contexts structurally and to enforce the linear usage of channels
+by checking their syntactical occurrence in processes. Interestingly, the
+approach relies on the \emph{explicit naming of continuations} so as to
+precisely account for the number of times a channel is actually used.
 
@@
 contexts whereby channels are associated not just with a type but also with a
-\emph{tag} (or \emph{multiplicity}) from a given resource algebra for describing
-their availability. This approach streamlines context splitting in the sense
-that contexts can be treated in an essentially structural way, except for tags
-which are conveninently combined using operations from the resource algebra.
+\emph{tag} (or \emph{multiplicity}) from a given resource algebra. The tag
+summarises the number of allowed usages of a particular channel, including the
+possibility that the channel is not available. This approach streamlines context
+splitting since contexts can be treated in an essentially structural way, except
+for tags which are conveninently combined using operations from the resource
+algebra.
 
@@
 
-\Cref{tab:comparison} summarises the comparison of our formalisation with the
-aforementioned ones. The three leftmost colums list the reference papers (sorted
-by publication date), the proof assistant and the formalised calculus.
+\Cref{tab:comparison} reports a summary of the aforementioned formalisations and
+our own. The three leftmost colums list the reference papers (sorted by
+publication date), the proof assistant and the formalised calculus.
 %
@@
 
-The rightmost columns of \Cref{tab:comparison} report the \emph{size} (in
-kilobytes) of the formalisations as rough (and possibly questionable) estimates
-of their complexity. Instead of counting line numbers, which may be affected by
-code indentation styles and syntactical constraints of the proof assistant being
+The rightmost columns of \Cref{tab:comparison} report the size (in kilobytes) of
+the formalisations as rough (and possibly questionable) estimates of their
+complexity. Instead of counting line numbers, which may be affected by code
+indentation styles and syntactical constraints of the proof assistant being
 used, we have counted the number of characters after comments have been removed
@@
 whole range of sizes. Our formalisation is based on context splitting and it is
-the only one that covers \CP in full, yet it is by far the smallest one.
+the only one that supports all the featurs of \CP, yet it is also the smallest
+one.
 
 The largest formalisations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are also
-the ones that adopt a locally nameless representation of channels. In this
+the ones that adopt a locally nameless representation of channels. In these
 formalisations channels are represented in two different ways, depending on

### 2025-12-25 paper/calculus.tex
global revision

@@
 connectives $A \With B$ and $\A \Plus B$ describe protocols having two possible
-continuations of type $A$ and $B$ which are selected/offered; the multiplicative
-connectives $A \Par B$ and $A \Ten B$ describe bifurcating protocols with two
-continuations of type $A$ and $B$. The additive constants $\Zero$ and $\Top$ do
-not describe useful protocols, but they can play the role of smallest/largest
-type in type systems that support a notion of subtyping~\citep{HornePadovani24}. 
+continuations of type $A$ and $B$ which are respectively offered/selected; the
+multiplicative connectives $A \Par B$ and $A \Ten B$ describe bifurcating
+protocols with two continuations of type $A$ and $B$ that are respectively
+received and sent. The additive constants $\Zero$ and $\Top$ do not describe
+useful protocols, but they can play the role of smallest/largest type in type
+systems that support a notion of subtyping~\citep{HornePadovani24}. 
 %
@@
 %
-A link $\Link\x\y$ denotes the merging of the channels $x$ and $y$, so that
-every message sent on one of the channels is forwarded to the other. As
-discussed in the literature~\cite{LindleyMorris15}, this form is useful for
-modeling higher-order sessions, that is the exchange of an existing channel on
-another channel.
+A link $\Link\x\y$ denotes the merging of the channels $x$ and $y$, so that the
+message sent on one of the channels is forwarded to the other. As discussed in
+the literature~\cite{LindleyMorris15}, this form is useful for modeling
+higher-order sessions, that is the exchange of an existing channel on another
+channel.
 %
@@
 describe the input/output of a label $\InTag_i$ along with a fresh continuation
-$z$ on the channel $x$. The receiver continues as $Q_i$ depending on the
-received label $\InTag_i$. In practice it is useful to consider arbitrary
-(domain-specific) sets of labels instead of the canonical $\InTag_1$ and
-$\InTag_2$. This generalisation has no substantial impact on the formalisation,
-so we stick to binary choices in line with the usual presentations of linear
-logic.
+$z$ on the channel $x$. The receiver continues as $Q_i$ depending on the label
+$\InTag_i$. In practice it is useful to consider arbitrary (domain-specific)
+sets of labels instead of the canonical $\InTag_1$ and $\InTag_2$. This
+generalisation has no substantial impact on the formalisation, so we stick to
+binary choices in line with the usual presentations of linear logic.
 %
@@
     \label{ex:booleans}
-    We can represent the type of a boolean value being produced as the
-    proposition $\Bool \eqdef \One \Plus \One$ and that of a boolean value being
-    consumed as its dual $\dual\Bool = \Bot \With \Bot$. Following these types,
-    the boolean constants can be modeled by the processes
+    We can represent the type of a boolean value being produced as the protocol
+    $\Bool \eqdef \One \Plus \One$ and that of a boolean value being consumed as
+    its dual $\dual\Bool = \Bot \With \Bot$. Following these types, the boolean
+    constants can be modeled by the processes
     \[
@@
     \]
-    where each channel is used multiple times to indicate the various
+    where each channel is used multiple times to indicate the sequence of
     input/output actions pertaining to the same session.
@@
 rules are key to float input/output actions to the top-level of a process, so
-that they can interact with corresponding complementary actions in the context
-surrounding the process.
+that they can interact with corresponding complementary actions in the
+surrounding context.
 %
@@
 Note also that there are two versions of \SForkL and \SForkR depending on which
-of the two continuations (either $P$ or $Q$) contains free occurrences of the
+of the two continuations (either $P$ or $Q$) contains a free occurrence of the
 restricted channel $x$.
@@
 every simplification is welcome, although standard presentations of classical
-linear logic admit such law~\cite{Wadler14}.
+linear logic admit such law~\cite{Wadler14,Doumane17}.
 %
@@
 a session $x$ when a termination signal is exchanged on $x$. \RSelect and \RFork
-respectively model the communication of a tag $\InTag_i$ and of a fresh channel
-$y$. \RExists models the instantiation of a polymorphic variable $X$ as the
-communication of a type $A$. We write $Q\subst{A}\X$ for the process obtained by
-replacing every free occurrence of the type variable $X$ with $A$. Each of these
-communications consumes the channel $x$ and possibly creates and transfers a
-\emph{fresh continuation} $z$ on which the rest of the conversation takes place,
-in line with the encoding of sessions into the linear
+respectively model the communication of a tag $\InTag_i$ and of a pair of fresh
+continuation channels. \RExists models the instantiation of a polymorphic
+variable $X$ as the communication of a type $A$. We write $Q\subst{A}\X$ for the
+process obtained by replacing every free occurrence of the type variable $X$
+with $A$. Each of these communications consumes the channel $x$ and possibly
+creates and transfers \emph{fresh continuations} on which the rest of the
+conversation takes place, in line with the encoding of sessions into the linear
 $\pi$-calculus~\cite{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
@@
 proof rules of classical linear logic, the reader may refer to the standard
-literature on linear logic or propositions as sessions~\citep{Wadler14} for
-their interpretation.
+literature on propositions as sessions~\citep{Wadler14} for their
+interpretation.
 
@@
 
-    The server accepts requests on the shared channel $x$. Each request
-    established a new session, initially represented by the channel $y$. The
-    session consists in receiving the type $X$ of the value being echoed
-    followed by the value itself $u$. The value is then sent back unchanged and
-    the session is terminated. Note the succession of different (linear)
-    channels $y, y', y'', y'''$ that chain the session together.
+    The server accepts requests on the shared channel $x$. Each request creates
+    a new session, initially represented by the channel $y$. The session
+    consists in receiving the type $X$ of the value being echoed followed by the
+    value itself $u$. The value is then sent back unchanged and the session is
+    terminated. Note the succession of different (linear) channels $y, y', y'',
+    y'''$ that chain the session together.
 
@@
 a thread is either a link or a process that starts with an input/output action
-of some sort. Note that a thread may contain cuts, but these cuts must all
+of some sort. Note that a thread may contain cuts, but these cuts must be
 guarded underneath the topmost action prefix of the thread.

### 2025-12-25 paper/introduction.tex
revision
global revision

@@
 These models often make use of distinctive features that set them apart from the
-formal models of \emph{sequential} programming languages, such as the adoption
-of substructural (linear, affine) type systems, the dynamic scope of first-class
+formal models of sequential programming languages, such as the adoption of
+substructural (linear, affine) type systems, the dynamic scope of first-class
 channels in systems of communicating processes, the need for coinductive
 definitions and proof methods for describing and reasoning on possibly infinite
-behaviors.
+behaviours.
 %
@@
 properties of a program; (2) the type of a session endpoint is \emph{updated}
-after each use to reflect the state of the interaction it describes; (3) peer
+after each use to reflect the state of the protocol it describes; (3) peer
 session endpoints are meant to be used in complementary ways so as to guarantee
@@
 for a \emph{single} communication. Structured communications involving the
-exchange of several messages within the same session is modeled in \LCP using
+exchange of several messages within the same session are modeled in \LCP using
 \emph{explicit continuations} according to the well-known encoding of (binary)

### 2025-12-25 paper/abstract.tex
expanded abstract

--- 2025-12-24 paper/abstract.tex
@@
 \abstract{%
-    We describe the Agda formalisation of a process calculus based on classical
-    linear logic that supports the modeling of binary sessions. The
-    formalisation turns out to be remarkably compact despite the use of
-    ``plain'' context splitting for the proper management of linear channels. We
-    argue that the logical nature of the calculus and the use of explicit
-    continuations are contributing factors to the simplicity of its
-    formalisation.
+    The formalisation of session calculi is considered challenging due to the
+    management of channels, which are linear resources that cannot be discarded
+    or duplicated at will and whose type changes over time, as input/output
+    operations are performed on them.
+    %
+    Context splitting, the ``naive'' channel management technique directly
+    related to the way session type theories are usually sketched with pen and
+    paper, is often considered a hindrance and a notable source of complexity,
+    to the point that several alternative approaches have been recently
+    proposed.
+    %
+    In this paper we describe the Agda formalisation of a process calculus based
+    on classical linear logic that supports the modeling of binary sessions
+    through their encoding with explicit continuation channels. The
+    formalisation turns out to be remarkably compact despite the adoption of
+    context splitting. We argue that the logical nature of the calculus and the
+    use of explicit continuations are contributing factors to the simplicity of
+    its formalisation.
     %

### 2025-12-25 paper/Congruence.lagda.tex
global revision

@@
 
-The datatype for $\pcong$ has one constructor for each of the structural
-precongruence rules in \Cref{tab:semantics}. Since many aspects recur
-repeatedly, we will focus on just a few representative rules starting from
-\SComm.
+The datatype for $\pcong$ has one constructor for each structural pre-congruence
+rule in \Cref{tab:semantics}. Since many aspects recur repeatedly, we illustrate
+the implementation of just a few representative rules starting from \SComm.
 
@@
 There are two non-trivial aspects worth commenting. The first one concerns the
-proof $(\AgdaInductiveConstructor{>}~q)$ used in the
+proof $\AgdaInductiveConstructor{>}~q$ used in the
 \AgdaInductiveConstructor{wait} process before structural pre-congruence is
@@
   \item The \SWait rule is applicable only provided that the channel restricted
-    by the cut (say $x$, of type $A$) is different from the channel (say $y$, of
-    type $\Bot$) consumed by the \AgdaInductiveConstructor{wait} process.
+    by the cut (say $x$, of type $A$) is different from the channel  consumed by
+    the \AgdaInductiveConstructor{wait} process (say $y$, of type $\Bot$).
     %
@@
     \Cref{tab:semantics}) imposing that the type $A$ in front of
-    $A~\AgdaInductiveConstructor{∷}~\Context_1$ necessarily comes from the right
-    partition of the splitting $[\Bot] + A~\AgdaInductiveConstructor{∷}~\Delta$
-    through the use of \AgdaInductiveConstructor{>}.
+    $A~\AgdaInductiveConstructor{∷}~\Context_1$ goes to the right partition of
+    the splitting $[\Bot] + A~\AgdaInductiveConstructor{∷}~\Delta$ through the
+    use of \AgdaInductiveConstructor{>}.
 \end{enumerate}
@@
 the form $\Cut[A]\x{\Fork\y\u\v{P}{Q}}{R}$ when $x \in \fn{P}$. We capture this
-condition by means of the splitting $(\AgdaInductiveConstructor{<}~r)$, meaning
+condition by means of the splitting $\AgdaInductiveConstructor{<}~r$, meaning
 that the type $A$ of $x$ ends up in the typing context for $P$ and not in the
@@
 \AgdaInductiveConstructor{s-fork-r}, not shown here but similar to
-\AgdaInductiveConstructor{s-fork-l} except that
-$(\AgdaInductiveConstructor{<}~r)$ is replaced by
-$(\AgdaInductiveConstructor{>}~r)$.
+\AgdaInductiveConstructor{s-fork-l} except that $\AgdaInductiveConstructor{<}~r$
+is replaced by $\AgdaInductiveConstructor{>}~r$.
 

### 2025-12-25 paper/Context.lagda.tex
global revision

@@
 
-We are going to adopt a nameless representation of channels hence typing
-contexts are represented using lists of types, where the (polymorphic)
+We are going to adopt a nameless representation of channels. Hence, typing
+contexts are represented as lists of types, where the (polymorphic)
 \AgdaDatatype{List} type and its constructors \AgdaInductiveConstructor{[]} and
@@
 \AgdaModule{Data.List} of Agda's standard library. We will keep using
-$\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts as in
-\Cref{sec:typing-rules}, even though in the Agda formalisation they are lists
-instead of finite maps.
+$\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts, even
+though in the Agda formalisation they are lists and not finite maps as in
+\Cref{sec:typing-rules}.
 
@@
 and \AgdaInductiveConstructor{>} to indicate whether the type is meant to be
-placed in $\ContextD$ or $\ContextE$.
+placed in $\ContextD$ or in $\ContextE$.
 %
@@
 \begin{code}
-  splitting-example₁ : (A ∷ B ∷ C ∷ D ∷ []) ≃ (B ∷ []) + (A ∷ C ∷ D ∷ [])
+  splitting-example₁ : (A ∷ B ∷ C ∷ D ∷ []) ≃ [ B ] + (A ∷ C ∷ D ∷ [])
   splitting-example₁ = > < > > •
@@
 If $P$ and $Q$ are predicates over contexts, the predicate $P \sand Q$ holds for
-those contexts $\ContextC$ that can be split into $\ContextD$ and $\ContextE$
-such that $P$ holds for $\ContextD$ and $Q$ holds for $\ContextE$.
+those contexts $\ContextC$ that can be split into $\ContextD$ and $\ContextE$ so
+that $P$ holds for $\ContextD$ and $Q$ holds for $\ContextE$.
 %
@@
 To conclude the implementation of typing contexts, we define a predicate
-\AgdaDatatype{Un} that holds for \emph{unrestricted} contexts, namely those
-contexts solely made of types of the form $?A$. We will need this predicate in
-the definition of a server, which must obey the typing rule \ServerRule of
-\Cref{tab:typing-rules}.
+\AgdaDatatype{Un} that holds for \emph{unrestricted} contexts, those solely made
+of types of the form $?A$. We need this predicate in the definition of a server,
+which must comply with the typing rule \ServerRule in \Cref{tab:typing-rules}.
 

### 2025-12-25 paper/Examples.lagda.tex
global revision

@@
 
-In this section we revisit and expand the sample processes discussed in
-\Cref{ex:booleans,ex:echo} and show their encoding in our formalisation. The
-encoding of $\Bool$ is straightforward.
+In this section we revisit the processes discussed in \Cref{ex:booleans,ex:echo}
+and show their encoding in our formalisation. The encoding of $\Bool$ is
+straightforward.
 
@@
 
-The function \AgdaFunction{\$} is defined in Agda's standard library and denotes
-function application. We use it here as ``separator'' to save a few parentheses.
+The function \AgdaFunction{\$} (defined in Agda's standard library) is just a
+visible function application operator. We use it here as ``separator'' to save a
+few parentheses.
 %

### 2025-12-25 paper/Permutations.lagda.tex
global revision

@@
 
-According to our nameless representation of channels, it is the \emph{order} of
-channels in a typing context $\Context$ that determines the binding location in
-the structure of a process. When the binding structure of a process changes, \eg
-because a structural pre-congruence rule is applied, or when a channel
-substitution occurs, \cf the right-hand side of the \LinkRule reduction in
-\Cref{tab:semantics}, the typing context must be suitably rearranged. Such
+According to our nameless representation of channels, it is the \emph{position}
+of a type in a typing context $\Context$ that determines the location of its
+binder in the structure of a process. When the binding structure of a process
+changes, \eg because a structural pre-congruence rule is applied, or when a
+channel substitution occurs, \cf the right-hand side of the \LinkRule reduction
+in \Cref{tab:semantics}, the typing context must be suitably rearranged. Such
 rearrangement is in fact a \emph{permutation} of the elements of $\Context$.
@@
 Agda's standard library. We have preferred defining our own notion of
-permutation for improved simplicity and convenience: the
-\AgdaInductiveConstructor{swap} constructor does not need a sub-permutation for
-the tail of the typing context, which can always be performed, if needed,
-combining \AgdaInductiveConstructor{swap} with \AgdaInductiveConstructor{prep}
-and \AgdaInductiveConstructor{trans}. Also, and more importantly, our data type
+permutation for simplicity and convenience: the \AgdaInductiveConstructor{swap}
+constructor does not need a sub-permutation for the tail of the typing context,
+which can always be performed, if needed, combining
+\AgdaInductiveConstructor{swap} with \AgdaInductiveConstructor{prep} and
+\AgdaInductiveConstructor{trans}. Also, and more importantly, our data type
 \AgdaDatatype{\_↭\_} is monomorphic (it does not need to relate arbitrary lists)

### 2025-12-25 paper/Process.lagda.tex
global revision

@@
 Since $x$ is the most recently introduced channel, the types $A$ and $\dual{A}$
-must be added \emph{at the front} of $\ContextD$ and $\ContextE$, which we do by
+is added \emph{at the front} of $\ContextD$ and $\ContextE$, which we do by
 means of the functions $(A~\AgdaInductiveConstructor{∷\_})$ and
@@
 list permutations basically correspond to channel renaming, we can read this
-property as the fact that typing is preserved by (injective) name
-substitutions.
+property as the fact that typing is preserved by (bijective) name substitutions.
 

### 2025-12-25 paper/Reduction.lagda.tex
global revision

@@
 between processes that are well typed in the very same typing context. Thus, the
-definition of reduction embeds the subject reduction result.
+definition of reduction embeds the subject reduction result (\Cref{thm:red}).
 
@@
 sub-process of the cut (that is the link $\Link\x\y$) and the splitting
-$(\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{>}~\AgdaInductiveConstructor{•})$
+$\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{>}~\AgdaInductiveConstructor{•}$
 to which \AgdaInductiveConstructor{link} is applied is structured consistently
@@
 \AgdaInductiveConstructor{wait} constructor uses a splitting proof of the form
-$(\AgdaInductiveConstructor{<}~q_0)$ to be sure that the referenced channel is
+$\AgdaInductiveConstructor{<}~q_0$ to be sure that the referenced channel is
 also the restricted one, and therefore matches the one of the

### 2025-12-25 paper/Type.lagda.tex
global revision

@@
 The representation of types is standard, with one constructor for each of the
-forms described in \Cref{sec:types}. We start by defining \emph{pre-types} as an
-indexed data type $\AgdaDatatype{PreType}~n$ where $n$ is the number of bound
-type variables and we use de Bruijn indices for denoting type variables. In this
-way, we make sure that pre-types are well scoped.
+forms described in \Cref{sec:types}. We start by defining an indexed data type
+$\AgdaDatatype{PreType}~n$ to represents (\LCP) types in the scope of $n$
+quantifiers and we use elements of $\AgdaDatatype{Fin}~n$ as de Bruijn indices
+for the quantified type variables. In this way, we make sure that pre-types are
+well scoped.
 
@@
 their dual, and the quantifiers \AgdaInductiveConstructor{`∀} and
-\AgdaInductiveConstructor{`∃} which increase the number of bound type variables
-in the scoped pre-type.
+\AgdaInductiveConstructor{`∃} which increase the number of quantifiers in the
+scoped pre-type.
 
@@
 
-This property is so important in the rest of the formalisation that it is worth
-defining an implicit rewriting rule so that Agda can autonomously apply it
-whenever possible. This is achieved by means of the following pragma
-directive.\footnote{The directive is effective provided that the option
-\texttt{{-}{-}rewriting} is enabled, either globally when invoking Agda or
-within a pragma in the module's source code.}
+This property is important in the rest of the formalisation so we define an
+implicit rewriting rule that Agda can autonomously apply whenever possible. This
+is achieved by means of the following pragma directive.\footnote{The directive
+is effective provided that the option \texttt{{-}{-}rewriting} is enabled,
+either globally when invoking Agda or within a pragma in the module's source
+code.}
 
@@
 Next we define the function \AgdaFunction{subst} that simultaneously substitutes
-the type variables of a (pre)type with other (pre)types. Even though in practice
-we will always substitute one variable at a time, it is technically easier to
-substitutes \emph{all} variables of a (pre)type. The definition of
-\AgdaFunction{subst} relies on some auxiliary functions for \emph{renaming} type
-variables and \emph{lifting} substitutions across quantifiers. All such
-definitions are straightforward adaptations of those described by
-\citet{KokkeSiekWadler20}.
+the type variables of a pre-type with other pre-types. In practice we will
+always substitute one variable at a time, but it is technically easier to
+substitute \emph{all} variables of a pre-type, possibly with themselves. The
+definition of \AgdaFunction{subst} relies on some auxiliary functions for
+\emph{renaming} type variables and \emph{lifting} substitutions across
+quantifiers. These functions are straightforward adaptations of those described
+by \citet{KokkeSiekWadler20}.
 
@@
 Among all substitutions, we will use the one that substitutes the 0-indexed type
-variable with a (pre)type. It is convenient to introduce this substitution once
+variable with a pre-type. It is convenient to introduce this substitution once
 and for all, which we do here.
@@
 
-At last we are ready to define the type of closed pre-types, those
-having no free type variables. We call \AgdaDatatype{Type} closed pre-types.
-From now on, we will seldom use pre-types again.
+We call \AgdaDatatype{Type} closed pre-types, those having no free type
+variables. From now on, we will seldom use pre-types again.
 

### 2025-12-25 paper/calculus.tex
global revision

@@
 
-The \emph{linearity challenge}~\citep{CarboneEtAl24} aims at formalising two
-essential properties of well-typed processes: (1) typing is preserved by
-reductions; (2) the peer endpoints of the same session are always used in
-complementary ways. This latter property is called \emph{well formedness} in the
-challenge. In this work we also consider \emph{deadlock freedom}, which is more
-general than well formedness and holds for \Calculus since its type system is
-based on linear logic. We now formulate these properties using the notation
-developed so far.
+The linearity challenge~\citep{CarboneEtAl24} aims at formalising two essential
+properties of well-typed processes: (1) typing is preserved by reductions; (2)
+the peer endpoints of the same session are always used in complementary ways.
+This latter property is called \emph{well formedness} in the challenge. In this
+work we also consider \emph{deadlock freedom}, which is more general than well
+formedness and holds for \Calculus since its type system is based on linear
+logic. We now formulate these properties using the notation developed so far.
 

### 2025-12-25 paper/conclusion.tex
global revision

@@
 %
-It is quite interesting to note that, among the approaches that have been
-proposed to overcome context splitting, the one of \citet{SanoKavanaghPientka23}
-makes use of explicit continuations.
+It is quite striking that, among the approaches that have been proposed to
+overcome context splitting, the one of \citet{SanoKavanaghPientka23} makes use
+of explicit continuations.
 
@@
 have already extended \LCP with support for infinite types and recursive
-processes (the source code of this extension is publicly available in a sibling
-folder of the one for \LCP). In the future, it would be interesting to formalise
-the strong normalisation property of \LCP as a consequence of cut elimination of
-classical linear logic.
+processes (the source code of this extension is publicly available in the same
+repository as the one that contains the formalisation of \LCP). In the future,
+it would be interesting to formalise the strong normalisation property of \LCP
+as a consequence of cut elimination of classical linear logic.
 
@@
 (those connecting exactly two processes), but there are also formalisations of
-\emph{multiparty sessions}, notably those of \citet{JacobsBalzerKrebbers22a} and
-\citet{TiroreBengstonCarbone25}, which are significantly more complex than those
-of binary sessions. Just to give an example, the formalisation by
-\citet{TiroreBengstonCarbone25} amounts to more than 1Mb of Coq code.
-Incidentally, multiparty sessions are not supported by a strong connection with
-linear logic and there is no known continuation-passing encoding of multiparty
-sessions. Hence, it is unclear whether the considerations made for this work
-also apply to the multiparty case.
+\emph{multiparty sessions}, notably those by \citet{JacobsBalzerKrebbers22a} and
+\citet{TiroreBengstonCarbone25}, which can be significantly more complex than
+those of binary sessions. The formalisation by \citet{JacobsBalzerKrebbers22a}
+amounts to 173kb and the one by \citet{TiroreBengstonCarbone25} to more than 1Mb
+of Coq code. Also in these cases, the formalisation based on (virtual)
+continuations~\citep{JacobsBalzerKrebbers22a} happens to be substantially
+smaller. Whether this is a coincidence or further evidence of the effectiveness
+of the continuation-based approaches is left for future work.

### 2025-12-25 paper/related.tex
global revision

@@
 \citet{CastroFerreiraYoshida20} describe EMTST, a library for the formalisation
-of session type systems which includes as case studies the session calculus of
+of session type systems that includes as case studies the session calculus of
 \citet{HondaVasconcelosKubo98} (called ``original system'') and a revised
@@
 precisely account for the number of times a channel is actually used.
+\citet{SanoKavanaghPientka23} do not connect their technique with the
+continuation-passing encoding of binary sessions, but \emph{a posteriori} the
+analogies between the two approaches are evident.
 
@@
 \Cref{tab:comparison} reports a summary of the aforementioned formalisations and
-our own. The three leftmost colums list the reference papers (sorted by
+also our own. The three leftmost colums list the reference papers (sorted by
 publication date), the proof assistant and the formalised calculus.
@@
 because, while not being strictly based on the linear $\pi$-calculus, it makes
-use of explicit continuations for defining the linearity predicates that check
-the linear usage of channels. Finally, all the approaches based on \GV are also
-marked with \MAYBE. In principle, \emph{none} of these calculi makes use of
-continuations. However, \GV is designed in such a way that each session
-operation acting on a channel $s$ returns the result of the operation (if any)
-\emph{along with the same channel $s$}. In this way, the type of $s$ can be
+use of explicit continuations for defining the predicates that check the linear
+usage of channels. Finally, all the approaches based on \GV are also marked with
+\MAYBE. In principle, \emph{none} of these calculi makes use of continuations.
+However, \GV is designed in such a way that each operation acting on a channel
+$s$ is a function that returns the result of the operation (if any) \emph{along
+with the same channel $s$}. In this way, the type of $s$ can be conveniently
 ``updated'' to take into account the effect of the operation. As observed by
-\citet{Padovani17}, this way of modeling the communication primitives is
-virtually indistinguishable from one making use of explicit continuation
-channels.
+\citet{Padovani17}, this semantics of the communication primitives is virtually
+indistinguishable from one making use of explicit continuation channels.
 
@@
 formalisations differ widely for a variety of entangled factors including the
-reference calculus, supported features, proof assistant,
-process/expression/channel representation, handling of linearity. Also, most
+features of calculus, the used proof assistant, the representation of
+processes/expressions/channels, the handling of linearity. Also, most
 formalisations predate the publication of the linearity
@@
 %
-The use of context splitting, which is very well represented, does not seem to
-be a good indicator of the complexity of the formalisation, since it spans the
-whole range of sizes. Our formalisation is based on context splitting and it is
-the only one that supports all the featurs of \CP, yet it is also the smallest
-one.
+The adoption of context splitting, which is very well represented, does not seem
+to be a good indicator of the complexity of the formalisation. Indeed, the
+formalisations based on context splitting span the whole range of sizes, from
+the largest by \citet{GotoEtAl16} (543kb) to our own (21kb, without the proof of
+deadlock freedom) which is also the only one supporting all the featurs of \CP.
 
-The largest formalisations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are also
-the ones that adopt a locally nameless representation of channels. In these
+The two largest formalisations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are the
+ones that adopt a locally nameless representation of channels. In these
 formalisations channels are represented in two different ways, depending on
@@
 
-There is no strong evidence that the intrinsically typed representation of
+We found no strong evidence that the intrinsically typed representation of
 processes reduces the size of the formalisation. As observed in
@@
 
-Finally, it seems that the use of (explicit) continuations helps taming the
+Finally, it seems that the use of (explicit) continuations affects the
 complexity of the formalisation more than anything else. Indeed, the six

### 2025-12-26 paper/calculus.tex
LCC

@@
-\section{Linear \CP}
+\section{A Linear Calculus of Continuations}
 \label{sec:calculus}
 
-In this section we give a cursory presentation of \LCP.
+In this section we give a cursory presentation of \Calculus starting from its
+types (\Cref{sec:types}), which coincide with classical linear logic
+propositions, then moving on to the syntax of processes (\Cref{sec:syntax}),
+their reduction semantics (\Cref{sec:semantics}), the typing rules
+(\Cref{sec:typing-rules}) and the formulation of the properties ensured by the
+type system (\Cref{sec:properties}).
+
+As we have anticipated in \Cref{sec:introduction}, \Calculus is closely related
+to \CP~\citep{Wadler14,GayVasconcelos25} except that \Calculus features linear
+channels instead of sessions.
 %
-As we have anticipated in \Cref{sec:introduction}, \LCP is closely related to
-\CP~\citep{Wadler14,GayVasconcelos25} except that \LCP features linear channels
-instead of sessions. A session consisting of several exchanged messages is
-encoded using linear channels by pairing each message with a fresh continuation
-channel on which the rest of the session takes place.
-%
-This way of encoding sessions has been first described by \citet[extended
-version]{Kobayashi02b} and later studied by \citet{DardhaGiachinoSangiorgi17}.
-
-We present \LCP starting from its types (\Cref{sec:types}), which coincide with
-classical linear logic propositions, then moving on to the syntax of processes
-(\Cref{sec:syntax}), their reduction semantics (\Cref{sec:semantics}), the
-typing rules (\Cref{sec:typing-rules}) and the formulation of the properties
-ensured by the type system (\Cref{sec:properties}).
+We refer the reader to the literature on
+\CP~\cite{Wadler14,LindleyMorris15,GayVasconcelos25} and other session calculi
+based on linear logic~\cite{CairesPfenning10,CairesPfenningToninho16} for a
+thorough introduction to these models.
 
@@
 
-The interpretation of linear logic propositions as protocols is fairly
-usual~\cite{Wadler14}: the multiplicative constants $\Bot$ and $\One$ describe
-protocols that terminate with the input/output of a signal; the additive
-connectives $A \With B$ and $\A \Plus B$ describe protocols having two possible
-continuations of type $A$ and $B$ which are respectively offered/selected; the
-multiplicative connectives $A \Par B$ and $A \Ten B$ describe bifurcating
-protocols with two continuations of type $A$ and $B$ that are respectively
-received and sent. The additive constants $\Zero$ and $\Top$ do not describe
-useful protocols, but they can play the role of smallest/largest type in type
-systems that support a notion of subtyping~\citep{HornePadovani24}. 
+The interpretation of linear logic propositions as behaviours is fairly
+standard. Constants and connectives describe \emph{linear channels}, which must
+be used for a single communication, whereas the modalities $\OfCourse$ and
+$\WhyNot$ describe \emph{shared channels}.
 %
-The quantifiers $\forall X.A$ and $\exists X.A$ describe the input/output of a
-type $X$ with a continuation $A$ and are useful to describe parametric protocol
-polymorphism.
+The multiplicative constants $\Bot$ and $\One$ describe channels used for
+receiving/sending an empty message (without continuations).
+%
+The additive connectives $A \With B$ and $A \Plus B$ describe  channels used for
+receiving/sending either a continuation of type $A$ or a continuation of type
+$B$. The sender selects one of the two possibilities, while the receiver offers
+both.
+%
+The multiplicative connectives $A \Par B$ and $A \Ten B$ describe channels used
+for receiving/sending two continuations, one of type $A$ and the other of type
+$B$.
+%
+The additive constants can play the role of smallest/largest type in type
+systems that support a notion of subtyping~\citep{HornePadovani24}. We will not
+be using them in this work.
+%
+The quantifiers $\forall X.A$ and $\exists X.A$ describe channels used for
+receiving/sending a type $X$ along with a continuation of type $A$. They are
+useful to describe parametric protocol polymorphism.
 %
@@
 %
-A link $\Link\x\y$ denotes the merging of the channels $x$ and $y$, so that the
+A link $\Link\x\y$ denotes the merging of the channels $x$ and $y$, so that each
 message sent on one of the channels is forwarded to the other. As discussed in
@@
 The processes $\Wait\x.P$ and $\Close\x$ respectively model the input and output
-of a termination signal on the channel $x$. The latter process has no
-continuation, while the former process continues as $P$ once the signal has been
-received.
+of an empty message on the channel $x$. The latter process has no continuation,
+while the former process continues as $P$ once the message has been received.
 %
@@
 %
-The processes $\Case\x\z{Q_1}{Q_2}$ and $\Select\x{\InTag_i}\z.P$ for $i=1,2$
-describe the input/output of a label $\InTag_i$ along with a fresh continuation
-$z$ on the channel $x$. The receiver continues as $Q_i$ depending on the label
-$\InTag_i$. In practice it is useful to consider arbitrary (domain-specific)
-sets of labels instead of the canonical $\InTag_1$ and $\InTag_2$. This
-generalisation has no substantial impact on the formalisation, so we stick to
-binary choices in line with the usual presentations of linear logic.
+The process $\Case\x\z{Q_1}{Q_2}$ offers a choice on channel $x$ and continues
+as either $Q_1$ or $Q_2$ with $z$ bound to the received continuation depending
+on which branch is selected.
+%
+The process $\Select\x{\InTag_i}\z.P$ performs a choice (represented by a label
+$\InTag_i$ with $i=1,2$) and sends a fresh continuation $z$ on the channel $x$.
 %
 The processes $\Join\x\y\z.R$ and $\Fork\x\y\z{P}{Q}$ describe the input/output
-of a pair of fresh channels $y$ and $z$ on the channel $x$. The former process
-has a single continuation $R$ that can use both $y$ and $z$. The latter process
-forks into $P$ and $Q$, each using $y$ and $z$ respectively. These forms are
-often used in conjunction with links to model higher-order sessions.
+of two fresh continuations channels $y$ and $z$ on the channel $x$. The receiver
+can use $y$ and $z$ in whatever order. The sender forks into $P$ and $Q$, each
+using $y$ and $z$ respectively. These forms are often used in conjunction with
+links to model higher-order sessions.
 %
@@
 Next we have process forms dealing with shared (non-linear) channels. The
-$\Server\x\y.P$ and $\Client\x\y.P$ processes respectively denote \emph{servers}
+processes $\Server\x\y.P$ and $\Client\x\y.P$ respectively denote \emph{servers}
 and \emph{clients} acting on the shared channel $x$. Each request (from a
@@
     \label{ex:booleans}
-    We can represent the type of a boolean value being produced as the protocol
+    We can represent the protocol of a boolean value being produced as the type
     $\Bool \eqdef \One \Plus \One$ and that of a boolean value being consumed as
@@
     %
-    In general, the \CP version of an \LCP process can be obtained by reusing
+    In general, the \CP version of an \Calculus process can be obtained by reusing
     the same channel $x$ in place of the continuation $z$ in \Cref{tab:syntax}.
@@
 
-The operational semantics of \Calculus is shown in~\Cref{tab:semantics} and
-is given by two relations: a \emph{structural pre-congruence} relation $\pcong$
-and a \emph{reduction} relation $\red$. Structural pre-congruence relates
-processes that we consider to be essentially indistinguishable. Reduction
-relates processes in which a communication occurs. In some presentations of
-linear logic~\cite{BaeldeDoumaneSaurin16,Doumane17}, these relations are
-respectively dubbed \emph{external} and \emph{internal} reductions. Let us
-describe the two relations more in detail.
+The operational semantics of \Calculus is shown in~\Cref{tab:semantics} and is
+given by two relations: a \emph{structural pre-congruence} relation $\pcong$ and
+a \emph{reduction} relation $\red$. Structural pre-congruence relates processes
+that we consider to be essentially indistinguishable. Reduction relates
+processes in which a communication occurs. Let us describe the two relations
+more in detail.
 
@@
 associative. While this is a sound property of parallel composition, it turns
-out to be unnecessary for proving the properties of \LCP that we consider. We
+out to be unnecessary for proving the properties of \Calculus that we consider. We
 leave it out in light of the forthcoming formalisation of the calculus, in which
@@
 %
-The reduction rules \RLink, \RClose, \RSelect, \RFork, \RExists, \RConnect,
-\RWeaken and \RContract are in one-to-one correspondence with the well-known
-reductions of session calculi and the principal cut reductions of linear logic.
-\RLink eliminates a link $\Link\x\y$ by substituting $y$ for $x$. \RClose closes
-a session $x$ when a termination signal is exchanged on $x$. \RSelect and \RFork
-respectively model the communication of a tag $\InTag_i$ and of a pair of fresh
-continuation channels. \RExists models the instantiation of a polymorphic
-variable $X$ as the communication of a type $A$. We write $Q\subst{A}\X$ for the
-process obtained by replacing every free occurrence of the type variable $X$
-with $A$. Each of these communications consumes the channel $x$ and possibly
-creates and transfers \emph{fresh continuations} on which the rest of the
-conversation takes place, in line with the encoding of sessions into the linear
-$\pi$-calculus~\cite{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
+The reduction rules \RLink, \RClose, \RSelect, \RFork, \RExists and \RConnect
+are in one-to-one correspondence with the well-known reductions of session
+calculi and the principal cut reductions of linear logic. These reductions erase
+the topmost cut and replace it with zero, one or two new cuts, depending on the
+number of continuations that are exchanged. Note that \RLink eliminates a link
+$\Link\x\y$ by substituting $y$ for $x$ in the scope of $x$ and \RClose models
+the communication of an empty message (without continuations), hence no new cut
+is present in the reduct. \RSelect, \RFork and \RConnect are standard and
+\RExists models the instantiation of a polymorphic variable $X$ as the
+communication of a type $A$. We write $Q\subst{A}\X$ for the process obtained by
+replacing every free occurrence of the type variable $X$ with $A$.
 %
-The rules \RConnect, \RWeaken and \RContract describe interactions on shared
-channels and respectively represent the connection between a server and a
-client, the disposal of an unused server, and the duplication of a server. In
-\RWeaken and \RContract we use some convenient (albeit slightly informal)
-notation for denoting sequences of (pairwise distinct) channels and prefixes. In
-particular, $\seqof\z$ stands for a sequence $z_1, \dots, z_n$ of channels,
-$\Weaken{\seqof\z}.Q$ stands for $\Weaken{z_1}\dots\Weaken{z_n}.Q$ and
+The reduction rules \RWeaken and \RContract respectively represent the disposal
+of an unused server and the duplication of a server. In these rules we use some
+convenient (albeit slightly informal) notation for denoting sequences of
+(pairwise distinct) channels and prefixes. In particular, $\seqof\z$ stands for
+a sequence $z_1, \dots, z_n$ of channels, $\Weaken{\seqof\z}.Q$ stands for
+$\Weaken{z_1}\dots\Weaken{z_n}.Q$ and
 $\Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.R$ stands for
@@
 reduction up to structural pre-congruence.
+
+\newcommand\MonoEchoS{\mathit{Server}}
+\newcommand\MonoEchoC{\mathit{Client}}
+\begin{example}
+    \label{ex:neg}
+    %
+    We illustrate an example of higher-order session by modeling an echo server
+    that consumes boolean values and sends them back unchanged. The protocol
+    implemented by the server is described by the type $\OfCourse(\dual\Bool
+    \Par (\Bool \Ten \One))$ where the modality~$\OfCourse$ indicates that the
+    server is able to accept an unbounded number of requests and the type
+    $\dual\Bool \Par (\Bool \Ten \One)$ describes the sequence of actions
+    performed by the server at each connection with a client: the server first
+    consumes a boolean value (say $u$, of type $\dual\Bool$), then produces
+    another boolean value (say $v$, of type $\Bool$) and finally sends an empty
+    message. We can model the server and a possible client thus:
+    \begin{align*}
+        \MonoEchoS(x) & \eqdef
+        \Server\x\y.
+        \Join\y\u{y'}.
+        \Fork{y'}\v{y''}{
+            \Link\v\u
+        }{
+            \Close{y''}
+        } \\
+        \MonoEchoC(x,z) & \eqdef
+        \Client\x\y.
+        \Fork\y\u{y'}{
+            \TrueP(u)
+        }{
+            \Join{y'}\v{y''}.
+            \Wait{y''}.
+            \Link\v\z
+        }
+    \end{align*}
+
+    Notice the use of continuations for chaining communications together. In the
+    server, the link $\Link\v\u$ merges $v$ and $u$ so as to send the same
+    channel $u$ received from the client. In the client, the link $\Link\v\z$
+    ``assigns'' the message $v$ received from the server to the free channel
+    $z$, which represents the result of the interaction.
+    %
+    If we write $\reds$ for the reflexive, transitive closure of $\red$ it is
+    easy to verify that $\Cut\x{\MonoEchoC(x)}{\MonoEchoS(x)} \red^* \TrueP(z)$,
+    as expected.
+    %
+    \eoe
+\end{example}
 

### 2025-12-26 paper/introduction.tex
LCC

@@
 %
-Sessions and session types~\citep{Honda93,HondaVasconcelosKubo98,HuttelEtAl16}
-are established abstractions for the static analysis of distributed programs
-based on peer-to-peer communication channels. Every session type system revolves
-around three key ideas: (1) session endpoints are \emph{linear resources} that
-cannot be discarded or duplicated without compromising some safety and liveness
-properties of a program; (2) the type of a session endpoint is \emph{updated}
-after each use to reflect the state of the protocol it describes; (3) peer
-session endpoints are meant to be used in complementary ways so as to guarantee
-the absence of communication errors and, to some extent, progress of the
+Sessions and session
+types~\citep{Honda93,HondaVasconcelosKubo98,HuttelEtAl16,GayVasconcelos25} are
+established abstractions for the static analysis of distributed programs based
+on peer-to-peer communications. Every session type system revolves around three
+key ideas: (1) session endpoints are \emph{linear resources} that cannot be
+discarded or duplicated without compromising some safety and liveness properties
+of a program; (2) the type of a session endpoint is \emph{updated} after each
+use to reflect the state of the protocol it describes; (3) peer session
+endpoints are meant to be used in complementary ways so as to guarantee the
+absence of communication errors and, to some extent, progress of the
 interaction.
@@
 %
-More specifically, we describe a calculus called \LCP whose type system
-coincides with the proof system of classical linear logic and that features
-\emph{linear channels} instead of sessions. Linear channels are meant to be used
-for a \emph{single} communication. Structured communications involving the
-exchange of several messages within the same session are modeled in \LCP using
-\emph{explicit continuations} according to the well-known encoding of (binary)
-sessions into the linear
-$\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
+More specifically, we describe the Linear Calculus of Continuations (\LCC) whose
+type system coincides with the proof system of classical linear logic and that
+features \emph{linear channels} instead of sessions. Linear channels are meant
+to be used for a \emph{single} communication. Structured conversations can be
+modeled in \LCC using \emph{explicit continuations} as in the well-known
+encoding of (binary) sessions into the linear
+$\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}:
+each message exchanged on a linear channel may include one or two fresh channels
+-- the \emph{continuations} -- on which the rest of the conversation takes
+place.
 %
-Overall, \LCP is a linear version of \CP~\cite{Wadler14} in which sessions are
-encoded instead of being featured natively.
+Overall, \LCC is nothing but a lower-level, linear version of
+\CP~\cite{Wadler14}, Wadler's calculus of sessions based on classical linear
+logic, such that sessions are encoded instead of being featured natively.
 
-The logical foundations of \LCP and the use of explicit continuations play an
+The logical foundations of \LCC and the use of explicit continuations play an
 important role in taming the complexity of the formalisation.

### 2025-12-26 paper/calculus.tex
shortened poly echo server

@@
 
-\newcommand\MonoEchoS{\mathit{Server}}
-\newcommand\MonoEchoC{\mathit{Client}}
 \begin{example}
@@
     \begin{align*}
-        \MonoEchoS(x) & \eqdef
+        \ServerP(x) & \eqdef
         \Server\x\y.
@@
         } \\
-        \MonoEchoC(x,z) & \eqdef
+        \ClientP(x,z) & \eqdef
         \Client\x\y.
@@
     If we write $\reds$ for the reflexive, transitive closure of $\red$ it is
-    easy to verify that $\Cut\x{\MonoEchoC(x)}{\MonoEchoS(x)} \red^* \TrueP(z)$,
+    easy to verify that $\Cut\x{\ClientP(x)}{\ServerP(x)} \red^* \TrueP(z)$,
     as expected.
@@
 x_1 : \WhyNot A_1, \dots, x_n : \WhyNot A_n$ where all the types in its range
-are prefixed by the $\WhyNot$ modality. We call these types \emph{unrestricted},
-since they are used to denote shared channels that can be (explicitly) discarded
-and duplicated at will.
+are prefixed by the modality~$\WhyNot$. We call these types \emph{unrestricted}
+because they are used to denote shared channels that can be (explicitly)
+discarded and duplicated.
 
@@
             }{
-                \wtp{P}{x : \WhyNot A, \Context}
+                \wtp{\Weaken\x.P}{x : \WhyNot A, \Context}
             }
@@
             }{
-                \wtp{P}{x : \WhyNot A, \Context}
+                \wtp{\Contract\x\y\z.P}{x : \WhyNot A, \Context}
             }
@@
 well typed in the context $\Context$. Equivalently, the judgment indicates that
-the sequent $\wtp{}\Context$ is derivable and $P$ is a \emph{proof term}
-corresponding to the derivation for $\wtp{}\Context$.
+the sequent $\wtp{}\Context$ is derivable and $P$ is a proof term corresponding
+to the derivation for $\wtp{}\Context$.
 %
-The typing rules are shown in \Cref{tab:typing-rules}. Since they are in
-one-to-one correspondence with -- and have exactly the same structure of -- the
-proof rules of classical linear logic, the reader may refer to the standard
-literature on propositions as sessions~\citep{Wadler14} for their
-interpretation.
+The typing rules are shown in \Cref{tab:typing-rules}. They are in one-to-one
+correspondence with -- and have exactly the same structure of -- the proof rules
+of classical linear logic.
+%
+The reader may refer to the standard literature on propositions as
+sessions~\citep{Wadler14,LindleyMorris15,GayVasconcelos25} for the
+interpretation of the rules. The only relevant difference with the typing rules
+in these papers is that the premises mentioning the continuation channel $z$
+actually refer to the same channel $x$ on which the process in the conclusion is
+acting.
 
@@
     \label{ex:echo}
-    In this example we define a polymorphic echo server:
+    Looking at \Cref{ex:echo} we notice that the server does not make any
+    assumption on the type of the values it receives and sends. Therefore, we
+    can define a polymorphic version of the echo server that works for every
+    message type and not just for the booleans.
+    %
+    The polymorphic version of the echo server is defined below:
     \[
-        \EchoP(x) \eqdef
+        \ServerP(x) \eqdef
         \Server\x\y.
@@
 
-    The server accepts requests on the shared channel $x$. Each request creates
-    a new session, initially represented by the channel $y$. The session
-    consists in receiving the type $X$ of the value being echoed followed by the
-    value itself $u$. The value is then sent back unchanged and the session is
-    terminated. Note the succession of different (linear) channels $y, y', y'',
-    y'''$ that chain the session together.
-
-    We establish that $\EchoP(x)$ is well typed with the following derivation
+    The only difference with respect to the server in \Cref{ex:echo} is that now
+    the process receives the type $X$ of the messages to be processed and then
+    continues as before.
+    %
+    We establish that $\ServerP(x)$ is well typed with the following derivation
     \[
@@
                             }{
-                                u : X,
-                                v : \dual\X
+                                u : \dual\X,
+                                v : X
                             }
@@
                         }{
-                            u : X,
-                            y'' : \dual\X \Ten \One
+                            u : \dual\X,
+                            y'' : X \Ten \One
                         }
@@
                     }{
-                        y' : X \Par (\dual\X \Ten \One)
+                        y' : \dual\X \Par (X \Ten \One)
                     }
@@
                 }{
-                    y : \forall\X.X \Par (\dual\X \Ten \One)
+                    y : \forall\X.\dual\X \Par (X \Ten \One)
                 }
@@
             \wtp{
-                \EchoP(x)
+                \ServerP(x)
             }{
-                x : \OfCourse(\forall\X.X \Par (\dual\X \Ten \One))
+                x : \OfCourse(\forall\X.\dual\X \Par (X \Ten \One))
             }

### 2025-12-26 paper/calculus.tex
changed label

@@
                     & | & \Case\x\z{P}{Q}         & \text{case} \\
-                    & | & \Select\x{\InTag_i}\z.P & \text{left/right selection} \\
+                    & | & \Select\x{\InTag_i}\z.P & \text{select} \\
                     & | & \Join\x\y\z.P           & \text{join} \\

### 2025-12-26 paper/Type.lagda.tex
LCP => LCC

@@
 forms described in \Cref{sec:types}. We start by defining an indexed data type
-$\AgdaDatatype{PreType}~n$ to represents (\LCP) types in the scope of $n$
+$\AgdaDatatype{PreType}~n$ to represents (\LCC) types in the scope of $n$
 quantifiers and we use elements of $\AgdaDatatype{Fin}~n$ as de Bruijn indices

### 2025-12-26 paper/conclusion.tex
LCP => LCC

@@
 
-We have presented a formalisation of \LCP, a variant of the linear
+We have presented a formalisation of \LCC, a variant of the linear
 $\pi$-calculus~\cite{KobayashiPierceTurner99} supported by the same type system
-of \CP~\cite{Wadler14}. Binary sessions are modeled in \LCP using the
+of \CP~\cite{Wadler14}. Binary sessions are modeled in \LCC using the
 continuation-passing encoding described by \citet[extended
@@
 Under the assumption that the size of a formalisation is a good measure of its
-complexity, the formalisation of \LCP can be qualified as ``simple'' compared to
+complexity, the formalisation of \LCC can be qualified as ``simple'' compared to
 the existing ones.
@@
 
-We remark that \LCP and the calculus described in the linearity
+We remark that \LCC and the calculus described in the linearity
 challenge~\cite{CarboneEtAl24} are incomparable in terms of expressiveness. On
 the one hand, the challenge only considers a minimal calculus of first-order,
-monomorphic sessions while \LCP is a fully-fledged calculus that supports
+monomorphic sessions while \LCC is a fully-fledged calculus that supports
 linear, shared, higher-order, polymorphic channels. On the other hand, the
-logical foundations of \LCP prevent the modeling of deadlocked processes and of
+logical foundations of \LCC prevent the modeling of deadlocked processes and of
 sequential processes that own both endpoints of a session. These processes can
@@
 %
-Because of its logical foundations, we like to think of \LCP as a ``canonical''
+Because of its logical foundations, we like to think of \LCC as a ``canonical''
 calculus that deserves its own space in the context of the linearity challenge
@@
 
-The simple formalisation of \LCP is a good basis for further investigations. We
-have already extended \LCP with support for infinite types and recursive
+The simple formalisation of \LCC is a good basis for further investigations. We
+have already extended \LCC with support for infinite types and recursive
 processes (the source code of this extension is publicly available in the same
-repository as the one that contains the formalisation of \LCP). In the future,
-it would be interesting to formalise the strong normalisation property of \LCP
+repository as the one that contains the formalisation of \LCC). In the future,
+it would be interesting to formalise the strong normalisation property of \LCC
 as a consequence of cut elimination of classical linear logic.

### 2025-12-26 paper/introduction.tex
LCP => LCC

@@
 
-The formalisation of \LCP that we obtain is both the most complete (in terms of
+The formalisation of \LCC that we obtain is both the most complete (in terms of
 features supported by the calculus) and the most streamlined (in terms of code
@@
 %
-\Cref{sec:calculus} describes the syntax and the operational semantics of \LCP
+\Cref{sec:calculus} describes the syntax and the operational semantics of \LCC
 and states the properties of well-typed processes that we formalise and prove,

### 2025-12-26 paper/related.tex
LCP => LCC

@@
 average size of around 62kb -- are all based on continuations, whether they are
-explicit (\LinearPI, \SCP, \LCP) or ``virtual'' (\GV), while the remaining ones
+explicit (\LinearPI, \SCP, \LCC) or ``virtual'' (\GV), while the remaining ones
 are 263kb on average. This observation is in agreement with the considerations

### 2025-12-26 paper/Context.lagda.tex
global revision

@@
 We are going to adopt a nameless representation of channels. Hence, typing
-contexts are represented as lists of types, where the (polymorphic)
-\AgdaDatatype{List} type and its constructors \AgdaInductiveConstructor{[]} and
+contexts are represented as lists of types, where the (polymorphic) type
+\AgdaDatatype{List} and its constructors \AgdaInductiveConstructor{[]} and
 \AgdaInductiveConstructor{\_∷\_} are defined in the module
@@
 
-Note that, when specifying a splitting $\Splitting\ContextC{[A]}\ContextE$ where
-the left partition is a singleton $[A]$, it may be convenient to use
-\AgdaFunction{≫} as a shortcut for a sequence of applications of
-\AgdaInductiveConstructor{>} once the $A$ type has been reached in $\ContextC$.
-For instance, \AgdaFunction{splitting-example₁} can be written equivalently and
-in a more compact way as shown below. More usages of \AgdaFunction{≫} will be
-provided in \Cref{sec:examples}.
+When proving a splitting $\Splitting\ContextC{[A]}\ContextE$ where the left
+partition is a singleton $[A]$, it may be convenient to use \AgdaFunction{≫} as
+a shortcut for a sequence of applications of \AgdaInductiveConstructor{>} once
+the $A$ type has been reached in $\ContextC$. For instance,
+\AgdaFunction{splitting-example₁} can be written equivalently and in a more
+compact way as shown below. More usages of \AgdaFunction{≫} will be provided in
+\Cref{sec:examples}.
 
@@
 
-Hereafter we will make extensive use of predicates over contexts. For this
-reason, it is worth recalling a number of definitions from the
-\AgdaModule{Relation.Unary} module of Agda's standard library in
-\Cref{tab:predicates}.
-
-Context splitting is pervasive in the subsequent development. For this reason,
-we define a few abstractions that allow us to hide context splittings and
-enhance code readability, at least in some cases.
+From now on we will make extensive use of predicates over contexts. For this
+reason, it is worth recalling in \Cref{tab:predicates} a number of definitions
+from the module \AgdaModule{Relation.Unary} of Agda's standard library.
 %
-Inspired by separation logic~\cite{OHearnPym99} and
-following~\citet{RouvoetPoulsenKrebbersVisser20}, we define the \emph{separating
+We begin using these definitions for building a few abstractions inspired to
+separation logic~\cite{OHearnPym99} that allow us to hide context splittings, at
+least in some cases.
+%
+Following~\citet{RouvoetPoulsenKrebbersVisser20}, we define the \emph{separating
 conjunction} $P \sand Q$ of two predicates $P$ and $Q$ over contexts:
@@
 
-It is easy to prove that $\Splitting\ContextC\ContextD\ContextE$ is unrestricted
-if so are $\ContextD$ and $\ContextE$:
+It is easy to prove that $\ContextC$ is unrestricted if so are $\ContextD$ and
+$\ContextE$ when $\Splitting\ContextC\ContextD\ContextE$:
 

### 2025-12-26 paper/Permutations.lagda.tex
global revision

@@
 channel substitution occurs, \cf the right-hand side of the \LinkRule reduction
-in \Cref{tab:semantics}, the typing context must be suitably rearranged. Such
-rearrangement is in fact a \emph{permutation} of the elements of $\Context$.
+in \Cref{tab:semantics}, $\Context$ must be suitably rearranged to agree with
+the updated binding structure. Such rearrangement is in fact a
+\emph{permutation} of the elements of $\Context$.
 
@@
 
-Just like permutations in Agda's standard library, also \AgdaDatatype{\_↭\_}
-enjoys a number of expected properties, such as symmetry. In the following we
-will make use of a specific property that relates context splitting, list
-concatenation \AgdaFunction{\_++\_} and permutations: if
+It is easy to see that \AgdaDatatype{\_↭\_} is an equivalence relation. In the
+following we also use another specific property of permutations related to
+context splitting and list concatenation \AgdaFunction{\_++\_}: if
 $\Splitting\ContextC\ContextD\ContextE$, then $\ContextC$ is a permutation of

### 2025-12-26 paper/Process.lagda.tex
global revision

@@
 witnessing that their type is present in the typing context. This is known as
-\emph{co-de Bruijn syntax}~\cite{McBride18}, according to which the typing
-context associated with a term (a process, a channel) only contains the types of
-the channels that actually occur within the term. For this reason, typing
-contexts are split eagerly, according to the structure of processes, to make
-sure that channels are appropriately (and above all linearly) distributed among
+\emph{co-de Bruijn syntax}~\cite{McBride18}, whereby the typing context
+associated with a term (a process, a channel) only contains the types of the
+channels that actually occur within the term. For this reason, typing contexts
+are split eagerly, according to the structure of processes, to make sure that
+channels are appropriately (and above all linearly) distributed among
 sub-processes, so that each channel is used exactly once.
@@
 
-The constructor \AgdaInductiveConstructor{link} models a process of the form
-$\Link\x\y$. This process is well typed in a context of the form $x : A, y :
-\dual{A}$, namely a context satisfying the predicate
+The constructor \AgdaInductiveConstructor{link} builds a link $\Link\x\y$. This
+process is well typed in a context of the form $x : A, y : \dual{A}$, namely a
+context satisfying the predicate
 $\AgdaDatatype{Ch}~A~\AgdaDatatype{∗}~\AgdaDatatype{Ch}~(\AgdaFunction{dual}~A)$
 which we see on the left-hand side of \AgdaFunction{⇒}.
-%
-The constructor \AgdaInductiveConstructor{cut} models a cut $\Cut[A]\x{P}{Q}$.
-For this process to be well typed in a context $\Context$, it must be the case
-that $\Splitting\Context\ContextD\ContextE$ so that $P$ and $Q$ are well typed
-in the contexts $x : A, \ContextD$ and $x : \dual{A}, \ContextE$ respectively.
-Since $x$ is the most recently introduced channel, the types $A$ and $\dual{A}$
-is added \emph{at the front} of $\ContextD$ and $\ContextE$, which we do by
-means of the functions $(A~\AgdaInductiveConstructor{∷\_})$ and
-$(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷\_})$ (these are partial
+
+The constructor \AgdaInductiveConstructor{cut} builds a cut $\Cut[A]\x{P}{Q}$.
+This process is well typed in a context $\Context$ if
+$\Splitting\Context\ContextD\ContextE$ so that $P$ and $Q$ are well typed in the
+contexts $x : A, \ContextD$ and $x : \dual{A}, \ContextE$, which are obtained
+from $\ContextD$ and $\ContextE$ by adding the bindings $x : A$ and $x :
+\dual{A}$, respectively. Since $x$ is the most recently introduced channel, the
+types $A$ and $\dual{A}$ are added \emph{in front} of $\ContextD$ and
+$\ContextE$, which we do by means of the functions
+$(A~\AgdaInductiveConstructor{∷\_})$ and
+$(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷\_})$. These are partial
 applications of the constructor $\AgdaInductiveConstructor{\_∷\_}$ for lists to
-which we have supplied the left operand).
+which we have supplied the left operand.
 
@@
 
-For example, the constructor \AgdaInductiveConstructor{fail}, which corresponds
-to a process of the form $\Fail\x$, requires the context $\Context$ to satisfy
-the predicate
+For example, the constructor \AgdaInductiveConstructor{fail}, which builds a
+process $\Fail\x$, requires the context $\Context$ to satisfy the predicate
 $\AgdaDatatype{Ch}~\AgdaInductiveConstructor{⊤}~\AgdaDatatype{∗}~\AgdaFunction{U}$,
@@
 
-The constructor \AgdaInductiveConstructor{wait} corresponds to a process of the
-form $\Wait\x.P$ and requires $\Context$ to (separately) satisfy
+The constructor \AgdaInductiveConstructor{wait}, which builds a process
+$\Wait\x.P$, requires $\Context$ to (separately) satisfy
 $\AgdaDatatype{Ch}~\AgdaInductiveConstructor{⊥}$, that is the channel $x$ on
@@
 
-The constructor \AgdaInductiveConstructor{close}, which corresponds to a process
-of the form $\Close\x$, requires the typing context to be the singleton list
-$[\One]$.
+The constructor \AgdaInductiveConstructor{close}, which builds a process
+$\Close\x$, requires the typing context to be the singleton list $[\One]$.
 
-Let us move on to the processes that produce a continuation channel. For
-example, the constructor \AgdaInductiveConstructor{case} corresponds to a
-process of the form $\Case\x\y{P}{Q}$, where both $P$ and $Q$ use the
-continuation channel $y$. In this case $\Context$ must satisfy the predicate
+Let us move on to the forms that produce continuation channels. As an example,
+the constructor \AgdaInductiveConstructor{case} builds a process
+$\Case\x\y{P}{Q}$, where both $P$ and $Q$ use the continuation channel $y$. In
+this case $\Context$ must satisfy the predicate
 \[
@@
 the requirement that the type of $x$ must be of the form $A \With B$ and should
-be found in $\Context$. In other words, it must be the case that
-$\Splitting\Context{[A \With B]}\ContextD$ for some $\ContextD$. Now $P$ and $Q$
-must be well typed in the context $\ContextD$ augmented with the association $y
-: A$ and $y : B$, respectively, whence the use of \AgdaFunction{∩} to verify a
-(non-separating) conjunction of the predicates
+be found in $\Context$. In other words, $\Splitting\Context{[A \With
+B]}\ContextD$ for some $\ContextD$. Now $P$ and $Q$ must be well typed in the
+context $\ContextD$ augmented with the association $y : A$ and $y : B$,
+respectively, whence the use of \AgdaFunction{∩} to verify a (non-separating)
+conjunction of the predicates
 $(A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}$ and
@@
 The interpretation of the remaining constructors is analogous, so we only
-comment \AgdaInductiveConstructor{all}, which represents a polymorphic process
+comment \AgdaInductiveConstructor{all}, which builds a process
 $\Forall\x\X\y.P$. This constructor models the continuation $P$ using
-higher-order abstract syntax: witnessing that a context $\Context$ satisfies the
+higher-order abstract syntax: proving that a context $\Context$ satisfies the
 predicate
@@
 applied to $\Context$. Note that the side condition $X \not\in \fv\Context$ of
-\ForallRule is enforced \emph{by definition}:
+\ForallRule is trivially enforced \emph{by definition}:
 %

### 2025-12-26 paper/Type.lagda.tex
global revision

@@
 
-It is straightforward to prove that duality is a an involution.
+It is straightforward to prove that duality is an involution.
 
@@
 is effective provided that the option \texttt{{-}{-}rewriting} is enabled,
-either globally when invoking Agda or within a pragma in the module's source
-code.}
+either globally when invoking Agda or within an \AgdaPragma{OPTIONS} pragma
+directive in the module's source code.}
 
@@
 the type variables of a pre-type with other pre-types. In practice we will
-always substitute one variable at a time, but it is technically easier to
-substitute \emph{all} variables of a pre-type, possibly with themselves. The
-definition of \AgdaFunction{subst} relies on some auxiliary functions for
-\emph{renaming} type variables and \emph{lifting} substitutions across
-quantifiers. These functions are straightforward adaptations of those described
-by \citet{KokkeSiekWadler20}.
+always substitute one variable at a time, but it is technically easier to define
+\AgdaFunction{subst} so that it accepts a function substituting \emph{all}
+variables of a pre-type, possibly with themselves. The definition of
+\AgdaFunction{subst} relies on some auxiliary functions for \emph{renaming} type
+variables and \emph{lifting} substitutions across quantifiers. These functions
+are straightforward adaptations of those described by \citet{KokkeSiekWadler20}.
 

### 2025-12-26 paper/Congruence.lagda.tex
revision

@@
 $B~\AgdaInductiveConstructor{∷}~A~\AgdaInductiveConstructor{∷}~\ContextD$, where
-the type of the continuation $z$ comes before the type of the restricted channel
-$x$.
+the type $B$ of the continuation $z$ comes \emph{before} the type $A$ of the
+restricted channel $x$.
 %
-After structural precongruence is applied, however, the type $B$ of the
+After structural pre-congruence is applied, however, the type of the
 continuation channel $z$ ends up behind that of the restricted channel $x$,
-because now the two channels are introduced in the opposite order. Therefore, we
-need to rename the channels in $P$ so that it is well typed in the context
+because now $z$ and $x$ are introduced in the opposite order. Therefore, we need
+to rename the channels in $P$ so that it is well typed in the context
 $A~\AgdaInductiveConstructor{∷}~B~\AgdaInductiveConstructor{∷}~\ContextD$. Such
@@
 the form $\Cut[A]\x{\Fork\y\u\v{P}{Q}}{R}$ when $x \in \fn{P}$. We capture this
-condition by means of the splitting $\AgdaInductiveConstructor{<}~r$, meaning
+condition by means of the splitting $\AgdaInductiveConstructor{<}~r$, implying
 that the type $A$ of $x$ ends up in the typing context for $P$ and not in the
@@
 The symmetric rule \SForkR is modeled by another constructor
-\AgdaInductiveConstructor{s-fork-r}, not shown here but similar to
+\AgdaInductiveConstructor{s-fork-r}, which is similar to
 \AgdaInductiveConstructor{s-fork-l} except that $\AgdaInductiveConstructor{<}~r$

### 2025-12-26 paper/DeadlockFreedom.lagda.tex
revision

@@
 
-As we have seen in \Cref{sec:calculus}, the deadlock freedom result
-(\Cref{thm:df}) rests on some notions and predicates about processes which we
-now formalise in Agda. First of all we need to define the notion of
-\emph{thread}, that is any process other than a cut. It is convenient to provide
-a more fine-grained classification of threads, distinguishing between links,
-input/output actions, and sometimes also whether such actions operate on free or
-bound channels. We define predicates for each of these process classes:
+As we have seen in \Cref{sec:calculus}, the deadlock freedom property and
+\Cref{thm:df} rest on some notions and predicates about processes which must be
+formalised in Agda. First of all we need to define the notion of \emph{thread},
+that is any process other than a cut. It is convenient to provide a more
+fine-grained classification of threads, distinguishing between links and
+input/output actions and sometimes also on whether such actions operate on free
+or bound channels. We define predicates for each of these classes:
 
@@
 
-These predicates are relatively uninteresting since they are essentially
-isomorphic to the relevant fragments of the \AgdaDatatype{Proc} datatype. The
-only aspect that is worth pointing out here is that in \AgdaDatatype{Input},
-\AgdaDatatype{Output} and \AgdaDatatype{Server} the channel being acted upon by
-the thread is the \emph{first} in $\Context$, hence it is the \emph{most
-recently introduced} channel, whereas in \AgdaDatatype{Delayed} and
-\AgdaDatatype{DelayedServer} the channel is older. This allows us to distinguish
-those threads that, in the context of a cut, operate on the channel bound by the
-cut or on a free channel. To clarify this aspect, consider the
-\AgdaInductiveConstructor{wait} constructor in \AgdaDatatype{Input} and in
+The implementation of these predicates is not interesting since it is
+essentially isomorphic to the relevant fragments of the \AgdaDatatype{Proc}
+datatype. The only aspect that is worth pointing out here is that in
+\AgdaDatatype{Input}, \AgdaDatatype{Output} and \AgdaDatatype{Server} the
+channel being acted upon by the thread is the \emph{first} in $\Context$, hence
+it is the \emph{most recently introduced} channel, whereas in
+\AgdaDatatype{Delayed} and \AgdaDatatype{DelayedServer} the channel is not the
+first. This allows us to distinguish those threads that, in the context of a
+cut, operate on the channel bound by the cut or on a free channel. To clarify
+this aspect, we show the implementation of the constructor
+\AgdaInductiveConstructor{wait} in \AgdaDatatype{Input} and in
 \AgdaDatatype{Delayed}. In the former predicate we have
@@
 \end{code}
-where the use of the \AgdaInductiveConstructor{<} constructor indicates that the
+where the use of the constructor \AgdaInductiveConstructor{<} indicates that the
 thread operates on the most recent channel. In the latter predicate we have
@@
 \end{code}
-where the use of the \AgdaInductiveConstructor{>} constructor indicates that the
+where the use of the constructor \AgdaInductiveConstructor{>} indicates that the
 thread operates on a channel other than the most recent one. Note that here we
@@
 
-Now the \AgdaDatatype{Thread} predicate is simply the disjoint union of all the
+The predicate \AgdaDatatype{Thread} is simply the disjoint union of all the
 previous ones.
@@
 
-We define observability, reducibility and aliveness in the expected way:
+Observability, reducibility and aliveness are defined in the expected way:
 
@@
 In order to prove that every (well-typed) process is alive, it is convenient to
-also define a ``canonical'' form for cuts, that is a form that matches at least
-one of the left-hand-sides of the rules for structural pre-congruence or
-reduction (\Cref{tab:semantics}). This is the notion where the fine-grained
-classification of threads introduced earlier comes into play.
+define a ``canonical'' form for cuts, that is a form that matches at least one
+of the l.h.s of one of the rules for structural pre-congruence or reduction in
+\Cref{tab:semantics}. This is the notion where the fine-grained classification
+of threads introduced earlier comes into play.
 

### 2025-12-26 paper/Reduction.lagda.tex
revision

--- 2025-12-25 paper/Reduction.lagda.tex
@@
 
-Just like structural precongruence, reduction is formalised as a binary relation
-between processes that are well typed in the very same typing context. Thus, the
-definition of reduction embeds the subject reduction result.
+Just like structural pre-congruence, reduction is formalised as a binary
+relation between processes that are well typed in the same typing context. Thus,
+the definition of reduction embeds subject reduction (\Cref{thm:red}).
 
@@
 sub-process of the cut (that is the link $\Link\x\y$) and the splitting
-$(\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{>}~\AgdaInductiveConstructor{•})$
+$\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{>}~\AgdaInductiveConstructor{•}$
 to which \AgdaInductiveConstructor{link} is applied is structured consistently
 with the syntax of the link $\Link\x\y$ being reduced, which is oriented so that
-the restricted channel $x$ occurs on the left. 
+the restricted channel $x$ is on the left. 
 %
 The process $P$ has type
-$\AgdaDatatype{Proc}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\ContextD)$,
+$\AgdaDatatype{Proc}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\ContextD)$
 and turns into $P\subst\y\x$ after the reduction. The type
@@
 
-While the \AgdaInductiveConstructor{close} constructor implicitly refers to the
-only free channel occurring in a process of the form $\Close\x$, the
-\AgdaInductiveConstructor{wait} constructor uses a splitting proof of the form
-$(\AgdaInductiveConstructor{<}~q_0)$ to be sure that the referenced channel is
+While the constructor \AgdaInductiveConstructor{close} implicitly refers to the
+only free channel occurring in a process of the form $\Close\x$, the constructor
+\AgdaInductiveConstructor{wait} uses a splitting proof of the form
+$\AgdaInductiveConstructor{<}~q_0$ to make sure that the referenced channel is
 also the restricted one, and therefore matches the one of the

### 2025-12-26 paper/Safety.lagda.tex
global revision

--- 2025-12-24 paper/Safety.lagda.tex
@@
 
-Note the presence of two constructors \AgdaInductiveConstructor{cut-l} and
-\AgdaInductiveConstructor{cut-r} which distinguish on whether the hole is found
-in the left (respectively, right) sub-context, as per the grammar of reduction
-contexts in \Cref{sec:properties}.
+The constructor \AgdaInductiveConstructor{hole} builds a hole, as the name
+implies.
+%
+The constructors \AgdaInductiveConstructor{cut-l} and
+\AgdaInductiveConstructor{cut-r} build reduction contexts where the hole is
+found in the left (respectively, right) sub-term of a cut, as per the grammar of
+reduction contexts given in \Cref{sec:properties}.
 
@@
 \end{code}
+
+This notion of process substitution preserves typing by construction thanks to
+the fact that both processes and reduction contexts are intrinsically typed.
 

### 2025-12-26 paper/Examples.lagda.tex
revision

@@
 open import Data.Sum using (inj₁; inj₂)
-open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
+open import Data.Product using (_×_; _,_; ∃; ∃-syntax; curry)
 open import Data.Fin using (zero; suc; #_)
@@
 
-In this section we revisit the processes discussed in \Cref{ex:booleans,ex:echo}
-and show their encoding in our formalisation. The encoding of $\Bool$ is
-straightforward.
+In this section we revisit and expand the processes discussed in
+\Cref{ex:booleans,ex:echo} and show their encoding in our formalisation. The
+encoding of $\Bool$ is straightforward
 
@@
 \end{code}
-
-The boolean constants are encoded thus:
+and the boolean constants are encoded thus:
 \begin{code}
@@
 
-We can take advantage of the host language for programming higher-order
-processes. For example, we can define a conditional process thus:
+We take advantage of the host language for programming higher-order processes.
+For example, we can define a conditional process thus:
 
@@
 The function \AgdaFunction{\$} (defined in Agda's standard library) is just a
-visible function application operator. We use it here as ``separator'' to save a
-few parentheses.
+visible function application operator. We use it as a ``separator'' so that we
+can flatten deeply nested expressions and save a bunch of parentheses.
 %
-Note the use of \AgdaInductiveConstructor{cut}s for combining a process that
-produces a boolean value with a process that consumes a boolean value and the
-use of \AgdaInductiveConstructor{link} for copying a boolean value from one
-channel to another. 
-%
-Here we have chosen to define the disjunction \AgdaFunction{||} from the
-conjunction \AgdaFunction{\&\&} and negation \AgdaFunction{!!} using De Morgan's
-laws.
+We have chosen to define the disjunction \AgdaFunction{||} from the conjunction
+\AgdaFunction{\&\&} and negation \AgdaFunction{!!} using De Morgan's laws.
 
@@
 
-For the encoding of the echo server (\Cref{ex:echo}), recall that the server is
-polymorphic in the sense that the type $X$ of the value it receives is
-arbitrary. The notation $\AgdaFunction{\#}~n$ is defined in Agda's standard
-library and creates an element of \AgdaDatatype{Fin} from the natural number
-$n$.
+For the encoding of the polymorphic echo server (\Cref{ex:echo}), we start by
+encoding its type $\OfCourse(\forall\X.\dual\X \Par (X \Ten \One))$:
 
 \begin{code}
-Echo : let X = var (# 0) in
-       Proc [ `! (`∀ (X ⅋ (dual X ⊗ 𝟙))) ]
-Echo = curry∗ server ch (< ≫)
-             ( un-[]
-             , curry∗ all ch (< ≫) λ X →
+EchoServerT : Type
+EchoServerT = `! (`∀ (rav (# 0) ⅋ (var (# 0) ⊗ 𝟙)))
+\end{code}
+
+The notation $\AgdaFunction{\#}~n$ is defined in Agda's standard library and
+creates an element of \AgdaDatatype{Fin} from the natural number $n$. In
+\AgdaFunction{EchoServer}, we use it to represent the de Bruijn index of the
+type variable $X$.
+%
+We now encode the server
+\begin{code}
+Server : Proc [ EchoServerT ]
+Server = curry (curry∗ server ch (< ≫)) un-[] $
+         curry∗ all ch (< ≫) λ X →
          curry∗ join ch (< ≫) $
          curry∗ (curry∗ fork ch (< ≫)) (curry∗ link ch (< > •) ch) (< ≫) $
-               close ch )
+         close ch
 \end{code}
-
-We can test the server by connecting it with a \AgdaInductiveConstructor{client}
-that sends true to it.
-
+and the client that sends true to it
 \begin{code}
-Echo-True : Proc [ 𝔹 ]
-Echo-True = curry∗ cut Echo ≫ $
-            curry∗ client ch (< ≫) $
-            curry∗ (ex {_} {dual 𝔹}) ch (< ≫) $
+Client : Proc (dual EchoServerT ∷ 𝔹 ∷ [])
+Client = curry∗ client ch (< ≫) $
+         curry∗ (ex {_} {𝔹}) ch (< ≫) $
          curry∗ (curry∗ fork ch (< ≫)) True ≫ $
@@
 
-If we ask Agda to evaluate this process we obtain \AgdaFunction{True}, as
-expected.
-
-From these examples it should be clear that writing \Calculus processes in Agda
-is not entirely straightforward and requires a little effort. The lack of names
-(for channels) clearly obfuscates the code, but the context splittings that must
-be provided are usually small and fairly easy to understand. Most importantly,
-the use of \AgdaFunction{≫} helps truncating context splittings as soon as the
-seeked channel has been reached.
-%
-Either way, the formalisation is meant to study the meta-theory of \Calculus
-rather than implementing a user-friendly DSL for it.
+To test these definitions, we define a main process that connects server and
+client
+\begin{code}
+Main : Proc [ 𝔹 ]
+Main = curry∗ cut Client (< •) Server
+\end{code}
+and then ask Agda to normalize \AgdaFunction{Main}, which yields
+\AgdaFunction{True} as expected.

### 2025-12-26 paper/conclusion.tex
global revision

@@
 
-We have presented a formalisation of \LCC, a variant of the linear
-$\pi$-calculus~\cite{KobayashiPierceTurner99} supported by the same type system
-of \CP~\cite{Wadler14}. Binary sessions are modeled in \LCC using the
-continuation-passing encoding described by \citet[extended
-version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}.
+We have presented a formalisation of \LCC, a linear calculus of continuations
+closely related to the linear $\pi$-calculus~\cite{KobayashiPierceTurner99} and
+supported by the same type system of \CP~\cite{Wadler14}. Binary sessions can be
+modeled in \LCC using the continuation-passing encoding described by
+\citet[extended version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}.
 %
 Under the assumption that the size of a formalisation is a good measure of its
-complexity, the formalisation of \LCC can be qualified as ``simple'' compared to
-the existing ones.
+complexity, the formalisation of \LCC qualifies as the simplest compared to the
+existing ones. This result should also be weighed considering the richness of
+\LCC in terms of supported features and provable properties.
+
+Our work casts some doubts on the actual contribution of context splitting to
+the complexity of a formalisation and hints instead at explicit continuations in
+the underlying model as a simplifying factor.
 %
-In this respect, our work casts some doubts on the actual contribution of
-context splitting to the complexity of a formalisation and hints instead at
-explicit continuations in the underlying model as a simplifying factor.
-%
-It is quite striking that, among the approaches that have been proposed to
-overcome context splitting, the one of \citet{SanoKavanaghPientka23} makes use
-of explicit continuations.
+It is quite remarkable that, among the alternative approaches that have been
+proposed to avoid context splitting, the one of \citet{SanoKavanaghPientka23}
+based on linearity predicates makes use of explicit continuations.
 
-We remark that \LCC and the calculus described in the linearity
+The calculus of linear continuations and the calculus described in the linearity
 challenge~\cite{CarboneEtAl24} are incomparable in terms of expressiveness. On
@@
 %
-Because of its logical foundations, we like to think of \LCC as a ``canonical''
-calculus that deserves its own space in the context of the linearity challenge
-alongside with (but not in substitution of) more traditional session calculi.
+Because of its logical foundations, we think that \LCC deserves its own space in
+the context of the linearity challenge alongside with (but not in substitution
+of) more traditional session calculi.
 
-The simple formalisation of \LCC is a good basis for further investigations. We
-have already extended \LCC with support for infinite types and recursive
-processes (the source code of this extension is publicly available in the same
-repository as the one that contains the formalisation of \LCC). In the future,
-it would be interesting to formalise the strong normalisation property of \LCC
-as a consequence of cut elimination of classical linear logic.
+The simple formalisation of \LCC is a good starting point for further
+developments. We have already extended \LCC with support for infinite types and
+recursive processes (the source code of this extension is publicly available in
+the same repository as the one that contains the formalisation of \LCC). In the
+future, it would be interesting to formalise the strong normalisation property
+of \LCC as a consequence of cut elimination of classical linear logic.
 
@@
 smaller. Whether this is a coincidence or further evidence of the effectiveness
-of the continuation-based approaches is left for future work.
+of the continuation-based approaches is left as future work.

### 2025-12-26 paper/related.tex
global revision

@@
 
-Motivated by the technical difficulties arising from context splits,
+Motivated by the technical difficulties arising from context splittings,
 \citet{SanoKavanaghPientka23} define a structural version of \CP using an
-alternative approach based on \emph{linearity predicates}. The key idea is to
-treat typing contexts structurally and to enforce the linear usage of channels
-by checking their syntactical occurrence in processes. Interestingly, the
-approach relies on the \emph{explicit naming of continuations} so as to
-precisely account for the number of times a channel is actually used.
-\citet{SanoKavanaghPientka23} do not connect their technique with the
-continuation-passing encoding of binary sessions, but \emph{a posteriori} the
-analogies between the two approaches are evident.
+approach based on \emph{linearity predicates}. The key idea is to treat typing
+contexts structurally and to enforce the linear usage of channels by checking
+their syntactical occurrence in processes. Interestingly, this approach relies
+on the \emph{explicit naming of continuations} so as to precisely account for
+the number of times a channel is actually used. \citet{SanoKavanaghPientka23} do
+not connect their technique with the continuation-passing encoding of binary
+sessions, but the analogies are evident even though the role of continuations
+differs.
 
-\citet{ZackonEtAl25} describe yet another approach to the handling of typing
-contexts whereby channels are associated not just with a type but also with a
-\emph{tag} (or \emph{multiplicity}) from a given resource algebra. The tag
-summarises the number of allowed usages of a particular channel, including the
-possibility that the channel is not available. This approach streamlines context
-splitting since contexts can be treated in an essentially structural way, except
-for tags which are conveninently combined using operations from the resource
-algebra.
+\citet{ZackonEtAl25} describe yet another approach for managing typing contexts
+whereby channels are associated not just with a type but also with a \emph{tag}
+(or \emph{multiplicity}) from a given resource algebra. The tag summarises the
+number of allowed usages of a particular channel, including the possibility that
+the channel is not available. This approach streamlines context splitting since
+contexts can be treated in an essentially structural way, except for tags which
+are conveninently combined using operations from the resource algebra.
 
@@
 \Cref{tab:comparison} reports a summary of the aforementioned formalisations and
-also our own. The three leftmost colums list the reference papers (sorted by
-publication date), the proof assistant and the formalised calculus.
+also our own. The three leftmost columns list the reference papers (sorted by
+publication date), the used proof assistant and the formalised calculus.
 %
 For simplicity, we have grouped the calculi into classes identified by an
-acronym. Some acronyms, like \CP~\cite{Wadler14} and
-\GV~\cite{GayVasconcelos10,Wadler14} are well known in the literature on session
-types. It should be noted that the actual calculus being formalised usually
-differs from (typically, is a strict subset of) the one identified by the
-acronym, but it retains at least some of its key features. We use \SessionPI to
-denote (variants of) the session-based $\pi$-calculus presented by
-\citet{HondaVasconcelosKubo98} and \LinearPI to denote (variants of) the linear
-$\pi$-calculus~\cite{KobayashiPierceTurner99}, into which binary sessions can be
-encoded using explicit continuation channels as described by \citet[extended
-version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}. Finally, we use
-\SCP for the structural version of \CP introduced by
-\citet{SanoKavanaghPientka23} and \Calculus for our calculus, a linear version
-of \CP with explicit continuations.
+acronym. The acronyms \CP~\cite{Wadler14} and
+\GV~\cite{GayVasconcelos10,Wadler14} refer to well-known calculi in the
+literature on session types. It should be noted that the actual calculus being
+formalised usually differs from (typically, is a strict subset of) the one
+identified by the acronym, but it retains at least some of its key features.
+%
+The acronym \SessionPI refers to (variants of) the session-based $\pi$-calculus
+presented by \citet{HondaVasconcelosKubo98} while \LinearPI refers to (variants
+of) the linear $\pi$-calculus~\cite{KobayashiPierceTurner99}, into which binary
+sessions can be encoded using explicit continuation channels as described by
+\citet[extended version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}.
+Finally, \SCP is the structural version of \CP introduced by
+\citet{SanoKavanaghPientka23} and \Calculus is our calculus.
 
@@
 for our own, they all differ from the one stated in the linearity
-challenge~\citep{CarboneEtAl24} when available.
+challenge~\citep{CarboneEtAl24}.
 %
-Some formalisation define \emph{libraries} which could be reused in different
+Some formalisations define \emph{libraries} which could be reused in different
 contexts. In these cases, the size of the library is reported separately next to
@@
 the largest by \citet{GotoEtAl16} (543kb) to our own (21kb, without the proof of
-deadlock freedom) which is also the only one supporting all the featurs of \CP.
+deadlock freedom) which is also the only one supporting all the features of \CP.
 
@@
 datatypes to be defined, but the definitions are also more involved because they
-incorporate invariants and the proofs of typing preservation. In the end, it
-seems plausible to speculate that the effort for representing processes, types
-and typing rules is not substantially different overall.
+incorporate invariants and the proofs of typing preservation. It seems plausible
+to speculate that the effort for representing processes, types and typing rules
+is not substantially different overall.
 

### 2025-12-26 paper/Examples.lagda.tex
fixed process names

@@
 \begin{code}
-EchoServerT : Type
-EchoServerT = `! (`∀ (rav (# 0) ⅋ (var (# 0) ⊗ 𝟙)))
+ServerT : Type
+ServerT = `! (`∀ (rav (# 0) ⅋ (var (# 0) ⊗ 𝟙)))
 \end{code}
@@
 The notation $\AgdaFunction{\#}~n$ is defined in Agda's standard library and
-creates an element of \AgdaDatatype{Fin} from the natural number $n$. In
-\AgdaFunction{EchoServer}, we use it to represent the de Bruijn index of the
-type variable $X$.
+creates an element of \AgdaDatatype{Fin} from the natural number $n$.
 %
@@
 \begin{code}
-Server : Proc [ EchoServerT ]
+Server : Proc [ ServerT ]
 Server = curry (curry∗ server ch (< ≫)) un-[] $
@@
 \begin{code}
-Client : Proc (dual EchoServerT ∷ 𝔹 ∷ [])
+Client : Proc (dual ServerT ∷ 𝔹 ∷ [])
 Client = curry∗ client ch (< ≫) $
@@
 
-To test these definitions, we define a main process that connects server and
-client
+We define a main process that connects server and client
 \begin{code}

### 2025-12-27 paper/abstract.tex
tuning

@@
 \abstract{%
-    The formalisation of session calculi is considered challenging due to the
-    management of channels, which are linear resources that cannot be discarded
-    or duplicated at will and whose type changes over time, as input/output
+    The formalisation of session calculi is considered challenging because of
+    the management of channels, which are linear resources that cannot be
+    discarded or duplicated and whose type changes over time, as input/output
     operations are performed on them.
@@
     Context splitting, the ``naive'' channel management technique directly
-    related to the way session type theories are usually sketched with pen and
+    related to the way session type theories are usually written with pen and
     paper, is often considered a hindrance and a notable source of complexity,

### 2025-12-27 paper/related.tex
tuning

@@
 %
-With leftover typing a process is typed with respect to an \emph{input context},
-which describes the overall available channels, and an \emph{output context},
-which describes the residual channels not consumed by the process. In this way,
-it is possible to ``concatenate'' typing judgments instead of splitting typing
-contexts beforehand. As \citet{ZalakainDardha21} nicely summarise, ``context
+With leftover typing the judgments have the form $\wtl{P}\ContextC\ContextD$
+where a process $P$ is typed with respect to an input context $\ContextC$, which
+describes the overall available channels, and a context of leftovers
+$\ContextD$, which describes the residual channels not consumed by the process.
+In this way, it is possible to ``concatenate'' typing judgments by matching the
+leftovers in one judgment with the input context of the subsequent one, with no
+need for splitting. As \citet{ZalakainDardha21} nicely summarise, ``context
 splits contain usage information that is already present in processes,'' and
@@
 not connect their technique with the continuation-passing encoding of binary
-sessions, but the analogies are evident even though the role of continuations
-differs.
+sessions~\cite{Kobayashi02b,DardhaGiachinoSangiorgi17}, but the analogies are
+evident even though the role of continuations differs.
 
@@
 %
-Some formalisations define \emph{libraries} which could be reused in different
-contexts. In these cases, the size of the library is reported separately next to
-the size of the part of the development that uses it.
+Some
+formalisations~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,JacobsBalzerKrebbers22,ZackonEtAl25}
+define \emph{libraries} which can be reused in different contexts. In these
+cases, the size of the library is reported separately next to the size of the
+part of the development that uses it.
 
@@
 processes reduces the size of the formalisation. As observed in
-\Cref{sec:process-agda}, this choice may help reducing the overall number of
-datatypes to be defined, but the definitions are also more involved because they
-incorporate invariants and the proofs of typing preservation. It seems plausible
-to speculate that the effort for representing processes, types and typing rules
-is not substantially different overall.
+\Cref{sec:process-agda}, this choice helps reducing the overall number of
+datatypes to be defined and makes some results trivial (\eg
+\Cref{thm:type-safety} formalised by \AgdaFunction{type-safety}), but the
+definitions are also more involved because they incorporate invariants and the
+proofs of typing preservation. It is plausible to speculate that the effort for
+representing processes, types and typing rules is not substantially different
+overall.
 
@@
 smallest formalisations (excluding the deadlock freedom results) -- with an
-average size of around 62kb -- are all based on continuations, whether they are
-explicit (\LinearPI, \SCP, \LCC) or ``virtual'' (\GV), while the remaining ones
-are 263kb on average. This observation is in agreement with the considerations
-we made earlier in this paper concerning the difference between \emph{updating}
-bindings versus \emph{removing and adding} bindings in a typing context. The
-adoption of continuations enables a clean linear management of channels and
-avoids the need for strong updates on the typing context.
+average size of around 62kb -- are all based on continuations, no matter if they
+are explicit (\LinearPI, \SCP, \LCC) or ``virtual'' (\GV), while the remaining
+ones are 263kb on average. This observation is in agreement with the
+considerations we made earlier in this paper concerning the difference between
+\emph{updating} bindings versus \emph{removing and adding} bindings in a typing
+context. The use of continuations enables a cleaner management of typing
+contexts since channels become true ``use-once'' resources and there is no need
+for strong updates.
 

### 2025-12-27 paper/related.tex
more related

@@
 
-\citet{Thiemann19} formalises a subset of \GV~\citep{GayVasconcelos10,Wadler14},
-a functional language extended with session communication primitives, along with
+\citet{Thiemann19} formalises a subset of \GV~\citep{GayVasconcelos10}, a
+functional language extended with session communication primitives, along with
 an interpreter. \citet{CicconePadovani20} have taken inspiration from this work
@@
 \begin{table}[t]
+    \newcommand\T{\hspace{1.25ex}}
     \newcommand{\textlabel}[1]{\rotatebox{90}{#1}}
-    \caption{\label{tab:comparison} A summary of different formalisations of binary session calculi. All sizes are in kb.}
-    \begin{minipage}{.9\textwidth}
-    \begin{tabular}{@{}r@{~}@{~}c@{~~}c@{~}|@{~}c@{~~}c@{~~}c@{~~}c@{~}|@{~}r@{~~}r@{~~}r@{~~}r@{~~}r@{~}l@{}}
+    \caption{\label{tab:comparison} A summary of different formalisations of
+    binary session calculi (sizes in kb).}
+    \begin{minipage}{.95\textwidth}
+    \begin{tabular}{@{}r@{\T}c@{\T}c@{\T}c@{\T}|@{\T}c@{\T}c@{\T}c@{\T}c@{\T}|@{\T}r@{\T}r@{\T}r@{\T}r@{\T}r@{\,}l@{}}
         \textbf{Formalisation} &
@@
         \textlabel{Calculus} &
+        \textlabel{Cut} &
         \textlabel{Channel linearity} &
@@
         \SessionPI &
+        \NO &
         splits &
@@
         \GV &
+        \NO &
         splits &
@@
         \LinearPI & 
+        \NO &
         splits &
@@
         \LinearPI &
+        \NO &
         leftovers &
@@
         \textbf{90} &
-        \footnote{Includes shared channels and the library for algebra of types.}
+        \footnote{Includes shared channels and the library for the algebra of types.}
         \\
@@
         \GV &
+        \YES &
         splits &
@@
         \SCP &
+        \YES &
         predicates &
@@
         \CP &
+        \YES &
         tags &
@@
         \Calculus &
+        \YES &
         splits &
@@
 \Cref{tab:comparison} reports a summary of the aforementioned formalisations and
-also our own. The three leftmost columns list the reference papers (sorted by
-publication date), the used proof assistant and the formalised calculus.
+also our own. The four leftmost columns list the reference papers (sorted by
+publication date), the used proof assistant, the formalised calculus, and
+whether the calculus features a cut in the style of linear logic, as opposed to
+separate restriction and parallel composition.
 %
@@
 \GV~\cite{GayVasconcelos10,Wadler14} refer to well-known calculi in the
-literature on session types. It should be noted that the actual calculus being
-formalised usually differs from (typically, is a strict subset of) the one
-identified by the acronym, but it retains at least some of its key features.
+literature on session types. The acronym \SessionPI refers to (variants of) the
+session-based $\pi$-calculus presented by \citet{HondaVasconcelosKubo98} while
+\LinearPI refers to (variants of) the linear
+$\pi$-calculus~\cite{KobayashiPierceTurner99}. Finally, \SCP is the structural
+version of \CP introduced by \citet{SanoKavanaghPientka23} and \Calculus is our
+calculus.
 %
-The acronym \SessionPI refers to (variants of) the session-based $\pi$-calculus
-presented by \citet{HondaVasconcelosKubo98} while \LinearPI refers to (variants
-of) the linear $\pi$-calculus~\cite{KobayashiPierceTurner99}, into which binary
-sessions can be encoded using explicit continuation channels as described by
-\citet[extended version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}.
-Finally, \SCP is the structural version of \CP introduced by
-\citet{SanoKavanaghPientka23} and \Calculus is our calculus.
+It should be noted that the actual calculus being formalised usually differs
+from (typically, is a strict subset of) the one identified by the acronym and
+that the same acronym may identify different calculi. For example, the first
+version of \GV described by \citet{GayVasconcelos10} is not based on linear
+logic, whereas later versions~\cite{Wadler14} used \eg by
+\citet{JacobsBalzerKrebbers22} are.
+%
+In any case, the acronym is meant to give an idea of the overall flavour of the
+calculus.
 
 The four middle columns of \Cref{tab:comparison} report the relevant qualitative
-aspects of the formalisations, namely the technique used to enforce the linear
-usage of channels, the technique used to encode channels, whether the
-formalisation is intrinsically or extrinsically typed, and whether the calculus
-makes use of continuation channels. Concerning the use of continuations, the
-approaches based directly on the linear $\pi$-calculus (into which sessions can
-be encoded) are marked with {\YES} and those based on a calculus with native
-sessions are marked with {\NO}. The calculus \SCP is marked with {\MAYBE}
-because, while not being strictly based on the linear $\pi$-calculus, it makes
-use of explicit continuations for defining the predicates that check the linear
-usage of channels. Finally, all the approaches based on \GV are also marked with
-\MAYBE. In principle, \emph{none} of these calculi makes use of continuations.
-However, \GV is designed in such a way that each operation acting on a channel
-$s$ is a function that returns the result of the operation (if any) \emph{along
-with the same channel $s$}. In this way, the type of $s$ can be conveniently
-``updated'' to take into account the effect of the operation. As observed by
+aspects of the formalisations, namely the technique for the management of typing
+contexts, the representation of channels, whether the formalisation is
+intrinsically or extrinsically typed, and whether the calculus makes use of
+continuation channels. Concerning the use of continuations, the approaches based
+directly on the linear $\pi$-calculus (into which sessions can be encoded) are
+marked with {\YES} and those based on a calculus with native sessions are marked
+with {\NO}. The calculus \SCP is marked with {\MAYBE} because, while not
+directly inspired to the linear $\pi$-calculus, it makes use of explicit
+continuations for defining the predicates that check the linear usage of
+channels. Finally, all the approaches based on \GV are also marked with \MAYBE.
+In principle, \emph{none} of these calculi makes use of continuations. However,
+\GV is designed in such a way that each operation acting on a channel $s$ is a
+function that returns the result of the operation (if any) \emph{along with the
+same channel $s$}. In this way, the type of $s$ can be conveniently ``updated''
+to take into account the effect of the operation. As observed by
 \citet{Padovani17}, this semantics of the communication primitives is virtually
@@
 various approaches in addressing the linearity challenge because the
-formalisations differ widely for a variety of entangled factors including the
-features of calculus, the used proof assistant, the representation of
-processes/expressions/channels, the handling of linearity. Also, most
+formalisations differ widely for a variety of entangled factors. Also, most
 formalisations predate the publication of the linearity
@@
 
+Using the cut in the style of linear logic instead of separate restriction and
+parallel composition simplifies the representation of channels (or session
+endpoints). All the formalisations of calculi that adopt the cut tend to be
+small (if we exclude the libraries), but this is not a general rule.
+
 Finally, it seems that the use of (explicit) continuations affects the
@@
 context. The use of continuations enables a cleaner management of typing
-contexts since channels become true ``use-once'' resources and there is no need
-for strong updates.
+contexts since linear channels become true ``use-once'' resources and there is
+no need for strong updates of their type.
 

### 2025-12-27 paper/related.tex
revised related

@@
 $\pi$-calculus~\cite{KobayashiPierceTurner99} that supports dependent types, so
-as to enable the description of communication protocols whose structure depends
-on the content of messages.
+as to enable the description of communication protocols whose structure may
+depend on the content of messages.
 
@@
 \citet{RouvoetPoulsenKrebbersVisser20} present a library of abstractions
-inspired to separation logic aiding the formalisation of interpreters for models
-supported by linear type systems. One of the presented case studies is the
+inspired to separation logic aiding the formalisation of interpreters for
+languages with linear resources. One of the presented case studies is the
 formalisation of a fragment of \GV~\citep{GayVasconcelos10,Wadler14}. The
@@
 the constructors for the representation of processes in \Cref{sec:process-agda}
-have been heavily inspired by this work.
+have been adapted from this work.
 
@@
 %
-With leftover typing the judgments have the form $\wtl{P}\ContextC\ContextD$
-where a process $P$ is typed with respect to an input context $\ContextC$, which
-describes the overall available channels, and a context of leftovers
-$\ContextD$, which describes the residual channels not consumed by the process.
-In this way, it is possible to ``concatenate'' typing judgments by matching the
-leftovers in one judgment with the input context of the subsequent one, with no
-need for splitting. As \citet{ZalakainDardha21} nicely summarise, ``context
-splits contain usage information that is already present in processes,'' and
-therefore can be omitted. This is true provided that channels are named
-(\citet{ZalakainDardha21} use de Bruijn indices to this aim). In fact, the co-de
-Bruijn representation of processes~\cite{McBride18}, whereby channels are
-nameless and context splitting is performed eagerly, can be seen as the ``dual
-approach'' of leftover typing: channel names provide information that is already
-present in their (singleton) typing context, hence they can be omitted from
-contexts and processes.
+Typing judgments have the form $\wtl{P}\ContextC\ContextD$ so that a process $P$
+is typed with respect to an input context $\ContextC$, which describes all the
+available channels, and a context of leftovers $\ContextD$, which describes the
+residual channels not consumed by the process. In this way, it is possible to
+``concatenate'' typing judgments by matching the leftovers in one judgment with
+the input context of the subsequent one, with no need for splitting. As
+\citet{ZalakainDardha21} nicely summarise, context splittings are not necessary
+because they ``contain usage information that is already present in processes.''
+This is true provided that channels are named (\citet{ZalakainDardha21} use de
+Bruijn indices to this aim). In fact, the co-de Bruijn representation of
+processes~\cite{McBride18}, whereby channels are nameless and context splitting
+is performed eagerly, can be seen as the ``dual approach'' of leftover typing:
+channel names provide information that is already present in their (singleton)
+typing context, hence they can be omitted from contexts and processes.
 
@@
     \begin{minipage}{.95\textwidth}
-    \begin{tabular}{@{}r@{\T}c@{\T}c@{\T}c@{\T}|@{\T}c@{\T}c@{\T}c@{\T}c@{\T}|@{\T}r@{\T}r@{\T}r@{\T}r@{\T}r@{\,}l@{}}
-        \textbf{Formalisation} &
-        \textlabel{Proof assistant} &
+    \begin{tabular}{@{}l@{\T}c@{\T}c@{\T}c@{\T}c@{\T}|c@{\T}c@{\T}c@{\T}|@{\T}r@{\T}r@{\T}r@{\T}r@{\T}r@{\,}l@{}}
+        \text{Reference paper} &
+        \text{Prover} &
         \textlabel{Calculus} &
-        \textlabel{Cut} &
-        \textlabel{Channel linearity} &
-        \textlabel{Channel names} &
+        \textlabel{Cuts} &
+        \textlabel{Continuations} &
+        \text{Linearity} &
+        \text{Channels} &
         \textlabel{Intrinsically typed} &
-        \textlabel{Continuations} &
         % \textlabel{Shared channels} &
@@
         \NO &
+        \NO &
         splits &
         loc. nameless &
-        \NO &
         \NO &
@@
         \NO &
+        \MAYBE &
         splits &
@@
         \YES &
-        \MAYBE &
         % \YES &
@@
         \YES &
-        \MAYBE &
         % \NO &
@@
         \NO &
+        \YES &
         splits &
         co-de Bruijn &
-        \YES &
         \YES &
@@
         \NO &
+        \YES &
         leftovers &
@@
         \NO &
-        \YES &
         % \YES &
@@
         \YES &
+        \MAYBE &
         splits &
@@
         \NO &
-        \MAYBE &
         % \NO &
@@
         \YES &
+        \MAYBE &
         predicates &
@@
         \YES &
+        \NO &
         tags &
         HOAS &
-        \NO &
         \NO &
@@
         \\
-        this work &
+        this &
         Agda &
@@
         \YES &
+        \YES &
         splits &
         co-de Bruijn &
-        \YES &
         \YES &
@@
 
-\Cref{tab:comparison} reports a summary of the aforementioned formalisations and
-also our own. The four leftmost columns list the reference papers (sorted by
-publication date), the used proof assistant, the formalised calculus, and
-whether the calculus features a cut in the style of linear logic, as opposed to
-separate restriction and parallel composition.
+\Cref{tab:comparison} reports a summary of the aforementioned formalisations
+(sorted by publication date) and also our own. The first five columns identify
+the calculus being formalised specifying its reference paper, the used proof
+assistant and an acronym that gives an idea of the overall flavour of the
+calculus.
 %
-For simplicity, we have grouped the calculi into classes identified by an
-acronym. The acronyms \CP~\cite{Wadler14} and
-\GV~\cite{GayVasconcelos10,Wadler14} refer to well-known calculi in the
-literature on session types. The acronym \SessionPI refers to (variants of) the
-session-based $\pi$-calculus presented by \citet{HondaVasconcelosKubo98} while
-\LinearPI refers to (variants of) the linear
+The acronyms \CP~\cite{Wadler14,GayVasconcelos25} and
+\GV~\cite{GayVasconcelos10,Wadler14} are well-known in the literature on session
+types. The acronym \SessionPI refers to (variants of) the session-based
+$\pi$-calculus presented by \citet{HondaVasconcelosKubo98} while \LinearPI
+refers to (variants of) the linear
 $\pi$-calculus~\cite{KobayashiPierceTurner99}. Finally, \SCP is the structural
@@
 %
-It should be noted that the actual calculus being formalised usually differs
-from (typically, is a strict subset of) the one identified by the acronym and
-that the same acronym may identify different calculi. For example, the first
-version of \GV described by \citet{GayVasconcelos10} is not based on linear
-logic, whereas later versions~\cite{Wadler14} used \eg by
-\citet{JacobsBalzerKrebbers22} are.
+We emphasize that the actual calculus being formalised usually differs from
+(typically, is a strict subset of) the one identified by the acronym and that
+the same acronym may refer to different versions of the same calculus.
 %
-In any case, the acronym is meant to give an idea of the overall flavour of the
-calculus.
+We also indicate whether the calculus features \emph{cuts} (that is, the
+combination of restriction and parallel composition in correspondence with the
+cut of linear logic) and continuations.
 
-The four middle columns of \Cref{tab:comparison} report the relevant qualitative
-aspects of the formalisations, namely the technique for the management of typing
-contexts, the representation of channels, whether the formalisation is
-intrinsically or extrinsically typed, and whether the calculus makes use of
-continuation channels. Concerning the use of continuations, the approaches based
-directly on the linear $\pi$-calculus (into which sessions can be encoded) are
-marked with {\YES} and those based on a calculus with native sessions are marked
-with {\NO}. The calculus \SCP is marked with {\MAYBE} because, while not
-directly inspired to the linear $\pi$-calculus, it makes use of explicit
-continuations for defining the predicates that check the linear usage of
-channels. Finally, all the approaches based on \GV are also marked with \MAYBE.
-In principle, \emph{none} of these calculi makes use of continuations. However,
-\GV is designed in such a way that each operation acting on a channel $s$ is a
-function that returns the result of the operation (if any) \emph{along with the
-same channel $s$}. In this way, the type of $s$ can be conveniently ``updated''
-to take into account the effect of the operation. As observed by
-\citet{Padovani17}, this semantics of the communication primitives is virtually
-indistinguishable from one making use of explicit continuation channels.
+Concerning the use of continuations, the approaches based directly on the linear
+$\pi$-calculus (into which sessions can be encoded) are marked with {\YES} and
+those based on a calculus with native sessions are marked with {\NO}. The
+calculus \SCP is marked with {\MAYBE} because, while not directly inspired to
+the linear $\pi$-calculus, it makes use of explicit continuations for defining
+the predicates that check the linear usage of channels. Finally, all the
+approaches based on \GV are also marked with \MAYBE. In principle, \emph{none}
+of these calculi makes use of continuations. However, \GV is designed in such a
+way that each operation acting on a channel $s$ is a function that returns the
+result of the operation (if any) \emph{along with the same channel $s$}. In this
+way, the type of $s$ can be conveniently ``updated'' to take into account the
+effect of the operation. As observed by \citet{Padovani17}, this semantics of
+the communication primitives is virtually indistinguishable from one making use
+of explicit continuation channels.
 
-The rightmost columns of \Cref{tab:comparison} report the size (in kilobytes) of
-the formalisations as rough (and possibly questionable) estimates of their
-complexity. Instead of counting line numbers, which may be affected by code
-indentation styles and syntactical constraints of the proof assistant being
-used, we have counted the number of characters after comments have been removed
-and spaces have been squeezed.\footnote{Sequences of two or more consecutive
-space-like characters are collapsed into a single space. The squeezing is
-obtained by running the command \texttt{tr -s [:space:] file} on Unix-like
-systems.}
+The three middle columns of \Cref{tab:comparison} report the relevant
+qualitative aspects of the formalisations, namely the management of typing
+contexts, the representation of channels and whether the formalisation is
+intrinsically or extrinsically typed.
+
+The rightmost columns report the size (in kilobytes) of the formalisations as
+rough (and possibly questionable) estimates of their complexity. Papers
+describing formalisations typically use ``lines of code'' as a measure of
+development effort, but the number of lines may be affected by code indentation
+styles and syntactical constraints of the proof assistant being used. For this
+reason, we have preferred to count the total number of characters after comments
+have been removed and spaces have been squeezed.\footnote{Sequences of two or
+more consecutive space-like characters are collapsed into a single space. The
+squeezing is obtained by running the command \texttt{tr -s [:space:] file} on
+Unix-like systems.}
 %
-The reported sizes account for the source code of the formalisation excluding
+The reported sizes account for the source code of the formalisations excluding
 examples and any safety result, if present. We have noticed that the
@@
 
-Based on the available data, we can make the following observations.
+Looking on the available data, we can make the following observations.
 %
@@
 
-We found no strong evidence that the intrinsically typed representation of
+There is no strong evidence that the intrinsically typed representation of
 processes reduces the size of the formalisation. As observed in
@@
 definitions are also more involved because they incorporate invariants and the
-proofs of typing preservation. It is plausible to speculate that the effort for
-representing processes, types and typing rules is not substantially different
-overall.
+proofs of typing preservation. We speculate that the effort for representing
+processes, types and typing rules is not substantially impacted overall.
 
@@
 
-Finally, it seems that the use of (explicit) continuations affects the
+Finally, it appears that the use of (explicit) continuations is related to the
 complexity of the formalisation more than anything else. Indeed, the six
@@
 are explicit (\LinearPI, \SCP, \LCC) or ``virtual'' (\GV), while the remaining
-ones are 263kb on average. This observation is in agreement with the
-considerations we made earlier in this paper concerning the difference between
-\emph{updating} bindings versus \emph{removing and adding} bindings in a typing
-context. The use of continuations enables a cleaner management of typing
-contexts since linear channels become true ``use-once'' resources and there is
-no need for strong updates of their type.
-
-% It is also interesting to note that whereby \SCP, which the
-% authors~\citet{SanoKavanaghPientka23} do not relate directly to the linear
-% $\pi$-calculus, relies on explicit continuations to tame the complexity of
-% resource management. Our comparison suggests that this aspect, more than the
-% structural treatment of contexts,
-
-% \citet{SanoKavanaghPientka23} do not relate it directly to the linear
-% $\pi$-calculus.
-
-% più che il trattamento strutturale dei contesti (ed evitare lo splitting) è
-% l'uso delle continuazione a fare la differenza
+ones are 263kb on average. This observation agrees with the considerations made
+earlier in this paper concerning the difference between \emph{updating} bindings
+versus \emph{removing and adding} bindings in a typing context. The use of
+continuations enables a cleaner management of typing contexts since linear
+channels are true ``use-once'' resources and there is no need to update their
+type.

### 2025-12-28 paper/conclusion.tex
tuning related and repo references

@@
 recursive processes (the source code of this extension is publicly available in
-the same repository as the one that contains the formalisation of \LCC). In the
+the same repository as the one that contains the formalisation of \LCC~\cite{PadovaniRaffaelli25}). In the
 future, it would be interesting to formalise the strong normalisation property

### 2025-12-28 paper/introduction.tex
tuning related and repo references

@@
 type system coincides with the proof system of classical linear logic and that
-features \emph{linear channels} instead of sessions. Linear channels are meant
-to be used for a \emph{single} communication. Structured conversations can be
-modeled in \LCC using \emph{explicit continuations} as in the well-known
+features \emph{linear channels} instead of sessions. Unlike session endpoints,
+which can be used multiple times albeit sequentially, linear channels are meant
+to be used \emph{once}. \LCC retains the expressiveness of modeling structured
+conversations using \emph{explicit continuations}, just like in the well-known
 encoding of (binary) sessions into the linear
@@
 %
-Overall, \LCC is nothing but a lower-level, linear version of
-\CP~\cite{Wadler14}, Wadler's calculus of sessions based on classical linear
-logic, such that sessions are encoded instead of being featured natively.
+Overall, \LCC is nothing but a low-level version of \CP~\cite{Wadler14} --
+Wadler's calculus of sessions based on classical linear logic -- such that
+sessions are encoded instead of being featured natively.
 
@@
 The formalisation has been checked with Agda 2.8.0 and the code is available on
-GitHub at \url{https://github.com/boystrange/LinearityChallenge/tree/main/lcp}.
+in a public repository on GitHub~\cite{PadovaniRaffaelli25}.

### 2025-12-28 paper/related.tex
tuning related and repo references

@@
 languages with linear resources. One of the presented case studies is the
-formalisation of a fragment of \GV~\citep{GayVasconcelos10,Wadler14}. The
-separating conjunction introduced in \Cref{sec:context-agda} and the typing of
-the constructors for the representation of processes in \Cref{sec:process-agda}
-have been adapted from this work.
+formalisation of a fragment of \GV~\citep{GayVasconcelos10,Wadler14}. 
+%
+Unlike the other formalisations we are discussing,
+\citet{RouvoetPoulsenKrebbersVisser20} do not define a small-step semantics for
+\GV which explains (in part) the small size of their development. However, since
+their formalisation of \GV is intrinsically typed, the interpreter preserves
+typing and therefore proves a form of subject reduction property.
+%
+The separating conjunction introduced in \Cref{sec:context-agda} and the typing
+of the constructors for the representation of processes in
+\Cref{sec:process-agda} have been adapted from this work of
+\citet{RouvoetPoulsenKrebbersVisser20}.
 

### 2025-12-28 paper/related.tex
revised related

@@
     \newcommand{\textlabel}[1]{\rotatebox{90}{#1}}
-    \caption{\label{tab:comparison} A summary of different formalisations of
+    \caption{\label{tab:comparison} Overview of different formalisations of
     binary session calculi (sizes in kb).}
@@
 
-\Cref{tab:comparison} reports a summary of the aforementioned formalisations
-(sorted by publication date) and also our own. The first five columns identify
-the calculus being formalised specifying its reference paper, the used proof
-assistant and an acronym that gives an idea of the overall flavour of the
-calculus.
+\Cref{tab:comparison} shows an overview of the aforementioned formalisations
+(sorted by publication date) including our own. The first five columns identify
+the calculus being formalised. We provide its reference paper, the language in
+which it is formalised and an acronym that gives an idea of the flavour of the
+calculus. We also specify whether the calculus features \emph{cuts} (that is,
+the combination of restriction and parallel composition that corresponds with
+the cut of linear logic) and continuations.
 %
-The acronyms \CP~\cite{Wadler14,GayVasconcelos25} and
-\GV~\cite{GayVasconcelos10,Wadler14} are well-known in the literature on session
-types. The acronym \SessionPI refers to (variants of) the session-based
-$\pi$-calculus presented by \citet{HondaVasconcelosKubo98} while \LinearPI
-refers to (variants of) the linear
-$\pi$-calculus~\cite{KobayashiPierceTurner99}. Finally, \SCP is the structural
-version of \CP introduced by \citet{SanoKavanaghPientka23} and \Calculus is our
-calculus.
+\CP~\cite{Wadler14,GayVasconcelos25} and \GV~\cite{GayVasconcelos10,Wadler14}
+are well-known acronyms in the literature on session types. \SessionPI refers to
+(variants of) the session-based $\pi$-calculus presented by
+\citet{HondaVasconcelosKubo98} while \LinearPI refers to (variants of) the
+linear $\pi$-calculus~\cite{KobayashiPierceTurner99}. Finally, \SCP is the
+structural version of \CP introduced by \citet{SanoKavanaghPientka23} and
+\Calculus is our calculus.
 %
@@
 (typically, is a strict subset of) the one identified by the acronym and that
-the same acronym may refer to different versions of the same calculus.
-%
-We also indicate whether the calculus features \emph{cuts} (that is, the
-combination of restriction and parallel composition in correspondence with the
-cut of linear logic) and continuations.
+the same acronym may refer to different versions of the same calculus. In
+particular, \GV in a logical setting is described by \citet{Wadler14} but its
+first (non-logical) version is due to \citet{GayVasconcelos10}.
 
@@
 way that each operation acting on a channel $s$ is a function that returns the
-result of the operation (if any) \emph{along with the same channel $s$}. In this
-way, the type of $s$ can be conveniently ``updated'' to take into account the
-effect of the operation. As observed by \citet{Padovani17}, this semantics of
-the communication primitives is virtually indistinguishable from one making use
-of explicit continuation channels.
+result of the operation (if present) \emph{along with the same channel $s$}. In
+this way, the type of $s$ can be conveniently ``updated'' to take into account
+the effect of the operation. As observed by \citet{Padovani17}, this semantics
+of the communication primitives is virtually indistinguishable from one making
+use of explicit continuation channels.
 

### 2025-12-28 paper/Examples.lagda.tex
tuning
remark in example and revised related

@@
 \begin{code}
-!!_ : Proc [ 𝔹 ] → Proc [ 𝔹 ]
-!!_ B = curry∗ cut B ≫ (If False Else True)
+!! : Proc [ 𝔹 ] → Proc [ 𝔹 ]
+!! B = curry∗ cut B ≫ (If False Else True)
 
@@
             curry∗ cut B ≫ $
-            If (curry∗ link ch (< ≫) ch) Else (Drop False)
-A || B   = !! ((!! A) && (!! B))
+            If curry∗ link ch (< ≫) ch Else (Drop False)
+A || B   =  !! (!! A && !! B)
 \end{code}
@@
 The function \AgdaFunction{\$} (defined in Agda's standard library) is just a
-visible function application operator. We use it as a ``separator'' so that we
-can flatten deeply nested expressions and save a bunch of parentheses.
+low-precedence, visible function application operator. We use it as a separator
+to flatten deeply nested expressions and save a bunch of parentheses.
 %
-We have chosen to define the disjunction \AgdaFunction{||} from the conjunction
-\AgdaFunction{\&\&} and negation \AgdaFunction{!!} using De Morgan's laws.
+For the sake of illustration, we have chosen to define the disjunction
+\AgdaFunction{||} from the conjunction \AgdaFunction{\&\&} and negation
+\AgdaFunction{!!} using De Morgan's laws.
 
-If we wish to test our definitions, we can implement a simple evaluator using
-the deadlock freedom property. We have not proved a termination result, but
-since linear logic enjoys cut elimination we can safely annotate the evaluator
-as terminating.
+To test our definitions we implement a simple evaluator using the deadlock
+freedom property. We have not proved a termination result, but since linear
+logic enjoys cut elimination we can safely annotate the evaluator as
+terminating.
 
@@
 
-The notation $\AgdaFunction{\#}~n$ is defined in Agda's standard library and
-creates an element of \AgdaDatatype{Fin} from the natural number $n$.
+The notation $\AgdaFunction{\#}~n$ (defined in Agda's standard library) creates
+an element of \AgdaDatatype{Fin} from the natural number $n$. Here it is used to
+create the de Bruijn index of the type variable $X$.
 %
@@
 
-We define a main process that connects server and client
+To test our definitions we compose client and server in parallel
 \begin{code}
@@
 \AgdaFunction{True} as expected.
+
+\begin{remark}
+    Writing processes in Agda would be easier if we could avoid using
+    \AgdaFunction{curry∗}, namely if the constructors of the data type
+    \AgdaDatatype{Proc} were naturally curried.
+    %
+    This is possible, but only if we are willing to sacrifice some clarity. To
+    illustrate the problem with an example, below is the fully curried
+    constructor \AgdaInductiveConstructor{fork} of a hypothetical data type
+    \AgdaDatatype{Proc'}, obtained by expanding the definition of separating
+    conjunction:
+\begin{code}[hide]
+data Proc' : Context → Set where
+\end{code}
+\begin{code}
+  fork : ∀{A B Γ Δ Θ Θ₁ Θ₂} → Ch (A ⊗ B) Δ → Γ ≃ Δ + Θ →
+         Proc' (A ∷ Θ₁) → Θ ≃ Θ₁ + Θ₂ → Proc' (B ∷ Θ₂) → Proc' (A ⊗ B ∷ Γ)
+\end{code}
+
+    This version of \AgdaInductiveConstructor{fork} is way less readable than
+    the one we gave in \Cref{sec:process-agda} because of the (now visible)
+    context splittings. We can recover some clarity and still obtain a curried
+    version of \AgdaInductiveConstructor{fork} using the magic wand instead of
+    Agda's native arrow:
+\begin{code}[hide]
+data Proc'' : Context → Set where
+\end{code}
+\begin{code}
+  fork : ∀{A B} →
+         ∀[ Ch (A ⊗ B) ⇒ (A ∷_) ⊢ Proc'' ─∗ (B ∷_) ⊢ Proc'' ─∗ Proc'' ]
+\end{code}
+
+    However, now in each constructor we have to make a distinction between the
+    first arrow \AgdaFunction{⇒} and the following magic wands \AgdaFunction{─∗}
+    to account for the appropriate amount of context splittings. We found this
+    formulation of the constructors harder to explain and motivate in
+    \Cref{sec:process-agda}.
+    %
+    In the end, none of the alternative definitions of the data type
+    \AgdaDatatype{Proc} was fully satisfactory and we preferred the cleanest and
+    most readable version at the expense of additional clutter in this section.
+    %
+    \eoe
+\end{remark}

### 2025-12-28 paper/abstract.tex
remark in example and revised related

@@
     %
-    Context splitting, the ``naive'' channel management technique directly
-    related to the way session type theories are usually written with pen and
-    paper, is often considered a hindrance and a notable source of complexity,
-    to the point that several alternative approaches have been recently
-    proposed.
+    Context splitting, the channel management technique directly related to the
+    way session type theories are usually written with pen and paper, is often
+    considered a hindrance and a notable source of complexity, to the point that
+    several alternative approaches have been recently proposed.
     %

### 2025-12-28 paper/related.tex
remark in example and revised related

@@
 various approaches in addressing the linearity challenge because the
-formalisations differ widely for a variety of entangled factors. Also, most
-formalisations predate the publication of the linearity
-challenge~\cite{CarboneEtAl24}, hence they are not necessarily targeted to
-solving that particular problem. However, some of
-them~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,SanoKavanaghPientka23,ZackonEtAl25}
-are explicitly motivated by the complexity arising from the linear usage of
-resources/channels and from context splitting in particular.
-
-Looking on the available data, we can make the following observations.
+formalisations differ widely for a variety of entangled factors.
+%
+% Also, most formalisations predate the publication of the linearity
+% challenge~\cite{CarboneEtAl24}, hence they are not necessarily targeted to
+% solving that particular problem. However, some of
+% them~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,SanoKavanaghPientka23,ZackonEtAl25}
+% are explicitly motivated by the complexity arising from the linear usage of
+% resources/channels and from context splitting in particular.
+%
+Looking at the available data, we can make the following observations.
 %
@@
 
-The two largest formalisations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are the
-ones that adopt a locally nameless representation of channels. In these
+The two largest formalisations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are
+also the ones that adopt a locally nameless representation of channels. In these
 formalisations channels are represented in two different ways, depending on
-whether they are free or bound. This necessarily entails some duplication of
-effort as well as some transformation machinery between the two representations
-that contribute to the complexity of the formalisation.
+whether they are free or bound. This entails some duplication of effort as well
+as some transformation machinery between the two representations.
 %

### 2025-12-28 paper/Examples.lagda.tex
tuning

@@
 
-    However, now in each constructor we have to make a distinction between the
-    first arrow \AgdaFunction{⇒} and the following magic wands \AgdaFunction{─∗}
-    to account for the appropriate amount of context splittings. We found this
-    formulation of the constructors harder to explain and motivate in
-    \Cref{sec:process-agda}.
+    However, now we have to make a distinction between the first arrow
+    \AgdaFunction{⇒} and the following magic wands to account for the
+    appropriate amount of context splittings. We found this formulation of the
+    constructors harder to explain and motivate in \Cref{sec:process-agda}.
     %
-    In the end, none of the alternative definitions of the data type
-    \AgdaDatatype{Proc} was fully satisfactory and we preferred the cleanest and
-    most readable version at the expense of additional clutter in this section.
+    Since none of the alternative definitions of \AgdaDatatype{Proc} was fully
+    satisfactory, we preferred the cleanest and most readable version of the
+    data type at the expense of additional clutter in this section.
     %

### 2025-12-28 paper/conclusion.tex
revised conclusions

--- 2025-12-26 paper/conclusion.tex
@@
 \citet[extended version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}.
-%
-Under the assumption that the size of a formalisation is a good measure of its
-complexity, the formalisation of \LCC qualifies as the simplest compared to the
-existing ones. This result should also be weighed considering the richness of
-\LCC in terms of supported features and provable properties.
 
-Our work casts some doubts on the actual contribution of context splitting to
-the complexity of a formalisation and hints instead at explicit continuations in
-the underlying model as a simplifying factor.
-%
-It is quite remarkable that, among the alternative approaches that have been
-proposed to avoid context splitting, the one of \citet{SanoKavanaghPientka23}
-based on linearity predicates makes use of explicit continuations.
-
-The calculus of linear continuations and the calculus described in the linearity
-challenge~\cite{CarboneEtAl24} are incomparable in terms of expressiveness. On
-the one hand, the challenge only considers a minimal calculus of first-order,
-monomorphic sessions while \LCC is a fully-fledged calculus that supports
-linear, shared, higher-order, polymorphic channels. On the other hand, the
-logical foundations of \LCC prevent the modeling of deadlocked processes and of
-sequential processes that own both endpoints of a session. These processes can
-be modeled in the calculus of the challenge, where parallel composition and name
-restriction are distinct constructs as in the $\pi$-calculus.
+The linear calculus of continuations and the calculus described in the linearity
+challenge~\cite{CarboneEtAl24} are incomparable in terms of expressiveness. The
+challenge only considers a minimal calculus of first-order, monomorphic sessions
+while \LCC supports linear, shared, higher-order, polymorphic channels. At the
+same time, the logical foundations of \LCC prevent the modeling of deadlocked
+processes, of sequential processes that own both endpoints of a session and in
+general of cyclic network topologies. Such processes can be modeled in the
+calculus of the challenge, where parallel composition and name restriction are
+distinct constructs as in the $\pi$-calculus.
 %
@@
 
-The simple formalisation of \LCC is a good starting point for further
+Considering the richness of \LCC in terms of features and proved properties, the
+small (and possibly simple) formalisation of \LCC casts some doubts on the
+actual role of context splitting as a source of unnecessary complexity. We
+perceive far greater benefits from the adoption of a model with explicit
+continuations and we find hints of this also beyond our work.
+%
+Among the alternative approaches that have been proposed to overcome the
+difficulties of context splitting, the one of \citet{SanoKavanaghPientka23}
+makes use of explicit continuations.
+%
+In a different setting, the continuation-based encoding of binary sessions has
+greatly simplified the problem of session type reconstruction~\cite{Padovani17}.
+
+The compact formalisation of \LCC is a good starting point for further
 developments. We have already extended \LCC with support for infinite types and
 recursive processes (the source code of this extension is publicly available in
-the same repository as the one that contains the formalisation of \LCC). In the
-future, it would be interesting to formalise the strong normalisation property
-of \LCC as a consequence of cut elimination of classical linear logic.
+the same repository as the one that contains the formalisation of
+\LCC~\cite{PadovaniRaffaelli25}). In the future, it would be interesting to
+formalise the strong normalisation property of \LCC as a consequence of cut
+elimination of classical linear logic.
 

### 2025-12-28 paper/related.tex
revision

@@
 
-We have compared our formalisation with others that have appeared in the
-literature and for which we were able to retrieve the source code. We have
-looked at those formalisations that include the following features:
-\begin{itemize}
-\item reduction/evaluation semantics for a calculus with support for binary
-    sessions, either natively or through their encoding in terms of linear
-    channels~\cite{KobayashiPierceTurner99,Kobayashi02b};
-\item a type system along with the usual typing preservation results;
-\item optionally, a proof of global progress/deadlock freedom.
-\end{itemize}
+We have compared our formalisation with others of typed calculi that support
+binary sessions either natively or through their encoding using
+continuations~\cite{KobayashiPierceTurner99,Kobayashi02b}.
 
@@
 
-\citet{ZackonEtAl25} describe yet another approach for managing typing contexts
-whereby channels are associated not just with a type but also with a \emph{tag}
-(or \emph{multiplicity}) from a given resource algebra. The tag summarises the
-number of allowed usages of a particular channel, including the possibility that
-the channel is not available. This approach streamlines context splitting since
-contexts can be treated in an essentially structural way, except for tags which
-are conveninently combined using operations from the resource algebra.
+\citet{ZackonEtAl25} describe a typing context management technique where
+channels are associated not just with a type but also with an element -- called
+\emph{tag} -- of a given resource algebra that summarises the number of allowed
+usages of a particular channel, including the possibility that the channel is
+not available. This approach streamlines context splitting since contexts can be
+treated in an essentially structural way, except for tags which are
+conveninently combined using operations from the resource algebra.
 
@@
 (sorted by publication date) including our own. The first five columns identify
-the calculus being formalised. We provide its reference paper, the language in
+the calculus being formalised. We provide its reference paper, the prover in
 which it is formalised and an acronym that gives an idea of the flavour of the
 calculus. We also specify whether the calculus features \emph{cuts} (that is,
-the combination of restriction and parallel composition that corresponds with
-the cut of linear logic) and continuations.
+the combination of restriction and parallel composition corresponding with the
+cut of linear logic) and continuations.
 %
@@
 (typically, is a strict subset of) the one identified by the acronym and that
-the same acronym may refer to different versions of the same calculus. In
-particular, \GV in a logical setting is described by \citet{Wadler14} but its
+the same acronym may sometimes refer to different versions of the same calculus.
+In particular, \GV in a logical setting is described by \citet{Wadler14} but its
 first (non-logical) version is due to \citet{GayVasconcelos10}.
@@
 qualitative aspects of the formalisations, namely the management of typing
-contexts, the representation of channels and whether the formalisation is
-intrinsically or extrinsically typed.
+contexts, the representation of channels and whether processes are intrinsically
+or extrinsically typed.
 

### 2025-12-28 paper/calculus.tex
tuning

@@
 The process $\Case\x\z{Q_1}{Q_2}$ offers a choice on channel $x$ and continues
-as either $Q_1$ or $Q_2$ with $z$ bound to the received continuation depending
-on which branch is selected.
+as either $Q_1$ or $Q_2$ with $z$ bound to the received continuation channel
+depending on which branch is selected.
 %
@@
 The notions of free and bound channels are fairly standard, bearing in mind that
-output prefixes bind continuation channels in (some) continuation processes. In
-particular, $\Select\x{\InTag_i}\z.P$ binds $z$ in $P$, while
-$\Fork\x\y\z{P}{Q}$ binds $y$ in $P$ but not in $Q$ and $z$ in $Q$ but not in
-$P$.
+output prefixes bind continuation channels in (some) continuation processes. For
+instance, $\Select\x{\InTag_i}\z.P$ binds $z$ in $P$, while $\Fork\x\y\z{P}{Q}$
+binds $y$ in $P$ but not in $Q$ and $z$ in $Q$ but not in $P$.
 %
@@
 the topmost cut and replace it with zero, one or two new cuts, depending on the
-number of continuations that are exchanged. Note that \RLink eliminates a link
-$\Link\x\y$ by substituting $y$ for $x$ in the scope of $x$ and \RClose models
-the communication of an empty message (without continuations), hence no new cut
-is present in the reduct. \RSelect, \RFork and \RConnect are standard and
-\RExists models the instantiation of a polymorphic variable $X$ as the
+number of continuation channels that are exchanged. Note that \RLink eliminates
+a link $\Link\x\y$ by substituting $y$ for $x$ in the scope of $x$ and \RClose
+models the communication of an empty message (without continuations), hence no
+new cut is present in the reduct. \RSelect, \RFork and \RConnect are standard
+and \RExists models the instantiation of a polymorphic variable $X$ as the
 communication of a type $A$. We write $Q\subst{A}\X$ for the process obtained by

### 2025-12-28 paper/conclusion.tex
tuning

@@
 The compact formalisation of \LCC is a good starting point for further
-developments. We have already extended \LCC with support for infinite types and
-recursive processes (the source code of this extension is publicly available in
-the same repository as the one that contains the formalisation of
+developments. We have already extended \LCC with support for coinductive (\ie
+possibly infinite) types and recursive processes (the source code of this
+extension is publicly available in the same repository of
 \LCC~\cite{PadovaniRaffaelli25}). In the future, it would be interesting to
@@
 
-In this work we have focused on models and theories of \emph{binary sessions}
-(those connecting exactly two processes), but there are also formalisations of
+In this work we have focused on models of \emph{binary sessions} (those
+connecting exactly two processes), but there are also formalisations of
 \emph{multiparty sessions}, notably those by \citet{JacobsBalzerKrebbers22a} and

### 2025-12-28 paper/introduction.tex
tuning

@@
 type system coincides with the proof system of classical linear logic and that
-features \emph{linear channels} instead of sessions. Unlike session endpoints,
-which can be used multiple times albeit sequentially, linear channels are meant
-to be used \emph{once}. \LCC retains the expressiveness of modeling structured
-conversations using \emph{explicit continuations}, just like in the well-known
-encoding of (binary) sessions into the linear
+features \emph{linear channels} instead of sessions. While a session endpoint
+can be used \emph{multiple times} (sequentially), linear channels must be used
+\emph{exactly once}. \LCC retains the expressiveness of other session calculi
+using \emph{explicit continuations}, as in the well-known encoding of (binary)
+sessions into the linear
 $\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}:
@@
 Wadler's calculus of sessions based on classical linear logic -- such that
-sessions are encoded instead of being featured natively.
+sessions can be encoded instead of being featured natively.
 

### 2025-12-28 paper/Examples.lagda.tex
tuning

@@
 \begin{remark}
-    Writing processes in Agda would be easier if we could avoid using
-    \AgdaFunction{curry∗}, namely if the constructors of the data type
-    \AgdaDatatype{Proc} were naturally curried.
+    Writing processes in Agda would be more pleasant if the constructors of the
+    data type \AgdaDatatype{Proc} were naturally curried, instead of currying
+    them on demand with \AgdaFunction{curry∗} as we do here.
     %
-    This is possible, but only if we are willing to sacrifice some clarity. To
-    illustrate the problem with an example, below is the fully curried
-    constructor \AgdaInductiveConstructor{fork} of a hypothetical data type
-    \AgdaDatatype{Proc'}, obtained by expanding the definition of separating
-    conjunction:
+    Below is the naturally curried constructor \AgdaInductiveConstructor{fork}
+    of a hypothetical data type \AgdaDatatype{Proc'}, obtained by expanding the
+    definition of separating conjunction:
 \begin{code}[hide]
@@
 
-    This version of \AgdaInductiveConstructor{fork} is way less readable than
-    the one we gave in \Cref{sec:process-agda} because of the (now visible)
-    context splittings. We can recover some clarity and still obtain a curried
-    version of \AgdaInductiveConstructor{fork} using the magic wand instead of
-    Agda's native arrow:
+    This version of \AgdaInductiveConstructor{fork} is fully curried, but also
+    less readable than the one we gave in \Cref{sec:process-agda} because of the
+    (now visible) context splittings. We can recover some clarity and still
+    obtain a curried version of \AgdaInductiveConstructor{fork} using
+    (literally) the magic wand:
 \begin{code}[hide]
@@
 
-    However, now we have to make a distinction between the first arrow
-    \AgdaFunction{⇒} and the following magic wands to account for the
-    appropriate amount of context splittings. We found this formulation of the
-    constructors harder to explain and motivate in \Cref{sec:process-agda}.
+    However, the first arrow must be a plain implication \AgdaFunction{⇒} and
+    not a magic wand to account for the appropriate amount of context
+    splittings. We found this formulation of the constructors harder to explain
+    and motivate in \Cref{sec:process-agda}.
     %
     Since none of the alternative definitions of \AgdaDatatype{Proc} was fully
-    satisfactory, we preferred the cleanest and most readable version of the
-    data type at the expense of additional clutter in this section.
+    satisfactory, we preferred the most elegant version of the data type at the
+    expense of additional clutter in this section.
     %

### 2025-12-28 paper/conclusion.tex
tuning

@@
 while \LCC supports linear, shared, higher-order, polymorphic channels. At the
-same time, the logical foundations of \LCC prevent the modeling of deadlocked
-processes, of sequential processes that own both endpoints of a session and in
-general of cyclic network topologies. Such processes can be modeled in the
-calculus of the challenge, where parallel composition and name restriction are
-distinct constructs as in the $\pi$-calculus.
+same time, the calculus of the challenge allows the modeling of sequential
+processes owning both endpoints of a session and in general of cyclic network
+topologies, none of which can be modeled in \LCC because of its tight
+correspondence with linear logic.
 %
-Because of its logical foundations, we think that \LCC deserves its own space in
-the context of the linearity challenge alongside with (but not in substitution
-of) more traditional session calculi.
+We think that \LCC deserves its own space in the context of the linearity
+challenge alongside with (but not in substitution of) more traditional session
+calculi.
 
 Considering the richness of \LCC in terms of features and proved properties, the
-small (and possibly simple) formalisation of \LCC casts some doubts on the
-actual role of context splitting as a source of unnecessary complexity. We
-perceive far greater benefits from the adoption of a model with explicit
-continuations and we find hints of this also beyond our work.
+simple formalisation of \LCC casts some doubts on the actual role of context
+splitting as a source of complexity. We perceive more tangible benefits from the
+adoption of a calculus with explicit continuations where channels are linear in
+a literal sense.
 %
-Among the alternative approaches that have been proposed to overcome the
-difficulties of context splitting, the one of \citet{SanoKavanaghPientka23}
-makes use of explicit continuations.
-%
-In a different setting, the continuation-based encoding of binary sessions has
-greatly simplified the problem of session type reconstruction~\cite{Padovani17}.
+In this respect, we find it intriguing that, among the alternative approaches
+that have been proposed to overcome the difficulties of context splitting, the
+one by \citet{SanoKavanaghPientka23} makes key use of explicit continuations and
+also that, in a different setting, the continuation-based encoding of binary
+sessions has greatly simplified the problem of session type
+reconstruction~\cite{Padovani17}.
 
@@
 developments. We have already extended \LCC with support for coinductive (\ie
-possibly infinite) types and recursive processes (the source code of this
-extension is publicly available in the same repository of
-\LCC~\cite{PadovaniRaffaelli25}). In the future, it would be interesting to
-formalise the strong normalisation property of \LCC as a consequence of cut
-elimination of classical linear logic.
+possibly infinite) types and recursive processes (this extension is in \LCC's
+public repository~\cite{PadovaniRaffaelli25}). In the future, it would be
+interesting to formalise the strong normalisation property of \LCC as a
+consequence of cut elimination of classical linear logic.
 
@@
 smaller. Whether this is a coincidence or further evidence of the effectiveness
-of the continuation-based approaches is left as future work.
+of the continuation-based approaches is left for future investigations.

### 2025-12-29 paper/Process.lagda.tex
tuning
global revision

@@
 $\Forall\x\X\y.P$. This constructor models the continuation $P$ using
-higher-order abstract syntax: proving that a context $\Context$ satisfies the
-predicate
+higher-order abstract syntax (HOAS): proving that a context $\Context$ satisfies
+the predicate
 \[
@@
 \]
-applied to $\Context$. Note that the side condition $X \not\in \fv\Context$ of
-\ForallRule is trivially enforced \emph{by definition}:
+applied to $\Context$. Note that the side condition $X \not\in \fv\Context$
+of\Luca{Da definire} \ForallRule is trivially enforced \emph{by definition}:
 %

### 2025-12-29 paper/calculus.tex
global revision

--- 2025-12-26 paper/calculus.tex
@@
 In this section we give a cursory presentation of \Calculus starting from its
-types (\Cref{sec:types}), which coincide with classical linear logic
-propositions, then moving on to the syntax of processes (\Cref{sec:syntax}),
-their reduction semantics (\Cref{sec:semantics}), the typing rules
-(\Cref{sec:typing-rules}) and the formulation of the properties ensured by the
-type system (\Cref{sec:properties}).
+types (\Cref{sec:types}) then moving on to the syntax of processes
+(\Cref{sec:syntax}), their reduction semantics (\Cref{sec:semantics}), the
+typing rules (\Cref{sec:typing-rules}) and the formulation of the properties
+ensured by the type system (\Cref{sec:properties}).
 
@@
 %
+The additive constants describe unusable channels. They can play the role of
+smallest/largest type in type systems that support a notion of
+subtyping~\citep{HornePadovani24}, but will mostly ignore them in this work.
+%
 The additive connectives $A \With B$ and $A \Plus B$ describe  channels used for
@@
 $B$.
-%
-The additive constants can play the role of smallest/largest type in type
-systems that support a notion of subtyping~\citep{HornePadovani24}. We will not
-be using them in this work.
 %
@@
 The notions of \emph{free type variables}, of \emph{duality} and of \emph{type
-substitution} are standard~\cite{Wadler14}. In particular, we write $\fv{A}$ for
-the set of type variables occurring free in $A$, we write $\dual{A}$ for the
-dual of $A$ and $A\subst{B}X$ for the type obtained by replacing the free
-occurrences of $X$ in $A$ with $B$.
+substitution} are standard~\cite{Wadler14}. In particular, we write $\dual{A}$
+for the dual of $A$ and $A\subst{B}X$ for the type obtained by replacing the
+free occurrences of $X$ in $A$ with $B$.
 
@@
 message sent on one of the channels is forwarded to the other. As discussed in
-the literature~\cite{LindleyMorris15}, this form is useful for modeling
-higher-order sessions, that is the exchange of an existing channel on another
+the literature~\cite{LindleyMorris15} and illustrated in \Cref{ex:neg}, this
+form is useful for modeling the exchange of an existing channel on another
 channel.
@@
 The processes $\Wait\x.P$ and $\Close\x$ respectively model the input and output
-of an empty message on the channel $x$. The latter process has no continuation,
-while the former process continues as $P$ once the message has been received.
+of an empty message on the channel $x$. The latter process terminates after the
+message has been sent, while the former process continues as $P$ once the
+message has been received.
 %
@@
 The process $\Case\x\z{Q_1}{Q_2}$ offers a choice on channel $x$ and continues
-as either $Q_1$ or $Q_2$ with $z$ bound to the received continuation depending
-on which branch is selected.
+as either $Q_1$ or $Q_2$ depending on which branch is selected with $z$ bound to
+the received continuation channel.
 %
 The process $\Select\x{\InTag_i}\z.P$ performs a choice (represented by a label
-$\InTag_i$ with $i=1,2$) and sends a fresh continuation $z$ on the channel $x$.
+$\InTag_i$ with $i=1,2$) and sends a fresh continuation channel $z$ on the
+channel $x$.
 %
@@
 can use $y$ and $z$ in whatever order. The sender forks into $P$ and $Q$, each
-using $y$ and $z$ respectively. These forms are often used in conjunction with
-links to model higher-order sessions.
+using $y$ and $z$ respectively.
 %
@@
 The notions of free and bound channels are fairly standard, bearing in mind that
-output prefixes bind continuation channels in (some) continuation processes. In
-particular, $\Select\x{\InTag_i}\z.P$ binds $z$ in $P$, while
-$\Fork\x\y\z{P}{Q}$ binds $y$ in $P$ but not in $Q$ and $z$ in $Q$ but not in
-$P$.
+output prefixes bind continuation channels in (some) continuation processes. For
+instance, $\Select\x{\InTag_i}\z.P$ binds $z$ in $P$, while $\Fork\x\y\z{P}{Q}$
+binds $y$ in $P$ but not in $Q$ and binds $z$ in $Q$ but not in $P$.
 %
@@
         \NotP(x,y) \eqdef
-        \BigCase\x{x'}{
+        \Case\x{x'}{
             \Wait{x'}.\FalseP(y)
@@
 The operational semantics of \Calculus is shown in~\Cref{tab:semantics} and is
-given by two relations: a \emph{structural pre-congruence} relation $\pcong$ and
-a \emph{reduction} relation $\red$. Structural pre-congruence relates processes
-that we consider to be essentially indistinguishable. Reduction relates
-processes in which a communication occurs. Let us describe the two relations
+given by two relations: a \emph{structural pre-congruence} relation $\pcong$,
+which relates essentially indistinguishable processes, and a \emph{reduction}
+relation $\red$, which models communications. Let us describe the two relations
 more in detail.
@@
 leave it out in light of the forthcoming formalisation of the calculus, in which
-every simplification is welcome, although standard presentations of classical
-linear logic admit such law~\cite{Wadler14,Doumane17}.
+every simplification is welcome.
 %
-Another rule that deserves attention is \SServer. In this case, a server prefix
-$\Server\y\u$ can be pulled out of a cut only provided that the other process in
-the cut is also a server on the channel restricted by the cut.
+Another rule that deserves attention is \SServer. In this case, a cut can be
+pushed underneath a server prefix $\Server\y\u$ only provided that the other
+process in the cut is also a server on the channel restricted by the cut.
 
-Reduction is defined by the \rulename{r-*} rules.
+Reduction is defined by the \rulename{r-*} rules, most of them coinciding with
+principal cut reductions of linear logic.
 %
-The reduction rules \RLink, \RClose, \RSelect, \RFork, \RExists and \RConnect
-are in one-to-one correspondence with the well-known reductions of session
-calculi and the principal cut reductions of linear logic. These reductions erase
-the topmost cut and replace it with zero, one or two new cuts, depending on the
-number of continuations that are exchanged. Note that \RLink eliminates a link
-$\Link\x\y$ by substituting $y$ for $x$ in the scope of $x$ and \RClose models
-the communication of an empty message (without continuations), hence no new cut
-is present in the reduct. \RSelect, \RFork and \RConnect are standard and
-\RExists models the instantiation of a polymorphic variable $X$ as the
-communication of a type $A$. We write $Q\subst{A}\X$ for the process obtained by
-replacing every free occurrence of the type variable $X$ with $A$.
+The rules \RLink, \RClose, \RSelect, \RFork and \RExists erase the topmost cut
+and replace it with zero, one or two new cuts, depending on the number of
+continuation channels that are exchanged.
 %
-The reduction rules \RWeaken and \RContract respectively represent the disposal
-of an unused server and the duplication of a server. In these rules we use some
-convenient (albeit slightly informal) notation for denoting sequences of
-(pairwise distinct) channels and prefixes. In particular, $\seqof\z$ stands for
-a sequence $z_1, \dots, z_n$ of channels, $\Weaken{\seqof\z}.Q$ stands for
-$\Weaken{z_1}\dots\Weaken{z_n}.Q$ and
-$\Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.R$ stands for
-$\Contract{z_1}{z_1'}{z_1''}\dots\Contract{z_n}{z_n'}{z_n''}.R$.
+Note that the rule \RLink eliminates a link $\Link\x\y$ by substituting $y$ for
+$x$ in the scope of $x$ and \RClose models the communication of an empty message
+(without continuations). The rule \RExists models the instantiation of a
+polymorphic variable $X$ as the communication of a type $A$. We write
+$Q\subst{A}\X$ for the process obtained by replacing every free occurrence of
+the type variable $X$ with $A$.
+%
+The rule \RConnect models the connection between a client and a server, whereas
+the rules \RWeaken and \RContract respectively model the disposal of an unused
+server and the duplication of a server. In these rules we use some slightly
+informal notation for denoting sequences of (pairwise distinct) channels and
+prefixes. In particular, $\seqof\z$ stands for a sequence $z_1, \dots, z_n$ of
+channels, $\Weaken{\seqof\z}.Q$ stands for a sequence
+$\Weaken{z_1}\dots\Weaken{z_n}.Q$ of weakening prefixes and
+$\Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.R$ stands for a sequence
+$\Contract{z_1}{z_1'}{z_1''}\dots\Contract{z_n}{z_n'}{z_n''}.R$ of contraction
+prefixes.
 
@@
     %
-    We illustrate an example of higher-order session by modeling an echo server
-    that consumes boolean values and sends them back unchanged. The protocol
-    implemented by the server is described by the type $\OfCourse(\dual\Bool
-    \Par (\Bool \Ten \One))$ where the modality~$\OfCourse$ indicates that the
-    server is able to accept an unbounded number of requests and the type
-    $\dual\Bool \Par (\Bool \Ten \One)$ describes the sequence of actions
-    performed by the server at each connection with a client: the server first
-    consumes a boolean value (say $u$, of type $\dual\Bool$), then produces
-    another boolean value (say $v$, of type $\Bool$) and finally sends an empty
-    message. We can model the server and a possible client thus:
+    In this example we illustrate the role of links for the communication of
+    free channels by modeling an echo server that consumes boolean values and
+    sends them back unchanged.
+    %
+    The protocol of the server we want to model is described by the type
+    $\OfCourse(\dual\Bool \Par (\Bool \Ten \One))$ where the
+    modality~$\OfCourse$ indicates that the server is able to accept an
+    unbounded number of requests and the type $\dual\Bool \Par (\Bool \Ten
+    \One)$ describes the sequence of actions performed by the server at each
+    connection with a client: the server first consumes a boolean value (say
+    $u$, of type $\dual\Bool$), then produces another boolean value (say $v$, of
+    type $\Bool$) and finally sends an empty message. We can model the server
+    and a possible client thus:
     \begin{align*}
@@
     If we write $\reds$ for the reflexive, transitive closure of $\red$ it is
-    easy to verify that $\Cut\x{\ClientP(x)}{\ServerP(x)} \red^* \TrueP(z)$,
-    as expected.
+    easy to verify that $\Cut\x{\ClientP(x)}{\ServerP(x)} \red^* \TrueP(z)$.
     %
@@
 
-We use typing contexts (\ie sequents) to keep track of the type of channels.
-Typing contexts are finite maps from channels to types written as $x_1 : A_1,
-\dots, x_n : A_n$ and ranged over by $\ContextC$ and $\ContextD$. We write
-$\ContextC, \ContextD$ for the union of $\ContextC$ and $\ContextD$ when they
-have disjoint domains. We write $\WhyNot\Context$ for some context $\Context =
-x_1 : \WhyNot A_1, \dots, x_n : \WhyNot A_n$ where all the types in its range
-are prefixed by the modality~$\WhyNot$. We call these types \emph{unrestricted}
-because they are used to denote shared channels that can be (explicitly)
-discarded and duplicated.
+We use typing contexts (\ie sequents) to keep track of the type of channels in
+processes. Typing contexts are finite maps from channels to types written as
+$x_1 : A_1, \dots, x_n : A_n$ and ranged over by $\ContextC$ and $\ContextD$. We
+write $\ContextC, \ContextD$ for the union of $\ContextC$ and $\ContextD$ when
+they have disjoint domains. We write $\WhyNot\Context$ for some context
+$\Context = x_1 : \WhyNot A_1, \dots, x_n : \WhyNot A_n$ where all the types in
+its range are prefixed by the modality~$\WhyNot$. We call these types
+\emph{unrestricted} because they are used to denote shared channels that can be
+(explicitly) discarded and duplicated.
 
@@
             }
-            ~X\not\in\fn\Context
+            ~X\not\in\fv\Context
             \and
@@
 sessions~\citep{Wadler14,LindleyMorris15,GayVasconcelos25} for the
-interpretation of the rules. The only relevant difference with the typing rules
-in these papers is that the premises mentioning the continuation channel $z$
-actually refer to the same channel $x$ on which the process in the conclusion is
-acting.
+interpretation of the rules. The only relevant difference with our typing rules
+is that the premises mentioning the continuation channel $z$ actually refer to
+the same channel $x$ on which the process in the conclusion is acting.
+%
+The side condition $X \not\in \fv\Context$ in the rule \ExistsRule checks that
+the type variable $X$ does not occur free in the types of $\Context$ and
+therefore can be generalised.
 
@@
 with a process of the form $\Wait\x.P$) and, more generally, it is unable to
-interact regardless of the context in which it occurs because the sub-processes
-it contains are blocked on the channel $x$ that is restricted by the cut.
+interact regardless of the context in which it is placed because the
+sub-processes it contains are blocked on the channel $x$ that is restricted by
+the cut.
 %

### 2025-12-29 paper/introduction.tex
global revision

@@
 The Concurrent Calculi Formalisation Benchmark~\citep{CarboneEtAl24} is a
-collection of challenges concerning the mechanisation of formal models of
+collection of challenges concerning the mechanisation of core models of
 concurrent and distributed programming languages.
@@
 These models often make use of distinctive features that set them apart from the
-formal models of sequential programming languages, such as the adoption of
+models of sequential programming languages, such as the adoption of
 substructural (linear, affine) type systems, the dynamic scope of first-class
@@
 %
-The linearity challenge is motivated by the observation that the proper
-management of linear resources in a formalisation often requires a large number
-of auxiliary definitions and technical results that divert from the main problem
+The linearity challenge is based on the observation that the proper management
+of linear resources in a formalisation often requires a large number of
+auxiliary definitions and technical results that divert from the main problem
 under study~\cite{CarboneEtAl24}. One of the alleged culprits of such complexity
@@
 \emph{exactly once}. \LCC retains the expressiveness of other session calculi
-using \emph{explicit continuations}, as in the well-known encoding of (binary)
-sessions into the linear
-$\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}:
-each message exchanged on a linear channel may include one or two fresh channels
--- the \emph{continuations} -- on which the rest of the conversation takes
-place.
+thanks to \emph{explicit continuations}, which enable the encoding of (binary)
+sessions is terms of linear
+channels~\citep{Kobayashi02b,DardhaGiachinoSangiorgi17}: each message exchanged
+on a linear channel may include one or two fresh channels -- the
+\emph{continuations} -- on which the rest of the conversation takes place.
 %
@@
 %
-Using a calculus based on linear logic prevents \emph{by construction} the same
-(sequential) process to own both endpoints of a session. This situation is
-clearly undesirable since it does not correspond to a useful pattern of
+Working with a calculus based on linear logic prevents \emph{by construction}
+the same (sequential) process to own both endpoints of a session. This situation
+is clearly undesirable since it does not correspond to a useful pattern of
 interaction (every meaningful session requires its endpoints to be used by
@@
 %
-\Cref{sec:agda} presents the Agda formalisation and illustrates its key aspects
-with particular emphasis on the representation of channels and of typing
-contexts. We assume that the reader is somewhat familiar about Agda but we
-recall the lesser known definitions from Agda's standard library.
+\Cref{sec:agda} illustrates the key aspects of the Agda formalisation with
+particular emphasis on the representation of channels and of typing contexts. We
+assume that the reader is somewhat familiar about Agda but we recall the lesser
+known definitions from Agda's standard library.
 %

### 2025-12-29 paper/Context.lagda.tex
global revision

@@
 and \AgdaInductiveConstructor{>} to indicate whether the type is meant to be
-placed in $\ContextD$ or in $\ContextE$.
+placed in $\ContextD$ or in $\ContextE$. Once we reach the end of the typing
+context, we use the constructor \AgdaInductiveConstructor{•} to build the
+trivial splitting of the empty context into two empty partitions.
 %
@@
 
-% It is often the case that the context $\ContextD$ in a splitting
-% $\Splitting\ContextC\ContextD\ContextE$ is a singleton list $[A]$. We use this
-% particular splitting to describe the presence of a channel of type $A$ in
-% $\ContextC$ and giving a name $\ContextE$ to the residual context. For this
-% particular case, we introduce the dedicated notation
-% $\SimpleSplitting\ContextC{A}\ContextE$ which we find clearer.
-
-% \begin{code}
-% _∋_⊳_ : Context → Type → Context → Set
-% Γ ∋ A ⊳ Δ = Γ ≃ [ A ] + Δ
-% \end{code}
 \begin{code}[hide]
@@
 
+The empty context is trivially unrestricted. A non-empty context is unrestricted
+if its head has the form $\WhyNot A$ for some $A$ and provided that its tail is
+unrestricted as well.
+%
 It is easy to prove that $\ContextC$ is unrestricted if so are $\ContextD$ and

### 2025-12-29 paper/agda.tex
LCP => LCC
global revision
tuning related and repo references

@@
 
-In this section we describe the formalisation of \LCP in Agda. We present in
-detail only some key parts of the code including the representation of types and
-processes, the definition of the operational semantics and the statement of the
-main results. The interested reader may find the complete source code in the
-public repository.
+In this section we describe the formalisation of \LCC in Agda. Each of the
+following sub-sections matches one of the modules of the formalisation. We
+present in detail only some key parts of the code including the representation
+of types and processes, the definition of the operational semantics and the
+statement of the main results. The complete source code is available in \LCC's
+public repository~\cite{PadovaniRaffaelli25}.
 

### 2025-12-29 paper/Congruence.lagda.tex
global revision

@@
 proof $\AgdaInductiveConstructor{>}~q$ used in the
-\AgdaInductiveConstructor{wait} process before structural pre-congruence is
-applied. To understand the meaning of this proof, we must recall three key
-elements:
+\AgdaInductiveConstructor{wait} process. To understand the meaning of this
+proof, we must recall three key elements:
 \begin{enumerate}
@@
 $\Select\y\LeftTag\z.P$, is found under a cut for $x : A$ and is using some
-channel $y : B \Plus C$ to send the tag $\InTag_1$.
+channel $y : B \Plus C$ to select $\InTag_1$.
 %
-Unlike \AgdaInductiveConstructor{wait}, which consumes the channel it operates
-on, the continuation process $P$ uses a fresh continuation channel $z : B$.
+The continuation process $P$ uses a fresh continuation channel $z : B$.
 Therefore, $P$ is required to be well typed in the context
 $B~\AgdaInductiveConstructor{∷}~A~\AgdaInductiveConstructor{∷}~\ContextD$, where
-the type $B$ of the continuation $z$ comes \emph{before} the type $A$ of the
-restricted channel $x$.
+the type $B$ of $z$ comes \emph{before} the type $A$ of $x$ since $z$ is
+introduced later than $x$.
 %
@@
 
-Recall from \Cref{tab:semantics} that we want to apply this rule on a process of
-the form $\Cut[A]\x{\Fork\y\u\v{P}{Q}}{R}$ when $x \in \fn{P}$. We capture this
-condition by means of the splitting $\AgdaInductiveConstructor{<}~r$, implying
-that the type $A$ of $x$ ends up in the typing context for $P$ and not in the
-one for $Q$.
+Recall from \Cref{tab:semantics} that we allow using this rule on a process of
+the form $\Cut[A]\x{\Fork\y\u\v{P}{Q}}{R}$ when $x \in \fn{P}$. We capture the
+condition $x\in\fn{P}$ by means of the splitting
+$\AgdaInductiveConstructor{<}~r$, implying that the type $A$ of $x$ ends up in
+the typing context for $P$ and not in the one for $Q$.
 %
@@
 side conditions of the rules in \Cref{tab:semantics}. We should also bear in
-mind that, using an intrinsically-typed representation of processes,
-pre-congruence and reduction rules (which we describe next) incorporate the
-proof that they preserve typing.
+mind that, using an intrinsically-typed representation of processes, we have
+already taken care of the proof that structural pre-congruence preserves typing.

### 2025-12-29 paper/Reduction.lagda.tex
global revision

@@
 
-The splitting $p$ indicates that $y$ (of type $\dual{A}$) occurs on the left
+The splitting $p$ indicates that $y$ (of type $\dual{A}$) occurs in the left
 sub-process of the cut (that is the link $\Link\x\y$) and the splitting
@@
 to which \AgdaInductiveConstructor{link} is applied is structured consistently
-with the syntax of the link $\Link\x\y$ being reduced, which is oriented so that
-the restricted channel $x$ is on the left. 
+with the syntax of the link being reduced, which is oriented so that the
+restricted channel $x$ is on the left. 
 %

### 2025-12-29 paper/related.tex
global revision

@@
 \Cref{thm:type-safety} formalised by \AgdaFunction{type-safety}), but the
-definitions are also more involved because they incorporate invariants and the
-proofs of typing preservation. We speculate that the effort for representing
-processes, types and typing rules is not substantially impacted overall.
+definitions are also more involved because they incorporate invariants and bits
+of the proofs of typing preservation. We speculate that the effort for
+representing processes, types and typing rules is not substantially impacted
+overall.
 

### 2025-12-29 paper/related.tex
final revision?

@@
 \citet{RouvoetPoulsenKrebbersVisser20} do not define a small-step semantics for
-\GV which explains (in part) the small size of their development. However, since
-their formalisation of \GV is intrinsically typed, the interpreter preserves
-typing and therefore proves a form of subject reduction property.
+\GV but their formalisation is intrinsically typed, hence the interpreter
+preserves proves a form of subject reduction property.
 %
-The separating conjunction introduced in \Cref{sec:context-agda} and the typing
-of the constructors for the representation of processes in
-\Cref{sec:process-agda} have been adapted from this work of
-\citet{RouvoetPoulsenKrebbersVisser20}.
+The separating conjunction defined in \Cref{sec:context-agda} and the typing of
+the constructors for the representation of processes in \Cref{sec:process-agda}
+have been adapted from this work of \citet{RouvoetPoulsenKrebbersVisser20}.
 
@@
 \citet{ZackonEtAl25} describe a typing context management technique where
-channels are associated not just with a type but also with an element -- called
-\emph{tag} -- of a given resource algebra that summarises the number of allowed
+channels are associated not just with a type but also with a \emph{tag}, that is
+an element of a given resource algebra that summarises the number of allowed
 usages of a particular channel, including the possibility that the channel is
@@
 calculus. We also specify whether the calculus features \emph{cuts} (that is,
-the combination of restriction and parallel composition corresponding with the
-cut of linear logic) and continuations.
+the combination of restriction and parallel composition corresponding to the cut
+of linear logic) and continuations.
 %
@@
 the predicates that check the linear usage of channels. Finally, all the
-approaches based on \GV are also marked with \MAYBE. In principle, \emph{none}
-of these calculi makes use of continuations. However, \GV is designed in such a
-way that each operation acting on a channel $s$ is a function that returns the
+approaches based on \GV are also marked with \MAYBE. Officially, \emph{none} of
+these calculi makes use of continuations. However, \GV is designed in such a way
+that each operation acting on a channel $s$ is a function that returns the
 result of the operation (if present) \emph{along with the same channel $s$}. In
@@
 rough (and possibly questionable) estimates of their complexity. Papers
-describing formalisations typically use ``lines of code'' as a measure of
+describing formalisations typically report the ``lines of code'' as a measure of
 development effort, but the number of lines may be affected by code indentation
@@
 The reported sizes account for the source code of the formalisations excluding
-examples and any safety result, if present. We have noticed that the
-interpretation of ``safety'' varies widely among the formalisations and, except
-for our own, they all differ from the one stated in the linearity
+examples and any safety result, if present. We have excluded safety results
+because their meaning varies widely across the formalisations and, except for
+our own, they all differ from the one stated in the linearity
 challenge~\citep{CarboneEtAl24}.
@@
 are explicit (\LinearPI, \SCP, \LCC) or ``virtual'' (\GV), while the remaining
-ones are 263kb on average. This observation agrees with the considerations made
-earlier in this paper concerning the difference between \emph{updating} bindings
-versus \emph{removing and adding} bindings in a typing context. The use of
-continuations enables a cleaner management of typing contexts since linear
-channels are true ``use-once'' resources and there is no need to update their
-type.
+ones are 263kb on average. At the very least, the use of continuations enables a
+cleaner management of typing contexts since linear channels are true
+``use-once'' resources and there is no need to update their type.

### 2025-12-31 paper/cover-letter.tex
drafted cover letter

\documentclass[12pt]{letter}

\usepackage{a4wide}

\begin{document}
\begin{letter}{}
\opening{Dear Editor,}

I am submitting a manuscript entitled ``A Continuation-Based Solution of the
Linearity Challenge'' to the Journal of Automated Reasoning.

This paper addresses the \emph{linearity challenge}, which appears in a recently
published collection of problems concerning the formalisation of concurrent and
distributed models of computation. The ultimate goal of the challenge (and, more
generally, of the benchmark) is to ``foster the adoption of proof mechanisation
in future research on concurrency'' pretty much like the POPLMark challenge did
for sequential models of programming languages. The linearity challenge is
specifically concerned with the formalisation of a calculus of sessions, where
session endpoints are linear resources that cannot be duplicated or discarded.

The paper describes a particular solution of the linearity challenge in which we
formalise a low level model of computation whose type system corresponds to the
proof system of classical linear logic and into which sessions can be encoded
instead of being featured natively. The paper includes a thorough comparison
with related formalisations of linear/session calculi. Despite the low level of
abstraction of the calculus and the full range of features supported by its type
system, we show that the calculus allows for a remarkably simple formalisation.

The paper has not been published elsewhere. The formalisation started as a
master thesis project of Claudia Raffaelli, a former student of mine and
co-author of the paper. After her graduation, we kept refining and extending the
formalisation once we realised its potential, and eventually decided to write a
paper to share our results and findings with the community.

Thank you for managing our manuscript and considering it for review.

\closing{Best regards,}

Prof. Luca Padovani \\
Department of Computer Science and Engineering \\
University of Bologna \\
ITALY

\end{letter}
\end{document}
